### 2019-10-17 doc/Casey1.Rmd
added manuscript current draft, set up directory structure

---
title: "Distinct communities of snow algae bloom at different elevations"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
date: "`r format(Sys.time())`"
output: 
  word_document:
    # toc: yes
    pandoc_args:
      - '--lua-filter=lua-filters-master/scholarly-metadata.lua'
      - '--lua-filter=lua-filters-master/author-info-blocks.lua'
    reference_docx: /home/cengstro/projects/sa_rbcl/Casey1/tmp/word-template.docx
  pdf_document: 
    number_sections: yes
    # toc: yes
    # toc_depth: 4
    pandoc_args:
      - '--lua-filter=lua-filters-master/scholarly-metadata.lua'
      - '--lua-filter=lua-filters-master/author-info-blocks.lua'
bibliography: /home/cengstro/ownCloud/mendeley_library/bibtex_by_grp/casey1.bib
abstract: |
  *Abstract*: Snow algae grow in melting summer snow across polar and alpine regions, forming green or red colored blooms on the snow surface which can reduce snow surface albedo and increases snow melt rate. Despite over a century of study, the we know little of the diversity of algae species that cause snow algae blooms. We used light microscopy and high-throughput amplicon sequencing of the plastid *rbc*L gene and 18S rDNA to survey snow algae blooms in the Coast Range of British Columbia, Canada. We found distinct communities at different elevations. Our findings suggest that some species of snow algae are widely dispersed, but environmental conditions likely select which snow algae species will proliferate.

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>
<br>
<br>
<!-- Our findings highlight the need to consider the species composition of snow algae blooms in future research of this poorly understood microbiome. -->

### *Keywords*: snow algae rbcL 18S amplicon microbiome alpine community

### *Run title*: Elevation patterns in snow algae communities

### *To be submitted to*: 1) Frontiers in Micb 2) Env Micb 3) Microbial Ecology

<!-- what we now know: (Old studies suggested... here we are blowing open the door. ) -->
<!-- - there are different microbiomes -->
<!-- - alpine vs subalpine -->
<!-- - communities are not distinct, but a gradient!! -->

<!-- completely novel -->
<!-- - chloromonas krienitzii blooms in snow runnels (more seen below treeline), green below, orange above -->
<!-- - chlainomonas is widespread in all blooms -->
<!-- - muramotoi in North America? (if this is a reliable genetic marker) -->
<!-- - Some unknown Chloromonas OTUs with no match in ref databases -->
<!-- - distinct clade Chlainomonas above treeline -->
<!-- - sphere krienitzii, oval, same? or closely related... -->
<!-- - Chlainomonas more at surface than depth -->

<!-- 18S data -->
<!-- write discussion -->

##### pagebreak

# Introduction {-}

Snow algae are cold-adapted microalgae that bloom in melting snow [@Leya2013]. Red pigments produced by some species of snow algae can color vast areas of polar and alpine snowfields, which decreases snow surface albedo, by one estimate enough to substantially increase the rate of snow melt [@Ganey2017]. Loss of snow cover can contribute to drought [@Staudinger2014;@Barnhart2016] and exposes darker surfaces upon meltout, contributing to snow-albedo feedback [@Dery2007]. Thus, snow algae could play an important role in the cryosphere, yet little is known about what species of algae are present, or differences in community composition between snow algae blooms.

Until recently, snow algae species were defined and identified on the basis of morphology. Morphology can be deceptive when identifying microalgae, as the same species can look completely different from one life stage to the next [@Shah2016;@Salome2019]. In the lab, cultures of *Chloromonas* are all morphologically indistinct green biflagellates, but field samples of this genus are dominated by cells with yellow or orange pigment and have thick cell walls with flanges or spikes [@Matsuzaki2019;@Prochazkova2019a]. Cells formerly defined by morphological similarity can be genetically distinct, for example cell types referred to as *Chloromonas cf. nivalis* are now split into multiple species [@Matsuzaki2019;@Prochazhova2019a].
<!-- based on this paragraph, I would not trust the morphology categories in my thing. Ill take em out anyway, but... chlainomonas is distinct, chloro distinct -->

<!-- get reader up to date on snow algae taxonomy... two out of three are poorly explored -->
To date, genetic studies have identified three genera of algae that cause most snow algae blooms: *Chloromonas*, *Sanguina*, and *Chlainomonas*. Only *Chloromonas* contains cultured representatives, with _ named species on GenBank. *Chloromonas* are not known to produce the red pigment. *Sanguina* was recently designated, and contains just two species, one of which *Sanguina nivaloides* is found in alpine and polar snowfields worldwide, and is likely synonymous with some of *Chlamydomonas cf. nivalis* [@Prochazkova2018c]. *Chlainomonas* are less studied, but appear in field samples as large (up to 40 $\mu$m diameter) quadriflagellate oval red cells. In spite of their striking morphological differences, *rbc*L and 18S phylogeny places them within the *Chloromonas* clade.

<!-- how communities vary from patch to patch. -->

<!-- Much of the recent snow algae literature has focused on the eye-catching red snow prevalent throughout polar and alpine regions [@Segawa2018;@Leya2013].  -->

The habitat distribution of snow algae are not well established. Amplicon high-throughput sequencing (HTS) shows that red snow from sites across the Arctic are dominated by reads with high similarity to *Sanguina*, with low relative abundance of *Chloromonas polyptera*, *Chloromonas cf. nivalis*, and *Raphidonema* [@Lutz2016a]. @Ganey2018 suggests that the red pigment of snow algae may be a special adaptation to drier snow, while green snow is restricted to moister areas, although red snow caused by *Chlainomonas* is only reported from waterlogged snow overlying alpine lakes [@Novis2008;@Prochazkova2018c]. Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a]. In the Arctic, one green snow patch was dominated by *Microglena* [@Lutz2015]. Microscopy suggests differences in the snow algae community above and below treeline: two studies found red snow containing *Sanguina*-like cells to be prevalent in the alpine zone above treeline, while green and orange snows caused by *Chloromonas cf. brevispina* and *Chloromonas cf. nivalis* were prevalent in the subalpine zone below treeline [@Hoham1979a;@Nedbalova2008a].

<!-- clear objectives section -->
 Whether snow algae blooms are typically dominated by a single species, or contain a diversity of species remains an open question, as high-resolution genetic comparison between communities is lacking.

Thus far, efforts to characterize snow algae communities by amplicon high-throughput sequencing (HTS) have been limited by ambiguous or nonexistent reference sequences and high 18S similarity between snow algae species [@Lutz2018]. To robustly compare snow algae community comparison side-by-side we used rbcL and 18S amplicon HTS and microscopy to survey 35 snow algae communities in our local mountains, to see what species were present, to compare community composition between samples, and see how they are distributed across the landscape.

<!-- These blooms are often dominated by red spherical cells, which until recently were referred to in the literature as *Chlamydomonas cf. nivalis*, but recent taxonomy classifies as *Sanguina nivaloides*, although synonymy was not established [@Prochazkova2019].  -->
<!-- What species? Are snow algae communities distinct, or overlapping gradient?  -->

<!-- Snow algae diversity is poorly understood, and whether different species inhabit different niches remains an open question.  -->

<!-- A diversity of algae species and cell morphologies are reported from snow algae blooms, yet little is known about the habitat distribution of snow algae species. -->
<!-- why is this interesting? -->
<!-- make it obvious what the paper is about early as possible -->

<!-- yet community composition is perceived to be relatively homogeneous between sites [@Lutz2016a]; whether there may be different types of snow algae blooms has not been adequately explored. -->

<!-- but their impact on alpine food webs and nutrient cycling has not been explored. -->

<!-- Until recently, snow algae species were defined on the basis of cell morphology. -->

<!-- *Sanguina* currently has only two named species, , and . The competitive exclusion principle suggests that within the melting snowpack microbiome there exist different niches, for example different snow moisture content, shade, or nutrient input from trees [@Hoham1976].  -->

<!-- Chloro polyptera absent in rbcL. could we detect this in our samples by 18S? how similar is it to other Chloromonas? -->

<!-- Red cells of *Chlamydomonas cf. nivalis* can be either cosmopolitan or endemic between the poles [@Segawa2018]. -->

<!-- What species of Chloromonas? (Chlainomonas within)  -->

<!-- What potentially novel species look similar, but are genetically distinct? -->

<!-- Are all snow algae communities the same, or different?  Do the differences in cell morphology correspond to genetically different species? What described species are found in the snow? Which of these may be unknown species that looked similar in previous surveys?  -->
<!-- What do we now know? Chlainomonas hi low, above and below treeline specific otus, Chloro krienitzii is widespread in runnels, esp below treeline, green below turning orange on surface. Sanguina less common below treeline.  -->

<!-- what we now know: (Old studies suggested... here we are blowing open the door. ) -->

<!-- - gradient, dom morph vs mix -->
<!--   - chlainomonas is widespread in all blooms -->
<!--   - X17 -->

<!-- - alpine vs subalpine -->
<!--   - Some unknown Chloromonas OTUs with no match in ref databases -->
<!--   - distinct clade Chlainomonas above treeline -->

<!-- - krienitzii -->
<!--   - in snow runnels (more seen below treeline), green below, orange above -->
<!--   - sphere krienitzii, oval, same? or closely related... -->
<!--   - Chlainomonas more at surface than depth -->

# Results {-}

**A gradient between snow algae communities**. Variation in snow algae communities was primarily driven by the presence or absence of *Sanguina nivaloides*, *Chloromonas krienitzii*, *Chlainomonas rubra*, or *Chlainomonas* OTU3 (Fig. 2a). <!--to use named OTUs, or not. The line drawn by our classifier is also arbitrary, depending on database etc but less so than me circling some dots. But I could make a definition in methods, refer to OTU by the most abundant species in it--> However, most samples contained a mix of species; few samples were relatively "pure". However, sample comparison with weighted UniFrac [@Lozupone2005] found little difference between *Chlainomonas* and *Chloromonas* dominated sites, and the presence or absence of *Sanguina* was the driving factor, which was generally found in higher relative abundance at higher elevation (Fig. 2b).

Although *Chlainomonas* is reported to only grow in slushy snow above alpine tarns [@Novis2008;@Prochazkova2018c], we found *Chlainomonas* was present in low relative abundance at most sites sampled. Only one site high in Chlainomonas was near water (S9), the other Chlainomonas dominated sites were not near water, nor was the snow noticeably mushier than the surrounding snow. This does not rule out the possibility that these sites contained high moisture at some point; further sampling is needed to determine its habitat distribution.

<!-- One *Chlainomonas* dominated sample (S9) was from snow near the edge of a pool, but we also collected *Chlainomonas* at sites where we did not see evidence of water nearby (S13, G16.1), and one from a pink snow runnel (S6) that looked the same as the runnels dominated by *C. krienitzii*. *Chlainomonas* dominated samples appeared indistinguishable to pink snow caused by *Sanguina* to the naked eye, although it was less frequently collected than *Sanguina*. This suggests that *Chlainomonas* could play a role in albedo-driven snow melt previously ascribed to *Chlamydomonas cf. nivalis* or *Sanguina*. *Chlainomonas* was not previously reported to be so prevalent in snow algae blooms of mixed composition. This could be because *Chlainomonas* is not distinguishable from *Chloromonas* with 18S primers in previous alpine HTS surveys [@Brown2016a,@Hamilton2017]. It is not reported from either pole, although this could be due to lack of sampling as well as lack of detect ability with HTS. -->

#

**Different alpine and subalpine snow algae communities**. *Sanguina* was more abundant in alpine sites, as well as one OTU of *Chlainomonas* and two OTUs of *Chloromonas*, while *Chloromonas krienitzii* and Chlainomonas OTU6 were found in high relative abundance below treeline (Fig. 2, 3). This consistent with the microscopy observations of @Hoham1979a and @Nedbalova2008a, who found primarily *Chlamydomonas cf. nivalis* above treeline and *Chloromonas* below treeline. However, in microscopy surveys we infrequently observed *Sanguina* below treeline, and *Chloromonas krienitzii* above treeline. We found higher diversity in samples from higher elevation, both at the ASV level and OTU level<!--Chao1 w fossil.-->, with several OTUs that were not found at lower elevations. *Raphidonema* and related ASVs were restricted to high alpine sites, with the highest relative abundance occurring in samples from August and September. This fits with the interpretation of @Stibal2005, who suggest that *Raphidonema* is an opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow. However, rare taxa should not be discounted as playing an important role in ecosystem functioning [@Jousset2017].

Snowpack above and below treeline are very different habitats, with forest providing shade (and therefore less evaporation). The red pigment of *Sanguina* and *Chlainomonas* (astaxanthin) is likely an adaptation to high light [@Gorton2003;@Holzinger2006], and it has been suggested that *Sanguina* is adapted to the drier snowpacks that form at higher elevations [@Dial2018c].

#

**_Chloromonas krienitzii_ is green below, turns orange at surface**. On initial visits to Chloromonas krienitzii dominated sites, green snow was hidden 2 to 5 cm below the surface of the runnel (Supp. Fig. 4). Green snow was noticeably moister than surrounding snow, in some locations containing ice lenses not found in the surrounding snowpack. By June 6 the surface of the runnel had turned orange while the snow below remained green, which persisted on three subsequent visits until the snow melted. Cell morphology at the surface closely resembled *Chloromonas cf. brevispina* “Trochiscia” [@Matsuzaki2015] (Fig. 4<!-- which panel-->), while green subsurface snow was dominated by green biflagellate or bipolar quadriflagellate cells (Fig. 4). Both green and orange samples were dominated by reads assigned to *C. krienitzii*, with lower relative abundance of *Chlainomonas*, which was more abundant at the surface (Supp. Fig. 3). <!--We were surprised to find that --><!--H5 was dominated by *Chlainomonas* reads, but this sample contained few cells matching this morphology.-->

One possibility is that C. krienitzii cells orange pigmentation is triggered or an adaptation to higher light levels at the surface. Green cells below the surface would be protected from UV radiation<!--cite -->. Possibly, the green cells maintain an optimal light level through phototaxis, similar to marine phytoplankton<!--cite-->. Perhaps higher elevations typically have fewer low angle runnels that would allow a moist enough environment for *C. krienitzii*. It seems unlikely that nutrients from conifer needles would explain this, as several of our *C. krienitzii* dominated sites were in full sunlight, far from any trees.

#

**Snow algae blooms harbor novel diversity**. Several OTUs were not assigned to species level due to low similarity (Fig. 1). Two OTUs of Chloromonas, one OTU of Chlainomonas, and one OTU ("Algae OTU8") that had best BLAST matches to both Trebouxiophyceae and Chlorophyceae algae. Algae OTU8 has a best BLAST score of only 88%; this could represent a free-living algae found in snow, or possibly an endosymbiotic algae from a ciliate, as we noted many ciliates with green organelles in our microscopy samples. Sample X17 contained was dominated by spherical red cells superficially similar to *Sanguina*, this sample was dominated by reads of Chloromonas OTU7 and Algae OTU8. Cells in this sample had thicker cell wall, and some cells were slightly square-shaped (Fig. 4). 18S algae reads from this sample were equivocal, with best BLAST matches to both *Raphidonema* and *Chloromonas*. Currently there are just two species *Chlainomonas* in reference databases, but our results suggest there may be genetically distinct populations *Chlainomonas* at higher elevations (Fig. 1, Supp. Fig. 1, Supp. Fig 2). Although some of the high elevations ASVs were classified as *Chlainomonas rubra*, because we only included this species in our custom snow algae reference this could have biased our classifier, possibly causing over-classification [@Murali2018].

Although our rbcL primers successfully detected distinct OTUs, this is likely an underestimate of the true diversity. We did not detect any clustering within *Sanguina* using either 18S or rbcL, however when we included the homologous rbcL fragment of *S. aurentia* in our analysis it did not form a distinct OTU <!--this is where I am quesy about out ghetto OTU method-->. Furthermore, rbcL is highly differentiated in *Chloromonas* relative to other microalgae [@Nozaki2002], which could partly explain why *Sanguina* is less differentiated than *Chloromonas* or *Chlainomonas* (Supp. Fig. 1 and 2).

<!-- Runnels could provide habitat, e.g. liquid water for motility, nutrient mobilization, alternatively the algae could passively accumulate due to gravity. These green cells could swim up from the base of the snowpack, or be passively washed into the runnels from the surface. We observed some green snow ice lenses below these runnels, which could be due to higher snow moisture in the runnel, or due to the algae themselves through antifreeze production or albedo.  -->

<!-- Although all samples dominated by *C. krienitzii* were taken from elevations near or below treeline, our microscopy observations suggest that *C. krienitzii* blooms infrequently at higher elevations as well. Why *C. krienitzii* dominates below treeline is a mystery, since we observed it in high light clearings.  -->
<!-- Exposure to light can trigger astaxanthin production in Haematococcus pluvialis [@Shah2016], -->

# Discussion {-}

We found snow algae communities were typically dominated by either *Sanguina*, *Chlainomonas* or *Chloromonas*, with secondary abundance of *Chlainomonas* or *Chloromonas* in most samples. Most samples contained a sizable proportion of secondary or tertiary OTUs, although some some contained nearly 100% *Chlainomonas rubra*, *Sanguina nivaloides*, or *Chloromonas krienitzii*. The drivers of these differences between communities remain unclear; it is likely that within the melting snowpack there may be distinct microhabitats that impacts snow algae distribution, with varying water content with topography, different light levels in shade or at depth [@Gorton2007], different daily freeze-thaw cycles, or possibly varying nutrient levels with proximity to vegetation [@Hoham1976] or bedrock [@Kol1968a]. Further work needs to be done to uncover the micro-habitat niches within the snowpack that allow this differentiation, and what novel species the snow microbiome might contain.

<!-- Where to put these? -->
<!-- cross validation -->
<!-- Sample composition varied between our three data sets (rbcL, 18S, and cell count), although some robust correlations were found between them. *Sanguina* relative abundance was strongly correlated between data sets, between rbcL and 18s (Spearman's rho=0.78, p<0.001) and rbcL and "*Sanguina*-like" cell morphologies (Spearman's rho=0.92, p<0.001). *Chlainomonas* was strongly correlated with "*Chlainomonas*-like" cells (Spearman's rho=0.7, p<0.001), but IDTaxa did not identify any *Chlainomonas* in our 18S library (in spite of the fact that we included one sample in our 18S reference database). *Chloromonas krienitzii* was not correlated with cell count data (because green cells), nor any 18S ASVs or OTUs. Koliellaceae was not strongly correlated between 18S and rbcL datasets by either relative abundance data or presence absence. No strong correlations were found between Unk8 and taxa or OTUs in other datasets. Sample S4 was dominated by cell morphologies we classified as *Chloromonas cf. brevispina*, and this sample was dominated by reads best matching *Chloromonas krienitzii*. *Sanguina* and *Trebouxiophyceae* was in proportionally higher among 18S algae reads than in rbcL (respectively 40% vs 25%, and 0.9% vs 2%), while *Chloromonas* was lower in 18S than rbcL *Chloromonas* and *Chlainomonas* combined (24% vs 45%).  -->

<!-- mention our brevispina sample -->
<!-- Cell morphology also suggests rbcL underestimates diversity. Sample S4 was dominated by *C. krienitzii* reads, but the cell morphologies dominant in this samples were similar to *Chloromonas cf. brevispina* Cryocystis-type [@Matsuzaki2015], which is a distinct but closely related species to *C. krienitzii*. Consistent with @Nedbalova2008a, we found these cells in bronze snow in well-shaded forest sites. Likely our short-read rbcL fragment would not offer enough genetic resolution to differentiate between the two, with the genetic "signal" of *C. cf. brevispina* getting lost in the "noise" of haplotype variation in *C. krienitzii*. -->

<!-- primer bias -->
<!-- Primer bias likely impacted our estimates of relative abundance. Samples containing 80-90% *Sanguina* 18S were only 50-60% *Sanguina* rbcL. This is likely because we designed our rbcL primers off of *Chloromonas* and *Chlainomonas*, as *Sanguina* sequences were not available at the time, thus leading to their disproportionate representation. Perhaps the three uppermost outliers in Figure 2 (K7, K19.1, and G16.2) could be because these samples contained very little *Chloromonas* initially. Although *Chloromonas* and *Chlainomonas* may be artificially inflated by rbcL, they are likely a significant secondary component of *Sanguina* blooms, as our 18S data suggests, as well as microscopy. Other studies also found that *Chlamydomonas cf. nivalis* blooms contained lower abundance of *Chloromonas* [@Lutz2016a;@Terashima2017]. Furthermore extraction bias could likewise have resulted in over-representation of *Chloromonas* and *Chlainomonas*, since we observed that *Sanguina* were difficult to lyse by light microscopy cell counts, and contained lower DNA yields per dry weight than samples containing green cells. -->

<!-- Our rbcL amplicon library was dominated by reads assigned to *Chloromonas*, *Sanguina*, and *Chlainomonas* (Fig. 1). The most abundant species were *Chloromonas krienitzii*, *Chlainomonas rubra*, and *Sanguina nivaloides*. <!--exact percentage?--> <!--X% of reads were assigned to genus, and X% of reads were assigned to species. Many ASVs were not assigned to genus level by our taxonomic classifier IDTaxa, but were nevertheless closely related to *Chloromonas* (Fig. 1, Supp. Fig. 1 and 2). Although one OTU "Chloro7" was distant from other *Chloromonas* ASVs (Supp. Fig. 2), best BLAST matches were all *Chloromonas* snow algae (best match 92%). Similarly, Chloro9 had best BLAST matches to Chloromonas, with a best match of 94%. Another OTU "Unk8" was only assigned to domain level by IDTaxa, but top BLAST hits included algae from both Chlamydomonadaceae and Trebouxiophyceae, with a best match of 88% similarity. <!--This cluster did not correlate with any 18S OTUs. However, another unidentified clade (labeled as “Chlorophyceae X” in Figure 1) was highly correlated with an 18S OTU that was similarly placed taxonomically.-->

<!-- Did 18S identify any of the unknown clusters? What was in the unknown clusters? -->

<!-- Three green snow samples from above treeline were dominated by *Chloromonas* reads.  -->
<!-- Two Sanguina dominated samples contained high proportion of smaller orange cells... -->

<!-- move some of this to discussion? -->

 <!--However, the low abundance community was less differentiated between samples<!-- too discussion-y?-->  <!--How to mention the higher diversity at higher elevation? Chao with fossil pkg? There was greater genetic variation between high elevation samples than there was between lower elevation samples (Fig. 2b) this is because there were more samples? more sites? hard to make comparisons. <!-- ... becayse Chloromonas and Chlainomonas are closely related-->

<!-- should I go into Chlain genetics, above and below treeline? morph differences? If I get into the morphology, I am characterising new species. Maybe reviewers will want me to go back and flesh out the details of the unknown species, but for now I won't mention it to provoke their curiosity -->

<!-- todo, get 18S clade -> OTU that correlates to Chloro 7. get correl. -->

<!-- remind of the problem: we have all this diversity, where is it? what are the differences? -->

<!--Chloro3, what to make of this. Are we suggesting it is a single species? (NO)-->
<!-- why do we care if there are new species?-->

<!--minor point, whittle down-->

<!-- Red, orange, or green snow can be caused by different species of snow algae. Much of the snow algae literature has focussed on the eye-catching red snow prevalent throughout polar and alpine regions caused by red spherical cells [@Leya2013;@Segawa2018;@Lutz2016a]. Until recently, these cells were referred to in the literature as *Chlamydomonas cf. nivalis*, but were recently re-classified as *Sanguina nivaloides* [@Prochazkova2019]. Red snow can also be caused by *Chlainomonas*, whose distinctive large (40 $\mu$m diameter) quadriflagellate red cells can be found in waterlogged snow over alpine lakes [@Novis2008;@Prochazkova2018c]. Snow algae that lack secondary pigments can color snow green, which occurs in slushy wet snow in the Arctic and Antarctic [@Fogg1967;@Lutz2015;@Remias2013], or in subalpine forests [@Hoham1979b]. Green snow is now thought to be caused by different genera than in red snow, including *Microglena* [@Lutz2015], and various *Chloromonas* [Hoham1979b;@Remias2013a;@Terashima2017]. Two studies (in USA and Slovakia) found green and orange snow to be more common below treeline, while red snow is prevalent in the alpine [@Hoham1979a;@Nedbalova2008a]. <!-- add element of unknown to this, emphasize how little we know-->

<!-- Although  -->

<!-- Cell morphology can be ambiguous in snow algae taxonomy; different species can look nearly identical, and the same species can look completely different depending on the life stage. Recent molecular work shows that strains formerly lumped together as *Chloromonas. cf. nivalis* contain four genetically-distinct lineages [@Matsuzaki2019], while cells described as *C. cf. brevispina* contain at least two genetically distinct clades [@Matsuzaki2015].  -->

<!-- S.aurantia-->
<!-- sky27 and gar 2 couldbe had fewer Chloromonas, so there were no Chloromonas to preferentially amplify -->
<!-- haplotypes, a single ASV difference may not be biologically meaningful -->
<!--future directions: what is the unknowns? -->
<!-- Below treeline, trees shade snowpack and provide a potential source of nutrients through conifer needles [@Hoham1976]; furthermore the window of opportunity for snow algae growth is much shorter as snow melts out sooner at low elevations. -->

<!-- Problems with snow algae 18S amplicon high-throughput sequencing (HTS). Although these studies report taxonomy to the species level, outdated annotation in reference databases, algorithms prone to overclassification [@Murali2018], lack of cross-referenced marker genes and microscopy samples, primers targeting highly conserved regions, and arbitrary OTU thresholds [@Callahan2017] could have hidden or mislabelled taxa, potentially obscuring variation between samples.  -->
<!-- - Most species of *Chloromonas* are known from a single sample.  -->

<!-- blooms side by side different community, which sample? -->
<!-- rbcL underestimates non Chloromonas, less differentiated within Sanguina -->
 <!-- nes18.01, has chlainomonas, but visually more sanguina. primer bias, check 18s.  -->
 <!-- are the citrine cells sanguina? -->
 <!-- return to this in discussion, we found that green snow near red snow was Chloromonas, but we did see some green cells below the surface in transition to red, similar to reported in Prochazkova2019. Where these cells come from is a enduring mystery, how the bloom is seeded. We observed red snow in patches on the saddles, with no green below. The snow was shallow below the alpine green Chloromonas. -->

# Methods {-}

**Field sampling and microscopy.** We collected snow algae from mountains near Vancouver, British Columbia throughout the summer of 2018. To capture the extent of snow algae diversity in our local mountains we collected as many samples as possible from different elevations, dates, mountains, and micro-habitats within the snow <!--insert suppl table-->. In total we collected 310 colored snow samples from 13 different mountains on 33 different dates from elevations between 880 m and 2150 m above sea level. In early season we sampled at lower elevations, moving uphill as the snow algae bloom progressed upwards in elevation [@Takeuchi2013]. We observed the first small spots of red snow above treeline on June 20.

We scooped samples from visibly colored snow into 50 mL centrifuge tubes using sterile technique. To prevent melting en route to the lab we stored tubes in a bag of snow. Back in the lab, we melted each sample at room temperature on the bench and removed a 1 mL aliquot for light microscopy. Immediately after, samples were stored at -20 &deg;C for up to eight months until DNA extraction. Most samples were viewed using light microscopy, characterizing the dominant cell morphology in each sample. We prepared a slide of cells fixed in 2% gluteraldehyde, and quantified the most common cell morphologies with a cell count to 100, classifying each cell as either *Sanguina*, *Chloromonas cf. nivalis*, *Chloromonas cf. brevispina*, *Chloromonas krienitzii* *Chlainomonas*, or "other".

#

**DNA extraction and amplicon library preparation.** We selected 35 out of 310 samples for rbcL and 18s high-throughput amplicon sequencing. We chose these samples to represent the variation in date, elevation, geographic location, snow color, micro-habitat, and cell morphology. We freeze-dried these samples for up to 48 hours until until samples appeared completely desiccated. To allow water vapor to escape sample tubes while freeze drying we poked holes in each lid; as a safeguard against cross-contamination from airborne snow algae powder we wrapped tubes individually with paper towels. We mini-pestled 5 mg of sample at room temperature to physically rupture cell walls. Based on our observation that samples dominated by *Sanguina* cell morphologies had lower DNA yields, and had many intact cells following various lysis methods, we used 20 mg of these samples to increase our DNA yield.

To extract DNA from the crushed cells we added 800 mL 1x CTAB extraction buffer [@ColdSp2009], 1% B-mercaptoethanol, 5 $\mu$L each of proteinase K and RNAase A, and incubated these at 65 &deg;C for 30 minutes. We centrifuged samples at 10,000 g for 3 minutes to pellet cell debris, then added 700 $\mu$L of supernatant to an equal volume of 24:1 chloroform:isoamyl alcohol. We inverted samples to mix and centrifuged again at 12,000 g for 10 minutes. We precipitated DNA by transferring the top layer to 700 $\mu$L ice-cold ethanol, which we gently mixed, and spun over Qiagen DNA columns for 30 s at 15,000 g. We washed columns twice with 70% ethanol, and finally dissolved our DNA by spinning with 50 $\mu$L of sterile TE buffer. As a negative control we processed a sterile distilled water sample alongside each batch, treating it exactly the same as the other tubes.

We used rbcL as a marker gene, which offers high resolution between microalgae species [@Zou2016], and has reference data for snow algae available on GenBank. We also sequenced each sample with 18S primers, due to it's coverage across a wide range of taxa, and its ubiquity in reference databases. We designed rbcL primers to target an approximately 400 bp section of this gene, based on 20 snow algae rbcL GenBank sequences from *Chloromonas* and *Chlainomonas* (GenBank accession numbers AB434272.1, LC012752.1, LC012747.1, AF517072.1, LC012738.1, LC012739.1, AB434267.1, EU030690.1, LC360494.1, AJ001878.1, AB022225.1, DQ885964.2, DQ885962.1, AJ001879.1, AB022226.1, AB022530.1, LC012751.1, AB504764.1, EU030689.1, AB101508.1). *Sanguina* sequences were not included because they were not available at the time. We designed primers with PRIME(+) from the GCG package. These primers were rbcL369F (5’-GAA CGT GAC AAA TTA AAC AAA-3’) and rbcL870R (5’-ACC WGA YAD ACG WAG AGC TT-3’). To target 18S we used Euk1181 (5'-TTA ATT TGA CTC AAC RCG GG-3') and Euk1624 (5'-CGG GCG GTG TGT ACA AAG G-3') [@Wang2014].

We constructed our amplicon library using a two-step PCR [@Meyer2010]. In the first PCR we amplified template DNA using our primers attached to a universal adapter, and in the second PCR we re-amplified that product to attach a 6 bp index to the universal adapter at the 3’ end. The first PCR total volume was 25 $\mu$L, consisting of 1 $\mu$L template, 12.5 $\mu$L Q5 high-fidelity 2X MM (New England BioLabs), 1.25 $\mu$L each of forward and reverse primer, and 9 $\mu$L of ddH2O. The second PCR was the same except we reduced our reaction volume to 20 $\mu$L by using only 5 $\mu$L of ddH2O. The cycling conditions were the same for both primer pairs for the first PCR, with an initial denaturation at 98 &deg;C for 30 s, followed by 30 cycles of 98 &deg;C for 5 s, 58 &deg;C for 10 s, and 72 &deg;C for 25 s, with a final extension at 72 &deg;C for 2 minutes. For the second indexing PCR we started with an initial denaturation at 98 &deg;C for 30 s, then 10 cycles of 98 &deg;C for 10 s, 65 &deg;C for 30 s, and 72 &deg;C for 30 s with a final denaturation of 72 &deg;C for 5 min. After each PCR, we purified using Agencourt AMPure XP kit (Beckman Coulter). We quantified final DNA concentration with Qubit (Thermo Fisher), and standardized sample concentration for pooling. The pooled library was then loaded and run on an Illumina MiSeq V3 kit.

#

**Bioinformatic processing.** Reads were demultiplexing using cutadapt [@Martin2011]. Samples were filtered and trimmed, errors removed, dereplicated, pair end reads merged, and chimeras removed in DADA2 [@Callahan2017]. <!-- singletons, failed samples removed, remove sequences based on negative controls, adjust for 0's, -->

We assigned taxonomy using a custom reference databases for both rbcL and 18S, since snow algae are not well-represented in SILVA or PR2, and there is no curated database for rbcL. We made our rbcL database by downloading 30,865 algae rbcL sequences from GenBank with the query “rbcl[gene] AND (algae OR chlorophyta OR trebouxiophyceae) NOT (18s OR ribosomal OR psaB OR atpB) AND 300:2000[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion”. We removed ambiguous snow algae annotations based on the most recent snow algae phylogeny [@Matsuzaki2019]. Because entries labelled as *Chloromonas cf. brevispina* and *Chloromonas cf. nivalis* are polyphyletic we re-labeled these as “unassigned *Chloromonas*”. <!--Where multiple sequences were present for a single species we generated a consensus sequence to avoid imbalance where some groups have far more representatives than others [@Murali2018] --><!--To  re-labeled GenBank accessions LC012733.1 and LC012735.1 as Chloromonas krienitzii, LC012734.1 as Chloromonas “Cryocystis”-type, and LC012744.1 to Chloromonas miwae, and LC012743 and LC012745 to “Chloromonas Group_B” 6`.--> Species designated by morphology alone that were not genetically different from other species in their genera were annotated only to genus level. <!-- all Chlainomonas to genus level only --> We assigned rbcL ASV taxonomy using IDTaxa with threshold = 50.

To assign 18S reads we initially ran our ASVs against SILVA [@Quast2013], and selected ASVs assigned to Chlorophyta. We then ran this subset of ASVs on our custom 18S snow algae database based on the following GenBank query: "(18S OR ribosomal)[gene] AND (chloromonas OR chlainomonas OR sanguina OR raphidonema OR KMY-2018) NOT (rbcL OR psaB OR atpB) AND 300:3500[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion". This database was edited to reflect the most recent snow algae taxonomy and remove ambiguous annotations [@Matsuzaki2019;@Prochazkova2018c]. We assigned taxonomy using IDTaxa with threshold = 50.

We assigned OTUs with t-SNE, based on visually distinct clusters [@Graham-Rowe1999] <!--BibTeX header messed up-->(Fig. 1). We split the largest cluster (containing the vast majority of the *Chloromonas* reads) into three clusters, one containing all ASVs assigned to *Chloromonas krienitzii*, and one containing all ASVs assigned to *Chloromonas muramotoi* and *Chloromonas hohamii*. To highlight our observation that one clade of *Chlainomonas* were only found above treeline (Suppl. Fig. 1), we split *Chlainomonas* into two OTUs. The OTUs produced by this naive method were roughly consistent across different methods of visualization (Suppl. Fig. 1 and 2). <!--Other OTU classifiers such swarm [@Mahe2014] are optimized for 18S data, and we found swarm did a poor job of clustering when visualised with MDS, t-SNE, and phylogenetic trees, with various values of d. <!--and CROP [@Chen2013].-->
<!--OTU clustering with t-SNE: based on poor results with other classifiers, we used tsne. since tsne is known to find patterns where there are none depending on the perplexity, we ran this test with perplexity ranging from 5-100, and found the same large scale clusters each time. -->

**Software information.** All analysis was conducted in R 3.6.1 with tidyverse 1.2.1, with dada2 1.12.1, DECIPHER 2.12.0, Rtsne 0.15, GUniFrac 1.1, and vegan 2.5-6. Phylogenetic analysis was carried out in IQTree [@Nguyen2015] and visualized with ggtree [@Yu2017].

**Data availability.** All raw fastq files are freely available on the European Nucleotide Archive under the project accession PRJEB34539. All scripts are available at cengstro.github.io/projects/sa_rbcl.

##### pagebreak

# Acknowledgments {-}

We wish to thank Leah Tooman (Simon Fraser University) for assistance with sequencing, and Chris Rushton (Simon Fraser University) for assistance with bioinformatics. <!--We also wish to thank Forest Rohwer (UCSD) for the inspiration that led to this project.--> This project was funded with a Sector Innovation Grant from Genome BC (SIP016), and a NSERC Individual Discovery Grant, both awarded to LQ.

# Author Contributions {-}

<!--

When determining authorship the following criteria should be observed:

 - Substantial contributions to the conception or design of the work; or the
   acquisition, analysis, or interpretation of data for the work; AND
 - Drafting the work or revising it critically for important intellectual
   content; AND
 - Final approval of the version to be published ; AND
 - Agreement to be accountable for all aspects of the work in ensuring that
   questions related to the accuracy or integrity of any part of the work are
   appropriately investigated and resolved.

Contributors who meet fewer than all 4 of the above criteria for authorship
should not be listed as authors, but they should be acknowledged.
(http://www.icmje.org/roles_a.html)

-->

CE, LQ, and KY designed this study. Samples collected by CE with assistance from KY and LQ. KY and CE prepped samples for sequencing. All analyses were completed by CE and KY. All authors discussed the results and contributed to the final manuscript. CE wrote the manuscript with major input from LQ and KY.
<!-- The statement about the authors and contributors can be up to several sentences -->
<!-- long, describing the tasks of individual authors referred to by their initials -->
<!-- and should be included at the end of the manuscript before the References -->
<!-- section. -->

# Additional information {-}

### *Supplementary information*.

### *Competing financial interests.* The authors declare that the research was conducted in the absence of any
commercial or financial relationships that could be construed as a potential
conflict of interest.

##### pagebreak

# References {-}
<div id="refs"></div>

# Figures {-}

<br>
<br>

![Figure 1. t-SNE clustering of snow algae rbcL amplicon sequence variants (ASVs). Colored points correspond to genus and species level assignment, grey points were not assigned at or below genus level. Labelled ellipses show OTUs. Perplexity = 30, theta = 0.5](/home/cengstro/projects/sa_rbcl/figs/asv_tsne_fig.png)

<!-- todo: italics, underscores, ChlorKri5, ChlorMura5, put ChloroKri4 on top of Chloro3, so its obvious that these aren't part of both OTUs.-->

##### pagebreak

![Figure 2. Sample ordination plots. A. NMDS comparing sample similarity using OTUs, shaded by elevation. Vectors indicate relative contribution of taxa driving differences between samples. Stress = 0.183. B. NMDS showing weighted UniFrac distances between snow algae samples. Stress = 0.11.](/home/cengstro/projects/sa_rbcl/figs/sample_ordination.png){width=13.01cm, height=18.56cm}

<!-- add dom morph groupings for UniFrac to connect with heatmap.  adjust size.  -->
<!-- change axis labs of b, MDS to NMDS-->

##### pagebreak

![Figure 3. Relative abundance heatmap of snow algae OTUs in each sample. Weight of color corresponds to the relative abundance of that OTU within a sample. Samples (rows) are grouped by the dominant cell morphology within that sample as determined by light microscopy. Within each morphology group samples are ordered by elevation, from low on the bottom to high on top. Sample IDs follow the convention: letters represent the mountain, and the number represents the order in which it was collected. Samples collected on the same day are labelled as .1, .2, and so on.](/home/cengstro/projects/sa_rbcl/figs/heatmap.png){width=12.88cm, height=17.17cm}

<!-- add elev range, date range. dom cell morph text to grey. lump chlainomonas categories (let it be a surprise)-->
<!-- add treeline -->
```{r, echo = F, include=F}
# otu_heat2 %>% 
#   group_by(bloom_type) %>% 
#   summarise(min(elev_m), max(elev_m))
# otu_heat2 %>% 
#   group_by(bloom_type) %>% 
#   summarise(min(date), max(date))
```

##### pagebreak

![Figure 4. Select microphotographs of snow algae. All photos are taken with light microscopy, except (f) which was taken with DIC. Scale bars represent 30 $\mu$m. (a) H5 (b) S10.2. (c) S4. (d) K2. (e) S10.1. (f) G16.1. (g) N12.1. (h) P20.1.](/home/cengstro/projects/sa_rbcl/figs/fig4.png){width=13.51cm, height=17.49cm}
<!-- order by elev, add sax1 -->

<!-- ![Figure 5. Relative abundance barplots comparing community composition in surface orange snow and green snow below in three samples.] -->
<!-- choose photo, microphotos, rethink colors (make consistant w heatmap) -->

##### pagebreak

# Supplementary Data {-}

![Supplementary Figure 1. Maximum likelihood tree (IQTree) of snow algae rbcL ASVs. Colors correspond to OTU. Edges with bootstrap values < 80 are displayed as dotted lines.](/home/cengstro/projects/sa_rbcl/figs/ggtree.png)

##### pagebreak

![Supplementary Figure 2. MDS plot of rbcL ASVs, colored by taxa. Ellipses contain all ASVs within an OTU. Stress =  ](/home/cengstro/projects/sa_rbcl/figs/asv_mds.png)

<!-- redo with NMDS, include stress vals -->
##### pagebreak

![Supplementary Figure 3. Relative abundance barplots comparing community composition in surface orange snow and subsurface green snow. * designates depth and surface samples collected on different days.](/home/cengstro/projects/sa_rbcl/figs/draft1/depth.png)

##### pagebreak

<!-- ![Supplementary Figure 4. Photos showing location of *Chloromonas krienitzii* dominated samples.]() -->

<!-- word to template: *[:alpha:] to \*, and *[:space:] to } -->

### 2019-10-17 doc/tmp/word-template.Rmd
added manuscript current draft, set up directory structure

---
title: "word-template"
author: "Casey Engstrom"
date: "09/09/2019"
output: 
  word_document:
    reference_docx: word-template.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2019-10-17 doc/Casey1.Rmd
intro

@@
 Snow algae are cold-adapted microalgae that bloom in melting snow [@Leya2013].
-Red pigments produced by some species of snow algae can color vast areas of polar and alpine snowfields, which decreases snow surface albedo, by one estimate enough to substantially increase the rate of snow melt [@Ganey2017].
-Loss of snow cover can contribute to drought [@Staudinger2014;@Barnhart2016] and exposes darker surfaces upon meltout, contributing to snow-albedo feedback [@Dery2007].
-Thus, snow algae could play an important role in the cryosphere, yet little is known about what species of algae are present, or differences in community composition between snow algae blooms.
+Red pigments produced by some species of snow algae can color vast areas of polar and alpine snowfields, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017].
+Loss of snow cover can lead to snow drought [@Staudinger2014;@Barnhart2016], and upon meltout exposes darker surfaces below which contributes to snow-albedo feedback [@Dery2007].
+Thus, snow algae could play an important role in cryosphere regulation, yet the species composition within snow algae blooms remains poorly studied.
+Snow algae are the primary producers in a microbiome that includes grazers and heterotrophic fungi and bacteria (ref), and can be considered ecosystem engineers through their impacts on snow melt and by supplying carbon to this poorly understood microbiome.
+Knowing the identity and community composition of the algae themselves is of primary importance.
 
 Until recently, snow algae species were defined and identified on the basis of morphology.
-Morphology can be deceptive when identifying microalgae, as the same species can look completely different from one life stage to the next [@Shah2016;@Salome2019].
-In the lab, cultures of *Chloromonas* are all morphologically indistinct green biflagellates, but field samples of this genus are dominated by cells with yellow or orange pigment and have thick cell walls with flanges or spikes [@Matsuzaki2019;@Prochazkova2019a].
-Cells formerly defined by morphological similarity can be genetically distinct, for example cell types referred to as *Chloromonas cf.
+Morphology can be deceptive when identifying microalgae, as the same species can look completely different from one life stage to the next [@Shah2016;@Salome2019]; conversely, different species can look nearly identical.
+In the lab cultures of *Chloromonas* are green biflagellates, but in the field the same species contain yellow or orange pigment, and thick cell walls with flanges or spikes.
+Different species were thought to have diagnostic traits such as pigments or cell walls, but genetic studies show that morphological similarity may be misleading in this genus; for example, cell types formerly known as *Chloromonas cf.
 nivalis* are now split into multiple species [@Matsuzaki2019;@Prochazhova2019a].
@@
 <!-- get reader up to date on snow algae taxonomy... two out of three are poorly explored -->
-To date, genetic studies have identified three genera of algae that cause most snow algae blooms: *Chloromonas*, *Sanguina*, and *Chlainomonas*.
-Only *Chloromonas* contains cultured representatives, with _ named species on GenBank.
-*Chloromonas* are not known to produce the red pigment.
-*Sanguina* was recently designated, and contains just two species, one of which *Sanguina nivaloides* is found in alpine and polar snowfields worldwide, and is likely synonymous with some of *Chlamydomonas cf.
+To date, genetic studies have identified three main genera of algae that cause snow algae blooms: *Sanguina*, *Chlainomonas*, and *Chloromonas*.
+Neither of the first two have cultured representatives, but both produce the red pigment astaxanthin (refs).
+*Sanguina* was only recently designated and contains just two species, one of which *Sanguina nivaloides* is found in alpine and polar snowfields worldwide, and is likely synonymous with some of *Chlamydomonas cf.
 nivalis* [@Prochazkova2018c].
-*Chlainomonas* are less studied, but appear in field samples as large (up to 40 $\mu$m diameter) quadriflagellate oval red cells.
-In spite of their striking morphological differences, *rbc*L and 18S phylogeny places them within the *Chloromonas* clade.
+*Chlainomonas* also produce red pigment, but are physically much larger than *Sanguina* (up to 40 $\mu$m diameter).
+Two species are described on the basis of morphology, although analysis of the genetic differences between these species is lacking.
+Although *Chlainomonas* are morphologically distinct from all known *Chloromonas*, *rbc*L and 18S phylogeny places them within the *Chloromonas* clade.
+*Chloromonas* are highly diverse, with at least 13 named species [@Matsuzaki2019].
 
@@
 
-The habitat distribution of snow algae are not well established.
-Amplicon high-throughput sequencing (HTS) shows that red snow from sites across the Arctic are dominated by reads with high similarity to *Sanguina*, with low relative abundance of *Chloromonas polyptera*, *Chloromonas cf.
-nivalis*, and *Raphidonema* [@Lutz2016a].
-@Ganey2018 suggests that the red pigment of snow algae may be a special adaptation to drier snow, while green snow is restricted to moister areas, although red snow caused by *Chlainomonas* is only reported from waterlogged snow overlying alpine lakes [@Novis2008;@Prochazkova2018c].
+Whether different species of snow algae are restricted to a particular microhabitat within the melting snowpack remains unresolved.
+Liquid water availability and nutrient limitation likely play a role in determining where snow algae blooms, as water and fertilizer application increased cell counts of *Chlamydomonas cf.
+nivalis* in experimental manipulation plots in Alaska [@Ganey2017].
 Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a].
-In the Arctic, one green snow patch was dominated by *Microglena* [@Lutz2015].
-Microscopy suggests differences in the snow algae community above and below treeline: two studies found red snow containing *Sanguina*-like cells to be prevalent in the alpine zone above treeline, while green and orange snows caused by *Chloromonas cf.
+Two studies using microscopy found that red snow containing *Chlamydomonas cf.
+nivalis* to be prevalent in the alpine zone above treeline, while green and orange snows caused by *Chloromonas cf.
 brevispina* and *Chloromonas cf.
 nivalis* were prevalent in the subalpine zone below treeline [@Hoham1979a;@Nedbalova2008a].
+@Ganey2018 hypothesized the red pigment of snow algae may be a special adaptation to drier snow, while green snow is restricted to snow with higher moisture content; however red snow caused by *Chlainomonas* is only reported from waterlogged snow overlying alpine lakes [@Novis2008;@Prochazkova2018c].
 
-<!-- clear objectives section -->
- Whether snow algae blooms are typically dominated by a single species, or contain a diversity of species remains an open question, as high-resolution genetic comparison between communities is lacking.
+Microscopic observation of snow algae blooms shows snow algae blooms can be either morphologically diverse [@Lutz2018] or highly similar [@Remias2013].
+18S amplicon high-throughput sequencing (HTS) shows that snow algae may contain more diversity than meets the eye: although red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a].
+18S is highly conserved and therefore suited to survey broadly across a variety of taxa (ref); however due to its high similarity between closely related species it is poorly suited to differentiating between closely related taxa.
+The *rbc*L plastid gene coding for the large subunit of RuBisCO is highly differentiated among photosynthetic organisms, and contains reference sequences for snow algae, making it an ideal complement to survey snow algae communities.
 
-Thus far, efforts to characterize snow algae communities by amplicon high-throughput sequencing (HTS) have been limited by ambiguous or nonexistent reference sequences and high 18S similarity between snow algae species [@Lutz2018].
-To robustly compare snow algae community comparison side-by-side we used rbcL and 18S amplicon HTS and microscopy to survey 35 snow algae communities in our local mountains, to see what species were present, to compare community composition between samples, and see how they are distributed across the landscape.
+In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS.
+Our aims were to: 1) see what species were present, 2) compare community composition between samples, 3) and see how they are distributed across the landscape.
 

### 2019-10-20 doc/Casey1.Rmd
re-wrote intro thru line 85

@@
 
-Snow algae are cold-adapted microalgae that bloom in melting snow [@Leya2013].
+Snow algae are cold-adapted microalgae that bloom in melting summer snow [@Leya2013].
 Red pigments produced by some species of snow algae can color vast areas of polar and alpine snowfields, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017].
-Loss of snow cover can lead to snow drought [@Staudinger2014;@Barnhart2016], and upon meltout exposes darker surfaces below which contributes to snow-albedo feedback [@Dery2007].
-Thus, snow algae could play an important role in cryosphere regulation, yet the species composition within snow algae blooms remains poorly studied.
-Snow algae are the primary producers in a microbiome that includes grazers and heterotrophic fungi and bacteria (ref), and can be considered ecosystem engineers through their impacts on snow melt and by supplying carbon to this poorly understood microbiome.
-Knowing the identity and community composition of the algae themselves is of primary importance.
+Depleted snowpack threatens water supplies and can lead to drought [@Staudinger2014;@Barnhart2016], and earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007]; thus, snow algae could play an important role in snow melt regulation.
+Furthermore, snow algae are the primary producers in a microbiome that includes grazers, fungi, and bacteria [@Hoham2001], but whether there are ecological distinctions between types of snow algae blooms has been poorly studied.
+The first step in understanding this microbiome is defining the algae species that are present in snow algae blooms, and their community structure.
 
 Until recently, snow algae species were defined and identified on the basis of morphology.
-Morphology can be deceptive when identifying microalgae, as the same species can look completely different from one life stage to the next [@Shah2016;@Salome2019]; conversely, different species can look nearly identical.
-In the lab cultures of *Chloromonas* are green biflagellates, but in the field the same species contain yellow or orange pigment, and thick cell walls with flanges or spikes.
-Different species were thought to have diagnostic traits such as pigments or cell walls, but genetic studies show that morphological similarity may be misleading in this genus; for example, cell types formerly known as *Chloromonas cf.
+Morphology can be deceptive when identifying microalgae, as the same species can look completely different from one life stage to the next [@Shah2016;@Salome2019].
+In the lab, cultures of *Chloromonas* are green biflagellates, but in the field the same species contain yellow or orange pigment, and thick cell walls with flanges or spikes.
+Conversely, different genetic species can look nearly identical: for example, cell types formerly known as *Chloromonas cf.
 nivalis* are now split into multiple species [@Matsuzaki2019;@Prochazhova2019a].
@@
 To date, genetic studies have identified three main genera of algae that cause snow algae blooms: *Sanguina*, *Chlainomonas*, and *Chloromonas*.
-Neither of the first two have cultured representatives, but both produce the red pigment astaxanthin (refs).
+Neither of the first two have cultured representatives, but both produce the red pigment astaxanthin [@Remias2005;Remias2016].
 *Sanguina* was only recently designated and contains just two species, one of which *Sanguina nivaloides* is found in alpine and polar snowfields worldwide, and is likely synonymous with some of *Chlamydomonas cf.
@@
 *Chlainomonas* also produce red pigment, but are physically much larger than *Sanguina* (up to 40 $\mu$m diameter).
-Two species are described on the basis of morphology, although analysis of the genetic differences between these species is lacking.
-Although *Chlainomonas* are morphologically distinct from all known *Chloromonas*, *rbc*L and 18S phylogeny places them within the *Chloromonas* clade.
-*Chloromonas* are highly diverse, with at least 13 named species [@Matsuzaki2019].
-
+Although *Chlainomonas* look very different from *Chloromonas*, *rbc*L and 18S phylogeny places them within the *Chloromonas* clade [@Novis2008].
+*Chloromonas* are highly diverse, with at least 13 named species [@Matsuzaki2019]; in contrast, only two species are currently described in *Chlainomonas* [@Novis2008].
 <!-- how communities vary from patch to patch. -->
@@
 
-Whether different species of snow algae are restricted to a particular microhabitat within the melting snowpack remains unresolved.
-Liquid water availability and nutrient limitation likely play a role in determining where snow algae blooms, as water and fertilizer application increased cell counts of *Chlamydomonas cf.
-nivalis* in experimental manipulation plots in Alaska [@Ganey2017].
+Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013].
+In this micro-habitat, temperature is nearly constant, but snow moisture, pH, nutrients, and light can differ greatly within the snowpack depending on location of snow and time of year [@Loranger1988;@Kuhn2001;@Gorton2007;@Ganey2017].
+Differences in snow algae community composition could offer insight into the environmental factors that determine snow algae distribution.
 Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a].
 Two studies using microscopy found that red snow containing *Chlamydomonas cf.
-nivalis* to be prevalent in the alpine zone above treeline, while green and orange snows caused by *Chloromonas cf.
+nivalis* was prevalent in the above timberline, while green and orange snows caused by *Chloromonas cf.
 brevispina* and *Chloromonas cf.
-nivalis* were prevalent in the subalpine zone below treeline [@Hoham1979a;@Nedbalova2008a].
-@Ganey2018 hypothesized the red pigment of snow algae may be a special adaptation to drier snow, while green snow is restricted to snow with higher moisture content; however red snow caused by *Chlainomonas* is only reported from waterlogged snow overlying alpine lakes [@Novis2008;@Prochazkova2018c].
+nivalis* were prevalent in conifer forests below timberline [@Hoham1979a;@Nedbalova2008a].
+Differences in snow moisture could influence which species of snow algae grow; several species of snow algae are anecdotally reported to only inhabit very saturated, waterlogged snow [@Lutz2015;@Remias2013;@Novis2008;@Prochazkova2018c], while  *Chlamydomonas cf.
+nivalis* is hypothesized to be specially adapted to well-drained, drier snow [@Ganey2018].
 
-Microscopic observation of snow algae blooms shows snow algae blooms can be either morphologically diverse [@Lutz2018] or highly similar [@Remias2013].
+<!-- Liquid water availability and nutrient limitation likely play a role in determining where snow algae blooms, as both water and NPK fertilizer increased cell counts of *Chlamydomonas cf.
+nivalis* in one study in Alaska .
+-->
+
+<!-- im not doing network analysis, this is community structure -->
+Microscopic observation shows that snow algae blooms can contain a diversity of cell morphologies in the same sample [@Lutz2018], but the genetic community structure of snow algae may not correlate with what is seen under the microscope.
 18S amplicon high-throughput sequencing (HTS) shows that snow algae may contain more diversity than meets the eye: although red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a].
-18S is highly conserved and therefore suited to survey broadly across a variety of taxa (ref); however due to its high similarity between closely related species it is poorly suited to differentiating between closely related taxa.
+18S is highly conserved and therefore suited to survey broadly across a variety of taxa; however, due to its high similarity between closely related species short segments of thie gene are poorly-differentiated among closely related taxa (ref).
 The *rbc*L plastid gene coding for the large subunit of RuBisCO is highly differentiated among photosynthetic organisms, and contains reference sequences for snow algae, making it an ideal complement to survey snow algae communities.
@@
 Our aims were to: 1) see what species were present, 2) compare community composition between samples, 3) and see how they are distributed across the landscape.
+We found ...
 

### 2019-10-20 doc/Casey1.Rmd
asdfasdfasdf

@@
 
-<!-- im not doing network analysis, this is community structure -->
+<!-- im not doing community structure / network analysis (yet) -->
+<!-- what is the point of this para? rbcL, our hot new method, is the shit -->
 Microscopic observation shows that snow algae blooms can contain a diversity of cell morphologies in the same sample [@Lutz2018], but the genetic community structure of snow algae may not correlate with what is seen under the microscope.
@@
 In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS.
-Our aims were to: 1) see what species were present, 2) compare community composition between samples, 3) and see how they are distributed across the landscape.
+Our aims were to: 1) see what species were present, 2) compare community composition between samples.
 We found ...

### 2019-10-24 code/asv_viz.Rmd
transfer files over from old rproj

---
title: "ASV visualization"
---

```{r libs, echo=F}
library(tidyverse)
library(here)
library(DECIPHER)
library(ggforce)
library(Rtsne)
library(ape)
library(ggtree)
library(treeio)
library(dbscan)
library(ggthemes)
library(ggpubr)
# library(plotly)
# library(crosstalk)
# library(DT)
```

```{r set}
theme_set(theme_few())
set.seed(123)
```

```{r read}
rbcl_asv_id_key <- read_csv(here("output","keys","rbcl_asv_key.csv"))
rbcl_rel_abund <- read_csv(here("data","clean","rbcl_rel_abund_clean.csv"))
idtax <- read_csv(here("data","clean","rbcl_taxonomy_clean.csv"))
```

Align rbcL ASVs
```{r align}
# view alignments, make distance matrix
asv.dss <- rbcl_asv_id_key$asv %>%
  DNAStringSet()
names(asv.dss) <- rbcl_asv_id_key$asv_id
asv.aligns <- asv.dss %>% 
  DECIPHER::AlignSeqs(verbose=F)

asv.aligns %>% DECIPHER::BrowseSeqs(highlight = T) # 600 has a ~70 bp insert, 607 ~45 bp deletion, 508 ~110 bp deletion
length(asv.aligns)
```

Make distance matrix
```{r dists}
dists <-  asv.aligns %>% 
  DECIPHER::DistanceMatrix(type="matrix", correction = "JC")
```

t-SNE with distance matrix
```{r tsne}
perplexity <- 30

tsne <- dists %>% 
  Rtsne(pca=F, perplexity = perplexity)

tsne <- tsne$Y %>% 
  as_tibble()

ggplot(tsne, aes(x=V1,y=V2))+
  geom_point()
```

Add taxonomy assignments, see if t-SNE agrees with the reference-based clustering of IDTaxa
```{r genus_tsne}
tsne <- tsne %>% 
  add_column(asv_id = colnames(dists)) %>% # label each point by it's ASV ID
  left_join(idtax, by="asv_id")

ggplot(tsne, aes(x=V1,y=V2))+
  geom_point(aes(color = genus))
```
Seperates out the Sanguia, Chlainomonas. Lumps all Trebouxiophyceae together. Lots of clustering within Chloromonas, some big clusters with only one point assigned to Chloromonas.

check at species level
```{r sp_tsne}
ggplot(tsne, aes(x=V1,y=V2))+
  geom_point(aes(color = species))
```

Make it look nice, show both genus and species level assignments
```{r tsne-plot1}

# make new categorical to tell ggplot what taxonomy to plot
# include both species and genus level assignments
# lump low abundance Trebouxiophyceae
treb <- c("Myrmecia","Trebouxia","Coccomyxa","Dictyochloropsis","Diplosphaera")
tsne <- tsne %>% 
    mutate(display_taxa = case_when(genus %in% treb ~"other Trebouxiophyceae",
                                  genus == "Stichococcus" ~ "Stichococcus",#########not sure why I need this...
                                  id_level == "species" ~ assignment,
                                  id_level == "genus" ~ assignment))

# order
taxa.order <- c("Chloromonas","Chloromonas_muramotoi","Chloromonas_hohamii","Chloromonas_krienitzii","Chloromonas_Group_B","Raphidonema_nivale","Stichococcus","other Trebouxiophyceae","Chlainomonas", "Chlainomonas_rubra","Sanguina","Sanguina_aurantia","Sanguina_nivaloides" )

tsne_plot <- ggplot(tsne, aes(x=V1, y=V2)) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne, is.na(display_taxa)), 
             color = "grey",alpha = 0.8, size = 3) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne, !is.na(display_taxa)), 
             aes(color = display_taxa %>% fct_relevel(taxa.order), alpha = 0.8), size = 3)  
tsne_plot
```

cluster t-SNE into OTUs
```{r otu-cluster}
eps <- 4

tsne.mat <- tsne %>% 
  select(1,2) %>% 
  as.matrix()

dbscan.out <- dbscan::dbscan(tsne.mat, eps=eps) # eps=4 gives best results

tsne <- tsne %>% 
  add_column(otu = as.character(dbscan.out$cluster)) # tsne is in order of asv_id

ggplot(tsne, aes(x=V1, y=V2)) +
  geom_point(aes(color = otu))
```

manually add *Chloromonas krienitzii* cluster, since this is highly abundant, and split *Chlainomonas* into two, based on the observation that the two clades of *Chlainomonas* seperate out by elevation
```{r edit-otus}
tsne <- tsne %>%
  mutate(otu = case_when((V2 < (-5) & V1 < -17) ~ "9", # chlain "high"
                         (V2 < 14  & V2 > 8 & V1 > (-20) & V1 < (-12)) ~ "10",# krienitzii
                         TRUE ~ otu))

tsne %>% 
  select(display_taxa, otu) %>% 
  distinct() %>% 
  arrange(otu)

# name each OTU
otu_names <- tibble(otu = as.character(1:10), otu_name = c("Chloromonas spp. C","Sanguina A","Chlainomonas B", "Chloromonas muramotoi E","Chloromonas G","Chloromonas I","Algae H","Trebouxiophyceae J","Chlainomonas F","Chloromonas krienitzii D"))

tsne <- tsne %>% 
  left_join(otu_names, by="otu")

asv_otu_key <- tsne %>% 
  select(asv_id, otu, otu_name)

tsne_plot <- tsne_plot +
  ggforce::geom_mark_ellipse(data = tsne, aes(fill=otu_name, label=otu_name))

# set

# extract the legend
tsne_plot %>% 
  get_legend() %>% 
  as_ggplot()
ggsave(here("figs","asv-cluster","asv-legend.pdf"), height = 7, width = 3)

# plot and save the main plot
tsne_plot +
  theme(legend.position = "none")
ggsave(here("figs","asv-cluster","asv-cluster.pdf"), height = 5, width = 5)
```

# Phylogenetic tree

Write fasta for IQTree web serv
```{r}
asv.aligns %>% writeXStringSet(here("output","tree","rbcl","rbcl_aligns.fasta"))
```

Run IQtree
```{r}
rbcl.tree <- read.tree(here("output","tree","rbcl","asv613.fasta.treefile"))

idtax %>% filter(species == "Raphidonema_nivale")
# re-root tree
rbcl.tree <- rbcl.tree %>% root(outgroup = "559", resolve.root=T)

rbcl_ggtree <- ggtree(rbcl.tree)

# parse bootstrap values
rbcl_ggtree_data <- rbcl_ggtree$data %>% 
  filter(isTip == FALSE) %>% 
  separate(label, c("boot", "bayes"), sep = "/") %>% #split label into boots and bayes vals
  filter(boot > 80) %>%
  mutate(boot = round(as.numeric(boot), digits = 0),
         insignificant_edge = if_else(boot > 80, "black", "red"))# round boot to nearest integer

# Plot trees
rbcl_ggtree %<+% select(tsne, asv_id, display_taxa) + # attach taxonomic assignment for tip label
  geom_tippoint(aes(color=display_taxa, alpha=0.7), size=3) + # 
  geom_text2(data=rbcl_ggtree_data, aes(label=paste0(boot,";",bayes)),  vjust= 0.01, hjust = 1, size=3) + # add bootstrap vals
  theme(legend.position = c(0.2,0.7), 
        legend.title = element_blank(), 
        legend.key = element_blank())

rbcl_ggtree %<+% asv_otu_key + # attach OTU
  geom_tippoint(aes(color=otu_name, alpha=0.7), size=3) + # 
  geom_text2(data=rbcl_ggtree_data, aes(label=paste0(boot,";",bayes)),  vjust= 0.01, hjust = 1, size=3) + # add bootstrap vals
  theme(legend.position = c(0.2,0.7), 
        legend.title = element_blank(), 
        legend.key = element_blank())

# color edge red if boot < 80
# ggtree_conf <- rbcl_ggtree$data %>% 
#   separate(label, c("boot", "bayes"), sep = "/") %>% #split label into boots and bayes vals
#   mutate(insignificant_edge = case_when((boot < 80 & isTip == FALSE)~"dotted", 
#                                         TRUE~"solid"),
#          asv_id = case_when(isTip == TRUE~boot,
#                             TRUE~NA_character_)) %>% # round boot to nearest integer
#   left_join(otu_tbl, by="asv_id")

# rbcl_ggtree %<+% ggtree_conf +
#   aes(linetype = I(insignificant_edge)) +
#   geom_tippoint(aes(color=otu),alpha=0.7, size=3) + #
#   theme(legend.position = c(0.2,0.7), 
#         legend.title = element_blank(), 
#         legend.key = element_blank()) 
# ggsave("figs/ggtree.pdf", height = 10, width=7.5)
```

```{r output}

write_csv(asv_otu_key, here("output","keys","asv_otu_key.csv"))
```

# TO do
- add MDS
- point size proportional to relative abundance
- ggtree
- show different values of perplexity yield comparable results

### 2019-10-24 doc/Casey1.Rmd
transfer files over from old rproj

@@
 Red pigments produced by some species of snow algae can color vast areas of polar and alpine snowfields, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017].
-Depleted snowpack threatens water supplies and can lead to drought [@Staudinger2014;@Barnhart2016], and earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007]; thus, snow algae could play an important role in snow melt regulation.
-Furthermore, snow algae are the primary producers in a microbiome that includes grazers, fungi, and bacteria [@Hoham2001], but whether there are ecological distinctions between types of snow algae blooms has been poorly studied.
+Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], and loss of snowpack reservoirs can threaten water supplies, leading to drought [@Staudinger2014;@Barnhart2016].
+Thus, snow algae could play an important role in snow melt regulation.
+Snow algae are the primary producers in a microbiome that includes grazers, fungi, and bacteria [@Hoham2001], but whether this microbiome can be considered as a single microbiome, or whether there are distinctions between types of snow algae blooms has been poorly studied.
 The first step in understanding this microbiome is defining the algae species that are present in snow algae blooms, and their community structure.

### 2019-10-24 code/wrangle.R

@@ -1,0 +1,337 @@
+# Wrangle rbcL and 18S ASV tables and taxonomy
+
+library(tidyverse)
+
+
+# data in ------------
+seq.rbcl.2019 <- read.csv("dat/ky_rbcL_seqtab.nochim.txt", sep=" ", stringsAsFactors = F)
+idtax <- read_csv("dat/assignments_9-18.csv") # thresh = 50
+
+asv.tbl.18s <- read.csv("dat/ky_18s_seqtab.nochim_April2019.txt", sep=" ", stringsAsFactors = F)
+s18_taxa_silva <- read_csv("dat/18s_assignments_silva.csv") # thresh 40
+
+field <- read_csv("dat/field_2018.csv")
+cellct <- read_csv("dat/cellct_2018.csv")
+
+
+# rbcL relative abund --------------------
+
+unfilt_rbcl_rel_abund <- seq.rbcl.2019 %>%
+  t() %>% 
+  as_tibble(rownames = "asv") %>% 
+  gather(key = "sample_id", val = "n_reads", -asv)
+
+# format sample ids properly
+unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
+  mutate(sample_id = str_extract(sample_id, ".*(?=_S.*_)") %>% # remove underscores
+           str_replace_all("-","."), # replace hyphen
+         protocol = str_extract(sample_id, "(?<=[:alpha:]{3}18\\..{2}\\.p)[:digit:]"), # parse protocol version used
+         sample_id = str_replace(sample_id, ".p[:digit:]",""), # remove p1
+         protocol = replace_na(protocol, 3)) # replace NA with 3
+
+# compute new variable "rel_abund" 
+############### consider normalizing with metagenomeSeq ( i think this weights for 0s somehow...)
+
+unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
+  group_by(sample_id, protocol) %>% 
+  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
+  ungroup()
+
+# Filter out controls, and samples with low reads ---------------------------
+
+# set aside and remove protocol controls
+protocol.ids <- unfilt_rbcl_rel_abund %>% 
+  select(sample_id, protocol) %>% 
+  distinct() %>% 
+  dplyr::count(sample_id) %>% 
+  filter(n>1) %>% # choose if multiple protocols associated with a sample
+  pull(sample_id)
[…]

### 2019-10-24 code/community_comparison.Rmd
transfer files over from old rproj

---
title: "Community comparison"
output: html_notebook
---

```{r libs, echo=F}
library(tidyverse)
library(GUniFrac)
library(vegan)
library(ggthemes)
```

```{r set}
theme_set(theme_few())
set.seed(123)
```

```{r input, echo=F}
rbcl_rel_abund <- read_csv(here("data","clean","rbcl_rel_abund_clean.csv"))
idtax <- read_csv(here("data","clean","rbcl_taxonomy_clean.csv"))
field <- read_csv(here("data","clean","field_clean.csv"))

rbcl_asv_id_key <- read_csv(here("output","keys","rbcl_asv_key.csv"))
rbcl_asv_otu_key <- read_csv(here("output","keys","asv_otu_key.csv"))
sample_aliases <- read_csv(here("data","sample_aliases.csv"))
rbcl.tree <- read.tree(here("output","tree","rbcl","asv613.fasta.treefile"))
```

Unifrac rbcL
```{r unifrac}
rbcl.asv.rel.abund.wide <- rbcl_rel_abund %>% 
  select(asv_id, sample_id, rel_abund) %>% 
  pivot_wider(names_from = asv_id, values_from = rel_abund, values_fill = list(rel_abund = 0)) %>% 
  column_to_rownames("sample_id")

# rooted tree required, root by one of my Raphidonema asvs
rbcl.tree <- rbcl.tree %>% root(outgroup = "559", resolve.root=T)

unifracs <- rbcl.asv.rel.abund.wide %>% 
  GUniFrac::GUniFrac(rbcl.tree, alpha=c(0, 0.5, 1))
unifrac <- unifracs$unifracs[, , "d_1"]

# plot results as NMDS
nmds <- metaMDS(unifrac, autotransform = F)
nmds # stress
unifrac_plot_data <- nmds %>% 
  scores() %>% 
  as_tibble(rownames = "sample_id") %>% 
  left_join(field, by="sample_id")

# plot unifrac, colored by elevation
uf_plot <- ggplot(unifrac_plot_data, aes(x=NMDS2, y=NMDS1)) +
  geom_point(aes(color=elev_m), size=4) +
  scale_color_gradientn(colours = terrain.colors(8)[1:7]) +
  labs(color="Elevation (m)") +
  geom_text(data = unifrac_plot_data %>%   # add sample ID labels
              left_join(sample_aliases, by="sample_id"),
            aes(label = alias), size=3)  ######### change to alias for final version. edit overlap txt in svg
uf_plot
```

# rbcL relative abundance plot
```{r barplot}
rbcl_stack_barplot_tbl <- rbcl_rel_abund %>% 
  left_join(rbcl_asv_otu_key, by="asv_id") %>% 
  left_join(field, by="sample_id") %>% 
  left_join(sample_aliases, by="sample_id") %>% 
  select(alias, rel_abund, otu_name, elev_m) %>% 
  group_by(alias, otu_name, elev_m) %>% 
  summarise(rel_abund = sum(rel_abund)) %>% 
  ungroup()

rbcl_stack_barplot_tbl %>% distinct(otu_name)

# sanity check samples rel abund sum to 1
rbcl_stack_barplot_tbl %>%
  group_by(alias) %>% 
  summarise(sum(rel_abund))

# order OTUs
# otu.order <- c("Chloromonas", "Sanguina A","Trebouxiophyceae J")

stacked_barplot <- ggplot(rbcl_stack_barplot_tbl, aes(x = alias %>% fct_reorder(elev_m), y = rel_abund, fill=otu_name)) +
  geom_bar(position="stack",stat="identity") +
  coord_flip() +
  ylab("Sample ID") +
  xlab("Relative abundance") +
  labs(fill = "OTU")
stacked_barplot

```

```{r output}
ggsave(plot = uf_plot, filename = here("figs","rbcl_unifrac_nmds.pdf"), height = 6, width = 8)
ggsave(plot = stacked_barplot, filename = here("figs","rbcl_otu_rel_abund.pdf"), height = 6, width = 8)
```

### 2019-10-24 code/map.Rmd
transfer files over from old rproj

---
title: "Snow algae 2018 map"

---

```{r libs}
library(tidyverse)
library(lubridate)
library(ggmap)
library(ggdark) # ggplot themes
library(googleway) # for getting geographic data from google
library(ggsn) # add symbols to plot
```

```{r data_in}
field <- read_csv(here("data","clean","field_clean.csv"))
```

```{r settings}
key="AIzaSyBJsmyAKTkTg9LT-HdGCCdZdbE2Hjwm-L8"
register_google(key = key) 
theme_set(dark_theme_classic())
```

```{r map_setup}
algae_map <- get_googlemap(c(-123.004722, 49.850556), zoom = 7, maptype = "satellite")

# add in Vancouver and Squamish
cities <- c("Vancouver, BC, Canada", "Squamish, BC, Canada")
# get coordinates from google
google_output <- cities %>% 
  map(~googleway::google_geocode(.x, key=key))
glimpse(google_output)

# functions to reach into the list and extract wanted elements
city_name <- function(x) x$results$address_components[[1]][1,2]
lat <- function(x) x$results$geometry$location[1,1]
lon <- function(x) x$results$geometry$location[1,2]

# make tbl with city and coords
city_coords <- google_output %>% {
  tibble(
    name = map_chr(., .f=city_name),
    lat = map_dbl(., .f=lat),
    lon = map_dbl(., .f=lon)
  )
}
```

```{r all_samples_map}
full_map <- ggmap(algae_map, extent = "device") +
  # jitter points to make it easier to read
  geom_jitter(data = field, 
              size = 2, alpha=0.5, 
              width = 0.05, height = 0.05, # specify jitter
              aes(color=elev_m)) +
  scale_color_gradientn(colours = terrain.colors(8)[1:7], limits = c(850,2200)) +
  geom_point(data = city_coords, shape = 15) +
  geom_text(data = city_coords, aes(label = name), hjust = 1.1)+
  labs(title = "Snow algae sample sites, summer 2018", color="Elevation (m)")

full_map

ggsave(filename = here::here("figs","map",paste0("full_map.png")),
       plot = full_map, height = 15, width = 20, units = "cm")
```

```{r plot_function}
plot_map <- function(my_month){
  # input: chr with capitalized complete month name, eg "May","June"
  # output: ggplot showing the location of samples from that month
  filtered_sample_coords <- field %>% 
    filter( lubridate::month(date, label=TRUE, abbr=FALSE) == my_month)

  ggmap(algae_map, extent = "device") +
    # jitter points to make it easier to read
    geom_jitter(data = filtered_sample_coords, 
                size = 2, alpha=0.7, 
                width = 0.05, height = 0.05, # specify jitter
                aes(color=elev_m)) +
    scale_color_gradientn(colours = terrain.colors(8)[1:7], limits = c(850,2200)) +
    geom_point(data = city_coords, shape = 15) +
    geom_text(data = city_coords, aes(label = name), hjust = 1.1)+
    labs(title = my_month, color="Elevation (m)")}

#test
plot_map("May")
```

```{r save_maps}
save_map <- function(my_month, num){
  map <- plot_map(my_month)

  ggsave(filename = here::here("figs","map",paste0("map_",num,"_", my_month,".png")),
       plot = map, height = 15, width = 20, units = "cm")
}

# generate plots
save_map("May",1)
save_map("June",2)
save_map("July",3)
save_map("August",4)
save_map("September",5)

```

### 2019-10-28 doc/Casey1.Rmd
re-wrote abstract and title

@@
 ---
-title: "Distinct communities of snow algae bloom at different elevations"
+title: "Distinct snow algae communities along an elevational gradient"
 author:
@@
 abstract: |
-  *Abstract*: Snow algae grow in melting summer snow across polar and alpine regions, forming green or red colored blooms on the snow surface which can reduce snow surface albedo and increases snow melt rate.
-Despite over a century of study, the we know little of the diversity of algae species that cause snow algae blooms.
-We used light microscopy and high-throughput amplicon sequencing of the plastid *rbc*L gene and 18S rDNA to survey snow algae blooms in the Coast Range of British Columbia, Canada.
-We found distinct communities at different elevations.
-Our findings suggest that some species of snow algae are widely dispersed, but environmental conditions likely select which snow algae species will proliferate.
+  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate.
+Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch within the same mountain, and varies with elevation and date.
+We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada.
+We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m), with a distinct clades of *Chlainomonas* found only in our high elevation sites.
+High elevation samples, which were collected later in the season, countained higher phylogenetic diversity---including several potentially novel clades of algae.
+Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface.
+Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome.
 
@@
 
-### *Run title*: Elevation patterns in snow algae communities
+### *Run title*: Elevational patterns in snow algae communities
 
-### *To be submitted to*: 1) Frontiers in Micb 2) Env Micb 3) Microbial Ecology
+### *To be submitted to*: 1) Nat Comm, 2) Frontiers in Micb 3) Env Micb 4) Microbial Ecology
 
@@
 
-**Different alpine and subalpine snow algae communities**.
+**Snow algae community composition co-varies with date and elevation**.
 *Sanguina* was more abundant in alpine sites, as well as one OTU of *Chlainomonas* and two OTUs of *Chloromonas*, while *Chloromonas krienitzii* and Chlainomonas OTU6 were found in high relative abundance below treeline (Fig.

### 2019-10-29 code/scratch/not_used.Rmd
organized project directory

---
title: "not_used"
author: "Casey Engstrom"
date: "21/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

trouble getting anything useful out of ggsn::scalebar()
```{r}
  ggmap(algae_map, extent = "device") +
    # jitter points to make it easier to read
    geom_jitter(data = field, 
                size = 2, alpha=0.7, 
                width = 0.05, height = 0.05, # specify jitter
                aes(color=elev_m)) +
    scale_color_gradientn(colours = terrain.colors(8)[1:7], limits = c(850,2200)) +
    geom_point(data = city_coords, shape = 15) +
    geom_text(data = city_coords, aes(label = name), hjust = 1.1)+
    labs(title = "my_month", color="Elevation (m)")+
    scalebar(x.min = -123, x.max = -124,
             y.min = 48, y.max = 50,
             dist = 50, dist_unit = "km",
             height = 0.05,
             st.bottom = FALSE, st.color = "white",
             box.color = "white",
             transform = TRUE, model = "WGS84",
             vjust = -0.5, size=0.5)
```

### 2019-10-29 code/wrangle/normalize.rel_abund_tbl.R

@@ -1,0 +1,31 @@
+library(tidyverse)
+library(here)
+library(metagenomeSeq)
+
+seq.rbcl.2019 <- read.csv(here("data","raw","ky_rbcL_seqtab.nochim.txt"), sep=" ", stringsAsFactors = F)
+
+# format for metagenomeSeq
+MR.rbcl <- rbcl_rel_abund %>% 
+  select(-rel_abund) %>% 
+  spread(sample_id, n_reads, fill = 0) %>% # adds 0s back in
+  column_to_rownames("asv_id") %>% 
+  newMRexperiment()
+
+rbcl_norm <- MR.rbcl %>% 
+  cumNorm(p = cumNormStatFast(MR.rbcl)) %>% 
+  MRcounts(norm=T) %>% 
+  as_tibble(rownames = "asv_id") %>% 
+  gather("sample_id","norm_reads", -asv_id)
+
+rbcl_norm <- rbcl_norm %>% 
+  filter(norm_reads!=0) %>% 
+  group_by(sample_id) %>% 
+  mutate(norm_rel_abund = norm_reads/sum(norm_reads))
+
+rbcl_rel_abund %>% 
+  left_join(rbcl_norm, by=c("sample_id","asv_id")) %>% 
+  select(asv_id, sample_id, rel_abund, norm_rel_abund) %>% 
+  mutate(same = if_else(rel_abund == norm_rel_abund, T, F),
+         diff = rel_abund - norm_rel_abund) %>% 
+  arrange(-diff)
+filter(same==F)

### 2019-10-29 code/wrangle/tidy.rbcl_rel_abund.R

@@ -1,0 +1,31 @@
+library(tidyverse)
+library(here)
+library(metagenomeSeq)
+
+seq.rbcl.2019 <- read.csv(here("data","raw","ky_rbcL_seqtab.nochim.txt"), sep=" ", stringsAsFactors = F) ##### KY saved as txt %^&*
+
+# to long format
+unfilt_rbcl_rel_abund <- seq.rbcl.2019 %>%
+  t() %>% 
+  as_tibble(rownames = "asv") %>% 
+  gather(key = "sample_id", val = "n_reads", -asv)
+
+# format sample ids properly
+unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
+  mutate(sample_id = str_extract(sample_id, ".*(?=_S.*_)") %>% # remove underscores
+           str_replace_all("-","."), # replace hyphen
+         protocol = str_extract(sample_id, "(?<=[:alpha:]{3}18\\..{2}\\.p)[:digit:]"), # parse protocol version used
+         sample_id = str_replace(sample_id, ".p[:digit:]",""), # remove p1
+         protocol = replace_na(protocol, 3)) # replace NA with 3
+
+# compute new variable "rel_abund"
+
+
+
+
+
+unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
+  group_by(sample_id, protocol) %>% 
+  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
+  ungroup()
+unfilt_rbcl_rel_abund

### 2019-10-29 code/community_comparison.Rmd
organized project directory

@@
 library(tidyverse)
+library(here)
 library(GUniFrac)
 library(vegan)
+library(vegalite)
+library(fossil)
+library(picante)
+
+# ggplot themes
 library(ggthemes)
+library(ggdark)
 ```
@@
 ```{r set}
-theme_set(theme_few())
-set.seed(123)
+# theme_set(ggdark::dark_theme_classic()) # for presentation figures
+theme_set(theme_classic())
+
 ```
@@
 
+set.seed(123)
 unifracs <- rbcl.asv.rel.abund.wide %>% 
@@
   geom_point(aes(color=elev_m), size=4) +
-  scale_color_gradientn(colours = terrain.colors(8)[1:7]) +
-  labs(color="Elevation (m)") +
-  geom_text(data = unifrac_plot_data %>%   # add sample ID labels
-              left_join(sample_aliases, by="sample_id"),
-            aes(label = alias), size=3)  ######### change to alias for final version. edit overlap txt in svg
+  scale_color_gradientn(colours = terrain.colors(8)[1:8]) +
+  labs(color="Elevation (m)") #+
+  # geom_text(data = unifrac_plot_data %>%   # add sample ID labels
+              # left_join(sample_aliases, by="sample_id"),
+            # aes(label = alias), size=3,
+            # position = position_nudge(y = -0.025))  ######### change to alias for final version. edit overlap txt in svg
 uf_plot
@@
 # rbcL relative abundance plot
-```{r barplot}
-rbcl_stack_barplot_tbl <- rbcl_rel_abund %>% 
+```{r barplot, include=F}
+rbcl_rel_abund_tbl <- rbcl_rel_abund %>%
   left_join(rbcl_asv_otu_key, by="asv_id") %>%
@@
 
-rbcl_stack_barplot_tbl %>% distinct(otu_name)
+rbcl_rel_abund_tbl %>% distinct(otu_name)
 
 # sanity check samples rel abund sum to 1
-rbcl_stack_barplot_tbl %>%
+rbcl_rel_abund_tbl %>%
   group_by(alias) %>%
@@
 # order OTUs
-# otu.order <- c("Chloromonas", "Sanguina A","Trebouxiophyceae J")
+otu.order <- c("Sanguina A","Chloromonas I","Chloromonas G","Chloromonas muramotoi E","Chloromonas krienitzii D","Chloromonas spp. C","Chlainomonas B","Chlainomonas F","Algae H", "Trebouxiophyceae J")
 
-stacked_barplot <- ggplot(rbcl_stack_barplot_tbl, aes(x = alias %>% fct_reorder(elev_m), y = rel_abund, fill=otu_name)) +
+# specify colors for OTUs
+#           Sang         Chlo I    Chl G      Chl mura    Chl kri   chlor sp  chlain B    chlain F   Alg H     Treb J
+colors <- c("#e31a1c", "#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#00BFC4", "#cab2d6", "#6a3d9a", "#fdbf6f", "#ff7f00")
+
+stacked_barplot <- ggplot(rbcl_rel_abund_tbl,
+                          aes(x = alias %>% fct_reorder(elev_m),
+                              y = rel_abund,
+                              fill=otu_name %>% fct_relevel(otu.order))) +
   geom_bar(position="stack",stat="identity") +
   coord_flip() +
-  ylab("Sample ID") +
-  xlab("Relative abundance") +
-  labs(fill = "OTU")
+  xlab("Sample ID") +
+  ylab("Relative abundance") +
+  labs(fill = "OTU") +
+  scale_fill_manual(values = colors)
 stacked_barplot
 
+# elevation sidebar
+ggplot(rbcl_rel_abund_tbl,
+       aes(x = alias %>% fct_reorder(elev_m),
+           y = elev_m,
+           fill = elev_m)) +
+  geom_bar(stat="identity") +
+  coord_flip() +
+  scale_fill_gradientn(colors = terrain.colors(8)[1:7])
 ```
 
+rbcL heatmap
+```{r heatmap, fig.height=6, fig.width=6}
+
+heatmap <- ggplot(rbcl_rel_abund_tbl, aes(y = alias %>% fct_reorder(elev_m),
+                               x = otu_name %>% fct_relevel(otu.order))) +  
+  geom_tile(aes(alpha = rel_abund, fill = otu_name %>% fct_relevel(otu.order)), color = "white") + 
+  scale_alpha(range = c(0.1, 1)) +
+  ylab("Sample ID") +
+  xlab("OTU") +
+  labs(fill = "OTU", alpha="Relative abundance") +
+  scale_x_discrete(position = "bottom") +
+  theme(axis.text.x = element_text(hjust=1,angle=45)) +
+  guides(fill=FALSE)
+heatmap
+```
+
+# Diversity
+
+Chao
+
+```{r chao1}
+# asv-based chao1
+rbcl_rel_abund_matrix <- rbcl_rel_abund %>%
+  pivot_wider(id_cols = -rel_abund, names_from = sample_id, values_from = n_reads, values_fill = list(rel_abund = 0)) %>% 
+  column_to_rownames("asv_id") %>% 
+  as.matrix() %>% 
+  t()
+
+# fossil::chao1(rbcl_rel_abund_matrix)
+
+# re-root tree
+rbcl.tree <- rbcl.tree %>% root(outgroup = "559", resolve.root=T)
+
+faiths <- picante::pd(rbcl_rel_abund_matrix, rbcl.tree) %>% 
+  # attach elevation data
+  rownames_to_column("sample_id") %>% 
+  left_join(field, by="sample_id") %>% 
+  select(sample_id, PD, date, elev_m)
+ # plot PD as function of elevation
+faiths
+
+faith_date_plot <- ggplot(faiths, aes(y=PD, x=date))+
+  geom_point() +
+  geom_smooth(method = "lm")+
+  ylab("Faith's phylogenetic diversity (PD)") +
+  xlab("Date")
+faith_date_plot
+
+lm_out <- lm(PD ~ date + elev_m, data = faiths)
+summary(lm_out)
+
+# just elev
+lm_out1 <- lm(log(PD) ~ elev_m, data = faiths) %>% # log transform to normalize
+  broom::tidy()
+
+elev_slope <- lm_out1 %>% 
+  filter(term == "elev_m") %>% 
+  pull(estimate)
+elev_p <- lm_out1 %>% 
+  filter(term == "elev_m") %>% 
+  pull(p.value)
+
+ggplot(faiths, aes(y=log(PD), x=elev_m))+
+  geom_point() +
+  geom_smooth(method = "lm")+
+  ylab("log Faith's phylogenetic diversity (PD)") +
+  xlab("Elevation (m)")
+
+# just date
+lm_out2 <- lm(PD ~ date, data = faiths) %>% 
+  broom::tidy()
+
+date_slope <- lm_out2 %>% 
+  filter(term == "date") %>% 
+  pull(estimate)
+date_p <- lm_out2 %>% 
+  filter(term == "date") %>% 
+  pull(p.value)
+
+# check resids
+hist(resid(lm_out1))
+
+# calculate co-linearity
+car::vif(lm_out)
+```
+
+Later dates and higher elevations had higher within-sample diversity.
+Phylogenetic diversity (PD) [@Faith1992] increased linearly with date by `r date_slope` per day (p = `r date_p`), and log-transformed PD increased by `r exp(elev_slope)` per metre gain in elevation (p = `r elev_p`).
+(Supplemental Figure X)
+
 ```{r output}
-ggsave(plot = uf_plot, filename = here("figs","rbcl_unifrac_nmds.pdf"), height = 6, width = 8)
-ggsave(plot = stacked_barplot, filename = here("figs","rbcl_otu_rel_abund.pdf"), height = 6, width = 8)
+ggsave(plot = uf_plot, filename = here("figs","rbcl_unifrac_nmds.png"), height = 4, width = 5)
+
+path <- here("figs/heatmap_dark.png")
+ggsave(plot=heatmap, filename=path, height = 8, width = 6)
+
+ggsave(plot=faith_date_plot, filename = here("figs/faith_diversity_date.png"), height = 7, width = 7)
 ```

### 2019-10-29 code/wrangle/wrangle_18s_rel_abund.Rmd
organized project directory

---
title: "Wrangle 18s rel abund data"
output: html_notebook
---

```{r libs, results="hide"}
library(tidyverse)
library(here)
library(metagenomeSeq)
```

Read in file, and view contents

```{r input, results="hide"}
path_18s_ra <- here("data/raw/ky_18s_seqtab.nochim_April2019.txt")
raw_18s_ra <- read.csv(rel_abund_18S_path, sep=" ", stringsAsFactors = F)

# also read in the rbcl rel abund tidy data, since I'm only using the 18s data as a cross reference I can filter by this
path_rbcl_ra <- here("data/clean/rbcl_rel_abund_clean.csv")
clean_rbcl_ra <- read_csv(path_rbcl_ra)
```

Object naming convention: 
- first what the objects is in its local context, then any globel identifiers
- I'll refer to 18S relative abundance table object as "18s_ra"

Rename sample IDs
- remove MiSeq lane identifier ("S" followed by a number)
```{r}
head(raw_18s_ra)
renamed_sids_18s_ra <- raw_18s_ra %>%
  as_tibble(rownames = "sample_id") %>% 
  mutate(sample_id = sample_id %>% 
           str_remove("_S.*_") %>% # remove _S*_ ending
           str_replace_all("-",".")) #replace hyphen with period
head(renamed_sids_18s_ra)
```

Transpose so each ASV a row, each sample a column
```{r}
transposed_18s_ra <- renamed_sids_18s_ra %>% 
  column_to_rownames("sample_id") %>% 
  t() %>% 
  as_tibble(rownames = "asv_sequence")
head(transposed_18s_ra)
```

Remove actual sequences from tbl, set aside in a seperate tbl

First check each ASV unique
```{r asv_seq_key}
selected_asvs <- transposed_18s_ra %>% 
  select(asv_sequence)
distinct_asvs <- transposed_18s_ra %>% 
  distinct(asv_sequence)
identical(selected_asvs, distinct_asvs)
```

Make a key table
```{r}
asv_seq_key_18s <- distinct_asvs %>% 
  rownames_to_column("asv_id") %>% 
  mutate(asv_id = asv_id %>% 
           as.character() %>% 
           str_pad(width=3, pad="0"))
glimpse(asv_seq_key_18s)
```

Remove the sequence from the main table
```{r}
seq_removed_18s_ra <- asv_seq_key_18s %>% 
  left_join(transposed_18s_ra, by="asv_sequence") %>% # left_join ensures that we are matching id with sequence by the master key
  select(-asv_sequence) # remove sequence, put asv_id first

head(seq_removed_18s_ra)
```

Normalize

Do this before we start to filter and select names

Convert to metagemoneSeq format
```{r normalize}
# mr_obj_18s_ra <- seq_removed_18s_ra %>% 
#   column_to_rownames("asv_id") %>% 
#   newMRexperiment()
# 
# mr_obj_18s_ra
```

```{r normalize2}
# 
# # calculate percentile by which to normalize
# p <- cumNormStatFast(mr_obj_18s_ra)
# 
# 
# # normed_18s_ra <- 
#   
# mr_obj_18s_ra %>% 
#   
#   
#   cumNorm(
#   MRcounts(norm=T) %>% 
#   as_tibble(rownames = "asv_id")
# 
# %>% 
#   gather("sample_id","norm_reads", -asv_id)
# 
# rbcl_norm <- rbcl_norm %>% 
#   filter(norm_reads!=0) %>% 
#   group_by(sample_id) %>% 
#   mutate(norm_rel_abund = norm_reads/sum(norm_reads))
# 
# rbcl_rel_abund %>% 
#   left_join(rbcl_norm, by=c("sample_id","asv_id")) %>% 
#   select(asv_id, sample_id, rel_abund, norm_rel_abund) %>% 
#   mutate(same = if_else(rel_abund == norm_rel_abund, T, F),
#          diff = rel_abund - norm_rel_abund) %>% 
#   arrange(-diff)
# filter(same==F)
```

Convert to long format
```{r long}
long_18s_ra <- seq_removed_18s_ra %>% 
  pivot_longer(cols = -asv_id, names_to = "sample_id", values_to = "n_reads") %>% 
  filter(n_reads != 0) # now that its in long format we can remove 0s

glimpse(long_18s_ra)

```

Calculate relative abundance as percentage
```{r}
as_percent_18s_ra <- long_18s_ra %>% 
  group_by(sample_id) %>% 
  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
  ungroup()
glimpse(as_percent_18s_ra)
```

# Set aside negative controls

look for duplicate sample ids
```{r}
as_percent_18s_ra %>% 
  distinct(sample_id)
```

set aside duplicate sample ids
```{r}
duplicate_sample_18s_ra <- as_percent_18s_ra %>%
  filter(sample_id %>% str_detect("gar18.01|sax18.01|sey18.25|sey18.74|sky18.12"))
glimpse(duplicate_sample_18s_ra)
```

remove duplicates from my main analysis tbl
```{r}
no_dups_18s_ra <- as_percent_18s_ra %>% 
  filter(!(sample_id %in% c("gar18.01.p2", "sax18.01.p1", "sey18.25.p1", "sey18.74.p1", "sky18.12.p1")))
glimpse(no_dups_18s_ra)
```

remove the "p" from the sample id
```{r}
sample_id_fixed_18s_ra <- no_dups_18s_ra %>% 
  mutate(sample_id = str_remove(sample_id, "\\.p[:digit:]"))
glimpse(sample_id_fixed_18s_ra)
```

remove negative controls
```{r}
clean_18s_ra <- sample_id_fixed_18s_ra %>% 
  filter(sample_id %in% clean_rbcl_ra$sample_id)

glimpse(clean_18s_ra)
```

Check number of reads per sample
```{r}
clean_18s_ra %>% 
  group_by(sample_id) %>% 
  summarise(sum_n = sum(n_reads)) %>% 
  arrange(sum_n)
```

write clean csv as output
```{r}
clean_18s_ra_path <- here("data/clean/clean_18s_ra.csv")
write_csv(clean_18s_ra, clean_18s_ra_path)
```

### 2019-10-29 manuscript/Casey1.Rmd
organized project directory

--- 2019-10-28 doc/Casey1.Rmd
@@
     pandoc_args:
-      - '--lua-filter=lua-filters-master/scholarly-metadata.lua'
-      - '--lua-filter=lua-filters-master/author-info-blocks.lua'
+      - '--lua-filter=templates/scholarly-metadata.lua'
+      - '--lua-filter=templates/author-info-blocks.lua'
     reference_docx: /home/cengstro/projects/sa_rbcl/Casey1/tmp/word-template.docx
@@
     pandoc_args:
-      - '--lua-filter=lua-filters-master/scholarly-metadata.lua'
-      - '--lua-filter=lua-filters-master/author-info-blocks.lua'
+      - '--lua-filter=templates/scholarly-metadata.lua'
+      - '--lua-filter=templates/author-info-blocks.lua'
 bibliography: /home/cengstro/ownCloud/mendeley_library/bibtex_by_grp/casey1.bib

### 2019-10-29 code/wrangle.R

@@ -1,337 +1,0 @@
-# Wrangle rbcL and 18S ASV tables and taxonomy
-
-library(tidyverse)
-
-
-# data in ------------
-seq.rbcl.2019 <- read.csv("dat/ky_rbcL_seqtab.nochim.txt", sep=" ", stringsAsFactors = F)
-idtax <- read_csv("dat/assignments_9-18.csv") # thresh = 50
-
-asv.tbl.18s <- read.csv("dat/ky_18s_seqtab.nochim_April2019.txt", sep=" ", stringsAsFactors = F)
-s18_taxa_silva <- read_csv("dat/18s_assignments_silva.csv") # thresh 40
-
-field <- read_csv("dat/field_2018.csv")
-cellct <- read_csv("dat/cellct_2018.csv")
-
-
-# rbcL relative abund --------------------
-
-unfilt_rbcl_rel_abund <- seq.rbcl.2019 %>%
-  t() %>% 
-  as_tibble(rownames = "asv") %>% 
-  gather(key = "sample_id", val = "n_reads", -asv)
-
-# format sample ids properly
-unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
-  mutate(sample_id = str_extract(sample_id, ".*(?=_S.*_)") %>% # remove underscores
-           str_replace_all("-","."), # replace hyphen
-         protocol = str_extract(sample_id, "(?<=[:alpha:]{3}18\\..{2}\\.p)[:digit:]"), # parse protocol version used
-         sample_id = str_replace(sample_id, ".p[:digit:]",""), # remove p1
-         protocol = replace_na(protocol, 3)) # replace NA with 3
-
-# compute new variable "rel_abund" 
-############### consider normalizing with metagenomeSeq ( i think this weights for 0s somehow...)
-
-unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
-  group_by(sample_id, protocol) %>% 
-  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
-  ungroup()
-
-# Filter out controls, and samples with low reads ---------------------------
-
-# set aside and remove protocol controls
-protocol.ids <- unfilt_rbcl_rel_abund %>% 
-  select(sample_id, protocol) %>% 
-  distinct() %>% 
-  dplyr::count(sample_id) %>% 
-  filter(n>1) %>% # choose if multiple protocols associated with a sample
-  pull(sample_id)
[…]

### 2019-10-29 doc/lua-filters-master/minted/run_minted_tests.py

@@ -1,522 +1,0 @@
-#!/usr/bin/env python
-
-"""
-Unit tests for the pandoc minted.lua filter.
-"""
-
-# Lint this file with: flake8 --max-line-length=80
-import os
-import string
-import subprocess
-import sys
-import textwrap
-
-code_block = textwrap.dedent('''
-    ## A Code Block
-
-    ```{.cpp}
-    auto mult = []<typename T, typename U>(T const & x, U const & y) {
-        return x * y;
-    };
-    ```
-''')
-"""
-The base CodeBlock code.  {.cpp} is used as a replacement marker in most tests!
-"""
-
-inline_delims = '|!@#^&*-=+' + string.digits + string.ascii_letters
-inline_code = textwrap.dedent('''
-    ## Inline Code
-
-    `#include <type_traits>`{.cpp}
-    C and C++ use `{` and `}` to delimit scopes.
-    Some other special characters:
-    These check bypass: `~!@#$%^&*()-=_+[]\\{}|;\':",./<>?`
-    These check regular inline: ''' + ' '.join(
-    '`{' + inline_delims[:i] + '`' for i in range(len(inline_delims))
-))
-"""
-The base Code code.  {.cpp} is used as a replacement marker in most tests!
-"""
-
-
-def run_pandoc(pandoc_args, stdin):
-    """Run pandoc with the specified arguments, returning the output."""
-    # The input / output should be small enough for these tests that buffer
-    # overflows should not happen.
-    pandoc_proc = subprocess.Popen(
-        ["pandoc"] + pandoc_args,
-        stdout=subprocess.PIPE, stderr=subprocess.PIPE, stdin=subprocess.PIPE
-    )
-
-    # Python 3.x and later require communicating with bytes.
-    if sys.version_info[0] >= 3:
-        stdin = bytes(stdin, "utf-8")
-
-    stdout, stderr = pandoc_proc.communicate(input=stdin)
-    if pandoc_proc.returncode != 0:
-        sys.stderr.write("Non-zero exit code of {ret} from pandoc!\n".format(
-            ret=pandoc_proc.returncode
-        ))
-        sys.stderr.write("pandoc stderr: {stderr}".format(
-            stderr=stderr.decode("utf-8")
-        ))
-        sys.exit(1)
-
-    return stdout.decode("utf-8")
-
-
-def fail_test(test_name, messages, ansi_color_code="31"):
-    """
-    Print failure message and ``sys.exit(1)``.
-
-    ``test_name`` (str)
[…]

### 2019-10-29 code/wrangle/wrangle_cellct.Rmd
initial commit

---
title: "Clean cell count data"
output: html_notebook
---

```{r}
library(tidyverse)
library(here)
```

```{r}
path_cellct <- here("data/raw/cellct_2018.csv")
raw_cellct <- read_csv(path_cellct)
glimpse(raw_cellct)
```

```{r}
thresh <- 50 # set this much lower than 100, to keep some counts with high fungi relative abundance

cellct <- raw_cellct %>%
  # drop non-algae categories
  select(-three_arm:-biofilm) %>% 
  pivot_longer(cols = -sample_id, names_to="morpho_sp", values_to="count") %>%
  filter(count != 0) %>% 
  drop_na() %>% 
  mutate(morpho_sp = case_when( str_detect(morpho_sp, "green")~"Green cell", # lump all green categories
                                morpho_sp=="orb"~"Sanguina nivaloides", 
                                morpho_sp=="ruby"~"Sanguina nivaloides", # assuming Sanguina grows turrets 
                                morpho_sp=="balloon"~"Chlainomonas",
                                morpho_sp=="tangerine"~"Chloromonas krienitzii",
                                morpho_sp=="hedgehog"~"Chloromonas cf. brevispina",
                                morpho_sp=="lemon"~"Chloromonas cf. nivalis",
                                morpho_sp=="spiny"~"Other",
                                morpho_sp=="oval"~"Other",
                                morpho_sp=="zuke"~"Other",
                                morpho_sp=="halo"~"Other",
                                morpho_sp=="citrine"~"Sanguina nivaloides",
                                TRUE ~ morpho_sp)) %>% 
  # drop samples with count < thresh
  group_by(sample_id) %>% 
  mutate(count_total = sum(count)) %>% 
  filter(count_total > thresh) %>% 
  # calculate percentage
  mutate(percent_of_sample = count / count_total) %>% 
  ungroup()

head(cellct)
```

```{r}
path_to_write <- here("data/clean/clean.cellct.csv")
write_csv(cellct, path_to_write)
```

### 2019-10-29 code/cellcount.Rmd
init

---
title: "Cellct analysis"
output: html_notebook
---

```{r}
library(tidyverse)
```

```{r}
path_cellct <- here("data/clean/clean.cellct.csv")
cellct <- read_csv(path_cellct)
path_field <- here("data/clean/field_clean.csv")
field <- read_csv(path_field)
path_alias <- here("data/sample_aliases.csv")
alias <- read_csv(path_alias)
```

Add site metadata
```{r}
p <- cellct %>% 
  left_join(field, by="sample_id") %>% 
  left_join(alias, by="sample_id")
p
```

Plot as a heatmap
```{r, fig.height=12, fig.width=7}

morpho_order <- c("Green cell",
                  "Sanguina 'citrine'",
                  "Sanguina nivaloides",
                  "Sanguina nivaloides 'turret'",
                  "Green cell",
                  "Chlainomonas",
                  "Chloromonas krienitzii",
                  "Chloromonas cf. brevispina",
                  "Chloromonas cf. nivalis",
                  "Chloromonas 'spiny'",
                  "Chloromonas 'oval'",
                  "Chloromonas 'zuke'"
                  )

ggplot(p, aes(y=alias %>% fct_reorder(elev_m), 
              x=morpho_sp)) +
  geom_tile(aes(alpha = percent_of_sample, fill = morpho_sp )) + # color = "white" %>% fct_reorder(morpho_order)
  scale_alpha(range = c(0.1, 1)) +
  ylab("Sample ID") +
  xlab("Cell morphology") +
  labs(fill = "Cell morphology", alpha="Relative abundance") +
  scale_x_discrete(position = "bottom") +
  theme(axis.text.x = element_text(hjust=1,angle=45)) +
  guides(fill=FALSE)
```

### 2019-10-29 manuscript/Casey1.Rmd
removed comments from ms

@@
 <br>
-<!-- Our findings highlight the need to consider the species composition of snow algae blooms in future research of this poorly understood microbiome. -->
 
@@
 ### *To be submitted to*: 1) Nat Comm, 2) Frontiers in Micb 3) Env Micb 4) Microbial Ecology
-
-<!-- what we now know: (Old studies suggested... here we are blowing open the door. ) -->
-<!-- - there are different microbiomes -->
-<!-- - alpine vs subalpine -->
-<!-- - communities are not distinct, but a gradient!! -->
-
-<!-- completely novel -->
-<!-- - chloromonas krienitzii blooms in snow runnels (more seen below treeline), green below, orange above -->
-<!-- - chlainomonas is widespread in all blooms -->
-<!-- - muramotoi in North America? (if this is a reliable genetic marker) -->
-<!-- - Some unknown Chloromonas OTUs with no match in ref databases -->
-<!-- - distinct clade Chlainomonas above treeline -->
-<!-- - sphere krienitzii, oval, same? or closely related... -->
-<!-- - Chlainomonas more at surface than depth -->
-
-<!-- 18S data -->
-<!-- write discussion -->
 
@@
 nivalis* are now split into multiple species [@Matsuzaki2019;@Prochazhova2019a].
-<!-- based on this paragraph, I would not trust the morphology categories in my thing. Ill take em out anyway, but... chlainomonas is distinct, chloro distinct -->
 
-<!-- get reader up to date on snow algae taxonomy... two out of three are poorly explored -->
 To date, genetic studies have identified three main genera of algae that cause snow algae blooms: *Sanguina*, *Chlainomonas*, and *Chloromonas*.
@@
 *Chloromonas* are highly diverse, with at least 13 named species [@Matsuzaki2019]; in contrast, only two species are currently described in *Chlainomonas* [@Novis2008].
-<!-- how communities vary from patch to patch. -->
-
-<!-- Much of the recent snow algae literature has focused on the eye-catching red snow prevalent throughout polar and alpine regions [@Segawa2018;@Leya2013].  -->
 
@@
 
-<!-- Liquid water availability and nutrient limitation likely play a role in determining where snow algae blooms, as both water and NPK fertilizer increased cell counts of *Chlamydomonas cf.
-nivalis* in one study in Alaska .
--->
-
-<!-- im not doing community structure / network analysis (yet) -->
-<!-- what is the point of this para? rbcL, our hot new method, is the shit -->
 Microscopic observation shows that snow algae blooms can contain a diversity of cell morphologies in the same sample [@Lutz2018], but the genetic community structure of snow algae may not correlate with what is seen under the microscope.
@@
 
-<!-- These blooms are often dominated by red spherical cells, which until recently were referred to in the literature as *Chlamydomonas cf.
-nivalis*, but recent taxonomy classifies as *Sanguina nivaloides*, although synonymy was not established [@Prochazkova2019].
--->
-<!-- What species? Are snow algae communities distinct, or overlapping gradient?  -->
-
-<!-- Snow algae diversity is poorly understood, and whether different species inhabit different niches remains an open question.  -->
-
-<!-- A diversity of algae species and cell morphologies are reported from snow algae blooms, yet little is known about the habitat distribution of snow algae species. -->
-<!-- why is this interesting? -->
-<!-- make it obvious what the paper is about early as possible -->
-
-<!-- yet community composition is perceived to be relatively homogeneous between sites [@Lutz2016a]; whether there may be different types of snow algae blooms has not been adequately explored. -->
-
-<!-- but their impact on alpine food webs and nutrient cycling has not been explored. -->
-
-<!-- Until recently, snow algae species were defined on the basis of cell morphology. -->
-
-<!-- *Sanguina* currently has only two named species, , and .
-The competitive exclusion principle suggests that within the melting snowpack microbiome there exist different niches, for example different snow moisture content, shade, or nutrient input from trees [@Hoham1976].
--->
-
-<!-- Chloro polyptera absent in rbcL. could we detect this in our samples by 18S? how similar is it to other Chloromonas? -->
-
-<!-- Red cells of *Chlamydomonas cf. nivalis* can be either cosmopolitan or endemic between the poles [@Segawa2018]. -->
-
-<!-- What species of Chloromonas? (Chlainomonas within)  -->
-
-<!-- What potentially novel species look similar, but are genetically distinct? -->
-
-<!-- Are all snow algae communities the same, or different?
-Do the differences in cell morphology correspond to genetically different species?
-What described species are found in the snow?
-Which of these may be unknown species that looked similar in previous surveys?
--->
-<!-- What do we now know?
-Chlainomonas hi low, above and below treeline specific otus, Chloro krienitzii is widespread in runnels, esp below treeline, green below turning orange on surface.
-Sanguina less common below treeline.
--->
-
-<!-- what we now know: (Old studies suggested... here we are blowing open the door. ) -->
-
-<!-- - gradient, dom morph vs mix -->
-<!--   - chlainomonas is widespread in all blooms -->
-<!--   - X17 -->
-
-<!-- - alpine vs subalpine -->
-<!--   - Some unknown Chloromonas OTUs with no match in ref databases -->
-<!--   - distinct clade Chlainomonas above treeline -->
-
-<!-- - krienitzii -->
-<!--   - in snow runnels (more seen below treeline), green below, orange above -->
-<!--   - sphere krienitzii, oval, same? or closely related... -->
-<!--   - Chlainomonas more at surface than depth -->
-
 # Results {-}
@@
 2a).
-<!--to use named OTUs, or not.
-The line drawn by our classifier is also arbitrary, depending on database etc but less so than me circling some dots.
-But I could make a definition in methods, refer to OTU by the most abundant species in it--> However, most samples contained a mix of species; few samples were relatively "pure".
+However, most samples contained a mix of species; few samples were relatively "pure".
 However, sample comparison with weighted UniFrac [@Lozupone2005] found little difference between *Chlainomonas* and *Chloromonas* dominated sites, and the presence or absence of *Sanguina* was the driving factor, which was generally found in higher relative abundance at higher elevation (Fig.
@@
 This does not rule out the possibility that these sites contained high moisture at some point; further sampling is needed to determine its habitat distribution.
-
-<!-- One *Chlainomonas* dominated sample (S9) was from snow near the edge of a pool, but we also collected *Chlainomonas* at sites where we did not see evidence of water nearby (S13, G16.1), and one from a pink snow runnel (S6) that looked the same as the runnels dominated by *C.
-krienitzii*.
-*Chlainomonas* dominated samples appeared indistinguishable to pink snow caused by *Sanguina* to the naked eye, although it was less frequently collected than *Sanguina*.
-This suggests that *Chlainomonas* could play a role in albedo-driven snow melt previously ascribed to *Chlamydomonas cf.
-nivalis* or *Sanguina*.
-*Chlainomonas* was not previously reported to be so prevalent in snow algae blooms of mixed composition.
-This could be because *Chlainomonas* is not distinguishable from *Chloromonas* with 18S primers in previous alpine HTS surveys [@Brown2016a,@Hamilton2017].
-It is not reported from either pole, although this could be due to lack of sampling as well as lack of detect ability with HTS.
--->
 
@@
 brevispina* “Trochiscia” [@Matsuzaki2015] (Fig.
-4<!-- which panel-->), while green subsurface snow was dominated by green biflagellate or bipolar quadriflagellate cells (Fig.
+4), while green subsurface snow was dominated by green biflagellate or bipolar quadriflagellate cells (Fig.
 4).
@@
 3).
-<!--We were surprised to find that --><!--H5 was dominated by *Chlainomonas* reads, but this sample contained few cells matching this morphology.-->
 
@@
 
-<!-- Runnels could provide habitat, e.g.
-liquid water for motility, nutrient mobilization, alternatively the algae could passively accumulate due to gravity.
-These green cells could swim up from the base of the snowpack, or be passively washed into the runnels from the surface.
-We observed some green snow ice lenses below these runnels, which could be due to higher snow moisture in the runnel, or due to the algae themselves through antifreeze production or albedo.
--->
-
-<!-- Although all samples dominated by *C.
-krienitzii* were taken from elevations near or below treeline, our microscopy observations suggest that *C.
-krienitzii* blooms infrequently at higher elevations as well.
-Why *C.
-krienitzii* dominates below treeline is a mystery, since we observed it in high light clearings.
--->
-<!-- Exposure to light can trigger astaxanthin production in Haematococcus pluvialis [@Shah2016], -->
-
 # Discussion {-}
@@
 Further work needs to be done to uncover the micro-habitat niches within the snowpack that allow this differentiation, and what novel species the snow microbiome might contain.
-
-<!-- Where to put these? -->
-<!-- cross validation -->
-<!-- Sample composition varied between our three data sets (rbcL, 18S, and cell count), although some robust correlations were found between them.
-*Sanguina* relative abundance was strongly correlated between data sets, between rbcL and 18s (Spearman's rho=0.78, p<0.001) and rbcL and "*Sanguina*-like" cell morphologies (Spearman's rho=0.92, p<0.001).
-*Chlainomonas* was strongly correlated with "*Chlainomonas*-like" cells (Spearman's rho=0.7, p<0.001), but IDTaxa did not identify any *Chlainomonas* in our 18S library (in spite of the fact that we included one sample in our 18S reference database).
-*Chloromonas krienitzii* was not correlated with cell count data (because green cells), nor any 18S ASVs or OTUs.
-Koliellaceae was not strongly correlated between 18S and rbcL datasets by either relative abundance data or presence absence.
-No strong correlations were found between Unk8 and taxa or OTUs in other datasets.
-Sample S4 was dominated by cell morphologies we classified as *Chloromonas cf.
-brevispina*, and this sample was dominated by reads best matching *Chloromonas krienitzii*.
-*Sanguina* and *Trebouxiophyceae* was in proportionally higher among 18S algae reads than in rbcL (respectively 40% vs 25%, and 0.9% vs 2%), while *Chloromonas* was lower in 18S than rbcL *Chloromonas* and *Chlainomonas* combined (24% vs 45%).
--->
-
-<!-- mention our brevispina sample -->
-<!-- Cell morphology also suggests rbcL underestimates diversity.
-Sample S4 was dominated by *C.
-krienitzii* reads, but the cell morphologies dominant in this samples were similar to *Chloromonas cf.
-brevispina* Cryocystis-type [@Matsuzaki2015], which is a distinct but closely related species to *C.
-krienitzii*.
-Consistent with @Nedbalova2008a, we found these cells in bronze snow in well-shaded forest sites.
-Likely our short-read rbcL fragment would not offer enough genetic resolution to differentiate between the two, with the genetic "signal" of *C.
-cf.
-brevispina* getting lost in the "noise" of haplotype variation in *C.
-krienitzii*.
--->
-
-<!-- primer bias -->
-<!-- Primer bias likely impacted our estimates of relative abundance.
-Samples containing 80-90% *Sanguina* 18S were only 50-60% *Sanguina* rbcL.
-This is likely because we designed our rbcL primers off of *Chloromonas* and *Chlainomonas*, as *Sanguina* sequences were not available at the time, thus leading to their disproportionate representation.
-Perhaps the three uppermost outliers in Figure 2 (K7, K19.1, and G16.2) could be because these samples contained very little *Chloromonas* initially.
-Although *Chloromonas* and *Chlainomonas* may be artificially inflated by rbcL, they are likely a significant secondary component of *Sanguina* blooms, as our 18S data suggests, as well as microscopy.
-Other studies also found that *Chlamydomonas cf.
-nivalis* blooms contained lower abundance of *Chloromonas* [@Lutz2016a;@Terashima2017].
-Furthermore extraction bias could likewise have resulted in over-representation of *Chloromonas* and *Chlainomonas*, since we observed that *Sanguina* were difficult to lyse by light microscopy cell counts, and contained lower DNA yields per dry weight than samples containing green cells.
--->
-
-<!-- Our rbcL amplicon library was dominated by reads assigned to *Chloromonas*, *Sanguina*, and *Chlainomonas* (Fig.
-1).
-The most abundant species were *Chloromonas krienitzii*, *Chlainomonas rubra*, and *Sanguina nivaloides*.
-<!--exact percentage?--> <!--X% of reads were assigned to genus, and X% of reads were assigned to species.
-Many ASVs were not assigned to genus level by our taxonomic classifier IDTaxa, but were nevertheless closely related to *Chloromonas* (Fig.
-1, Supp.
-Fig.
-1 and 2).
-Although one OTU "Chloro7" was distant from other *Chloromonas* ASVs (Supp.
-Fig.
-2), best BLAST matches were all *Chloromonas* snow algae (best match 92%).
-Similarly, Chloro9 had best BLAST matches to Chloromonas, with a best match of 94%.
-Another OTU "Unk8" was only assigned to domain level by IDTaxa, but top BLAST hits included algae from both Chlamydomonadaceae and Trebouxiophyceae, with a best match of 88% similarity.
-<!--This cluster did not correlate with any 18S OTUs.
-However, another unidentified clade (labeled as “Chlorophyceae X” in Figure 1) was highly correlated with an 18S OTU that was similarly placed taxonomically.-->
-
-<!-- Did 18S identify any of the unknown clusters? What was in the unknown clusters? -->
-
-<!-- Three green snow samples from above treeline were dominated by *Chloromonas* reads.  -->
-<!-- Two Sanguina dominated samples contained high proportion of smaller orange cells... -->
-
-<!-- move some of this to discussion? -->
-
- <!--However, the low abundance community was less differentiated between samples<!-- too discussion-y?-->  <!--How to mention the higher diversity at higher elevation?
-Chao with fossil pkg?
-There was greater genetic variation between high elevation samples than there was between lower elevation samples (Fig.
-2b) this is because there were more samples?
-more sites?
-hard to make comparisons.
-<!-- ...
-becayse Chloromonas and Chlainomonas are closely related-->
-
-<!-- should I go into Chlain genetics, above and below treeline?
-morph differences?
-If I get into the morphology, I am characterising new species.
-Maybe reviewers will want me to go back and flesh out the details of the unknown species, but for now I won't mention it to provoke their curiosity -->
-
-<!-- todo, get 18S clade -> OTU that correlates to Chloro 7. get correl. -->
-
-<!-- remind of the problem: we have all this diversity, where is it? what are the differences? -->
-
-<!--Chloro3, what to make of this. Are we suggesting it is a single species? (NO)-->
-<!-- why do we care if there are new species?-->
-
-<!--minor point, whittle down-->
-
-<!-- Red, orange, or green snow can be caused by different species of snow algae.
-Much of the snow algae literature has focussed on the eye-catching red snow prevalent throughout polar and alpine regions caused by red spherical cells [@Leya2013;@Segawa2018;@Lutz2016a].
-Until recently, these cells were referred to in the literature as *Chlamydomonas cf.
-nivalis*, but were recently re-classified as *Sanguina nivaloides* [@Prochazkova2019].
-Red snow can also be caused by *Chlainomonas*, whose distinctive large (40 $\mu$m diameter) quadriflagellate red cells can be found in waterlogged snow over alpine lakes [@Novis2008;@Prochazkova2018c].
-Snow algae that lack secondary pigments can color snow green, which occurs in slushy wet snow in the Arctic and Antarctic [@Fogg1967;@Lutz2015;@Remias2013], or in subalpine forests [@Hoham1979b].
-Green snow is now thought to be caused by different genera than in red snow, including *Microglena* [@Lutz2015], and various *Chloromonas* [Hoham1979b;@Remias2013a;@Terashima2017].
-Two studies (in USA and Slovakia) found green and orange snow to be more common below treeline, while red snow is prevalent in the alpine [@Hoham1979a;@Nedbalova2008a].
-<!-- add element of unknown to this, emphasize how little we know-->
-
-<!-- Although  -->
-
-<!-- Cell morphology can be ambiguous in snow algae taxonomy; different species can look nearly identical, and the same species can look completely different depending on the life stage.
-Recent molecular work shows that strains formerly lumped together as *Chloromonas.
-cf.
-nivalis* contain four genetically-distinct lineages [@Matsuzaki2019], while cells described as *C.
-cf.
-brevispina* contain at least two genetically distinct clades [@Matsuzaki2015].
--->
-
-<!-- S.aurantia-->
-<!-- sky27 and gar 2 couldbe had fewer Chloromonas, so there were no Chloromonas to preferentially amplify -->
-<!-- haplotypes, a single ASV difference may not be biologically meaningful -->
-<!--future directions: what is the unknowns? -->
-<!-- Below treeline, trees shade snowpack and provide a potential source of nutrients through conifer needles [@Hoham1976]; furthermore the window of opportunity for snow algae growth is much shorter as snow melts out sooner at low elevations.
--->
-
-<!-- Problems with snow algae 18S amplicon high-throughput sequencing (HTS).
-Although these studies report taxonomy to the species level, outdated annotation in reference databases, algorithms prone to overclassification [@Murali2018], lack of cross-referenced marker genes and microscopy samples, primers targeting highly conserved regions, and arbitrary OTU thresholds [@Callahan2017] could have hidden or mislabelled taxa, potentially obscuring variation between samples.
--->
-<!-- - Most species of *Chloromonas* are known from a single sample.  -->
-
-<!-- blooms side by side different community, which sample? -->
-<!-- rbcL underestimates non Chloromonas, less differentiated within Sanguina -->
- <!-- nes18.01, has chlainomonas, but visually more sanguina. primer bias, check 18s.  -->
- <!-- are the citrine cells sanguina? -->
- <!-- return to this in discussion, we found that green snow near red snow was Chloromonas, but we did see some green cells below the surface in transition to red, similar to reported in Prochazkova2019.
-Where these cells come from is a enduring mystery, how the bloom is seeded.
-We observed red snow in patches on the saddles, with no green below.
-The snow was shallow below the alpine green Chloromonas.
--->
 

### 2019-10-29 code/make_sample_aliases.Rmd
init

---
title: "Generate sample aliases"
output: html_notebook
---

```{r}
library(tidyverse)
```

```{r}
path <- here("data/clean/field_clean.csv")
field <- read_csv(path)
```
Give each mountain a unique letter ID
```{r}
field_1 <- field %>% 
  select(sample_id, elev_m, date) %>% 
  mutate(mountain = str_sub(sample_id,0,3)) 
field_1 %>% 
  distinct(mountain) %>% 
  arrange(mountain)
```

Sax =X, Sey=S, Sky=K, and Stm=M
```{r}
field_2 <- field_1 %>% 
  mutate(mountain_letter = case_when(mountain == "sax"~"X",
                                     mountain == "sey"~"S",
                                     mountain == "sky"~"K",
                                     mountain == "stm"~"M",
                                     TRUE ~ str_sub(mountain, 0,1) %>% toupper()))
field_2
```

Now give each sample a unique identifier by date
```{r}
sample_dates <- field_2 %>%
  distinct(date) %>% 
  arrange(date) %>% 
  rownames_to_column("date_ord")

field_3 <- field_2 %>% 
  left_join(sample_dates, by="date")

field_3 %>% select(date_ord, date)
```

Add a decimal for samples collected at the same date
```{r}
# split df by date, order each date by elevation going down, then add this as decimal to the alias
alias_tbl <- field_3 %>% 
  mutate(alias = paste0(mountain_letter, date_ord)) %>% 
  arrange(alias, -elev_m) %>% 
  group_by(alias, add = T) %>% 
  group_split(alias) %>% 
  map(rownames_to_column) %>% 
  reduce(bind_rows) %>%
  ungroup() %>% 
  mutate(alias = paste0(alias, ".",rowname)) %>% 
  select(sample_id, alias)
```

```{r}
path_to_write <- here("data/sample_aliases.csv")
write_csv(alias_tbl, path_to_write)
```

### 2019-10-30 manuscript/Casey1.Rmd
re outlined intro, still mid-way thru

@@
 
-Snow algae are cold-adapted microalgae that bloom in melting summer snow [@Leya2013].
-Red pigments produced by some species of snow algae can color vast areas of polar and alpine snowfields, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017].
-Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], and loss of snowpack reservoirs can threaten water supplies, leading to drought [@Staudinger2014;@Barnhart2016].
-Thus, snow algae could play an important role in snow melt regulation.
-Snow algae are the primary producers in a microbiome that includes grazers, fungi, and bacteria [@Hoham2001], but whether this microbiome can be considered as a single microbiome, or whether there are distinctions between types of snow algae blooms has been poorly studied.
-The first step in understanding this microbiome is defining the algae species that are present in snow algae blooms, and their community structure.
+Snow algae are cold-adapted microalgae that bloom in melting summer snow throughout alpine and polar regions [@Leya2013].
+Red snow algae blooms can cover vast areas of snowfield, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017].
+Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], which can threaten water supplies [@Staudinger2014] and worsten drought [@Barnhart2016].
+Thus, snow algae could play an important role in snow melt regulation, but little is known about how these blooms form.
+We don't know the spatial extent of snow algae blooms worldwide, what conditions trigger snow algae bloom formation, or how cells arrive on the surface of the snow to "seed" the bloom.
+Critically, we lack basic understanding of the species composition and diversity in snow algae blooms, which is fundamental to answering these long-standing mysteries about snow algae blooms.
 
-Until recently, snow algae species were defined and identified on the basis of morphology.
-Morphology can be deceptive when identifying microalgae, as the same species can look completely different from one life stage to the next [@Shah2016;@Salome2019].
-In the lab, cultures of *Chloromonas* are green biflagellates, but in the field the same species contain yellow or orange pigment, and thick cell walls with flanges or spikes.
-Conversely, different genetic species can look nearly identical: for example, cell types formerly known as *Chloromonas cf.
-nivalis* are now split into multiple species [@Matsuzaki2019;@Prochazhova2019a].
+Historically snow algae were classified by cell morphology, however, morphology can be deceptive when identifying microalgae.
+Until recently, all red cells from red snow were referred to as *Chlamydomonas cf.
+nivalis* [@Remias2013;@Segawa2018], desipte the fact that this morphology-based species was thought to contain multiple distinct species [@Kol1961].
+Recent genetic work designated found that some of these cells were not closely related to *Chlamydomonas* at all, and were designated as a new genus and species: *Sanguina nivaloides* [@Prochazkova2018c].
+Other genetic work shows that morphology-based species of snow algae may in fact contain multiple genetically unrelated clades: for example, *Chloromonas cf.
+nivalis* contains at least four distinct clades [@Matsuzaki2019].
 
-To date, genetic studies have identified three main genera of algae that cause snow algae blooms: *Sanguina*, *Chlainomonas*, and *Chloromonas*.
-Neither of the first two have cultured representatives, but both produce the red pigment astaxanthin [@Remias2005;Remias2016].
-*Sanguina* was only recently designated and contains just two species, one of which *Sanguina nivaloides* is found in alpine and polar snowfields worldwide, and is likely synonymous with some of *Chlamydomonas cf.
-nivalis* [@Prochazkova2018c].
-*Chlainomonas* also produce red pigment, but are physically much larger than *Sanguina* (up to 40 $\mu$m diameter).
-Although *Chlainomonas* look very different from *Chloromonas*, *rbc*L and 18S phylogeny places them within the *Chloromonas* clade [@Novis2008].
-*Chloromonas* are highly diverse, with at least 13 named species [@Matsuzaki2019]; in contrast, only two species are currently described in *Chlainomonas* [@Novis2008].
+<!-- hammer this home: WE LACK BASIC UNDERSTANDING ABOUT SPECIES COMPOSITOIN AND DIVERSITY -->
 
+<!-- sequence cultures, and batch, and single cell, but this dosent tell us whats actually in the field -->
+More recently, snow algae taxonomy has
+
+<!-- high throughput: lacking ref data, no otu based, none in alpine environments (despite this tantalizing evidence of microscopy based, etc) -->
+
+<!-- But no one until now has used high throughput sequencing to compare snow algae blooms in alpine environments, or used higher resolution marker genes to do so -->
+
+<!-- so we did this... -->
+In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS.
+Our aims were to: 1) see what species were present, 2) compare community composition between samples.
+We found ...
+
+<!-- the red green problem -->
+The same species can look completely different depending on the life stage [@Shah2016;@Salome2019]; conversely, different species can look nearly identical.
+For example, cell morphologies known as *Chloromonas cf.
+nivalis* appear in at least four different lineages.
+One of these lineages, *Chloromonas muramotoi* appear as green biflagellates in culture, but in the field the same species features thick cell walls with flanges [@Matsuzaki2019].
+However, life cycle changes in the the cosmopolitan and widespread genus *Sanguina* have yet to be documented: despite over a century of study [@Kol1961] no one has sucessfully cultured this genus.
+
+<!-- habitat differences -->
 Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013].
@@
 
+<!-- habitat differences, short new form -->
+  Red snow is also occasionally reported to be caused by another red-pigmented algae: *Chlainomonas* [@Novis2008].
+Little is known about the distribution of this genus; one study suggested that *Chlainomonas* is limited to high-moisture snowpack above alpine lakes [@Prochazkova2018].
+Green snow may contain a variety of algae, notably *Chloromonas* (refs), as well as algae from other families including Trebouxiophyceae and Chrysophyceae (refs).
+Morphology-based cell counts suggest that green snow caused by *Chloromonas* is prevalent in forested areas and red snow containing *Chlamydomonas cf.
+nivalis* is prevalent in alpine areas [@Hoham1979a;@Nedbalova2008a]; however, morphology can be misleading when identifying microalgae.
+
+Although much of the snow algae literature focusses on eye-catching red snow, different species of snow algae can also color snow green [@Lutz2015].
+
+<!-- hts -->
 Microscopic observation shows that snow algae blooms can contain a diversity of cell morphologies in the same sample [@Lutz2018], but the genetic community structure of snow algae may not correlate with what is seen under the microscope.
@@
 
-In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS.
-Our aims were to: 1) see what species were present, 2) compare community composition between samples.
-We found ...
+Until recently, snow algae sequences were lacking in reference databases,
 
@@
 However, in microscopy surveys we infrequently observed *Sanguina* below treeline, and *Chloromonas krienitzii* above treeline.
-We found higher diversity in samples from higher elevation, both at the ASV level and OTU level<!--Chao1 w fossil.-->, with several OTUs that were not found at lower elevations.
+We found higher diversity in samples from higher elevation, both at the ASV level and OTU level <!--Chao1 w fossil.-->, with several OTUs that were not found at lower elevations.
 *Raphidonema* and related ASVs were restricted to high alpine sites, with the highest relative abundance occurring in samples from August and September.

### 2019-10-30 code/make_18s_algae_fasta.Rmd
init

---
title: "Select Algal 18S ASVs "
output: html_notebook
---

```{r}
library(tidyverse)
library(DECIPHER)
```

```{r}
key_path <- here("intermediate_output/asv_seq_key_18s.csv")
key <- read_csv(key_path)

path <- here("data/raw/18s_assignments_silva.csv")
assignments <- read_csv(path)
assignments
```

# Only Algae

Select Chlorophyta ASVs (according to SILVA)
```{r}
algae_18s_seqs <- assignments %>% 
  dplyr::rename(asv_sequence = asv) %>% 
  filter(V3 %>% str_detect("Chlorophyta")) %>% 
  # get the asv_id, for labelling the tree tips
  left_join(key, by="asv_sequence") %>% 
  select(asv_id, asv_sequence) %>% 
  deframe() %>% # to vector
  DNAStringSet() 
algae_18s_seqs
```
Align
```{r}
aligns <- algae_18s_seqs %>%
  AlignSeqs(verbose=F)

BrowseSeqs(aligns, highlight = T)
```

Remove ASVs with gaps
```{r}
gappy_asvs <- c("508", "570", "581", "586", "587", "591")

quality_aligns <- aligns[!(names(aligns) %in% gappy_asvs)]

# check
length(quality_aligns) == length(aligns) - length(gappy_asvs)
```

# try with all ASVs

Make another fasta with all 18S ASVs,
```{r}
seqs <- assignments %>% 
  dplyr::rename(asv_sequence = asv) %>% 
  # get the asv_id, for labelling the tree tips
  left_join(key, by="asv_sequence") %>% 
  select(asv_id, asv_sequence) %>% 
  deframe() %>% # to vector
  DNAStringSet()
seqs
```

Remove gappy seqs
```{r}
seqs <- seqs[!(names(seqs) %in% gappy_asvs)]
seqs
```

Align
```{r}
all_aligns <- seqs %>%
  AlignSeqs(verbose=F)

BrowseSeqs(all_aligns, highlight = T)
```

Not the best looking alignment I've ever seen, but roll with it

Write fastas
```{r}
path_write <- here("intermediate_output/algae_18s_seqs.fasta")
writeXStringSet(quality_aligns, path_write)

path_all <- here("intermediate_output/all_18s_aligns.fasta")
writeXStringSet(all_aligns, path_all)
```

### 2019-10-30 code/wrangle/clean_silva_tax_assignment.R
init

@@ -1,0 +1,31 @@
+ # wrangle silva 18s assignment table
+
+library(tidyverse)
+library(here)
+
+path <- here("intermediate_output/asv_seq_key_18s.csv")
+key <- read_csv(path)
+
+path2 <- here("data/raw/18s_assignments_silva.csv")
+silva <- read_csv(path2)
+
+clean_silva_18s_assignments <- silva %>% 
+  dplyr::rename(asv_sequence = asv) %>% 
+  left_join(key, by="asv_sequence") %>% 
+  select(-asv_sequence) %>% 
+  dplyr::rename(kingdom = V2, 
+                phylum = V3, 
+                class = V4,
+                order = V5,
+                family = V6,
+                genus = V7,
+                kingdom_conf = V21, 
+                phylum_conf = V31, 
+                class_conf = V41,
+                order_conf = V51,
+                family_conf = V61,
+                genus_conf = V71) %>% 
+  select(asv_id, kingdom:genus_conf)
+
+write_path <- here("data/clean/clean_silva_18s_assignments.csv")
+write_csv(clean_silva_18s_assignments, write_path)

### 2019-10-30 code/wrangle/filter_18s_ra_to_algae.R
init

@@ -1,0 +1,26 @@
+# this script takes in the clean algae 18s relative abundance table, and returns a filtered version that only contains the Chlorophyta assigned by IDTaxa using SILVA
+
+library(tidyverse)
+library(here)
+
+path <- here("data/clean/clean_18s_ra.csv")
+ra <- read_csv(path)
+
+path2 <- here("data/clean/clean_silva_18s_assignments.csv")
+silva <- read_csv(path2)
+
+filtered_rel_abund <- ra %>% 
+  left_join(silva, by="asv_id") %>% 
+  filter(phylum %>% str_detect("Chlorophyta"))
+
+# check
+filtered_rel_abund %>% 
+  distinct(asv_id, phylum)
+
+# remove the taxonomy columns
+filtered_rel_abund_final <- filtered_rel_abund %>% 
+  select(asv_id, sample_id, n_reads, rel_abund)
+
+# write file
+path_out <- here("data/clean/only_algae_18s_rel_abund.csv")
+write_csv(filtered_rel_abund_final, path_out)

### 2019-10-30 code/wrangle/wrangle_18s_rel_abund.Rmd
write out 18s asv_id - seq key file

@@
 write_csv(clean_18s_ra, clean_18s_ra_path)
+
+asv_path <- here("intermediate_output/asv_seq_key_18s.csv")
+write_csv(asv_seq_key_18s, asv_path)
 ```

### 2019-10-30 code/whole_18s_community_compare.Rmd
init

---
title: "Full 18s Community composition"
output: html_notebook
---

```{r libs, include=F}
library(tidyverse)
library(here)
library(GUniFrac)
library(ape)
library(ggtree)
library(ggrepel)
```

```{r input, include=F}
tree_path <- here("intermediate_output/tree/full_18s/all_18s_aligns.fasta.treefile")
all_18s_tree <- read.tree(tree_path)

ra_path <- here("data/clean/clean_18s_ra.csv")
ra <- read_csv(ra_path)

assignment_path <- here("data/clean/clean_silva_18s_assignments.csv")
assign <- read_csv(assignment_path)

field_path <- here("data/clean/field_clean.csv")
field <- read_csv(field_path)

alias_path <- here("data/sample_aliases.csv")
alias <- read_csv(alias_path)
```

```{r to_wide, include=F}
wide_ra <- ra %>% 
  select(asv_id, sample_id, rel_abund) %>% 
  filter(asv_id %in% all_18s_tree$tip.label) %>% # Unifrac needs relative abundance table to match the tips on the tree
  pivot_wider(names_from = asv_id, values_from = rel_abund, values_fill = list(rel_abund = 0)) %>% 
  column_to_rownames("sample_id")
head(wide_ra)
```

[Some background on Eukaryotic phylogeny](http://tolweb.org/Eukaryotes/3)
```{r root_tree, include=F}
# rooting by bacteria
root_id <- assign %>% 
  filter(kingdom=="Bacteria") %>% 
  head(1) %>% 
  pull(asv_id)

all_18s_tree <- all_18s_tree %>% root(outgroup = root, resolve.root=T)
```

```{r unifrac, include=F}
set.seed(123)
unifrac_out <- wide_ra %>% 
  GUniFrac::GUniFrac(all_18s_tree, alpha=c(0, 0.5, 1)) # ignore error message, we removed some otus from our tree, they should be automatically discluded

# select the weighted UniFrac matrix
weighted_unifrac_distm <- unifrac_out$unifracs[, , "d_1"]
dim(weighted_unifrac_distm)
```

```{r nmds_stress, include=F}
mono <- metaMDS(weighted_unifrac_distm)
mono
```

View plot of UniFrac NMDS
```{r, echo=F}
unifrac_plot_data <- mono %>%
  scores() %>%
  as_tibble(rownames = "sample_id") %>%
  left_join(field, by="sample_id")

# plot unifrac, colored by elevation
uf_plot <- ggplot(unifrac_plot_data, aes(x=NMDS2, y=NMDS1)) +
  geom_point(aes(color=elev_m), size=4, alpha=0.7) +
  scale_color_gradientn(colours = terrain.colors(8)[1:7]) +
  labs(color="Elevation (m)") +
  geom_text_repel(aes(label=sample_id)) +
  ggtitle("UniFrac NMDS of 18S amplicon data (only 2019 MiSeq run)")
  # geom_text(data = unifrac_plot_data %>%   # add sample ID labels
  #                   left_join(alias, by="sample_id"),
  #                   aes(label = sample_id), size=3,
  #                   position = position_nudge(y = -0.025))  ######### change to alias for final version. edit overlap txt in svg
uf_plot
```

  - Seymour samples clustered somewhat, except for sey18.25, sey18.63o
    - sey25 = brevispina, big biofilms
  - Hollyburn, 42b close to sey, but 21 and 47 were closer to high elevation samples
  - Sky12 closer to high elevation samples despite being from below treeline
  - Sax1, gar1, bre1, nes3 outliers from other high elevation samples

Check out sey18.25
```{r, echo=F}
ra %>% 
  filter(sample_id == "sey18.25") %>% 
  arrange(-rel_abund) %>% 
  left_join(assign, by="asv_id")
```

```{r, echo=F, fig.height=8, fig.width=10}
p <- ra %>% 
  left_join(assign, by="asv_id") %>% 
  group_by(phylum, sample_id) %>% 
  summarise(rel_abund = sum(rel_abund)) %>% 
  ungroup() %>% 
  left_join(field, by="sample_id") %>% 
  select(sample_id, phylum, rel_abund, date, elev_m) %>% 
  arrange(sample_id)

ggplot(p, aes(x=phylum, y=sample_id %>% fct_reorder(elev_m))) +
  geom_point(aes(color = phylum, size = rel_abund), alpha=0.7) +
  scale_size(range = c(1, 10)) +
  ylab("Sample ID") +
  xlab("OTU") +
  labs(fill = "OTU", alpha="Relative abundance", title = "Relative abundance plot of 18S phyla ") +
  scale_x_discrete(position = "bottom") +
  theme(axis.text.x = element_text(hjust=1,angle=45)) +
  scale_colour_discrete(guide = FALSE) # turn off legend
```

```{r, echo=FALSE}
view_phylum_table <- function(phylum){
  phylum <- enquo(phylum)
  ra %>% 
  left_join(assign, by="asv_id") %>% 
  filter(phylum == !!phylum) %>% 
  arrange(-rel_abund)
}
```

<br>
<br>
Table zoom in on the most abundance phyla

### Arthropods
```{r, echo=F}
view_phylum_table("Arthropoda")
```

  - mostly at higher elev red snow sites, and hol47 (which LQ collected, dirty red snow near stream)
  - hol47 arthropods are all Collembola, similar to other sites

### Ascomycota
```{r, echo=F}
view_phylum_table("Ascomycota")
```
  - unknown Cordycipitaceae, a family of parasitic fungi
  - ASV 4 BLAST 100% *Fusarium*, widely distributed in soil, saprophyta assoc w plants (although Wiki places in a different family, Nectriaceae)

### Basidiomycota
```{r, echo=F}
view_phylum_table("Basidiomycota")
```

  - sey7, sey25, hol21, pan.ro, wed2 have highest fungi
    - pan.ro:
      - once its off the snow, fungi jump in
      - high unknown Euks
    - 
  - Kriegeriales yeast
  - asv7 blast matches a bunch of fungi 100%

### Cercozoa
```{r, echo=F}
view_phylum_table("Cercozoa")
```
  - Glissomonads
    - ASV17 BLAST best strain match, 96% *Orciraptor agilis*, a parasite of algae

### Chlorophyta

####Trebouxiophyceae
```{r, echo=F}
ra %>% 
  left_join(assign, by="asv_id") %>% 
  filter(class == "Trebouxiophyceae") %>% 
  group_by(sample_id) %>% 
  summarise(rel_abund=sum(rel_abund)) %>% 
  arrange(-rel_abund)
```
  - sax18.01 outlier, much higher Trebouxiophyceae
    - could explain why an outlier in *rbc*L data?

####Chlorophyceae
```{r, echo=F}
ra %>% 
  left_join(assign, by="asv_id") %>% 
  filter(class == "Chlorophyceae") %>% 
  group_by(sample_id) %>% 
  summarise(rel_abund=sum(rel_abund)) %>% 
  arrange(rel_abund)
```
  - lowest Chlorophyceae is sey18.25, with 18%
  - highest " " is gar18.01, with 80%

### Chytridiomycota
```{r, echo=F}
view_phylum_table("Chytridiomycota")
```

### Ciliophora:
```{r, echo=F}
view_phylum_table("Ciliophora")
```
  - SILVA assigns most to *Stokesia*, but best BLAST match for asv5 is *Frontonia* 
  - highest in hol18.42b (depth)
      - and nes3, green snow sample

### Ochrophyta
```{r, echo=F}
view_phylum_table("Ochrophyta")
```

  - higher abundance than exected
  - Xanthophyceae at pan18.01, otherwise mostly *Hydrurus*
  - Chrysophyceae, Leya found this family causing green snow
  - seen in microscopy
  - < 1% rel abundance
  - could be wind deposited or grow in snow
    - known to grow in cold, clear waters

### Rotifera
```{r, echo=F}
view_phylum_table("Rotifera")
```
  - sey18.25: highest rotifers, matches photos. 
    - could explain why its an outlier in NMDS above
    -

# Check differences between surface and depth samples:
  -

# Check differences in high vs low diversity
  -

### 2019-10-30 code/community_comparison_18s.Rmd
remade relative abundance plot as faceted barplot

---
title: "18S Community Comparison"
output: html_notebook
---

```{r libs, results="hide"}
library(tidyverse)
library(here)
library(GUniFrac)
library(vegan)
library(vegalite)
library(fossil)
library(picante)

# ggplot themes
library(ggthemes)
library(ggdark)
```

```{r set}
theme_set(theme_classic())
```

```{r}
path_18s_ra <- here("data/clean/only_algae_18s_rel_abund.csv")
clean_18s_ra <- read_csv(path_18s_ra)

path_tree <- here("intermediate_output/tree/algae_18s/algae_18s_seqs.fasta.treefile")
tree_18s <- read.tree(path_tree)

path_full_18s_tree <- here("intermediate_output/tree/18s/")
full_tree_18s <- read.tree(path_full_18s_tree)

path_silva_assigns <- here("data/clean/clean_silva_18s_assignments.csv")
silva <- read_csv(path_silva_assigns)

path_field <- here("data/clean/field_clean.csv")
field <- read_csv(path_field)

path_alias <- here("data/sample_aliases.csv")
alias <- read_csv(path_alias)
```

# Unifrac Algae 18s

convert to wide format, asvs as cols
```{r unifrac}
wide_18s_ra <- clean_18s_ra %>% 
  select(asv_id, sample_id, rel_abund) %>% 
  filter(asv_id %in% tree_18s$tip.label) %>% # Unifrac needs relative abundance table to match the tips on the tree
  pivot_wider(names_from = asv_id, values_from = rel_abund, values_fill = list(rel_abund = 0)) %>% 
  column_to_rownames("sample_id")
head(wide_18s_ra)
```

Choose a Trebouxiophyceae to root tree by
```{r}
root <- silva %>% 
  filter(class == "Trebouxiophyceae") %>% 
  head(1) %>% 
  pull(1)

tree_18s <- tree_18s %>% root(outgroup = root, resolve.root=T)
```

run Unifrac
```{r}

set.seed(123)
unifrac_out <- wide_18s_ra %>% 
  GUniFrac::GUniFrac(tree_18s, alpha=c(0, 0.5, 1)) # ignore error message, we removed some otus from our tree, they should be automatically discluded

# select the weighted UniFrac matrix
weighted_unifrac_distm <- unifrac_out$unifracs[, , "d_1"]
dim(weighted_unifrac_distm)
```

Ordination (NMDS) of 18s UniFrac, check stress value
```{r}
mono <- metaMDS(weighted_unifrac_distm)
iso <- metaMDS(weighted_unifrac_distm, engine = "isoMDS")

mono
iso

```

stress too low to plot... with the few algae asvs, perhaps it all looks too similar

try with all 18s data, not just algae

```{r}
unifrac_plot_data <- nmds %>%
  scores() %>%
  as_tibble(rownames = "sample_id") %>%
  left_join(field, by="sample_id")

# plot unifrac, colored by elevation
uf_plot <- ggplot(unifrac_plot_data, aes(x=NMDS2, y=NMDS1)) +
  geom_point(aes(color=elev_m), size=4) +
  scale_color_gradientn(colours = terrain.colors(8)[1:8]) +
  labs(color="Elevation (m)") #+
  # geom_text(data = unifrac_plot_data %>%   # add sample ID labels
              # left_join(sample_aliases, by="sample_id"),
            # aes(label = alias), size=3,
            # position = position_nudge(y = -0.025))  ######### change to alias for final version. edit overlap txt in svg
uf_plot
```

### 2019-10-30 code/whole_18s_community_compare.Rmd
updated relative abundance plot

@@
 
-View plot of UniFrac NMDS
-```{r, echo=F}
+```{r unifrac_nmds, echo=F}
 unifrac_plot_data <- mono %>%
@@
 
-```{r, echo=F, fig.height=8, fig.width=10}
+```{r set_plot, echo=F}
 p <- ra %>% 
@@
   select(sample_id, phylum, rel_abund, date, elev_m) %>% 
-  arrange(sample_id)
+  arrange(sample_id) %>% 
+  filter(rel_abund >= 0.01) # only show phyla with >= 1% rel abund
 
-ggplot(p, aes(x=phylum, y=sample_id %>% fct_reorder(elev_m))) +
-  geom_point(aes(color = phylum, size = rel_abund), alpha=0.7) +
-  scale_size(range = c(1, 10)) +
-  ylab("Sample ID") +
-  xlab("OTU") +
-  labs(fill = "OTU", alpha="Relative abundance", title = "Relative abundance plot of 18S phyla ") +
-  scale_x_discrete(position = "bottom") +
-  theme(axis.text.x = element_text(hjust=1,angle=45)) +
-  scale_colour_discrete(guide = FALSE) # turn off legend
+# ggplot(p, aes(x=phylum, y=sample_id %>% fct_reorder(elev_m))) +
+#   geom_point(aes(color = phylum, size = rel_abund), alpha=0.7) +
+#   scale_size(range = c(1, 10)) +
+#   ylab("Sample ID") +
+#   xlab("OTU") +
+#   labs(fill = "OTU", alpha="Relative abundance", title = "Relative abundance plot of 18S phyla ") +
+#   scale_x_discrete(position = "bottom") +
+#   theme(axis.text.x = element_text(hjust=1,angle=45)) +
+#   scale_colour_discrete(guide = FALSE) # turn off legend
+```
+
+```{r facet_bar, echo=F, fig.width=10}
+cust_theme <- theme(legend.position = "none",
+                    panel.grid.major.x = element_blank(), # remove vertical lines in background
+                    panel.grid.minor.x = element_blank(),
+                    axis.line=element_blank(), # remove x axis
+                    axis.title.x=element_blank(), 
+                    axis.text.x=element_blank(),
+                    axis.ticks.x=element_blank(),
+                    panel.spacing = unit(-0.01, "lines"), # remove space between facet panels
+                    strip.text.x = element_text(angle = 90, hjust=1.0, vjust = 0.8), # rotate facet labels 90
+                    strip.background = element_blank()) # remove grey boxes around facet labs
+
+# order algae, fungi, protists, other
+phylum_ord <- c("Chlorophyta_ph", "Ochrophyta","Basidiomycota", "Ascomycota", "Chytridiomycota", "Ciliophora","Cercozoa","Protalveolata", "Phragmoplastophyta")
+
+ggplot(p, aes(x=sample_id %>% fct_reorder(elev_m), y=rel_abund)) +
+  geom_bar(aes(fill=phylum), stat="identity") + # classic stacked relative abundance plot
+  coord_flip() +
+  facet_grid(~phylum %>% fct_relevel(phylum_ord), switch = "x") +
+  labs(title = "Relative abundance plot of 18S phyla", x="Sample ID (from lowest to highest elevation)") +
+  cust_theme
 ```
@@
 <br>
-Table zoom in on the most abundance phyla
-
-### Arthropods
-```{r, echo=F}
-view_phylum_table("Arthropoda")
-```
-
-  - mostly at higher elev red snow sites, and hol47 (which LQ collected, dirty red snow near stream)
-  - hol47 arthropods are all Collembola, similar to other sites
-
-### Ascomycota
-```{r, echo=F}
-view_phylum_table("Ascomycota")
-```
-  - unknown Cordycipitaceae, a family of parasitic fungi
-  - ASV 4 BLAST 100% *Fusarium*, widely distributed in soil, saprophyta assoc w plants (although Wiki places in a different family, Nectriaceae)
-
-### Basidiomycota
-```{r, echo=F}
-view_phylum_table("Basidiomycota")
-```
-
-  - sey7, sey25, hol21, pan.ro, wed2 have highest fungi
-    - pan.ro:
-      - once its off the snow, fungi jump in
-      - high unknown Euks
-    - 
-  - Kriegeriales yeast
-  - asv7 blast matches a bunch of fungi 100%
-
-### Cercozoa
-```{r, echo=F}
-view_phylum_table("Cercozoa")
-```
-  - Glissomonads
-    - ASV17 BLAST best strain match, 96% *Orciraptor agilis*, a parasite of algae
 
@@
 
+### Ochrophyta
+```{r, echo=F}
+view_phylum_table("Ochrophyta")
+```
+
+  - higher abundance than exected
+  - Xanthophyceae at pan18.01, otherwise mostly *Hydrurus*
+  - Chrysophyceae, Leya found this family causing green snow
+  - seen in microscopy
+  - < 1% rel abundance
+  - could be wind deposited or grow in snow
+    - known to grow in cold, clear waters
+
+### Basidiomycota
+```{r, echo=F}
+view_phylum_table("Basidiomycota")
+```
+
+  - sey7, sey25, hol21, pan.ro, wed2 have highest fungi
+    - pan.ro:
+      - once its off the snow, fungi jump in
+      - high unknown Euks
+    - 
+  - Kriegeriales yeast
+  - asv7 blast matches a bunch of fungi 100%
+
+### Ascomycota
+```{r, echo=F}
+view_phylum_table("Ascomycota")
+```
+  - unknown Cordycipitaceae, a family of parasitic fungi
+  - ASV 4 BLAST 100% *Fusarium*, widely distributed in soil, saprophyta assoc w plants (although Wiki places in a different family, Nectriaceae)
+
 ### Chytridiomycota
@@
 
-### Ochrophyta
+### Cercozoa
 ```{r, echo=F}
-view_phylum_table("Ochrophyta")
+view_phylum_table("Cercozoa")
+```
+  - Glissomonads
+    - ASV17 BLAST best strain match, 96% *Orciraptor agilis*, a parasite of algae
+
+### Arthropods
+```{r, echo=F}
+view_phylum_table("Arthropoda")
 ```
 
-  - higher abundance than exected
-  - Xanthophyceae at pan18.01, otherwise mostly *Hydrurus*
-  - Chrysophyceae, Leya found this family causing green snow
-  - seen in microscopy
-  - < 1% rel abundance
-  - could be wind deposited or grow in snow
-    - known to grow in cold, clear waters
+  - mostly at higher elev red snow sites, and hol47 (which LQ collected, dirty red snow near stream)
+  - hol47 arthropods are all Collembola, similar to other sites
 

### 2019-10-30 code/rbcl_community_compare.Rmd
init, made a fresh rbcl community compare notebook

---
title: "rbcL community comparison"
output: html_notebook
---

```{r libs, include=F}
library(tidyverse)
library(here)
library(GUniFrac)
library(vegan)
library(ggrepel)
# library(vegalite)
# library(fossil)
library(picante)

# ggplot themes
library(ggthemes)
library(ggdark)
```

```{r input, echo=F}
rbcl_rel_abund <- read_csv(here("data/clean/rbcl_rel_abund_clean.csv"))
idtax <- read_csv(here("data/clean/rbcl_taxonomy_clean.csv"))
field <- read_csv(here("data/clean/field_clean.csv"))

rbcl_asv_id_key <- read_csv(here("intermediate_output/rbcl_asv_key.csv"))
rbcl_asv_otu_key <- read_csv(here("intermediate_output/asv_otu_key.csv"))
alias <- read_csv(here("data/sample_aliases.csv"))
rbcl.tree <- read.tree(here("intermediate_output/tree/rbcl/asv613.fasta.treefile"))
```

```{r unifrac, echo=F}
ra_wide <- rbcl_rel_abund %>% 
  select(asv_id, sample_id, rel_abund) %>% 
  pivot_wider(names_from = asv_id, values_from = rel_abund, values_fill = list(rel_abund = 0)) %>% 
  column_to_rownames("sample_id")

# root tree with Trebouxiophyceae
root <- idtax %>% 
  filter(class=="Trebouxiophyceae") %>% 
  head(1) %>% 
  pull(asv_id)

rbcl.tree <- rbcl.tree %>% root(outgroup = root, resolve.root=T)

# run weighted UniFrac
set.seed(123)
unifracs <- ra_wide %>% 
  GUniFrac::GUniFrac(rbcl.tree, alpha=1)
unifrac <- unifracs$unifracs[, , "d_1"]

# run NMDS ordination on results 
nmds <- metaMDS(unifrac, autotransform = F)

# plot
unifrac_plot_data <- nmds %>% 
  scores() %>% 
  as_tibble(rownames = "sample_id") %>% 
  left_join(field, by="sample_id") %>% 
  left_join(alias, by="sample_id")

# plot unifrac, colored by elevation
uf_plot <- ggplot(unifrac_plot_data, aes(x=NMDS2, y=NMDS1)) +
  geom_point(aes(color=elev_m), size=4) +
  scale_color_gradientn(colours = terrain.colors(8)[1:7]) +
  labs(color="Elevation (m)") +
  geom_text_repel(aes(label=alias))
uf_plot
```

```{r rel_abund_barplot, fig.width=10, fig.height=7, echo=F}
plot_data <- rbcl_rel_abund %>% 
  left_join(rbcl_asv_otu_key, by="asv_id") %>% 
  group_by(sample_id, otu_name) %>% 
  summarise(rel_abund = sum(rel_abund)) %>%   
  left_join(field, by="sample_id") %>%
  left_join(alias, by="sample_id") %>% 
  ungroup() %>% 
  mutate(rel_abund = if_else(rel_abund <0.01, 0.01, rel_abund)) # artificially boost rel abund <1% to 1% so its visible on plot

# check rel abund sums = 1
plot_data %>%
  group_by(alias) %>%
  summarise(sum(rel_abund))

cust_theme <- theme(legend.position = "none",
                    panel.grid.major.x = element_blank(), # remove vertical lines in background
                    panel.grid.minor.x = element_blank(),
                    axis.line=element_blank(), # remove x axis
                    axis.title.x=element_blank(), 
                    axis.text.x=element_blank(),
                    axis.ticks.x=element_blank(),
                    panel.spacing = unit(-0.01, "lines"), # remove space between facet panels
                    strip.text.x = element_text(angle = 90, hjust=1.0, vjust = 0.8), # rotate facet labels 90
                    strip.background = element_blank()) # remove grey boxes around facet labs

# order OTUs
otu_ord <- c("Sanguina A","Chloromonas I","Chloromonas G","Chloromonas muramotoi E","Chloromonas krienitzii D","Chloromonas spp. C","Chlainomonas B","Chlainomonas F","Algae H", "Trebouxiophyceae J")

plot <- ggplot(plot_data, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
  geom_bar(aes(fill=otu_name), stat="identity") + # classic stacked relative abundance plot
  coord_flip() +
  facet_grid(~otu_name %>% fct_relevel(otu_ord), switch = "x") +
  labs( x="Sample ID (from lowest to highest elevation)") +
  cust_theme
plot
# add sparkline on side for elevation
```

```{r faiths, echo=F}
re_wide_mat <- ra_wide %>% as.matrix()

faiths <- re_wide_mat %>% 
  picante::pd(rbcl.tree) %>% 
  # attach elevation data
  rownames_to_column("sample_id") %>% 
  left_join(field, by="sample_id") %>% 
  left_join(alias, by="sample_id")

faith_date_plot <- ggplot(faiths, aes(y=PD, x=date))+
  geom_point(aes(color=elev_m), size=4) +
  geom_smooth(method = "lm")+
  scale_color_gradientn(colours = terrain.colors(8)[1:7]) +
  geom_text_repel(aes(label=alias))+
  labs(title="Phylogenetic diversity of snow algae rbcL throughout the summer of 2018",y="Faith's phylogenetic diversity (PD)",x="Date", color="Elevation (m)")
faith_date_plot
```

```{r lm_faiths_pd, include=FALSE}
lm_out <- lm(PD ~ date + elev_m, data = faiths)
summary(lm_out)

# just elev
lm_out1 <- lm(log(PD) ~ elev_m, data = faiths) %>% # log transform to normalize
  broom::tidy()

elev_slope <- lm_out1 %>% 
  filter(term == "elev_m") %>% 
  pull(estimate)
elev_p <- lm_out1 %>% 
  filter(term == "elev_m") %>% 
  pull(p.value)

# just date
lm_out2 <- lm(PD ~ date, data = faiths) %>% 
  broom::tidy()

date_slope <- lm_out2 %>% 
  filter(term == "date") %>% 
  pull(estimate)
date_p <- lm_out2 %>% 
  filter(term == "date") %>% 
  pull(p.value)

# check resids
hist(resid(lm_out1))

# calculate co-linearity
car::vif(lm_out)
```

```{r output, include=F}
# ggsave
```

### 2019-10-31 code/depth_comparison.Rmd
split up wrangle tasks into scripts, made seperate csvs

---
title: "Depth sample comparison"
output: html_notebook
---

```{r}
library(tidyverse)
library(here)
```

```{r include=F}
rbcl_rel_abund <- read_csv(here("data/"))
s18_all_rel_abund

rbcl_tax
s18_tax
```

### 2019-10-31 code/whole_18s_community_compare.Rmd
split up wrangle tasks into scripts, made seperate csvs

@@
   facet_grid(~phylum %>% fct_relevel(phylum_ord), switch = "x") +
-  labs(title = "Relative abundance plot of 18S phyla", x="Sample ID (from lowest to highest elevation)") +
+  labs(title = "Relative abundance plot of 18S phyla (SILVA assignments)", x="Sample ID (from lowest to highest elevation)") +
   cust_theme

### 2019-10-31 code/make_sample_aliases.R

@@ -1,0 +1,39 @@
+# Make sample aliases
+
+library(tidyverse)
+
+field <- read_csv(here("data/02_tidied/field_clean.csv"))
+
+# each mountain gets a letter, each date gets a number, each 
+
+# assign unique mountain_letter code to each field site
+field_1 <- field %>% 
+  mutate(mountain_letter = case_when(str_detect(sample_id, "sax") ~ "X",
+                                     str_detect(sample_id, "sey") ~ "S",
+                                     str_detect(sample_id, "sky") ~ "K",
+                                     str_detect(sample_id, "stm") ~ "M",
+                                     TRUE ~ str_sub(sample_id, 0,1) %>% toupper()))
+
+
+# which visit to the mountain (only 1 visit for most sites)
+visit_number <- field_1 %>%
+  distinct(mountain_letter, date) %>% 
+  arrange(mountain_letter, date) %>% 
+  group_split(mountain_letter) %>% 
+  map_df(rownames_to_column, "visit_number") %>% 
+  mutate(alias = paste0(mountain_letter, visit_number))
+  
+field_2 <- field_1 %>% 
+  left_join(visit_number, by=c("mountain_letter", "date"))
+  
+# for each moutain visit, unique sample ID
+alias <- field_2 %>% 
+  arrange(-elev_m) %>% # from high to low since thats usually the order I collected in
+  group_split(alias) %>% 
+  map_df(rownames_to_column, "uid") %>% 
+  mutate(alias = paste0(alias, ".", uid)) %>% 
+  select(sample_id, alias)
+  
+  
+
+write_csv(alias, here("output/sample_aliases.csv"))

### 2019-10-31 code/wrangle/clean_silva_tax_assignment.R

@@ -1,31 +1,0 @@
- # wrangle silva 18s assignment table
-
-library(tidyverse)
-library(here)
-
-path <- here("intermediate_output/asv_seq_key_18s.csv")
-key <- read_csv(path)
-
-path2 <- here("data/raw/18s_assignments_silva.csv")
-silva <- read_csv(path2)
-
-clean_silva_18s_assignments <- silva %>% 
-  dplyr::rename(asv_sequence = asv) %>% 
-  left_join(key, by="asv_sequence") %>% 
-  select(-asv_sequence) %>% 
-  dplyr::rename(kingdom = V2, 
-                phylum = V3, 
-                class = V4,
-                order = V5,
-                family = V6,
-                genus = V7,
-                kingdom_conf = V21, 
-                phylum_conf = V31, 
-                class_conf = V41,
-                order_conf = V51,
-                family_conf = V61,
-                genus_conf = V71) %>% 
-  select(asv_id, kingdom:genus_conf)
-
-write_path <- here("data/clean/clean_silva_18s_assignments.csv")
-write_csv(clean_silva_18s_assignments, write_path)

### 2019-10-31 code/wrangle/tidy_rbcl_assignment.R

@@ -1,0 +1,90 @@
+# tidy rbcl assignment data
+
+library(tidyverse)
+library(here)
+
+raw_assignments <- read_csv(here("data/01_raw/rbcl_taxonomy_genbank.csv"))
+
+assignments <- raw_assignments %>% 
+  # rename columns
+  dplyr::rename(asv_sequence=asv,
+                domain=V2, 
+                kingdom=V3, 
+                phylum=V4,
+                class=V5,
+                order=V6, 
+                family=V7, 
+                genus=V8, 
+                species=V9, 
+                domainConf=V21,
+                kingdomConf=V31, 
+                phylumConf=V41, 
+                classConf = V51, 
+                orderConf=V61, 
+                familyConf=V71, 
+                genusConf=V81, 
+                speciesConf=V91) %>% 
+  # replace default IDTaxa output "unclassified" with NA
+  mutate_if(is.character, funs(replace(., str_detect(., "unclassified"), NA_character_))) %>% ### funs deprecated
+  # make new col detecting assignment level
+  mutate(id_level = case_when(is.na(domain) ~"root",
+                              is.na(kingdom) ~"domain",
+                              is.na(phylum)~"kingdom",
+                              is.na(class)~"phylum",
+                              is.na(order)~"class",
+                              is.na(family)~"order",
+                              is.na(genus)~"family",
+                              is.na(species)~"genus",
+                              TRUE ~ "species")) %>% 
+  # make new col with best assignment
+  mutate(assignment = case_when(id_level =="root"~"root",
+                                id_level == "domain"~ domain,
+                                id_level == "kingdom"~kingdom,
+                                id_level == "phylum"~phylum,
+                                id_level == "class"~class,
+                                id_level == "order"~order,
+                                id_level == "family"~family,
+                                id_level == "genus"~genus,
+                                id_level == "species"~species))
+
+assignments %>% view()
+assignments %>% nrow()
+
+
+
+# replace the actual sequence with a sequence identifier
+asv_id <- assignments %>% 
+  select(asv_sequence) %>% 
+  rownames_to_column("asv_id") %>% 
+  mutate(asv_id = str_pad(asv_id,3,pad="0"))
+
+
+# add asv_id to taxonomy tbl
+assigns_w_asv_id <- asv_id %>% 
[…]

### 2019-10-31 code/rbcl_community_compare.Rmd
split up wrangle tasks into scripts, made seperate csvs

@@
 
-```{r input, echo=F}
+```{r input, include=F}
 rbcl_rel_abund <- read_csv(here("data/clean/rbcl_rel_abund_clean.csv"))
@@
 
-```{r unifrac, echo=F}
+```{r unifrac, include=F}
 ra_wide <- rbcl_rel_abund %>% 
@@
 nmds <- metaMDS(unifrac, autotransform = F)
+```
 
+```{r unifrac_plot, echo=F}
 # plot
@@
 
-```{r rel_abund_barplot, fig.width=10, fig.height=7, echo=F}
+```{r setup_rel_abund_barplot, fig.width=10, fig.height=7, include=F}
 plot_data <- rbcl_rel_abund %>% 
@@
   summarise(sum(rel_abund))
+```
 
+```{r plot_barplot, fig.width=10, fig.height=7, echo=F}
 cust_theme <- theme(legend.position = "none",

### 2019-10-31 code/wrangle/wrangle_18s.Rmd
split up wrangle tasks into scripts, made seperate csvs

--- 2019-10-29 code/wrangle/wrangle_18s_rel_abund.Rmd
@@
 ---
-title: "Wrangle 18s rel abund data"
+title: "R Notebook"
 output: html_notebook
@@
 
-```{r libs, results="hide"}
+# asdf
+test
+
+```{r libs}
 library(tidyverse)
@@
 library(metagenomeSeq)
+print("a")
 ```
@@
 
-```{r input, results="hide"}
+Object naming convention: first what the objects is in its local context, then any globel identifiers
+
+I'll refer to 18S relative abundance table object as "18s_ra"
+
+```{r input}
 path_18s_ra <- here("data/raw/ky_18s_seqtab.nochim_April2019.txt")
 raw_18s_ra <- read.csv(rel_abund_18S_path, sep=" ", stringsAsFactors = F)
-
-# also read in the rbcl rel abund tidy data, since I'm only using the 18s data as a cross reference I can filter by this
-path_rbcl_ra <- here("data/clean/rbcl_rel_abund_clean.csv")
-clean_rbcl_ra <- read_csv(path_rbcl_ra)
-```
-
-Object naming convention: 
-- first what the objects is in its local context, then any globel identifiers
-- I'll refer to 18S relative abundance table object as "18s_ra"
-
-Rename sample IDs
-- remove MiSeq lane identifier ("S" followed by a number)
-```{r}
 head(raw_18s_ra)
-renamed_sids_18s_ra <- raw_18s_ra %>%
-  as_tibble(rownames = "sample_id") %>% 
-  mutate(sample_id = sample_id %>% 
-           str_remove("_S.*_") %>% # remove _S*_ ending
-           str_replace_all("-",".")) #replace hyphen with period
-head(renamed_sids_18s_ra)
 ```
@@
 ```{r}
-transposed_18s_ra <- renamed_sids_18s_ra %>% 
-  column_to_rownames("sample_id") %>% 
+transposed_18s_ra <- raw_18s_ra %>%
   t() %>% 
@@
            str_pad(width=3, pad="0"))
-glimpse(asv_seq_key_18s)
+nrow(asv_seq_key_18s)
 ```
@@
 long_18s_ra <- seq_removed_18s_ra %>% 
-  pivot_longer(cols = -asv_id, names_to = "sample_id", values_to = "n_reads") %>% 
-  filter(n_reads != 0) # now that its in long format we can remove 0s
+  pivot_longer(cols = -asv_id, names_to = "sample_id", values_to = "n_reads")
 
-glimpse(long_18s_ra)
+head(long_18s_ra)
+nrow(long_18s_ra)
+```
 
+Format sample IDs properly (thanks to our poor organization during sequencing)
+```{r reformat_sample_id}
+fixed_sample_ids_18s_ra <- long_18s_ra %>% 
+  mutate(sample_id = 
+           # remove underscores
+           str_extract(sample_id, ".*(?=_S.*_)") %>% 
+           # replace hyphen
+           str_replace_all("-","."), 
+         # parse protocol version used from the sample ID col
+         protocol = str_extract(sample_id, "(?<=[:alpha:]{3}18\\..{2}\\.p)[:digit:]"), 
+         # remove protocol identifier from the sample ID col
+         sample_id = str_replace(sample_id, ".p[:digit:]",""), 
+         # protocol 3 was default, so can assume that these remaining are p3
+         protocol = replace_na(protocol, 3))
+fixed_sample_ids_18s_ra %>% 
+  select(sample_id) %>% 
+  distinct()
 ```
@@
 ```{r}
-as_percent_18s_ra <- long_18s_ra %>% 
-  group_by(sample_id) %>% 
+as_percent_18s_ra <- fixed_sample_ids_18s_ra %>% 
+  group_by(sample_id, protocol) %>% 
   mutate(rel_abund = n_reads/sum(n_reads)) %>% 
   ungroup()
-glimpse(as_percent_18s_ra)
+as_percent_18s_ra
 ```
 
-# Set aside negative controls
-
-look for duplicate sample ids
-```{r}
-as_percent_18s_ra %>% 
-  distinct(sample_id)
-```
-
-set aside duplicate sample ids
-```{r}
-duplicate_sample_18s_ra <- as_percent_18s_ra %>%
-  filter(sample_id %>% str_detect("gar18.01|sax18.01|sey18.25|sey18.74|sky18.12"))
-glimpse(duplicate_sample_18s_ra)
-```
-
-remove duplicates from my main analysis tbl
-```{r}
-no_dups_18s_ra <- as_percent_18s_ra %>% 
-  filter(!(sample_id %in% c("gar18.01.p2", "sax18.01.p1", "sey18.25.p1", "sey18.74.p1", "sky18.12.p1")))
-glimpse(no_dups_18s_ra)
-```
-
-remove the "p" from the sample id
-```{r}
-sample_id_fixed_18s_ra <- no_dups_18s_ra %>% 
-  mutate(sample_id = str_remove(sample_id, "\\.p[:digit:]"))
-glimpse(sample_id_fixed_18s_ra)
-```
-
-remove negative controls
-```{r}
-clean_18s_ra <- sample_id_fixed_18s_ra %>% 
-  filter(sample_id %in% clean_rbcl_ra$sample_id)
-
-glimpse(clean_18s_ra)
-```
-
-Check number of reads per sample
-```{r}
-clean_18s_ra %>% 
-  group_by(sample_id) %>% 
-  summarise(sum_n = sum(n_reads)) %>% 
-  arrange(sum_n)
-```
-
-write clean csv as output
-```{r}
-clean_18s_ra_path <- here("data/clean/clean_18s_ra.csv")
-write_csv(clean_18s_ra, clean_18s_ra_path)
-```
+Set aside negative controls, and

### 2019-10-31 manuscript/Casey1.Rmd
split up wrangle tasks into scripts, made seperate csvs

@@
   *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate.
-Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch within the same mountain, and varies with elevation and date.
+Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch on the same mountain, and varies with elevation and date.
 We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada.
-We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m), with a distinct clades of *Chlainomonas* found only in our high elevation sites.
-High elevation samples, which were collected later in the season, countained higher phylogenetic diversity---including several potentially novel clades of algae.
+We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m).
+Samples from higher elevation had higher phylogenetic diversity, and included several potentially novel clades of algae.
 Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface.
@@
 <!-- so we did this... -->
-In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS.
+In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS (explain what these code for).
 Our aims were to: 1) see what species were present, 2) compare community composition between samples.

### 2019-10-31 code/wrangle/filter_18s_ra_to_algae.R

@@ -1,26 +1,44 @@
-# this script takes in the clean algae 18s relative abundance table, and returns a filtered version that only contains the Chlorophyta assigned by IDTaxa using SILVA
+# this script takes in the clean, filtered 18s data, and limits it taxonomically to just algae
 
 library(tidyverse)
 library(here)
 
-path <- here("data/clean/clean_18s_ra.csv")
-ra <- read_csv(path)
 
-path2 <- here("data/clean/clean_silva_18s_assignments.csv")
-silva <- read_csv(path2)
+depth <- read_csv(here("data/03_filtered/18s/depth_18s_rel_abund.csv"))
+main <- read_csv(here("data/03_filtered/18s/filtered_cleaned_18s_rel_abund.csv"))
+protocol <- read_csv(here("data/03_filtered/18s/protocol_comparison_18s_rel_abund.csv"))
+scrape <- read_csv(here("data/03_filtered/18s/scrape_comparison_18s_rel_abund.csv"))
+water <- read_csv(here("data/03_filtered/18s/water_controls_18s_rel_abund.csv"))
+white <- read_csv(here("data/03_filtered/18s/white_snow_18s_rel_abund.csv"))
 
-filtered_rel_abund <- ra %>% 
-  left_join(silva, by="asv_id") %>% 
-  filter(phylum %>% str_detect("Chlorophyta"))
+taxonomy <- read_csv(here("data/02_tidied/tidied_silva_18s_assignments.csv"))
 
-# check
-filtered_rel_abund %>% 
-  distinct(asv_id, phylum)
+algae_asv_ids <- taxonomy %>%
+  filter(phylum == "Chlorophyta") %>% 
+  pull(asv_id)
 
-# remove the taxonomy columns
-filtered_rel_abund_final <- filtered_rel_abund %>% 
-  select(asv_id, sample_id, n_reads, rel_abund)
+filter_to_algae <- function(tbl){
+  tbl %>% 
+    left_join(taxonomy, by="asv_id") %>% 
+    filter(asv_id %in% algae_asv_ids) %>% 
+    select(1:4)
+}
 
-# write file
-path_out <- here("data/clean/only_algae_18s_rel_abund.csv")
-write_csv(filtered_rel_abund_final, path_out)
+# run filters
+depth_2 <- depth %>% filter_to_algae()
+main_2 <- main %>% filter_to_algae()
[…]

### 2019-10-31 code/wrangle/filter_18s_rel_abund.R

@@ -1,0 +1,103 @@
+# filter 18s dataset to just algae, and to just the samples in our rbcL dataset
+# set aside other rel abund datasets
+
+library(tidyverse)
+library(here)
+
+
+# input -----------------
+rel_abund_in <- read_csv(here("data/02_tidied/tidied.18s_rel_abund.csv"))
+head(rel_abund_in)
+
+# protocols ------------------------
+# parsed protocol vers
+rel_abund_1 <- rel_abund_in %>%
+  mutate(protocol = sample_protocol_id %>% 
+           str_extract("(?<=\\.p)[:digit:]$"),
+         sample_id = sample_protocol_id %>% 
+           str_remove("\\.p[:digit:]$"),
+         # if NA is protocol v3
+         protocol = if_else(is.na(protocol), "3",protocol)) %>% 
+  select(sample_id, protocol, asv_id, n_reads, rel_abund)
+head(rel_abund_1)
+
+# make a seperate tbl for protocol comparison
+samples_w_mult_protocols <- rel_abund_1 %>% 
+  distinct(sample_id, protocol) %>% 
+  dplyr::count(sample_id) %>% 
+  filter(n>1) %>% # choose if multiple protocols associated with a sample
+  pull(sample_id)
+samples_w_mult_protocols
+
+protocol_comparison_rel_abund <- rel_abund_1 %>% 
+  filter(sample_id %in% samples_w_mult_protocols)
+protocol_comparison_rel_abund %>% distinct(sample_id)
+
+# only keep the most recent protocol in my samples
+protocol_comparison_rel_abund %>% 
+  distinct(sample_id, protocol) %>% 
+  arrange(sample_id, protocol)
+
+rel_abund_2 <- rel_abund_1 %>%
+  # filter out the lowest protocol vers used in the duplicate samples
+  filter(!(sample_id == "gar18.01" & protocol == "2"),
+         !(sample_id == "sax18.01" & protocol == "1"),
+         !(sample_id == "sey18.25" & protocol == "1"),
+         !(sample_id == "sey18.74" & protocol == "1"),
+         !(sample_id == "sky18.12" & protocol == "1") )
+
+# now can remove protocol from the tbl
+rel_abund_2.5 <- rel_abund_2 %>% 
+  select(-protocol)
+
[…]

### 2019-10-31 code/wrangle/filter_rbcl_rel_abund.R

@@ -1,0 +1,96 @@
+# filter rbcl relative abund
+
+library(tidyverse)
+
+# input -----------------
+rel_abund_rbcl <- read_csv(here("data/02_tidied/tidied.rbcl_rel_abund.csv"))
+
+
+# protocols ------------------------
+# parsed protocol vers
+rel_abund_rbcl_1 <- rel_abund_rbcl %>%
+  dplyr::rename(protocol_sample_id = sample_id) %>% 
+  mutate(protocol = protocol_sample_id %>% 
+           str_extract("(?<=\\.p)[:digit:]$"),
+         sample_id = protocol_sample_id %>% 
+           str_remove("\\.p[:digit:]$"),
+         # if NA is protocol v3
+         protocol = if_else(is.na(protocol), "3",protocol)) %>% 
+  select(sample_id, protocol, asv_id, n_reads, rel_abund)
+
+# make a seperate tbl for protocol comparison
+samples_w_mult_protocols <- rel_abund_rbcl_1 %>% 
+  distinct(sample_id, protocol) %>% 
+  dplyr::count(sample_id) %>% 
+  filter(n>1) %>% # choose if multiple protocols associated with a sample
+  pull(sample_id)
+
+protocol_comparison_rel_abund <- rel_abund_1 %>% 
+  filter(sample_id %in% samples_w_mult_protocols)
+protocol_comparison_rel_abund %>% distinct(sample_id)
+
+# only keep the most recent protocol in my samples
+protocol_comparison_rel_abund %>% 
+  distinct(sample_id, protocol) %>% 
+  arrange(sample_id, protocol)
+
+rel_abund_2 <- rel_abund_1 %>%
+  filter(!(sample_id == "gar18.01" & protocol == "2"),
+         !(sample_id == "sax18.01" & protocol == "1"),
+         !(sample_id == "sey18.25" & protocol == "1"),
+         !(sample_id == "sey18.74" & protocol == "1"),
+         !(sample_id == "sky18.12" & protocol == "1") )
+
+
+
+# water negative controls ---------------
+
+water_ctrls <- rel_abund_2 %>% 
+  filter(sample_id %>% str_detect("water"))
+rel_abund_3 <- rel_abund_2 %>% 
+  filter(!(sample_id %>% str_detect("water")))
+
+# white snow samples ---------------
[…]

### 2019-10-31 code/wrangle/tidy.rbcl_rel_abund.R

@@ -1,31 +1,0 @@
-library(tidyverse)
-library(here)
-library(metagenomeSeq)
-
-seq.rbcl.2019 <- read.csv(here("data","raw","ky_rbcL_seqtab.nochim.txt"), sep=" ", stringsAsFactors = F) ##### KY saved as txt %^&*
-
-# to long format
-unfilt_rbcl_rel_abund <- seq.rbcl.2019 %>%
-  t() %>% 
-  as_tibble(rownames = "asv") %>% 
-  gather(key = "sample_id", val = "n_reads", -asv)
-
-# format sample ids properly
-unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
-  mutate(sample_id = str_extract(sample_id, ".*(?=_S.*_)") %>% # remove underscores
-           str_replace_all("-","."), # replace hyphen
-         protocol = str_extract(sample_id, "(?<=[:alpha:]{3}18\\..{2}\\.p)[:digit:]"), # parse protocol version used
-         sample_id = str_replace(sample_id, ".p[:digit:]",""), # remove p1
-         protocol = replace_na(protocol, 3)) # replace NA with 3
-
-# compute new variable "rel_abund"
-
-
-
-
-
-unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
-  group_by(sample_id, protocol) %>% 
-  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
-  ungroup()
-unfilt_rbcl_rel_abund

### 2019-10-31 code/wrangle/tidy_18s_rel_abund.R

@@ -1,0 +1,44 @@
+# Tidy 18s relative abundance data
+
+library(tidyverse)
+library(here)
+
+rel_abund <- read.csv(here("data/01_raw/ky_18s_seqtab.nochim_April2019.txt"), sep=" ", stringsAsFactors = F)
+key <- read_csv(here("output/asv_seq_key_18s.csv"))
+
+# this is basically the same script as for rbcl... 
+
+# to long format
+rel_abund_1 <- rel_abund %>% 
+  t() %>% 
+  as_tibble(rownames = "asv_sequence") %>% 
+  pivot_longer(cols = -asv_sequence, names_to = "sample_protocol_id", values_to = "n_reads") %>% 
+  drop_na() %>% 
+  filter(n_reads !=0)
+head(rel_abund_1)
+
+# clean sample names
+rel_abund_2 <- rel_abund_1 %>%
+  mutate(sample_protocol_id = sample_protocol_id %>% 
+           str_extract(".*(?=_S.*_)") %>% # remove underscores
+           str_replace_all("-",".") %>%  # replace hyphen
+           tolower())
+head(rel_abund_2)
+
+# compute relative abundance
+rel_abund_3 <- rel_abund_2 %>% 
+  group_by(sample_protocol_id) %>% 
+  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
+  ungroup()
+
+rel_abund_3 %>% distinct(sample_protocol_id) %>% view()
+
+# set aside actual sequences, replace with asv_id
+rel_abund_4 <- rel_abund_3 %>% 
+  left_join(key, by="asv_sequence") %>% 
+  select(sample_protocol_id, asv_id, n_reads, rel_abund) %>% 
+  arrange(sample_protocol_id)
+head(rel_abund_4)
+
+
+write_csv(rel_abund_4, here("data/02_tidied/tidied.18s_rel_abund.csv"))

### 2019-10-31 code/wrangle/tidy_rbcl_rel_abund.R

@@ -1,0 +1,43 @@
+# tidy rbcl relative abundance data
+
+library(tidyverse)
+library(here)
+
+raw <- read.csv(here("data/01_raw/ky_rbcL_seqtab.nochim.txt"), sep=" ", stringsAsFactors = F)
+key <- read_csv(here("output/rbcl_asv_sequence_key.csv"))
+
+rel_abund_long <- raw %>%
+  t() %>% 
+  as_tibble(rownames = "asv_sequence") %>% 
+  gather(key = "sample_id", val = "n_reads", -asv_sequence) %>% 
+  drop_na() %>% 
+  filter(n_reads !=0)
+
+# format sample ids (keep protocol for now ".p1" at end on some duplicate sequenced samples)
+clean_sample_ids <- rel_abund_long %>%
+  mutate(sample_id = sample_id %>% 
+           str_extract(".*(?=_S.*_)") %>% # remove underscores
+           str_replace_all("-",".") %>%  # replace hyphen
+           tolower())
+
+# compute relative abundance
+percent_rel_abund <- clean_sample_ids %>% 
+  group_by(sample_id) %>% 
+  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
+  ungroup()
+
+percent_rel_abund %>% distinct(sample_id) %>% view()
+
+# set aside actual sequences, replace with asv_id
+added_asv_id <- percent_rel_abund %>% 
+  left_join(key, by="asv_sequence") %>% 
+  select(asv_id, sample_id, n_reads, rel_abund) %>% 
+  arrange(sample_id)
+
+key %>% nrow()
+
+added_asv_id %>% 
+  distinct(asv_id) %>% 
+  nrow()
+
+write_csv(added_asv_id, here("data/02_tidied/tidied.rbcl_rel_abund.csv"))

### 2019-10-31 code/wrangle/tidy_silva_tax_assignment.R

@@ -1,0 +1,73 @@
+ # wrangle silva 18s assignment table
+
+library(tidyverse)
+library(here)
+library(Biostrings)
+
+silva <- read_csv(here("data/01_raw/taxonomy_18s_silva.csv"))
+
+silva_1 <- silva %>% 
+  dplyr::rename(asv_sequence = asv,
+                kingdom = V2, 
+                phylum = V3, 
+                class = V4,
+                order = V5,
+                family = V6,
+                genus = V7,
+                kingdom_conf = V21, 
+                phylum_conf = V31, 
+                class_conf = V41,
+                order_conf = V51,
+                family_conf = V61,
+                genus_conf = V71) %>% 
+  # replace default IDTaxa output "unclassified" with NA
+  mutate_if(is.character, funs(replace(., str_detect(., "unclassified"), NA_character_))) %>% ### funs deprecated
+  # make new col detecting assignment level
+  mutate(id_level = case_when(is.na(kingdom) ~"unknown",
+                              is.na(phylum)~"kingdom",
+                              is.na(class)~"phylum",
+                              is.na(order)~"class",
+                              is.na(family)~"order",
+                              is.na(genus)~"family",
+                              !is.na(genus)~"genus")) %>% 
+  # make new col with best assignment
+  mutate(assignment = case_when(id_level == "unknown"~ "unknown",
+                                id_level == "kingdom"~kingdom,
+                                id_level == "phylum"~phylum,
+                                id_level == "class"~class,
+                                id_level == "order"~order,
+                                id_level == "family"~family,
+                                id_level == "genus"~genus))
+silva_1 %>% view()
+silva_1 %>% nrow()
+
+# fix some annotation quirks
+silva_2 <- silva_1 %>% 
+  mutate(phylum = str_remove(phylum, "_ph"),
+         family = str_remove(family, "_fa"),
+         class = if_else(str_detect(class, "Incertae"), NA_character_, class))
+
+
+silva_2 %>% view()
+
+
+# make a key of the sequences and asv_ids, and replace the actual sequence with a key
+key <- silva_2 %>% 
+  select(asv_sequence) %>% 
+  rownames_to_column("asv_id") %>% 
+  mutate(asv_id = str_pad(asv_id, 3, pad="0"))
+head(key)
+
+seqs_to_fasta <- key %>% 
+  deframe() %>% 
[…]

### 2019-10-31 code/wrangle/wrangle.R

@@ -313,25 +313,4 @@ field <- field %>%
 
 # cell count data -------
 
-# rename, select algae
-cellct <- cellct %>% 
-  rowwise() %>% 
-  mutate(green=sum(green_oval, green_sphere,misc_small_green,tiny_green, spiny, na.rm = T),) %>% #lump green
-  select(sample_id, orb, citrine, redwall, ruby, balloon,tangerine, hedgehog, lemon, oval, zuke, green) # select only algae
 
-# to long, drop na, calculate per_algae
-cellct <- cellct %>% 
-  pivot_longer(names_to="morpho_sp", values_to="count", -sample_id ) %>% 
-  drop_na() %>% 
-  group_by(sample_id) %>% 
-  mutate(percent = count / sum(count))
-
-# remove samples with less than 50 entries (some with many fungi will be lost)
-count.sum.less.50 <- cellct %>% 
-  group_by(sample_id) %>% 
-  summarise(sum=sum(count)) %>% 
-  filter(sum<50) %>% 
-  pull(sample_id)
-
-cellct <- cellct %>% 
-  filter(!(sample_id %in% count.sum.less.50))

### 2019-10-31 code/depth_comparison.Rmd
init

@@
 ```{r include=F}
-rbcl_rel_abund <- read_csv(here("data/"))
-s18_all_rel_abund
+rbcl_rel_abund <- read_csv(here("data/03_filtered/rbcl/depth_rbcl_rel_abund.csv"))
+s18_all_rel_abund <- read_csv(here("data/03_filtered/18s/depth_18s_rel_abund.csv"))
 
-rbcl_tax
-s18_tax
+rbcl_tax <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
+s18_tax <- read_csv(here("data/02_tidied/tidied_silva_18s_assignments.csv"))
+
+rbcl_otus <- read_csv(here("output/asv_otu_key.csv"))
+
+field <- read_csv(here("data/02_tidied/tidied_field.csv"))
 ```
+
+```{r rbcl_wrangle, include=F}
+rbcl <- rbcl_rel_abund %>% 
+  left_join(rbcl_tax, by="asv_id") %>% 
+  left_join(field, by="sample_id") %>% 
+  left_join(rbcl_otus, by="asv_id")
+glimpse(rbcl)
+rbcl %>% distinct(sample_id, alias, depth_cm)
+
+depth_sample_info <- rbcl %>% 
+  distinct(sample_id, alias, depth_cm) %>% 
+  mutate(depth_chr = if_else(is.na(depth_cm), "surface", "sub-surface"),
+         top_sample = case_when(alias == "H2.10"~"H2.10*", # add a star, same site, but different date
+                                alias == "H3.4"~"H2.10*",
+                                alias == "H4.3"~"H4.4",
+                                alias == "S9.2"~"S9.1",
+                                TRUE ~ alias)) %>% 
+  select(sample_id, top_sample, depth_chr)
+depth_sample_info
+
+rbcl_1 <- rbcl %>% 
+  left_join(depth_sample_info, by="sample_id") %>% 
+  group_by(top_sample, depth_chr, otu_name) %>% 
+  summarise(rel_abund= sum(rel_abund)) %>% 
+  ungroup()
+
+rbcl_1
+```
+
+```{r}
+ggplot(rbcl_1, aes(x=top_sample, y=rel_abund, fill = otu_name))+
+  geom_bar(stat="identity") +
+  facet_grid(rows = vars(depth_chr %>% fct_relevel("surface","sub-surface"))) +
+  labs(x="Sample ID", y="Relative abundance", fill="OTU")
+```
+
+```{r 18s_wrangle, include=F}
+s18 <- s18_all_rel_abund %>% 
+  left_join(s18_tax, by="asv_id") %>% 
+  left_join(field, by="sample_id") %>% 
+  left_join(depth_sample_info, by="sample_id") %>% 
+  select(top_sample, depth_chr, phylum:genus, rel_abund)
+
+glimpse(s18)
+```
+
+```{r}
+ggplot(s18, aes(x=top_sample, y=rel_abund, fill = phylum))+
+  geom_bar(stat="identity") +
+  facet_grid(rows = vars(depth_chr %>% fct_relevel("surface","sub-surface"))) +
+  labs(x="Sample ID", y="Relative abundance", fill="OTU")
+```
+More fungi on surface, one sample with more rotifers on surface, more ciliates on surface also

### 2019-10-31 code/make_sample_aliases.R
deleted old wrangle files, (now moved into seperate tidy, filter, files) as well as make sample aliases (now part of the field tidy file)

@@ -1,39 +1,0 @@
-# Make sample aliases
-
-library(tidyverse)
-
-field <- read_csv(here("data/02_tidied/field_clean.csv"))
-
-# each mountain gets a letter, each date gets a number, each 
-
-# assign unique mountain_letter code to each field site
-field_1 <- field %>% 
-  mutate(mountain_letter = case_when(str_detect(sample_id, "sax") ~ "X",
-                                     str_detect(sample_id, "sey") ~ "S",
-                                     str_detect(sample_id, "sky") ~ "K",
-                                     str_detect(sample_id, "stm") ~ "M",
-                                     TRUE ~ str_sub(sample_id, 0,1) %>% toupper()))
-
-
-# which visit to the mountain (only 1 visit for most sites)
-visit_number <- field_1 %>%
-  distinct(mountain_letter, date) %>% 
-  arrange(mountain_letter, date) %>% 
-  group_split(mountain_letter) %>% 
-  map_df(rownames_to_column, "visit_number") %>% 
-  mutate(alias = paste0(mountain_letter, visit_number))
-  
-field_2 <- field_1 %>% 
-  left_join(visit_number, by=c("mountain_letter", "date"))
-  
-# for each moutain visit, unique sample ID
-alias <- field_2 %>% 
-  arrange(-elev_m) %>% # from high to low since thats usually the order I collected in
-  group_split(alias) %>% 
-  map_df(rownames_to_column, "uid") %>% 
-  mutate(alias = paste0(alias, ".", uid)) %>% 
-  select(sample_id, alias)
-  
-  
-
-write_csv(alias, here("output/sample_aliases.csv"))

### 2019-10-31 code/wrangle/normalize.rel_abund_tbl.R
deleted old wrangle files, (now moved into seperate tidy, filter, files) as well as make sample aliases (now part of the field tidy file)

@@ -1,31 +1,0 @@
-library(tidyverse)
-library(here)
-library(metagenomeSeq)
-
-seq.rbcl.2019 <- read.csv(here("data","raw","ky_rbcL_seqtab.nochim.txt"), sep=" ", stringsAsFactors = F)
-
-# format for metagenomeSeq
-MR.rbcl <- rbcl_rel_abund %>% 
-  select(-rel_abund) %>% 
-  spread(sample_id, n_reads, fill = 0) %>% # adds 0s back in
-  column_to_rownames("asv_id") %>% 
-  newMRexperiment()
-
-rbcl_norm <- MR.rbcl %>% 
-  cumNorm(p = cumNormStatFast(MR.rbcl)) %>% 
-  MRcounts(norm=T) %>% 
-  as_tibble(rownames = "asv_id") %>% 
-  gather("sample_id","norm_reads", -asv_id)
-
-rbcl_norm <- rbcl_norm %>% 
-  filter(norm_reads!=0) %>% 
-  group_by(sample_id) %>% 
-  mutate(norm_rel_abund = norm_reads/sum(norm_reads))
-
-rbcl_rel_abund %>% 
-  left_join(rbcl_norm, by=c("sample_id","asv_id")) %>% 
-  select(asv_id, sample_id, rel_abund, norm_rel_abund) %>% 
-  mutate(same = if_else(rel_abund == norm_rel_abund, T, F),
-         diff = rel_abund - norm_rel_abund) %>% 
-  arrange(-diff)
-filter(same==F)

### 2019-10-31 code/wrangle/wrangle.R

@@ -1,316 +1,0 @@
-# Wrangle rbcL and 18S ASV tables and taxonomy
-
-library(tidyverse)
-
-
-# data in ------------
-seq.rbcl.2019 <- read.csv("dat/ky_rbcL_seqtab.nochim.txt", sep=" ", stringsAsFactors = F)
-idtax <- read_csv("dat/assignments_9-18.csv") # thresh = 50
-
-asv.tbl.18s <- read.csv("dat/ky_18s_seqtab.nochim_April2019.txt", sep=" ", stringsAsFactors = F)
-s18_taxa_silva <- read_csv("dat/18s_assignments_silva.csv") # thresh 40
-
-field <- read_csv("dat/field_2018.csv")
-cellct <- read_csv("dat/cellct_2018.csv")
-
-
-# rbcL relative abund --------------------
-
-unfilt_rbcl_rel_abund <- seq.rbcl.2019 %>%
-  t() %>% 
-  as_tibble(rownames = "asv") %>% 
-  gather(key = "sample_id", val = "n_reads", -asv)
-
-# format sample ids properly
-unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
-  mutate(sample_id = str_extract(sample_id, ".*(?=_S.*_)") %>% # remove underscores
-           str_replace_all("-","."), # replace hyphen
-         protocol = str_extract(sample_id, "(?<=[:alpha:]{3}18\\..{2}\\.p)[:digit:]"), # parse protocol version used
-         sample_id = str_replace(sample_id, ".p[:digit:]",""), # remove p1
-         protocol = replace_na(protocol, 3)) # replace NA with 3
-
-# compute new variable "rel_abund" 
-############### consider normalizing with metagenomeSeq ( i think this weights for 0s somehow...)
-
-unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
-  group_by(sample_id, protocol) %>% 
-  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
-  ungroup()
-
-# Filter out controls, and samples with low reads ---------------------------
-
-# set aside and remove protocol controls
-protocol.ids <- unfilt_rbcl_rel_abund %>% 
-  select(sample_id, protocol) %>% 
-  distinct() %>% 
-  dplyr::count(sample_id) %>% 
-  filter(n>1) %>% # choose if multiple protocols associated with a sample
-  pull(sample_id)
[…]

### 2019-10-31 code/wrangle/tidy_cell_count.R
init

@@ -1,0 +1,46 @@
+# tidy cell count data
+
+library(tidyverse)
+library(here)
+
+raw_cellct <- read_csv(here("data/01_raw/cellct_2018.csv"))
+glimpse(raw_cellct)
+
+
+
+thresh <- 50 # set this much lower than 100, to keep some counts with high fungi relative abundance
+
+cellct <- raw_cellct %>%
+  # drop non-algae categories
+  select(-three_arm:-biofilm) %>% 
+  # make long format
+  pivot_longer(cols = -sample_id, names_to="morpho_sp", values_to="count") %>%
+  # remove 0's
+  filter(count != 0) %>% 
+  drop_na() %>% 
+  mutate(morpho_sp = case_when( str_detect(morpho_sp, "green")~"Green cell", # lump all green categories
+                                morpho_sp=="orb"~"Sanguina nivaloides", 
+                                morpho_sp=="ruby"~"Sanguina nivaloides", # assuming Sanguina grows turrets 
+                                morpho_sp=="balloon"~"Chlainomonas",
+                                morpho_sp=="tangerine"~"Chloromonas krienitzii",
+                                morpho_sp=="hedgehog"~"Chloromonas cf. brevispina",
+                                morpho_sp=="lemon"~"Chloromonas cf. nivalis",
+                                morpho_sp=="spiny"~"Other",
+                                morpho_sp=="oval"~"Other",
+                                morpho_sp=="zuke"~"Other",
+                                morpho_sp=="halo"~"Other",
+                                morpho_sp=="citrine"~"Sanguina nivaloides",
+                                TRUE ~ morpho_sp)) %>% 
+  # drop samples with count < thresh
+  group_by(sample_id) %>% 
+  mutate(count_total = sum(count)) %>% 
+  filter(count_total > thresh) %>% 
+  # calculate percentage
+  mutate(percent_of_sample = count / count_total) %>% 
+  ungroup()
+
+head(cellct)
+
+
+
+write_csv(cellct, here("data/02_tidied/tidied.cell_count.csv"))

### 2019-10-31 code/wrangle/tidy_field.R
init, includes sample aliases

@@ -1,0 +1,64 @@
+# tidy field data, adds sample aliases
+
+library(tidyverse)
+library(here)
+
+
+
+field <- read_csv(here("data/01_raw/field_2018.csv"))
+
+head(field)
+
+
+# clean up a few small details
+field_1 <- field %>%
+  mutate(
+    # lump some of the color snow categories
+    color_snow = case_when(color_snow %in% c("white","yellow") ~ "other",
+                           color_snow %in% c("purple","red") ~ "pink", 
+                           is.na(color_snow) ~ "other",
+                           TRUE ~ color_snow),
+    # change NA habitat to other
+    habitat = if_else(is.na(habitat), "other", habitat)
+  )
+
+
+
+# make sample aliases ----------------------
+# each mountain gets a letter, each date gets a number, each sample gets unique ID 
+# in the format B1.05 where B is mountain, 1 is the first visit, and 5 is the 5th sample collected that day
+
+# assign unique mountain_letter code to each field site
+field_2 <- field_1 %>% 
+  mutate(mountain_letter = case_when(str_detect(sample_id, "sax") ~ "X",
+                                     str_detect(sample_id, "sey") ~ "S",
+                                     str_detect(sample_id, "sky") ~ "K",
+                                     str_detect(sample_id, "stm") ~ "M",
+                                     TRUE ~ str_sub(sample_id, 0,1) %>% toupper()))
+field_2 %>% select(mountain_letter, sample_id)
+
+# which visit to the mountain (only 1 visit for most sites)
+visit_number <- field_2 %>%
+  distinct(mountain_letter, date) %>% 
+  arrange(mountain_letter, date) %>% 
+  group_split(mountain_letter) %>% 
+  map_df(rownames_to_column, "visit_number") %>% 
+  mutate(alias_1 = paste0(mountain_letter, visit_number))
+
+field_3 <- field_2 %>% 
+  left_join(visit_number, by=c("mountain_letter", "date"))
+field_3 %>% distinct(sample_id, date, mountain_letter, visit_number, alias_1) %>% arrange(date) %>% view()
+
+# for each moutain visit, unique sample ID
+field_4 <- field_3 %>% 
+  arrange(-elev_m) %>% # from high to low since thats usually the order I collected in
+  group_split(alias_1) %>% 
+  map_df(rownames_to_column, "uid") %>% 
+  mutate(alias = paste0(alias_1, ".", uid))
[…]

### 2019-10-31 code/wrangle/old/wrangle.Rmd
moved these into "old" in case I need anything

---
title: "Wrangle"
output: html_notebook
---

```{r}
library(tidyverse)
library(here)
```
Read in DADA2 output files from server, and sample site data
```{r input}
# rbcl asv table, and taxonomy table
seq.rbcl.2019 <- read.csv(here("data","raw","ky_rbcL_seqtab.nochim.txt"), sep=" ", stringsAsFactors = F) ##### KY saved as txt %^&*
idtax <- read_csv(here("data","raw","assignments_9-18.csv")) # thresh = 50

# 18s
asv.tbl.18s <- read.csv(here("data","raw","ky_18s_seqtab.nochim_April2019.txt"), sep=" ", stringsAsFactors = F)
s18_taxa_silva <- read_csv(here("data","raw","18s_assignments_silva.csv")) # thresh 40 ########### redo with thresh = 50 for consistancy

# site metadata
field <- read_csv(here("data","raw","field_2018.csv"))
```

Wrangle rbcL rel abund
```{r rbcl_rel_abund}
# rbcL relative abund --------------------

unfilt_rbcl_rel_abund <- seq.rbcl.2019 %>%
  t() %>% 
  as_tibble(rownames = "asv") %>% 
  gather(key = "sample_id", val = "n_reads", -asv)

# format sample ids properly
unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
  mutate(sample_id = str_extract(sample_id, ".*(?=_S.*_)") %>% # remove underscores
           str_replace_all("-","."), # replace hyphen
         protocol = str_extract(sample_id, "(?<=[:alpha:]{3}18\\..{2}\\.p)[:digit:]"), # parse protocol version used
         sample_id = str_replace(sample_id, ".p[:digit:]",""), # remove p1
         protocol = replace_na(protocol, 3)) # replace NA with 3

# compute new variable "rel_abund" 
############### consider normalizing with metagenomeSeq ( i think this weights for 0s somehow...)

unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
  group_by(sample_id, protocol) %>% 
  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
  ungroup()
unfilt_rbcl_rel_abund
```

Seperate out controls, and samples with low reads
```{r controls}
# set aside and remove protocol controls
protocol.ids <- unfilt_rbcl_rel_abund %>% 
  select(sample_id, protocol) %>% 
  distinct() %>% 
  dplyr::count(sample_id) %>% 
  filter(n>1) %>% # choose if multiple protocols associated with a sample
  pull(sample_id)
protocol_comparison <- unfilt_rbcl_rel_abund %>% 
  filter(sample_id %in% protocol.ids)

# manually look for "protocol" samples to remove
protocol_comparison %>% 
  select(sample_id, protocol) %>% 
  distinct() %>% 
  group_by(sample_id) %>% 
  mutate(min=as.numeric(protocol) %>% min() %>% as.character(), 
         is_min = if_else(protocol == min, TRUE, FALSE)) %>% 
  filter(is_min == TRUE) %>% 
  select(1,2)
# ... and remove them from the main tbl manually
rbcl_rel_abund <- unfilt_rbcl_rel_abund %>%
  filter(!(sample_id == "gar18.01" & protocol == "2"),
         !(sample_id == "sax18.01" & protocol == "1"),
         !(sample_id == "sey18.25" & protocol == "1"),
         !(sample_id == "sey18.74" & protocol == "1"),
         !(sample_id == "sky18.12" & protocol == "1") )

# water negative controls
water_ctrls <- rbcl_rel_abund %>% 
  filter(sample_id %>% str_detect("water"))
rbcl_rel_abund <- rbcl_rel_abund %>% 
  filter(!(sample_id %>% str_detect("water")))

# white snow samples
white_snow_samples <- rbcl_rel_abund %>% 
  filter(sample_id %in% c("bre18.17", "bre18.22.bb"))
rbcl_rel_abund <- rbcl_rel_abund %>% 
  filter( !(sample_id %in% c("bre18.17", "bre18.22.bb")))

# scraped vs unscraped comparison
scrape_comparison <- rbcl_rel_abund %>% 
  filter(sample_id %>% str_detect("sky18.24"))
rbcl_rel_abund <- rbcl_rel_abund %>% 
  filter(!(sample_id == "sky18.24sc"))

# set aside depth samples
depth_sample_rel_abund <- rbcl_rel_abund %>% 
  filter(sample_id %in% c("hol18.42", "hol18.42b", "sey18.66", "sey18.66b", "hol18.21", "hol18.39d"))
rbcl_rel_abund <- rbcl_rel_abund %>% 
  filter(!(sample_id %in% c("hol18.42", "sey18.66b","hol18.39d")))

# remove samples with low read count
rbcl_rel_abund %>% 
  group_by(sample_id, protocol) %>% 
  summarise(sum=sum(n_reads)) %>% 
  arrange(sum)
# remove failed sample lib18.01
rbcl_rel_abund <- rbcl_rel_abund %>% 
  filter(sample_id != "lib18.01" )

# remove outlier sey18.22a, likely cross contam
rbcl_rel_abund <- rbcl_rel_abund %>% 
  filter(sample_id != "sey18.22a")

# now remove rows that contain 0s
rbcl_rel_abund <- rbcl_rel_abund %>% 
  filter(n_reads!=0)

rbcl_rel_abund <- rbcl_rel_abund %>% 
  select(-protocol)

rbcl_rel_abund
```

Wrangle rbcL taxonomy table
```{r taxonomy}
idtax <- idtax %>% 
  dplyr::rename(domain=V2, kingdom=V3, phylum=V4,class=V5 ,order=V6, family=V7, genus=V8, species=V9) %>% 
  dplyr::rename(domainConf=V21,kingdomConf=V31, phylumConf=V41, classConf = V51, orderConf=V61, familyConf=V71, genusConf=V81, speciesConf=V91) %>% 
  mutate_if(is.character, funs(replace(., str_detect(., "unclassified"), NA_character_))) %>% 
  mutate(id_level = case_when(is.na(domain) ~"root",
                              is.na(kingdom) ~"domain",
                              is.na(phylum)~"kingdom",
                              is.na(class)~"phylum",
                              is.na(order)~"class",
                              is.na(family)~"order",
                              is.na(genus)~"family",
                              is.na(species)~"genus",
                              TRUE ~ "species")) %>% 
  mutate(assignment = case_when(id_level =="root"~"root",
                                id_level == "domain"~ domain,
                                id_level == "kingdom"~kingdom,
                                id_level == "phylum"~phylum,
                                id_level == "class"~class,
                                id_level == "order"~order,
                                id_level == "family"~family,
                                id_level == "genus"~genus,
                                id_level == "species"~species))
idtax %>% nrow() # 652 ASVs
# remove ASVs not found in snow samples
idtax <- idtax %>% 
  filter(asv %in% rbcl_rel_abund$asv)
idtax %>% nrow()
```

For convenience, assign each rbcL ASV a numerical ID
```{r asv_key}
# make rbcl asv id key
rbcl_asv_id_key <- idtax %>% 
  select(asv) %>% 
  rownames_to_column("asv_id") %>% 
  mutate(asv_id = str_pad(asv_id,3,pad="0")) %>% 
  arrange(asv_id)

# add in asv_id to asv tbl
rbcl_rel_abund <- rbcl_asv_id_key %>% 
  right_join(rbcl_rel_abund, by="asv") %>% 
  arrange(asv_id) %>% 
  select(-asv)

# add asv_id to taxonomy tbl
idtax <- rbcl_asv_id_key %>% 
  right_join(idtax, by="asv") %>% 
  select(-asv) %>% 
  arrange(asv_id)

idtax %>% distinct(genus)
```

Manually change some of the taxonomy that was incorrectly labelled
```{r}
idtax <- idtax %>% 
  mutate(genus = case_when(str_detect(genus,"Stichococcus")~"Stichococcus",
                           str_detect(genus,"Myrmecia")~"Myrmecia",
                           str_detect(genus, "Dicty")~"Dictyochloropsis",
                           str_detect(genus,"Trebouxia")~"Trebouxia",
                           str_detect(family,"Stichococcus")~"Stichococcus", # a mis-alignment of columns in the ref db for this one
                           TRUE~genus),
         species = case_when(str_detect(genus,"Stichococcus")~NA_character_, # not named species, removing
                           TRUE~species),
         id_level = case_when(str_detect(genus,"Stichococcus")~"genus", # and update this
                              TRUE~id_level))
# species list
idtax %>% distinct(species)
```

# 18S ---------------------------

Wrangle 18S rel abundance tables
```{r 18S_wrangle}
unfilt_18s_ra_2019 <- asv.tbl.18s %>%
  t() %>% 
  as_tibble(rownames = "asv")

# change column names
colnames(unfilt_18s_ra_2019)[-1] <- colnames(unfilt_18s_ra_2019)[-1] %>% 
  str_remove("_S.*_") %>% # remove _S*_ ending
  str_replace_all("-",".") #replace hyphen with period

# remove duplocate protocols, remove p from name
colnames(unfilt_18s_ra_2019) # look for dups
unfilt_18s_ra_2019 <- unfilt_18s_ra_2019 %>% 
  select(-gar18.01.p2, -sax18.01.p1, -sey18.25.p1, -sey18.74.p1, -sky18.12.p1)

# now remove p from sample Id
colnames(unfilt_18s_ra_2019)[-1] <- colnames(unfilt_18s_ra_2019)[-1] %>% 
  str_remove("\\.p[:digit:]")

# tidy long form, remove remaining samples not in rbcl data set (water, etc)
s18_rel_abund <- unfilt_18s_ra_2019 %>% 
  gather(key = "sample_id", val = "n_reads", -asv) %>% 
  filter(sample_id %in% rbcl_rel_abund$sample_id)

# add percent asv
s18_rel_abund <- s18_rel_abund %>% 
  group_by(sample_id) %>% 
  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
  ungroup()

# n reads per sample
s18_rel_abund %>% 
  group_by(sample_id) %>% 
  summarise(sum_n = sum(n_reads)) %>% 
  arrange(sum_n) # max 76k, min 21k reads per sample

s18_rel_abund <- s18_rel_abund %>% 
  filter(sample_id != "gar18.04") %>% # remove gar4, no reads (?)
  filter(rel_abund != 0) # remove 0s
s18_rel_abund
```

Wrangle 18S taxonomy
```{r}
s18_taxa_silva <- s18_taxa_silva %>% 
  dplyr::rename(domain=V2,phylum=V3, class=V4 ,order=V5, family=V6, genus=V7, domainConf=V21,phylumConf=V31, classConf=V41, orderConf = V51, familyConf=V61 ,genusConf=V71) %>% 
  mutate_if(is.character, funs(replace(., str_detect(., "unclassified"), NA_character_))) %>% 
  mutate(id_level = case_when(is.na(domain) ~"none",
                              is.na(phylum) ~"domain",
                              is.na(class)~"phylum",
                              is.na(order)~"class",
                              is.na(family)~"order",
                              is.na(genus)~"family",
                              TRUE ~ "genus")) %>% 
  mutate(assignment = case_when(id_level == "none"~ "none",
                                id_level == "domain"~ domain,
                                id_level == "phylum"~phylum,
                                id_level == "class"~class,
                                id_level == "order"~order,
                                id_level == "family"~family,
                                id_level == "genus"~genus,
                                TRUE ~ "ERROR"))
s18_taxa_silva
##### some weird additions to the 18s taxonomy tbl from SILVA, _fa, _ge, only in some cases...
```

For convenience, assign each 18S ASV a numerical ID
```{r}
# make rbcl asv id key
s18_asv_id_key <- s18_taxa_silva %>% 
  select(asv) %>% 
  rownames_to_column("asv_id") %>% 
  mutate(asv_id = str_pad(asv_id,3,pad="0")) %>% 
  arrange(asv_id)

# add in asv_id to asv tbl
s18_rel_abund <- s18_asv_id_key %>% 
  right_join(s18_rel_abund, by="asv") %>% # right_join just a shortcut so that asv_id is first column
  arrange(asv_id) %>% 
  select(-asv)

# add asv_id to taxonomy tbl
s18_taxa_silva <- s18_asv_id_key %>% 
  right_join(s18_taxa_silva, by="asv") %>% 
  select(-asv) %>% 
  arrange(asv_id)

```

Wrangle field data
```{r field}

field <- field %>%
  mutate(
    # make new col with 3 letter mountain code
    mountain = str_sub(sample_id, 1,3), 
    # lump some of the color snow categories
    color_snow = case_when(color_snow %in% c("white","yellow") ~ "other",
                           color_snow %in% c("purple","red") ~ "pink", 
                           is.na(color_snow) ~ "other",
                           TRUE ~ color_snow),
    # change NA habitat to other
    habitat = if_else(is.na(habitat), "other", habitat)
  )
```

```{r output}
rbcl_rel_abund %>% write_csv(here("data","clean","rbcl_rel_abund_clean.csv"))
idtax %>% write_csv(here("data","clean","rbcl_taxonomy_clean.csv"))
rbcl_asv_id_key %>% write_csv(here("data","clean","rbcl_asv_key.csv"))

write_csv(field, here("data","clean","field_clean.csv"))
```

### 2019-10-31 code/wrangle/old/wrangle.R

@@ -1,0 +1,316 @@
+# Wrangle rbcL and 18S ASV tables and taxonomy
+
+library(tidyverse)
+
+
+# data in ------------
+seq.rbcl.2019 <- read.csv("dat/ky_rbcL_seqtab.nochim.txt", sep=" ", stringsAsFactors = F)
+idtax <- read_csv("dat/assignments_9-18.csv") # thresh = 50
+
+asv.tbl.18s <- read.csv("dat/ky_18s_seqtab.nochim_April2019.txt", sep=" ", stringsAsFactors = F)
+s18_taxa_silva <- read_csv("dat/18s_assignments_silva.csv") # thresh 40
+
+field <- read_csv("dat/field_2018.csv")
+cellct <- read_csv("dat/cellct_2018.csv")
+
+
+# rbcL relative abund --------------------
+
+unfilt_rbcl_rel_abund <- seq.rbcl.2019 %>%
+  t() %>% 
+  as_tibble(rownames = "asv") %>% 
+  gather(key = "sample_id", val = "n_reads", -asv)
+
+# format sample ids properly
+unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
+  mutate(sample_id = str_extract(sample_id, ".*(?=_S.*_)") %>% # remove underscores
+           str_replace_all("-","."), # replace hyphen
+         protocol = str_extract(sample_id, "(?<=[:alpha:]{3}18\\..{2}\\.p)[:digit:]"), # parse protocol version used
+         sample_id = str_replace(sample_id, ".p[:digit:]",""), # remove p1
+         protocol = replace_na(protocol, 3)) # replace NA with 3
+
+# compute new variable "rel_abund" 
+############### consider normalizing with metagenomeSeq ( i think this weights for 0s somehow...)
+
+unfilt_rbcl_rel_abund <- unfilt_rbcl_rel_abund %>% 
+  group_by(sample_id, protocol) %>% 
+  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
+  ungroup()
+
+# Filter out controls, and samples with low reads ---------------------------
+
+# set aside and remove protocol controls
+protocol.ids <- unfilt_rbcl_rel_abund %>% 
+  select(sample_id, protocol) %>% 
+  distinct() %>% 
+  dplyr::count(sample_id) %>% 
+  filter(n>1) %>% # choose if multiple protocols associated with a sample
+  pull(sample_id)
[…]

### 2019-10-31 code/2018_sample_map.Rmd
init

--- 2019-10-24 code/map.Rmd
@@
 ---
-title: "Snow algae 2018 map"
-
+title: "Map of 2018 snow algae sample locations"
+output: html_notebook
 ---
 
-```{r libs}
+```{r libs, include=F}
 library(tidyverse)
@@
 library(ggmap)
-library(ggdark) # ggplot themes
-library(googleway) # for getting geographic data from google
-library(ggsn) # add symbols to plot
+
+library(ggdark)
 ```
 
-```{r data_in}
-field <- read_csv(here("data","clean","field_clean.csv"))
+```{r include=F}
+field <- read_csv(here::here("data/02_tidied/tidied_field.csv"))
+cities <- read_csv(here::here("output/map/city_coords.csv"))
+
+load(here::here("output/map/raw_google_map.rda")) # loads .rda called "algae_map"... not sure how avoid using.rda format, maybe png?
 ```
 
-```{r settings}
-key="AIzaSyBJsmyAKTkTg9LT-HdGCCdZdbE2Hjwm-L8"
-register_google(key = key) 
-theme_set(dark_theme_classic())
-```
+```{r all_samples_map, echo=F}
+# one point per mountain, with moutain code, colored by mean elevation
+# site tbl: each mtn, total samples, visits, n sequenced samples
 
-```{r map_setup}
-algae_map <- get_googlemap(c(-123.004722, 49.850556), zoom = 7, maptype = "satellite")
-
-# add in Vancouver and Squamish
-cities <- c("Vancouver, BC, Canada", "Squamish, BC, Canada")
-# get coordinates from google
-google_output <- cities %>% 
-  map(~googleway::google_geocode(.x, key=key))
-glimpse(google_output)
-
-# functions to reach into the list and extract wanted elements
-city_name <- function(x) x$results$address_components[[1]][1,2]
-lat <- function(x) x$results$geometry$location[1,1]
-lon <- function(x) x$results$geometry$location[1,2]
-
-# make tbl with city and coords
-city_coords <- google_output %>% {
-  tibble(
-    name = map_chr(., .f=city_name),
-    lat = map_dbl(., .f=lat),
-    lon = map_dbl(., .f=lon)
-  )
-}
-```
-
-```{r all_samples_map}
 full_map <- ggmap(algae_map, extent = "device") +
   # jitter points to make it easier to read
-  geom_jitter(data = field, 
+  geom_point(data = field, 
               size = 2, alpha=0.5, 
@@
 full_map
-
-ggsave(filename = here::here("figs","map",paste0("full_map.png")),
-       plot = full_map, height = 15, width = 20, units = "cm")
 ```
@@
 
-```{r save_maps}
+```{r save_maps, include=F}
+
+# ggsave(filename = here("")
+
+# for the month by month maps
 save_map <- function(my_month, num){
@@
 
-# generate plots
 save_map("May",1)

### 2019-10-31 code/cellcount.Rmd
updated paths for files to read, minor tweaks of plot

@@
 
-```{r}
+```{r, include=F}
 library(tidyverse)
+library(here)
 ```
 
-```{r}
-path_cellct <- here("data/clean/clean.cellct.csv")
-cellct <- read_csv(path_cellct)
-path_field <- here("data/clean/field_clean.csv")
-field <- read_csv(path_field)
-path_alias <- here("data/sample_aliases.csv")
-alias <- read_csv(path_alias)
+```{r, include=F}
+cellct <- read_csv(here("data/02_tidied/tidied.cell_count.csv"))
+field <- read_csv(here("data/02_tidied/tidied_field.csv"))
+rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
 ```
 
-Add site metadata
-```{r}
+```{r heatmap, echo=F, fig.height=10, fig.width=6}
 p <- cellct %>% 
-  left_join(field, by="sample_id") %>% 
-  left_join(alias, by="sample_id")
-p
-```
-
-Plot as a heatmap
-```{r, fig.height=12, fig.width=7}
+  left_join(field, by="sample_id")
 
@@
 
+my_theme <- theme(axis.text.x = element_text(hjust=1,angle=45))
+
 ggplot(p, aes(y=alias %>% fct_reorder(elev_m), 
               x=morpho_sp)) +
-  geom_tile(aes(alpha = percent_of_sample, fill = morpho_sp )) + # color = "white" %>% fct_reorder(morpho_order)
+  geom_tile(aes(alpha = percent_of_sample, fill = morpho_sp)) + # color = "white" %>% fct_reorder(morpho_order)
+  labs(title="Cell count relative abundance", x="Cell morphology", y= "Sample ID", fill = "Cell morphology", alpha="Relative abundance") +
+  guides(fill=F) +
   scale_alpha(range = c(0.1, 1)) +
-  ylab("Sample ID") +
-  xlab("Cell morphology") +
-  labs(fill = "Cell morphology", alpha="Relative abundance") +
-  scale_x_discrete(position = "bottom") +
-  theme(axis.text.x = element_text(hjust=1,angle=45)) +
-  guides(fill=FALSE)
+  my_theme
 ```

### 2019-10-31 code/wrangle/tidy_field.R
updated read paths, add info whether sequenced or not

@@ -6,12 +6,13 @@ library(here)
 
 
 field <- read_csv(here("data/01_raw/field_2018.csv"))
+seq <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
 
 head(field)
 
 
 # clean up a few small details
-field_1 <- field %>%
+field_added_other_category <- field %>%
   mutate(
     # lump some of the color snow categories
     color_snow = case_when(color_snow %in% c("white","yellow") ~ "other",
@@ -19,8 +20,17 @@ field_1 <- field %>%
                            is.na(color_snow) ~ "other",
                            TRUE ~ color_snow),
     # change NA habitat to other
-    habitat = if_else(is.na(habitat), "other", habitat)
-  )
+    habitat = if_else(is.na(habitat), "other", habitat))
+
+# add info whether the sample was sequenced or not
+samples_w_seq <- seq %>% 
+  distinct(sample_id) %>% 
+  pull()
+
+field_have_seq <- field_added_other_category %>% 
+  mutate(have_seq = if_else(sample_id %in% samples_w_seq, TRUE, FALSE))
+
+
 
 
 
@@ -29,36 +39,40 @@ field_1 <- field %>%
 # in the format B1.05 where B is mountain, 1 is the first visit, and 5 is the 5th sample collected that day
 
 # assign unique mountain_letter code to each field site
-field_2 <- field_1 %>% 
+field_mtn_letter <- field_have_seq %>% 
   mutate(mountain_letter = case_when(str_detect(sample_id, "sax") ~ "X",
                                      str_detect(sample_id, "sey") ~ "S",
                                      str_detect(sample_id, "sky") ~ "K",
                                      str_detect(sample_id, "stm") ~ "M",
                                      TRUE ~ str_sub(sample_id, 0,1) %>% toupper()))
-field_2 %>% select(mountain_letter, sample_id)
+field_mtn_letter %>% select(mountain_letter, sample_id)
 
 # which visit to the mountain (only 1 visit for most sites)
-visit_number <- field_2 %>%
+visit_number <- field_mtn_letter %>%
   distinct(mountain_letter, date) %>% 
   arrange(mountain_letter, date) %>% 
   group_split(mountain_letter) %>% 
   map_df(rownames_to_column, "visit_number") %>% 
   mutate(alias_1 = paste0(mountain_letter, visit_number))
 
[…]

### 2019-10-31 code/misc/attempt_conditoinally_bold_x_axis_labs.R
init

@@ -1,0 +1,55 @@
+# toy -------------
+data <- tibble(labs = c("Oranges", "Apples", "Cucumbers"), counts = c(5, 10, 12), bold = c(T, F, F))
+
+# breaks <- levels(data$labs)
+# str(breaks)
+# Reference breaks by name
+
+ggplot(data = data) + 
+  geom_bar(aes(x = labs, y = counts), stat="identity") + 
+  # scale_x_discrete(label = labels, breaks = breaks) +
+  theme(=
+          
+
+
+
+# my case -------------
+
+samples_w_seq <- rel_abund %>% 
+  distinct(sample_id) %>% 
+  pull()
+
+p <- cellct %>% 
+  left_join(field, by="sample_id") %>% 
+  mutate(have_seq = if_else(sample_id %in% samples_w_seq, TRUE, FALSE))
+
+morpho_order <- c("Green cell",
+                  "Sanguina 'citrine'",
+                  "Sanguina nivaloides",
+                  "Sanguina nivaloides 'turret'",
+                  "Green cell",
+                  "Chlainomonas",
+                  "Chloromonas krienitzii",
+                  "Chloromonas cf. brevispina",
+                  "Chloromonas cf. nivalis",
+                  "Chloromonas 'spiny'",
+                  "Chloromonas 'oval'",
+                  "Chloromonas 'zuke'"
+)
+
+to_bold <- p %>% 
+  distinct(sample_id, have_seq, elev_m) %>% 
+  arrange(elev_m) %>% 
+  pull(have_seq)
+
+my_theme <- theme(axis.text.x = element_text(hjust=1,angle=45),
+                  axis.text.y = element_text(face=if_else(to_bold == T, "bold", "plain"),
+                                             color = if_else(to_bold == T, "black", "grey")))
+
+ggplot(p, aes(y=sample_id %>% fct_reorder(elev_m),
+              x=morpho_sp)) +
+  geom_tile(aes(alpha = percent_of_sample, fill = morpho_sp)) + # color = "white" %>% fct_reorder(morpho_order)
+  labs(title="Cell count relative abundance", x="Cell morphology", y= "Sample ID", fill = "Cell morphology", alpha="Relative abundance") +
+  guides(fill=F) +
+  scale_alpha(range = c(0.1, 1)) +
+  my_theme

### 2019-10-31 code/misc/download_google_map_info.R
init

@@ -1,0 +1,44 @@
+# get google map
+
+library(tidyverse)
+library(here)
+library(googleway)
+library(ggmap)
+
+
+# get base map --------------
+key="AIzaSyBJsmyAKTkTg9LT-HdGCCdZdbE2Hjwm-L8"
+register_google(key = key) 
+
+algae_map <- get_googlemap(c(-123.004722, 49.850556), zoom = 7, maptype = "satellite")
+
+
+# get city data ------------------
+cities <- c("Vancouver, BC, Canada", "Squamish, BC, Canada")
+
+# get coordinates from google
+google_output <- cities %>% 
+  map(~googleway::google_geocode(.x, key=key))
+glimpse(google_output)
+
+# functions to reach into the list and extract wanted elements
+city_name <- function(x) x$results$address_components[[1]][1,2]
+lat <- function(x) x$results$geometry$location[1,1]
+lon <- function(x) x$results$geometry$location[1,2]
+
+# make tbl with city and coords
+city_coords <- google_output %>% {
+  tibble(
+    name = map_chr(., .f=city_name),
+    lat = map_dbl(., .f=lat),
+    lon = map_dbl(., .f=lon)
+  )
+}
+
+head(city_coords)
+
+# save data
+dir.create(here::here("output/map"))
+
+save(algae_map, file = here::here("output/map/raw_google_map.rda"))
+write_csv(city_coords, here::here("output/map/city_coords.csv"))

### 2019-10-31 code/misc/make_sample_list.R

@@ -1,0 +1,20 @@
+# which samples are in the main, and depth analysis (for looking for photos)
+library(tidyverse)
+library(here)
+
+main <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
+depth <- read_csv(here("data/03_filtered/rbcl/depth_rbcl_rel_abund.csv"))
+field <- read_csv(here("data/02_tidied/tidied_field.csv"))
+
+main %>% 
+  bind_rows(depth) %>% 
+  distinct(sample_id) %>% 
+  view()
+
+main %>% 
+  distinct(sample_id) %>% 
+  left_join(field, by="sample_id") %>% 
+  arrange(-elev_m) %>% 
+  select(sample_id, alias, elev_m) %>% 
+  view()
+  

### 2019-10-31 code/misc/make_session_info_table.R

@@ -1,0 +1,32 @@
+# print session info
+
+library(sessioninfo)
+library(here)
+library(tidyverse)
+library(kableExtra)
+
+my_session_info <- sessioninfo::session_info()
+  
+my_session_info$platform %>% 
+  unlist() %>% 
+  enframe() %>% 
+  kable() %>% 
+  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F, position = "left")
+
+
+
+software <- my_session_info$packages %>% 
+  as_tibble() %>% 
+  select(package, loadedversion) %>% 
+  dplyr::rename(version = loadedversion)
+
+cedar_info <- tibble(package = c(""),
+                     version = c(""))
+
+software <- software %>% 
+  bind_rows(cedar_info)
+
+software %>% 
+  kable() %>% 
+  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F, position = "left")
+

### 2019-10-31 code/rbcl_asv_clustering.Rmd
new version of asv clustering, copied over from old asv_viz

--- 2019-10-24 code/asv_viz.Rmd
@@
 ---
-title: "ASV visualization"
+title: "rbcL ASV clustering"
+output: html_notebook
 ---
 
-```{r libs, echo=F}
+```{r libs, include=F}
 library(tidyverse)
@@
 library(dbscan)
+
+library(ggpubr)
 library(ggthemes)
-library(ggpubr)
-# library(plotly)
-# library(crosstalk)
-# library(DT)
+library(ggdark)
 ```
 
-```{r set}
-theme_set(theme_few())
-set.seed(123)
+```{r read, include=F}
+asv_key <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
+rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
+taxonomy <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
 ```
 
-```{r read}
-rbcl_asv_id_key <- read_csv(here("output","keys","rbcl_asv_key.csv"))
-rbcl_rel_abund <- read_csv(here("data","clean","rbcl_rel_abund_clean.csv"))
-idtax <- read_csv(here("data","clean","rbcl_taxonomy_clean.csv"))
-```
+```{r asv_distance_matrix, include=F}
 
-Align rbcL ASVs
-```{r align}
-# view alignments, make distance matrix
-asv.dss <- rbcl_asv_id_key$asv %>%
-  DNAStringSet()
-names(asv.dss) <- rbcl_asv_id_key$asv_id
-asv.aligns <- asv.dss %>% 
+aligns <- asv_key %>%
+  filter(asv_id %in% rel_abund$asv_id) %>% # only want asvs used in the main analysis
+  deframe() %>% 
+  DNAStringSet() %>% 
   DECIPHER::AlignSeqs(verbose=F)
 
-asv.aligns %>% DECIPHER::BrowseSeqs(highlight = T) # 600 has a ~70 bp insert, 607 ~45 bp deletion, 508 ~110 bp deletion
-length(asv.aligns)
-```
+# check alignments
+aligns %>% DECIPHER::BrowseSeqs(highlight = T) # ASV 639 has a ~70 bp insert, 646 ~45 bp deletion, 535 ~110 bp deletion
+length(aligns)
 
-Make distance matrix
-```{r dists}
-dists <-  asv.aligns %>% 
+dists <-  aligns %>% 
   DECIPHER::DistanceMatrix(type="matrix", correction = "JC")
@@
 
-t-SNE with distance matrix
-```{r tsne}
+```{r tsne_1, echo=F}
 perplexity <- 30
+
+set.seed(123)
 
@@
 ggplot(tsne, aes(x=V1,y=V2))+
-  geom_point()
+  geom_point() +
+  labs(title="rbcL ASV clustering with t-SNE")
 ```
 
-Add taxonomy assignments, see if t-SNE agrees with the reference-based clustering of IDTaxa
-```{r genus_tsne}
+```{r genus_tsne, echo=F}
 tsne <- tsne %>% 
   add_column(asv_id = colnames(dists)) %>% # label each point by it's ASV ID
-  left_join(idtax, by="asv_id")
+  left_join(taxonomy, by="asv_id")
 
 ggplot(tsne, aes(x=V1,y=V2))+
-  geom_point(aes(color = genus))
+  geom_point(aes(color = genus)) +
+  labs(title = "Points labeled by genus")
 ```
-Seperates out the Sanguia, Chlainomonas. Lumps all Trebouxiophyceae together. Lots of clustering within Chloromonas, some big clusters with only one point assigned to Chloromonas.
 
-check at species level
-```{r sp_tsne}
+```{r sp_tsne, echo=F}
 ggplot(tsne, aes(x=V1,y=V2))+
-  geom_point(aes(color = species))
+  geom_point(aes(color = species)) +
+  labs(title="Pts labeled by species")
 ```
@@
 Make it look nice, show both genus and species level assignments
-```{r tsne-plot1}
+```{r tsne-plot1, echo=F}
 
@@
 
-tsne_plot <- ggplot(tsne, aes(x=V1, y=V2)) +
+ggplot(tsne, aes(x=V1, y=V2)) +
   # plot points not assigned to genus or species
@@
   geom_point(data = filter(tsne, !is.na(display_taxa)), 
-             aes(color = display_taxa %>% fct_relevel(taxa.order), alpha = 0.8), size = 3)  
-tsne_plot
+             aes(color = display_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
+  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
+  theme_few()
 ```
 
-cluster t-SNE into OTUs
-```{r otu-cluster}
+```{r otu-cluster, echo=F}
 eps <- 4
@@
   as.matrix()
+
+set.seed(123)
 
@@
 ggplot(tsne, aes(x=V1, y=V2)) +
-  geom_point(aes(color = otu))
+  geom_point(aes(color = otu)) +
+  labs(title="dbscan clustering")
 ```
 
-manually add *Chloromonas krienitzii* cluster, since this is highly abundant, and split *Chlainomonas* into two, based on the observation that the two clades of *Chlainomonas* seperate out by elevation
-```{r edit-otus}
+Manually edit the clustering:
+  - *Chloromonas krienitzii* gets its own cluster, since this is highly abundant
+  - split *Chlainomonas* into two, based on the observation in IQTree the two clades of *Chlainomonas* seperate out by elevation
+
+```{r edit-otus, echo=F}
 tsne <- tsne %>%
@@
 
-tsne_plot <- tsne_plot +
-  ggforce::geom_mark_ellipse(data = tsne, aes(fill=otu_name, label=otu_name))
+ggplot(tsne, aes(x=V1, y=V2)) +
+  # plot points not assigned to genus or species
+  geom_point(data = filter(tsne, is.na(display_taxa)), 
+             color = "grey",alpha = 0.8, size = 3) +
+  # color points that were assigned to genus or species
+  geom_point(data = filter(tsne, !is.na(display_taxa)), 
+             aes(color = display_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
+  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
+  theme_few() +
+  ggforce::geom_mark_ellipse(data = tsne, aes(fill=otu_name, label=otu_name), show.legend = FALSE)
 
-# set
+```
+
+```{r include_rel_abund, echo=F, fig.width=8, fig.height=6}
+total_rel_abund <- rel_abund %>% 
+  group_by(asv_id) %>% 
+  summarise(global_rel_abund = sum(rel_abund))
+
+tsne_2 <- tsne %>% 
+  left_join(total_rel_abund, by="asv_id")
+
+ggplot(tsne_2, aes(x=V1, y=V2)) +
+  ggforce::geom_mark_ellipse(data = tsne, aes(fill=otu_name, label=otu_name), show.legend = FALSE) +
+  # plot points not assigned to genus or species
+  geom_point(data = filter(tsne_2, is.na(display_taxa)), 
+             aes(size = global_rel_abund),
+             color = "grey",alpha = 0.7) +
+  # color points that were assigned to genus or species
+  geom_point(data = filter(tsne_2, !is.na(display_taxa)), 
+             aes(color = display_taxa %>% fct_relevel(taxa.order),
+                 size = global_rel_abund), 
+             alpha = 0.7) +
+  labs(title = "rbcL ASV t-SNE clustering", color="ASV taxonomic assignment") +
+  theme_few() +
+  scale_size(range = c(1, 6), guide="none")
+```
+
+```{r tsne-output, echo=F}
+
+tsne_plot +
+  theme(legend.position = "none")
 
@@
   get_legend() %>% 
-  as_ggplot()
-ggsave(here("figs","asv-cluster","asv-legend.pdf"), height = 7, width = 3)
+  as_ggplot
 
-# plot and save the main plot
-tsne_plot +
-  theme(legend.position = "none")
-ggsave(here("figs","asv-cluster","asv-cluster.pdf"), height = 5, width = 5)
+ggsave(here("figs","asv-cluster","asv-legend-dark.pdf"), height = 7, width = 3)
+ggsave(here("figs","asv-cluster","asv-cluster-dark.pdf"), height = 5, width = 5)
+```
+
+```{r asv_mds, echo=F}
+
 ```
@@
 Run IQtree
-```{r}
+```{r tree, fig.width=8, fig.height=20}
 rbcl.tree <- read.tree(here("output","tree","rbcl","asv613.fasta.treefile"))

### 2019-10-31 code/scratch/asv_viz.Rmd
moved this to scratch

--- 2019-10-24 code/asv_viz.Rmd
@@
 library(ggpubr)
+library(ggdark)
 # library(plotly)
@@
 theme_set(theme_few())
-set.seed(123)
 ```
@@
 perplexity <- 30
+
+set.seed(123)
 
@@
 
+set.seed(123)
+
 dbscan.out <- dbscan::dbscan(tsne.mat, eps=eps) # eps=4 gives best results
@@
   get_legend() %>% 
-  as_ggplot()
-ggsave(here("figs","asv-cluster","asv-legend.pdf"), height = 7, width = 3)
+  as_ggplot 
+ggsave(here("figs","asv-cluster","asv-legend-dark.pdf"), height = 7, width = 3)
 
@@
   theme(legend.position = "none")
-ggsave(here("figs","asv-cluster","asv-cluster.pdf"), height = 5, width = 5)
+ggsave(here("figs","asv-cluster","asv-cluster-dark.pdf"), height = 5, width = 5)
 ```

### 2019-11-01 code/rbcl_asv_clustering.Rmd
added mds

@@
 library(dbscan)
+library(ggrepel)
 
@@
 set.seed(123)
-
-tsne <- dists %>% 
+tsne_full_output <- dists %>% 
   Rtsne(pca=F, perplexity = perplexity)
 
-tsne <- tsne$Y %>% 
+tsne_pts <- tsne_full_output$Y %>% 
   as_tibble()
 
-ggplot(tsne, aes(x=V1,y=V2))+
+ggplot(tsne_pts, aes(x=V1,y=V2))+
   geom_point() +
@@
 ```{r genus_tsne, echo=F}
-tsne <- tsne %>% 
+tsne_pts_taxonomy <- tsne_pts %>% 
   add_column(asv_id = colnames(dists)) %>% # label each point by it's ASV ID
@@
 
-ggplot(tsne, aes(x=V1,y=V2))+
+ggplot(tsne_pts_taxonomy, aes(x=V1,y=V2))+
   geom_point(aes(color = genus)) +
@@
 ```{r sp_tsne, echo=F}
-ggplot(tsne, aes(x=V1,y=V2))+
+ggplot(tsne_pts_taxonomy, aes(x=V1,y=V2))+
   geom_point(aes(color = species)) +
@@
 
-# make new categorical to tell ggplot what taxonomy to plot
-# include both species and genus level assignments
-# lump low abundance Trebouxiophyceae
-treb <- c("Myrmecia","Trebouxia","Coccomyxa","Dictyochloropsis","Diplosphaera")
-tsne <- tsne %>% 
-    mutate(display_taxa = case_when(genus %in% treb ~"other Trebouxiophyceae",
-                                  genus == "Stichococcus" ~ "Stichococcus",#########not sure why I need this...
-                                  id_level == "species" ~ assignment,
-                                  id_level == "genus" ~ assignment))
-
-# order
 taxa.order <- c("Chloromonas","Chloromonas_muramotoi","Chloromonas_hohamii","Chloromonas_krienitzii","Chloromonas_Group_B","Raphidonema_nivale","Stichococcus","other Trebouxiophyceae","Chlainomonas", "Chlainomonas_rubra","Sanguina","Sanguina_aurantia","Sanguina_nivaloides" )
 
-ggplot(tsne, aes(x=V1, y=V2)) +
+ggplot(tsne_pts_taxonomy, aes(x=V1, y=V2)) +
   # plot points not assigned to genus or species
-  geom_point(data = filter(tsne, is.na(display_taxa)), 
-             color = "grey",alpha = 0.8, size = 3) +
+  geom_point(data = filter(tsne_pts_taxonomy, is.na(plot_taxa)), 
+             color = "grey", alpha = 0.8, size = 3) +
   # color points that were assigned to genus or species
-  geom_point(data = filter(tsne, !is.na(display_taxa)), 
-             aes(color = display_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
+  geom_point(data = filter(tsne_pts_taxonomy, !is.na(plot_taxa)), 
+             aes(color = plot_taxa %>% fct_relevel(taxa.order)), 
+             alpha = 0.8, size = 3) +
   labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
@@
 
-tsne.mat <- tsne %>% 
+tsne_mat <- tsne_pts_taxonomy %>% 
   select(1,2) %>% 
@@
 
-dbscan.out <- dbscan::dbscan(tsne.mat, eps=eps) # eps=4 gives best results
+dbscan_out <- dbscan::dbscan(tsne_mat, eps=eps) # eps=4 gives best results
 
-tsne <- tsne %>% 
-  add_column(otu = as.character(dbscan.out$cluster)) # tsne is in order of asv_id
+tsne_taxa_otu <- tsne_pts_taxonomy %>% 
+  add_column(otu = as.character(dbscan_out$cluster)) # tsne is in order of asv_id
 
-ggplot(tsne, aes(x=V1, y=V2)) +
+ggplot(tsne_taxa_otu, aes(x=V1, y=V2)) +
   geom_point(aes(color = otu)) +
@@
 
-```{r edit-otus, echo=F}
-tsne <- tsne %>%
+```{r edit-otus, echo=F, fig.width=8, fig.height=6}
+tsne_edited_otus <- tsne_taxa_otu %>%
   mutate(otu = case_when((V2 < (-5) & V1 < -17) ~ "9", # chlain "high"
@@
 
-tsne %>% 
-  select(display_taxa, otu) %>% 
+tsne_edited_otus %>% 
+  select(plot_taxa, otu) %>% 
   distinct() %>% 
@@
 # name each OTU
-otu_names <- tibble(otu = as.character(1:10), otu_name = c("Chloromonas spp.
-C","Sanguina A","Chlainomonas B", "Chloromonas muramotoi E","Chloromonas G","Chloromonas I","Algae H","Trebouxiophyceae J","Chlainomonas F","Chloromonas krienitzii D"))
+otu_names <- tibble(otu = as.character(1:10), 
+                    otu_name = c("Chloromonas spp. C",
+                                 "Sanguina A","Chlainomonas B",
+                                 "Chloromonas muramotoi E",
+                                 "Chloromonas G",
+                                 "Chloromonas I",
+                                 "Algae H",
+                                 "Trebouxiophyceae J",
+                                 "Chlainomonas F",
+                                 "Chloromonas krienitzii D"))
 
-tsne <- tsne %>% 
+tsne_otu_names <- tsne_edited_otus %>% 
   left_join(otu_names, by="otu")
 
-asv_otu_key <- tsne %>% 
+# make a key for downstream use (write to file at end of script)
+asv_otu_key <- tsne_otu_names %>% 
   select(asv_id, otu, otu_name)
 
-ggplot(tsne, aes(x=V1, y=V2)) +
+ggplot(tsne_otu_names, aes(x=V1, y=V2)) +
   # plot points not assigned to genus or species
-  geom_point(data = filter(tsne, is.na(display_taxa)), 
+  geom_point(data = filter(tsne_otu_names, is.na(plot_taxa)), 
              color = "grey",alpha = 0.8, size = 3) +
   # color points that were assigned to genus or species
-  geom_point(data = filter(tsne, !is.na(display_taxa)), 
-             aes(color = display_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
+  geom_point(data = filter(tsne_otu_names, !is.na(plot_taxa)), 
+             aes(color = plot_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
   labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
   theme_few() +
-  ggforce::geom_mark_ellipse(data = tsne, aes(fill=otu_name, label=otu_name), show.legend = FALSE)
-
+  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), show.legend = FALSE)
 ```
@@
 
-tsne_2 <- tsne %>% 
+tsne_rel_abund <- tsne_otu_names %>% 
   left_join(total_rel_abund, by="asv_id")
 
-ggplot(tsne_2, aes(x=V1, y=V2)) +
-  ggforce::geom_mark_ellipse(data = tsne, aes(fill=otu_name, label=otu_name), show.legend = FALSE) +
+final_tsne <- ggplot(tsne_rel_abund, aes(x=V1, y=V2)) +
+  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), 
+                             show.legend = FALSE, 
+                             alpha=0.2, 
+                             linetype="blank", 
+                             expand = unit(2, "mm")) +
   # plot points not assigned to genus or species
-  geom_point(data = filter(tsne_2, is.na(display_taxa)), 
+  geom_point(data = filter(tsne_rel_abund, is.na(plot_taxa)), 
              aes(size = global_rel_abund),
@@
   # color points that were assigned to genus or species
-  geom_point(data = filter(tsne_2, !is.na(display_taxa)), 
-             aes(color = display_taxa %>% fct_relevel(taxa.order),
+  geom_point(data = filter(tsne_rel_abund, !is.na(plot_taxa)), 
+             aes(color = plot_taxa %>% fct_relevel(taxa.order),
                  size = global_rel_abund), 
@@
   labs(title = "rbcL ASV t-SNE clustering", color="ASV taxonomic assignment") +
-  theme_few() +
-  scale_size(range = c(1, 6), guide="none")
+  ggthemes::theme_few() +
+  scale_size(range = c(1, 6), guide="none") ################################################# not changing the scale size
+final_tsne
 ```
@@
 
-tsne_plot +
-  theme(legend.position = "none")
-
-# extract the legend
-tsne_plot %>% 
-  get_legend() %>% 
-  as_ggplot
-
-ggsave(here("figs","asv-cluster","asv-legend-dark.pdf"), height = 7, width = 3)
-ggsave(here("figs","asv-cluster","asv-cluster-dark.pdf"), height = 5, width = 5)
+# finel_tsne +
+#   theme(legend.position = "none")
+# 
+# # extract the legend
+# tsne_plot %>% 
+#   get_legend() %>% 
+#   as_ggplot 
+# 
+# ggsave(here("figs","asv-cluster","asv-legend-dark.pdf"), height = 7, width = 3)
+# ggsave(here("figs","asv-cluster","asv-cluster-dark.pdf"), height = 5, width = 5)
 ```
 
-```{r asv_mds, echo=F}
+```{r asv_mds, echo=F, fig.width=10, fig.height=8}
+mds_coords <- dists %>% 
+  cmdscale() %>% # ok to use euclidian dist, no null values when comparing relatedness
+  as_tibble() %>% 
+  add_column(asv_id = colnames(dists), .before=1)
+
+mds_asv_plot <- mds_coords %>% 
+  left_join(taxonomy, by="asv_id") %>% 
+  left_join(asv_otu_key, by="asv_id") %>% 
+  left_join(total_rel_abund, by="asv_id") %>% 
+  filter(!(asv_id %in% c("639", "534", "646"))) # remove two trebouxiophyceae outliers
+
+ggplot(mds_asv_plot, aes(V1, V2)) + 
+  ggforce::geom_mark_ellipse(data = mds_asv_plot %>% 
+                               filter(otu_name != "Trebouxiophyceae J"), # dont plot the ellipse around other treb
+                             aes(fill = otu_name, label=otu_name),
+                             show.legend = FALSE, 
+                             alpha=0.2, 
+                             linetype="blank",
+                             expand = unit(2, "mm")) +
+  geom_point(aes(alpha = 0.7, color = plot_taxa, size = global_rel_abund)) +
+  labs(title = "MDS clustering rbcL ASVs", color="ASV taxonomic assignment") +
+  scale_size(range = c(1, 6), guide="none") +
+  scale_alpha(guide="none")+
+  theme_few() #+
+  # geom_text(aes(label = asv_id)) # too see outliers, asv_id=c("639", "534")
+```
+
+```{r write_mds, include=F}
 

### 2019-11-01 code/wrangle/tidy_rbcl_assignment.R
added a column, "plot_taxa", basically lumps together a bunch of Trebouxiophyceae for downstream

@@ -2,6 +2,7 @@
 
 library(tidyverse)
 library(here)
+library(Biostrings)
 
 raw_assignments <- read_csv(here("data/01_raw/rbcl_taxonomy_genbank.csv"))
 
@@ -81,10 +82,30 @@ fixed_annotations <- assigns_w_asv_id %>%
                              TRUE~species),
          id_level = case_when(str_detect(genus,"Stichococcus")~"genus", # and update this
                               TRUE~id_level))
+
+
+# make new categorical to tell ggplot what taxonomy to plot
+# include both species and genus level assignments
+# lump low abundance Trebouxiophyceae
+misc_treb <- c("Myrmecia","Trebouxia","Coccomyxa","Dictyochloropsis","Diplosphaera")
+plot_taxonomy <- fixed_annotations %>% 
+  mutate(plot_taxa = case_when(genus %in% misc_treb ~"other Trebouxiophyceae",
+                                  genus == "Stichococcus" ~ "Stichococcus",
+                                  id_level == "species" ~ assignment,
+                                  id_level == "genus" ~ assignment))
+
 # species list
-fixed_annotations %>% distinct(genus)
-fixed_annotations %>% distinct(species)
+plot_taxonomy %>% distinct(genus)
+plot_taxonomy %>% distinct(species)
+plot_taxonomy %>% distinct(plot_taxa)
 
 
-write_csv(fixed_annotations, here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
-write_csv(asv_id, here("output/rbcl_asv_sequence_key.csv"))
+
+write_csv(plot_taxonomy, here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
+write_csv(asv_id, here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
+
+# make this into a fasta as well
+asv_id %>% 
+  deframe() %>% 
+  DNAStringSet() %>% 
+  writeXStringSet(here("data/asv_seq_data/rbcl.fasta"))

### 2019-11-01 manuscript/Casey1.Rmd
wrote the touchy feely field bit, rough draft

@@
 # Introduction {-}
+
+where we were: there are red and green blooms
+where we are: there are multiple types of red and green blooms
 
@@
 # Results {-}
+
+### Morphologically diverse snow algae blooms across a variety of habitats
+
+Snow algae was widespread in our study area, beginning on May _ when we first observed snow algae blooming in the trees.
+Over the course of the season, we collected 309 samples from 33 dates and 13 moutains.
+We did not observe pink snow above treeline until June 6, despite several visits above treeline before that (site data table).
+We observed snow algae until the first snows of winter, although by later in the season most of the snow had melted out at lower elevations.
+
+Visually, several patterns in snow algae site characteristics and cell morphology were apparent.
+Most frequently, we observed pink or red snow, typically in areas recieving high sunlight, dominated by cells that looked similar to *Sanguina nivaloides*, frequently containing green cells of varying shapes and sizes.
+However, red snow could also be dominated by *Chlainomonas* morphologies, in red snow that was visually indistinguishible from red snow containing primarily *S.
+nivaloides*.
+*Chlainomonas* was frequently mixed in with *S.
+nivaloides* as well (Fig.
+1d).
+We also occasionally observed green snow above treeline, which was dominated by green or yellow oval cells with flanges, similar to those described for *Chloromonas muramotoi*.
+Colored snow often appeared in snow runnels, frequently at elevations near or below treeline.
+These runnels containing colored snow formed linear streaks, often with orange on the surface and green snow just below the surface (Fig.
+1b).
+At two sites like this that we visited repeatedly, the runnel began with green snow hidden 1-2 cm below the surface, but on subsequent visits the surface snow had changed color to orange.
+The subsurface snow contained green, often flagellated algae, while surface snow was dominated by orange cells resembling *Chloro.
+krienitzii* (Fig.
+1e) [@Matsuzaki2015].
+Below treeline, orange and green snow was common, often dominated by orange oval cells with short spines, similar to *Chloromonas cf.
+brevispina* (Fig.
+1g).
+Snow dominated by *Chloromonas cf.
+brevispina* and *Chloromonas cf.
+nivalis* was abundnat in shady areas below treeline, often in distinctively bronze colored snow (supp data).
+
+### Diverse *rbc*L reveals species-level snow algae taxonomy
+
+### Snow algae community composition varies with date and elevation
+
+This is supported by our 18S and cell count data. Although we found Chlainomonas
+
+### High elevation, high diversity
+
+### Green snow algae turns orange at the surface
 

### 2019-11-02 code/scratch/tree_scratch.R
scratch init

@@ -1,0 +1,17 @@
+# color edge red if boot < 80
+# ggtree_conf <- rbcl_ggtree$data %>%
+#   separate(label, c("boot", "bayes"), sep = "/") %>% #split label into boots and bayes vals
+#   mutate(insignificant_edge = case_when((boot < 80 & isTip == FALSE)~"dotted",
+#                                         TRUE~"solid"),
+#          asv_id = case_when(isTip == TRUE~boot,
+#                             TRUE~NA_character_)) %>% # round boot to nearest integer
+#   left_join(otus, by="asv_id")
+# 
+# 
+# 
+# rbcl_ggtree %<+% ggtree_conf +
+#   aes(linetype = I(insignificant_edge)) +
+#   geom_tippoint(aes(color=otu_name),alpha=0.7, size=3) + #
+#   theme(legend.position = c(0.2,0.7),
+#         legend.title = element_blank(),
+#         legend.key = element_blank())

### 2019-11-02 code/alignment/rbcl_alignment.R
init

@@ -1,0 +1,43 @@
+# prep rbcL fasta for IQTree
+
+library(tidyverse)
+library(here)
+library(DECIPHER)
+
+
+rbcl <- readDNAStringSet(here("data/asv_seq_data/rbcl.fasta"))
+asv_ids_main_analysis <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv")) %>% distinct(asv_id) %>% pull()
+
+# all alignments ------------
+aligns <- rbcl %>% 
+  AlignSeqs(verbose = F)
+
+BrowseSeqs(aligns, highlight = T)
+
+
+# remove gaps alignments ------------------
+gappy_seqs <- c("534", "639", "646")
+
+length(aligns)
+aligns_no_gaps <- aligns[!(names(aligns) %in% gappy_seqs)]
+length(aligns)
+
+BrowseSeqs(aligns_no_gaps, highlight = T)
+
+# just alignments in "main" analysis --------------
+length(rbcl)
+main_aligns <- rbcl[ names(rbcl) %in% asv_ids_main_analysis ] %>% 
+  AlignSeqs(verbose = F)
+length(main_aligns)
+
+
+
+
+aligns %>% 
+  writeXStringSet(filepath = here("output/tree/rbcl/rbcl_aligns_all.fasta"))
+
+aligns_no_gaps %>% 
+  writeXStringSet(filepath = here("output/tree/rbcl/rbcl_aligns_indels_removed.fasta"))
+
+main_aligns %>% 
+  writeXStringSet(filepath = here("output/alignments/rbcl_aligns_in_main.fasta"))

### 2019-11-02 code/rbcl_asv_clustering.Rmd
removed Chlainomonas high OTU

@@
 ```{r read, include=F}
-asv_key <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
+
 rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
 taxonomy <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
+
+aligns <- readDNAStringSet(here("output/alignments/rbcl_aligns_in_main.fasta"))
+tree <- read.tree(file=here("output/tree/rbcl/asv613.fasta.treefile")) # this is the tree with only the ASVs that appear in the filtered rel abund dataset (not including protocol stuff, etc)
 ```
@@
 ```{r asv_distance_matrix, include=F}
-
-aligns <- asv_key %>%
-  filter(asv_id %in% rel_abund$asv_id) %>% # only want asvs used in the main analysis
-  deframe() %>% 
-  DNAStringSet() %>% 
-  DECIPHER::AlignSeqs(verbose=F)
-
-# check alignments
-aligns %>% DECIPHER::BrowseSeqs(highlight = T) # ASV 639 has a ~70 bp insert, 646 ~45 bp deletion, 535 ~110 bp deletion
-length(aligns)
+BrowseSeqs(aligns, highlight = T)
 
@@
   geom_point(data = filter(tsne_pts_taxonomy, is.na(plot_taxa)), 
-             color = "grey", alpha = 0.8, size = 3) +
+             color = "grey", alpha = 0.8) +
   # color points that were assigned to genus or species
@@
              aes(color = plot_taxa %>% fct_relevel(taxa.order)), 
-             alpha = 0.8, size = 3) +
-  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
-  theme_few()
+             alpha = 0.8) +
+  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment")
 ```
@@
   - *Chloromonas krienitzii* gets its own cluster, since this is highly abundant
-  - split *Chlainomonas* into two, based on the observation in IQTree the two clades of *Chlainomonas* seperate out by elevation
 
-```{r edit-otus, echo=F, fig.width=8, fig.height=6}
+```{r edit-otus, echo=F}
 tsne_edited_otus <- tsne_taxa_otu %>%
-  mutate(otu = case_when((V2 < (-5) & V1 < -17) ~ "9", # chlain "high"
-                         (V2 < 14  & V2 > 8 & V1 > (-20) & V1 < (-12)) ~ "10",# krienitzii
+  mutate(otu = case_when((V2 < 14  & V2 > 8 & V1 > (-20) & V1 < (-12)) ~ "9",# krienitzii
                          TRUE ~ otu))
 
+otus <- tsne_edited_otus %>% 
+  select(asv_id, otu)
+
+# get names for each otu
+rel_abund %>% 
+  left_join(otus, by="asv_id") %>% 
+  left_join(taxonomy, by="asv_id") %>% 
+  distinct(otu, plot_taxa) %>% 
+  arrange(otu)
+
+# get rel abund of each otu
+rel_abund %>% 
+  left_join(otus, by="asv_id") %>% 
+  group_by(otu) %>% 
+  summarise(rel_abund = sum(rel_abund)) %>% 
+  arrange(-rel_abund) %>% 
+  select(-rel_abund) %>% 
+  add_column(letter = LETTERS[1:9])
+
+# name each OTU, lettered in order of rel abund
+otu_names <- tibble(otu = as.character(1:9), 
+                    otu_name = c("Chloromonas spp. C",
+                                 "Sanguina B",
+                                 "Chlainomonas A",
+                                 "Chloromonas muramotoi E",
+                                 "Chloromonas F",
+                                 "Chloromonas H",
+                                 "Algae G",
+                                 "Trebouxiophyceae I",
+                                 "Chloromonas krienitzii D"))
+
+otus_asv_name <- otus %>% 
+  left_join(otu_names, by="otu") %>% 
+  select(asv_id, otu_name)
+```
+
+```{r echo=F, fig.width=8, fig.height=6}
 tsne_edited_otus %>% 
@@
   arrange(otu)
-
-# name each OTU
-otu_names <- tibble(otu = as.character(1:10), 
-                    otu_name = c("Chloromonas spp. C",
-                                 "Sanguina A","Chlainomonas B",
-                                 "Chloromonas muramotoi E",
-                                 "Chloromonas G",
-                                 "Chloromonas I",
-                                 "Algae H",
-                                 "Trebouxiophyceae J",
-                                 "Chlainomonas F",
-                                 "Chloromonas krienitzii D"))
-
-tsne_otu_names <- tsne_edited_otus %>% 
-  left_join(otu_names, by="otu")
 
@@
 ggplot(tsne_otu_names, aes(x=V1, y=V2)) +
+  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), alpha=0.3, show.legend = FALSE) +
   # plot points not assigned to genus or species
@@
   labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
-  theme_few() +
-  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), show.legend = FALSE)
+  theme_few() 
 ```
 
-```{r include_rel_abund, echo=F, fig.width=8, fig.height=6}
+```{r include_rel_abund, echo=F}
 total_rel_abund <- rel_abund %>% 
@@
              aes(size = global_rel_abund),
-             color = "grey",alpha = 0.7) +
+             color = "grey",alpha = 0.5) +
   # color points that were assigned to genus or species
@@
                  size = global_rel_abund), 
-             alpha = 0.7) +
-  labs(title = "rbcL ASV t-SNE clustering", color="ASV taxonomic assignment") +
+             alpha = 0.5) +
+  labs(color="ASV taxonomic assignment") +
   ggthemes::theme_few() +
-  scale_size(range = c(1, 6), guide="none") ################################################# not changing the scale size
+  scale_size(range = c(1, 6), guide="none") + ################################################# not changing the scale size
+  theme(legend.position = "none")
 final_tsne
@@
 
-```{r tsne-output, echo=F}
-
-# finel_tsne +
-#   theme(legend.position = "none")
-# 
-# # extract the legend
-# tsne_plot %>% 
-#   get_legend() %>% 
-#   as_ggplot 
-# 
-# ggsave(here("figs","asv-cluster","asv-legend-dark.pdf"), height = 7, width = 3)
-# ggsave(here("figs","asv-cluster","asv-cluster-dark.pdf"), height = 5, width = 5)
-```
-
-```{r asv_mds, echo=F, fig.width=10, fig.height=8}
+```{r asv_mds, echo=F}
 mds_coords <- dists %>% 
@@
 
-ggplot(mds_asv_plot, aes(V1, V2)) + 
+mds <- ggplot(mds_asv_plot, aes(V1, V2)) + 
   ggforce::geom_mark_ellipse(data = mds_asv_plot %>% 
@@
   geom_point(aes(alpha = 0.7, color = plot_taxa, size = global_rel_abund)) +
-  labs(title = "MDS clustering rbcL ASVs", color="ASV taxonomic assignment") +
   scale_size(range = c(1, 6), guide="none") +
   scale_alpha(guide="none")+
-  theme_few() #+
+  theme_few() +
+  labs(color="ASV taxonomic assignment")
   # geom_text(aes(label = asv_id)) # too see outliers, asv_id=c("639", "534")
-```
-
-```{r write_mds, include=F}
-
+mds
 ```
@@
 
-Write fasta for IQTree web serv
-```{r}
-asv.aligns %>% writeXStringSet(here("output","tree","rbcl","rbcl_aligns.fasta"))
-```
-
 Run IQtree
 ```{r tree, fig.width=8, fig.height=20}
-rbcl.tree <- read.tree(here("output","tree","rbcl","asv613.fasta.treefile"))
+# root tree with Trebouxiophyceae
+root <- taxonomy %>% 
+  filter(class=="Trebouxiophyceae") %>% 
+  head(1) %>% 
+  pull(asv_id)
 
-idtax %>% filter(species == "Raphidonema_nivale")
-# re-root tree
-rbcl.tree <- rbcl.tree %>% root(outgroup = "559", resolve.root=T)
+rbcl.tree <- rbcl.tree %>% root(outgroup = root, resolve.root=T)
 
@@
 # Plot trees
-rbcl_ggtree %<+% select(tsne, asv_id, display_taxa) + # attach taxonomic assignment for tip label
-  geom_tippoint(aes(color=display_taxa, alpha=0.7), size=3) + # 
+
+# tips colored by taxonomy
+taxonomy_tree <- rbcl_ggtree %<+% taxonomy + 
+  geom_tippoint(aes(color=plot_taxa, alpha=0.7), size=3) + # 
   geom_text2(data=rbcl_ggtree_data, aes(label=paste0(boot,";",bayes)),  vjust= 0.01, hjust = 1, size=3) + # add bootstrap vals
@@
 
-rbcl_ggtree %<+% asv_otu_key + # attach OTU
+# clades annotated by OTU
+otu_tree <- rbcl_ggtree %<+% asv_otu_key + 
   geom_tippoint(aes(color=otu_name, alpha=0.7), size=3) + # 
@@
 
-# color edge red if boot < 80
-# ggtree_conf <- rbcl_ggtree$data %>% 
-#   separate(label, c("boot", "bayes"), sep = "/") %>% #split label into boots and bayes vals
-#   mutate(insignificant_edge = case_when((boot < 80 & isTip == FALSE)~"dotted", 
-#                                         TRUE~"solid"),
-#          asv_id = case_when(isTip == TRUE~boot,
-#                             TRUE~NA_character_)) %>% # round boot to nearest integer
-#   left_join(otu_tbl, by="asv_id")
+otu_tree
 
-# rbcl_ggtree %<+% ggtree_conf +
-#   aes(linetype = I(insignificant_edge)) +
-#   geom_tippoint(aes(color=otu),alpha=0.7, size=3) + #
-#   theme(legend.position = c(0.2,0.7), 
-#         legend.title = element_blank(), 
-#         legend.key = element_blank()) 
-# ggsave("figs/ggtree.pdf", height = 10, width=7.5)
 ```
 
+todo: 
+- better color choices
+- circular tree
+
 ```{r output}
+write_csv(otus_asv_name, path = here("output/asv_otu_key.csv"))
 
-write_csv(asv_otu_key, here("output","keys","asv_otu_key.csv"))
+ggsave(filename = here("manuscript/figs/2_taxonomy/tsne.png"), plot=final_tsne, height = 6, width = 8)
+ggsave(filename = here("manuscript/figs/2_taxonomy/mds.png"), plot=mds,  height = 6, width = 8)
+ggsave(filename = here("manuscript/figs/tree.png"), plot=otu_tree, width=8, height = 20)
 ```

### 2019-11-02 code/rbcl_community_compare.Rmd
combined UniFrac and rel abund into multi panel

@@
 ---
+
+```{r, echo=FALSE}
+knitr::opts_chunk$set(fig.path='figs/', echo=FALSE, warning=FALSE, message=FALSE)
+```
 
@@
 library(picante)
+library(gridExtra)
 
@@
 ```{r input, include=F}
-rbcl_rel_abund <- read_csv(here("data/clean/rbcl_rel_abund_clean.csv"))
-idtax <- read_csv(here("data/clean/rbcl_taxonomy_clean.csv"))
-field <- read_csv(here("data/clean/field_clean.csv"))
+rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv")) # not looking at the depth samples, controls, etc
+tax <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
+field <- read_csv(here("data/02_tidied/tidied_field.csv"))
 
-rbcl_asv_id_key <- read_csv(here("intermediate_output/rbcl_asv_key.csv"))
-rbcl_asv_otu_key <- read_csv(here("intermediate_output/asv_otu_key.csv"))
-alias <- read_csv(here("data/sample_aliases.csv"))
-rbcl.tree <- read.tree(here("intermediate_output/tree/rbcl/asv613.fasta.treefile"))
+asv_seqs <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
+otus <- read_csv(here("output/asv_otu_key.csv"))
+
+rbcl.tree <- read.tree(here("output/tree/rbcl/all_asvs/rbcl_aligns_all.fasta.treefile")) ### this tree should only contain the asvs we found in the "main" sites, not looking at controls, etc
 ```
@@
 ```{r unifrac, include=F}
-ra_wide <- rbcl_rel_abund %>% 
-  select(asv_id, sample_id, rel_abund) %>% 
+
+ra_wide <- rel_abund %>% 
+  select(sample_id, asv_id, rel_abund) %>% 
+  arrange(asv_id) %>% 
   pivot_wider(names_from = asv_id, values_from = rel_abund, values_fill = list(rel_abund = 0)) %>% 
@@
 # root tree with Trebouxiophyceae
-root <- idtax %>% 
+root <- tax %>% 
   filter(class=="Trebouxiophyceae") %>% 
@@
 rbcl.tree <- rbcl.tree %>% root(outgroup = root, resolve.root=T)
+
+# check for differences between asvs in tree and rel abund table (all names must appear in tree)
+asvs_rel_abund <- colnames(ra_wide)
+asvs_tree <- rbcl.tree$tip.label %>% sort()
+
+length(asvs_rel_abund)
+length(asvs_tree)
+
+setdiff(asvs_rel_abund, asvs_tree) # output elts unique to asvs_rel_abund
 
@@
   as_tibble(rownames = "sample_id") %>% 
-  left_join(field, by="sample_id") %>% 
-  left_join(alias, by="sample_id")
+  left_join(field, by="sample_id")
 
@@
 
-```{r setup_rel_abund_barplot, fig.width=10, fig.height=7, include=F}
-plot_data <- rbcl_rel_abund %>% 
-  left_join(rbcl_asv_otu_key, by="asv_id") %>% 
+```{r setup_rel_abund_barplot, include=F}
+plot_data <- rel_abund %>% 
+  left_join(otus, by="asv_id") %>% 
   group_by(sample_id, otu_name) %>% 
@@
   left_join(field, by="sample_id") %>%
-  left_join(alias, by="sample_id") %>% 
   ungroup() %>% 
@@
 
-```{r plot_barplot, fig.width=10, fig.height=7, echo=F}
+```{r plot_barplot, echo=F}
 cust_theme <- theme(legend.position = "none",
@@
 
+otu_names %>% distinct(otu_name)
+
 # order OTUs
-otu_ord <- c("Sanguina A","Chloromonas I","Chloromonas G","Chloromonas muramotoi E","Chloromonas krienitzii D","Chloromonas spp. C","Chlainomonas B","Chlainomonas F","Algae H", "Trebouxiophyceae J")
+otu_ord <- c("Chlainomonas A","Sanguina B", "Chloromonas spp. C","Chloromonas krienitzii D","Chloromonas muramotoi E","Chloromonas F", "Algae G","Chloromonas H", "Trebouxiophyceae I")
 
-plot <- ggplot(plot_data, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
+barplot <- ggplot(plot_data, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
   geom_bar(aes(fill=otu_name), stat="identity") + # classic stacked relative abundance plot
@@
   facet_grid(~otu_name %>% fct_relevel(otu_ord), switch = "x") +
-  labs( x="Sample ID (from lowest to highest elevation)") +
+  labs( x="Sample ID") +
   cust_theme
-plot
-# add sparkline on side for elevation
+barplot
+
+```
+
+```{r combine, fig.width=7, fig.height=10}
+combined_fig_3 <- grid.arrange(uf_plot, barplot, nrow=2)
 ```
@@
   rownames_to_column("sample_id") %>% 
-  left_join(field, by="sample_id") %>% 
-  left_join(alias, by="sample_id")
+  left_join(field, by="sample_id")
 
@@
 ```{r output, include=F}
-# ggsave
+ggsave(filename = here("manuscript/figs/3_community/combined_fig_3.pdf"), plot=combined_fig_3, height=10, width=7)
+ggsave(filename = here("manuscript/figs/supplementary/diversity.png"), plot = faith_date_plot, height = 5, width = 7)
 ```

### 2019-11-02 manuscript/figures.Rmd
init

---
title: "Figures"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=12, fig.height=8, fig.path='figs/', echo=FALSE, warning=FALSE, message=FALSE)
```

```{r input, include=F}
# field <- knitr::include_graphics("temp.png")
asvs <- include_graphics("figs/2_taxonomy/combined.png")
communities <- include_graphics("figs/3_community/combined.png")
# depth
```

```{r pressure, echo=FALSE, fig.cap="A caption", out.width = '100%'}
communities
```

Figure 1.

### 2019-11-02 manuscript/make_word_output.R
init

@@ -1,0 +1,21 @@
+library(here)
+library(bookdown)
+library(rmarkdown)
+
+setwd(here("manuscript/")) # bookdown wants project root to be where the .Rmd files are
+
+# input
+file <- "manuscript.Rmd"
+
+# output
+render(input=file,
+       output_format = "bookdown::word_document2",
+       output_file = paste0("casey1_", format(Sys.Date(), ("%Y-%m-%d")), ".docx"),
+       output_dir = "drafts_outbox",
+       output_options = list(reference_docx = "templates/word-template.docx"))
+
+
+setwd(here()) # return to the project root
+
+
+# can use bookdown::preview_chapter(), https://eddjberry.netlify.com/post/writing-your-thesis-with-bookdown/

### 2019-11-02 manuscript/manuscript.Rmd
init, copied over from Casey.Rmd. Removing figs from this file, put in figures.Rmd

--- 2019-11-01 manuscript/Casey1.Rmd
@@
   - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
-date: "`r format(Sys.time())`"
 output: 
-  word_document:
-    # toc: yes
+  bookdown::word_document2:
+    fig_caption: yes
     pandoc_args:
@@
       - '--lua-filter=templates/author-info-blocks.lua'
-    reference_docx: /home/cengstro/projects/sa_rbcl/Casey1/tmp/word-template.docx
+    reference_docx: templates/word-template.docx
   pdf_document: 
+    fig_caption: yes
     number_sections: yes
-    # toc: yes
-    # toc_depth: 4
-    pandoc_args:
-      - '--lua-filter=templates/scholarly-metadata.lua'
-      - '--lua-filter=templates/author-info-blocks.lua'
-bibliography: /home/cengstro/ownCloud/mendeley_library/bibtex_by_grp/casey1.bib
+csl: templates/nature-communications.csl
+bibliography: casey1.bib
 abstract: |
@@
 Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome.
+---
 
----
-```{r setup, include=FALSE}
+```{r include=FALSE}
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-<br>
-<br>
-<br>
-
 ### *Keywords*: snow algae rbcL 18S amplicon microbiome alpine community
 
-### *Run title*: Elevational patterns in snow algae communities
-
-### *To be submitted to*: 1) Nat Comm, 2) Frontiers in Micb 3) Env Micb 4) Microbial Ecology
+### *Run title*: Snow algae communities along elevational...
 
@@
 
-### Morphologically diverse snow algae blooms across a variety of habitats
+## Morphologically diverse snow algae blooms across a variety of habitats
 
@@
 Over the course of the season, we collected 309 samples from 33 dates and 13 moutains.
-We did not observe pink snow above treeline until June 6, despite several visits above treeline before that (site data table).
+We did not observe pink snow above treeline until June 6, despite several visits above treeline before that (S2).
 We observed snow algae until the first snows of winter, although by later in the season most of the snow had melted out at lower elevations.
@@
 
-### Diverse *rbc*L reveals species-level snow algae taxonomy
+## Diverse *rbc*L reveals species-level snow algae taxonomy
 
-### Snow algae community composition varies with date and elevation
+Our *rbc*L amplicon library was dominated by sequences closely related to *Chloromonas*, *Chlainomonas*, and *Sanguina* (Fig. 2).
 
-This is supported by our 18S and cell count data. Although we found Chlainomonas
+60% and 30% of *rbc*L sequences were assigned to genus and species level, while only % and % of our 18S amplicon library was assigned respectively.
+However, our 18S library generally supported our rbcL findings: samples dominated by *Sanguina* were consistant between markers, however samples containing *Chlainomonas* rbcL were assigned to *Chloromonas* 18S, which could be due to lack of differentiation between the the two genera using 18S (Supp.
+Fig.
+X).
 
-### High elevation, high diversity
+*Chloromonas* was the most diverse
 
-### Green snow algae turns orange at the surface
+Algae OTU8 has a best BLAST score of 88%, with top hits in both Trebouxiophyceae and Chlorophyceae.
+
+Currently there are just two species *Chlainomonas* in reference databases, but our results suggest there may be genetically distinct populations *Chlainomonas* at higher elevations (Fig.
+1, Supp.
+Fig.
+1, Supp.
+Fig 2).
+
+There were fewer *rbc*L ASVs in *Sanguina* than in *Chloromonas* and *Chlainomonas*, and there were no visible sub-clusters within *Sanguina* in our t-SNE or phylogenetic tree (Fig. 2, S).
+
+## Snow algae community composition varies with date and elevation
+
+This is supported by our 18S and cell count data. Although we found Chlainomonas (Fig. 3))
+
+Sites at higher elevation had higher *rbc*L diversity.
+
+## Green snow algae turns orange at the surface
 
@@
 
-<!--
-
-When determining authorship the following criteria should be observed:
-
- - Substantial contributions to the conception or design of the work; or the
-   acquisition, analysis, or interpretation of data for the work; AND
- - Drafting the work or revising it critically for important intellectual
-   content; AND
- - Final approval of the version to be published ; AND
- - Agreement to be accountable for all aspects of the work in ensuring that
-   questions related to the accuracy or integrity of any part of the work are
-   appropriately investigated and resolved.
-
-Contributors who meet fewer than all 4 of the above criteria for authorship
-should not be listed as authors, but they should be acknowledged.
-(http://www.icmje.org/roles_a.html)
-
--->
-
 CE, LQ, and KY designed this study.
@@
 CE wrote the manuscript with major input from LQ and KY.
-<!-- The statement about the authors and contributors can be up to several sentences -->
-<!-- long, describing the tasks of individual authors referred to by their initials -->
-<!-- and should be included at the end of the manuscript before the References -->
-<!-- section. -->
 
@@
 <div id="refs"></div>
-
-# Figures {-}
-
-<br>
-<br>
-
-![Figure 1.
-t-SNE clustering of snow algae rbcL amplicon sequence variants (ASVs).
-Colored points correspond to genus and species level assignment, grey points were not assigned at or below genus level.
-Labelled ellipses show OTUs.
-Perplexity = 30, theta = 0.5](/home/cengstro/projects/sa_rbcl/figs/asv_tsne_fig.png)
-
-<!-- todo: italics, underscores, ChlorKri5, ChlorMura5, put ChloroKri4 on top of Chloro3, so its obvious that these aren't part of both OTUs.-->
-
-##### pagebreak
-
-![Figure 2.
-Sample ordination plots.
-A.
-NMDS comparing sample similarity using OTUs, shaded by elevation.
-Vectors indicate relative contribution of taxa driving differences between samples.
-Stress = 0.183.
-B.
-NMDS showing weighted UniFrac distances between snow algae samples.
-Stress = 0.11.](/home/cengstro/projects/sa_rbcl/figs/sample_ordination.png){width=13.01cm, height=18.56cm}
-
-<!-- add dom morph groupings for UniFrac to connect with heatmap.  adjust size.  -->
-<!-- change axis labs of b, MDS to NMDS-->
-
-##### pagebreak
-
-![Figure 3.
-Relative abundance heatmap of snow algae OTUs in each sample.
-Weight of color corresponds to the relative abundance of that OTU within a sample.
-Samples (rows) are grouped by the dominant cell morphology within that sample as determined by light microscopy.
-Within each morphology group samples are ordered by elevation, from low on the bottom to high on top.
-Sample IDs follow the convention: letters represent the mountain, and the number represents the order in which it was collected.
-Samples collected on the same day are labelled as .1, .2, and so on.](/home/cengstro/projects/sa_rbcl/figs/heatmap.png){width=12.88cm, height=17.17cm}
-
-<!-- add elev range, date range. dom cell morph text to grey. lump chlainomonas categories (let it be a surprise)-->
-<!-- add treeline -->
-```{r, echo = F, include=F}
-# otu_heat2 %>% 
-#   group_by(bloom_type) %>% 
-#   summarise(min(elev_m), max(elev_m))
-# otu_heat2 %>% 
-#   group_by(bloom_type) %>% 
-#   summarise(min(date), max(date))
-```
-
-##### pagebreak
-
-![Figure 4.
-Select microphotographs of snow algae.
-All photos are taken with light microscopy, except (f) which was taken with DIC.
-Scale bars represent 30 $\mu$m.
-(a) H5 (b) S10.2.
-(c) S4.
-(d) K2.
-(e) S10.1.
-(f) G16.1.
-(g) N12.1.
-(h) P20.1.](/home/cengstro/projects/sa_rbcl/figs/fig4.png){width=13.51cm, height=17.49cm}
-<!-- order by elev, add sax1 -->
-
-<!-- ![Figure 5. Relative abundance barplots comparing community composition in surface orange snow and green snow below in three samples.] -->
-<!-- choose photo, microphotos, rethink colors (make consistant w heatmap) -->
-
-##### pagebreak
-
-# Supplementary Data {-}
-
-![Supplementary Figure 1.
-Maximum likelihood tree (IQTree) of snow algae rbcL ASVs.
-Colors correspond to OTU.
-Edges with bootstrap values < 80 are displayed as dotted lines.](/home/cengstro/projects/sa_rbcl/figs/ggtree.png)
-
-##### pagebreak
-
-![Supplementary Figure 2. MDS plot of rbcL ASVs, colored by taxa. Ellipses contain all ASVs within an OTU. Stress =  ](/home/cengstro/projects/sa_rbcl/figs/asv_mds.png)
-
-<!-- redo with NMDS, include stress vals -->
-##### pagebreak
-
-![Supplementary Figure 3.
-Relative abundance barplots comparing community composition in surface orange snow and subsurface green snow.
-* designates depth and surface samples collected on different days.](/home/cengstro/projects/sa_rbcl/figs/draft1/depth.png)
-
-##### pagebreak
-
-<!-- ![Supplementary Figure 4. Photos showing location of *Chloromonas krienitzii* dominated samples.]() -->
-
-<!-- word to template: *[:alpha:] to \*, and *[:space:] to } -->

### 2019-11-02 manuscript/supplemental.Rmd
init

---
title: "Supplemental Figures"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r table}
knitr::kable(mtcars[1:5,, 1:5], caption = "A table caption")
```

### 2019-11-02 code/rbcl_asv_clustering.Rmd
Chloromonas krienitzii not really an OTU, change labelling system

@@
 
-Manually edit the clustering:
-  - *Chloromonas krienitzii* gets its own cluster, since this is highly abundant
+Manually edit the clustering to show *Chloromonas krienitzii* on relative abundance barplot
 
@@
 tsne_edited_otus <- tsne_taxa_otu %>%
-  mutate(otu = case_when((V2 < 14  & V2 > 8 & V1 > (-20) & V1 < (-12)) ~ "9",# krienitzii
+  mutate(otu = case_when((plot_taxa == "Chloromonas_krienitzii") ~ "9",
                          TRUE ~ otu))
@@
                                  "Chlainomonas A",
-                                 "Chloromonas muramotoi E",
-                                 "Chloromonas F",
-                                 "Chloromonas H",
-                                 "Algae G",
-                                 "Trebouxiophyceae I",
-                                 "Chloromonas krienitzii D"))
+                                 "Chloromonas muramotoi D",
+                                 "Chloromonas E",
+                                 "Chloromonas G",
+                                 "Algae F",
+                                 "Trebouxiophyceae H",
+                                 "Chloromonas krienitzii"))
 
@@
              alpha = 0.5) +
-  labs(color="ASV taxonomic assignment") +
+  labs(color="ASV taxonomic assignment", tag="A") +
   ggthemes::theme_few() +
@@
   left_join(total_rel_abund, by="asv_id") %>% 
-  filter(!(asv_id %in% c("639", "534", "646"))) # remove two trebouxiophyceae outliers
+  filter(!(asv_id %in% c("639", "534", "646"))) # remove Trebouxiophyceae outliers
 
@@
   theme_few() +
-  labs(color="ASV taxonomic assignment")
+  labs(color="ASV taxonomic assignment", tag="B")
   # geom_text(aes(label = asv_id)) # too see outliers, asv_id=c("639", "534")
@@
 ```
-
-# TO do
-- add MDS
-- point size proportional to relative abundance
-- ggtree
-- show different values of perplexity yield comparable results

### 2019-11-02 manuscript/manuscript.Rmd
results complete

@@
 
-## Morphologically diverse snow algae blooms across a variety of habitats
+## Field and microscopy observations of distinct snow algae bloom types
 
-Snow algae was widespread in our study area, beginning on May _ when we first observed snow algae blooming in the trees.
-Over the course of the season, we collected 309 samples from 33 dates and 13 moutains.
-We did not observe pink snow above treeline until June 6, despite several visits above treeline before that (S2).
-We observed snow algae until the first snows of winter, although by later in the season most of the snow had melted out at lower elevations.
+We observed pink and green patches of snow algae in forested sites beginning on May 14.
+Snow algae was widespread in our study area, and we collected 309 samples from 33 dates and 13 moutains over the course of the summer (S1).
+We observed an elevational shift in snow algae throughout the season: forested sites bloomed earlier in the season and melted out sooner, while high elevation sites did not bloom until later in the season, but persisted until the first snowfall of winter 2019 (S2).
 
-Visually, several patterns in snow algae site characteristics and cell morphology were apparent.
-Most frequently, we observed pink or red snow, typically in areas recieving high sunlight, dominated by cells that looked similar to *Sanguina nivaloides*, frequently containing green cells of varying shapes and sizes.
-However, red snow could also be dominated by *Chlainomonas* morphologies, in red snow that was visually indistinguishible from red snow containing primarily *S.
-nivaloides*.
-*Chlainomonas* was frequently mixed in with *S.
-nivaloides* as well (Fig.
-1d).
-We also occasionally observed green snow above treeline, which was dominated by green or yellow oval cells with flanges, similar to those described for *Chloromonas muramotoi*.
-Colored snow often appeared in snow runnels, frequently at elevations near or below treeline.
-These runnels containing colored snow formed linear streaks, often with orange on the surface and green snow just below the surface (Fig.
-1b).
-At two sites like this that we visited repeatedly, the runnel began with green snow hidden 1-2 cm below the surface, but on subsequent visits the surface snow had changed color to orange.
-The subsurface snow contained green, often flagellated algae, while surface snow was dominated by orange cells resembling *Chloro.
-krienitzii* (Fig.
-1e) [@Matsuzaki2015].
-Below treeline, orange and green snow was common, often dominated by orange oval cells with short spines, similar to *Chloromonas cf.
-brevispina* (Fig.
-1g).
-Snow dominated by *Chloromonas cf.
-brevispina* and *Chloromonas cf.
-nivalis* was abundnat in shady areas below treeline, often in distinctively bronze colored snow (supp data).
+Often, different colors of snow and habitats contained different cell morphologies.
+In forest sites, orange or green snow was typically dominated by green or yellow-orange ovate cells with short spines (*Chloromonas cf.
+brevispina*; Fig.
+1).
+In snow runnels in forest clearings we observed a shift over repeated site visits; on initial visits snow runnels contained green snow hidden from sight 2 - 5 cm below the surface (Fig.
+1b, S4_), on subsequent visits the surface of these runnels was orange and contained cell morphologies resembling *Chloromonas krienitzii* (Fig 1e, S3_, S4_).
+Red or pink snow was mostly found in open areas with high sunlight, and contained cells similar in appearance to *Sanguina nivaloides* or *Chlainomonas rubra*, in many sites mixed in with green cells of variable morphology (Fig 1).
+*Sanguina nivaloides* morphologies were primarily at high elevation sites, although we did observe several blooms dominated by this morphospecies in shaded forest sites.
 
-## Diverse *rbc*L reveals species-level snow algae taxonomy
+## Diverse *rbc*L reveals snow algae taxonomy with high-resolution
 
-Our *rbc*L amplicon library was dominated by sequences closely related to *Chloromonas*, *Chlainomonas*, and *Sanguina* (Fig. 2).
+Samples were dominated by green algae (Chlorophytes), primarily the genera *Chloromonas*, *Chlainomonas*, and *Sanguina* (Fig.
+2).
+Trebouxiophyceae (Chlorophytes) such as *Raphidonema* were present in low abundance primarily at high elevation sites (Fig.
+2), as well as low abundance of several Ochrophytes such as *Hydrurus* at several low elevation sites (S5).
+One *rbc*L OTU defied taxonomic classification, with a best BLAST score of 88%, and top hits in both Trebouxiophyceae and Chlorophyceae (Algae H, Fig.
+2).
 
-60% and 30% of *rbc*L sequences were assigned to genus and species level, while only % and % of our 18S amplicon library was assigned respectively.
-However, our 18S library generally supported our rbcL findings: samples dominated by *Sanguina* were consistant between markers, however samples containing *Chlainomonas* rbcL were assigned to *Chloromonas* 18S, which could be due to lack of differentiation between the the two genera using 18S (Supp.
-Fig.
-X).
+Our taxa-specific *rbc*L primers showed considerable variation within *Chloromonas* and the closely-related *Chlainomonas* (Fig 2, S6).
+We found several distinct clusters among *Chloromonas* ASVs not matching any known species on GenBank (I, G, Fig.
+2), but the majority of *Chloromonas* ASVs did not form distinct clusters (*Chloromonas spp* C, Fig.
+2).
 
-*Chloromonas* was the most diverse
-
-Algae OTU8 has a best BLAST score of 88%, with top hits in both Trebouxiophyceae and Chlorophyceae.
-
-Currently there are just two species *Chlainomonas* in reference databases, but our results suggest there may be genetically distinct populations *Chlainomonas* at higher elevations (Fig.
-1, Supp.
-Fig.
-1, Supp.
-Fig 2).
-
-There were fewer *rbc*L ASVs in *Sanguina* than in *Chloromonas* and *Chlainomonas*, and there were no visible sub-clusters within *Sanguina* in our t-SNE or phylogenetic tree (Fig. 2, S).
+Our three relative abundance data sets were generally consistant.
+18S, *rbc*L, and cell count data were correlated for *Sanguina* (stats).
+*Chloromonas* and *Chlainomonas* cell counts were correlated with *rbc*L (stats, stats); however, 18S classified samples dominated by *Chlainomonas* as containing *Chloromonas* (S5), likely due to high target region similarity (S7).
+However, in at least one case cell morphology was a poor predictor of taxonomic composition.
+We were surprised to find sample X1.1 contained few *Sanguina* reads (Fig.
+3b, S5b) despite being dominated by red cells similar of similar morphology to *Sanguina* (Fig.
+1g, S4_).
 
@@
 
-This is supported by our 18S and cell count data. Although we found Chlainomonas (Fig. 3))
-
-Sites at higher elevation had higher *rbc*L diversity.
+We found snow algae community composition was strongly influenced by elevation.
+Low elevation sites were compositionally similar, while high elevation sites were more diverse (Fig.
+3a, S8).
+Low elevation sites were dominated by *Chloromonas krienitzii* and *Chlainomonas rubra*, while *Sanguina nivaloides* was nearly exclusive to sites above treeline (Fig.
+3b), along with one OTU of *Chloromonas*, and *Raphidonema* (H, I, Fig.
+3b).
+Several sites above treeline were more similar to low elevation sites, including patches of green snow above treeline dominated by *Chloromonas* (N1.5, G1.4, Fig.
+3) and red snow patches dominated by *Chlainomonas* (G1.2, P1.9, Fig.
+3).
 
@@
 
-**A gradient between snow algae communities**.
-Variation in snow algae communities was primarily driven by the presence or absence of *Sanguina nivaloides*, *Chloromonas krienitzii*, *Chlainomonas rubra*, or *Chlainomonas* OTU3 (Fig.
-2a).
-However, most samples contained a mix of species; few samples were relatively "pure".
-However, sample comparison with weighted UniFrac [@Lozupone2005] found little difference between *Chlainomonas* and *Chloromonas* dominated sites, and the presence or absence of *Sanguina* was the driving factor, which was generally found in higher relative abundance at higher elevation (Fig.
-2b).
+At several sites dominated by *Chloromonas krienitzii* we sampled both the surface orange snow as well as the underlying green snow (Fig.
+4a) to see if the difference in color and cell morphology was due to different species, or different life stages of the same species inhabiting different habitats.
+The surface samples contained primarily orange cells resembling *Chloromonas krienitzii*, while the subsurface samples contained green cells including biflagellates and polar quadriflagellates (Fig.
+4b, c).
+Both surface and subsurface samples were dominated by *Chloromonas krienitzii* reads (Fig.
+4d).
+Unexpectedly, we found high relative abundance of *Chlainomonas* in surface samples, and in one subsurface sample, which we did not detect in cell counts.
 
-Although *Chlainomonas* is reported to only grow in slushy snow above alpine tarns [@Novis2008;@Prochazkova2018c], we found *Chlainomonas* was present in low relative abundance at most sites sampled.
-Only one site high in Chlainomonas was near water (S9), the other Chlainomonas dominated sites were not near water, nor was the snow noticeably mushier than the surrounding snow.
-This does not rule out the possibility that these sites contained high moisture at some point; further sampling is needed to determine its habitat distribution.
-
-#
-
-**Snow algae community composition co-varies with date and elevation**.
-*Sanguina* was more abundant in alpine sites, as well as one OTU of *Chlainomonas* and two OTUs of *Chloromonas*, while *Chloromonas krienitzii* and Chlainomonas OTU6 were found in high relative abundance below treeline (Fig.
-2, 3).
-This consistent with the microscopy observations of @Hoham1979a and @Nedbalova2008a, who found primarily *Chlamydomonas cf.
-nivalis* above treeline and *Chloromonas* below treeline.
-However, in microscopy surveys we infrequently observed *Sanguina* below treeline, and *Chloromonas krienitzii* above treeline.
-We found higher diversity in samples from higher elevation, both at the ASV level and OTU level <!--Chao1 w fossil.-->, with several OTUs that were not found at lower elevations.
-*Raphidonema* and related ASVs were restricted to high alpine sites, with the highest relative abundance occurring in samples from August and September.
-This fits with the interpretation of @Stibal2005, who suggest that *Raphidonema* is an opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow.
-However, rare taxa should not be discounted as playing an important role in ecosystem functioning [@Jousset2017].
-
-Snowpack above and below treeline are very different habitats, with forest providing shade (and therefore less evaporation).
-The red pigment of *Sanguina* and *Chlainomonas* (astaxanthin) is likely an adaptation to high light [@Gorton2003;@Holzinger2006], and it has been suggested that *Sanguina* is adapted to the drier snowpacks that form at higher elevations [@Dial2018c].
-
-#
-
-**_Chloromonas krienitzii_ is green below, turns orange at surface**.
-On initial visits to Chloromonas krienitzii dominated sites, green snow was hidden 2 to 5 cm below the surface of the runnel (Supp.
-Fig.
-4).
-Green snow was noticeably moister than surrounding snow, in some locations containing ice lenses not found in the surrounding snowpack.
-By June 6 the surface of the runnel had turned orange while the snow below remained green, which persisted on three subsequent visits until the snow melted.
-Cell morphology at the surface closely resembled *Chloromonas cf.
-brevispina* “Trochiscia” [@Matsuzaki2015] (Fig.
-4), while green subsurface snow was dominated by green biflagellate or bipolar quadriflagellate cells (Fig.
-4).
-Both green and orange samples were dominated by reads assigned to *C.
-krienitzii*, with lower relative abundance of *Chlainomonas*, which was more abundant at the surface (Supp.
-Fig.
-3).
-
-One possibility is that C.
-krienitzii cells orange pigmentation is triggered or an adaptation to higher light levels at the surface.
-Green cells below the surface would be protected from UV radiation<!--cite -->.
-Possibly, the green cells maintain an optimal light level through phototaxis, similar to marine phytoplankton<!--cite-->.
-Perhaps higher elevations typically have fewer low angle runnels that would allow a moist enough environment for *C.
-krienitzii*.
-It seems unlikely that nutrients from conifer needles would explain this, as several of our *C.
-krienitzii* dominated sites were in full sunlight, far from any trees.
-
-#
-
-**Snow algae blooms harbor novel diversity**.
-Several OTUs were not assigned to species level due to low similarity (Fig.
-1).
-Two OTUs of Chloromonas, one OTU of Chlainomonas, and one OTU ("Algae OTU8") that had best BLAST matches to both Trebouxiophyceae and Chlorophyceae algae.
-Algae OTU8 has a best BLAST score of only 88%; this could represent a free-living algae found in snow, or possibly an endosymbiotic algae from a ciliate, as we noted many ciliates with green organelles in our microscopy samples.
-Sample X17 contained was dominated by spherical red cells superficially similar to *Sanguina*, this sample was dominated by reads of Chloromonas OTU7 and Algae OTU8.
-Cells in this sample had thicker cell wall, and some cells were slightly square-shaped (Fig.
-4).
-18S algae reads from this sample were equivocal, with best BLAST matches to both *Raphidonema* and *Chloromonas*.
-Currently there are just two species *Chlainomonas* in reference databases, but our results suggest there may be genetically distinct populations *Chlainomonas* at higher elevations (Fig.
-1, Supp.
-Fig.
-1, Supp.
-Fig 2).
-Although some of the high elevations ASVs were classified as *Chlainomonas rubra*, because we only included this species in our custom snow algae reference this could have biased our classifier, possibly causing over-classification [@Murali2018].
-
-Although our rbcL primers successfully detected distinct OTUs, this is likely an underestimate of the true diversity.
-We did not detect any clustering within *Sanguina* using either 18S or rbcL, however when we included the homologous rbcL fragment of *S.
-aurentia* in our analysis it did not form a distinct OTU <!--this is where I am quesy about out ghetto OTU method-->.
-Furthermore, rbcL is highly differentiated in *Chloromonas* relative to other microalgae [@Nozaki2002], which could partly explain why *Sanguina* is less differentiated than *Chloromonas* or *Chlainomonas* (Supp.
-Fig.
-1 and 2).
-
-# Discussion {-}
-
-We found snow algae communities were typically dominated by either *Sanguina*, *Chlainomonas* or *Chloromonas*, with secondary abundance of *Chlainomonas* or *Chloromonas* in most samples.
-Most samples contained a sizable proportion of secondary or tertiary OTUs, although some some contained nearly 100% *Chlainomonas rubra*, *Sanguina nivaloides*, or *Chloromonas krienitzii*.
-The drivers of these differences between communities remain unclear; it is likely that within the melting snowpack there may be distinct microhabitats that impacts snow algae distribution, with varying water content with topography, different light levels in shade or at depth [@Gorton2007], different daily freeze-thaw cycles, or possibly varying nutrient levels with proximity to vegetation [@Hoham1976] or bedrock [@Kol1968a].
-Further work needs to be done to uncover the micro-habitat niches within the snowpack that allow this differentiation, and what novel species the snow microbiome might contain.
+# Discussion
 
@@
 brevispina*, *Chloromonas krienitzii* *Chlainomonas*, or "other".
+
+Cite refs for why we are calling them that.
 
@@
 We split the largest cluster (containing the vast majority of the *Chloromonas* reads) into three clusters, one containing all ASVs assigned to *Chloromonas krienitzii*, and one containing all ASVs assigned to *Chloromonas muramotoi* and *Chloromonas hohamii*.
-To highlight our observation that one clade of *Chlainomonas* were only found above treeline (Suppl.
-Fig.
-1), we split *Chlainomonas* into two OTUs.
-The OTUs produced by this naive method were roughly consistent across different methods of visualization (Suppl.
+The OTUs produced by this naive method were consistent across different methods of visualization (Suppl.
 Fig.
@@
 -->
+
+UniFrac [@Lozupone2005]
 

### 2019-11-02 code/cellcount.Rmd
minor changes

@@
                   "Sanguina nivaloides",
-                  "Sanguina nivaloides 'turret'",
+                  "Sanguina nivaloides 'ruby'",
                   "Green cell",
@@
   geom_tile(aes(alpha = percent_of_sample, fill = morpho_sp)) + # color = "white" %>% fct_reorder(morpho_order)
-  labs(title="Cell count relative abundance", x="Cell morphology", y= "Sample ID", fill = "Cell morphology", alpha="Relative abundance") +
+  labs(x="Cell morphology", y= "Sample ID", fill = "Cell morphology", alpha="Relative abundance") +
   guides(fill=F) +

### 2019-11-02 code/depth_comparison.Rmd
ordered otus

@@
 ```{r}
-ggplot(rbcl_1, aes(x=top_sample, y=rel_abund, fill = otu_name))+
+
+otu_ord <- c("Chlainomonas A","Sanguina B", "Other Chloromonas","Chloromonas krienitzii","Chloromonas muramotoi C","Chloromonas D", "Algae E","Chloromonas F", "Trebouxiophyceae G")
+
+depth <- ggplot(rbcl_1, aes(x=top_sample, y=rel_abund, fill = otu_name %>% fct_relevel(otu_ord)))+
   geom_bar(stat="identity") +
   facet_grid(rows = vars(depth_chr %>% fct_relevel("surface","sub-surface"))) +
-  labs(x="Sample ID", y="Relative abundance", fill="OTU")
+  labs(x="Sample ID", y="Relative abundance", fill="OTU", tag="D")
+depth
 ```
@@
 More fungi on surface, one sample with more rotifers on surface, more ciliates on surface also
+
+```{r output, include=FALSE}
+ggsave(filename = here("manuscript/figs/4_depth_comparison/rbcl.pdf"), plot=depth)
+```

### 2019-11-02 code/rbcl_asv_clustering.Rmd
renamed OTUs

@@
 otu_names <- tibble(otu = as.character(1:9), 
-                    otu_name = c("Chloromonas spp. C",
+                    otu_name = c("Other Chloromonas",
                                  "Sanguina B",
                                  "Chlainomonas A",
-                                 "Chloromonas muramotoi D",
-                                 "Chloromonas E",
-                                 "Chloromonas G",
-                                 "Algae F",
-                                 "Trebouxiophyceae H",
+                                 "Chloromonas muramotoi C",
+                                 "Chloromonas D",
+                                 "Chloromonas F",
+                                 "Algae E",
+                                 "Trebouxiophyceae G",
                                  "Chloromonas krienitzii"))

### 2019-11-02 code/rbcl_community_compare.Rmd
minor changes, make multi panel plot

@@
   scale_color_gradientn(colours = terrain.colors(8)[1:7]) +
-  labs(color="Elevation (m)") +
+  labs(color="Elevation (m)", tag="A") +
   geom_text_repel(aes(label=alias))
@@
 
-# check rel abund sums = 1
+# check rel abund sums ~= 1 (will be slightly higher, due to rounding)
 plot_data %>%
@@
 
-# order OTUs
-otu_ord <- c("Chlainomonas A","Sanguina B", "Chloromonas spp. C","Chloromonas krienitzii D","Chloromonas muramotoi E","Chloromonas F", "Algae G","Chloromonas H", "Trebouxiophyceae I")
+otu_ord <- c("Chlainomonas A","Sanguina B", "Other Chloromonas","Chloromonas krienitzii","Chloromonas muramotoi C","Chloromonas D", "Algae E","Chloromonas F", "Trebouxiophyceae G")
 
 barplot <- ggplot(plot_data, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
-  geom_bar(aes(fill=otu_name), stat="identity") + # classic stacked relative abundance plot
+  geom_bar(aes(fill=otu_name %>% fct_relevel(otu_ord)), stat="identity") + # classic stacked relative abundance plot
   coord_flip() +
   facet_grid(~otu_name %>% fct_relevel(otu_ord), switch = "x") +
-  labs( x="Sample ID") +
+  labs( x="Sample ID", tag="B") +
   cust_theme
@@
 ```{r combine, fig.width=7, fig.height=10}
-combined_fig_3 <- grid.arrange(uf_plot, barplot, nrow=2)
+combined_fig_3 <- gridExtra::grid.arrange(uf_plot, barplot, nrow=2)
 ```
@@
 faith_date_plot
+```
+
+```{r}
+faith_elev_plot <- ggplot(faiths, aes(y=PD, x=elev_m))+
+  geom_point() +
+  geom_smooth(method = "lm")+
+  geom_text_repel(aes(label=alias))+
+  labs(title="Phylogenetic diversity of snow algae rbcL vs elevation",y="Faith's phylogenetic diversity (PD)",x="Elevation (m)")
+faith_elev_plot
 ```

### 2019-11-02 code/whole_18s_community_compare.Rmd
new paths for input files

--- 2019-10-30 code/whole_18s_community_compare.Rmd
@@
 ```{r input, include=F}
-tree_path <- here("intermediate_output/tree/full_18s/all_18s_aligns.fasta.treefile")
-all_18s_tree <- read.tree(tree_path)
-
-ra_path <- here("data/clean/clean_18s_ra.csv")
-ra <- read_csv(ra_path)
-
-assignment_path <- here("data/clean/clean_silva_18s_assignments.csv")
-assign <- read_csv(assignment_path)
-
-field_path <- here("data/clean/field_clean.csv")
-field <- read_csv(field_path)
-
-alias_path <- here("data/sample_aliases.csv")
-alias <- read_csv(alias_path)
+all_18s_tree <- read.tree(here("output/tree/full_18s/all_18s_aligns.fasta.treefile"))
+ra <- read_csv(here("data/03_filtered/18s/filtered_cleaned_18s_rel_abund.csv"))
+assign <- read_csv(here("data/02_tidied/tidied_silva_18s_assignments.csv"))
+field <- read_csv(here("data/02_tidied/tidied_field.csv"))
 ```
@@
 ```{r nmds_stress, include=F}
+set.seed(123)
 mono <- metaMDS(weighted_unifrac_distm)
@@
 # order algae, fungi, protists, other
-phylum_ord <- c("Chlorophyta_ph", "Ochrophyta","Basidiomycota", "Ascomycota", "Chytridiomycota", "Ciliophora","Cercozoa","Protalveolata", "Phragmoplastophyta")
+phylum_ord <- c("Chlorophyta", "Ochrophyta","Basidiomycota", "Ascomycota", "Chytridiomycota", "Ciliophora","Cercozoa","Protalveolata", "Phragmoplastophyta")
 
-ggplot(p, aes(x=sample_id %>% fct_reorder(elev_m), y=rel_abund)) +
+rel_abund_phyla <- ggplot(p, aes(x=sample_id %>% fct_reorder(elev_m), y=rel_abund)) +
   geom_bar(aes(fill=phylum), stat="identity") + # classic stacked relative abundance plot
@@
   facet_grid(~phylum %>% fct_relevel(phylum_ord), switch = "x") +
-  labs(title = "Relative abundance plot of 18S phyla", x="Sample ID (from lowest to highest elevation)") +
+  labs(title = "Relative abundance plot of 18S phyla (SILVA assignments)", x="Sample ID (from lowest to highest elevation)") +
   cust_theme
+rel_abund_phyla
 ```
@@
   -
+
+```{r output, include=F}
+dir.create(here("manuscript/figs/S5_18s"))
+ggsave(filename = here("manuscript/figs/S5_18s/18s_rel_abund.pdf"), plot=rel_abund_phyla)
+```

### 2019-11-02 code/wrangle/tidy_field.R
keep column whether sample was sequenced or not

@@ -67,9 +67,9 @@ field_sample_uid <- field_visit_num %>%
   mutate(alias = paste0(alias_1, ".", uid))
 field_sample_uid %>% distinct(sample_id, alias, date) %>% arrange(date) %>% view()
 
-# remove some of the unnessecary columns
+# remove some of the unused columns
 field_removed_extra_cols <- field_sample_uid %>% 
-  select(sample_id, alias, date, lat, lon, elev_m, color_snow, habitat, substrate, depth_cm, dom_morph)
+  select(sample_id, alias, date, lat, lon, elev_m, color_snow, habitat, substrate, depth_cm, dom_morph, have_seq)
 
 
 

### 2019-11-02 manuscript/supplemental.Rmd
describe each sup fig

@@
 
+Supplemental figures:
+ S1) Map of sample locations
+ S2) Table of sample site information
+ S3) Photos of field sites
+ S4) Microphotos of snow algae samples
+ S5) Plots of snow algae 18S community composition. A. All 18S, Silva assignments. B. Chlorophyta assignments with custom reference database. 
+ S6) Maximum likelihood tree of *rbc*L amplicon
+ S7) Plot of base differences between Genbank 18S and *rbc*L segments
+ S8) Phylogenetic diversity of snow algae.
+A.
+Plot of Faith's phylogenetic diversity in snow algae communities by date and elevation B.
+Linear model results comparing date and elevation as explanatory variables.
+
 ```{r table}

### 2019-11-03 manuscript/manuscript.Rmd
new abstract, title brainstorm, bit more intro

@@
 ---
-title: "Distinct snow algae communities along an elevational gradient"
+title: "Snow algae community composition along an elevational gradient"
+# distinct yet overlapping categories of snow algae community 
+# Classification of snow algae communities along an elevational gradient
+# Elevation influences snow algae community composition
+# 
 author:
@@
   *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate.
-Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch on the same mountain, and varies with elevation and date.
-We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada.
-We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m).
-Samples from higher elevation had higher phylogenetic diversity, and included several potentially novel clades of algae.
-Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface.
-Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome.
+The factors that determine when and where a snow algae bloom will form are poorly understood; fundamentally, we need to know what algae species are present in this poorly understood microbiome.
+We used a novel *rbc*L and 18S amplicon high-throughput sequencing (HTS) approach to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada.
+We found distinct communities above and below treeline, with *Chloromonas* prevalent in forested sites and *Sanguina* above treeline, while *Chlainomonas* was abundant across all elevations.
+Alpha diversity was variable; surprisingly, samples from highest elevations contained the highest diversity.
+We observed the snow algae *Chloromonas krienitzii* transition from green to orange over the course of the summer, and our observations suggest that different life stages of this species occupy distinct habitats above and below the snow surface.
+The snow algae microbiome is more diverse than previously thought, and future work to understand the biological mechanisms of snow algae blooms must take into account variation in community composition.
 ---
@@
 
-where we were: there are red and green blooms
-where we are: there are multiple types of red and green blooms
-
 Snow algae are cold-adapted microalgae that bloom in melting summer snow throughout alpine and polar regions [@Leya2013].
@@
 Thus, snow algae could play an important role in snow melt regulation, but little is known about how these blooms form.
-We don't know the spatial extent of snow algae blooms worldwide, what conditions trigger snow algae bloom formation, or how cells arrive on the surface of the snow to "seed" the bloom.
-Critically, we lack basic understanding of the species composition and diversity in snow algae blooms, which is fundamental to answering these long-standing mysteries about snow algae blooms.
+We don't know how cells colonize the white snow to initiate the bloom, what triggers snow algae to divide and produce pigment, or why one patch of snow turns red while the neighboring snow remains white.
+Critically, we lack basic understanding of the species composition and diversity of snow algae blooms, which is fundamental to answering long-standing mysteries about snow algae bloom formation.
 
-Historically snow algae were classified by cell morphology, however, morphology can be deceptive when identifying microalgae.
-Until recently, all red cells from red snow were referred to as *Chlamydomonas cf.
-nivalis* [@Remias2013;@Segawa2018], desipte the fact that this morphology-based species was thought to contain multiple distinct species [@Kol1961].
-Recent genetic work designated found that some of these cells were not closely related to *Chlamydomonas* at all, and were designated as a new genus and species: *Sanguina nivaloides* [@Prochazkova2018c].
-Other genetic work shows that morphology-based species of snow algae may in fact contain multiple genetically unrelated clades: for example, *Chloromonas cf.
-nivalis* contains at least four distinct clades [@Matsuzaki2019].
+<!-- red green problem -->
+Despite over a century of study, no one has sucessfully cultured the snow algae that produce red secondary pigment, nor directly observed these genera dividing in the field.
+Consequently, we know little of the life cycle of red-pigmented snow algae.
+In related green algae, cells produce secondary pigments in just one life stage, remaining green for much of their life cycle [@Shah2009;@Salome2019; thus red snow algae could originate from green cells.
+Green snow is frequently found in waterlogged snow, sometimes alongside red snow, leading some to hypothesize that green snow could be an earlier life stage of red snow.
+Two studies comparing green and red snow using amplicon high-throughput sequencing (HTS) found they were caused by different species [@Lutz2016;@Terashoma2016]; however it remains an open question whether green snow can develop into red snow.
 
-<!-- hammer this home: WE LACK BASIC UNDERSTANDING ABOUT SPECIES COMPOSITOIN AND DIVERSITY -->
+<!-- chlainomonas -->
+Currently just two genera of snow algae are known to produce red pigment: *Sanguina* and *Chlainomonas*.
+*Sanguina* is cosmopolitan and widespread in summer snow in alpine and polar regions [@Prochazkova2019a], however the distinctive large red cells of *Chlainomonas* (Cr, Fig.
+1d) are only known from a few alpine sites in Europe, North America, and New Zealand [@Novis2008;@Prochazkova2018c].
+On the basis of microscopy and Sanger sequencing of field samples, these studies suggested that *Chlainomonas* blooms are limited to highly saturated, waterlogged snow overlying mountain lakes.
+*Chlainomonas* has not been detected in amplicon HTS surveys of snow algae blooms [@Terashima2017;@Lutz2017], and little is known of it's distribution, and relative abundance in the red snow micobiome.
 
-<!-- sequence cultures, and batch, and single cell, but this dosent tell us whats actually in the field -->
-More recently, snow algae taxonomy has
+<!-- habitat pattens -->
+Few studies have addressed the habitat differences that give rise to different communities of snow algae to develop.
 
-<!-- high throughput: lacking ref data, no otu based, none in alpine environments (despite this tantalizing evidence of microscopy based, etc) -->
+Morphology-based cell counts suggest that green snow caused by *Chloromonas* is morprevalent in forested areas, and red snow containing cells similar in appearance to *Sanguina* prevalent in alpine areas [@Hoham1979a;@Nedbalova2008a].
 
-<!-- But no one until now has used high throughput sequencing to compare snow algae blooms in alpine environments, or used higher resolution marker genes to do so -->
+<!-- composition -->
+The snow algae literature suggests that snow algae blooms are typically dominated by a single species. Microscopy finds that
 
-<!-- so we did this... -->
+Green snow can be caused by a variety of snow algae, including a diversity of *Chloromonas* species [@Remias2013a;@Terashima2017], *Microglena* [@Lutz2015], as well as Ochrophytes [@Remias2013;@Remias2019].
+Some *Chloromonas* produce pigments that turn snow orange, but this genus lacks the bright red astaxanthin characteristic of *Chlainomonas* and *Sanguina*.
+
+<!-- WHERE THEY ARE FOUND< HABITAT -->
+<!-- Where are snow algae species found? Do different habitats host different snow algae species? -->
+<!--   - why we care -->
+<!--   - what is unknown -->
+
+Chloromonas mostly known in culture, most diversity
+Chlainomonas few reports, found in NA, only in mountains not polar (but they arent looking)
+Sanguina cosmopolitan
+
+Cosmopolitan
+
+co-occurance:
+red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a].
+
+To understand the conditions that permit snow algae to grow, we need to define what species we are talking about; different species could have different habitat requirements.
+Before we develop testable hypotheses e.g.
+this species requires more of this nutrient, we need to see what is out there, and what species are found together, and under what conditions.
+
+Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013].
+In this micro-habitat, temperature is nearly constant, but snow moisture, pH, nutrients, and light can differ greatly within the snowpack depending on location of snow and time of year [@Loranger1988;@Kuhn2001;@Gorton2007;@Ganey2017].
+
+Although HTS found similar red snow, they lacked resolution, and used closed reference. Could be mislabelling OTUs. 
+They did not look at green snow, only alpine and polar red snow.
+
+Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a]
+
+, [@Lutz2015;@Remias2013], while  *Chlamydomonas cf.
+nivalis* is hypothesized to be specially adapted to well-drained, drier snow [@Ganey2018].
+Green snow is thought to likewise require high moisture content.
+
+<!-- WHEN/WHY THEY TRANSITION FROM GREEN TO RED<  -->
+<!-- Under what condions do snow algae develop pigment?  -->
+
+to crack the mystery of how and when snow algae divide, and produce pigment, we need to first must define what species we are talking about. 
+we need to know what species are in a bloom before we can figure out the red green problem. 
+No one has observed green snow transitioning to red in the field
+No one has cultured the algae that produce red pigment *Sanguina* and *Chlainomonas*
+  - despite over a century of study [@Kol1968] no one has sucessfully cultured this genus.
+  - microscopy pigment intermediated suggests they develop red pigment
+  - life cycle, related species are green algae that produce pigment in the cyst
+    - C rein
+    - H pluv
+Red and green snow could be the same species, or different species. 
+The few studies to compare red and green snow found that they were caused by different species: Lutz found green snow caused by Microglena, Terashima green snow caused by Chloromonas. 
+Are there blooms that transition from green to red? under what consitions do they develop pigment?
+
+<!-- how my methods will answer 1) whats out there 2) where is it 3) do snow algae transition from green to red -->
+<!-- brief overview of findings -->
+
 In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS (explain what these code for).
@@
 We found ...
-
-<!-- the red green problem -->
-The same species can look completely different depending on the life stage [@Shah2016;@Salome2019]; conversely, different species can look nearly identical.
-For example, cell morphologies known as *Chloromonas cf.
-nivalis* appear in at least four different lineages.
-One of these lineages, *Chloromonas muramotoi* appear as green biflagellates in culture, but in the field the same species features thick cell walls with flanges [@Matsuzaki2019].
-However, life cycle changes in the the cosmopolitan and widespread genus *Sanguina* have yet to be documented: despite over a century of study [@Kol1961] no one has sucessfully cultured this genus.
-
-<!-- habitat differences -->
-Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013].
-In this micro-habitat, temperature is nearly constant, but snow moisture, pH, nutrients, and light can differ greatly within the snowpack depending on location of snow and time of year [@Loranger1988;@Kuhn2001;@Gorton2007;@Ganey2017].
-Differences in snow algae community composition could offer insight into the environmental factors that determine snow algae distribution.
-Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a].
-Two studies using microscopy found that red snow containing *Chlamydomonas cf.
-nivalis* was prevalent in the above timberline, while green and orange snows caused by *Chloromonas cf.
-brevispina* and *Chloromonas cf.
-nivalis* were prevalent in conifer forests below timberline [@Hoham1979a;@Nedbalova2008a].
-Differences in snow moisture could influence which species of snow algae grow; several species of snow algae are anecdotally reported to only inhabit very saturated, waterlogged snow [@Lutz2015;@Remias2013;@Novis2008;@Prochazkova2018c], while  *Chlamydomonas cf.
-nivalis* is hypothesized to be specially adapted to well-drained, drier snow [@Ganey2018].
-
-<!-- habitat differences, short new form -->
-  Red snow is also occasionally reported to be caused by another red-pigmented algae: *Chlainomonas* [@Novis2008].
-Little is known about the distribution of this genus; one study suggested that *Chlainomonas* is limited to high-moisture snowpack above alpine lakes [@Prochazkova2018].
-Green snow may contain a variety of algae, notably *Chloromonas* (refs), as well as algae from other families including Trebouxiophyceae and Chrysophyceae (refs).
-Morphology-based cell counts suggest that green snow caused by *Chloromonas* is prevalent in forested areas and red snow containing *Chlamydomonas cf.
-nivalis* is prevalent in alpine areas [@Hoham1979a;@Nedbalova2008a]; however, morphology can be misleading when identifying microalgae.
-
-Although much of the snow algae literature focusses on eye-catching red snow, different species of snow algae can also color snow green [@Lutz2015].
-
-<!-- hts -->
-Microscopic observation shows that snow algae blooms can contain a diversity of cell morphologies in the same sample [@Lutz2018], but the genetic community structure of snow algae may not correlate with what is seen under the microscope.
-18S amplicon high-throughput sequencing (HTS) shows that snow algae may contain more diversity than meets the eye: although red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a].
 18S is highly conserved and therefore suited to survey broadly across a variety of taxa; however, due to its high similarity between closely related species short segments of thie gene are poorly-differentiated among closely related taxa (ref).
@@
 
-Until recently, snow algae sequences were lacking in reference databases,
+Are there distinct communities of snow algae?
+  - does green snow turn red?
+  -
 
@@
 We observed pink and green patches of snow algae in forested sites beginning on May 14.
-Snow algae was widespread in our study area, and we collected 309 samples from 33 dates and 13 moutains over the course of the summer (S1).
+Snow algae was widespread in our study area, and we collected 309 samples from 33 dates and 13 moutains over the course of the summer (S1, S2).
 We observed an elevational shift in snow algae throughout the season: forested sites bloomed earlier in the season and melted out sooner, while high elevation sites did not bloom until later in the season, but persisted until the first snowfall of winter 2019 (S2).
 
-Often, different colors of snow and habitats contained different cell morphologies.
+We observed some consistant patterns between cell morphologies and field site characteristics.
 In forest sites, orange or green snow was typically dominated by green or yellow-orange ovate cells with short spines (*Chloromonas cf.
@@
 2).
-Trebouxiophyceae (Chlorophytes) such as *Raphidonema* were present in low abundance primarily at high elevation sites (Fig.
-2), as well as low abundance of several Ochrophytes such as *Hydrurus* at several low elevation sites (S5).
+Trebouxiophyceae (Chlorophytes) such as *Raphidonema* were present in low abundance (Fig.
+2), as well as low abundance of Ochrophytes (primarily *Hydrurus*) at several low elevation sites (S5).
 One *rbc*L OTU defied taxonomic classification, with a best BLAST score of 88%, and top hits in both Trebouxiophyceae and Chlorophyceae (Algae H, Fig.
@@
 *Chloromonas* and *Chlainomonas* cell counts were correlated with *rbc*L (stats, stats); however, 18S classified samples dominated by *Chlainomonas* as containing *Chloromonas* (S5), likely due to high target region similarity (S7).
-However, in at least one case cell morphology was a poor predictor of taxonomic composition.
-We were surprised to find sample X1.1 contained few *Sanguina* reads (Fig.
+However, in some samples cell morphology was a poor predictor of taxonomic composition.
+For example, we were surprised to find sample X1.1 contained few *Sanguina* reads (Fig.
 3b, S5b) despite being dominated by red cells similar of similar morphology to *Sanguina* (Fig.
@@
 We found snow algae community composition was strongly influenced by elevation.
-Low elevation sites were compositionally similar, while high elevation sites were more diverse (Fig.
+Low elevation sites were compositionally similar, while high elevation samples showed inter-site variability (UniFrac Fig.
 3a, S8).
@@
 
-## Green snow algae turns orange at the surface
+We found *Chlainomonas* was widespread and abundant in both alpine and subalpine snowpack.
+
+Co-occurance
+
+## A green snow algae turns orange at the surface
 
@@
 # Discussion
+
+We found snow algae communities were distinct, dominated by one of
+
+- different communities
+  - fields dominated by Chlainomonas, Sanguina, Chloromonas in microscopy studies
+    - differences between above and below treeline community in microscopy studies
+    - side-by-side comparison lacking, only amplicon studies targeting polar red snow, low taxonomic resolution
+      - unable to recognize Chlainomonas with 18S primers, whether this genera is widespread in melting snow worldwide
+  - Chloromonas commonly occurs with Sanguina
+    - we report several likely novel taxa of Chloromonas that are widespread throughout snow algae blooms
+
+- We noted higher diversity in
+
+- what are the drivers of the community differences
+  - cell count, larger survey sample size supports our findings of different community types
+  - 
+- what our observations suggest for Chloro krienitzii life cycle
+  - one explaination, light triggers pigment production
+  -
+
+- chlainomonas widespread in low relative abundance
+  - not just in wet lake areas
+  - hitherto unrecognized major component of the snow algae community
+
+- bias
+  - extraction
+  - primer bias
+    - Because our rbcL primers were designed off of *Chloromonas* and *Chlainomonas*, it is possible they would have missed anything not closely related.
+However, they were able to detect both *Sanguina*, and even algae in a different class, Trebouxiophyceae, so although primer bias is always possible it at least should be able to detect Chlorophyceae and Trebouxiophyceae.
+  - false 0s?
+  - cross contam? (below detection threshold of qubit)
+    - sequencing of negative controls, there is stuff in there
+    -
+
+Although *Chlainomonas* is reported to only grow in slushy snow above alpine tarns [@Novis2008;@Prochazkova2018c], we found *Chlainomonas* was present in low relative abundance at most sites sampled.
+Only one site high in Chlainomonas was near water (S9), the other Chlainomonas dominated sites were not near water, nor was the snow noticeably mushier than the surrounding snow.
+This does not rule out the possibility that these sites contained high moisture at some point; further sampling is needed to determine its habitat distribution.
+
+Snowpack above and below treeline are very different habitats, with forest providing shade (and therefore less evaporation).
+The red pigment of Sanguina and Chlainomonas (astaxanthin) is likely an adaptation to high light (Gorton and Vogelmann 2003; Holzinger and Lütz 2006), and it has been suggested that Sanguina is adapted to the drier snowpacks that form at higher elevations (Dial, Ganey, and Skiles 2018).
+
+Raphidonema and related ASVs were restricted to high alpine sites, with the highest relative abundance occurring in samples from August and September.
+This fits with the interpretation of Stibal and Elster (2005), who suggest that Raphidonema is an opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow.
+However, rare taxa should not be discounted as playing an important role in ecosystem functioning (Jousset et al.
+2017).
+
+Hoham, Roemer, and Mullet (1979) and Nedbalová and Lukavský (2008), who found primarily Chlamydomonas cf.
+nivalis above treeline and Chloromonas below treeline.
+However, in microscopy surveys we infrequently observed Sanguina below treeline, and Chloromonas krienitzii above treeline.
 
@@
 We washed columns twice with 70% ethanol, and finally dissolved our DNA by spinning with 50 $\mu$L of sterile TE buffer.
-As a negative control we processed a sterile distilled water sample alongside each batch, treating it exactly the same as the other tubes.
+As a negative control we processed a sterile distilled water sample alongside each batch, treating it exactly the same as the other tubes; all negative controls did not contain Qubit-detectable levels of DNA.
 

### 2019-11-04 manuscript/manuscript.Rmd
rough draft of intro complete

@@
 We observed the snow algae *Chloromonas krienitzii* transition from green to orange over the course of the summer, and our observations suggest that different life stages of this species occupy distinct habitats above and below the snow surface.
-The snow algae microbiome is more diverse than previously thought, and future work to understand the biological mechanisms of snow algae blooms must take into account variation in community composition.
+The snow algae microbiome is more diverse than previously thought, and future work must take into account variable community composition in snow algae blooms.
 ---
@@
 We don't know how cells colonize the white snow to initiate the bloom, what triggers snow algae to divide and produce pigment, or why one patch of snow turns red while the neighboring snow remains white.
-Critically, we lack basic understanding of the species composition and diversity of snow algae blooms, which is fundamental to answering long-standing mysteries about snow algae bloom formation.
+Critically, we lack basic understanding of the species composition and diversity present in snow algae blooms, which is fundamental to resolving long-standing mysteries about snow algae bloom formation.
 
-<!-- red green problem -->
-Despite over a century of study, no one has sucessfully cultured the snow algae that produce red secondary pigment, nor directly observed these genera dividing in the field.
-Consequently, we know little of the life cycle of red-pigmented snow algae.
-In related green algae, cells produce secondary pigments in just one life stage, remaining green for much of their life cycle [@Shah2009;@Salome2019; thus red snow algae could originate from green cells.
+Snow algae blooms can color snow red, orange, or green, and can one or contain multiple species of algae.
+Most snow algae blooms are caused by green algae of class Chlorophyceae [@Leya2013], although blooms caused by Trebouxiophyceae [@Lutz2015] or Chrysophycae [@Remias2013;@Remias2019] have been reported.
+Among the Chlorophyceae, both *Sanguina nivaloides* [@Prochazkova2019a] or *Chlainomonas* [@Prochazkova2018c] produce the red pigment astaxanthin respoponsible for the red color of some snow algae blooms.
+Orange snow is reported to be caused by *Sanguina aurentia* [@Prochazkova2019a] and some species of *Chloromonas* [@Prochazkova2019c;@Remias2013a], while algae lacking secondary pigments color the snow green, such as *Chloromonas pichinchae* [@Hoham1975].
+However, some species are known from only one or two samples, and the recently described *Chlainomonas* and *Sanguina* are poorly explored, with only two species in each.
+
+Snow algae community composition is largely unknown.
+Some studies batch sequencing samples, and microscopy suggests that blooms are dominated by a single species.
+However, 18S amplicon surveys of red snow from across the Arctic [@Lutz2016a] and Japan [@Terashima2017] show that that snow algae blooms typically contain several secondary species; in addition to the dominant red snow algae, these studies found lower relative abundance of *Chloromonas*, as well as *Raphidonema* in Arctic sites.
+33 red snow samples from across the Arctic were relatively homogeneous in community composition [@Lutz2016a], but genetic comparison of the morphologically diverse snow algae communities found in mountains from temperate latitudes [@Nedbalova2008a] is lacking.
+
+Despite over a century of study, no one has sucessfully cultured the snow algae that produce red secondary pigments.
+In many green algae cells produce secondary pigments in just one life stage, remaining green for much of their life cycle [@Shah2009;@Salome2019; thus red snow algae could originate from green cells.
 Green snow is frequently found in waterlogged snow, sometimes alongside red snow, leading some to hypothesize that green snow could be an earlier life stage of red snow.
-Two studies comparing green and red snow using amplicon high-throughput sequencing (HTS) found they were caused by different species [@Lutz2016;@Terashoma2016]; however it remains an open question whether green snow can develop into red snow.
+Two studies comparing green and red snow using amplicon high-throughput sequencing (HTS) found they were caused by different species [@Lutz2015;@Terashoma2016]; however it remains an open question whether green snow can develop into red snow.
 
-<!-- chlainomonas -->
-Currently just two genera of snow algae are known to produce red pigment: *Sanguina* and *Chlainomonas*.
-*Sanguina* is cosmopolitan and widespread in summer snow in alpine and polar regions [@Prochazkova2019a], however the distinctive large red cells of *Chlainomonas* (Cr, Fig.
-1d) are only known from a few alpine sites in Europe, North America, and New Zealand [@Novis2008;@Prochazkova2018c].
-On the basis of microscopy and Sanger sequencing of field samples, these studies suggested that *Chlainomonas* blooms are limited to highly saturated, waterlogged snow overlying mountain lakes.
-*Chlainomonas* has not been detected in amplicon HTS surveys of snow algae blooms [@Terashima2017;@Lutz2017], and little is known of it's distribution, and relative abundance in the red snow micobiome.
-
-<!-- habitat pattens -->
-Few studies have addressed the habitat differences that give rise to different communities of snow algae to develop.
-
-Morphology-based cell counts suggest that green snow caused by *Chloromonas* is morprevalent in forested areas, and red snow containing cells similar in appearance to *Sanguina* prevalent in alpine areas [@Hoham1979a;@Nedbalova2008a].
-
-<!-- composition -->
-The snow algae literature suggests that snow algae blooms are typically dominated by a single species. Microscopy finds that
-
-Green snow can be caused by a variety of snow algae, including a diversity of *Chloromonas* species [@Remias2013a;@Terashima2017], *Microglena* [@Lutz2015], as well as Ochrophytes [@Remias2013;@Remias2019].
-Some *Chloromonas* produce pigments that turn snow orange, but this genus lacks the bright red astaxanthin characteristic of *Chlainomonas* and *Sanguina*.
-
-<!-- WHERE THEY ARE FOUND< HABITAT -->
-<!-- Where are snow algae species found? Do different habitats host different snow algae species? -->
-<!--   - why we care -->
-<!--   - what is unknown -->
-
-Chloromonas mostly known in culture, most diversity
-Chlainomonas few reports, found in NA, only in mountains not polar (but they arent looking)
-Sanguina cosmopolitan
-
-Cosmopolitan
-
-co-occurance:
-red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a].
-
-To understand the conditions that permit snow algae to grow, we need to define what species we are talking about; different species could have different habitat requirements.
-Before we develop testable hypotheses e.g.
-this species requires more of this nutrient, we need to see what is out there, and what species are found together, and under what conditions.
-
-Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013].
-In this micro-habitat, temperature is nearly constant, but snow moisture, pH, nutrients, and light can differ greatly within the snowpack depending on location of snow and time of year [@Loranger1988;@Kuhn2001;@Gorton2007;@Ganey2017].
-
-Although HTS found similar red snow, they lacked resolution, and used closed reference. Could be mislabelling OTUs. 
-They did not look at green snow, only alpine and polar red snow.
-
-Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a]
-
-, [@Lutz2015;@Remias2013], while  *Chlamydomonas cf.
-nivalis* is hypothesized to be specially adapted to well-drained, drier snow [@Ganey2018].
-Green snow is thought to likewise require high moisture content.
-
-<!-- WHEN/WHY THEY TRANSITION FROM GREEN TO RED<  -->
-<!-- Under what condions do snow algae develop pigment?  -->
-
-to crack the mystery of how and when snow algae divide, and produce pigment, we need to first must define what species we are talking about. 
-we need to know what species are in a bloom before we can figure out the red green problem. 
-No one has observed green snow transitioning to red in the field
-No one has cultured the algae that produce red pigment *Sanguina* and *Chlainomonas*
-  - despite over a century of study [@Kol1968] no one has sucessfully cultured this genus.
-  - microscopy pigment intermediated suggests they develop red pigment
-  - life cycle, related species are green algae that produce pigment in the cyst
-    - C rein
-    - H pluv
-Red and green snow could be the same species, or different species. 
-The few studies to compare red and green snow found that they were caused by different species: Lutz found green snow caused by Microglena, Terashima green snow caused by Chloromonas. 
-Are there blooms that transition from green to red? under what consitions do they develop pigment?
-
-<!-- how my methods will answer 1) whats out there 2) where is it 3) do snow algae transition from green to red -->
-<!-- brief overview of findings -->
-
-In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS (explain what these code for).
-Our aims were to: 1) see what species were present, 2) compare community composition between samples.
-We found ...
-18S is highly conserved and therefore suited to survey broadly across a variety of taxa; however, due to its high similarity between closely related species short segments of thie gene are poorly-differentiated among closely related taxa (ref).
-The *rbc*L plastid gene coding for the large subunit of RuBisCO is highly differentiated among photosynthetic organisms, and contains reference sequences for snow algae, making it an ideal complement to survey snow algae communities.
-
-Are there distinct communities of snow algae?
-  - does green snow turn red?
-  -
+In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy and 18S and *rbc*L amplicon HTS.
+The *rbc*L gene (coding for the large subunit of ruBiSCo) specifically targets phototrophs, and is more differentiated than the traditional 18S ribosomal marker gene, thus allowing us to specifically target snow algae.
+Our goal was to see what species were present, and how community composition varies over elevation.
 
@@
     -
+
+<!-- 3 HABITAT-->
+
+Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013].
+In this micro-habitat temperature is nearly constant, but snow moisture, pH, nutrients, and light can differ greatly within the snowpack depending on location of snow and time of day and year [@Loranger1988;@Kuhn2001;@Gorton2007;@Ganey2017].
+Thus
+
+Morphology-based cell counts suggest that green snow caused by *Chloromonas* is morprevalent in forested areas, while *Chlamydomonas cf.
+nivalis* was prevalent in alpine areas [@Hoham1979a;@Nedbalova2008a].
+However, different species of microalgae can look nearly identical, and the same species can look completely depending on the life stage [@Shah2009;@Salome2019]
+
+ *Sanguina* is cosmopolitan and widespread in summer snow in alpine and polar regions [@Prochazkova2019a], however the distinctive large red cells of *Chlainomonas* (Cr, Fig.
+1d) are only known from a few alpine sites in Europe, North America, and New Zealand [@Novis2008;@Prochazkova2018c].
+On the basis of microscopy and Sanger sequencing of field samples, these studies suggested that *Chlainomonas* blooms are limited to highly saturated, waterlogged snow overlying mountain lakes.
+*Chlainomonas* has not been detected in amplicon HTS surveys of snow algae blooms [@Terashima2017;@Lutz2017], and little is known of it's distribution, and relative abundance in the red snow micobiome.
+ <!-- x -->
 

### 2019-11-04 manuscript/manuscript.Rmd
discussion complete, spot edits in methods

@@
 ---
-title: "Snow algae community composition along an elevational gradient"
+title: "Contrasting snow algae communities along an elevational gradient"
 # distinct yet overlapping categories of snow algae community 
@@
 We used a novel *rbc*L and 18S amplicon high-throughput sequencing (HTS) approach to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada.
-We found distinct communities above and below treeline, with *Chloromonas* prevalent in forested sites and *Sanguina* above treeline, while *Chlainomonas* was abundant across all elevations.
+We found distinct communities at different elevations, with *Chloromonas* prevalent at lower elevations and *Sanguina* above treeline, while *Chlainomonas* was abundant across all elevations.
 Alpha diversity was variable; surprisingly, samples from highest elevations contained the highest diversity.
@@
 
-### *Run title*: Snow algae communities along elevational...
+### *Run title*: Contrasting snow algae communities
 
@@
 Some studies batch sequencing samples, and microscopy suggests that blooms are dominated by a single species.
-However, 18S amplicon surveys of red snow from across the Arctic [@Lutz2016a] and Japan [@Terashima2017] show that that snow algae blooms typically contain several secondary species; in addition to the dominant red snow algae, these studies found lower relative abundance of *Chloromonas*, as well as *Raphidonema* in Arctic sites.
+The red cells of *Sanguina nivaloides* are widespread and cosmopolitan [@Prochazkova2019a;@Segawa2018], but 18S amplicon surveys of red snow from across the Arctic [@Lutz2016a] and Japan [@Terashima2017] show that that snow algae blooms typically contain several secondary species; in addition to the dominant red snow algae (likely *Sangina*, as it had not been described at the time), these studies found lower relative abundance of *Chloromonas*, as well as *Raphidonema* in Arctic sites.
 33 red snow samples from across the Arctic were relatively homogeneous in community composition [@Lutz2016a], but genetic comparison of the morphologically diverse snow algae communities found in mountains from temperate latitudes [@Nedbalova2008a] is lacking.
@@
 Our taxa-specific *rbc*L primers showed considerable variation within *Chloromonas* and the closely-related *Chlainomonas* (Fig 2, S6).
-We found several distinct clusters among *Chloromonas* ASVs not matching any known species on GenBank (I, G, Fig.
+We found several distinct clusters among *Chloromonas* amplicon sequence variants (ASVs) not matching any known species on GenBank (I, G, Fig.
 2), but the majority of *Chloromonas* ASVs did not form distinct clusters (*Chloromonas spp* C, Fig.
@@
 
-We found snow algae community composition was strongly influenced by elevation.
+We found snow algae community composition was strongly influenced by elevation across all three datasets: *rbc*L, 18S, and cell count.
+*Sanguina* was prevalent at higher elevations, and *Chloromonas* dominant at lower elevations (Fig.
+1g, Fig.
+3, S5).
 Low elevation sites were compositionally similar, while high elevation samples showed inter-site variability (UniFrac Fig.
 3a, S8).
-Low elevation sites were dominated by *Chloromonas krienitzii* and *Chlainomonas rubra*, while *Sanguina nivaloides* was nearly exclusive to sites above treeline (Fig.
+Lower elevation sites were dominated by *Chloromonas krienitzii* and *Chlainomonas rubra*, while *Sanguina nivaloides* was only abundant in samples from above treeline (Fig.
 3b), along with one OTU of *Chloromonas*, and *Raphidonema* (H, I, Fig.
@@
 
-We found *Chlainomonas* was widespread and abundant in both alpine and subalpine snowpack.
+*Chlainomonas* was abundant across all elevations.
+Some samples were nearly completely dominated by *Chlainomonas*, but it was widespread in samples containing *Sanguina* and *Chloromonas krienitzii* as well (Fig.
+3b).
 
-Co-occurance
+## A green snow algae turning orange at the surface
 
-## A green snow algae turns orange at the surface
-
-At several sites dominated by *Chloromonas krienitzii* we sampled both the surface orange snow as well as the underlying green snow (Fig.
-4a) to see if the difference in color and cell morphology was due to different species, or different life stages of the same species inhabiting different habitats.
+Snow algae blooms dominated by *Chloromonas krienitzii* formed in low angle runnels, often in clearings at elevations near treeline.
+Based on our observation that the subsurface snow was green, and surface snow turned orange through the course of the season we sampled both orange and green snow to see if they were the same species (Fig.
+4a).
 The surface samples contained primarily orange cells resembling *Chloromonas krienitzii*, while the subsurface samples contained green cells including biflagellates and polar quadriflagellates (Fig.
@@
 
-We found snow algae communities were distinct, dominated by one of
+We found snow algae communities were highly variable from site to site, with communities at lower elevation dominated by *Chloromonas*, and higher elevation sites dominated by *Sanguina*.
+Microscopy studies of snow algae in USA and Slovakia found similar results, with *Chloromonas* is prevalent in forested areas and *Chlamydomonas cf.
+nivalis* above treeline [@Hoham1979a;@Nedbalova2008a].
+Our cell counts, 18S, and *rbc*L data support this finding.
 
-- different communities
-  - fields dominated by Chlainomonas, Sanguina, Chloromonas in microscopy studies
-    - differences between above and below treeline community in microscopy studies
-    - side-by-side comparison lacking, only amplicon studies targeting polar red snow, low taxonomic resolution
-      - unable to recognize Chlainomonas with 18S primers, whether this genera is widespread in melting snow worldwide
-  - Chloromonas commonly occurs with Sanguina
-    - we report several likely novel taxa of Chloromonas that are widespread throughout snow algae blooms
+The obvious question is: what environmental factors driving those differences in community composition?
+Further research will be needed to determine the specific environmental factors driving community differences, but our field observations does provide some testable hypotheses.
+In forested areas, green cells would be shaded from damaging solar radiation, while above treeline pigmented cells of *Sanguina* and *Chloromonas* are likely protected by their astaxanthin pigment [@Gorton2003;Holzinger2006].
+This fits with our observation that green cells of *Chloromonas krienitzii* were hidden below the snow surface, and pigmented cells were more prevalent on the surface (Fig.
+4); possibly pigment production is triggered as snow level melts lower, and cells are exposed to higher light at the surface.
+However, this does not explain why *Chloromonas krienitzii* was not found at higher elevations: our *C.
+krienitzii* dominated samples were taken from sites that were ~100 m from the nearest tree, and were not notabley shaded by topography in any way.
+Anecdotally, we observed *C.
+krienitzi* developed over running water, so their distribution could be linked to the presence of streams below the snowpack.
 
-- We noted higher diversity in
+Our results show that *Chlainomonas* is an important, hitherto unrecognized component of the snow algae microbiome.
+Prior to this study, *Chlainomonas* were thought to only bloom in highly saturated, waterlogged snow overlying mountain lakes [@Novis2008;@Prochazkova2018c], but we found *Chlainomonas* at most sites we sampled.
+Only one site high in *Chlainomonas* was near water (S9), the other *Chlainomonas* dominated sites were not near water, nor was the snow noticeably mushier than the surrounding snow.
+This does not rule out the possibility that these sites contained high moisture at some point; further sampling is needed to determine its habitat distribution, but the broad distribution of *Chlainomonas* in most samples suggests that this taxa may be an important component of the red snow microbiome.
 
-- what are the drivers of the community differences
-  - cell count, larger survey sample size supports our findings of different community types
-  - 
-- what our observations suggest for Chloro krienitzii life cycle
-  - one explaination, light triggers pigment production
-  -
+In contrast to typical alpine ecosystems with lower diversity at higher elevations, we observed the highest diversity at the highest elevation sites (W1.1, W1.2, P1.5, S8).
+These sites were our only samples collected on glaciers, and were also collected latest in the summer.
+These samples contained the highest proportion of *Raphidonema* (Trebouxiophyceae I, Fig.
+3), which is phylogenetically distant from Chlorophyceae, and thus could contribute heavily to phylogenetic diversity metrics, however Shannon and Simpson diversity was also highest for these three samples, which only took into account the evenness of OTU abundance, so the presence of *Raphidonema* alone cannot explain the even spread of OTUs at these sites.
+@Stibal2005 suggest that *Raphidonema* is a slow-growing opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow.
+Perhaps it's slow growth in snow means it is only found later in the season; or sites late in the season were closer to bare ground that seeded populations of this algae.
 
-- chlainomonas widespread in low relative abundance
-  - not just in wet lake areas
-  - hitherto unrecognized major component of the snow algae community
+Our use of cross-referenced 18S, *rbc*L, and cell count data uniquely allowed us to compare our findings between datasets.
+*Sanguina* was highly correlated between all three.
+One notable difference was that no reads were assigned to *Chlainomonas* from our 18S library, despite this genera being highly abundant in *rbc*L.
+Samples contaiing primarily *Chlainomonas* in our *rbc*L and cell count data were marked as containing only *Chloromonas* in 18S.
+At the time of publication, only two 18S *Chlainomonas* were present on GenBank (MF803743.1, MF803745.1); possibly our *Chlainomonas* sequences are different enough from the reference that they were mis-labelled as *Chloromonas*, especially likely given how closely related the two genera are.
+Samples dominated by *Sanguina* in our 18S library contained lower abundance in *rbc*L, which is likely due to primer bias: because our custom *rbc*L primers were designed to target *Chloromonas* and *Chlainomonas* (*Sanguina* not being available at the time), this likely led to a over-representation of these taxa in our *rbc*L library.
 
-- bias
-  - extraction
-  - primer bias
-    - Because our rbcL primers were designed off of *Chloromonas* and *Chlainomonas*, it is possible they would have missed anything not closely related.
-However, they were able to detect both *Sanguina*, and even algae in a different class, Trebouxiophyceae, so although primer bias is always possible it at least should be able to detect Chlorophyceae and Trebouxiophyceae.
-  - false 0s?
-  - cross contam? (below detection threshold of qubit)
-    - sequencing of negative controls, there is stuff in there
-    -
-
-<!-- 3 HABITAT-->
-
-Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013].
-In this micro-habitat temperature is nearly constant, but snow moisture, pH, nutrients, and light can differ greatly within the snowpack depending on location of snow and time of day and year [@Loranger1988;@Kuhn2001;@Gorton2007;@Ganey2017].
-Thus
-
-Morphology-based cell counts suggest that green snow caused by *Chloromonas* is morprevalent in forested areas, while *Chlamydomonas cf.
-nivalis* was prevalent in alpine areas [@Hoham1979a;@Nedbalova2008a].
-However, different species of microalgae can look nearly identical, and the same species can look completely depending on the life stage [@Shah2009;@Salome2019]
-
- *Sanguina* is cosmopolitan and widespread in summer snow in alpine and polar regions [@Prochazkova2019a], however the distinctive large red cells of *Chlainomonas* (Cr, Fig.
-1d) are only known from a few alpine sites in Europe, North America, and New Zealand [@Novis2008;@Prochazkova2018c].
-On the basis of microscopy and Sanger sequencing of field samples, these studies suggested that *Chlainomonas* blooms are limited to highly saturated, waterlogged snow overlying mountain lakes.
-*Chlainomonas* has not been detected in amplicon HTS surveys of snow algae blooms [@Terashima2017;@Lutz2017], and little is known of it's distribution, and relative abundance in the red snow micobiome.
- <!-- x -->
-
-Although *Chlainomonas* is reported to only grow in slushy snow above alpine tarns [@Novis2008;@Prochazkova2018c], we found *Chlainomonas* was present in low relative abundance at most sites sampled.
-Only one site high in Chlainomonas was near water (S9), the other Chlainomonas dominated sites were not near water, nor was the snow noticeably mushier than the surrounding snow.
-This does not rule out the possibility that these sites contained high moisture at some point; further sampling is needed to determine its habitat distribution.
-
-Snowpack above and below treeline are very different habitats, with forest providing shade (and therefore less evaporation).
-The red pigment of Sanguina and Chlainomonas (astaxanthin) is likely an adaptation to high light (Gorton and Vogelmann 2003; Holzinger and Lütz 2006), and it has been suggested that Sanguina is adapted to the drier snowpacks that form at higher elevations (Dial, Ganey, and Skiles 2018).
-
-Raphidonema and related ASVs were restricted to high alpine sites, with the highest relative abundance occurring in samples from August and September.
-This fits with the interpretation of Stibal and Elster (2005), who suggest that Raphidonema is an opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow.
-However, rare taxa should not be discounted as playing an important role in ecosystem functioning (Jousset et al.
-2017).
-
-Hoham, Roemer, and Mullet (1979) and Nedbalová and Lukavský (2008), who found primarily Chlamydomonas cf.
-nivalis above treeline and Chloromonas below treeline.
-However, in microscopy surveys we infrequently observed Sanguina below treeline, and Chloromonas krienitzii above treeline.
+# Conclusions
 
@@
 
-**Field sampling and microscopy.** We collected snow algae from mountains near Vancouver, British Columbia throughout the summer of 2018.
-To capture the extent of snow algae diversity in our local mountains we collected as many samples as possible from different elevations, dates, mountains, and micro-habitats within the snow <!--insert suppl table-->.
+## Field sampling and microscopy
+
+We collected snow algae from mountains near Vancouver, British Columbia throughout the summer of 2018.
+To capture the extent of snow algae diversity in our local mountains we collected as many samples as possible from different elevations, dates, mountains, and micro-habitats within the snow (S2).
 In total we collected 310 colored snow samples from 13 different mountains on 33 different dates from elevations between 880 m and 2150 m above sea level.
-In early season we sampled at lower elevations, moving uphill as the snow algae bloom progressed upwards in elevation [@Takeuchi2013].
-We observed the first small spots of red snow above treeline on June 20.
+In early season we sampled at lower elevations, moving uphill as the snow algae bloom progressed upwards in elevation.
 
@@
 Most samples were viewed using light microscopy, characterizing the dominant cell morphology in each sample.
-We prepared a slide of cells fixed in 2% gluteraldehyde, and quantified the most common cell morphologies with a cell count to 100, classifying each cell as either *Sanguina*, *Chloromonas cf.
-nivalis*, *Chloromonas cf.
-brevispina*, *Chloromonas krienitzii* *Chlainomonas*, or "other".
+We prepared a slide of cells fixed in 2% gluteraldehyde, and quantified the most common cell morphologies with a cell count to 100, classifying each cell as either *Sanguina nivaloides* [@Prochazkova2019a], *Chloromonas cf.
+nivalis* [@Matsuzaki2019;@Prochazkova2017], *Chloromonas cf.
+brevispina* [@Nedbalova2008, @Matsuzaki2015], *Chloromonas krienitzii* [@Matsuzaki2015] *Chlainomonas rubra* [@Novis2008], or "other" based on similarity to published photographs.
 
-Cite refs for why we are calling them that.
+## DNA extraction and amplicon library preparation
 
-#
-
-**DNA extraction and amplicon library preparation.** We selected 35 out of 310 samples for rbcL and 18s high-throughput amplicon sequencing.
+We selected 35 out of 310 samples for rbcL and 18s high-throughput amplicon sequencing.
 We chose these samples to represent the variation in date, elevation, geographic location, snow color, micro-habitat, and cell morphology.
@@
 We mini-pestled 5 mg of sample at room temperature to physically rupture cell walls.
-Based on our observation that samples dominated by *Sanguina* cell morphologies had lower DNA yields, and had many intact cells following various lysis methods, we used 20 mg of these samples to increase our DNA yield.
+Based on our observation that samples dominated by *S.
+nivaloides* cell morphologies had lower DNA yields, and had many intact cells following various lysis methods, we used 20 mg of these samples to increase our DNA yield.
 
@@
 
-#
+## Bioinformatic processing
 
-**Bioinformatic processing.** Reads were demultiplexing using cutadapt [@Martin2011].
-Samples were filtered and trimmed, errors removed, dereplicated, pair end reads merged, and chimeras removed in DADA2 [@Callahan2017].
+Reads were demultiplexing using cutadapt [@Martin2011].
+Samples were filtered and trimmed, errors removed, dereplicated, pair end reads merged, and chimeras removed following the default pipeline of DADA2 [@Callahan2017].
 <!-- singletons, failed samples removed, remove sequences based on negative controls, adjust for 0's, -->
 
-We assigned taxonomy using a custom reference databases for both rbcL and 18S, since snow algae are not well-represented in SILVA or PR2, and there is no curated database for rbcL.
+We assigned taxonomy for both *rbc*L and 18S ASVs using IDTaxa with threshold set to 50 [@Murali2018].
+We made custom reference databases for each to include snow algae GenBank sequences.
 We made our rbcL database by downloading 30,865 algae rbcL sequences from GenBank with the query “rbcl[gene] AND (algae OR chlorophyta OR trebouxiophyceae) NOT (18s OR ribosomal OR psaB OR atpB) AND 300:2000[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion”.
@@
 nivalis* are polyphyletic we re-labeled these as “unassigned *Chloromonas*”.
-<!--Where multiple sequences were present for a single species we generated a consensus sequence to avoid imbalance where some groups have far more representatives than others [@Murali2018] --><!--To  re-labeled GenBank accessions LC012733.1 and LC012735.1 as Chloromonas krienitzii, LC012734.1 as Chloromonas “Cryocystis”-type, and LC012744.1 to Chloromonas miwae, and LC012743 and LC012745 to “Chloromonas Group_B” 6`.--> Species designated by morphology alone that were not genetically different from other species in their genera were annotated only to genus level.
-<!-- all Chlainomonas to genus level only --> We assigned rbcL ASV taxonomy using IDTaxa with threshold = 50.
+To assign 18S reads we initially ran our ASVs against SILVA [@Quast2013], and then re-ran ASVs assigned to Chlorophyta on our custom 18S snow algae database based on the following GenBank query: "(18S OR ribosomal)[gene] AND (chloromonas OR chlainomonas OR sanguina OR raphidonema OR KMY-2018) NOT (rbcL OR psaB OR atpB) AND 300:3500[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion".
+This database was edited to reflect the most recent snow algae taxonomy and remove ambiguous annotations [@Matsuzaki2019;@Prochazkova2018c].
 
-To assign 18S reads we initially ran our ASVs against SILVA [@Quast2013], and selected ASVs assigned to Chlorophyta.
-We then ran this subset of ASVs on our custom 18S snow algae database based on the following GenBank query: "(18S OR ribosomal)[gene] AND (chloromonas OR chlainomonas OR sanguina OR raphidonema OR KMY-2018) NOT (rbcL OR psaB OR atpB) AND 300:3500[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion".
-This database was edited to reflect the most recent snow algae taxonomy and remove ambiguous annotations [@Matsuzaki2019;@Prochazkova2018c].
-We assigned taxonomy using IDTaxa with threshold = 50.
+As most OTU clustering algorithms are optimized for 18S data, we chose to assign high level OTUs based on the results of t-SNE clustering [@Graham-Rowe1999] (Fig.
+2a).
+We tested this with perplexity values ranging from 1 to 100, and found comparable results with perplexity ranging from 10 to 50.
+We compared sample similarity using UniFrac with default settings [@Lozupone2005].
 
-We assigned OTUs with t-SNE, based on visually distinct clusters [@Graham-Rowe1999] <!--BibTeX header messed up-->(Fig.
-1).
-We split the largest cluster (containing the vast majority of the *Chloromonas* reads) into three clusters, one containing all ASVs assigned to *Chloromonas krienitzii*, and one containing all ASVs assigned to *Chloromonas muramotoi* and *Chloromonas hohamii*.
-The OTUs produced by this naive method were consistent across different methods of visualization (Suppl.
-Fig.
-1 and 2).
-<!--Other OTU classifiers such swarm [@Mahe2014] are optimized for 18S data, and we found swarm did a poor job of clustering when visualised with MDS, t-SNE, and phylogenetic trees, with various values of d.
-<!--and CROP [@Chen2013].-->
-<!--OTU clustering with t-SNE: based on poor results with other classifiers, we used tsne.
-since tsne is known to find patterns where there are none depending on the perplexity, we ran this test with perplexity ranging from 5-100, and found the same large scale clusters each time.
--->
-
-UniFrac [@Lozupone2005]
-
-**Software information.** All analysis was conducted in R 3.6.1 with tidyverse 1.2.1, with dada2 1.12.1, DECIPHER 2.12.0, Rtsne 0.15, GUniFrac 1.1, and vegan 2.5-6.
-Phylogenetic analysis was carried out in IQTree [@Nguyen2015] and visualized with ggtree [@Yu2017].
-
-**Data availability.** All raw fastq files are freely available on the European Nucleotide Archive under the project accession PRJEB34539.
-All scripts are available at cengstro.github.io/projects/sa_rbcl.
+For software information, see Supplementary Table 9.
+All raw fastq files are freely available on the European Nucleotide Archive under the project accession PRJEB34539.
+All scripts are available at https://github.com/cengstro/bc_snow_algae_amplicon.
 
@@
 We wish to thank Leah Tooman (Simon Fraser University) for assistance with sequencing, and Chris Rushton (Simon Fraser University) for assistance with bioinformatics.
-<!--We also wish to thank Forest Rohwer (UCSD) for the inspiration that led to this project.--> This project was funded with a Sector Innovation Grant from Genome BC (SIP016), and a NSERC Individual Discovery Grant, both awarded to LQ.
+This project was funded with a Sector Innovation Grant from Genome BC (SIP016), and a NSERC Individual Discovery Grant, both awarded to LQ.
 
@@
 
-### *Supplementary information*.
-
-### *Competing financial interests.* The authors declare that the research was conducted in the absence of any
-commercial or financial relationships that could be construed as a potential
-conflict of interest.
+The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
 

### 2019-11-04 manuscript/manuscript.Rmd
conclusions done

@@
 
+Our findings highlight the remaining unexplored diversity in the snow algae microbiome.
+One abundant OTU of algae, "Algae H" is widespread and abundant, but poorly matches all *rbc*L GenBank sequences.
+However, out 18S sequences did not detect any such cluster.
+Non-specific priming on an unsequenced gene is one possibility, another is that this represents a novel clade of snow algae, or perhaps an endosymbiotic algae within the ciliates that were abundant in many of our samples.
+Interestingly, the sample with the highest relative abundance of this OTU (X1.1) had red cells that were visually distinct from other *Sanguina*, with thicker cell walls that were slightly less than perfectly spherical in some cells (S4); clearly, further investigation is needed.
+
 # Conclusions
+
+Our side by side comparison of snow algae communities shows that they are more diverse than previously thought, and suggests that some snow algae may be adapted to specific micro-habitats such as shade and moisture within the melting snowpack.
+As the different snow algae species may have varying habitat preferences, different taxa could have different life cycles that permit them to survive in their niche, thus future work to understand the biological mechanisms of snow algae blooms should take community composition into account.
+Different communities of snow algae will likely affect albedo and local snowpack chemistry in different ways [@Lutz2015], and so different snow algae communities could impact the microbiome and greater ecosystem in unique ways.
 
@@
 
-We collected snow algae from mountains near Vancouver, British Columbia throughout the summer of 2018.
+We collected snow algae from mountains near Vancouver, British Columbia, Canada throughout the summer of 2018.
 To capture the extent of snow algae diversity in our local mountains we collected as many samples as possible from different elevations, dates, mountains, and micro-habitats within the snow (S2).
@@
 Immediately after, samples were stored at -20 &deg;C for up to eight months until DNA extraction.
-Most samples were viewed using light microscopy, characterizing the dominant cell morphology in each sample.
-We prepared a slide of cells fixed in 2% gluteraldehyde, and quantified the most common cell morphologies with a cell count to 100, classifying each cell as either *Sanguina nivaloides* [@Prochazkova2019a], *Chloromonas cf.
+
+We used light microscopy to characterize the cell morphologies in each sample.
+We prepared a slide of cells fixed in 2% gluteraldehyde, and at 400x magnification starting from the center of the slide counted up to 100 cells.
+We classified each cell as either *Sanguina nivaloides* [@Prochazkova2019a], *Chloromonas cf.
 nivalis* [@Matsuzaki2019;@Prochazkova2017], *Chloromonas cf.
-brevispina* [@Nedbalova2008, @Matsuzaki2015], *Chloromonas krienitzii* [@Matsuzaki2015] *Chlainomonas rubra* [@Novis2008], or "other" based on similarity to published photographs.
+brevispina* [@Nedbalova2008, @Matsuzaki2015], *Chloromonas krienitzii* [@Matsuzaki2015] *Chlainomonas rubra* [@Novis2008], or "Other" based on similarity to published photographs.
 
@@
 
-We selected 35 out of 310 samples for rbcL and 18s high-throughput amplicon sequencing.
+We selected 33 out of 310 samples for rbcL and 18s high-throughput amplicon sequencing.
 We chose these samples to represent the variation in date, elevation, geographic location, snow color, micro-habitat, and cell morphology.
@@
 Samples were filtered and trimmed, errors removed, dereplicated, pair end reads merged, and chimeras removed following the default pipeline of DADA2 [@Callahan2017].
-<!-- singletons, failed samples removed, remove sequences based on negative controls, adjust for 0's, -->
+<!-- singletons, failed samples removed, remove sequences based on negative controls, adjust for 0's, sequencing depth, -->
 

### 2019-11-04 functions/ggplot_functions.R
init

@@ -1,0 +1,36 @@
+library(tidyverse)
+
+##### none of this is working
+
+otus %>% distinct(otu_name)
+
+
+otu_ord <- c("Chlainomonas A","Sanguina B", "Other Chloromonas","Chloromonas krienitzii","Chloromonas muramotoi C","Chloromonas D", "Algae E","Chloromonas F", "Trebouxiophyceae G")
+
+barplot <- 
+
+
+plot_rel_abund_bar <- function(rel_abund_tbl, species, species_order, sample_order){
+  # input: long form tibble with
+  cust_theme <- theme(legend.position = "none",
+                      panel.grid.major.x = element_blank(), # remove vertical lines in background
+                      panel.grid.minor.x = element_blank(),
+                      axis.line=element_blank(), # remove x axis
+                      axis.title.x=element_blank(), 
+                      axis.text.x=element_blank(),
+                      axis.ticks.x=element_blank(),
+                      panel.spacing = unit(-0.01, "lines"), # remove space between facet panels
+                      strip.text.x = element_text(angle = 90, hjust=1.0, vjust = 0.8), # rotate facet labels 90
+                      strip.background = element_blank()) # remove grey boxes around facet labs
+  
+  ggplot(rel_abund_tbl, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
+    geom_bar(aes(fill=otu_name %>% fct_relevel(otu_ord)), stat="identity") + # classic stacked relative abundance plot
+    coord_flip() +
+    facet_grid(~otu_name %>% fct_relevel(otu_ord), switch = "x") +
+    labs( x="Sample ID", tag="B") +
+    cust_theme
+}
+
+plot_rel_abund_stacked_bar <- function(){
+  
+}

### 2019-11-04 code/2018_sample_map.Rmd
minor, renamed obj

@@
   scale_color_gradientn(colours = terrain.colors(8)[1:7], limits = c(850,2200)) +
-  geom_point(data = city_coords, shape = 15) +
-  geom_text(data = city_coords, aes(label = name), hjust = 1.1)+
+  geom_point(data = cities, shape = 15) +
+  geom_text(data = cities, aes(label = name), hjust = 1.1)+
   labs(title = "Snow algae sample sites, summer 2018", color="Elevation (m)")
@@
 
-# ggsave(filename = here("")
+ggsave(filename = here::here("manuscript/figs/map.png"), plot=full_map)
 

### 2019-11-04 code/cellcount.Rmd
changes

@@
 
-```{r heatmap, echo=F, fig.height=10, fig.width=6}
+```{r heatmap, echo=F, fig.height=5, fig.width=3}
 p <- cellct %>% 
@@
 
-morpho_order <- c("Green cell",
-                  "Sanguina 'citrine'",
-                  "Sanguina nivaloides",
-                  "Sanguina nivaloides 'ruby'",
-                  "Green cell",
-                  "Chlainomonas",
-                  "Chloromonas krienitzii",
-                  "Chloromonas cf. brevispina",
-                  "Chloromonas cf. nivalis",
-                  "Chloromonas 'spiny'",
-                  "Chloromonas 'oval'",
-                  "Chloromonas 'zuke'"
-                  )
+short_names <- p %>% 
+  mutate(short_morpho_name = case_when(morpho_sp=="Chlainomonas"~"Cr",
+                                       morpho_sp=="Chloromonas cf. brevispina"~"Cb",
+                                       morpho_sp=="Chloromonas cf. nivalis"~"Cn",
+                                       morpho_sp=="Chloromonas krienitzii"~"Ck",
+                                       morpho_sp=="Green cell"~"G",
+                                       morpho_sp=="Other"~"O",
+                                       morpho_sp=="Sanguina nivaloides"~"Sn"))
 
-my_theme <- theme(axis.text.x = element_text(hjust=1,angle=45))
+my_theme <- theme(axis.title.y =element_blank(),
+                  axis.text.y=element_blank(),
+                  axis.ticks.y=element_blank(), 
+                  legend.position = "none")
 
-ggplot(p, aes(y=alias %>% fct_reorder(elev_m), 
-              x=morpho_sp)) +
+cellct_heatmap <- ggplot(short_names, aes(y=alias %>% fct_reorder(elev_m), 
+              x=short_morpho_name)) +
   geom_tile(aes(alpha = percent_of_sample, fill = morpho_sp)) + # color = "white" %>% fct_reorder(morpho_order)
-  labs(x="Cell morphology", y= "Sample ID", fill = "Cell morphology", alpha="Relative abundance") +
-  guides(fill=F) +
+  labs(x="Cell morphology", y= "Sample ID (in order of elevation)", fill = "Cell morphology", alpha="Relative abundance") +
   scale_alpha(range = c(0.1, 1)) +
+  guides(alpha=FALSE)+
   my_theme
+
+cellct_heatmap
 ```
+
+```{r, elevation_bar, echo=F}
+elevation_ranges <- short_names %>% 
+  distinct(elev_m, sample_id) %>% 
+  arrange(elev_m)
+
+ggplot(elevation_ranges, aes(x=sample_id %>% fct_reorder(elev_m), y=elev_m))+
+  geom_bar(stat = "identity")+
+  coord_flip()
+```
+
+```{r}
+ggsave(filename=here("manuscript/figs/1_field/cellct_heatmap.pdf"), plot=cellct_heatmap, height = 7, width = 2)
+```

### 2019-11-04 code/community_comparison_18s.Rmd
split up wrangle tasks into scripts, made seperate csvs
path stuff

@@
 ---
-title: "18S Community Comparison"
+title: "Just algae 18S community comparison"
 output: html_notebook
@@
 
-```{r libs, results="hide"}
+```{r libs, include=F}
 library(tidyverse)
@@
 
-```{r set}
-theme_set(theme_classic())
-```
-
-```{r}
-path_18s_ra <- here("data/clean/only_algae_18s_rel_abund.csv")
-clean_18s_ra <- read_csv(path_18s_ra)
-
-path_tree <- here("intermediate_output/tree/algae_18s/algae_18s_seqs.fasta.treefile")
-tree_18s <- read.tree(path_tree)
-
-path_full_18s_tree <- here("intermediate_output/tree/18s/")
-full_tree_18s <- read.tree(path_full_18s_tree)
+```{r, include=F}
+rel_abund <- read_csv(here("data/03_filtered/18s_algae/18s_algae.filtered_cleaned_rel_abund.csv"))
+tree_18s <- read.tree(here("output/tree/algae_18s/algae_18s_seqs.fasta.treefile"))
 
@@
 field <- read_csv(path_field)
-
-path_alias <- here("data/sample_aliases.csv")
-alias <- read_csv(path_alias)
 ```

### 2019-11-04 code/rbcl_community_compare.Rmd
minor

@@
 # run NMDS ordination on results 
+set.seed(123)
 nmds <- metaMDS(unifrac, autotransform = F)
@@
 ```{r setup_rel_abund_barplot, include=F}
-plot_data <- rel_abund %>% 
+otu_rel_abund <- rel_abund %>% 
   left_join(otus, by="asv_id") %>% 
@@
   left_join(field, by="sample_id") %>%
-  ungroup() %>% 
+  ungroup()
+
+otu_rel_abund_round_to_1_percent <- otu_rel_abund %>% 
   mutate(rel_abund = if_else(rel_abund <0.01, 0.01, rel_abund)) # artificially boost rel abund <1% to 1% so its visible on plot
@@
 # check rel abund sums ~= 1 (will be slightly higher, due to rounding)
-plot_data %>%
+otu_rel_abund_round_to_1_percent %>%
   group_by(alias) %>%
@@
 
-otu_names %>% distinct(otu_name)
+otus %>% distinct(otu_name)
 
@@
 
-barplot <- ggplot(plot_data, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
+barplot <- ggplot(otu_rel_abund_round_to_1_percent, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
   geom_bar(aes(fill=otu_name %>% fct_relevel(otu_ord)), stat="identity") + # classic stacked relative abundance plot
@@
 
+Faith's Phylogenetic Diversity, sum of branch lengths in a sample's phylogenetic tree (does not take relative abundance into account)
 ```{r faiths, echo=F}
-re_wide_mat <- ra_wide %>% as.matrix()
+ra_wide_mat <- ra_wide %>% as.matrix()
 
-faiths <- re_wide_mat %>% 
+faiths <- ra_wide_mat %>% 
   picante::pd(rbcl.tree) %>% 
@@
 
-```{r}
+```{r, echo=F}
 faith_elev_plot <- ggplot(faiths, aes(y=PD, x=elev_m))+
@@
 faith_elev_plot
+```
+
+Shannon and Simpson diversity indices (these take relative abundance into account, but require OTUs)
+```{r, include=F}
+otu_rel_abund_mat <- otu_rel_abund %>% 
+  select(sample_id, otu_name, rel_abund) %>% 
+  pivot_wider(names_from = otu_name, values_from = rel_abund, values_fill = list(rel_abund=0)) %>% 
+  column_to_rownames("sample_id") %>% 
+  as.matrix()
+
+# use ra_wide_mat for asv based diversity
+shannon <- otu_rel_abund_mat %>% 
+  diversity(index = "shannon") %>% 
+  enframe() %>% 
+  dplyr::rename(sample_id = name, shannon = value)
+simpson <- otu_rel_abund_mat %>% 
+  diversity(index = "simpson") %>% 
+  enframe() %>% 
+  dplyr::rename(sample_id = name, simpson = value)
+otu_diversity <- shannon %>% 
+  left_join(simpson, by="sample_id") %>% 
+  left_join(field, by="sample_id")
+
+# plot diversity of samples
+ggplot(otu_diversity, aes(x=elev_m, y=shannon))+
+  geom_point() +
+  geom_text_repel(aes(label=alias)) +
+  geom_smooth(method = "lm")
+```
+
+Chao
+```{r}
+# otu_rel_abund_mat %>% 
+#   mutate
+#   # head(1) %>% 
+#   as.matrix() %>% 
+#   fossil::chao1()
 ```
@@
 ```{r output, include=F}
-ggsave(filename = here("manuscript/figs/3_community/combined_fig_3.pdf"), plot=combined_fig_3, height=10, width=7)
+ggsave(filename = here("manuscript/figs/3_community/combined_fig_3.png"), plot=combined_fig_3, height=10, width=7)
 ggsave(filename = here("manuscript/figs/supplementary/diversity.png"), plot = faith_date_plot, height = 5, width = 7)

### 2019-11-04 code/wrangle/tidy_18s_algae_taxonomy.R
init

@@ -1,0 +1,45 @@
+# tidy 18s taxonomy-- just snow algae
+library(tidyverse)
+library(here)
+
+taxonomy_tbl <- read_csv(here("data/01_raw/taxonomy_18s_genbank_snow_algae.csv"))
+asv_seq_data <- read_csv(here("data/asv_seq_data/asv_seq_key_18s.csv"))
+
+
+
+taxonomy_tbl %>% distinct(V2, V3, V4, V5, V6, V7, V8, V9) %>% view()
+
+
+  
+# rename cols in taxonomy tbl
+cleaned <- taxonomy_tbl %>% 
+  select(-V2, -V3, -V4, -V21, -V31, -V41) %>% # only interested in assignments at class level or below (Chlorophyceae vs Trebouxiophyceae)
+  dplyr::rename(asv_sequence = asv,
+                class = V5,
+                order = V6,
+                family = V7, 
+                genus = V8, 
+                species = V9,
+                class_conf = V51,
+                order_conf = V61,
+                family_conf = V71, 
+                genus_conf = V81, 
+                species_conf = V91) %>% 
+  mutate_at(c("class","order","family","genus","species"),
+            ~str_replace(., pattern="unclassified", replacement = NA_character_)) %>% 
+  filter(!(is.na(class)))
+
+
+cleaned %>% view()
+
+# replace sequence with asv_id
+cleaned %>% 
+  left_join(asv_seq_data, by="asv_sequence") %>% view()
+  select(asv_id, class:species_conf)
+
+  
+  
+### problem: the sequences in silva (that we used to generate the key) do not match any of the sequences in the 18s
+  
+  
+  

### 2019-11-04 code/wrangle/tidy_silva_tax_assignment.R
minor

@@ -67,6 +67,7 @@ silva_3 <- key %>%
   left_join(silva_2, by="asv_sequence") %>% 
   select(-asv_sequence)
 
+silva_3
 
 write_csv(key, here("output/asv_seq_key_18s.csv"))
 writeXStringSet(seqs_to_fasta, here("output/all_18s.fasta"))

### 2019-11-04 manuscript/figures.Rmd
added all figs for v1

@@
 library(knitr)
-opts_chunk$set(fig.width=12, fig.height=8, fig.path='figs/', echo=FALSE, warning=FALSE, message=FALSE)
+library(tidyverse)
+library(here)
+opts_chunk$set(fig.width=12, fig.height=8, fig.path='figs/', fig.align= "center", echo=FALSE, warning=FALSE, message=FALSE)
 ```
@@
 ```{r input, include=F}
-# field <- knitr::include_graphics("temp.png")
-asvs <- include_graphics("figs/2_taxonomy/combined.png")
-communities <- include_graphics("figs/3_community/combined.png")
-# depth
+field <- include_graphics(here("manuscript/figs/1_field/field.png"))
+asvs <- include_graphics(here("manuscript/figs/2_taxonomy/combined.png"))
+communities <- include_graphics(here("manuscript/figs/3_community/combined_fig_3.png"))
+depth <- include_graphics(here("manuscript/figs/4_depth_comparison/depth_fig.png"))
+
+field_cap <- "Representative photos and microphotos of snow algae.
+A.
+Red snow bloom above treeline, sample G1.1.
+B.
+Green snow hidden below surface, same sample site as S9.
+C.
+Bronze snow spots below trees.
+D.
+Red snow containing *Sanguina nivaloides* and *Chlainomonas rubra* morphologies.
+All scale bars 30 um.
+E.
+Orange-green snow containing *Chloromonas krienitzii* and *Chloromonas cf.
+nivalis*.
+F.
+Snow algae from green snow, including *Chloromonas cf.
+brevispina* and *Chlainomonas rubra*.
+G.
+Heatmap of cell count data, with samples on the y axis arranged from low to high elevation, and cell type on the x axis."
+
+asv_cap <- "A.
+t-SNE clustering of snow algae *rbc*L ASVs.
+Each point represents an ASV, and is colored by taxonomic assignment, and sized by total relative abundance.
+OTU labels are superimposed on clusters.
+B.
+MDS clustering of the same ASVs as in A, showing genetic distnaces between ASVs.
+Stress = 0.13"
+
+community_cap <-"Snow algae community composition.
+A.
+NMDS clustering of snow algae sample UniFrac, colored by elevation.
+B.
+Relative abundance barplot of OTUs present in snow algae samples.
+Y-axis is ordered from lowest to highest sample elevation."
+
+depth_cap <- "A.
+Representative field photo of snow runnel, with orange on surface and green below.
+B.
+Microphoto of orange snow on surface of sample S9.1.
+C.
+Microphoto of snow algae below surface of sample S9.1.
+D.
+Relative abundance barplot of OTUs in three snow runnels.
+Sample H2.10 depth and surface samples were taken 1 week apart, other depth/surface pairs are from the same date."
 ```
 
-```{r pressure, echo=FALSE, fig.cap="A caption", out.width = '100%'}
+```{r, fig.cap=field_cap}
+field
+```
+```{r, fig.cap=asv_cap}
+asvs
+```
+
+```{r, fig.cap=community_cap}
 communities
@@
 
-Figure 1.
+```{r, fig.cap=depth_cap, out.width = '100%'}
+depth
+```

### 2019-11-04 manuscript/manuscript.Rmd
first draft to LQ

@@
 Red snow algae blooms can cover vast areas of snowfield, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017].
-Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], which can threaten water supplies [@Staudinger2014] and worsten drought [@Barnhart2016].
+Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], which can threaten water supplies [@Staudinger2014] and worsen drought [@Barnhart2016].
 Thus, snow algae could play an important role in snow melt regulation, but little is known about how these blooms form.
@@
 Snow algae blooms can color snow red, orange, or green, and can one or contain multiple species of algae.
-Most snow algae blooms are caused by green algae of class Chlorophyceae [@Leya2013], although blooms caused by Trebouxiophyceae [@Lutz2015] or Chrysophycae [@Remias2013;@Remias2019] have been reported.
-Among the Chlorophyceae, both *Sanguina nivaloides* [@Prochazkova2019a] or *Chlainomonas* [@Prochazkova2018c] produce the red pigment astaxanthin respoponsible for the red color of some snow algae blooms.
+Most snow algae blooms are caused by green algae of class Chlorophyceae [@Leya2013], although blooms caused by Trebouxiophyceae [@Lutz2015] or Chrysophyceae [@Remias2013;@Remias2019] have been reported.
+Among the Chlorophyceae, both *Sanguina nivaloides* [@Prochazkova2019a] or *Chlainomonas* [@Prochazkova2018c] produce the red pigment astaxanthin responsible for the red color of some snow algae blooms.
 Orange snow is reported to be caused by *Sanguina aurentia* [@Prochazkova2019a] and some species of *Chloromonas* [@Prochazkova2019c;@Remias2013a], while algae lacking secondary pigments color the snow green, such as *Chloromonas pichinchae* [@Hoham1975].
@@
 Some studies batch sequencing samples, and microscopy suggests that blooms are dominated by a single species.
-The red cells of *Sanguina nivaloides* are widespread and cosmopolitan [@Prochazkova2019a;@Segawa2018], but 18S amplicon surveys of red snow from across the Arctic [@Lutz2016a] and Japan [@Terashima2017] show that that snow algae blooms typically contain several secondary species; in addition to the dominant red snow algae (likely *Sangina*, as it had not been described at the time), these studies found lower relative abundance of *Chloromonas*, as well as *Raphidonema* in Arctic sites.
+The red cells of *Sanguina nivaloides* are widespread and cosmopolitan [@Prochazkova2019a;@Segawa2018], but 18S amplicon surveys of red snow from across the Arctic [@Lutz2016a] and Japan [@Terashima2017] show that that snow algae blooms typically contain several secondary species; in addition to the dominant red snow algae (likely *Sanguina*, as it had not been described at the time), these studies found lower relative abundance of *Chloromonas*, as well as *Raphidonema* in Arctic sites.
 33 red snow samples from across the Arctic were relatively homogeneous in community composition [@Lutz2016a], but genetic comparison of the morphologically diverse snow algae communities found in mountains from temperate latitudes [@Nedbalova2008a] is lacking.
 
-Despite over a century of study, no one has sucessfully cultured the snow algae that produce red secondary pigments.
-In many green algae cells produce secondary pigments in just one life stage, remaining green for much of their life cycle [@Shah2009;@Salome2019; thus red snow algae could originate from green cells.
+Despite over a century of study, no one has successfully cultured the snow algae that produce red secondary pigments.
+In many green algae cells produce secondary pigments in just one life stage, remaining green for much of their life cycle [@Shah2009;@Salome2019]; thus red snow algae could originate from green cells.
 Green snow is frequently found in waterlogged snow, sometimes alongside red snow, leading some to hypothesize that green snow could be an earlier life stage of red snow.
@@
 In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy and 18S and *rbc*L amplicon HTS.
-The *rbc*L gene (coding for the large subunit of ruBiSCo) specifically targets phototrophs, and is more differentiated than the traditional 18S ribosomal marker gene, thus allowing us to specifically target snow algae.
+The *rbc*L gene (coding for the large subunit of Rubisco) specifically targets phototrophs, and is more differentiated than the traditional 18S ribosomal marker gene, thus allowing us to specifically target snow algae.
 Our goal was to see what species were present, and how community composition varies over elevation.
@@
 We observed pink and green patches of snow algae in forested sites beginning on May 14.
-Snow algae was widespread in our study area, and we collected 309 samples from 33 dates and 13 moutains over the course of the summer (S1, S2).
+Snow algae was widespread in our study area, and we collected 309 samples from 33 dates and 13 mountains over the course of the summer (S1, S2).
 We observed an elevational shift in snow algae throughout the season: forested sites bloomed earlier in the season and melted out sooner, while high elevation sites did not bloom until later in the season, but persisted until the first snowfall of winter 2019 (S2).
 
-We observed some consistant patterns between cell morphologies and field site characteristics.
+We observed some consistent patterns between cell morphologies and field site characteristics.
 In forest sites, orange or green snow was typically dominated by green or yellow-orange ovate cells with short spines (*Chloromonas cf.
@@
 We found several distinct clusters among *Chloromonas* amplicon sequence variants (ASVs) not matching any known species on GenBank (I, G, Fig.
-2), but the majority of *Chloromonas* ASVs did not form distinct clusters (*Chloromonas spp* C, Fig.
+2), but the majority of *Chloromonas* ASVs did not form distinct clusters (Other *Chloromonas*, Fig.
 2).
 
-Our three relative abundance data sets were generally consistant.
+Our three relative abundance data sets were generally consistent.
 18S, *rbc*L, and cell count data were correlated for *Sanguina* (stats).
@@
 However, this does not explain why *Chloromonas krienitzii* was not found at higher elevations: our *C.
-krienitzii* dominated samples were taken from sites that were ~100 m from the nearest tree, and were not notabley shaded by topography in any way.
+krienitzii* dominated samples were taken from sites that were ~100 m from the nearest tree, and were not notably shaded by topography in any way.
 Anecdotally, we observed *C.
-krienitzi* developed over running water, so their distribution could be linked to the presence of streams below the snowpack.
+krienitzii* developed over running water, so their distribution could be linked to the presence of streams below the snowpack.
 
@@
 One notable difference was that no reads were assigned to *Chlainomonas* from our 18S library, despite this genera being highly abundant in *rbc*L.
-Samples contaiing primarily *Chlainomonas* in our *rbc*L and cell count data were marked as containing only *Chloromonas* in 18S.
-At the time of publication, only two 18S *Chlainomonas* were present on GenBank (MF803743.1, MF803745.1); possibly our *Chlainomonas* sequences are different enough from the reference that they were mis-labelled as *Chloromonas*, especially likely given how closely related the two genera are.
+Samples containing primarily *Chlainomonas* in our *rbc*L and cell count data were marked as containing only *Chloromonas* in 18S.
+At the time of publication, only two 18S *Chlainomonas* were present on GenBank (MF803743.1, MF803745.1); possibly our *Chlainomonas* sequences are different enough from the reference that they were mislabelled as *Chloromonas*, especially likely given how closely related the two genera are.
 Samples dominated by *Sanguina* in our 18S library contained lower abundance in *rbc*L, which is likely due to primer bias: because our custom *rbc*L primers were designed to target *Chloromonas* and *Chlainomonas* (*Sanguina* not being available at the time), this likely led to a over-representation of these taxa in our *rbc*L library.
@@
 *Sanguina* sequences were not included because they were not available at the time.
-We designed primers with PRIME(+) from the GCG package.
-These primers were rbcL369F (5’-GAA CGT GAC AAA TTA AAC AAA-3’) and rbcL870R (5’-ACC WGA YAD ACG WAG AGC TT-3’).
+We designed primers using the primer design tool from the Eurofins Genomics website (www.eurofinsgenomics.eu/en/ecom/tools/pcr-primer-design/).
+The resultant primers were rbcL369F (5’-GAA CGT GAC AAA TTA AAC AAA-3’) and rbcL870R (5’-ACC WGA YAD ACG WAG AGC TT-3’).
 To target 18S we used Euk1181 (5'-TTA ATT TGA CTC AAC RCG GG-3') and Euk1624 (5'-CGG GCG GTG TGT ACA AAG G-3') [@Wang2014].
@@
 
-Reads were demultiplexing using cutadapt [@Martin2011].
+Reads were demultiplexing using CUTADAPT [@Martin2011].
 Samples were filtered and trimmed, errors removed, dereplicated, pair end reads merged, and chimeras removed following the default pipeline of DADA2 [@Callahan2017].

### 2019-11-04 manuscript/supplemental.Rmd
added 3 figs

--- 2019-11-02 manuscript/supplemental.Rmd
@@
 ---
-title: "Supplemental Figures"
-output: word_document
+title: "Supplementary materials"
+output: pdf_document
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+```{r,, include=F}
+knitr::opts_chunk$set(echo = FALSE)
 ```
 
-```{r table}
-knitr::kable(mtcars[1:5,, 1:5], caption = "A table caption")
+```{r setup, include=FALSE}
+library(here)
+library(tidyverse)
+library(kableExtra)
 ```
+
+```{r, include=F}
+map <- knitr::include_graphics(here::here("manuscript/figs/map.png"))
+field <- read_csv(here::here("data/02_tidied/tidied_field.csv"))
+field_photos <- knitr::include_graphics(here::here("manuscript/figs/S3_site_field_photos/field_collage.png"))
+
+map_cap <- "S1. Map showing locations of snow algae sampling locations in summer 2018."
+table_cap <- "S2. Table of sample site information of sequenced snow algae samples."
+field_photo_cap <- "S3. Photos of snow algae field sites."
+```
+
+## Figures
+ 1. Map of sample locations
+ 2. Photos of field sites
+ 3. Microphotos of snow algae samples
+ 4. Plots of snow algae 18S community composition. A. All 18S, Silva assignments. B. Chlorophyta assignments with custom reference database.
+ 5. Maximum likelihood tree of *rbc*L amplicon
+ 6. Plot of base differences between Genbank 18S and *rbc*L segments
+ 7. Phylogenetic diversity of snow algae.
+
+## Tables
+ 1. Table of sample site information
+ 2. Software information
+
+```{r, fig.cap=map_cap}
+map
+```
+
+```{r, fig.cap=field_photo_cap}
+field_photos
+```
+
+```{r}
+field %>% 
+   filter(have_seq==T) %>% 
+   mutate(notes = if_else(sample_id %in% c("pan18.ro","bre18.rd"), "collected from rock",NA_character_)) %>% 
+   select(alias, date, elev_m, lat, lon,color_snow, habitat, notes) %>% 
+   arrange(date) %>% 
+   kable(caption = table_cap) %>% 
+   kable_styling(bootstrap_options = "striped")
+
+```

### 2019-11-05 code/community_comparison_18s.Rmd
ran 18s snow algae assignments locally

@@
 rel_abund <- read_csv(here("data/03_filtered/18s_algae/18s_algae.filtered_cleaned_rel_abund.csv"))
+algae_taxonomy <- read_csv(here("output/algae_18s_assignments_snow_algae_genbank.csv"))
+algae_select_taxonomy <- read_csv(here("output/algae_18s_assignments_w_select_snow_algae.csv"))
+
 tree_18s <- read.tree(here("output/tree/algae_18s/algae_18s_seqs.fasta.treefile"))
+field <- read_csv(here("data/02_tidied/tidied_field.csv"))
+```
 
-path_silva_assigns <- here("data/clean/clean_silva_18s_assignments.csv")
-silva <- read_csv(path_silva_assigns)
+```{r, include=FALSE}
+taxonomy <- algae_taxonomy
 
-path_field <- here("data/clean/field_clean.csv")
-field <- read_csv(path_field)
+algae_percent_rel_abund <- rel_abund %>% 
+  # calculate rel abund as % of Chlorophyta reads
+  group_by(sample_id) %>% 
+  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
+  ungroup()
+
+best_assignment_rel_abund <- algae_percent_rel_abund %>% 
+  # calculate relative abundance per genus
+  left_join(taxonomy, by="asv_id") %>% 
+  group_by(sample_id, best_assignment) %>% 
+  summarise(rel_abund = sum(rel_abund)) %>% 
+  ungroup() %>% 
+  left_join(field, by="sample_id")
+
+barplot <- ggplot(best_assignment_rel_abund, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
+  geom_bar(aes(fill=best_assignment), stat="identity") + # classic stacked relative abundance plot
+  coord_flip() +
+  labs(title="Assignment with unedited genbank downloads",
+       fill = "18S taxonomic assigmnent",
+       y="Relative abundance out of Chlorophyta reads",
+       x="Sample ID (in order of increasing elevation)")
+barplot
+
+# desired categories: lump C nivalis in Chloromonas, lump
+
 ```
+
+```{r}
+taxonomy <- algae_select_taxonomy
+
+best_assignment_rel_abund <- algae_percent_rel_abund %>% 
+  # calculate relative abundance per genus
+  left_join(taxonomy, by="asv_id") %>% 
+  group_by(sample_id, best_assignment) %>% 
+  summarise(rel_abund = sum(rel_abund)) %>% 
+  ungroup() %>% 
+  left_join(field, by="sample_id")
+
+barplot_select <- ggplot(best_assignment_rel_abund, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
+  geom_bar(aes(fill=best_assignment), stat="identity") + # classic stacked relative abundance plot
+  coord_flip() +
+  labs(title="Assignment with select database",
+       fill = "18S taxonomic assigmnent",
+       y="Relative abundance out of Chlorophyta reads",
+       x="Sample ID (in order of increasing elevation)")
+barplot_select
+
+```
+
+Try with consensus sequence for each known snow algae species.
+
+Takehome: 
+  - X1.1 has high Trebouxiophyceae of some sort, Koliellaceae
+  - W1.1 Chlorogonium (Haematococcaceae), 
+  - similar patterns otherwise as rbcL, except no Chlainomonas
+
+```{r}
+algae_percent_rel_abund %>% 
+  filter(sample_id == "sax18.01") %>% 
+  arrange(-rel_abund)
+```
+
+top two ASVs in sax18.01 blast as koliellaceae and chloromonas
 
@@
 ```
+
+```{r}
+ggsave(here("manuscript/figs/18s_algae_rel_abund.png"), plot = barplot)
+```

### 2019-11-05 code/assign_18s_algae_to_custom.R
ran 18s snow algae assignments locally

@@ -1,0 +1,73 @@
+# assign 18s algae to custom database
+
+library(tidyverse)
+library(DECIPHER)
+library(here)
+
+algae_18s <- readDNAStringSet(here("data/asv_seq_data/algae_18s.fasta")) 
+reference <- readDNAStringSet(here("reference_database/snow_algae_18s_taxonomy.fasta"))
+curated_ref <- readDNAStringSet(here("reference_database/select_snow_algae_refs.fasta"))
+
+
+
+# train classifier and assign taxonomy
+trained_reference <- curated_ref %>% 
+  LearnTaxa(names(curated_ref))
+
+assignments <- IdTaxa(algae_18s, trained_reference, threshold = 50)
+
+
+# convert to tbl
+taxonomy_tbl <- tibble(list = assignments) %>% 
+  add_column(asv_id = names(assignments)) %>% 
+  hoist(list,
+        taxon = "taxon", 
+        conf = "confidence") %>% 
+  hoist(taxon,
+        root = 1,
+        domain = 2,
+        kingdom = 3,
+        phylum = 4,
+        class=5,
+        order=6,
+        family=7,
+        genus=8,
+        species=9) %>% 
+  hoist(conf,
+        root_conf = 1,
+        domain_conf = 2,
+        kingdom_conf = 3,
+        phylum_conf = 4,
+        class_conf=5,
+        order_conf=6,
+        family_conf=7,
+        genus_conf=8,
+        species_conf=9)
+
+taxonomy_select <- taxonomy_tbl %>% 
+  select(asv_id, phylum, class, order, family, genus, species, phylum_conf, class_conf, order_conf, family_conf, genus_conf, species_conf)
+
+label_na <- function(x){ 
+  if_else(str_detect(x,"unclass"), NA_character_, x)
+  }
+taxonomy_na_labeled <- taxonomy_select %>% 
+  mutate_at(2:7, label_na)
+
+best_assignment <- taxonomy_na_labeled %>% 
+  # make new col detecting assignment level
+  mutate(id_level = case_when(is.na(class)~"phylum",
+                              is.na(order)~"class",
+                              is.na(family)~"order",
+                              is.na(genus)~"family",
+                              is.na(species)~"genus",
+                              TRUE~"species")) %>% 
+  # make new col with best assignment
+  mutate(best_assignment = case_when(id_level == "class"~class,
+                                     id_level == "order"~order,
+                                     id_level == "family"~family,
+                                     id_level == "genus"~genus,
[…]

### 2019-11-05 reference_database/FlatToFastaIdTaxa.py
assign 18s algae to genbank custom database

@@ -1,0 +1,102 @@
+#!/usr/bin/env python
+
+import argparse
+import os
+
+
+def is_valid_file(file, parser):
+    """
+    Checks to determine if the specified file exists
+
+    :param file: A string containing a filepath
+    :param parser: An argparse.ArgumentParser object
+    :return: file, if the file exists
+    :raises: parser.error, if the file does not exists
+    """
+
+    if os.path.exists(file):
+        return file
+    else:
+        raise parser.error("Unable to locate %s: No such file or directory" % file)
+
+
+def get_args():
+    """
+    Processes command line arguments
+
+    :return: A namespace containing command line arguments
+    """
+    parser = argparse.ArgumentParser(description="Converts one or more NCBI FLAT files into a single FASTA file")
+    parser.add_argument("-i", "--input", metavar="FLAT", required=True, nargs="+",
+                        type=lambda x: is_valid_file(x, parser), help="Input FLAT file(s)")
+    parser.add_argument("-o", "--output", metavar="FASTA", required=True, help="Output merged FASTA file")
+    return parser.parse_args()
+
+
+def main(args=None):
+    if args is None:
+        args = get_args()
+
+    # Process each FLAT file
+    with open(args.output, "w") as o:
+        for file in args.input:
+            with open(file) as f:
+
+                sequence = ""
+                taxonomy = ""
+                organism = ""
+                accession = ""
+
+                while True:
+                    line = f.readline()
+                    if line == "":  # There are no more lines in this file
+                        break
+                    line = line.strip("\n").strip("\n")  # Remove newline
+
+                    # Parse the required attributes from the FLAT file
+                    if line.startswith("ORIGIN"):  # Sequence
+
+                        while True:
+                            line = f.readline()
+                            line = line.strip("\r").strip("\n")  # Remove newline
+                            if line == "//":  # The // is a control character to end the FLAT entry
+                                # Format all the file features into a FASTA entry
+                                # The format is >taxonomy; Organism; Accession
+                                fasta_name = ">Root;" + taxonomy + organism.replace(" ", "_") + ";" + accession
+                                o.write(fasta_name)
+                                o.write(os.linesep)
+
+                                # Sequence
+                                o.write(sequence)
+                                o.write(os.linesep)
+
+                                # Reset variables
+                                sequence = ""
+                                taxonomy = ""
+                                organism = ""
+                                accession = ""
+                                break
+
+                            # Remove all spaces and numbers
+                            line = line.replace(" ", "")
[…]

### 2019-11-05 code/rbcl_asv_clustering.Rmd
minor changes, re-named var

@@
   select(asv_id, otu_name)
+
+tsne_added_names <- tsne_edited_otus %>% 
+  left_join(otus_asv_name, by="asv_id")
 ```
@@
 ```{r echo=F, fig.width=8, fig.height=6}
-tsne_edited_otus %>% 
-  select(plot_taxa, otu) %>% 
-  distinct() %>% 
-  arrange(otu)
-
-# make a key for downstream use (write to file at end of script)
-asv_otu_key <- tsne_otu_names %>% 
-  select(asv_id, otu, otu_name)
-
-ggplot(tsne_otu_names, aes(x=V1, y=V2)) +
+ggplot(tsne_added_names, aes(x=V1, y=V2)) +
   ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), alpha=0.3, show.legend = FALSE) +
   # plot points not assigned to genus or species
-  geom_point(data = filter(tsne_otu_names, is.na(plot_taxa)), 
+  geom_point(data = filter(tsne_added_names, is.na(plot_taxa)), 
              color = "grey",alpha = 0.8, size = 3) +
   # color points that were assigned to genus or species
-  geom_point(data = filter(tsne_otu_names, !is.na(plot_taxa)), 
+  geom_point(data = filter(tsne_added_names, !is.na(plot_taxa)), 
              aes(color = plot_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
@@
 
-tsne_rel_abund <- tsne_otu_names %>% 
+tsne_rel_abund <- tsne_added_names %>% 
   left_join(total_rel_abund, by="asv_id")
@@
   left_join(taxonomy, by="asv_id") %>% 
-  left_join(asv_otu_key, by="asv_id") %>% 
+  left_join(otus_asv_name, by="asv_id") %>% 
   left_join(total_rel_abund, by="asv_id") %>% 
@@
 Run IQtree
-```{r tree, fig.width=8, fig.height=20}
+```{r tree, fig.width=7, fig.height=10}
 # root tree with Trebouxiophyceae
@@
 # clades annotated by OTU
-otu_tree <- rbcl_ggtree %<+% asv_otu_key + 
+otu_tree <- rbcl_ggtree %<+% otus_asv_name + 
   geom_tippoint(aes(color=otu_name, alpha=0.7), size=3) + # 
@@
 
+taxonomy_tree
 otu_tree
@@
 ggsave(filename = here("manuscript/figs/2_taxonomy/mds.png"), plot=mds,  height = 6, width = 8)
-ggsave(filename = here("manuscript/figs/tree.png"), plot=otu_tree, width=8, height = 20)
+
+ggsave(filename = here("manuscript/figs/otu_tree.png"), plot=otu_tree, width=7, height = 10)
+ggsave(filename = here("manuscript/figs/taxonomy_tree.png"), plot=taxonomy_tree, width=7, height = 10) 
 ```

### 2019-11-05 code/rbcl_community_compare.Rmd
minor changes, re-made diversity plots as 2 panel

@@
   geom_text_repel(aes(label=alias))+
-  labs(title="Phylogenetic diversity of snow algae rbcL vs elevation",y="Faith's phylogenetic diversity (PD)",x="Elevation (m)")
+  labs(y="Faith's phylogenetic diversity (PD)",x="Elevation (m)", tag="B")
 faith_elev_plot
@@
 # plot diversity of samples
-ggplot(otu_diversity, aes(x=elev_m, y=shannon))+
+shannon <- ggplot(otu_diversity, aes(x=elev_m, y=shannon))+
   geom_point() +
   geom_text_repel(aes(label=alias)) +
-  geom_smooth(method = "lm")
+  geom_smooth(method = "lm")+
+  labs(y="Shannon diversity", tag="A")+
+  theme(axis.title.x = element_blank())
+shannon
+```
+
+```{r diversity_plots, include=F}
+diversity_plots <- gridExtra::grid.arrange(shannon, faith_elev_plot, nrow=2)
+
 ```
@@
 ggsave(filename = here("manuscript/figs/3_community/combined_fig_3.png"), plot=combined_fig_3, height=10, width=7)
-ggsave(filename = here("manuscript/figs/supplementary/diversity.png"), plot = faith_date_plot, height = 5, width = 7)
+ggsave(filename = here("manuscript/figs/diversity.png"), plot = diversity_plots, height = 10, width = 7)
 ```

### 2019-11-05 code/alignment/rbcl_alignment.R
deleted files

@@ -1,43 +1,0 @@
-# prep rbcL fasta for IQTree
-
-library(tidyverse)
-library(here)
-library(DECIPHER)
-
-
-rbcl <- readDNAStringSet(here("data/asv_seq_data/rbcl.fasta"))
-asv_ids_main_analysis <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv")) %>% distinct(asv_id) %>% pull()
-
-# all alignments ------------
-aligns <- rbcl %>% 
-  AlignSeqs(verbose = F)
-
-BrowseSeqs(aligns, highlight = T)
-
-
-# remove gaps alignments ------------------
-gappy_seqs <- c("534", "639", "646")
-
-length(aligns)
-aligns_no_gaps <- aligns[!(names(aligns) %in% gappy_seqs)]
-length(aligns)
-
-BrowseSeqs(aligns_no_gaps, highlight = T)
-
-# just alignments in "main" analysis --------------
-length(rbcl)
-main_aligns <- rbcl[ names(rbcl) %in% asv_ids_main_analysis ] %>% 
-  AlignSeqs(verbose = F)
-length(main_aligns)
-
-
-
-
-aligns %>% 
-  writeXStringSet(filepath = here("output/tree/rbcl/rbcl_aligns_all.fasta"))
-
-aligns_no_gaps %>% 
-  writeXStringSet(filepath = here("output/tree/rbcl/rbcl_aligns_indels_removed.fasta"))
-
-main_aligns %>% 
-  writeXStringSet(filepath = here("output/alignments/rbcl_aligns_in_main.fasta"))

### 2019-11-05 code/misc/compare_18s_rbcl_snow_algae_diversity.R
init, carried over from old project

@@ -1,0 +1,232 @@
+# n differences between rbcl species on genbank
+
+# align snow algae rbcl, align, trim to region between our primers, calculate distance matrix (in num bp diff) for each
+library(tidyverse)
+library(DECIPHER)
+library(broom)
+library(here)
+library(magrittr)
+
+
+
+gb <- readDNAStringSet(here("reference_database/genbank_snow_algae_fastas/snow_algae_rbcl.fasta")) # max 2000 bp len
+s18 <- readDNAStringSet(here("reference_database/genbank_snow_algae_fastas/chloromonas18s.fasta"))
+my_asvs <- readDNAStringSet(here("data/asv_seq_data/rbcl_main.fasta"))
+my_18s_asvs <- readDNAStringSet(here("data/asv_seq_data/algae_18s_seqs.fasta"))
+
+
+
+
+
+
+length(gb)
+length(s18)
+
+# shorten names
+short_names <- names(gb) %>%
+  str_split(" ") %>% 
+  purrr::map(extract, 1:3) %>% # select first name elements
+  purrr::map(paste, collapse = "_")  %>% # seperate by underscores
+  unlist()
+
+# remove N gaps
+gb <- gb %>%
+  str_replace_all("N", '') %>%
+  DNAStringSet()
+
+# assign short names
+names(gb) <- short_names # add names AFTER removng gaps (converting XStringSet to vector will remove the names)
+
+# align asvs
+aligns <- gb %>% 
+  append(my_asvs[1:50]) %>% # add my seqs to see where to trim
+  AlignSeqs()
+aligns %>% BrowseSeqs(highlight = T) # check
+
+# trim to region between primers
+names <- names(aligns)
+aligns <- aligns %>% 
+  str_sub(484, 896) %>% 
+  DNAStringSet()
+names(aligns) <- names
+aligns %>% BrowseSeqs(highlight = T) # check
+
+# remove low coverage genbank refs, and remove my primers
+aligns <- aligns[1:80]
+aligns <- aligns[-str_which(names(aligns), "AB434274.1|AB434275.1")]
+
+# filter aligns to one of each species
+unique_sp_indices <- names(aligns) %>% 
+  str_split("_") %>% 
+  purrr::map(magrittr::extract, 3) %>%
+  unlist() %>% 
+  enframe() %>% 
+  group_by(value) %>% 
+  sample_n(1) %>% #grab one of each species
+  ungroup() %>% 
[…]

### 2019-11-05 code/wrangle/make_fastas_and_alignments.R
init, new fasta generation script

@@ -1,0 +1,85 @@
+# make fastas of sequences, aligned and unaligned
+
+library(tidyverse)
+library(here)
+library(DECIPHER)
+
+s18_asvs <- read_csv(here("data/asv_seq_data/asv_seq_key_18s.csv"))
+rbcl_asvs <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
+
+algae_18s_taxonomy <- read_csv(here("data/02_tidied/tidied_silva_18s_assignments.csv"))
+rbcl_main_rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv")) 
+
+
+  
+# convert to DNAStringSet
+convert_csv_to_dss <- function(csv){
+  csv %>% 
+    deframe() %>% 
+    DNAStringSet()
+}
+
+s18_dss <- s18_asvs %>% convert_csv_to_dss()
+rbcl_dss <- rbcl_asvs %>% convert_csv_to_dss()
+rbcl_main_dss <- 
+
+# also get a fasta of just the algae 18S reads
+algae_18s_asv_ids <- algae_18s_taxonomy %>% 
+  filter(phylum == "Chlorophyta") %>% 
+  pull(asv_id)
+
+s18_algae_dss <- s18_asvs %>% 
+  filter(asv_id %in% algae_18s_asv_ids) %>% 
+  convert_csv_to_dss()
+
+# ... and a fasta of the rbcl reads, just in the "main" analysis
+rbcl_main_asvs <- rbcl_main_rel_abund %>% 
+  distinct(asv_id) %>% 
+  pull()
+
+nrow(rbcl_asvs)
+rbcl_main_dss <- rbcl_asvs %>% 
+  filter(asv_id %in% rbcl_main_asvs) %>% 
+  convert_csv_to_dss()
+length(rbcl_main_dss)
+
+# align (~30 seconds per alignment)
+s18_aligns <- s18_dss %>% AlignSeqs(verbose=F)
+rbcl_aligns <- rbcl_dss %>% AlignSeqs(verbose=F)
+s18_algae_aligns <- s18_algae_dss %>% AlignSeqs(verbose = F)
+rbcl_main_aligns <- rbcl_main_dss %>% AlignSeqs(verbose = F)
+
+BrowseSeqs(s18_aligns, highlight = T)
+BrowseSeqs(rbcl_aligns, highlight = T)
+BrowseSeqs(s18_algae_aligns, highlight = T)
[…]

### 2019-11-05 code/wrangle/tidy_18s_algae_taxonomy.R
delete

@@ -1,45 +1,0 @@
-# tidy 18s taxonomy-- just snow algae
-library(tidyverse)
-library(here)
-
-taxonomy_tbl <- read_csv(here("data/01_raw/taxonomy_18s_genbank_snow_algae.csv"))
-asv_seq_data <- read_csv(here("data/asv_seq_data/asv_seq_key_18s.csv"))
-
-
-
-taxonomy_tbl %>% distinct(V2, V3, V4, V5, V6, V7, V8, V9) %>% view()
-
-
-  
-# rename cols in taxonomy tbl
-cleaned <- taxonomy_tbl %>% 
-  select(-V2, -V3, -V4, -V21, -V31, -V41) %>% # only interested in assignments at class level or below (Chlorophyceae vs Trebouxiophyceae)
-  dplyr::rename(asv_sequence = asv,
-                class = V5,
-                order = V6,
-                family = V7, 
-                genus = V8, 
-                species = V9,
-                class_conf = V51,
-                order_conf = V61,
-                family_conf = V71, 
-                genus_conf = V81, 
-                species_conf = V91) %>% 
-  mutate_at(c("class","order","family","genus","species"),
-            ~str_replace(., pattern="unclassified", replacement = NA_character_)) %>% 
-  filter(!(is.na(class)))
-
-
-cleaned %>% view()
-
-# replace sequence with asv_id
-cleaned %>% 
-  left_join(asv_seq_data, by="asv_sequence") %>% view()
-  select(asv_id, class:species_conf)
-
-  
-  
-### problem: the sequences in silva (that we used to generate the key) do not match any of the sequences in the 18s
-  
-  
-  

### 2019-11-05 code/wrangle/old/tidy_18s_algae_taxonomy.R
init

@@ -1,0 +1,55 @@
+# tidy 18s taxonomy-- just snow algae
+library(tidyverse)
+library(here)
+
+taxonomy_tbl <- read_csv(here("data/01_raw/taxonomy_18s_genbank_snow_algae.csv"))
+asv_seq_data <- read_csv(here("data/asv_seq_data/asv_seq_key_18s.csv"))
+
+
+
+taxonomy_tbl %>% distinct(V2, V3, V4, V5, V6, V7, V8, V9) %>% view()
+
+
+  
+# rename cols in taxonomy tbl
+cleaned <- taxonomy_tbl %>% 
+  select(-V2, -V3, -V4, -V21, -V31, -V41) %>% # only interested in assignments at class level or below (Chlorophyceae vs Trebouxiophyceae)
+  dplyr::rename(asv_sequence = asv,
+                class = V5,
+                order = V6,
+                family = V7, 
+                genus = V8, 
+                species = V9,
+                class_conf = V51,
+                order_conf = V61,
+                family_conf = V71, 
+                genus_conf = V81, 
+                species_conf = V91) %>% 
+  mutate_at(c("class","order","family","genus","species"),
+            ~str_replace(., pattern="unclassified", replacement = NA_character_)) %>% 
+  filter(!(is.na(class)))
+
+
+cleaned %>% view()
+
+
+### testing
+tax <- cleaned %>% select(asv_sequence) %>% pull()
+key <- asv_seq_data %>% select(asv_sequence) %>% pull()
+
+setdiff(tax, key)
+setdiff(key, tax)
+
+# they're 100% different
+
+
+
+
+
+# replace sequence with asv_id
+cleaned %>% 
+  left_join(asv_seq_data, by="asv_sequence") %>% view()
+  select(asv_id, class:species_conf)
+
+  
+  

### 2019-11-05 code/cellcount.Rmd
added elevation side panel

@@
 library(here)
+library(gridExtra)
 ```
@@
 
+distinct(cellct, sample_id) %>% nrow()
+
 short_names <- p %>% 
@@
 
-my_theme <- theme(axis.title.y =element_blank(),
-                  axis.text.y=element_blank(),
+my_theme <- theme(axis.text.y=element_blank(),
                   axis.ticks.y=element_blank(), 
@@
   geom_tile(aes(alpha = percent_of_sample, fill = morpho_sp)) + # color = "white" %>% fct_reorder(morpho_order)
-  labs(x="Cell morphology", y= "Sample ID (in order of elevation)", fill = "Cell morphology", alpha="Relative abundance") +
+  labs(x="Cell morphology", y= "Elevation", fill = "Cell morphology", alpha="Relative abundance") +
   scale_alpha(range = c(0.1, 1)) +
-  guides(alpha=FALSE)+
   my_theme
@@
 
-```{r, elevation_bar, echo=F}
+```{r elevation_bar, echo=F, fig.height=5, fig.width=1}
 elevation_ranges <- short_names %>% 
@@
 
-ggplot(elevation_ranges, aes(x=sample_id %>% fct_reorder(elev_m), y=elev_m))+
-  geom_bar(stat = "identity")+
-  coord_flip()
+my_theme <- theme(axis.title.x =element_blank(),
+                  axis.text.x=element_blank(),
+                  axis.ticks.x=element_blank(),
+                  axis.title.y =element_blank(),
+                  axis.text.y=element_blank(),
+                  axis.ticks.y=element_blank(),
+                  panel.grid.major = element_blank(),
+                  panel.grid.minor = element_blank(),
+                  panel.background = element_blank(),
+                  legend.position = "none")
+
+elev <- ggplot(elevation_ranges, aes(x=sample_id %>% fct_reorder(elev_m), y=2200-elev_m, fill=elev_m))+
+  geom_bar(width=1, stat = "identity")+
+  scale_fill_gradientn(colors = terrain.colors(8)[1:7]) +
+  coord_flip() +
+  my_theme +
+  # labs(y="Elevation (m)")
+  annotate(geom="text", x=3, y=500, label="800 m",
+              color="black") +
+  annotate(geom="text", x=115, y=500, label="2200 m",
+              color="black")
+
+elev
 ```
@@
 ```{r}
-ggsave(filename=here("manuscript/figs/1_field/cellct_heatmap.pdf"), plot=cellct_heatmap, height = 7, width = 2)
+dir.create(here("manuscript/figs/supp/cellct"))
+ggsave(filename=here("manuscript/figs/supp/cellct/cellct_heatmap.pdf"), plot=cellct_heatmap, height = 7, width = 3)
+ggsave(here("manuscript/figs/supp/cellct/elevations.pdf"), plot=elev, height=7, width=1)
 ```

### 2019-11-05 manuscript/supplemental.Rmd
supplemental figs ready for lynne v1

@@
 library(kableExtra)
+library(knitr)
+library(sessioninfo)
 ```
@@
 field_photos <- knitr::include_graphics(here::here("manuscript/figs/S3_site_field_photos/field_collage.png"))
+micro_photos <- knitr::include_graphics(here::here("manuscript/figs/S4_site_micro_photos/micro.png"))
+s18 <- include_graphics(here("manuscript/figs/18s_algae_rel_abund.png"))
+tree <- include_graphics(here("manuscript/figs/otu_tree.png"))
+tree_taxonomy <- include_graphics(here("manuscript/figs/taxonomy_tree.png"))
+diversity <- include_graphics(here("manuscript/figs/diversity.png"))
+cellct <- include_graphics(here("manuscript/figs/supp/cellct/cellct.png"))
 
@@
 field_photo_cap <- "S3. Photos of snow algae field sites."
+micro_cap <- "Microphotos of snow algae samples included in sequencing run."
+s18_cap <- "Relative abundance barplot of 18S amplicon library"
+tree_cap <- "Maximum likelihood tree of rbcL amplicon library. Colors represent OTU assignments."
+tax_tree_cap <- "Maximum likelihood tree of rbcL amplicon library. Colors represent taxonomic assignments."
+diversity_cap <- "A. Shannon diversity of snow algae samples, plotted against sample elevation. Smooth method = linear model. B. Faith's phylogenetic diversity plotted against sample elevation."
+(ref:cellct_cap) "Heatmap showing cell count relative abundance from 117 snow algae samples.
+Cb = Chlainomonas cf.
+brevispina, Ck=Chloromonas krienitzii, Cn=Chloromonas cf.
+nivalis, Cr=Chlainomonas rubra, G=green cells, O=other, Sn=Sanguina nivaloides"
 ```
@@
 ```{r, fig.cap=map_cap}
+# zoom in
 map
@@
 ```{r, fig.cap=field_photo_cap}
+# order by date
+# sizes the same (crop template)
+# arrange in grid
 field_photos
+```
+
+```{r, fig.cap=micro_cap}
+#scale bar
+# # sizes the same (crop template)
+# arrange in grid
+micro_photos
+```
+
+```{r s18_rel_abund, fig.cap=s18_cap}
+s18 # should this go in the main fig next to the rbcl assignments?
+```
+
+```{r tree, fig.cap=tree_cap}
+tree
+```
+
+```{r tax_tree, fig.cap=tax_tree_cap}
+tree_taxonomy
+```
+
+```{r, fig.cap=diversity_cap}
+diversity
+```
+
+```{r, (ref:cellct_cap)}
+cellct
 ```
@@
    kable_styling(bootstrap_options = "striped")
+```
 
+```{r}
+my_session_info <- sessioninfo::session_info()
+
+my_session_info$platform %>% 
+  unlist() %>% 
+  enframe() %>% 
+  kable(caption = "Session information") %>% 
+  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F, position = "left")
 ```
+
+```{r}
+software <- my_session_info$packages %>% 
+  as_tibble() %>% 
+  select(package, loadedversion) %>% 
+  dplyr::rename(version = loadedversion)
+
+# cedar_info <- tibble(package = c(""),
+#                      version = c(""))
+
+# software <- software %>% 
+#   bind_rows(cedar_info)
+
+software %>% 
+  kable(caption=" R software package versions used") %>% 
+  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F, position = "left")
+```

### 2019-11-05 code/cellcount.Rmd
started to add cellct plot just for samples we sequenced

@@
 ```{r}
+cellct_in_seq <- short_names %>% 
+  filter(have_seq==T)
+
+cust_theme <- theme(legend.position = "none",
+                    panel.grid.major.x = element_blank(), # remove vertical lines in background
+                    panel.grid.minor.x = element_blank(),
+                    axis.line=element_blank(), # remove x axis
+                    axis.title.x=element_blank(), 
+                    axis.text.x=element_blank(),
+                    axis.ticks.x=element_blank(),
+                    panel.spacing = unit(-0.01, "lines"), # remove space between facet panels
+                    # strip.text.x = element_text(angle = 0, hjust=1.0, vjust = 0.8), # rotate facet labels 90
+                    strip.background = element_blank()) # remove grey boxes around facet labs
+
+cellct_order <- c("Cr","Sn","Cb","Ck","Cn","G","O")
+
+cellct_in_seq_bar <- ggplot(cellct_in_seq, aes(x=alias %>% fct_reorder(elev_m), y=percent_of_sample)) +
+  geom_bar(aes(fill=short_morpho_name), stat="identity") + # classic stacked relative abundance plot
+  coord_flip() +
+  facet_grid(~short_morpho_name, switch = "x") +
+  labs( x="Sample ID", tag="B") +
+  cust_theme
+
+cellct_in_seq_bar
+```
+
+```{r}
 dir.create(here("manuscript/figs/supp/cellct"))

### 2019-11-05 code/rbcl_asv_clustering.Rmd
aesthetics

@@
 
-```{r echo=F, fig.width=8, fig.height=6}
+```{r echo=F, fig.width=10, fig.height=10}
 ggplot(tsne_added_names, aes(x=V1, y=V2)) +
@@
 
-```{r include_rel_abund, echo=F}
+```{r include_rel_abund, echo=F, fig.width=5, fig.height=4}
 total_rel_abund <- rel_abund %>% 
@@
 tsne_rel_abund <- tsne_added_names %>% 
-  left_join(total_rel_abund, by="asv_id")
+  left_join(total_rel_abund, by="asv_id") %>% 
+  # remove C kri ellipse
+  mutate(otu_name = if_else(otu_name == "Chloromonas krienitzii", "Other Chloromonas", otu_name))
 
@@
                              show.legend = FALSE, 
-                             alpha=0.2, 
-                             linetype="blank", 
+                             alpha=0, 
+                             linetype="dashed", 
+                             label.margin = margin(1,1,1,1,"mm"),
+                             label.buffer = unit(1, "mm"),
+                             label.fontface = "plain",
+                             label.fontsize = 8,
+                             con.type = 0,
                              expand = unit(2, "mm")) +
-  # plot points not assigned to genus or species
-  geom_point(data = filter(tsne_rel_abund, is.na(plot_taxa)), 
-             aes(size = global_rel_abund),
-             color = "grey",alpha = 0.5) +
-  # color points that were assigned to genus or species
-  geom_point(data = filter(tsne_rel_abund, !is.na(plot_taxa)), 
-             aes(color = plot_taxa %>% fct_relevel(taxa.order),
-                 size = global_rel_abund), 
-             alpha = 0.5) +
-  labs(color="ASV taxonomic assignment", tag="A") +
+  geom_point(aes(color = plot_taxa %>% fct_relevel(taxa.order), size = global_rel_abund, alpha=0.5))+
+  labs(color="ASV taxonomic assignment", tag="A", title="t-SNE") +
   ggthemes::theme_few() +
-  scale_size(range = c(1, 6), guide="none") + ################################################# not changing the scale size
+  scale_size(range = c(2, 12), guide="none")
+
+tsne_no_legend <- final_tsne +
     theme(legend.position = "none")
-final_tsne
+tsne_no_legend
 ```
 
-```{r asv_mds, echo=F}
+```{r, fig.height=4, fig.width=2.2}
+p <- final_tsne +
+  guides(colour = guide_legend(override.aes = list(alpha = 1, size=3)), alpha="none") 
+legend <- p %>% 
+  get_legend() %>% 
+  as_ggplot()
+legend
+```
+
+```{r asv_mds, echo=F, fig.width=5, fig.height=4}
 mds_coords <- dists %>% 
@@
                              show.legend = FALSE, 
-                             alpha=0.2, 
-                             linetype="blank",
+                             alpha=0, 
+                             linetype="dashed", 
+                             label.margin = margin(1,1,1,1,"mm"),
+                             label.buffer = unit(1, "mm"),
+                             label.fontface = "plain",
+                             label.fontsize = 8,
+                             con.type = 0,
                              expand = unit(2, "mm")) +
-  geom_point(aes(alpha = 0.7, color = plot_taxa, size = global_rel_abund)) +
-  scale_size(range = c(1, 6), guide="none") +
-  scale_alpha(guide="none")+
-  theme_few() +
-  labs(color="ASV taxonomic assignment", tag="B")
+  geom_point(aes(color = plot_taxa %>% fct_relevel(taxa.order), size = global_rel_abund, alpha=0.5))+
+  labs(color="ASV taxonomic assignment", tag="B",title="MDS") +
+  ggthemes::theme_few() +
+  scale_size(range = c(2, 12), guide="none")+
+  guides(colour = guide_legend(override.aes = list(alpha = 1, size=3)), alpha="none") +
+  theme(legend.position = "none")
+
+  # geom_point(aes(alpha = 0.7, color = plot_taxa, size = global_rel_abund)) +
+  # scale_size(range = c(1, 6), guide="none") +
+  # scale_alpha(guide="none")+
+  # theme_few() +
+  # labs(color="ASV taxonomic assignment", tag="B")
   # geom_text(aes(label = asv_id)) # too see outliers, asv_id=c("639", "534")
 mds
+```
+
+```{r, fig.width=4, fig.height=6}
+asv_arrange <- grid.arrange(tsne_no_legend, mds, nrow=2)
 ```
@@
 
-ggsave(filename = here("manuscript/figs/2_taxonomy/tsne.png"), plot=final_tsne, height = 6, width = 8)
-ggsave(filename = here("manuscript/figs/2_taxonomy/mds.png"), plot=mds,  height = 6, width = 8)
+ggsave(filename = here("manuscript/figs/2_taxonomy/asvs.pdf"), plot=asv_arrange, height = 6, width = 4)
+ggsave(filename = here("manuscript/figs/2_taxonomy/asv_legend.pdf"), plot=legend,  height = 4, width = 2.2)
 

### 2019-11-05 manuscript/figures.Rmd
minor changes, new paths

@@
 title: "Figures"
-output: pdf_document
+output: bookdown::pdf_book
 ---
@@
 library(here)
-opts_chunk$set(fig.width=12, fig.height=8, fig.path='figs/', fig.align= "center", echo=FALSE, warning=FALSE, message=FALSE)
+opts_chunk$set(fig.path='figs/', fig.align= "center", echo=FALSE, warning=FALSE, message=FALSE)
 ```
@@
 ```{r input, include=F}
-field <- include_graphics(here("manuscript/figs/1_field/field.png"))
-asvs <- include_graphics(here("manuscript/figs/2_taxonomy/combined.png"))
+field <- include_graphics(here("manuscript/figs/1_field/field2.png"))
+asvs <- include_graphics(here("manuscript/figs/2_taxonomy/asvs_polished.png"))
 communities <- include_graphics(here("manuscript/figs/3_community/combined_fig_3.png"))
@@
 D.
-Red snow containing *Sanguina nivaloides* and *Chlainomonas rubra* morphologies.
+Red snow containing \\emph{Sanguina nivaloides} (Sn) and //emph{Chlainomonas rubra} (Cr) morphologies.
 All scale bars 30 um.
 E.
-Orange-green snow containing *Chloromonas krienitzii* and *Chloromonas cf.
-nivalis*.
+Orange-green snow containing \\emph{Chloromonas krienitzii} (Ck) and \\emph{Chloromonas cf.
+nivalis} (Cn).
 F.
-Snow algae from green snow, including *Chloromonas cf.
-brevispina* and *Chlainomonas rubra*.
-G.
-Heatmap of cell count data, with samples on the y axis arranged from low to high elevation, and cell type on the x axis."
+Snow algae from green snow, including \\emph{Chloromonas cf.
+brevispina} (Cb) and \\emph{Chlainomonas rubra} (Cr)."
 

### 2019-11-05 manuscript/supplemental.Rmd
added cellct

@@
 micro_photos <- knitr::include_graphics(here::here("manuscript/figs/S4_site_micro_photos/micro.png"))
+cellct <- include_graphics(here("manuscript/figs/supp/cellct/cellct.png"))
 s18 <- include_graphics(here("manuscript/figs/18s_algae_rel_abund.png"))
@@
 diversity <- include_graphics(here("manuscript/figs/diversity.png"))
-cellct <- include_graphics(here("manuscript/figs/supp/cellct/cellct.png"))
 
@@
 micro_cap <- "Microphotos of snow algae samples included in sequencing run."
+cellct_cap <- "Heatmap showing cell count relative abundance from 117 snow algae samples.
+Cb = \\emph{Chlainomonas cf.
+brevispina}, Ck=\\emph{Chloromonas krienitzii}, Cn=\\emph{Chloromonas cf.
+nivalis}, Cr=\\emph{Chlainomonas rubra}, G=green cells, O=other, Sn=\\emph{Sanguina nivaloides}"
 s18_cap <- "Relative abundance barplot of 18S amplicon library"
@@
 diversity_cap <- "A. Shannon diversity of snow algae samples, plotted against sample elevation. Smooth method = linear model. B. Faith's phylogenetic diversity plotted against sample elevation."
-(ref:cellct_cap) "Heatmap showing cell count relative abundance from 117 snow algae samples.
-Cb = Chlainomonas cf.
-brevispina, Ck=Chloromonas krienitzii, Cn=Chloromonas cf.
-nivalis, Cr=Chlainomonas rubra, G=green cells, O=other, Sn=Sanguina nivaloides"
+
 ```

### 2019-11-15 code/rbcl_asv_clustering.Rmd
changed colors, make fasta for BR and JW

@@
 library(ggrepel)
+library(RColorBrewer)
+library(scales)
 
@@
 library(ggdark)
+
+conflicted::conflict_scout()
+conflicted::conflict_prefer("filter","dplyr")
 ```
@@
 ```{r read, include=F}
-
 rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
@@
 tree <- read.tree(file=here("output/tree/rbcl/asv613.fasta.treefile")) # this is the tree with only the ASVs that appear in the filtered rel abund dataset (not including protocol stuff, etc)
+
+# actual seq data
+key <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
 ```
@@
 
-ggplot(tsne_pts, aes(x=V1,y=V2))+
-  geom_point() +
-  labs(title="rbcL ASV clustering with t-SNE")
-```
-
-```{r genus_tsne, echo=F}
 tsne_pts_taxonomy <- tsne_pts %>% 
@@
 
-ggplot(tsne_pts_taxonomy, aes(x=V1,y=V2))+
-  geom_point(aes(color = genus)) +
-  labs(title = "Points labeled by genus")
-```
-
-```{r sp_tsne, echo=F}
-ggplot(tsne_pts_taxonomy, aes(x=V1,y=V2))+
-  geom_point(aes(color = species)) +
-  labs(title="Pts labeled by species")
-```
-
-Make it look nice, show both genus and species level assignments
-```{r tsne-plot1, echo=F}
-
-taxa.order <- c("Chloromonas","Chloromonas_muramotoi","Chloromonas_hohamii","Chloromonas_krienitzii","Chloromonas_Group_B","Raphidonema_nivale","Stichococcus","other Trebouxiophyceae","Chlainomonas", "Chlainomonas_rubra","Sanguina","Sanguina_aurantia","Sanguina_nivaloides" )
+taxa.order <- c("Chloromonas","Chloromonas_krienitzii","Chloromonas_muramotoi","Chloromonas_hohamii","Chloromonas_Group_B","other Trebouxiophyceae","Raphidonema_nivale","Stichococcus","Chlainomonas","Chlainomonas_rubra","Sanguina","Sanguina_nivaloides","Sanguina_aurantia")
 
@@
              aes(color = plot_taxa %>% fct_relevel(taxa.order)), 
-             alpha = 0.8) +
-  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment")
+             alpha = 0.8)
 ```
@@
 tsne_edited_otus <- tsne_taxa_otu %>%
-  mutate(otu = case_when((plot_taxa == "Chloromonas_krienitzii") ~ "9",
+  mutate(otu = case_when(plot_taxa == "Chloromonas_krienitzii" ~ "9",
                          TRUE ~ otu))
@@
                                  "Chloromonas F",
-                                 "Algae E",
+                                 "Unknown E",
                                  "Trebouxiophyceae G",
@@
   left_join(otus_asv_name, by="asv_id")
-```
-
-```{r echo=F, fig.width=10, fig.height=10}
-ggplot(tsne_added_names, aes(x=V1, y=V2)) +
-  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), alpha=0.3, show.legend = FALSE) +
-  # plot points not assigned to genus or species
-  geom_point(data = filter(tsne_added_names, is.na(plot_taxa)), 
-             color = "grey",alpha = 0.8, size = 3) +
-  # color points that were assigned to genus or species
-  geom_point(data = filter(tsne_added_names, !is.na(plot_taxa)), 
-             aes(color = plot_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
-  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
-  theme_few() 
 ```
@@
                              linetype="dashed", 
-                             label.margin = margin(1,1,1,1,"mm"),
+                             label.margin = margin(0,0,0,0,"mm"),
                              label.buffer = unit(1, "mm"),
@@
                              label.fontsize = 8,
-                             con.type = 0,
-                             expand = unit(2, "mm")) +
+                             con.type = "none",
+                             expand = unit(2, "mm")) + # expansion of ellipse around points
   geom_point(aes(color = plot_taxa %>% fct_relevel(taxa.order), size = global_rel_abund, alpha=0.5))+
-  labs(color="ASV taxonomic assignment", tag="A", title="t-SNE") +
+  labs(color="ASV taxonomic assignment", tag="B", title="t-SNE") +
   ggthemes::theme_few() +
@@
 
+```{r, fig.width=6,fig.height=8, include=F}
+taxa.order
+display.brewer.all()
+scales::show_col(brewer_pal(palette = "Set1")(5))
+```
+
+13 taxa, 9 OTUs
+Taxa colors: (* if representative of OTU)
+- Chloromonas (5 species, 5 OTU)
+  - unspecified*
+  - krienitzii*
+  - muramotoi*
+  - hohamii (1) --- D
+  - Group B (1) --- F
+- Chlainomonas (2 sp, 1 otu)
+  - unspecified*
+  - rubra
+- Sanguina (3 sp, 1 otu)
+  - unspecified*
+  - nivaloides
+  - aurentia
+- Trebouxiophyceae (3 sp, 1 otu)
+  - unspecified*
+  - Raphidonema
+  - Stichococcus
+
+4 diverging colors for the different genera, use shades of colors for species. The darkest color is "unspecified", use this color for the OTUs.
+
+Strategy: use green and orange for Chloromonas, Red for Sanguina, Purple for Chlainomonas, blye for trebouxiophyceae
+
+```{r, fig.width=2, fig.height=2}
+chloromonas_color <- "#4daf4a" # green
+krienitzii_color <- "#ff7f00" # orange
+muramotoi_color <- "#d7d42e" # yellow
+# pick shades using https://www.color-hex.com/
+hohamii_and_D_color <- "#265725" # dark green
+group_B_and_F_color <- "#b25800" # brown
+chloromonas_colors <- c(chloromonas_color, krienitzii_color, muramotoi_color, hohamii_and_D_color,group_B_and_F_color)
+scales::show_col(chloromonas_colors)
+```
+
+```{r,fig.width=2, fig.height=2}
+trebouxiophyceae_color <- "#377eb8" # blu
+raphidonema_color <- "#9bbedb" # light
+stichococcus_color <- "#1b3f5c" # dark
+trebouxiophyceae_colors <- c(trebouxiophyceae_color,stichococcus_color,raphidonema_color)
+scales::show_col(trebouxiophyceae_colors)
+```
+
+```{r,fig.width=2, fig.height=2}
+chlainomonas_color <- "#984ea3" # purp
+rubra_color <- "#c194c7" #lilac
+chlainomonas_colors <- c(chlainomonas_color,rubra_color)
+scales::show_col(chlainomonas_colors)
+```
+
+```{r,fig.width=2, fig.height=2}
+# changed from colorbrewer red = too agressive
+sanguina_color <- "#ff534a"
+nivaloides_color <- "#ff8680"
+aurentia_color <- "#b23a33"
+sanguina_colors <- c(sanguina_color,nivaloides_color,aurentia_color)
+scales::show_col(sanguina_colors)
+```
+
+```{r,fig.width=2, fig.height=2}
+na_color <- "grey70"
+
+taxa_colors <- c(chloromonas_colors,trebouxiophyceae_colors,chlainomonas_colors, sanguina_colors,na_color)
+names(taxa_colors) <- taxa.order
+scales::show_col(taxa_colors)
+```
+
+```{r,fig.width=2, fig.height=2}
+otu_colors <- c(chlainomonas_color,sanguina_color,chloromonas_colors[1:2],"#173416",chloromonas_colors[4:5], trebouxiophyceae_color,na_color)
+# manually change muramotoi color to dark green so it shows up on the heatmap
+names(otu_colors) <- c("Chlainomonas A","Sanguina B","Other Chloromonas","Chloromonas krienitzii","Chloromonas muramotoi C","Chloromonas D","Chloromonas F","Trebouxiophyceae G","Unknown E")
+scales::show_col(otu_colors)
+```
+
+Change colors in plot
+```{r}
+tsne_no_legend_nice_colors <- tsne_no_legend +
+  scale_color_manual(values=taxa_colors, na.value=na_color)
+tsne_no_legend_nice_colors
+```
+
 ```{r, fig.height=4, fig.width=2.2}
 p <- final_tsne +
+  scale_color_manual(values=taxa_colors, na.value=na_color) +
   guides(colour = guide_legend(override.aes = list(alpha = 1, size=3)), alpha="none") 
@@
 ```{r asv_mds, echo=F, fig.width=5, fig.height=4}
+set.seed(123)
 mds_coords <- dists %>% 
@@
                              label.fontsize = 8,
-                             con.type = 0,
+                             con.type = "none",
                              expand = unit(2, "mm")) +
   geom_point(aes(color = plot_taxa %>% fct_relevel(taxa.order), size = global_rel_abund, alpha=0.5))+
-  labs(color="ASV taxonomic assignment", tag="B",title="MDS") +
+  labs(color="ASV taxonomic assignment", tag="A",title="MDS") +
   ggthemes::theme_few() +
@@
   guides(colour = guide_legend(override.aes = list(alpha = 1, size=3)), alpha="none") +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  scale_color_manual(values=taxa_colors, na.value=na_color)
 
-  # geom_point(aes(alpha = 0.7, color = plot_taxa, size = global_rel_abund)) +
-  # scale_size(range = c(1, 6), guide="none") +
-  # scale_alpha(guide="none")+
-  # theme_few() +
-  # labs(color="ASV taxonomic assignment", tag="B")
-  # geom_text(aes(label = asv_id)) # too see outliers, asv_id=c("639", "534")
 mds
@@
 ```{r, fig.width=4, fig.height=6}
-asv_arrange <- grid.arrange(tsne_no_legend, mds, nrow=2)
+asv_arranged <- gridExtra::grid.arrange(mds, tsne_no_legend, nrow=2)
 ```
@@
 
+# make Fastas for breanna and jordan
+```{r setup_fastas}
+rbcl_seqs_otus <- otus_asv_name %>% 
+  left_join(key, by="asv_id")
+
+chlain_seqs <- rbcl_seqs_otus %>% 
+  filter(otu_name =="Chlainomonas A") %>% 
+  select(-otu_name) %>% 
+  deframe() %>% 
+  DNAStringSet()
+names(chlain_seqs) <- paste0("field_miseq_",names(chlain_seqs))
+
+sang_seqs <- rbcl_seqs_otus %>% 
+  filter(otu_name =="Sanguina B") %>% 
+  select(-otu_name) %>% 
+  deframe() %>% 
+  DNAStringSet()
+names(sang_seqs) <- paste0("field_miseq_",names(sang_seqs))
+
+# also just the ones that matched the database
+sang_taxa_seqs <- taxonomy %>% 
+  filter(genus=="Sanguina") %>% 
+  select(asv_id) %>% 
+  left_join(key, by="asv_id") %>% 
+  mutate(asv_id = paste0("field_miseq_",asv_id)) %>% 
+  deframe() %>% 
+  DNAStringSet()
+
+chlain_taxa_seqs <- taxonomy %>% 
+  filter(genus=="Chlainomonas") %>% 
+  select(asv_id) %>% 
+  left_join(key, by="asv_id") %>% 
+  mutate(asv_id = paste0("field_miseq_",asv_id)) %>% 
+  deframe() %>% 
+  DNAStringSet()
+
+```
+
 ```{r output}
@@
 
-ggsave(filename = here("manuscript/figs/2_taxonomy/asvs.pdf"), plot=asv_arrange, height = 6, width = 4)
-ggsave(filename = here("manuscript/figs/2_taxonomy/asv_legend.pdf"), plot=legend,  height = 4, width = 2.2)
+# write the main ASV fig, w legend seperate
+ggsave(filename = here("manuscript/figs/main/taxonomy/asvs.pdf"), plot=asv_arranged, height = 6, width = 4)
+ggsave(filename = here("manuscript/figs/main/taxonomy/asv_legend.pdf"), plot=legend,  height = 4, width = 2.2)
 
+# write out the trees
 ggsave(filename = here("manuscript/figs/otu_tree.png"), plot=otu_tree, width=7, height = 10)
 ggsave(filename = here("manuscript/figs/taxonomy_tree.png"), plot=taxonomy_tree, width=7, height = 10)
+
+# write out otu colors
+otu_colors %>% 
+  enframe() %>% 
+  write_csv(here("output/etc/otu_colors.csv"))
+
+# sanguina fasta for breanna
+writeXStringSet(sang_seqs, here("output/taxa_filtered_fastas/sanguina_otu.fasta"))
+writeXStringSet(sang_taxa_seqs, here("output/taxa_filtered_fastas/sanguina_idtaxa.fasta"))
+
+# chlainomonas fasta for jordan
+writeXStringSet(chlain_seqs, here("output/taxa_filtered_fastas/chlainomonas_otu.fasta"))
+writeXStringSet(chlain_taxa_seqs, here("output/taxa_filtered_fastas/chlainomonas_idtaxa.fasta"))
 ```

### 2019-11-16 code/18s_asv_cluster.Rmd
not using this file, not sure why ites here

--- 2019-11-05 code/rbcl_asv_clustering.Rmd
@@
 ---
-title: "rbcL ASV clustering"
+title: "18S ASV clustering"
 output: html_notebook
@@
 
-```{r libs, include=F}
+```{r}
 library(tidyverse)
-library(here)
 library(DECIPHER)
-library(ggforce)
-library(Rtsne)
-library(ape)
-library(ggtree)
-library(treeio)
-library(dbscan)
-library(ggrepel)
 
-library(ggpubr)
-library(ggthemes)
-library(ggdark)
 ```
 
-```{r read, include=F}
+```{r}
+aligns <- readDNAStringSet(here("data/asv_seq_data/algae_18s_quality_aligns.fasta"))
 
-rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
-taxonomy <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
-
-aligns <- readDNAStringSet(here("output/alignments/rbcl_aligns_in_main.fasta"))
-tree <- read.tree(file=here("output/tree/rbcl/asv613.fasta.treefile")) # this is the tree with only the ASVs that appear in the filtered rel abund dataset (not including protocol stuff, etc)
+# taxonomy w genbank (multiple copies of each designation, no editing)
+algae_taxonomy <- read_csv(here("output/algae_18s_assignments_snow_algae_genbank.csv"))
 ```
 
-```{r asv_distance_matrix, include=F}
+```{r}
 BrowseSeqs(aligns, highlight = T)
@@
 
-```{r tsne_1, echo=F}
+```{r mds}
+set.seed(123)
+mds_coords <- dists %>% 
+  cmdscale() %>% # ok to use euclidian dist, no null values when comparing relatedness
+  as_tibble() %>% 
+  add_column(asv_id = colnames(dists), .before=1)
+
+mds_coords_w_taxonomy <- mds_coords %>% 
+  left_join(algae_taxonomy, by="asv_id")
+
+ggplot(mds_coords_w_taxonomy, aes(x=V1, y=V2))+
+  geom_point(aes(color=best_assignment), size=4)
+```
+
+```{r}
 perplexity <- 30
@@
   left_join(taxonomy, by="asv_id")
-
-ggplot(tsne_pts_taxonomy, aes(x=V1,y=V2))+
-  geom_point(aes(color = genus)) +
-  labs(title = "Points labeled by genus")
-```
-
-```{r sp_tsne, echo=F}
-ggplot(tsne_pts_taxonomy, aes(x=V1,y=V2))+
-  geom_point(aes(color = species)) +
-  labs(title="Pts labeled by species")
-```
-
-Make it look nice, show both genus and species level assignments
-```{r tsne-plot1, echo=F}
 
@@
              aes(color = plot_taxa %>% fct_relevel(taxa.order)), 
-             alpha = 0.8) +
-  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment")
+             alpha = 0.8)
 ```
-
-```{r otu-cluster, echo=F}
-eps <- 4
-
-tsne_mat <- tsne_pts_taxonomy %>% 
-  select(1,2) %>% 
-  as.matrix()
-
-set.seed(123)
-
-dbscan_out <- dbscan::dbscan(tsne_mat, eps=eps) # eps=4 gives best results
-
-tsne_taxa_otu <- tsne_pts_taxonomy %>% 
-  add_column(otu = as.character(dbscan_out$cluster)) # tsne is in order of asv_id
-
-ggplot(tsne_taxa_otu, aes(x=V1, y=V2)) +
-  geom_point(aes(color = otu)) +
-  labs(title="dbscan clustering")
-```
-
-Manually edit the clustering to show *Chloromonas krienitzii* on relative abundance barplot
-
-```{r edit-otus, echo=F}
-tsne_edited_otus <- tsne_taxa_otu %>%
-  mutate(otu = case_when((plot_taxa == "Chloromonas_krienitzii") ~ "9",
-                         TRUE ~ otu))
-
-otus <- tsne_edited_otus %>% 
-  select(asv_id, otu)
-
-# get names for each otu
-rel_abund %>% 
-  left_join(otus, by="asv_id") %>% 
-  left_join(taxonomy, by="asv_id") %>% 
-  distinct(otu, plot_taxa) %>% 
-  arrange(otu)
-
-# get rel abund of each otu
-rel_abund %>% 
-  left_join(otus, by="asv_id") %>% 
-  group_by(otu) %>% 
-  summarise(rel_abund = sum(rel_abund)) %>% 
-  arrange(-rel_abund) %>% 
-  select(-rel_abund) %>% 
-  add_column(letter = LETTERS[1:9])
-
-# name each OTU, lettered in order of rel abund
-otu_names <- tibble(otu = as.character(1:9), 
-                    otu_name = c("Other Chloromonas",
-                                 "Sanguina B",
-                                 "Chlainomonas A",
-                                 "Chloromonas muramotoi C",
-                                 "Chloromonas D",
-                                 "Chloromonas F",
-                                 "Algae E",
-                                 "Trebouxiophyceae G",
-                                 "Chloromonas krienitzii"))
-
-otus_asv_name <- otus %>% 
-  left_join(otu_names, by="otu") %>% 
-  select(asv_id, otu_name)
-
-tsne_added_names <- tsne_edited_otus %>% 
-  left_join(otus_asv_name, by="asv_id")
-```
-
-```{r echo=F, fig.width=8, fig.height=6}
-ggplot(tsne_added_names, aes(x=V1, y=V2)) +
-  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), alpha=0.3, show.legend = FALSE) +
-  # plot points not assigned to genus or species
-  geom_point(data = filter(tsne_added_names, is.na(plot_taxa)), 
-             color = "grey",alpha = 0.8, size = 3) +
-  # color points that were assigned to genus or species
-  geom_point(data = filter(tsne_added_names, !is.na(plot_taxa)), 
-             aes(color = plot_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
-  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
-  theme_few() 
-```
-
-```{r include_rel_abund, echo=F}
-total_rel_abund <- rel_abund %>% 
-  group_by(asv_id) %>% 
-  summarise(global_rel_abund = sum(rel_abund))
-
-tsne_rel_abund <- tsne_added_names %>% 
-  left_join(total_rel_abund, by="asv_id")
-
-final_tsne <- ggplot(tsne_rel_abund, aes(x=V1, y=V2)) +
-  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), 
-                             show.legend = FALSE, 
-                             alpha=0.2, 
-                             linetype="blank", 
-                             expand = unit(2, "mm")) +
-  # plot points not assigned to genus or species
-  geom_point(data = filter(tsne_rel_abund, is.na(plot_taxa)), 
-             aes(size = global_rel_abund),
-             color = "grey",alpha = 0.5) +
-  # color points that were assigned to genus or species
-  geom_point(data = filter(tsne_rel_abund, !is.na(plot_taxa)), 
-             aes(color = plot_taxa %>% fct_relevel(taxa.order),
-                 size = global_rel_abund), 
-             alpha = 0.5) +
-  labs(color="ASV taxonomic assignment", tag="A") +
-  ggthemes::theme_few() +
-  scale_size(range = c(1, 6), guide="none") + ################################################# not changing the scale size
-  theme(legend.position = "none")
-final_tsne
-```
-
-```{r asv_mds, echo=F}
-mds_coords <- dists %>% 
-  cmdscale() %>% # ok to use euclidian dist, no null values when comparing relatedness
-  as_tibble() %>% 
-  add_column(asv_id = colnames(dists), .before=1)
-
-mds_asv_plot <- mds_coords %>% 
-  left_join(taxonomy, by="asv_id") %>% 
-  left_join(otus_asv_name, by="asv_id") %>% 
-  left_join(total_rel_abund, by="asv_id") %>% 
-  filter(!(asv_id %in% c("639", "534", "646"))) # remove Trebouxiophyceae outliers
-
-mds <- ggplot(mds_asv_plot, aes(V1, V2)) + 
-  ggforce::geom_mark_ellipse(data = mds_asv_plot %>% 
-                               filter(otu_name != "Trebouxiophyceae J"), # dont plot the ellipse around other treb
-                             aes(fill = otu_name, label=otu_name),
-                             show.legend = FALSE, 
-                             alpha=0.2, 
-                             linetype="blank",
-                             expand = unit(2, "mm")) +
-  geom_point(aes(alpha = 0.7, color = plot_taxa, size = global_rel_abund)) +
-  scale_size(range = c(1, 6), guide="none") +
-  scale_alpha(guide="none")+
-  theme_few() +
-  labs(color="ASV taxonomic assignment", tag="B")
-  # geom_text(aes(label = asv_id)) # too see outliers, asv_id=c("639", "534")
-mds
-```
-
-# Phylogenetic tree
-
-Run IQtree
-```{r tree, fig.width=7, fig.height=10}
-# root tree with Trebouxiophyceae
-root <- taxonomy %>% 
-  filter(class=="Trebouxiophyceae") %>% 
-  head(1) %>% 
-  pull(asv_id)
-
-rbcl.tree <- rbcl.tree %>% root(outgroup = root, resolve.root=T)
-
-rbcl_ggtree <- ggtree(rbcl.tree)
-
-# parse bootstrap values
-rbcl_ggtree_data <- rbcl_ggtree$data %>% 
-  filter(isTip == FALSE) %>% 
-  separate(label, c("boot", "bayes"), sep = "/") %>% #split label into boots and bayes vals
-  filter(boot > 80) %>%
-  mutate(boot = round(as.numeric(boot), digits = 0),
-         insignificant_edge = if_else(boot > 80, "black", "red"))# round boot to nearest integer
-
-# Plot trees
-
-# tips colored by taxonomy
-taxonomy_tree <- rbcl_ggtree %<+% taxonomy + 
-  geom_tippoint(aes(color=plot_taxa, alpha=0.7), size=3) + # 
-  geom_text2(data=rbcl_ggtree_data, aes(label=paste0(boot,";",bayes)),  vjust= 0.01, hjust = 1, size=3) + # add bootstrap vals
-  theme(legend.position = c(0.2,0.7), 
-        legend.title = element_blank(), 
-        legend.key = element_blank())
-
-# clades annotated by OTU
-otu_tree <- rbcl_ggtree %<+% otus_asv_name + 
-  geom_tippoint(aes(color=otu_name, alpha=0.7), size=3) + # 
-  geom_text2(data=rbcl_ggtree_data, aes(label=paste0(boot,";",bayes)),  vjust= 0.01, hjust = 1, size=3) + # add bootstrap vals
-  theme(legend.position = c(0.2,0.7), 
-        legend.title = element_blank(), 
-        legend.key = element_blank())
-
-taxonomy_tree
-otu_tree
-
-```
-
-todo: 
-- better color choices
-- circular tree
-
-```{r output}
-write_csv(otus_asv_name, path = here("output/asv_otu_key.csv"))
-
-ggsave(filename = here("manuscript/figs/2_taxonomy/tsne.png"), plot=final_tsne, height = 6, width = 8)
-ggsave(filename = here("manuscript/figs/2_taxonomy/mds.png"), plot=mds,  height = 6, width = 8)
-
-ggsave(filename = here("manuscript/figs/otu_tree.png"), plot=otu_tree, width=7, height = 10)
-ggsave(filename = here("manuscript/figs/taxonomy_tree.png"), plot=taxonomy_tree, width=7, height = 10) 
-```

### 2019-11-16 code/cellcount.Rmd
test
added in cell counts for missing samples, re organized a few things how I look at it

@@
 
-```{r heatmap, echo=F, fig.height=5, fig.width=3}
-p <- cellct %>% 
+```{r heatmap, echo=F, fig.height=5, fig.width=6}
+cellct_tbl <- cellct %>% 
   left_join(field, by="sample_id")
 
-distinct(cellct, sample_id) %>% nrow()
+distinct(cellct_tbl, sample_id) %>% nrow()
+```
 
-short_names <- p %>% 
-  mutate(short_morpho_name = case_when(morpho_sp=="Chlainomonas"~"Cr",
-                                       morpho_sp=="Chloromonas cf. brevispina"~"Cb",
-                                       morpho_sp=="Chloromonas cf. nivalis"~"Cn",
-                                       morpho_sp=="Chloromonas krienitzii"~"Ck",
-                                       morpho_sp=="Green cell"~"G",
-                                       morpho_sp=="Other"~"O",
-                                       morpho_sp=="Sanguina nivaloides"~"Sn"))
+Heatmap for figure 4
+```{r, echo=F, fig.height=5, fig.width=6}
+
+cellct_tbl_w_seq <- cellct_tbl %>% 
+  filter(sample_id %in% samples_w_seq)
+
+samples_w_seq ## not sure where this comes from
+samples_with_cellcounts <- cellct_tbl_w_seq %>% distinct(sample_id) %>% pull(sample_id)
+setdiff(samples_w_seq, samples_with_cellcounts) # missing these
+
+cellct_tbl_w_seq %>% distinct(sample_id)
+cellct_tbl_w_seq %>% distinct(morpho_sp)
+
+my_theme <- theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+cellct_ord <- c("Chlainomonas rubra","Sanguina nivaloides","Chloromonas krienitzii","Chloromonas cf. brevispina","Chloromonas cf. nivalis", "Green cell","Other")
+
+cellct_tbl_w_seq %>% distinct(sample_id) %>% nrow()
+
+cellct_heatmap <- ggplot(cellct_tbl_w_seq, 
+                         aes(y=alias %>% fct_reorder(elev_m), 
+                             x=morpho_sp %>% fct_relevel(cellct_ord))) +
+  geom_tile(aes(alpha = percent_of_sample, fill = morpho_sp %>% fct_relevel(cellct_ord))) +
+  labs(x="Cell morphology", y= "Elevation") +
+  scale_alpha(range = c(0.1, 1)) +
+  my_theme +
+  guides(alpha="none",fill="none")
+
+cellct_heatmap
+```
+
+```{r, echo=F, fig.height=5, fig.width=6}
 
 my_theme <- theme(axis.text.y=element_blank(),
-                  axis.ticks.y=element_blank(), 
-                  legend.position = "none")
+                  axis.ticks.y=element_blank())
 
-cellct_heatmap <- ggplot(short_names, aes(y=alias %>% fct_reorder(elev_m), 
-              x=short_morpho_name)) +
-  geom_tile(aes(alpha = percent_of_sample, fill = morpho_sp)) + # color = "white" %>% fct_reorder(morpho_order)
-  labs(x="Cell morphology", y= "Elevation", fill = "Cell morphology", alpha="Relative abundance") +
-  scale_alpha(range = c(0.1, 1)) +
-  my_theme
+p2 <- p %>% 
+ filter(!(morpho_sp %in% c("Other","Green cell")))
+cellct_scatter <- ggplot(p2, aes(x=date, y=elev_m)) +
+  geom_jitter(aes(color=morpho_sp, size=percent_of_sample), alpha=0.5, width = 10, height=10) +
+  scale_size(range=c(0.5, 10)) +
+  labs(color="Morphospecies",size="Relative abundance",y="Elevation (m)",x="Date") +
+  guides(colour = guide_legend(override.aes = list(alpha = 1, size=3))) +
+  ggthemes::theme_few()
+cellct_scatter
 
-cellct_heatmap
+# pie charts?
 ```
@@
 ```{r}
-dir.create(here("manuscript/figs/supp/cellct"))
-ggsave(filename=here("manuscript/figs/supp/cellct/cellct_heatmap.pdf"), plot=cellct_heatmap, height = 7, width = 3)
+# dir.create(here("manuscript/figs/supp/cellct"))
+ggsave(filename=here("manuscript/figs/supp/cellct/cellct_dotplot.pdf"), plot=cellct_scatter, height = 5, width = 6)
 ggsave(here("manuscript/figs/supp/cellct/elevations.pdf"), plot=elev, height=7, width=1)
+
+saveRDS(cellct_heatmap, here("output/cellct_w_seq_heatmap.Rda"))
 ```

### 2019-11-16 code/community.Rmd
init, this makes the heatmap that compares 18s rbcl and cellcount. also the cellcount scatterpie.

---
title: "Community comparison"
output: html_notebook
---

```{r libs, include=F}
library(tidyverse)
library(here)

library(GUniFrac)
library(vegan)

library(scatterpie)
library(ggrepel)
library(gridExtra)
library(ggthemes)
library(ggdark)
```

```{r globals, include=F}
# define theme for all heatmaps
my_heat_theme <- theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x=element_blank())
```

```{r input, include=F}
rbcl_rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv")) # not looking at the depth samples, controls..
rbcl_tax <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
rbcl_asv_seqs <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
rbcl_otus <- read_csv(here("output/asv_otu_key.csv"))
otu_colors <- read_csv(here("output/etc/otu_colors.csv"))

s18_rel_abund <- read_csv(here("data/03_filtered/18s_algae/18s_algae.filtered_cleaned_rel_abund.csv"))
s18_algae_taxonomy <- read_csv(here("output/algae_18s_assignments_snow_algae_genbank.csv"))
# s18_algae_select_taxonomy <- read_csv(here("output/algae_18s_assignments_w_select_snow_algae.csv")) #### this didnt work very well, need to make consensus for each species

field <- read_csv(here("data/02_tidied/tidied_field.csv"))
cellct <- read_csv(here("data/02_tidied/tidied.cell_count.csv"))

rbcl_tree <- read.tree(here("output/tree/rbcl/all_asvs/rbcl_aligns_all.fasta.treefile")) ### this tree should only contain the asvs we found in the "main" sites, not looking at controls, etc
s18_tree <- read.tree(here("output/tree/algae_18s/algae_18s_seqs.fasta.treefile"))
```

```{r cellct_scatter, echo=F}
cellct_scatter_tbl <- cellct %>% 
  left_join(field, by="sample_id") %>% 
  # not showing all cells in cell count, only Sanguina, Chlainomonas, Chloro b and n
  filter(!(morpho_sp %in% c("Other","Green cell")))

cellct_scatter <- ggplot(cellct_scatter_tbl, aes(x=date, y=elev_m)) +
  geom_jitter(aes(color=morpho_sp, size=percent_of_sample), alpha=0.5, width = 10, height=10) +
  scale_size(range=c(0.5, 10)) +
  labs(color="Morphospecies",size="Relative abundance",y="Elevation (m)",x="Date") +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size=3))) +
  ggthemes::theme_few()
cellct_scatter
```

```{r cellct_scatterpie, echo=F}
cellct_wide <- cellct %>%
  pivot_wider(names_from = morpho_sp, values_from = percent_of_sample, values_fill = list(percent_of_sample=0)) %>% 
  left_join(field, by="sample_id") %>% 
  select(date, elev_m, `Chloromonas krienitzii`:Other) %>% 
  mutate(date=lubridate::yday(date)) %>% # convert to julian date
  rownames_to_column("rowname") %>% 
  as.data.frame()

celltypes <- cellct %>% filter(!is.na(morpho_sp)) %>% distinct(morpho_sp) %>% pull()

base_plot <- ggplot(field, aes(x=date, y=elev_m))

min_date <- min(cellct_wide$date)
max_date <- max(cellct_wide$date)
date_diff <- max_date - min_date
min_elev <- min(cellct_wide$elev_m)
max_elev <- max(cellct_wide$elev_m)
elev_diff <- max_elev - min_elev

cellct_converted <- cellct_wide %>% 
  mutate(elev_m = (elev_m - min_elev)/elev_diff, # express as fraction
         date = (date - min_date)/date_diff) # " "

scatterpie <- ggplot() +
  geom_scatterpie(aes(x=date, y=elev_m, group=elev_m, r=0.04), data=cellct_converted, cols=celltypes, color=NA, alpha=0.6)
scatterpie

```

```{r cellct_heat, echo=F}
# only plot the samples we want to compare
cellct_tbl_w_seq <- cellct %>% 
  filter(sample_id %in% rbcl_rel_abund$sample_id)

# sanity check
samples_with_cellcounts <- cellct_tbl_w_seq %>% distinct(sample_id) %>% pull(sample_id)
setdiff(rbcl_rel_abund$sample_id, samples_with_cellcounts) # samples we are missing cellcts for

# specify plotting order
cellct_ord <- c("Chlainomonas rubra","Sanguina nivaloides","Chloromonas krienitzii","Chloromonas cf. brevispina","Chloromonas cf. nivalis", "Green cell","Other")

# add metadata
cellct_field <- cellct_tbl_w_seq %>% 
  left_join(field, by="sample_id")

# set colors 
cellct_colors <- c("Chlainomonas rubra"="#bebada",
                   "Sanguina nivaloides"="#FCA69C",
                   "Chloromonas krienitzii"="#8dd3c7",
                   "Chloromonas cf. brevispina"="#8dd3c7",
                   "Chloromonas cf. nivalis"="#8dd3c7",
                   "Green cell"="grey70",
                   "Other"="grey70")

# plot
cellct_heatmap <- ggplot(cellct_field, 
                         aes(x=alias %>% fct_reorder(elev_m), 
                             y=morpho_sp %>% fct_relevel(cellct_ord))) +
  geom_tile(aes(alpha = percent_of_sample, fill = morpho_sp %>% fct_relevel(cellct_ord))) +
  labs(y="Cell morphology") +
  scale_alpha(range = c(0.01, 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x=element_blank()) +
  guides(alpha="none",fill="none") +
  my_heat_theme +
  scale_fill_manual(values = cellct_colors)
cellct_heatmap
```

```{r s18_heat_setup, include=F}
# calculate rel abund as % of Chlorophyta reads
s18_algae_rel_abund <- s18_rel_abund %>% 
  group_by(sample_id) %>% 
  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
  ungroup()

# check whats missing from this compared to rbcl
s18_sample_ids <- s18_algae_rel_abund %>% distinct(sample_id) %>% pull()
rbcl_sample_ids <- rbcl_rel_abund %>% distinct(sample_id) %>% pull()
setdiff(rbcl_sample_ids, s18_sample_ids)

# calculate relative abundance per genus
best_assignment_rel_abund <- s18_algae_rel_abund %>% 
  left_join(s18_algae_taxonomy, by="asv_id") %>% 
  group_by(sample_id, best_assignment) %>% 
  summarise(rel_abund = sum(rel_abund)) %>% 
  ungroup()

s18_rel_abund_lumped <- best_assignment_rel_abund %>% 
  mutate(lumped_best_assignment = case_when(best_assignment=="Sanguina_nivaloides"~"Sanguina nivaloides",
                                     best_assignment %>% str_detect("Chloromonas")~"Chloromonas",
                                     # best_assignment == "Chlorogonium"~"Chlorogonium",
                                     best_assignment=="Koliellaceae"~"Trebouxiophyceae",
                                     TRUE~"Other")) %>% 
  select(-best_assignment)

s18_rel_abund_lumped
```

```{r s18_heat_plot, include=F}
# specify plot order
assign_ord <- c("Sanguina nivaloides", "Chloromonas","Trebouxiophyceae","Other")

# add field data
rel_abund_field <- s18_rel_abund_lumped %>% 
  left_join(field, by="sample_id")

# set colors: sanguina, chloromonas
s18_heat_colors <- c("Sanguina nivaloides"="#FCA69C",
                     "Chloromonas" = "#8dd3c7",
                     "Trebouxiophyceae"="#80b1d3",
                     "Other" = "grey70")

# plot
s18_heatmap <- ggplot(rel_abund_field, 
                       aes(y=lumped_best_assignment %>% fct_relevel(assign_ord), 
                           x=alias %>% fct_reorder(elev_m))) +
  geom_tile(aes(alpha=rel_abund, fill=lumped_best_assignment %>% fct_relevel(assign_ord))) +
  labs(y="18S taxonomy") +
  scale_alpha(range = c(0.01, 1)) +
  guides(alpha="none", fill="none") +
  my_heat_theme +
  scale_fill_manual(values = s18_heat_colors)
s18_heatmap
```

```{r rbcl_heat, include=F}
# group by OTU
rbcl_rel_abund <- rbcl_rel_abund %>% 
  left_join(rbcl_otus, by="asv_id") %>% 
  group_by(sample_id, otu_name) %>% 
  summarise(rel_abund = sum(rel_abund)) %>%   
  ungroup()

# add grouping metadata
rbcl_rel_abund_field <- rbcl_rel_abund %>% 
  left_join(field, by="sample_id")

# specify colors
rbcl_otu_colors <- c("Chlainomonas A"="#bebada",
                     "Sanguina B"="#FCA69C",
                     "Other Chloromonas"="#8dd3c7",
                     "Chloromonas krienitzii"="#8dd3c7",
                     "Chloromonas muramotoi C"="#8dd3c7", 
                     "Chloromonas D"="#8dd3c7", 
                     "Chloromonas F"="#8dd3c7", 
                     "Trebouxiophyceae G"="#80b1d3", 
                     "Unknown E"="grey70")

# specify factor order
otu_ord <- c("Chlainomonas A","Sanguina B", "Other Chloromonas","Chloromonas krienitzii","Chloromonas muramotoi C","Chloromonas D","Chloromonas F", "Trebouxiophyceae G","Unknown E")

rbcl_heatmap <- ggplot(rbcl_rel_abund_field, aes(y=otu_name %>% fct_relevel(otu_ord), x=alias %>% fct_reorder(elev_m))) +
  geom_tile(aes(alpha=rel_abund, fill=otu_name %>% fct_relevel(otu_ord))) +
  labs(y="rbcL OTU") +
  scale_alpha(range = c(0.01, 1)) +
  my_heat_theme +
  guides(alpha="none", fill="none") +
  scale_fill_manual(values= rbcl_otu_colors)
rbcl_heatmap
```

```{r combined_heatmap_setup, echo=F, fig.width=7,fig.height=4}
# adjust tbls so they can be combined
cellct_combine <- cellct_tbl_w_seq %>% 
  mutate(marker = "Cell morphology") %>% 
  dplyr::rename(species=morpho_sp, rel_abund=percent_of_sample) %>% 
  select(sample_id, marker, species, rel_abund)

s18_combine <- s18_rel_abund_lumped %>% 
  mutate(marker = "18S") %>% 
  dplyr::rename(species=lumped_best_assignment) %>% 
  select(sample_id, marker, species, rel_abund)

rbcl_combine <- rbcl_rel_abund %>%
  mutate(marker="rbcL") %>% 
  dplyr::rename(species = otu_name) %>% 
  select(sample_id, marker, species, rel_abund)

# combine the tbls into one
combined <- bind_rows(cellct_combine, s18_combine, rbcl_combine) %>% 
  left_join(field, by="sample_id")

# specify factor order
species_ord <- c("Chlainomonas A","Chlainomonas rubra", 
                "Sanguina B", "Sanguina nivaloides",
                "Chloromonas krienitzii","Chloromonas muramotoi C", "Chloromonas D", "Chloromonas cf. brevispina", "Chloromonas cf. nivalis","Chloromonas F",
                "Other Chloromonas", "Chloromonas",
                "Trebouxiophyceae G", "Trebouxiophyceae",
                "Unknown E", "Green cell", "Other")

# specify colors
plot_colors <- c(rbcl_otu_colors,s18_heat_colors, cellct_colors)

# plot
community_heatmap <- ggplot(combined, aes(x=alias %>% fct_reorder(elev_m), 
                                          y = species %>% fct_relevel(species_ord))) + 
  geom_tile(aes(alpha=rel_abund, 
                fill=species %>% fct_relevel(species_ord) )) + 
  facet_grid(rows = vars(marker %>% fct_relevel(c("Cell morphology","18S","rbcL"))), 
             scales="free_y", space="free_y") +
  scale_alpha(range = c(0.01, 1)) +
  guides(alpha="none", fill="none") +
  scale_fill_manual(values = plot_colors) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.title.y=element_blank()) +
  labs(x="Sample ID")

community_heatmap
```

```{r output, echo=F}
ggsave(here::here("manuscript/figs/cellct_scatterpie.pdf"), plot=scatterpie, width=7, height=5)
ggsave(here::here("manuscript/figs/cellct_axis.pdf"), plot=base_plot, width=5, height=5)

ggsave(here("manuscript/figs/main/community/community_ggplot.pdf"), plot=community_heatmap, width=7, height=4)
```

### 2019-11-16 code/community_comparison_18s.Rmd
update

@@
 
-best_assignment_rel_abund <- algae_percent_rel_abund %>% 
+# check whats missing from this compared to rbcl
+
+rbcl_sample_ids <- algae_percent_rel_abund %>% distinct(sample_id) %>% pull()
+s18_sample_ids <- rel_abund_lumped %>% distinct(sample_id) %>% pull()
+setdiff(s18_sample_ids,rbcl_sample_ids)
+
+# add a empty row for gar18.04 (adding it upstream of left joining the field)
+
+## add in empty row for gar18.04
+gar4_NA_tibble <- tibble(sample_id="gar18.04",
+                         protocol=NA_character_, 
+                         asv_id=NA_character_, 
+                         n_reads=NA_character_,
+                         rel_abund=NA_character_ )
+
+rel_abund_w_gar4 <- algae_percent_rel_abund %>% 
+  bind_rows(gar4_NA_tibble)
+
+best_assignment_rel_abund <- rel_abund_w_gar4 %>% 
   # calculate relative abundance per genus
@@
   left_join(field, by="sample_id")
+
+view(best_assignment_rel_abund)
 
@@
 barplot
+```
 
-# desired categories: lump C nivalis in Chloromonas, lump
+```{r heatmap_18s, include=FALSE}
+# lump categories
+rel_abund_lumped <- best_assignment_rel_abund %>% 
+  mutate(lumped_best_assignment = case_when(best_assignment=="Sanguina_nivaloides"~"Sanguina nivaloides",
+                                     best_assignment %>% str_detect("Chloromonas")~"Chloromonas",
+                                     # best_assignment == "Chlorogonium"~"Chlorogonium",
+                                     best_assignment=="Koliellaceae"~"Trebouxiophyceae",
+                                     TRUE~"Other"))
+view(rel_abund_lumped)
 
+assign_ord <- c("Sanguina nivaloides", "Chloromonas","Trebouxiophyceae","Other")
+
+my_theme <- theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+s18_heatmap <- ggplot(rel_abund_lumped, 
+                       aes(x=lumped_best_assignment %>% fct_relevel(assign_ord), 
+                           y=alias %>% fct_reorder(elev_m))) +
+  geom_tile(aes(alpha=rel_abund, fill=lumped_best_assignment %>% fct_relevel(assign_ord))) +
+  labs(x="18S taxonomy", y= "Elevation") +
+  scale_alpha(range = c(0.01, 1)) +
+  my_theme +
+  guides(alpha="none", fill="none")
+s18_heatmap
+
+# manually remove the Other bar for G1.4
 ```
@@
 barplot_select
-
 ```
@@
 ggsave(here("manuscript/figs/18s_algae_rel_abund.png"), plot = barplot)
+saveRDS(s18_heatmap, here("output/s18_heatmap.Rda"))
 ```

### 2019-11-16 code/rbcl_community_compare.Rmd
updates... only using the NMDS UniFrac and the Diversity scatterplots from this script at this point, the final heatmap migrated into the community.Rmd since im comparing 18s and cellct in that one

@@
 rbcl.tree <- read.tree(here("output/tree/rbcl/all_asvs/rbcl_aligns_all.fasta.treefile")) ### this tree should only contain the asvs we found in the "main" sites, not looking at controls, etc
+
+# the cell counts and 18S rel abund data
+cellct_heatmap <- readRDS(here("output/cellct_w_seq_heatmap.Rda"))
+s18_heatmap <- readRDS(here("output/s18_heatmap.Rda"))
 ```
@@
 
-```{r unifrac_plot, echo=F}
+```{r unifrac_plot, echo=F, fig.width=5, fig.height=4}
 # plot
@@
   labs(color="Elevation (m)", tag="A") +
-  geom_text_repel(aes(label=alias))
+  geom_text_repel(aes(label=alias), size=3, segment.size = 0.3) +
+  theme_classic()
 uf_plot
@@
 
-otu_rel_abund_round_to_1_percent <- otu_rel_abund %>% 
-  mutate(rel_abund = if_else(rel_abund <0.01, 0.01, rel_abund)) # artificially boost rel abund <1% to 1% so its visible on plot
-
-# check rel abund sums ~= 1 (will be slightly higher, due to rounding)
-otu_rel_abund_round_to_1_percent %>%
-  group_by(alias) %>%
-  summarise(sum(rel_abund))
+# otu_rel_abund_round_to_1_percent <- otu_rel_abund %>% 
+#   mutate(rel_abund = if_else(rel_abund <0.01, 0.01, rel_abund)) # artificially boost rel abund <1% to 1% so its visible on plot
+# 
+# # check rel abund sums ~= 1 (will be slightly higher, due to rounding)
+# otu_rel_abund_round_to_1_percent %>%
+#   group_by(alias) %>%
+#   summarise(sum(rel_abund))
 ```
 
-```{r plot_barplot, echo=F}
-cust_theme <- theme(legend.position = "none",
-                    panel.grid.major.x = element_blank(), # remove vertical lines in background
-                    panel.grid.minor.x = element_blank(),
-                    axis.line=element_blank(), # remove x axis
-                    axis.title.x=element_blank(), 
-                    axis.text.x=element_blank(),
-                    axis.ticks.x=element_blank(),
-                    panel.spacing = unit(-0.01, "lines"), # remove space between facet panels
-                    strip.text.x = element_text(angle = 90, hjust=1.0, vjust = 0.8), # rotate facet labels 90
-                    strip.background = element_blank()) # remove grey boxes around facet labs
+```{r plot_barplot, echo=F, fig.width=2.9, fig.height=5}
+otu_ord <- c("Chlainomonas A","Sanguina B", "Other Chloromonas","Chloromonas krienitzii","Chloromonas muramotoi C","Chloromonas D", "Unknown E","Chloromonas F", "Trebouxiophyceae G")
 
-otus %>% distinct(otu_name)
+my_theme <- theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
-otu_ord <- c("Chlainomonas A","Sanguina B", "Other Chloromonas","Chloromonas krienitzii","Chloromonas muramotoi C","Chloromonas D", "Algae E","Chloromonas F", "Trebouxiophyceae G")
+otu_rel_abund %>% distinct(sample_id) %>% nrow()
 
-barplot <- ggplot(otu_rel_abund_round_to_1_percent, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
-  geom_bar(aes(fill=otu_name %>% fct_relevel(otu_ord)), stat="identity") + # classic stacked relative abundance plot
-  coord_flip() +
-  facet_grid(~otu_name %>% fct_relevel(otu_ord), switch = "x") +
-  labs( x="Sample ID", tag="B") +
-  cust_theme
-barplot
+rbcl_heatmap <- ggplot(otu_rel_abund, aes(x=otu_name %>% fct_relevel(otu_ord), y=alias %>% fct_reorder(elev_m))) +
+  geom_tile(aes(alpha=rel_abund, fill=otu_name %>% fct_relevel(otu_ord))) +
+  labs(x="rbcL OTU", y= "Elevation", fill = "rbcL OTU", alpha="Relative abundance") +
+  scale_alpha(range = c(0.01, 1)) +
+  my_theme +
+  guides(alpha="none", fill="none")
+rbcl_heatmap
+```
+
+```{r colors, include=F}
+# get colors, so they match across plots
+g <- ggplot_build(rbcl_heatmap)
+
+colors <- tibble(otu_name = otu_rel_abund$otu_name,
+                 color = g$data[[1]]$fill) %>% 
+  distinct(otu_name, color) %>% 
+  deframe()
+
+length(colors)
+colors
 
@@
 
-```{r combine, fig.width=7, fig.height=10}
-combined_fig_3 <- gridExtra::grid.arrange(uf_plot, barplot, nrow=2)
+Now add in the 18s and cellcount data side by side the rbcL OTU relative abundance data
+- set colors same
+- add in missing cellcount data
+- plot left to right (entire width of page): rbcl, cellct, 18s, with rbcl taking up most width
+- 
+```{r 18s_rel_abund, fig.width=1.7, fig.height=5}
+s18_heat <- s18_heatmap +
+  scale_fill_manual(values=c("#D39200", "#93AA00", "#FF61C3","#808080"))
+s18_heat
+```
+
+```{r cellct, fig.width=2.6, fig.height=5}
+cellct_heat <- cellct_heatmap +
+  scale_fill_manual(values = c("#F8766D","#D39200","#00BA38","#93AA00","#93AA00","#808080","#808080","#808080"))
+cellct_heat
+```
+
+```{r elev_strip, fig.width=1.5, fig.height=5}
+my_theme <- theme_classic() +
+  theme(axis.line=element_blank(),
+        axis.title.x = element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(),
+        axis.title.y = element_blank(),
+        axis.text.y=element_blank(),
+        axis.ticks.y=element_blank())
+
+elev_strip_tbl <- otu_rel_abund %>% 
+  distinct(alias, elev_m) %>% 
+  arrange(-elev_m) %>% 
+  add_column(x="column_1") %>% 
+  mutate(elev_m = elev_m %>% round(digits=-1))
+nrow(elev_strip_tbl)
+
+elev_strip <- ggplot(elev_strip_tbl, aes(x=x, y=alias %>% fct_reorder(elev_m))) +
+  geom_tile(aes(fill=elev_m)) +
+  scale_fill_gradientn(colours = terrain.colors(8)[1:7]) +
+  labs(x="Elevation (m)") +
+  my_theme 
+elev_strip
+```
+
+```{r elev_strip2, fig.width=0.5, fig.height=5}
+elev_strip2 <- elev_strip +
+  guides(fill="none") +
+  geom_text(aes(label=elev_m), size=3)
+elev_strip2
+```
+
+```{r snow_color_strip, fig.width=0.5, fig.height=5}
+color_strip_tbl <- otu_rel_abund %>% 
+  arrange(-elev_m) %>% 
+  distinct(alias, color_snow, elev_m) %>% 
+  add_column(x="column_1")
+
+color_strip_tbl %>% distinct(color_snow)
+
+color_strip <- ggplot(color_strip_tbl, aes(x=x, y=alias %>% fct_reorder(elev_m))) +
+  geom_tile(aes(fill=color_snow)) +
+  my_theme +
+  guides(fill="none") +
+  scale_fill_manual(values=c(pink="red",other="grey",green="green",orange="orange",rusty="purple"))
+color_strip
 ```
@@
 ```{r output, include=F}
-ggsave(filename = here("manuscript/figs/3_community/combined_fig_3.png"), plot=combined_fig_3, height=10, width=7)
+
+ggsave(here("manuscript/figs/main/community/rbcl.pdf"),plot=rbcl_heatmap, width = 2.9, height = 5)
+ggsave(here("manuscript/figs/main/community/18s.pdf"),plot=s18_heat, width = 1.7, height = 5)
+ggsave(here("manuscript/figs/main/community/cellct.pdf"),plot=cellct_heat, width = 2.6, height = 5)
+ggsave(here("manuscript/figs/main/community/elev.pdf"),plot=elev_strip, width = 1.5, height = 5)
+ggsave(here("manuscript/figs/main/community/elev2.pdf"),plot=elev_strip2, width = 0.5, height = 5)
+ggsave(here("manuscript/figs/main/community/snow_color.pdf"),plot=color_strip, width = 0.5, height = 5)
+
+ggsave(filename = here("manuscript/figs/main/uf_plot.png"), plot=uf_plot, width=5, height = 4)
 ggsave(filename = here("manuscript/figs/diversity.png"), plot = diversity_plots, height = 10, width = 7)

### 2019-11-16 code/misc/custom_ggplot_rel_abund_not_used.R
old script init

@@ -1,0 +1,40 @@
+# a few custom ggplot scripts set aside for later
+
+otu_rel_abund <- rel_abund %>% 
+  left_join(otus, by="asv_id") %>% 
+  group_by(sample_id, otu_name) %>% 
+  summarise(rel_abund = sum(rel_abund)) %>%   
+  left_join(field, by="sample_id") %>%
+  ungroup()
+
+otu_rel_abund_round_to_1_percent <- otu_rel_abund %>% 
+  mutate(rel_abund = if_else(rel_abund <0.01, 0.01, rel_abund)) # artificially boost rel abund <1% to 1% so its visible on plot
+
+# check rel abund sums ~= 1 (will be slightly higher, due to rounding)
+otu_rel_abund_round_to_1_percent %>%
+  group_by(alias) %>%
+  summarise(sum(rel_abund))
+
+cust_theme <- theme(legend.position = "none",
+                    panel.grid.major.x = element_blank(), # remove vertical lines in background
+                    panel.grid.minor.x = element_blank(),
+                    axis.line=element_blank(), # remove x axis
+                    axis.title.x=element_blank(), 
+                    axis.text.x=element_blank(),
+                    axis.ticks.x=element_blank(),
+                    panel.spacing = unit(-0.01, "lines"), # remove space between facet panels
+                    strip.text.x = element_text(angle = 90, hjust=1.0, vjust = 0.8), # rotate facet labels 90
+                    strip.background = element_blank()) # remove grey boxes around facet labs
+
+otus %>% distinct(otu_name)
+
+
+otu_ord <- c("Chlainomonas A","Sanguina B", "Other Chloromonas","Chloromonas krienitzii","Chloromonas muramotoi C","Chloromonas D", "Unknown E","Chloromonas F", "Trebouxiophyceae G")
+
+barplot <- ggplot(otu_rel_abund_round_to_1_percent, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
+  geom_bar(aes(fill=otu_name %>% fct_relevel(otu_ord)), stat="identity") + # classic stacked relative abundance plot
+  coord_flip() +
+  facet_grid(~otu_name %>% fct_relevel(otu_ord), switch = "x") +
+  labs( x="Sample ID", tag="B") +
+  cust_theme
+barplot

### 2019-11-16 code/misc/make_fasta_unknown_otu_asvs.R
init

@@ -1,0 +1,18 @@
+# make fasta of Unknown OTU
+
+library(tidyverse)
+library(DECIPHER)
+
+key <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
+otus <- read_csv(here("output/asv_otu_key.csv"))
+
+unknown_otu <- otus %>% 
+  filter(otu_name == "Unknown E") %>% 
+  pull(asv_id)
+
+unknown_seqs_DSS <- key %>% 
+  filter(asv_id %in% unknown_otu) %>% 
+  deframe() %>% 
+  DNAStringSet()
+
+writeXStringSet(unknown_seqs_DSS, here("output/unknown_otu_asvs.fasta"))

### 2019-11-16 code/wrangle/tidy_cell_count.R
updated

@@ -3,7 +3,7 @@
 library(tidyverse)
 library(here)
 
-raw_cellct <- read_csv(here("data/01_raw/cellct_2018.csv"))
+raw_cellct <- read_csv(here::here("data/01_raw/cellct_2018.csv"))
 glimpse(raw_cellct)
 
 
@@ -20,8 +20,8 @@ cellct <- raw_cellct %>%
   drop_na() %>% 
   mutate(morpho_sp = case_when( str_detect(morpho_sp, "green")~"Green cell", # lump all green categories
                                 morpho_sp=="orb"~"Sanguina nivaloides", 
-                                morpho_sp=="ruby"~"Sanguina nivaloides", # assuming Sanguina grows turrets 
-                                morpho_sp=="balloon"~"Chlainomonas",
+                                morpho_sp=="ruby"~"Sanguina nivaloides", # assuming Sanguina grows turrets
+                                morpho_sp=="balloon"~"Chlainomonas rubra",
                                 morpho_sp=="tangerine"~"Chloromonas krienitzii",
                                 morpho_sp=="hedgehog"~"Chloromonas cf. brevispina",
                                 morpho_sp=="lemon"~"Chloromonas cf. nivalis",
@@ -29,6 +29,7 @@ cellct <- raw_cellct %>%
                                 morpho_sp=="oval"~"Other",
                                 morpho_sp=="zuke"~"Other",
                                 morpho_sp=="halo"~"Other",
+                                morpho_sp=="unknown"~"Other",
                                 morpho_sp=="citrine"~"Sanguina nivaloides",
                                 TRUE ~ morpho_sp)) %>% 
   # drop samples with count < thresh
@@ -37,10 +38,21 @@ cellct <- raw_cellct %>%
   filter(count_total > thresh) %>% 
   # calculate percentage
   mutate(percent_of_sample = count / count_total) %>% 
-  ungroup()
+  ungroup() %>% 
+  # lump multiple rows of "Other" to same row
+  group_by(sample_id, morpho_sp) %>%
+  summarise(percent_of_sample=sum(percent_of_sample))
 
 head(cellct)
 
 
+# add in NA for wed2, sey7, and bre10
+# nas <- tibble(sample_id = c("wed18.02","sey18.07","bre18.10"),
+#               morpho_sp = NA_character_, 
+#               count=NA_character_, 
+#               count_total=NA_character_)
+# 
[…]

### 2019-11-16 code/wrangle/tidy_silva_tax_assignment.R
change path output

@@ -69,6 +69,6 @@ silva_3 <- key %>%
 
 silva_3
 
-write_csv(key, here("output/asv_seq_key_18s.csv"))
-writeXStringSet(seqs_to_fasta, here("output/all_18s.fasta"))
+write_csv(key, here("data/asv_seq_data/asv_seq_key_18s.csv"))
+writeXStringSet(seqs_to_fasta, here("data/asv_seq_data/all_18s.fasta"))
 write_csv(silva_3, here("data/02_tidied/tidied_silva_18s_assignments.csv"))

### 2019-11-16 manuscript/supplemental.Rmd
path updates, legend

--- 2019-11-05 manuscript/supplemental.Rmd
@@
 ```{r, include=F}
-map <- knitr::include_graphics(here::here("manuscript/figs/map.png"))
-field <- read_csv(here::here("data/02_tidied/tidied_field.csv"))
-field_photos <- knitr::include_graphics(here::here("manuscript/figs/S3_site_field_photos/field_collage.png"))
+map <- knitr::include_graphics(here::here("manuscript/figs/map.png")) # crop zoom, city font white, each site gets one point with box specifying elev range, n samples, dates, 
+field_tbl <- read_csv(here::here("data/02_tidied/tidied_field.csv")) #
+field_photos <- knitr::include_graphics(here::here("manuscript/figs/supp/field_photos/field_collage.png"))
 micro_photos <- knitr::include_graphics(here::here("manuscript/figs/S4_site_micro_photos/micro.png"))
-s18 <- include_graphics(here("manuscript/figs/18s_algae_rel_abund.png"))
-tree <- include_graphics(here("manuscript/figs/otu_tree.png"))
-tree_taxonomy <- include_graphics(here("manuscript/figs/taxonomy_tree.png"))
-diversity <- include_graphics(here("manuscript/figs/diversity.png"))
-cellct <- include_graphics(here("manuscript/figs/supp/cellct/cellct.png"))
+cellct <- include_graphics(here("manuscript/figs/supp/cellct_scatterpie/cellct_scatterpie.png")) # make pies circles circular! they're shightly squished lengthwise
+tree_otu <- include_graphics(here("manuscript/supp/figs/otu_tree.png"))
+tree_taxonomy <- include_graphics(here("manuscript/figs/supp/taxonomy_tree.png"))
+diversity <- include_graphics(here("manuscript/figs/supp/diversity.png"))
+depth <- include_graphics(here("manuscript/figs/supp/depth/depth_polished.png")) ###
 
-map_cap <- "S1. Map showing locations of snow algae sampling locations in summer 2018."
-table_cap <- "S2. Table of sample site information of sequenced snow algae samples."
-field_photo_cap <- "S3. Photos of snow algae field sites."
-micro_cap <- "Microphotos of snow algae samples included in sequencing run."
-s18_cap <- "Relative abundance barplot of 18S amplicon library"
+map_cap <- "Map showing locations of snow algae sampling locations in summer 2018."
+table_cap <- "Table of sample site information of sequenced snow algae samples."
+field_photo_cap <- "Photos of snow algae field sites, arranged from high to low elevation from top to bottom. Alphanumeric code in upper left indicates sample ID."
+micro_photo_cap <- "Microphotos of snow algae samples included in sequencing run."
+cellct_cap <- "Proportional abundance of snow algae cell morphologies, plotted against date and elevation."
 tree_cap <- "Maximum likelihood tree of rbcL amplicon library. Colors represent OTU assignments."
@@
 diversity_cap <- "A. Shannon diversity of snow algae samples, plotted against sample elevation. Smooth method = linear model. B. Faith's phylogenetic diversity plotted against sample elevation."
-(ref:cellct_cap) "Heatmap showing cell count relative abundance from 117 snow algae samples.
-Cb = Chlainomonas cf.
-brevispina, Ck=Chloromonas krienitzii, Cn=Chloromonas cf.
-nivalis, Cr=Chlainomonas rubra, G=green cells, O=other, Sn=Sanguina nivaloides"
+depth_cap <- "A.
+Representative field photo of snow runnel, with orange on surface and green below.
+B.
+Microphoto of orange snow on surface of sample S9.1.
+C.
+Microphoto of snow algae below surface of sample S9.1.
+D.
+Relative abundance barplot of OTUs in three snow runnels.
+Sample H2.10 depth and surface samples were taken 1 week apart, other depth/surface pairs are from the same date."
 ```

### 2019-11-16 manuscript/mbb801/MBB801_figs.Rmd
little write up for Marks class

---
title: "MBB801 figure assignment"
author: "Casey Engstrom"
date: "12/11/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fig. 1: Field site and microphotos

- consistent sized scale bars
- evenly spaced layout
- intuitive layout ("gestalt"--Wong 2010)
  - high elevation sites on top and vice versa
  - microphoto corresponds with the field photo on it's left

## Fig. 2: Cell morphologies by elevation

- intuitive layout 
  - elevation in y axis
  - time on x-axis
  - abundance is proportional to size
  - different categories of cell are different colors

## Fig. 3: *rbc*L ASV clustering

- intuitive layout
  - MDS ("traditional" clustering) on top, "newer" t-SNE clustering on bottom
  - seperate genera on legend with space
  - NA is grey
  - size proportional to relative abundance
- contrasting colors
  - each point color distinguishable in t-SNE

## Fig. 4: Heatmap of *rbc*L, 18S, and cell count relative abundance

- intuitive layout
  - elevation y axis
  - intuitive colors: 
    - "terrain" colors for elevation
    - actual color of the sampled snow 
    - consistant taxonomy colors; e.g. *Sanguina* is the same color in rbcL, 18S, and cell count
- because amplicon data is semi-quantative, using alpha (opacity) to represent relative abundance is appropriate and intuitive
- different colors allow for fast comparison between data sets for particular taxa
- although it makes heavy use of red and green, this is color blind friendly because only the intensity of the color matters within each column, and each column is labelled, thus alleviating the need for a color legend
- although Gehlenborg and Wong (2012) reccommend clustering by similarity, this would not be approproate for this presentation, because we want a fixed y axis factored by elevation, and the taxonomy columns should maintain the same order from rbcL, 18S and cell count. Furthermore, Figure 5 uses a more sophisticated method of clustering that is more appropriate to this data type than naive heirarchical clustering
- brewer palettes (Wong 2012)

## Fig. 5: *rbc*L uniFrac NMDS

- intuitive layout
  - same color scheme for elevation as in Fig. 4
  - unobtrusive font, staggered to avoid overplotting

### 2019-11-16 manuscript/manuscript.Rmd
minor edits

@@
 Among the Chlorophyceae, both *Sanguina nivaloides* [@Prochazkova2019a] or *Chlainomonas* [@Prochazkova2018c] produce the red pigment astaxanthin responsible for the red color of some snow algae blooms.
-Orange snow is reported to be caused by *Sanguina aurentia* [@Prochazkova2019a] and some species of *Chloromonas* [@Prochazkova2019c;@Remias2013a], while algae lacking secondary pigments color the snow green, such as *Chloromonas pichinchae* [@Hoham1975].
+Orange snow is reported to be caused by *Sanguina aurentia* [@Prochazkova2019a] and some species of *Chloromonas* [@Prochazkova2019a;@Remias2013a], while algae lacking secondary pigments color the snow green, such as *Chloromonas pichinchae* [@Hoham1975b].
 However, some species are known from only one or two samples, and the recently described *Chlainomonas* and *Sanguina* are poorly explored, with only two species in each.
@@
 Despite over a century of study, no one has successfully cultured the snow algae that produce red secondary pigments.
-In many green algae cells produce secondary pigments in just one life stage, remaining green for much of their life cycle [@Shah2009;@Salome2019]; thus red snow algae could originate from green cells.
+In model organism green algae *Chlamydomonas reinhardtii* cells produce secondary pigments in just one life stage, remaining green for much of their life cycle [@Salome2019], thus red snow algae could originate from green cells.
 Green snow is frequently found in waterlogged snow, sometimes alongside red snow, leading some to hypothesize that green snow could be an earlier life stage of red snow.
-Two studies comparing green and red snow using amplicon high-throughput sequencing (HTS) found they were caused by different species [@Lutz2015;@Terashoma2016]; however it remains an open question whether green snow can develop into red snow.
+Two studies comparing green and red snow using amplicon high-throughput sequencing (HTS) found they were caused by different species [@Lutz2015;@Terashima2017]; however it remains an open question whether green snow can develop into red snow.
 
@@
 Further research will be needed to determine the specific environmental factors driving community differences, but our field observations does provide some testable hypotheses.
-In forested areas, green cells would be shaded from damaging solar radiation, while above treeline pigmented cells of *Sanguina* and *Chloromonas* are likely protected by their astaxanthin pigment [@Gorton2003;Holzinger2006].
+In forested areas, green cells would be shaded from damaging solar radiation, while above treeline pigmented cells of *Sanguina* and *Chloromonas* are likely protected by their astaxanthin pigment [@Gorton2003;@Holzinger2006].
 This fits with our observation that green cells of *Chloromonas krienitzii* were hidden below the snow surface, and pigmented cells were more prevalent on the surface (Fig.
@@
 nivalis* [@Matsuzaki2019;@Prochazkova2017], *Chloromonas cf.
-brevispina* [@Nedbalova2008, @Matsuzaki2015], *Chloromonas krienitzii* [@Matsuzaki2015] *Chlainomonas rubra* [@Novis2008], or "Other" based on similarity to published photographs.
+brevispina* [@Nedbalova2008a;@Matsuzaki2015], *Chloromonas krienitzii* [@Matsuzaki2015] *Chlainomonas rubra* [@Novis2008], or "Other" based on similarity to published photographs.
 

### 2019-11-16 manuscript/figures.Rmd
update

--- 2019-11-02 manuscript/figures.Rmd
@@
 library(knitr)
-opts_chunk$set(fig.width=12, fig.height=8, fig.path='figs/', echo=FALSE, warning=FALSE, message=FALSE)
+library(tidyverse)
+library(here)
+library(grid)
+
+opts_chunk$set(fig.align= "center", echo=FALSE, warning=FALSE, message=FALSE)
 ```
@@
 ```{r input, include=F}
-# field <- knitr::include_graphics("temp.png")
-asvs <- include_graphics("figs/2_taxonomy/combined.png")
-communities <- include_graphics("figs/3_community/combined.png")
-# depth
+
+asv_similarity <- include_graphics(here("manuscript/figs/main/taxonomy/asvs2.png"))
+communities <- include_graphics(here("manuscript/figs/main/community/community_ggplot_polished_2.png"))
+unifrac <- include_graphics(here("manuscript/figs/main/uf_plot.png"))
+# depth <- include_graphics(here("manuscript/figs/depth_comparison/depth_fig.png"))
+
+field_cap <- "Representative photographs of snow algae.
+A.
+Red snow bloom above treeline at sample site G1.1.
+B.
+Microphotograph of red snow containing \\emph{Sanguina nivaloides} (Sn) and \\emph{Chlainomonas rubra} (Cr) cell morphologies.
+All scale bars 30 um, all microphotos under 63x objectives with DIC.
+C.
+Snow runnels form dendritic branching patterns on the snow surface, sample site S9.
+D.
+Microphotograph of orange snow from the surface of runnel containing \\emph{Chloromonas krienitzii} (Ck) and \\emph{Chloromonas cf.
+nivalis} (Cn).
+E.
+Bronze coloured snow spots below conifer canopy.
+F.
+Microphotograph of bronze snow containing \\emph{Chloromonas cf.
+brevispina} (Cb), \\emph{Chloromonas cf.
+nivalis} (Cn), and \\emph{Chlainomonas rubra} (Cr)."
+
+asv_cap <- "A.
+MDS clustering of \\emph{rbc}L ASVs.
+Color represents taxonomy, size proportional to relative abundance.
+Dotted ellipses show OTU clusters.
+Stress = 0.13.
+B.
+t-SNE clustering of snow algae \\emph{rbc}L ASVs.
+Perplexity = 30."
+
+heatmap_cap <- "Relative abundance heatmap of snow algae community composition using cell morphology, 18S, and \\emph{rbc}L.
+Samples on the x-axis are arranged from low to high elevation.
+Taxonomic designations for \\emph{rbc}L are based on \\emph{de novo} OTU clustering, while taxonomic assignments for 18S are reference-based.
+Samples lacking data for a particular dataset are shown as blank (e.g.
+G1.4 18S)."
+
+unifrac_cap <- "\\emph{rbc}L UniFrac similarity between snow algae samples. Points colored by elevation."
 ```
 
-```{r pressure, echo=FALSE, fig.cap="A caption", out.width = '100%'}
+```{r field, fig.cap = field_cap}
+include_graphics(here("manuscript/figs/main/field/field.png"))
+```
+
+```{r asvs, fig.cap = asv_cap, fig.width=5, fig.height=5}
+asv_similarity
+```
+
+```{r rel_abund, fig.cap = heatmap_cap}
 communities
@@
 
-Figure 1.
+```{r unifrac, fig.cap = unifrac_cap}
+unifrac
+```

### 2019-11-29 code/2018_sample_map.Rmd
red dots; field info table (save kable not working)

@@
 
+<style>
+table {
+  background-color: white !important;
+  color: black !important;
+}
+</style>
+<!-- format the table black and white -->
+
 ```{r libs, include=F}
@@
 library(ggmap)
+library(ggrepel)
+library(kableExtra)
+library(magick)
+library(formattable)
 
@@
 load(here::here("output/map/raw_google_map.rda")) # loads .rda called "algae_map"... not sure how avoid using.rda format, maybe png?
+```
+```{r}
+cities <- cities %>% 
+  filter(name == "Vancouver")
+```
+
+```{r}
+field_summary <-
+  field %>%
+  group_by(mountain) %>% 
+  drop_na(lat, lon) %>% 
+  arrange(date) %>% 
+  summarise(n=n(), 
+            mean_elev = mean(elev_m), 
+            first_visit=min(date) %>% format("%b %d"), 
+            last_visit= max(date) %>% format("%b %d"), 
+            n_visits= length(unique(date)),
+            lat=mean(lat),
+            lon=mean(lon),
+            date=min(date)) %>% 
+  arrange(date) %>% 
+  select(-date)
+
+# paste(field$date %>% month(label=TRUE), field$date %>%day())
 ```
@@
 full_map
+
+full_map_2 <-
+  ggmap(algae_map, extent = "device") +
+  geom_point(data = field_summary, aes(x=lon, y=lat),
+              size = 1, alpha=1, 
+              color = "red") +
+  geom_point(data = cities, shape = 15, color="white") +
+  geom_text(data = cities, aes(label = name), hjust = 1.1, color="white") #+
+  # geom_text_repel(data=field_summary,
+  #                 aes(x=lon, y=lat, label=mountain),
+  #                 size=3, vjust=-2, hjust=1) +
+  # plot transparent boxes for layers
+  # geom_label_repel(data=field_summary,
+  #            aes(label=mountain),
+  #            label.size = NA,
+  #            alpha=0.5, 
+  #             label.padding=.1, 
+  #             na.rm=TRUE,
+  #            seed=123) #+
+  #   # plot a second layer exactly the same, the labels for boxes
+  # geom_label_repel(data=field_summary,
+  #        aes(x=lon, y=lat, label=mountain), 
+  #        alpha=1, 
+  #        size=3,
+  #        segment.color="black",
+  #        segment.size = 1, 
+  #        seed=123)
+ # xlim = c(1, 20)
+full_map_2
+
+#missing enchantments point, not on map
+```
+
+```{r}
+# field_summary_kable <- 
+  field_summary %>% 
+  mutate(mean_elev = round(mean_elev,0),
+         last_visit = if_else(last_visit==first_visit,"",last_visit),
+         date_range= if_else(last_visit!="",paste(first_visit, last_visit, sep=" - "), first_visit)) %>% 
+  select(mountain, date_range,  mean_elev, n_visits,n) %>% 
+  dplyr::rename_all(str_to_title) %>%
+  dplyr::rename("Date"=Date_range,"Elevation"=Mean_elev, "N. visits"=N_visits, "Total samples"=N) %>% 
+  kable(booktabs=T, format = "html") %>%
+  kable_styling("striped",full_width = F) %>%
+  row_spec(0:14,color ="white",background ="black") %>% 
+  column_spec(1:5, width = "5em")
+
+# generated final via screenshot
+
+  # couldnt get to work
+  # tried 
+  # https://haozhu233.github.io/kableExtra/save_kable_and_as_image.html
+  # https://stackoverflow.com/questions/48834749/error-using-magick-r-to-import-pdf
+  # got image_read_pdf to work, still not authorized error for image_read
+  # https://stackoverflow.com/questions/42928765/convertnot-authorized-aaaa-error-constitute-c-readimage-453
+  # as_image()
+  # save_kable(file=paste0(getwd(),"/presentation/omics/field_tbl.pdf"))
+  # kable_as_image(filename=paste0(getwd(),"/presentation/omics/field_tbl.pdf"))
+field_summary_kable
 ```
@@
 
+# save output for omics presentation
+ggsave(filename = here::here("presentation/omics/map.png"))
+
 ```

### 2019-11-29 code/community.Rmd
dark theme for omics presentation

@@
         axis.title.x=element_blank())
+theme_set(dark_theme_light())
 ```
@@
 
-ggsave(here("manuscript/figs/main/community/community_ggplot.pdf"), plot=community_heatmap, width=7, height=4)
+# presentation output
+ggsave(here("presentation/omics/community_ggplot.pdf"), plot=community_heatmap, width=7, height=4)
 ```

### 2019-11-29 code/rbcl_asv_clustering.Rmd
changed object name
dark theme for omics presentation

@@
 
-conflicted::conflict_scout()
-conflicted::conflict_prefer("filter","dplyr")
+# conflicted::conflict_scout()
+# conflicted::conflict_prefer("filter","dplyr")
+```
+
+```{r}
+# theme_set(dark_theme_classic())
+theme_set(dark_theme_light())
 ```
@@
   labs(color="ASV taxonomic assignment", tag="B", title="t-SNE") +
-  ggthemes::theme_few() +
+  # ggthemes::theme_few() +
   scale_size(range = c(2, 12), guide="none")
@@
   labs(color="ASV taxonomic assignment", tag="A",title="MDS") +
-  ggthemes::theme_few() +
+  # ggthemes::theme_few() +
   scale_size(range = c(2, 12), guide="none")+
@@
 ```{r, fig.width=4, fig.height=6}
-asv_arranged <- gridExtra::grid.arrange(mds, tsne_no_legend, nrow=2)
+asv_arranged <- gridExtra::grid.arrange(mds, tsne_no_legend_nice_colors, nrow=2)
+```
+
+```{r, fig.width=7, fig.height=3.5}
+present_version <- gridExtra::grid.arrange(mds, tsne_no_legend_nice_colors, ncol=2)
 ```
@@
 
+# write black formatted versions for presentation
+ggsave(filename = here("presentation/omics/asvs.pdf"), plot=present_version, width = 7, height = 3.5)
+ggsave(filename = here("presentation/omics/asv_legend.pdf"), plot=legend,  height = 4, width = 2.2)
+
 # write out the trees
@@
 writeXStringSet(chlain_taxa_seqs, here("output/taxa_filtered_fastas/chlainomonas_idtaxa.fasta"))
+
 ```

### 2019-11-29 code/wrangle/tidy_field.R
added mtn names to field

@@ -67,12 +67,32 @@ field_sample_uid <- field_visit_num %>%
   mutate(alias = paste0(alias_1, ".", uid))
 field_sample_uid %>% distinct(sample_id, alias, date) %>% arrange(date) %>% view()
 
+# add column for mountain
+view(field_sample_uid)
+field_mountain_added <- field_sample_uid %>% 
+  mutate(mountain = case_when(mountain_letter=="A"~"Enchantments",
+                              mountain_letter=="B"~"Mt. Brew",
+                              mountain_letter=="F"~"Mt. Fromme",
+                              mountain_letter=="G"~"Opal Cone",
+                              mountain_letter=="H"~"Hollyburn",
+                              mountain_letter=="K"~"Sky Pilot",
+                              mountain_letter=="L"~"Washington Pass",
+                              mountain_letter=="M"~"St. Marks Summit",
+                              mountain_letter=="N"~"Mt. Rexford",
+                              mountain_letter=="P"~"Panorama Ridge",
+                              mountain_letter=="S"~"Mt. Seymour",
+                              mountain_letter=="T"~"Tricouni Pk.",
+                              mountain_letter=="W"~"Wedge Mtn.",
+                              mountain_letter=="X"~"Saxifrage Mtn."))
+
 # remove some of the unused columns
-field_removed_extra_cols <- field_sample_uid %>% 
-  select(sample_id, alias, date, lat, lon, elev_m, color_snow, habitat, substrate, depth_cm, dom_morph, have_seq)
+field_removed_extra_cols <- field_mountain_added %>% 
+  select(sample_id, alias, date, lat, lon, elev_m, color_snow, habitat, substrate, depth_cm, dom_morph, have_seq, mountain)
+
+view(field_removed_extra_cols)
 
 
 
 
 # write output ------------
-write_csv(field_removed_extra_cols, here("data/02_tidied/tidied_field.csv"))
+write_csv(field_removed_extra_cols, here::here("data/02_tidied/tidied_field.csv"))

### 2019-11-29 manuscript/mbb801/MBB801_figs.Rmd
added to final description for mbb801 figure project

@@
 
-## Fig. 2: Cell morphologies by elevation
+## Fig. 2: *rbc*L ASV clustering
 
 - intuitive layout
-  - elevation in y axis
-  - time on x-axis
-  - abundance is proportional to size
-  - different categories of cell are different colors
-
-## Fig. 3: *rbc*L ASV clustering
-
-- intuitive layout
-  - MDS ("traditional" clustering) on top, "newer" t-SNE clustering on bottom
+  - MDS ("traditional" clustering) on top, "sophisticated" t-SNE clustering on bottom
   - seperate genera on legend with space
@@
   - each point color distinguishable in t-SNE
+  - modified brewer palettes (Wong 2012)
 
-## Fig. 4: Heatmap of *rbc*L, 18S, and cell count relative abundance
+## Fig. 3: Heatmap of *rbc*L, 18S, and cell count relative abundance
 
 - intuitive layout
-  - elevation y axis
   - intuitive colors: 
-    - "terrain" colors for elevation
-    - actual color of the sampled snow 
     - consistant taxonomy colors; e.g. *Sanguina* is the same color in rbcL, 18S, and cell count
+- although I had initially wanted elevation to be on the y axis because it is intuitive, it makes sense to put taxonomy on y axis for readibility.
+The sample IDs are less important, and would be only referenced if someone wanted to cross reference this figure with another (e.g.
+fig 4, or in supp data), but the taxonomy is more important within this fig
 - because amplicon data is semi-quantative, using alpha (opacity) to represent relative abundance is appropriate and intuitive
@@
 
-## Fig. 5: *rbc*L uniFrac NMDS
+## Fig. 4: *rbc*L uniFrac NMDS
 

### 2019-12-20 manuscript/manuscript.Rmd
new title and abstract

@@
 ---
-title: "Contrasting snow algae communities along an elevational gradient"
-# distinct yet overlapping categories of snow algae community 
-# Classification of snow algae communities along an elevational gradient
-# Elevation influences snow algae community composition
-# 
+title: "Contrasting alpine and subalpine snow algae communities in British Columbia, Canada"
+# who grows in the snow? multi-hued snow algae communities
+# different coloured alpine and subalpine snows harbour constrasting snow algae communities
+# contrasting snow algae communities in di
+# not just dirty snow; alpine snow algae communities have different assemblages
+# green, orange, or red snow caused by distinct
+# snow algae communities causing colored snow in alpine and subalpine summer snow 
+# distinct snow algae communities across an elevational gradient
 author:
@@
 abstract: |
-  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate.
-The factors that determine when and where a snow algae bloom will form are poorly understood; fundamentally, we need to know what algae species are present in this poorly understood microbiome.
-We used a novel *rbc*L and 18S amplicon high-throughput sequencing (HTS) approach to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada.
-We found distinct communities at different elevations, with *Chloromonas* prevalent at lower elevations and *Sanguina* above treeline, while *Chlainomonas* was abundant across all elevations.
-Alpha diversity was variable; surprisingly, samples from highest elevations contained the highest diversity.
-We observed the snow algae *Chloromonas krienitzii* transition from green to orange over the course of the summer, and our observations suggest that different life stages of this species occupy distinct habitats above and below the snow surface.
-The snow algae microbiome is more diverse than previously thought, and future work must take into account variable community composition in snow algae blooms.
+  Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that reduce albedo and increase snow melt.
+Green, orange, and red patches of snow algae can harbour morphologically diverse communities, but despite their global ubiquity the community composition of snow algae blooms are poorly understood.
+We Illumina sequenced 18S and *rbcL* amplicons from 33 snow algae samples collected from a range of alpine and subalpine habitats in the Coast Range of British Columbia, Canada.
+Communities were dominated by algae closely related to *Sanguina* and *Chloromonas*, the latter including genetically distinct sub-groups including clusters matching *Chlainomonas* and several sub-groups that could represent potentially novel clades.
+Above treeline snow algae communities were dominated by *Sanguina*, while *Chloromonas krienitzii* was found in snow runnels near treeline, which transitioned from green to orange throughout the season.
+*Chlainomonas* was abundant in samples across all elevations.
+Our results suggest that while snow algae diversity at the genus level is relatively low much remains to be explored at the species level, and that communities of snow algae can differ greatly from patch to patch of snow.
 ---
@@
 
-### *Keywords*: snow algae rbcL 18S amplicon microbiome alpine community
+### *Keywords*: snow algae community microbiome amplicon rbcL 18S alpine
 

### 2019-12-21 manuscript/manuscript.Rmd
minor changes title and abstract

@@
 ---
-title: "Contrasting alpine and subalpine snow algae communities in British Columbia, Canada"
+title: "Contrasting snow algae communities in the mountains of British Columbia, Canada"
 # who grows in the snow? multi-hued snow algae communities
@@
 We Illumina sequenced 18S and *rbcL* amplicons from 33 snow algae samples collected from a range of alpine and subalpine habitats in the Coast Range of British Columbia, Canada.
-Communities were dominated by algae closely related to *Sanguina* and *Chloromonas*, the latter including genetically distinct sub-groups including clusters matching *Chlainomonas* and several sub-groups that could represent potentially novel clades.
-Above treeline snow algae communities were dominated by *Sanguina*, while *Chloromonas krienitzii* was found in snow runnels near treeline, which transitioned from green to orange throughout the season.
+Communities were dominated by algae closely related to *Sanguina* and *Chloromonas*, the latter including genetically distinct sub-groups of *Chlainomonas* and several potentially novel clades.
+Above tree line communities were dominated by *Sanguina*, and below tree line by *Chloromonas*.
 *Chlainomonas* was abundant in samples across all elevations.
-Our results suggest that while snow algae diversity at the genus level is relatively low much remains to be explored at the species level, and that communities of snow algae can differ greatly from patch to patch of snow.
+We also observed blooms of *Chloromonas krienitzii* transition from green to orange throughout the season.
+We found snow algae community composition was highly variable from patch to patch, even on the same mountain.
+Although relatively few genera dominated snow algae communities, we found high diversity at the species level.
 ---
@@
 
-Snow algae are cold-adapted microalgae that bloom in melting summer snow throughout alpine and polar regions [@Leya2013].
-Red snow algae blooms can cover vast areas of snowfield, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017].
+Snow algae are cold-adapted microalgae that bloom on melting summer snow throughout alpine and polar regions.
+Snow algae darken the snow surface and decrease albedo[@lutz_biogeography_2016;onuma_temporal_2019]---by one estimate, enough to substantially increase snow melt[@ganey_role_2017].
+Estimates of global coverage are lacking, but snow algae blooms can cover many square kilometers of glacier summer snow[@ganey_role_2017], are present on snow overlying sea ice[@gradinger_snow_1996], and are reported from every continent[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019].
+
+Sanger seq isolates
+
+Sanger seq field samples
+
+18S HTS
+
+<!-- re work this to follow the flow of my presentation -->
+Microscopy reveals a diversity of algae in patches of green, orange, or pink snow (Fig.
+1).
+Amplicon high-throughput sequencing (HTS) shows that most snow algae are typically Chlamydomonadales, although other orders are occasionally reported[@remias_two_2019;@lutz_integrated_2015;@soto_microbial_2020].
+Three genera are frequently found in colored snow.
+*Sanguina* are red or orange cells found across the Arctic, which was recently re-assigned to its own genus based on Sanger sequencing of field samples[@prochazkova_sanguina_2019].
+*Chloromonas* currently has 35 species listed on GenBank, 13 of which are snow algae[@matsuzaki_taxonomic_2019].
+Microscopy suggests *Chloromonas* is prevalent in forested sites[@hoham_distribution_1979,@nedbalova_ecology_2008], but recent 18S metabarcoding found *Chloromonas* was also widespread in Arctic red snow in low relative abundance[@lutz_biogeography_2016].
+The third genus *Chlainomonas* looks completely different from *Chloromonas*, but *rbcL* phylogeny suggests the two are closely related[@novis_two_2008].
+Both *Chlainomonas* and *Sanguina* are colored red by astaxanthin, but *Chlainomonas* cells are typically twice the diameter, with four flagella and a distinct cell wall[@remias_ecophysiology_2016] (Fig 2x).
+*Chlainomonas* and *Sanguina* remain uncultured[@prochazkova_ecophysiological_2018;@novis_two_2008].
+
+Microscopic identification is complicated by the multiple life stages typical of microalgae: the same species can look completely different between life stages; conversely, the same species can look nearly identical.
+The freshwater algae *Haematococcus pluvialis* undergoes mitosis as green cells, but stressors such as high light and nutrient starvation triggers formation of a dormant cyst with thick cell walls and astaxanthin pigment[@ota_carotenoid_2018;@kobayashi_light-independent_1997].
+Attempts to induce cyst development in snow algae remain unsuccessful; the bright pigments and ornamented cell walls of *Chloromonas* observed in the field bear little resemblance to the green cells in culture[@leya_response_2009].
+Amplicon studies have found that green snow is caused by different species than red snow[@lutz_integrated_2015;@terashima_microbial_2017;@soto_microbial_2020], so it remains unknown whether red snow species undergo a green life phase at all.
+
+Snow algae are the primary producer in a little-studied microbiome that includes protists, fungi, bacteria, and viruses[@naff_phylogeny_2013;@lutz_biogeography_2016;@hamilton_primary_2017;@terashima_microbial_2017;@hisakawa_metagenomic_2015].
+Despite nearly a century of microscopic study of red snow[@leya_snow_2013], how snow algae blooms develop remains a mystery.
+How do cells propagate in this harsh environment?
+What limits the extent of snow algae blooms?
+Critically, we lack basic understanding of snow algae diversity and life history.
+
+<!-- this from draft 1 of casey1, above is from MBB806 -->
  Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], which can threaten water supplies [@Staudinger2014] and worsen drought [@Barnhart2016].

### 2019-12-23 code/community.Rmd
stacked barplor with cellct, 18s, rbcL inprogress

@@
 library(ggrepel)
+library(grid)
+library(egg)
  library(gridExtra)
@@
 # specify factor order
-species_ord <- c("Chlainomonas A","Chlainomonas rubra", 
-                "Sanguina B", "Sanguina nivaloides",
-                "Chloromonas krienitzii","Chloromonas muramotoi C", "Chloromonas D", "Chloromonas cf. brevispina", "Chloromonas cf. nivalis","Chloromonas F",
-                "Other Chloromonas", "Chloromonas",
-                "Trebouxiophyceae G", "Trebouxiophyceae",
-                "Unknown E", "Green cell", "Other")
-
-# specify colors
-plot_colors <- c(rbcl_otu_colors,s18_heat_colors, cellct_colors)
+species_ord <- c("Chlainomonas A"="#984ea3","Chlainomonas rubra"="#984ea3", #purple
+                "Sanguina B"="#ff534a", "Sanguina nivaloides"="#ff534a", #red
+                "Chloromonas krienitzii"="#ff7f00","Chloromonas muramotoi C"="#173416", "Chloromonas D"="#265725", "Chloromonas cf.
+brevispina"="grey10", "Chloromonas cf.
+nivalis"="grey10","Chloromonas F"="#b25800", #greens
+                "Other Chloromonas"="#4daf4a", "Chloromonas"="#4daf4a", #light greens
+                "Trebouxiophyceae G"="#377eb8", "Trebouxiophyceae"="#377eb8", # blue
+                "Unknown E"="grey10", "Green cell"="grey10", "Other"="grey70") #greys
 
@@
 community_heatmap <- ggplot(combined, aes(x=alias %>% fct_reorder(elev_m), 
-                                          y = species %>% fct_relevel(species_ord))) + 
+                                          y = species %>% fct_relevel(names(species_ord)))) + 
   geom_tile(aes(alpha=rel_abund, 
-                fill=species %>% fct_relevel(species_ord) )) + 
+                fill=species %>% fct_relevel(names(species_ord) ))) + 
   facet_grid(rows = vars(marker %>% fct_relevel(c("Cell morphology","18S","rbcL"))), 
@@
   guides(alpha="none", fill="none") +
-  scale_fill_manual(values = plot_colors) +
+  scale_fill_manual(values = species_ord) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
@@
 community_heatmap
+```
+
+```{r rotated-stacked-bar, echo=F}
+# specify factor order, so Chlainomonas always on same side, etc
+species_colors <- c("Chlainomonas A"="#984ea3","Chlainomonas rubra"="#984ea3", #purple
+                "Sanguina B"="#ff534a", "Sanguina nivaloides"="#ff534a", #red
+                "Chloromonas krienitzii"="#ff7f00","Chloromonas muramotoi C"="#173416", "Chloromonas D"="#265725", "Chloromonas cf.
+brevispina"="forestgreen", "Chloromonas cf.
+nivalis"="lightgreen","Chloromonas F"="#b25800", #greens
+                "Other Chloromonas"="#4daf4a", "Chloromonas"="#4daf4a", #light greens
+                "Trebouxiophyceae G"="#377eb8", "Trebouxiophyceae"="#377eb8", # blue
+                "Unknown E"="grey40", "Green cell"="#b2d8b2", "Other"="grey70") #greys
+# backwards
+species_colors <- species_colors %>% rev()
+
+# plot
+stacked_bar <- ggplot(combined, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
+  geom_bar(aes(fill=species %>% fct_relevel(names(species_colors))), position="stack", stat="identity") +
+  coord_flip() +
+  facet_grid(cols = vars(marker %>% fct_relevel(c("Cell morphology","18S","rbcL"))), 
+             scales="free_y", space="free_y") +
+  theme_classic() +
+  labs(x="Sample ID", y="Relative abundance")+
+  scale_fill_manual(values = species_colors)
+
+stacked_bar +
+  theme(legend.position="none")
+```
+
+stacked bar legend
+```{r}
+legend <- stacked_bar %>% cowplot::get_legend()
+grid.newpage()
+grid.draw(legend)
+```
+
+dosent work in rmarkdown
+```{r, echo=F}
+gt <- ggplot_gtable(ggplot_build(stacked_bar))
+gtable_show_layout(gt)
+
+gt$widths[9] = 4*gt$widths[9]
+grid::grid.draw(gt)
 ```

### 2019-12-23 code/community.Rmd
change colors in stacked barplots

@@
 
-```{r rotated-stacked-bar, echo=F}
+```{r, fig.width=2, fig.height=2}
+chloromonas_color <- "#4daf4a" # green
+krienitzii_color <- "#ff7f00" # orange
+muramotoi_color <- "#d7d42e" # yellow
+# pick shades using https://www.color-hex.com/
+hohamii_and_D_color <- "#265725" # dark green
+group_B_and_F_color <- "#b25800" # brown
+chloromonas_colors <- c(chloromonas_color, krienitzii_color, muramotoi_color, hohamii_and_D_color,group_B_and_F_color)
+scales::show_col(chloromonas_colors)
+scales::show_col("#4aaf53")
+
+```
+
+```{r rotated-stacked-bar, echo=F, fig.width=7, fig.height=5}
+
 # specify factor order, so Chlainomonas always on same side, etc
@@
                 "Sanguina B"="#ff534a", "Sanguina nivaloides"="#ff534a", #red
-                "Chloromonas krienitzii"="#ff7f00","Chloromonas muramotoi C"="#173416", "Chloromonas D"="#265725", "Chloromonas cf.
-brevispina"="forestgreen", "Chloromonas cf.
-nivalis"="lightgreen","Chloromonas F"="#b25800", #greens
-                "Other Chloromonas"="#4daf4a", "Chloromonas"="#4daf4a", #light greens
-                "Trebouxiophyceae G"="#377eb8", "Trebouxiophyceae"="#377eb8", # blue
-                "Unknown E"="grey40", "Green cell"="#b2d8b2", "Other"="grey70") #greys
+                # Chloromonas: pastel greens / blues / yellows
+                "Chloromonas krienitzii"="#ff7f00","Chloromonas muramotoi C"="#87b1d4", "Chloromonas D"="#94cf92", "Chloromonas F"="#fdfd96", 
+                #    orange green                                      pale blu,                   blue green      yellow--only high elev
+                "Chloromonas cf. brevispina"="#80c786", "Chloromonas cf. nivalis"="#4aaf86", 
+                #                 lt green                            teal
+                "Other Chloromonas"="#4aaf53", "Chloromonas"="#4aaf53", 
+                # greyish light green
+                "Trebouxiophyceae G"="#173416", "Trebouxiophyceae"="#173416", # dark green, stand out
+                "Unknown E"="grey60", "Green cell"="#4aaf53", "Other"="grey30") #greys
+
+# teal #69c095
+# medium blue #6994c0
+
 # backwards
@@
   theme(legend.position="none")
+#18s: check
+# cellct: _
+# rbcl: _
 ```

### 2019-12-23 code/community.Rmd
final color tweak

@@
                 # Chloromonas: pastel greens / blues / yellows
-                "Chloromonas krienitzii"="#ff7f00","Chloromonas muramotoi C"="#87b1d4", "Chloromonas D"="#94cf92", "Chloromonas F"="#fdfd96", 
+                "Chloromonas krienitzii"="#ffa54c","Chloromonas muramotoi C"="#87b1d4", "Chloromonas D"="#94cf92", "Chloromonas F"="#fdfd96", 
                 #    orange green                                      pale blu,                   blue green      yellow--only high elev
@@
                 # greyish light green
-                "Trebouxiophyceae G"="#173416", "Trebouxiophyceae"="#173416", # dark green, stand out
-                "Unknown E"="grey60", "Green cell"="#4aaf53", "Other"="grey30") #greys
+                "Trebouxiophyceae G"="#6a2d2f", "Trebouxiophyceae"="#6a2d2f", # maroon, standout
+                "Unknown E"="grey60", "Green cell"="#4aaf53", "Other"="grey60") #greys
 

### 2019-12-23 manuscript/manuscript.Rmd
add to intro, made outline of hypothesis (comments)

@@
 Snow algae darken the snow surface and decrease albedo[@lutz_biogeography_2016;onuma_temporal_2019]---by one estimate, enough to substantially increase snow melt[@ganey_role_2017].
-Estimates of global coverage are lacking, but snow algae blooms can cover many square kilometers of glacier summer snow[@ganey_role_2017], are present on snow overlying sea ice[@gradinger_snow_1996], and are reported from every continent[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019].
+Estimates of global coverage are lacking, but snow algae blooms can cover many square kilometers of glacier summer snow[@ganey_role_2017], are present on snow overlying sea ice[@gradinger_snow_1996], and are reported from every continent[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019], yet their impact on biogeochemical cycles and food webs has not been explored.
+Snow algae has been observed throughout history from the ancient Romans[plinius_naturalis_1096] to Charles Darwin[@darwin_voyage_1959], yet despite nearly a century of microscopic study[@leya_snow_2013] we still lack the most basic understanding of snow algae blooms.
+How do algae colonise the surface of white snow in spring?
+How do cells propagate in this harsh environment?
+Are snow algae obligate symbionts of bacteria, like many other microalgae?[@seymour_zooming_2017] How do snow algae aquire nutrients?
+Central to answering these questions, we first need to define what species of algae cause snow algae blooms.
 
-Sanger seq isolates
+<!-- Sanger seq isolates -->
+<!-- Sanger seq field samples -->
+<!-- 18S HTS -->
 
-Sanger seq field samples
+<!-- rationale -->
+<!-- significance -->
+<!-- hypothesis -->
+<!-- methods -->
+<!-- expected results -->
+<!-- how this addresses the question -->
 
-18S HTS
+<!-- 1. are there distinct communities of snow algae? -->
+<!-- rationale: lit suggests chlaino, chloro, sanguina dominant, is there much overlap between the communities?
+Differences between high and low elevation in microscopy, are there genetic differences?
+This is an important first step before we measure habitat variables, will allow us to choose which variables to measure...
+-->
+<!-- significance. different community, different ecology and or life history. and the snow algae are the primary producers of the community. define the study system.  -->
+<!-- hypothesis: there are distinct communities of snow algae-->
+<!-- methods: cell count, 18s, and rbcL amplicon sequencing -->
+<!-- results: there is much overlap between communities, but many with dominant species.
+Some chloromonas, raphi only found with sanguina.
+Chlainomonas widespread.
+overall, no, there are not sharply delineated communities.
+-->
+<!-- new questions raised: what habitat differences are found with different communities?
+is bass becking applicable here?
+what ecological differences, life history differences characterise the different communities?
+-->
 
-<!-- re work this to follow the flow of my presentation -->
+Our knowledge of snow algae has deepened since the early days of morphology-based classification, yet still snow algae community composition at the species level remains unknown.
+Morphology can be misleading when identifying microalgae: the same species can look completely different at different life stages; conversely, different species can look nearly identical.
+
+The first efforts to use genetic sequencing to identify snow algae relied on Sanger sequencing of isolates. This approach has been fruitful *Chloromonas*[@matsuzaki_taxonomic_2019]
+
 Microscopy reveals a diversity of algae in patches of green, orange, or pink snow (Fig.
@@
 Snow algae are the primary producer in a little-studied microbiome that includes protists, fungi, bacteria, and viruses[@naff_phylogeny_2013;@lutz_biogeography_2016;@hamilton_primary_2017;@terashima_microbial_2017;@hisakawa_metagenomic_2015].
-Despite nearly a century of microscopic study of red snow[@leya_snow_2013], how snow algae blooms develop remains a mystery.
-How do cells propagate in this harsh environment?
-What limits the extent of snow algae blooms?
-Critically, we lack basic understanding of snow algae diversity and life history.
 

### 2019-12-23 manuscript/manuscript.Rmd
crunched out a rough sketch of intro...

@@
 Are snow algae obligate symbionts of bacteria, like many other microalgae?[@seymour_zooming_2017] How do snow algae aquire nutrients?
-Central to answering these questions, we first need to define what species of algae cause snow algae blooms.
+Central to answering these questions, we first need to define what species and communities of algae that cause snow algae blooms.
+
+What are snow algae?
+Leya defines snow algae as algae that can proliferate on the snow, not just algae that are present on the snow.
+Thus, cultured isolates taken from the snow surface are not necessarily snow algae by this definition, they could be not actively growing in the snow, and may not be the same species as the cells that are most abundant in snow samples.
+Many snow algae species are reported from Chlorophyceae, including the red cells of Sanguina and Chlainomonas, and Chloromonas, but snow algae in other families have been reported including Trebouxiophyceae, and Chrysophyceae.
+
+Lutz shows that all snow algae communities look the same in Arctic red snow, dominated by "uncultured Chlamydomonadaceae" with lesser amounts of Chloromonas and Trebouxiophyceae.
+The implication that red snow communities are similar is supported by Prochazkova, who through phylogenetic work suggest that the uncultured Chlamydomonadaceae of Lutz, and the Chlamydomonas nivalis of previous work should be regarded as a novel genus: Sanguina.
+Chlamydomonas nivalis haplotypes are globally distributed, possible leading to the false impression that snow algae communities are similar across the globe.
+
+However there are many other types of snow algae besides Sanguina, including field sequences from red snow that have not been assigned taxonomy [prochazkova].
+
+Although red snow algae are hypothesized to undergo a green life phase, no one has found green patches of snow with sequence matching red snow algae, nor cultured red snow algae in the lab.
+Field samples of green snow algae contain algae from taxa not known to produce red pigment, such as Trebouxiophyceae [lutz_omics], Chrysophyceae, or Chloromonas[soto, japan].
+The latter, Chloromonas, can produce secondary carotenoids[that leya paper, polyptera].
+There are  and patches of Bronze snow microscopically dominated by cells described as "Chloromonas nivalis" and "Chloromonas brevispina", found below treeline in microscopy surveys of the American southwest, and in the Giant Mtns of Czech Republic.
+
+There are a diversity of species of Chloromonas, which species are found where?
+Lutz and Terashima suggested that Chloromonas polyptera is globally distributed in Arctic, and in Japan, but 18S may not be sufficient to resolve closely related species, of which there are many of CHloromonas.
+
+Red snow can also be dominated by different communities in alpine areas, dominated by the large red quadriflagellate cells of Chlainomonas, which half the size of Sanguina, and although they are coloured by the same pigment are closely related to Chloromonas.
+Chlainomonas has not been detected in 18S amplicon sequencing studies of snow algae to date, whether it is found alongside other snow algae, or whether it is only found in relatively "pure" communities remains unknown.
+
+Several studies have found that Chlorophyceae is the main one, although Trebouxiophyceae, and Chrysophyceae also grow in snow, but no studies have found them growing in red snow.
+Whether these types of snow algae are as widespread as others
+
+Motility
+Secondary carotenoids
+
+is there overlap between the diferent types of snow algae communities?
 
@@
 
-<!-- 1. are there distinct communities of snow algae? -->
+<!-- are there distinct communities of snow algae? -->
 <!-- rationale: lit suggests chlaino, chloro, sanguina dominant, is there much overlap between the communities?
@@
 
-Our knowledge of snow algae has deepened since the early days of morphology-based classification, yet still snow algae community composition at the species level remains unknown.
-Morphology can be misleading when identifying microalgae: the same species can look completely different at different life stages; conversely, different species can look nearly identical.
-
-The first efforts to use genetic sequencing to identify snow algae relied on Sanger sequencing of isolates. This approach has been fruitful *Chloromonas*[@matsuzaki_taxonomic_2019]
-
 Microscopy reveals a diversity of algae in patches of green, orange, or pink snow (Fig.
@@
 Amplicon studies have found that green snow is caused by different species than red snow[@lutz_integrated_2015;@terashima_microbial_2017;@soto_microbial_2020], so it remains unknown whether red snow species undergo a green life phase at all.
-
-Snow algae are the primary producer in a little-studied microbiome that includes protists, fungi, bacteria, and viruses[@naff_phylogeny_2013;@lutz_biogeography_2016;@hamilton_primary_2017;@terashima_microbial_2017;@hisakawa_metagenomic_2015].
 

### 2019-12-23 code/community.Rmd
minor

@@
 # specify factor order, so Chlainomonas always on same side, etc
-species_colors <- c("Chlainomonas A"="#984ea3","Chlainomonas rubra"="#984ea3", #purple
+species_colors <- c("Chlainomonas A"="#ac71b5","Chlainomonas rubra"="#ac71b5", #purple
                 "Sanguina B"="#ff534a", "Sanguina nivaloides"="#ff534a", #red
@@
 # rbcl: _
+# red green colorblind similatur
 ```

### 2019-12-28 manuscript/manuscript.Rmd
re wrote intro

--- 2019-12-21 manuscript/manuscript.Rmd
@@
 ---
-title: "Contrasting snow algae communities in the mountains of British Columbia, Canada"
+title: "Overlapping snow algae communities in the alpine and subalpine zones of British Columbia"
 # who grows in the snow? multi-hued snow algae communities
@@
 We Illumina sequenced 18S and *rbcL* amplicons from 33 snow algae samples collected from a range of alpine and subalpine habitats in the Coast Range of British Columbia, Canada.
-Communities were dominated by algae closely related to *Sanguina* and *Chloromonas*, the latter including genetically distinct sub-groups of *Chlainomonas* and several potentially novel clades.
-Above tree line communities were dominated by *Sanguina*, and below tree line by *Chloromonas*.
+Communities were dominated by algae closely related to *Sanguina*, *Chloromonas*, and *Chlainomonas*, as well as potentially novel taxa.
+Above treeline communities were dominated by *Sanguina*, and below tree line by *Chloromonas*.
 *Chlainomonas* was abundant in samples across all elevations.
@@
 # Introduction {-}
+<!-- audience, problem -->
+Reduction in cryospheric albedo has become the focus of attention in recent years as an important feedback loop of global warming.
+Microbial growth on snow surface can reduce albedo, notably when snow algae form extensive red blooms on melting summer snow[@ganey_role_2017;@lutz_biogeography_2016;@onuma_temporal_2019].
+Snow algae blooms are found on every continent, and on snow overlying sea ice[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019;@gradinger_snow_1996].
+Despite their global ubiquity, we are only beginning to identify the various types of algae communities that cause snow algae blooms.
+<!-- why do we care if there are different communities as a climate scientist? maybe this should be a paper for snow phycologists? -->
+<!-- different communities could have different ecology and life history -->
+<!-- we need to define the community types to understand ecology of different blooms.  -->
 
-Snow algae are cold-adapted microalgae that bloom on melting summer snow throughout alpine and polar regions.
-Snow algae darken the snow surface and decrease albedo[@lutz_biogeography_2016;onuma_temporal_2019]---by one estimate, enough to substantially increase snow melt[@ganey_role_2017].
-Estimates of global coverage are lacking, but snow algae blooms can cover many square kilometers of glacier summer snow[@ganey_role_2017], are present on snow overlying sea ice[@gradinger_snow_1996], and are reported from every continent[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019].
+<!-- snow algae taxonomy -->
+Most snow algae are green algae (Chlorophyceae), although recent studies suggest that other families of algae can also cause snow algae blooms[@lutz_omics_2015;@soto_microbial_2020;@remias_two_2019].
+Within the Chlorophyceae *Sanguina nivaloides* is prevalent in red snow worldwide, but red snow also contains many other clades which have yet to be named[@prochazkova_sanguina_2019].
+Other Chlorophyceaen snow algae include *Chloromonas* and the closely-related *Chlainomonas*.
+Currently, *Sanguina* contains two species, *Chloromonas* fourteen, and *Chlainomonas* two[@matsuzaki_taxonomy_2019;@novis_two_2008].
+Only *Chloromonas* has been sucessfully cultivated in the lab.
 
-Sanger seq isolates
+<!-- microscopy suggests distinct communities -->
+Microscopy suggests that snow algae blooms are in many cases are dominated by single species.
+In the mountains of southwestern USA and Czech Republic red snow was prevelant above treeline, dominated by cells similar in appearance to *S.
+nivaloides*, while below treeline green or orange snow were prevalent dominated by cells similar to *Chloromonas*[@hoham_distribution_1979;nedbalova_ecology_2008].
+Red snow dominated by *Chlainomonas* type cells were reported from slushy snow overlying alpine lakes[@novis_two_2008;@remias_ecophysiology_2016].
+However, microscopic identification of snow algae is complicated by the fact that the same species can look completely different at different life stages.
+Isolates of *Chloromonas krienitzii* are green, bean-shaped biflagellates, but genetically identical isolated single-cells from field samples appear as orange spheres with thick cell walls, short spines, and no flagella[@matsuzaki_revisiting_2015].
+Conversely, different species can look nearly identical: recent phylogenetic work shows that four similar looking cells previously referred to as *Chloromonas cf.
+nivalis* actually contain multiple genetically distinct clades[@matsuzaki_taxonomic_2019].
 
-Sanger seq field samples
+<!-- but genetic comparison, most look the same? -->
+Although snow algae communities contain microscopically distinct snow algae community types, studies using 18S amplicon high-throughput sequencing (metabarcoding) suggests that snow algae communities are highly similar.
+33 red snow communities across the Arctic were remarkably similar: all were dominated by uncultured Chlamydomonadaceae, with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale*[@lutz_biogeography_2016].
+Another study using 18S metabarcoding of snow algae in the mountains of Japan found neighboring red and green snow patches with  highly overlapping communities, although the red snow snow contained one OTU not found in the green communities[@terashima_microbial_2017].
+However, side-by-side red and green snow in Greenland contained completely different communities, and the authors concluded that red and green snow were caused by a different species[@lutz_omics_2015].
 
-18S HTS
+Given how little is known about snow algae distribution and diversity, particularly in alpine habitats, our goal is to characterise the snow algae community structure in alpine and subalpine habitats.
+We surveyed snow algae communities from throughout the Coast Range of British Columbia, Canada using light microscopy and metabarcoding.
+18S is highly conserved, making it an ideal universal barcode for eukaryotes, but at lower taxonomic levels 18S is too similar to resolve closely related taxa.
+To complement our 18S sequencing, we also targeted the more differentiated *rbcL* coding for rubisco.
+By using three cross-referenced metrics of relative abundance, we were able to describe the community structure with high taxonomic resolution.
+We found communities above treeline contained a distinct set of taxa, and instead of sharply defined community types we found something more akin to a gradient, with varying relative abundances of different snow algae taxa.
 
-<!-- re work this to follow the flow of my presentation -->
-Microscopy reveals a diversity of algae in patches of green, orange, or pink snow (Fig.
-1).
-Amplicon high-throughput sequencing (HTS) shows that most snow algae are typically Chlamydomonadales, although other orders are occasionally reported[@remias_two_2019;@lutz_integrated_2015;@soto_microbial_2020].
-Three genera are frequently found in colored snow.
-*Sanguina* are red or orange cells found across the Arctic, which was recently re-assigned to its own genus based on Sanger sequencing of field samples[@prochazkova_sanguina_2019].
-*Chloromonas* currently has 35 species listed on GenBank, 13 of which are snow algae[@matsuzaki_taxonomic_2019].
-Microscopy suggests *Chloromonas* is prevalent in forested sites[@hoham_distribution_1979,@nedbalova_ecology_2008], but recent 18S metabarcoding found *Chloromonas* was also widespread in Arctic red snow in low relative abundance[@lutz_biogeography_2016].
-The third genus *Chlainomonas* looks completely different from *Chloromonas*, but *rbcL* phylogeny suggests the two are closely related[@novis_two_2008].
-Both *Chlainomonas* and *Sanguina* are colored red by astaxanthin, but *Chlainomonas* cells are typically twice the diameter, with four flagella and a distinct cell wall[@remias_ecophysiology_2016] (Fig 2x).
-*Chlainomonas* and *Sanguina* remain uncultured[@prochazkova_ecophysiological_2018;@novis_two_2008].
-
-Microscopic identification is complicated by the multiple life stages typical of microalgae: the same species can look completely different between life stages; conversely, the same species can look nearly identical.
-The freshwater algae *Haematococcus pluvialis* undergoes mitosis as green cells, but stressors such as high light and nutrient starvation triggers formation of a dormant cyst with thick cell walls and astaxanthin pigment[@ota_carotenoid_2018;@kobayashi_light-independent_1997].
-Attempts to induce cyst development in snow algae remain unsuccessful; the bright pigments and ornamented cell walls of *Chloromonas* observed in the field bear little resemblance to the green cells in culture[@leya_response_2009].
-Amplicon studies have found that green snow is caused by different species than red snow[@lutz_integrated_2015;@terashima_microbial_2017;@soto_microbial_2020], so it remains unknown whether red snow species undergo a green life phase at all.
-
-Snow algae are the primary producer in a little-studied microbiome that includes protists, fungi, bacteria, and viruses[@naff_phylogeny_2013;@lutz_biogeography_2016;@hamilton_primary_2017;@terashima_microbial_2017;@hisakawa_metagenomic_2015].
-Despite nearly a century of microscopic study of red snow[@leya_snow_2013], how snow algae blooms develop remains a mystery.
-How do cells propagate in this harsh environment?
-What limits the extent of snow algae blooms?
-Critically, we lack basic understanding of snow algae diversity and life history.
-
-<!-- this from draft 1 of casey1, above is from MBB806 -->
- Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], which can threaten water supplies [@Staudinger2014] and worsen drought [@Barnhart2016].
-Thus, snow algae could play an important role in snow melt regulation, but little is known about how these blooms form.
-We don't know how cells colonize the white snow to initiate the bloom, what triggers snow algae to divide and produce pigment, or why one patch of snow turns red while the neighboring snow remains white.
-Critically, we lack basic understanding of the species composition and diversity present in snow algae blooms, which is fundamental to resolving long-standing mysteries about snow algae bloom formation.
-
-Snow algae blooms can color snow red, orange, or green, and can one or contain multiple species of algae.
-Most snow algae blooms are caused by green algae of class Chlorophyceae [@Leya2013], although blooms caused by Trebouxiophyceae [@Lutz2015] or Chrysophyceae [@Remias2013;@Remias2019] have been reported.
-Among the Chlorophyceae, both *Sanguina nivaloides* [@Prochazkova2019a] or *Chlainomonas* [@Prochazkova2018c] produce the red pigment astaxanthin responsible for the red color of some snow algae blooms.
-Orange snow is reported to be caused by *Sanguina aurentia* [@Prochazkova2019a] and some species of *Chloromonas* [@Prochazkova2019a;@Remias2013a], while algae lacking secondary pigments color the snow green, such as *Chloromonas pichinchae* [@Hoham1975b].
-However, some species are known from only one or two samples, and the recently described *Chlainomonas* and *Sanguina* are poorly explored, with only two species in each.
-
-Snow algae community composition is largely unknown.
-Some studies batch sequencing samples, and microscopy suggests that blooms are dominated by a single species.
-The red cells of *Sanguina nivaloides* are widespread and cosmopolitan [@Prochazkova2019a;@Segawa2018], but 18S amplicon surveys of red snow from across the Arctic [@Lutz2016a] and Japan [@Terashima2017] show that that snow algae blooms typically contain several secondary species; in addition to the dominant red snow algae (likely *Sanguina*, as it had not been described at the time), these studies found lower relative abundance of *Chloromonas*, as well as *Raphidonema* in Arctic sites.
-33 red snow samples from across the Arctic were relatively homogeneous in community composition [@Lutz2016a], but genetic comparison of the morphologically diverse snow algae communities found in mountains from temperate latitudes [@Nedbalova2008a] is lacking.
-
-Despite over a century of study, no one has successfully cultured the snow algae that produce red secondary pigments.
-In model organism green algae *Chlamydomonas reinhardtii* cells produce secondary pigments in just one life stage, remaining green for much of their life cycle [@Salome2019], thus red snow algae could originate from green cells.
-Green snow is frequently found in waterlogged snow, sometimes alongside red snow, leading some to hypothesize that green snow could be an earlier life stage of red snow.
-Two studies comparing green and red snow using amplicon high-throughput sequencing (HTS) found they were caused by different species [@Lutz2015;@Terashima2017]; however it remains an open question whether green snow can develop into red snow.
-
-In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy and 18S and *rbc*L amplicon HTS.
-The *rbc*L gene (coding for the large subunit of Rubisco) specifically targets phototrophs, and is more differentiated than the traditional 18S ribosomal marker gene, thus allowing us to specifically target snow algae.
-Our goal was to see what species were present, and how community composition varies over elevation.
+<!--  This is an important first step before we measure habitat variables, will allow us to choose which variables to measure... -->
+<!-- significance. different community, different ecology and or life history. and the snow algae are the primary producers of the community. define the study system.  -->
+<!-- hypothesis: there are distinct communities of snow algae-->
+<!-- methods: cell count, 18s, and rbcL amplicon sequencing -->
+<!-- results: there is much overlap between communities, but many with dominant species.
+Some chloromonas, raphi only found with sanguina.
+Chlainomonas widespread.
+overall, no, there are not sharply delineated communities.
+-->
+<!-- new questions raised: what habitat differences are found with different communities?
+is bass becking applicable here?
+what ecological differences, life history differences characterise the different communities?
+-->
 
@@
 
-# Conclusions
+ Hypothetically, a positive feedback loop could permit snow algae blooms to propogate, in which snow algae melt snow, producing more liquid water that permits more snow algae growth[].
 
@@
 The resultant primers were rbcL369F (5’-GAA CGT GAC AAA TTA AAC AAA-3’) and rbcL870R (5’-ACC WGA YAD ACG WAG AGC TT-3’).
-To target 18S we used Euk1181 (5'-TTA ATT TGA CTC AAC RCG GG-3') and Euk1624 (5'-CGG GCG GTG TGT ACA AAG G-3') [@Wang2014].
+To target 18S we used Euk1181 (5'-TTA ATT TGA CTC AAC RCG GG-3') and Euk1624 (5'-CGG GCG GTG TGT ACA AAG G-3'), targeting the V7-V8 hypervariable regions[@Wang2014;@].
 

### 2019-12-28 manuscript/manuscript.Rmd
results and discussion rewrite for LQ

@@
     reference_docx: templates/word-template.docx
-  pdf_document: 
-    fig_caption: yes
-    number_sections: yes
+  # pdf_document: 
+  #   fig_caption: yes
+  #   number_sections: yes
 csl: templates/nature-communications.csl
-bibliography: casey1.bib
+bibliography: refs.bib
 abstract: |
   Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that reduce albedo and increase snow melt.
-Green, orange, and red patches of snow algae can harbour morphologically diverse communities, but despite their global ubiquity the community composition of snow algae blooms are poorly understood.
+Green, orange, and red patches of snow algae can harbor morphologically diverse communities, but despite their global ubiquity the community composition of snow algae blooms are poorly understood.
 We Illumina sequenced 18S and *rbcL* amplicons from 33 snow algae samples collected from a range of alpine and subalpine habitats in the Coast Range of British Columbia, Canada.
@@
 
-### *Run title*: Contrasting snow algae communities
+### *Run title*: Overlapping snow algae communities
 
@@
 Reduction in cryospheric albedo has become the focus of attention in recent years as an important feedback loop of global warming.
-Microbial growth on snow surface can reduce albedo, notably when snow algae form extensive red blooms on melting summer snow[@ganey_role_2017;@lutz_biogeography_2016;@onuma_temporal_2019].
-Snow algae blooms are found on every continent, and on snow overlying sea ice[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019;@gradinger_snow_1996].
-Despite their global ubiquity, we are only beginning to identify the various types of algae communities that cause snow algae blooms.
+Microbial growth on snow surface can reduce albedo, notably when snow algae form extensive blooms on melting summer snow[@ganey_role_2017;@lutz_biogeography_2016;@onuma_temporal_2019].
+Red snow algae blooms are found on every continent, and on snow overlying sea ice[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019;@gradinger_snow_1996].
+Despite their global ubiquity, the  we are only beginning to identify the various types of algae communities that cause snow algae blooms.
 <!-- why do we care if there are different communities as a climate scientist? maybe this should be a paper for snow phycologists? -->
@@
 <!-- we need to define the community types to understand ecology of different blooms.  -->
+<!-- it would be nice to know what the species are that form snow algae blooms, and whether different types of snow algae blooms exist.  -->
+<!-- understanding of the species that cause snow algae blooms is limited to  -->
 
 <!-- snow algae taxonomy -->
-Most snow algae are green algae (Chlorophyceae), although recent studies suggest that other families of algae can also cause snow algae blooms[@lutz_omics_2015;@soto_microbial_2020;@remias_two_2019].
-Within the Chlorophyceae *Sanguina nivaloides* is prevalent in red snow worldwide, but red snow also contains many other clades which have yet to be named[@prochazkova_sanguina_2019].
-Other Chlorophyceaen snow algae include *Chloromonas* and the closely-related *Chlainomonas*.
-Currently, *Sanguina* contains two species, *Chloromonas* fourteen, and *Chlainomonas* two[@matsuzaki_taxonomy_2019;@novis_two_2008].
-Only *Chloromonas* has been sucessfully cultivated in the lab.
+Most snow algae are green algae of class Chlorophyceae, although recent studies suggest that other classes of algae can also cause snow algae blooms[@lutz_integrated_2015;@soto_microbial_2020;@remias_two_2019].
+Within the Chlorophyceae *Sanguina nivaloides* is prevalent in red snow worldwide, but many red snow sequences fall into clades which have yet to be named[@prochazkova_sanguina_2019].
+Other Chlorophyceae include *Chloromonas* and the closely-related but morphologically distinct *Chlainomonas*.
+Currently just two species are described for *Sanguina*, while *Chloromonas* contains fourteen, and *Chlainomonas* also contains two[@matsuzaki_taxonomic_2019;@novis_two_2008].
+Of the three, only *Chloromonas* has been successfully cultivated in the lab.
 
@@
 Microscopy suggests that snow algae blooms are in many cases are dominated by single species.
-In the mountains of southwestern USA and Czech Republic red snow was prevelant above treeline, dominated by cells similar in appearance to *S.
-nivaloides*, while below treeline green or orange snow were prevalent dominated by cells similar to *Chloromonas*[@hoham_distribution_1979;nedbalova_ecology_2008].
+In the mountains of southwestern USA and Czech Republic red snow was prevalent above treeline, dominated by cells similar in appearance to *S.
+nivaloides*, while below treeline green or orange snow were prevalent dominated by cells similar to *Chloromonas*[@hoham_distribution_1979;@nedbalova_ecology_2008].
 Red snow dominated by *Chlainomonas* type cells were reported from slushy snow overlying alpine lakes[@novis_two_2008;@remias_ecophysiology_2016].
@@
 Another study using 18S metabarcoding of snow algae in the mountains of Japan found neighboring red and green snow patches with  highly overlapping communities, although the red snow snow contained one OTU not found in the green communities[@terashima_microbial_2017].
-However, side-by-side red and green snow in Greenland contained completely different communities, and the authors concluded that red and green snow were caused by a different species[@lutz_omics_2015].
+However, side-by-side red and green snow in Greenland contained completely different communities, and the authors concluded that red and green snow were caused by a different species[@lutz_integrated_2015].
 
-Given how little is known about snow algae distribution and diversity, particularly in alpine habitats, our goal is to characterise the snow algae community structure in alpine and subalpine habitats.
+Given how little is known about snow algae distribution and diversity, particularly in alpine habitats, our goal is to characterize the snow algae community structure in alpine and subalpine habitats.
 We surveyed snow algae communities from throughout the Coast Range of British Columbia, Canada using light microscopy and metabarcoding.
 18S is highly conserved, making it an ideal universal barcode for eukaryotes, but at lower taxonomic levels 18S is too similar to resolve closely related taxa.
-To complement our 18S sequencing, we also targeted the more differentiated *rbcL* coding for rubisco.
+To complement our 18S sequencing, we also targeted the more differentiated *rbcL* gene, coding for rubisco, thus only targeting the photosynthetic community.
 By using three cross-referenced metrics of relative abundance, we were able to describe the community structure with high taxonomic resolution.
@@
 is bass becking applicable here?
-what ecological differences, life history differences characterise the different communities?
+what ecological differences, life history differences characterize the different communities?
 -->
@@
 
-## Field and microscopy observations of distinct snow algae bloom types
+We collected 309 snow algae samples from 13 mountains throughout the summer of 2018.
+We first detected snow algae on May 18, and we continued to sample our low elevation sites at Seymour (S) and Hollyburn (H) until the snow had completely melted out by early July.
+We did not detect snow algae above treeline until June 20, and snow persisted at higher elevations throughout the entire summer.
 
-We observed pink and green patches of snow algae in forested sites beginning on May 14.
-Snow algae was widespread in our study area, and we collected 309 samples from 33 dates and 13 mountains over the course of the summer (S1, S2).
-We observed an elevational shift in snow algae throughout the season: forested sites bloomed earlier in the season and melted out sooner, while high elevation sites did not bloom until later in the season, but persisted until the first snowfall of winter 2019 (S2).
+We found morphologically distinct snow algae communities in different habitats.
+In shaded, forested sites we often found green or orange patches of snow, containing green or orange ovate cells with short spines resembling *Chloromonas cf.
+brevispina* (Fig.
+1c).
+In forest clearings we observed low-angle linear drainage channels or runnels containing high concentrations of snow algae.
+In May and early June these runnels contained green snow hidden 2 to 5 cm below the white snow surface, and on subsequent visits these same sites contained orange snow at the surface, with morphologies resembling *Chloromonas krienitzii* (Fig 1b).
+Red snow was prevalent above treeline, dominated by red cells similar in appearance to *Sanguina nivaloides*, although we also occasionally observed green snow patches above treeline.
+We found red snow dominated by *Chlainomonas rubra* morphologies in unshaded sites at all elevations.
 
-We observed some consistent patterns between cell morphologies and field site characteristics.
-In forest sites, orange or green snow was typically dominated by green or yellow-orange ovate cells with short spines (*Chloromonas cf.
-brevispina*; Fig.
+We chose 33 snow algae samples for 18S and *rbcL* amplicon Illumina sequencing.
+Both libraries were dominated by amplicon sequence variants (ASVs) that were assigned to Chlorophyta.
+Our 18S library contained 67 ASVs that were assigned to Chlorophyta and 7 ASVs to Ochrophyta, while our *rbcL* library contained 642 ASVs, of which 603 were assigned to Chlorophyta.
+The top genera were *Chloromonas*, *Chlainomonas*, and *Sanguina*, but most ASVs were not assigned to genus level (Fig.
+2).
+We used t-SNE to generate operational taxonomic units (OTUs), and found many OTUs that did not correspond to any known taxa on GenBank: several OTUs related to *Chloromonas*, and one Chlamydomonadaceae OTU that was only assigned to the species level.
+
+Above and below treeline communities contained some unique taxa, but most taxa did not exhibit a clear elevational trend (Fig.
+3).
+*Sanguina* was the dominant taxa in most samples taken from above treeline.
+Additionally, high elevation sites contained several taxa not found at low elevations: an unannotated OTU of *Chloromonas*, and ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1) were found at the three highest elevation sites.
+Green snow from alpine sites contained *Chloromonas* OTUs that were also found in sites below treeline.
+Low elevation sites contained *Chloromonas krienitzii*, but many sites were dominated by taxa that were found at all elevations, such as *Chlainomonas*.
+
+Cell counts, 18S, and *rbcL* datasets were generally consistent, but we did notice some discrepancies.
+*Chlainomonas* was more prevalent in our *rbcL* library than in cell counts, while *Sanguina* was underrepresented in *rbcL* compared to 18S and cell counts.
+We did not detect *Chlainomonas* in 18S.
+
+At runnel sites on Hollyburn and Seymour we observed snow algae blooms that changed from green to orange over the course of the summer (Fig.
 1).
-In snow runnels in forest clearings we observed a shift over repeated site visits; on initial visits snow runnels contained green snow hidden from sight 2 - 5 cm below the surface (Fig.
-1b, S4_), on subsequent visits the surface of these runnels was orange and contained cell morphologies resembling *Chloromonas krienitzii* (Fig 1e, S3_, S4_).
-Red or pink snow was mostly found in open areas with high sunlight, and contained cells similar in appearance to *Sanguina nivaloides* or *Chlainomonas rubra*, in many sites mixed in with green cells of variable morphology (Fig 1).
-*Sanguina nivaloides* morphologies were primarily at high elevation sites, although we did observe several blooms dominated by this morphospecies in shaded forest sites.
-
-## Diverse *rbc*L reveals snow algae taxonomy with high-resolution
-
-Samples were dominated by green algae (Chlorophytes), primarily the genera *Chloromonas*, *Chlainomonas*, and *Sanguina* (Fig.
-2).
-Trebouxiophyceae (Chlorophytes) such as *Raphidonema* were present in low abundance (Fig.
-2), as well as low abundance of Ochrophytes (primarily *Hydrurus*) at several low elevation sites (S5).
-One *rbc*L OTU defied taxonomic classification, with a best BLAST score of 88%, and top hits in both Trebouxiophyceae and Chlorophyceae (Algae H, Fig.
-2).
-
-Our taxa-specific *rbc*L primers showed considerable variation within *Chloromonas* and the closely-related *Chlainomonas* (Fig 2, S6).
-We found several distinct clusters among *Chloromonas* amplicon sequence variants (ASVs) not matching any known species on GenBank (I, G, Fig.
-2), but the majority of *Chloromonas* ASVs did not form distinct clusters (Other *Chloromonas*, Fig.
-2).
-
-Our three relative abundance data sets were generally consistent.
-18S, *rbc*L, and cell count data were correlated for *Sanguina* (stats).
-*Chloromonas* and *Chlainomonas* cell counts were correlated with *rbc*L (stats, stats); however, 18S classified samples dominated by *Chlainomonas* as containing *Chloromonas* (S5), likely due to high target region similarity (S7).
-However, in some samples cell morphology was a poor predictor of taxonomic composition.
-For example, we were surprised to find sample X1.1 contained few *Sanguina* reads (Fig.
-3b, S5b) despite being dominated by red cells similar of similar morphology to *Sanguina* (Fig.
-1g, S4_).
-
-## Snow algae community composition varies with date and elevation
-
-We found snow algae community composition was strongly influenced by elevation across all three datasets: *rbc*L, 18S, and cell count.
-*Sanguina* was prevalent at higher elevations, and *Chloromonas* dominant at lower elevations (Fig.
-1g, Fig.
-3, S5).
-Low elevation sites were compositionally similar, while high elevation samples showed inter-site variability (UniFrac Fig.
-3a, S8).
-Lower elevation sites were dominated by *Chloromonas krienitzii* and *Chlainomonas rubra*, while *Sanguina nivaloides* was only abundant in samples from above treeline (Fig.
-3b), along with one OTU of *Chloromonas*, and *Raphidonema* (H, I, Fig.
-3b).
-Several sites above treeline were more similar to low elevation sites, including patches of green snow above treeline dominated by *Chloromonas* (N1.5, G1.4, Fig.
-3) and red snow patches dominated by *Chlainomonas* (G1.2, P1.9, Fig.
-3).
-
-*Chlainomonas* was abundant across all elevations.
-Some samples were nearly completely dominated by *Chlainomonas*, but it was widespread in samples containing *Sanguina* and *Chloromonas krienitzii* as well (Fig.
-3b).
-
-## A green snow algae turning orange at the surface
-
-Snow algae blooms dominated by *Chloromonas krienitzii* formed in low angle runnels, often in clearings at elevations near treeline.
-Based on our observation that the subsurface snow was green, and surface snow turned orange through the course of the season we sampled both orange and green snow to see if they were the same species (Fig.
-4a).
-The surface samples contained primarily orange cells resembling *Chloromonas krienitzii*, while the subsurface samples contained green cells including biflagellates and polar quadriflagellates (Fig.
-4b, c).
-Both surface and subsurface samples were dominated by *Chloromonas krienitzii* reads (Fig.
-4d).
-Unexpectedly, we found high relative abundance of *Chlainomonas* in surface samples, and in one subsurface sample, which we did not detect in cell counts.
+We sequenced both green subsurface and orange surface snow at three different sites, and found both were dominated by *Chloromonas krienitzii*, although surface samples also contained some *Chlainomonas* (Supplementary _).
 
@@
 
-We found snow algae communities were highly variable from site to site, with communities at lower elevation dominated by *Chloromonas*, and higher elevation sites dominated by *Sanguina*.
-Microscopy studies of snow algae in USA and Slovakia found similar results, with *Chloromonas* is prevalent in forested areas and *Chlamydomonas cf.
-nivalis* above treeline [@Hoham1979a;@Nedbalova2008a].
-Our cell counts, 18S, and *rbc*L data support this finding.
-
-The obvious question is: what environmental factors driving those differences in community composition?
-Further research will be needed to determine the specific environmental factors driving community differences, but our field observations does provide some testable hypotheses.
-In forested areas, green cells would be shaded from damaging solar radiation, while above treeline pigmented cells of *Sanguina* and *Chloromonas* are likely protected by their astaxanthin pigment [@Gorton2003;@Holzinger2006].
-This fits with our observation that green cells of *Chloromonas krienitzii* were hidden below the snow surface, and pigmented cells were more prevalent on the surface (Fig.
-4); possibly pigment production is triggered as snow level melts lower, and cells are exposed to higher light at the surface.
-However, this does not explain why *Chloromonas krienitzii* was not found at higher elevations: our *C.
-krienitzii* dominated samples were taken from sites that were ~100 m from the nearest tree, and were not notably shaded by topography in any way.
-Anecdotally, we observed *C.
-krienitzii* developed over running water, so their distribution could be linked to the presence of streams below the snowpack.
-
-Our results show that *Chlainomonas* is an important, hitherto unrecognized component of the snow algae microbiome.
-Prior to this study, *Chlainomonas* were thought to only bloom in highly saturated, waterlogged snow overlying mountain lakes [@Novis2008;@Prochazkova2018c], but we found *Chlainomonas* at most sites we sampled.
-Only one site high in *Chlainomonas* was near water (S9), the other *Chlainomonas* dominated sites were not near water, nor was the snow noticeably mushier than the surrounding snow.
-This does not rule out the possibility that these sites contained high moisture at some point; further sampling is needed to determine its habitat distribution, but the broad distribution of *Chlainomonas* in most samples suggests that this taxa may be an important component of the red snow microbiome.
-
-In contrast to typical alpine ecosystems with lower diversity at higher elevations, we observed the highest diversity at the highest elevation sites (W1.1, W1.2, P1.5, S8).
-These sites were our only samples collected on glaciers, and were also collected latest in the summer.
-These samples contained the highest proportion of *Raphidonema* (Trebouxiophyceae I, Fig.
-3), which is phylogenetically distant from Chlorophyceae, and thus could contribute heavily to phylogenetic diversity metrics, however Shannon and Simpson diversity was also highest for these three samples, which only took into account the evenness of OTU abundance, so the presence of *Raphidonema* alone cannot explain the even spread of OTUs at these sites.
-@Stibal2005 suggest that *Raphidonema* is a slow-growing opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow.
-Perhaps it's slow growth in snow means it is only found later in the season; or sites late in the season were closer to bare ground that seeded populations of this algae.
-
-Our use of cross-referenced 18S, *rbc*L, and cell count data uniquely allowed us to compare our findings between datasets.
-*Sanguina* was highly correlated between all three.
-One notable difference was that no reads were assigned to *Chlainomonas* from our 18S library, despite this genera being highly abundant in *rbc*L.
-Samples containing primarily *Chlainomonas* in our *rbc*L and cell count data were marked as containing only *Chloromonas* in 18S.
-At the time of publication, only two 18S *Chlainomonas* were present on GenBank (MF803743.1, MF803745.1); possibly our *Chlainomonas* sequences are different enough from the reference that they were mislabelled as *Chloromonas*, especially likely given how closely related the two genera are.
-Samples dominated by *Sanguina* in our 18S library contained lower abundance in *rbc*L, which is likely due to primer bias: because our custom *rbc*L primers were designed to target *Chloromonas* and *Chlainomonas* (*Sanguina* not being available at the time), this likely led to a over-representation of these taxa in our *rbc*L library.
+We found that the alpine snow algae community can vary widely from site to site.
+Different taxa were dominant at different elevations, with alpine sites dominated several distinct taxa, most notably *Sanguina*.
+*Chloromonas krienitzii* was dominant in mid-elevation runnels appearing as green motile cells below the snow surface, and later in the season developing orange pigment at the surface.
+*Chloromonas* and *Chlainomonas* were abundant across all elevations, including many taxa that are not represented in GenBank.
 
 Our findings highlight the remaining unexplored diversity in the snow algae microbiome.
-One abundant OTU of algae, "Algae H" is widespread and abundant, but poorly matches all *rbc*L GenBank sequences.
-However, out 18S sequences did not detect any such cluster.
-Non-specific priming on an unsequenced gene is one possibility, another is that this represents a novel clade of snow algae, or perhaps an endosymbiotic algae within the ciliates that were abundant in many of our samples.
-Interestingly, the sample with the highest relative abundance of this OTU (X1.1) had red cells that were visually distinct from other *Sanguina*, with thicker cell walls that were slightly less than perfectly spherical in some cells (S4); clearly, further investigation is needed.
+Most of the diversity was only detected in *rbcL*: samples enriched in Chlamydomonadaceae E did not correspond to anything we detected in cell counts and 18S.
+One possibility is that this OTU represents a novel taxa of algae whose 18S is highly similar to Chloromonas.
+Another possibility is primer bias: perhaps this OTU was over-represented in *rbcL* and underrepresented in 18S.
 
- Hypothetically, a positive feedback loop could permit snow algae blooms to propogate, in which snow algae melt snow, producing more liquid water that permits more snow algae growth[].
+Previous studies suggested that *Chlainomonas* is restricted thought to waterlogged snow overlying mountain lakes, but our results suggest that it is widespread and abundant in all types of snow algae blooms[@novis_two_2008;@prochazkova_ecophysiological_2018].
+We did not detect any association with a particular habitat: only one site dominated by *Chlainomonas* was near water (S9), but other *Chlainomonas* sites were not notably wetter than the surrounding snow.
 
-Our side by side comparison of snow algae communities shows that they are more diverse than previously thought, and suggests that some snow algae may be adapted to specific micro-habitats such as shade and moisture within the melting snowpack.
-As the different snow algae species may have varying habitat preferences, different taxa could have different life cycles that permit them to survive in their niche, thus future work to understand the biological mechanisms of snow algae blooms should take community composition into account.
-Different communities of snow algae will likely affect albedo and local snowpack chemistry in different ways [@Lutz2015], and so different snow algae communities could impact the microbiome and greater ecosystem in unique ways.
+Although 18S, *rbcL*, and cell counts were generally consistent, we observed some discrepancies, notably *Sanguina* *rbcL* was underrepresented relative to 18S (Fig.
+3).
+Primer bias could partly explain this discrepancy, because we designed our *rbcL* primers off of *Chloromonas* sequences, as *Sanguina* reference data was not available at the time.
+Additionally, we observed that *Sanguina* cell morphologies were more resistant to lysis than other cell types in microscopy following lysis.
+*Chlainomonas* was overrepresented in *rbcL* relative to cell counts, possibly due to primer bias again, or higher rRNA gene copy number typical of larger cell volumes[@fu_single_2017].
+None of our 18S ASVs returned BLAST results for *Chlainomonas*, despite there being two representative GenBank sequences in our reference database (MF803743.1, MF803745.1).
+Possibly the *Chlainomonas* 18S sequence in our target region (V7-V8) is not different enough from closely-related *Chloromonas* to be detectable in this survey.
+
+Species composition is likely governed, at least in part, by different site habitat characteristics.
+The absence of *Sanguina* from low elevation sites could be indicative of habitat specificity: perhaps it is outcompeted at low elevation sites by taxa that are better adapted to the low light of the shade.
+Given the prevalence of airborne microalgae, it seems less likely that *Sanguina* is restricted in it's dispersal capabilities[@tesson_airborne_2016].
+Perhaps *Chloromonas krienitzii* dominates in low-angle runnels where moisture is sufficient; alternatively, we observed that many runnels overlay streams, which could relate to the dispersal mechanism of this species.
+We found *Raphidonema* only in high elevation sites that were sampled late in the growing season.
+@stibal_growth_2005 suggests that this is a slow-growing opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow.
+Perhaps sites late in the season were closer to bare ground containing source populations of this algae; alternately, perhaps this algae is slower to grow in cold temperatures on the snow, and is outcompeted by the better adapted *Sanguina*.
+
+Characterization of the snow algae communities allows us to define the study system, and serves as a foundation for understanding the ecology and life history of this globally significant microbiome.
+[...]
 
@@
 
-We assigned taxonomy for both *rbc*L and 18S ASVs using IDTaxa with threshold set to 50 [@Murali2018].
+We assigned taxonomy for both *rbcL* and 18S ASVs using IDTaxa with threshold set to 50 [@Murali2018].
 We made custom reference databases for each to include snow algae GenBank sequences.

### 2019-12-28 code/misc/top_treb.R
check out the composition of the trebouxiophyceae

@@ -1,0 +1,87 @@
+# what species of trebouxiophyceae in hi elev samples?
+
+library(tidyverse)
+library(here)
+library(DECIPHER)
+
+# rbcl
+rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
+taxonomy <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
+rbcl_key <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
+
+# 18S
+s18_rel_abund <- read_csv(here("data/03_filtered/18s/filtered_cleaned_18s_rel_abund.csv"))
+s18_taxonomy <- read_csv(here("data/02_tidied/tidied_silva_18s_assignments.csv"))
+s18_key <- read_csv(here("data/asv_seq_data/asv_seq_key_18s.csv"))
+
+#otus
+rbcl_otus <- read_csv(here("output/asv_otu_key.csv"))
+
+# blast the high elevation rbcl trebouxiophyceae
+high_samples <- c("wed18.01", "wed18.02", "pan18.01")
+
+
+treb_asvs <- taxonomy %>% 
+  filter(class == "Trebouxiophyceae") %>% 
+  pull(asv_id)
+
+my_samples_treb <- rel_abund %>% 
+  filter(sample_id %in% high_samples,
+         asv_id %in% treb_asvs,
+         rel_abund > 0.01) %>% 
+  left_join(taxonomy, by="asv_id")
+
+view(my_samples_treb)
+
+top_asvs <- my_samples_treb %>% 
+  pull(asv_id)
+
+rbcl_key %>% 
+  filter(asv_id %in% top_asvs) %>% 
+  deframe() %>% 
+  DNAStringSet() %>% 
+  writeXStringSet(filepath = here("data/asv_seq_data/etc/treb.fasta"))
+# blast the seqs
+# raphidonema longiseta and nivale are top two for these
+
+
+
+# ------------------
+# 18s
+
+s18_taxonomy %>% 
+  view()
+
+treb_tax <- s18_taxonomy %>% 
+  filter(class %in% c("Trebouxiophyceae", "Chrysophyceae")) %>% 
+  pull(asv_id)
+
+
+s18_rel_abund %>% 
+  filter(asv_id %in% treb_tax) %>% 
+  left_join(s18_taxonomy, by="asv_id") %>% 
[…]

### 2019-12-28 code/community.Rmd
minor

@@
 
-stacked_bar +
+stack_no_legend <- stacked_bar +
   theme(legend.position="none")
@@
 ```{r}
-legend <- stacked_bar %>% cowplot::get_legend()
-grid.newpage()
-grid.draw(legend)
+legend <- stacked_bar %>% 
+  cowplot::get_legend() %>% 
+  ggpubr::as_ggplot()
+legend
 ```
@@
 
+ggsave(here::here("manuscript/figs/stack.pdf"), plot=stack_no_legend, width=7, height=5)
+ggsave(here::here("manuscript/figs/stack_legend.pdf"), plot=legend, width=7, height=5)
+
 # presentation output

### 2020-01-07 manuscript/manuscript.Rmd
LQ comments back, redo abstract, p1 intro

@@
 ---
-title: "Overlapping snow algae communities in the alpine and subalpine zones of British Columbia"
+title: "Snow algae in the alpine and subalpine zones of British Columbia"
 # who grows in the snow? multi-hued snow algae communities
@@
   Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that reduce albedo and increase snow melt.
-Green, orange, and red patches of snow algae can harbor morphologically diverse communities, but despite their global ubiquity the community composition of snow algae blooms are poorly understood.
+Green, orange, and red patches of snow algae can harbor morphologically diverse algae, but despite their global ubiquity the algal species composition of blooms is largely unknown.
 We Illumina sequenced 18S and *rbcL* amplicons from 33 snow algae samples collected from a range of alpine and subalpine habitats in the Coast Range of British Columbia, Canada.
-Communities were dominated by algae closely related to *Sanguina*, *Chloromonas*, and *Chlainomonas*, as well as potentially novel taxa.
-Above treeline communities were dominated by *Sanguina*, and below tree line by *Chloromonas*.
-*Chlainomonas* was abundant in samples across all elevations.
-We also observed blooms of *Chloromonas krienitzii* transition from green to orange throughout the season.
-We found snow algae community composition was highly variable from patch to patch, even on the same mountain.
-Although relatively few genera dominated snow algae communities, we found high diversity at the species level.
+Predominant species were in, or closely related to, the genera *Sanguina*, *Chloromonas*, and *Chlainomonas*.
+The most common genus above treeline were *Sanguina*, while *Chloromonas* were dominant below treeline, and *Chlainomonas* were abundant in samples across all elevations.
+We found species composition was highly variable from patch to patch, even on the same mountain.
+Although snow algae in the Coast Range are represented by the same relatively few genera that dominate most reports of snow algal blooms around the globe, we found high diversity at the species level.
 ---
@@
 
-### *Keywords*: snow algae community microbiome amplicon rbcL 18S alpine
+### *Keywords*: snow algae, microbiome, amplicon, *rbcL*, 18S, alpine, Illumina, British Columbia
 
-### *Run title*: Overlapping snow algae communities
+### *Run title*: Alpine snow algae
 
@@
 <!-- audience, problem -->
-Reduction in cryospheric albedo has become the focus of attention in recent years as an important feedback loop of global warming.
+Each summer in polar and alpine snowfields worldwide, vast areas of snow surface are colored red by snow algae blooms.
+These extremophilic single-celled algae photosynthesize in near-freezing temperatures under high solar irradiance, and are the primary producers in a poorly-understood microbiome that includes bacteria, fungi, ciliates, and rotifers[@gorton_ultraviolet_2003;@lutz_biogeography_2016].
+Snow algae blooms occur on every continent and Arctic sea ice[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019;@gradinger_snow_1996]; in Alaska, snow algae spectral signatures were detected in one third of a 2,000 km^2^ icefield[@ganey_role_2017].
+In recent years snow algae have recieved attention for their impact on snow surface albedo: initial estimates suggests snow algae could increase snow melt rate by as much as 17%, potentially regulating spring snow melt and impacting global albedo[@ganey_role_2017].
+Despite their global ubiquity, we are only beginning to identify the algae that are the primary producers in these complex microbiomes.
+
+<!-- Reduction in cryospheric albedo has become the focus of attention in recent years as an important feedback loop of global warming.
 Microbial growth on snow surface can reduce albedo, notably when snow algae form extensive blooms on melting summer snow[@ganey_role_2017;@lutz_biogeography_2016;@onuma_temporal_2019].
-Red snow algae blooms are found on every continent, and on snow overlying sea ice[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019;@gradinger_snow_1996].
+.
 Despite their global ubiquity, the  we are only beginning to identify the various types of algae communities that cause snow algae blooms.
+-->
 <!-- why do we care if there are different communities as a climate scientist? maybe this should be a paper for snow phycologists? -->
@@
 <!-- snow algae taxonomy -->
-Most snow algae are green algae of class Chlorophyceae, although recent studies suggest that other classes of algae can also cause snow algae blooms[@lutz_integrated_2015;@soto_microbial_2020;@remias_two_2019].
-Within the Chlorophyceae *Sanguina nivaloides* is prevalent in red snow worldwide, but many red snow sequences fall into clades which have yet to be named[@prochazkova_sanguina_2019].
-Other Chlorophyceae include *Chloromonas* and the closely-related but morphologically distinct *Chlainomonas*.
-Currently just two species are described for *Sanguina*, while *Chloromonas* contains fourteen, and *Chlainomonas* also contains two[@matsuzaki_taxonomic_2019;@novis_two_2008].
-Of the three, only *Chloromonas* has been successfully cultivated in the lab.
+
+<!-- need to introduce the idea of patchiness, and idea of different colored snow -->
+Snow algae blooms can contain a mix of species, many of whose taxonomic identity is not well established.
+Green algae (Chlorophyceae) are often prevalent: *Sanguina nivaloides* is found in red snow samples worldwide[@prochazkova_sanguina_2019].
+Currently just two species are described for *Sanguina*---the other, *S.
+aurantia* is known from three orange snow samples from Svalbard.
+
+Snow algae form colored patches in otherwise white snowfields, growing in suncups
+
+Different patches of colored snow can contain different algae communities
+
+.
+Currently just two species are described for *Sanguina*, many genetic sequences retrieved from red snow have not been taxonomically assigned[@prochazkova_sanguina_2019].
+Other Chlorophyceae include *Chloromonas* (fourteen species identified to date) and the closely-related but morphologically distinct *Chlainomonas* (two species)[@matsuzaki_taxonomic_2019;@novis_two_2008].
+All species of *Chloromonas* have been cultivated in the lab, but *Sanguina* and *Chlainomonas* remain uncultured@matsuzaki_taxonomic_2019].
+
+The distribution of snow algae species are not understood,
+
+although two studies found snow algae blooms dominated by class Chrysophyceae[@soto_microbial_2020;@remias_two_2019], and one study found a green snow algae bloom dominated by Trebouxiophyceae in Greenland[@lutz_integrated_2015].
 
@@
 <div id="refs"></div>
+
+##### pagebreak
+
+# Figure legends {-}

### 2020-01-07 manuscript/manuscript.Rmd
responded to LQ results comments

@@
 We collected 309 snow algae samples from 13 mountains throughout the summer of 2018.
-We first detected snow algae on May 18, and we continued to sample our low elevation sites at Seymour (S) and Hollyburn (H) until the snow had completely melted out by early July.
-We did not detect snow algae above treeline until June 20, and snow persisted at higher elevations throughout the entire summer.
+At our low elevation sites (850 - 1500 m) we detected snow algae as early as May 18, but these sites were also the earliest to melt out.
+We did not detect snow algae at high elevation sites (1500 - 2200 m) until June 20, where both the snow and snow algae blooms persisted throughout the entire summer.
 
-We found morphologically distinct snow algae communities in different habitats.
-In shaded, forested sites we often found green or orange patches of snow, containing green or orange ovate cells with short spines resembling *Chloromonas cf.
-brevispina* (Fig.
+Using light microscopy we identified morphologically distinct snow algae in different habitats (Fig.
+1, S4).
+Red snow was prevalent in areas of high solar exposure; above treeline sites were often dominated by red cells similar in appearance to *Sanguina nivaloides*[@prochazkova_sanguina_2019] (Fig.
+1a,b).
+We found cells resembling *Chlainomonas rubra*[@novis_two_2008] at all elevations, in highest relative abundance in areas recieving direct sunlight.
+We occasionally observed green snow patches in alpine meadows recieving full sun, but more frequently observed green snow in well-shaded, forested sites.
+At four low-angle sites in clearings near treeline we observed snow melt channels or runnels containing high concentrations of snow algae (Fig.
 1c).
-In forest clearings we observed low-angle linear drainage channels or runnels containing high concentrations of snow algae.
-In May and early June these runnels contained green snow hidden 2 to 5 cm below the white snow surface, and on subsequent visits these same sites contained orange snow at the surface, with morphologies resembling *Chloromonas krienitzii* (Fig 1b).
-Red snow was prevalent above treeline, dominated by red cells similar in appearance to *Sanguina nivaloides*, although we also occasionally observed green snow patches above treeline.
-We found red snow dominated by *Chlainomonas rubra* morphologies in unshaded sites at all elevations.
+In May and early June these sites contained green snow hidden 2 to 5 cm below the white snow surface, but upon subsequent visits the same location contained orange snow at the surface, whose morphologies resembled published images of *Chloromonas krienitzii*[@matsuzaki_revisiting_2015] (Fig.
+1d).
+Forested sites at low elevations often contained green or orange patches of snow, containing green or orange ovate cells with short spines resembling *Chloromonas cf.
+brevispina*[@matsuzaki_revisiting_2015] (Fig.
+1e,f), although we also occasionally observed red snow at forested sites containing morphologies similar to *Sanguina nivaloides*.
 
-We chose 33 snow algae samples for 18S and *rbcL* amplicon Illumina sequencing.
+Of the 309 samples studied by light microscopy, we selected 33 snow algae samples for 18S and *rbcL* amplicon Illumina sequencing.
 Both libraries were dominated by amplicon sequence variants (ASVs) that were assigned to Chlorophyta.
-Our 18S library contained 67 ASVs that were assigned to Chlorophyta and 7 ASVs to Ochrophyta, while our *rbcL* library contained 642 ASVs, of which 603 were assigned to Chlorophyta.
-The top genera were *Chloromonas*, *Chlainomonas*, and *Sanguina*, but most ASVs were not assigned to genus level (Fig.
-2).
-We used t-SNE to generate operational taxonomic units (OTUs), and found many OTUs that did not correspond to any known taxa on GenBank: several OTUs related to *Chloromonas*, and one Chlamydomonadaceae OTU that was only assigned to the species level.
+Our 18S library contained 67 ASVs that were assigned to Chlorophyta and 7 ASVs to Ochrophyta, while our *rbcL* library contained 642 ASVs, of which 603 were assigned to Chlorophyta, the remaining 41 were Trebouxiophyceae.
+The top genera assigned by the taxonomy assignment algorithm IDTaxa[@murali_idtaxa_2018] were *Chloromonas*, *Chlainomonas*, and *Sanguina*; although most ASVs were not assigned to genus level by this algorithm, the majority of ASVs clustered near one of these genera in NMDS ordination (Fig.
+2a).
+We used t-SNE to generate operational taxonomic units (OTUs), and found that many OTUs that did not correspond to any known taxa on GenBank (Fig.
+2b).
+These included three OTUs that were closely related to *Chloromonas*, and one Chlamydomonadaceae OTU that was only assigned to the family level, meaning that it did not match any known algae genus on GenBank.
 
-Above and below treeline communities contained some unique taxa, but most taxa did not exhibit a clear elevational trend (Fig.
+Above and below treeline samples contained some unique taxa, but most taxa did not exhibit a clear elevational trend (Fig.
 3).
-*Sanguina* was the dominant taxa in most samples taken from above treeline.
+*Sanguina* was the dominant taxa in most samples from above treeline.
 Additionally, high elevation sites contained several taxa not found at low elevations: an unannotated OTU of *Chloromonas*, and ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1) were found at the three highest elevation sites.
-Green snow from alpine sites contained *Chloromonas* OTUs that were also found in sites below treeline.
+Green snow from high alpine sites contained *Chloromonas* OTUs that were also found in sites below treeline.
 Low elevation sites contained *Chloromonas krienitzii*, but many sites were dominated by taxa that were found at all elevations, such as *Chlainomonas*.
 
-Cell counts, 18S, and *rbcL* datasets were generally consistent, but we did notice some discrepancies.
+Cell counts, 18S, and *rbcL* datasets were generally consistent, but there were some interesting discrepancies.
 *Chlainomonas* was more prevalent in our *rbcL* library than in cell counts, while *Sanguina* was underrepresented in *rbcL* compared to 18S and cell counts.
-We did not detect *Chlainomonas* in 18S.
+We did not detect *Chlainomonas* with 18S sequencing.
 
@@
 1).
-We sequenced both green subsurface and orange surface snow at three different sites, and found both were dominated by *Chloromonas krienitzii*, although surface samples also contained some *Chlainomonas* (Supplementary _).
+We sequenced both green subsurface and orange surface snow at three different sites, and found both were dominated by *Chloromonas krienitzii*, although surface samples also contained *Chlainomonas* (Supplementary __).
 

### 2020-01-08 manuscript/manuscript.Rmd
pre KY edits

@@
 ---
-title: "Snow algae in the alpine and subalpine zones of British Columbia"
-# who grows in the snow? multi-hued snow algae communities
-# different coloured alpine and subalpine snows harbour constrasting snow algae communities
-# contrasting snow algae communities in di
-# not just dirty snow; alpine snow algae communities have different assemblages
-# green, orange, or red snow caused by distinct
-# snow algae communities causing colored snow in alpine and subalpine summer snow 
-# distinct snow algae communities across an elevational gradient
+title: "Diverse snow algae blooms in the alpine and subalpine zones of British Columbia"
 author:
@@
   #   number_sections: yes
-csl: templates/nature-communications.csl
+csl: frontiers.csl
 bibliography: refs.bib
 abstract: |
-  Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that reduce albedo and increase snow melt.
-Green, orange, and red patches of snow algae can harbor morphologically diverse algae, but despite their global ubiquity the algal species composition of blooms is largely unknown.
-We Illumina sequenced 18S and *rbcL* amplicons from 33 snow algae samples collected from a range of alpine and subalpine habitats in the Coast Range of British Columbia, Canada.
-Predominant species were in, or closely related to, the genera *Sanguina*, *Chloromonas*, and *Chlainomonas*.
-The most common genus above treeline were *Sanguina*, while *Chloromonas* were dominant below treeline, and *Chlainomonas* were abundant in samples across all elevations.
-We found species composition was highly variable from patch to patch, even on the same mountain.
-Although snow algae in the Coast Range are represented by the same relatively few genera that dominate most reports of snow algal blooms around the globe, we found high diversity at the species level.
+  Snow algae blooms cover vast areas of summer snowfields worldwide, and can reduce albedo and increase snow melt.
+Despite their global prevalence, little is known about the species that comprise snow algae blooms.
+We used 18S and *rbcL* high-throughput amplicon sequencing (metabarcoding) to survey 33 snow algae samples from alpine and subalpine habitats in the Coast Range of British Columbia, Canada.
+Species composition varied from bloom to bloom: *Sanguina* predominated above treeline, while *Chloromonas* were prevalent at lower elevations.
+*Chlainomonas* were abundant in samples across all elevations.
+The highest diversity was contained within the genus *Chloromonas*, which included operational taxonomic units (OTUs) that likely represent novel species of snow algae.
+???
+
 ---
@@
 
+<!-- search for ??? before send out -->
+
 ##### pagebreak
@@
 # Introduction {-}
-<!-- audience, problem -->
+<!-- global, vast blooms. albedo->snowmelt. what species? -->
 Each summer in polar and alpine snowfields worldwide, vast areas of snow surface are colored red by snow algae blooms.
-These extremophilic single-celled algae photosynthesize in near-freezing temperatures under high solar irradiance, and are the primary producers in a poorly-understood microbiome that includes bacteria, fungi, ciliates, and rotifers[@gorton_ultraviolet_2003;@lutz_biogeography_2016].
-Snow algae blooms occur on every continent and Arctic sea ice[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019;@gradinger_snow_1996]; in Alaska, snow algae spectral signatures were detected in one third of a 2,000 km^2^ icefield[@ganey_role_2017].
-In recent years snow algae have recieved attention for their impact on snow surface albedo: initial estimates suggests snow algae could increase snow melt rate by as much as 17%, potentially regulating spring snow melt and impacting global albedo[@ganey_role_2017].
-Despite their global ubiquity, we are only beginning to identify the algae that are the primary producers in these complex microbiomes.
+Such blooms are reported from every continent [@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019] and Arctic sea ice [@gradinger_snow_1996].
+They can be extensive: in Alaska, snow algae spectral signatures were detected in one third of a 1,900 km^2^ icefield [@ganey_role_2017].
+In recent years snow algae have received attention for their role in reducing snow surface albedo, which could substantially impact snow melt rates [@lutz_biogeography_2016;@ganey_role_2017].
+Thus, snow algae could impact the timing of spring melt and runoff, reducing water supplies in mountain snowpacks and glacier mass balance.
+Despite their global ubiquity, we are only beginning to identify the species that cause snow algae blooms.
 
-<!-- Reduction in cryospheric albedo has become the focus of attention in recent years as an important feedback loop of global warming.
-Microbial growth on snow surface can reduce albedo, notably when snow algae form extensive blooms on melting summer snow[@ganey_role_2017;@lutz_biogeography_2016;@onuma_temporal_2019].
-.
-Despite their global ubiquity, the  we are only beginning to identify the various types of algae communities that cause snow algae blooms.
--->
-<!-- why do we care if there are different communities as a climate scientist? maybe this should be a paper for snow phycologists? -->
-<!-- different communities could have different ecology and life history -->
-<!-- we need to define the community types to understand ecology of different blooms.  -->
-<!-- it would be nice to know what the species are that form snow algae blooms, and whether different types of snow algae blooms exist.  -->
-<!-- understanding of the species that cause snow algae blooms is limited to  -->
+<!-- why microscopy not reliable -->
+Much of the previous work on snow algae relied on cell morphology for identification, but microscopic identification of snow algae can be unreliable.
+The same species can look completely different with different environmental conditions of life stages.
+For example, cultured *Chloromonas krienitzii* cells are small green biflagellates, but field samples cells of this species are larger orange spheres with thick walls and short spines [@matsuzaki_revisiting_2015].
+In the freshwater green algae *Haematococcus pluvialis* red pigment is produced in response to stress conditions [@shah_astaxanthin_2016], and @muller_persistant_2001 hypothesized that red Arctic snowfields develops from green snow, but thus far genetic studies have failed to support this hypothesis [@terashima_microbial_2017;@lutz_integrated_2015].
+Different species of snow algae can look nearly identical: recent phylogenetic work found that four distinct field samples dominated by similar-looking cells previously referred to as *Chloromonas cf.
+nivalis* are actually genetically distinct clades [@matsuzaki_taxonomic_2019].
 
-<!-- snow algae taxonomy -->
+<!-- species from snow algae blooms, what is currently known -->
+A diversity of algae have been reported from snow algae blooms.
+Chlorophyceae are predominant in many blooms, including the genera *Sanguina*, *Chloromonas*, and *Chlainomonas*.
+Sanger sequencing of red snow samples has shown that *Sanguina nivaloides* is predominant in red snowfields worldwide [@prochazkova_sanguina_2019].
+The taxonomy of this genus is only recently established; while only one other species is ascribed to this genus, many sequences from red snow samples form an unnamed sister clade to *Sanguina*.
+The genus *Chloromonas* is speciose, with twelve species isolated from snow [@matsuzaki_taxonomic_2019], and different species can form green, orange, or brownish-red blooms on the snow surface [@remias_ecology_2018;@remias_insights_2013;@prochazkova_chloromonas_2018].
+Blooms of *Chlainomonas* have been found in red snow overlying alpine lakes in central Europe, western USA, and New Zealand [@novis_two_2008;@remias_ecophysiology_2016;@prochazkova_ecophysiological_2018].
+Two species are currently assigned to the genus *Chlainomonas*, but the taxonomy is not well-established.
+Non-Chlorophyceaen snow algae include Chrysophyceae found in Antarctica, the Alps, and Svalbard [@soto_microbial_2020;@remias_two_2019], and Trebouxiophyceae found in green snow [@lutz_integrated_2015].
 
-<!-- need to introduce the idea of patchiness, and idea of different colored snow -->
-Snow algae blooms can contain a mix of species, many of whose taxonomic identity is not well established.
-Green algae (Chlorophyceae) are often prevalent: *Sanguina nivaloides* is found in red snow samples worldwide[@prochazkova_sanguina_2019].
-Currently just two species are described for *Sanguina*---the other, *S.
-aurantia* is known from three orange snow samples from Svalbard.
+<!-- metabarcoding -->
+While many species of snow algae have been isolated or sequenced from snow, much less is known about the species composition of snow algae blooms: which species predominate, which species co-occur, and how species are distributed across the landscape.
+Previous studies using 18S amplicon high-throughput sequencing (metabarcoding) suggest that blooms are highly similar across continents.
+33 red snow communities across the Arctic were remarkably similar to one another: all were dominated by an "uncultured Chlamydomonadaceae 1 and 2" with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale* [@lutz_biogeography_2016].
+Similarly, another study using ITS2 metabarcoding found 24 red snow sites across the Arctic and Antarctic contained similar assemblages, dominated by "uncultured Chlamydomonadaceae A and B" with secondary abundance of *Raphidonema* and *Chloromonadinia* [@segawa_bipolar_2018].
+Other metabarcoding studies were limited to genus or family level taxonomic resolution [@hamilton_primary_2017].
+As 18S is highly conserved across eukaryotes, sequences can be too similar to resolve between closely related taxa.
 
-Snow algae form colored patches in otherwise white snowfields, growing in suncups
+<!-- our goal, what we did, what we found -->
+Virtually nothing is known about the regional variation in species composition of snow algae blooms.
+Our goal was to identify the algae in blooms on different mountains and elevations in the Coast Range of British Columbia, Canada.
+We used 18S and *rbcL* metabarcoding to assess species composition---the latter targeting a hypervariable region of *rbcL* (coding for large subunit of rubisco), thus targeting only photosynthetic species with high taxonomic resolution.
+Additionally, we used light microscopy to describe the relative abundance of different morphologies in each field sample.
+By using these three cross-referenced metrics of relative abundance we were able to describe the algal species assemblage with high accuracy.
+We discovered a much greater diversity of snow algae than expected.
+Although we did not find any differences in bloom composition between mountains, there were distinct blooms at high versus low elevation.
 
-Different patches of colored snow can contain different algae communities
+# Methods {-}
 
-.
-Currently just two species are described for *Sanguina*, many genetic sequences retrieved from red snow have not been taxonomically assigned[@prochazkova_sanguina_2019].
-Other Chlorophyceae include *Chloromonas* (fourteen species identified to date) and the closely-related but morphologically distinct *Chlainomonas* (two species)[@matsuzaki_taxonomic_2019;@novis_two_2008].
-All species of *Chloromonas* have been cultivated in the lab, but *Sanguina* and *Chlainomonas* remain uncultured@matsuzaki_taxonomic_2019].
+## Field sampling and microscopy {-}
 
-The distribution of snow algae species are not understood,
+We collected snow algae from sites throughout the Coast Range near Vancouver, British Columbia, Canada over the summer of 2018.
+To capture the extent of snow algae diversity in our local mountains we collected as many samples as possible from different elevations, dates, mountains, and micro-habitats within the snow (S2).
+In total we collected 310 colored snow samples from 13 different mountains on 33 different dates from elevations between 880 m and 2150 m above sea level.
+In early season we sampled at lower elevations, moving uphill as the snow algae bloom progressed upwards in elevation.
 
-although two studies found snow algae blooms dominated by class Chrysophyceae[@soto_microbial_2020;@remias_two_2019], and one study found a green snow algae bloom dominated by Trebouxiophyceae in Greenland[@lutz_integrated_2015].
+We scooped samples from visibly colored snow into 50 mL centrifuge tubes using sterile technique.
+To prevent melting en route to the lab we stored tubes in a bag of snow.
+Back in the lab, we melted each sample at room temperature on the bench and removed a 1 mL aliquot for light microscopy.
+Immediately after, samples were stored at -20 &deg;C for up to eight months until DNA extraction.
 
-<!-- microscopy suggests distinct communities -->
-Microscopy suggests that snow algae blooms are in many cases are dominated by single species.
-In the mountains of southwestern USA and Czech Republic red snow was prevalent above treeline, dominated by cells similar in appearance to *S.
-nivaloides*, while below treeline green or orange snow were prevalent dominated by cells similar to *Chloromonas*[@hoham_distribution_1979;@nedbalova_ecology_2008].
-Red snow dominated by *Chlainomonas* type cells were reported from slushy snow overlying alpine lakes[@novis_two_2008;@remias_ecophysiology_2016].
-However, microscopic identification of snow algae is complicated by the fact that the same species can look completely different at different life stages.
-Isolates of *Chloromonas krienitzii* are green, bean-shaped biflagellates, but genetically identical isolated single-cells from field samples appear as orange spheres with thick cell walls, short spines, and no flagella[@matsuzaki_revisiting_2015].
-Conversely, different species can look nearly identical: recent phylogenetic work shows that four similar looking cells previously referred to as *Chloromonas cf.
-nivalis* actually contain multiple genetically distinct clades[@matsuzaki_taxonomic_2019].
+We used light microscopy to characterize the cell morphologies in each sample within 24 hours following collection, with samples maintained on ice.
+We prepared a slide of cells fixed in 2% gluteraldehyde, and counted up to 100 cells at 400x.
+We classified each cell based on similarity to published photographs, as either *Sanguina nivaloides* [@prochazkova_sanguina_2019], *Chloromonas cf.
+nivalis* [@prochazkova_chloromonas_2018], *Chloromonas cf.
+brevispina* [@matsuzaki_revisiting_2015], *Chloromonas krienitzii* [@matsuzaki_revisiting_2015], *Chlainomonas rubra* [@novis_two_2008], and for cells that did not fall into one of there categories as either "green cell" or "other".
 
-<!-- but genetic comparison, most look the same? -->
-Although snow algae communities contain microscopically distinct snow algae community types, studies using 18S amplicon high-throughput sequencing (metabarcoding) suggests that snow algae communities are highly similar.
-33 red snow communities across the Arctic were remarkably similar: all were dominated by uncultured Chlamydomonadaceae, with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale*[@lutz_biogeography_2016].
-Another study using 18S metabarcoding of snow algae in the mountains of Japan found neighboring red and green snow patches with  highly overlapping communities, although the red snow snow contained one OTU not found in the green communities[@terashima_microbial_2017].
-However, side-by-side red and green snow in Greenland contained completely different communities, and the authors concluded that red and green snow were caused by a different species[@lutz_integrated_2015].
+## DNA extraction and amplicon library preparation {-}
 
-Given how little is known about snow algae distribution and diversity, particularly in alpine habitats, our goal is to characterize the snow algae community structure in alpine and subalpine habitats.
-We surveyed snow algae communities from throughout the Coast Range of British Columbia, Canada using light microscopy and metabarcoding.
-18S is highly conserved, making it an ideal universal barcode for eukaryotes, but at lower taxonomic levels 18S is too similar to resolve closely related taxa.
-To complement our 18S sequencing, we also targeted the more differentiated *rbcL* gene, coding for rubisco, thus only targeting the photosynthetic community.
-By using three cross-referenced metrics of relative abundance, we were able to describe the community structure with high taxonomic resolution.
-We found communities above treeline contained a distinct set of taxa, and instead of sharply defined community types we found something more akin to a gradient, with varying relative abundances of different snow algae taxa.
+We selected 33 out of 310 samples for rbcL and 18s high-throughput amplicon sequencing.
+We chose these samples to represent the variation in date, elevation, geographic location, snow color, micro-habitat, and cell morphology.
+We freeze-dried these samples for up to 48 hours until until samples appeared completely desiccated.
+We mini-pestled between 5 to 20 mg of sample at room temperature to physically rupture cell walls before incubation in lysis buffer.
 
-<!--  This is an important first step before we measure habitat variables, will allow us to choose which variables to measure... -->
-<!-- significance. different community, different ecology and or life history. and the snow algae are the primary producers of the community. define the study system.  -->
-<!-- hypothesis: there are distinct communities of snow algae-->
-<!-- methods: cell count, 18s, and rbcL amplicon sequencing -->
-<!-- results: there is much overlap between communities, but many with dominant species.
-Some chloromonas, raphi only found with sanguina.
-Chlainomonas widespread.
-overall, no, there are not sharply delineated communities.
--->
-<!-- new questions raised: what habitat differences are found with different communities?
-is bass becking applicable here?
-what ecological differences, life history differences characterize the different communities?
--->
+To extract DNA from the crushed cells we added 800 mL 1x CTAB extraction buffer [@noauthor_ctab_2009], 1% B-mercaptoethanol, 5 $\mu$L each of proteinase K and RNAase A, and incubated these at 65 &deg;C for 30 minutes.
+We spun samples at 10,000 g for 3 minutes to pellet cell debris, then added 700 $\mu$L of supernatant to an equal volume of 24:1 chloroform:isoamyl alcohol.
+We inverted samples to mix and centrifuged again at 12,000 g for 10 minutes.
+We precipitated DNA by transferring the top layer to 700 $\mu$L ice-cold ethanol, which we gently mixed, and spun over Qiagen DNA columns for 30 s at 15,000 g.
+We washed columns twice with 70% ethanol, and finally dissolved our DNA by spinning with 50 $\mu$L of sterile TE buffer.
+As a negative control we processed a sterile distilled water sample alongside each batch, treating it exactly the same as the other tubes; all negative controls did not contain Qubit-detectable levels of DNA.
+
+We designed custom primers to target a hypervariable region of *rbcL* of snow algae, along with established universal 18S V7-V8 primers Euk1181 and Euk1624 [@wang_optimal_2014].
+We designed *rbcL* primers using the Eurofins primer design tool [@eurofins_pcr_nodate] to target a 400 bp section of *rbcL* based off of 20 snow algae GenBank sequences (Supplemental table 3)---*Sanguina* sequences were not included because they were not available at the time of this study.
+The primer sequences can be seen in Supplemental table 4.
+
+We constructed our amplicon library using a standard two-step PCR approach [@meyer_illumina_2010].
+In the first PCR we amplified template DNA using our primers attached to a universal adapter, and in the second PCR we re-amplified that product to attach a 6 bp index to the universal adapter at the 3’ end.
+The first PCR total volume was 25 $\mu$L, consisting of 1 $\mu$L template, 12.5 $\mu$L Q5 high-fidelity 2X MM (NEB), 1.25 $\mu$L each of forward and reverse primer, and 9 $\mu$L of ddH2O.
+The second PCR was the same except we reduced our reaction volume to 20 $\mu$L by using only 5 $\mu$L of ddH2O.
+The cycling conditions were the same for both primer pairs for the first PCR, with an initial denaturation at 98 &deg;C for 30 s, followed by 30 cycles of 98 &deg;C for 5 s, 58 &deg;C for 10 s, and 72 &deg;C for 25 s, with a final extension at 72 &deg;C for 2 minutes.
+For the second indexing PCR we started with an initial denaturation at 98 &deg;C for 30 s, then 10 cycles of 98 &deg;C for 10 s, 65 &deg;C for 30 s, and 72 &deg;C for 30 s with a final denaturation of 72 &deg;C for 5 min.
+After each PCR, we purified using Agencourt AMPure XP kit (Beckman Coulter).
+We quantified final DNA concentration with Qubit (Thermo Fisher), and standardized sample concentration for pooling.
+The pooled library was then loaded and run on an Illumina MiSeq V3 kit.
+
+## Bioinformatic processing {-}
+
+Reads were demultiplexing using CUTADAPT [@martin_cutadapt_2011].
+Samples were filtered and trimmed, errors removed, dereplicated, pair end reads merged, and chimeras removed following the default pipeline of DADA2 [@callahan_dada2_2017].
+<!-- singletons, failed samples removed, remove sequences based on negative controls, adjust for 0's, sequencing depth, -->
+
+We assigned taxonomy for both *rbcL* and 18S amplicon sequence variants (ASVs) using IDTaxa with threshold set to 50 [@murali_idtaxa_2018].
+We made custom reference databases for each to include snow algae GenBank sequences.
+We made our rbcL database by downloading 30,865 algae rbcL sequences from GenBank with the query “rbcl[gene] AND (algae OR chlorophyta OR trebouxiophyceae) NOT (18s OR ribosomal OR psaB OR atpB) AND 300:2000[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion”.
+We removed ambiguous snow algae annotations: because entries labelled as *Chloromonas cf.
+brevispina* and *Chloromonas cf.
+nivalis* are polyphyletic [@matsuzaki_taxonomic_2019] we re-labeled these as “unassigned *Chloromonas*”.
+To assign 18S reads we initially ran our ASVs against SILVA [@quast_silva_2013], and then re-ran ASVs assigned to Chlorophyta on our custom 18S snow algae database based on the following GenBank query: "(18S OR ribosomal)[gene] AND (chloromonas OR chlainomonas OR sanguina OR raphidonema OR KMY-2018) NOT (rbcL OR psaB OR atpB) AND 300:3500[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion".
+
+As most operational taxonomic unit (OTU) clustering algorithms are optimized for 18S data, we visualized *rbcL* ASV clustering using t-SNE [@maaten_visualizing_2008] (Fig.
+2b).
+We tested this with perplexity values ranging from 1 to 100, and found comparable results with perplexity ranging from 10 to 50.
+We compared sample similarity using UniFrac with default settings [@lozupone_unifrac_2005].
+
+For software information, see Supplementary Table 9.
+All raw fastq files are freely available on the European Nucleotide Archive under the project accession PRJEB34539.
+All scripts are available at https://github.com/cengstro/bc_snow_algae_amplicon.
 
@@
 1, S4).
-Red snow was prevalent in areas of high solar exposure; above treeline sites were often dominated by red cells similar in appearance to *Sanguina nivaloides*[@prochazkova_sanguina_2019] (Fig.
+Red snow was prevalent in areas of high solar exposure; above treeline sites were often dominated by red cells similar in appearance to *Sanguina nivaloides* [@prochazkova_sanguina_2019] (Fig.
 1a,b).
-We found cells resembling *Chlainomonas rubra*[@novis_two_2008] at all elevations, in highest relative abundance in areas recieving direct sunlight.
-We occasionally observed green snow patches in alpine meadows recieving full sun, but more frequently observed green snow in well-shaded, forested sites.
+We found cells resembling *Chlainomonas rubra* [@novis_two_2008] at all elevations, in highest relative abundance in areas receiving direct sunlight.
+We occasionally observed green snow patches in alpine meadows receiving full sun, but more frequently observed green snow in well-shaded, forested sites.
 At four low-angle sites in clearings near treeline we observed snow melt channels or runnels containing high concentrations of snow algae (Fig.
 1c).
-In May and early June these sites contained green snow hidden 2 to 5 cm below the white snow surface, but upon subsequent visits the same location contained orange snow at the surface, whose morphologies resembled published images of *Chloromonas krienitzii*[@matsuzaki_revisiting_2015] (Fig.
+In May and early June these sites contained green snow hidden 2 to 5 cm below the white snow surface, but upon subsequent visits the same location contained orange snow at the surface, whose morphologies resembled published images of the *Chloromonas krienitzii* lineage [@matsuzaki_revisiting_2015] (Fig.
 1d).
 Forested sites at low elevations often contained green or orange patches of snow, containing green or orange ovate cells with short spines resembling *Chloromonas cf.
-brevispina*[@matsuzaki_revisiting_2015] (Fig.
+brevispina* [@matsuzaki_revisiting_2015] (Fig.
 1e,f), although we also occasionally observed red snow at forested sites containing morphologies similar to *Sanguina nivaloides*.
@@
 Our 18S library contained 67 ASVs that were assigned to Chlorophyta and 7 ASVs to Ochrophyta, while our *rbcL* library contained 642 ASVs, of which 603 were assigned to Chlorophyta, the remaining 41 were Trebouxiophyceae.
-The top genera assigned by the taxonomy assignment algorithm IDTaxa[@murali_idtaxa_2018] were *Chloromonas*, *Chlainomonas*, and *Sanguina*; although most ASVs were not assigned to genus level by this algorithm, the majority of ASVs clustered near one of these genera in NMDS ordination (Fig.
+The top genera assigned by the taxonomy assignment algorithm IDTaxa [@murali_idtaxa_2018] were *Chloromonas*, *Chlainomonas*, and *Sanguina*; although most ASVs were not assigned to genus level by this algorithm, the majority of ASVs clustered near one of these genera in NMDS ordination (Fig.
 2a).
@@
 3).
-*Sanguina* was the dominant taxa in most samples from above treeline.
-Additionally, high elevation sites contained several taxa not found at low elevations: an unannotated OTU of *Chloromonas*, and ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1) were found at the three highest elevation sites.
+*Sanguina* was the dominant taxa in most samples from above treeline, and was absent in sequences libraries from the samples below 1500 m.
+Additionally, these high elevation samples uniquely contained an unannotated OTU of *Chloromonas* (Fig.
+3, "Chloromonas C").
+The three highest elevation sites uniquely contained ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1)---all three samples were taken from snow overlying glacier .
 Green snow from high alpine sites contained *Chloromonas* OTUs that were also found in sites below treeline.
@@
 1).
-We sequenced both green subsurface and orange surface snow at three different sites, and found both were dominated by *Chloromonas krienitzii*, although surface samples also contained *Chlainomonas* (Supplementary __).
+We sequenced both green subsurface and orange surface snow at three different sites, and found both were dominated by *Chloromonas krienitzii*, although surface samples also contained *Chlainomonas* (Supplementary figure ???).
 
-# Discussion
+# Discussion {-}
 
-We found that the alpine snow algae community can vary widely from site to site.
-Different taxa were dominant at different elevations, with alpine sites dominated several distinct taxa, most notably *Sanguina*.
-*Chloromonas krienitzii* was dominant in mid-elevation runnels appearing as green motile cells below the snow surface, and later in the season developing orange pigment at the surface.
-*Chloromonas* and *Chlainomonas* were abundant across all elevations, including many taxa that are not represented in GenBank.
+We found that alpine snow algae bloom composition varied from site to site, with some elevational trends.
+*Sanguina* was dominant above treeline, while *Chloromonas krienitzii* was dominant in mid-elevation runnels, which initially appeared as green motile cells below the snow surface but developed into orange blooms at the snow surface.
+We found an unexpected diversity within *Chloromonas*, including many OTUs not represented in GenBank.
+Previous studies suggested that *Chlainomonas* is restricted to waterlogged snow overlying mountain lakes [@novis_two_2008;@prochazkova_ecophysiological_2018] but our results suggest that this genus can be widespread and abundant in many alpine red snow sites.
+While one *Chlainomonas*-dominant sample was located in waterlogged snow near the edge of a pool (sample S9), the other *Chlainomonas*-dominant sites were not notably wetter than the surrounding snow, nor overlying ice.
 
 Our findings highlight the remaining unexplored diversity in the snow algae microbiome.
-Most of the diversity was only detected in *rbcL*: samples enriched in Chlamydomonadaceae E did not correspond to anything we detected in cell counts and 18S.
-One possibility is that this OTU represents a novel taxa of algae whose 18S is highly similar to Chloromonas.
-Another possibility is primer bias: perhaps this OTU was over-represented in *rbcL* and underrepresented in 18S.
+*rbcL* revealed greater diversity than 18S, and many *rbcL* OTUs did not closely match any annotated GenBank sequences (Fig.
+2).
+Likely, the *Chloromonas* *rbcL* OTUs were lumped together in the 18S dataset due to high similarity.
+We found higher *rbcL* diversity in *Chloromonas* than *Sanguina*, which could be because *Chloromonas* lack a pyrenoid, and consequently has many non-synonymous mutations in the region of *rbcL* the codes for binding these proteins together to form a pyrenoid [@nozaki_differences_2002].
+One *rbcL* OTU, "Chlamydomonadaceae E", was only assigned to family level, which could represent a novel genus.
+However, this OTU did not correspond with any features in our 18S or cell count data.
+It seems unlikely that Chlamydomonadaceae E would have been lumped in with *Chloromonas* in 18S, given how distinct their *rbcL* sequences were from *Chloromonas*.
+Primers bias could explain this discrepancy: perhaps our *rbcL* primers over-represented this OTU compared to 18S.
 
-Previous studies suggested that *Chlainomonas* is restricted thought to waterlogged snow overlying mountain lakes, but our results suggest that it is widespread and abundant in all types of snow algae blooms[@novis_two_2008;@prochazkova_ecophysiological_2018].
-We did not detect any association with a particular habitat: only one site dominated by *Chlainomonas* was near water (S9), but other *Chlainomonas* sites were not notably wetter than the surrounding snow.
+Although 18S, *rbcL*, and cell counts were generally consistent, we observed some interesting discrepancies (Fig.
+3).
+We did not observe *Chlainomonas* in our 18S dataset, despite the fact that there were two representative sequences from this genus in our reference database (MF803743.1, MF803745.1).
+Likely, *Chlainomonas* 18S in our target region (V7-V8) is not sufficiently differentiated from *Chloromonas* to be distinguishable.
+Many samples contained high relative abundance of *Chlainomonas* *rbcL*, yet these samples contained lower concentration of *Chlainomonas* cell morphologies.
+Large cells such as *Chlainomonas* often have high rRNA gene copy number [@fu_single_2017], which could account for this discrepancy.
+*Sanguina* was found in the same samples, but relative abundance was generally higher in 18S and cell counts.
+Our *rbcL* primers were designed based on *Chloromonas* and *Chlainomonas* snow algae sequences (*Sanguina* sequence data was not available at the time), so primer bias could have resulted in over-representation of *Chloromonas* relative to *Sanguina*.
 
-Although 18S, *rbcL*, and cell counts were generally consistent, we observed some discrepancies, notably *Sanguina* *rbcL* was underrepresented relative to 18S (Fig.
+Variation in bloom composition could be due to a wide range of habitat features.
+*Sanguina* and *Chloromonas C* were limited to sites above 1500 m in full sunlight, but many low elevation sites also received full sunlight and did not contain these OTUs (Fig.
 3).
-Primer bias could partly explain this discrepancy, because we designed our *rbcL* primers off of *Chloromonas* sequences, as *Sanguina* reference data was not available at the time.
-Additionally, we observed that *Sanguina* cell morphologies were more resistant to lysis than other cell types in microscopy following lysis.
-*Chlainomonas* was overrepresented in *rbcL* relative to cell counts, possibly due to primer bias again, or higher rRNA gene copy number typical of larger cell volumes[@fu_single_2017].
-None of our 18S ASVs returned BLAST results for *Chlainomonas*, despite there being two representative GenBank sequences in our reference database (MF803743.1, MF803745.1).
-Possibly the *Chlainomonas* 18S sequence in our target region (V7-V8) is not different enough from closely-related *Chloromonas* to be detectable in this survey.
+Light intensity, snow moisture, or snow chemistry could all plausibly influence community composition.
+Intriguingly, we only observed *Raphidonema* at high-elevation glacier sites.
+In Svalbard *Raphidonema nivale* abundance increased on glacier surface snow following windstorms, and the authors suggest that this is a soil algae that grows sub-optimally on the snow surface following wind deposition [@stibal_growth_2005].
+Underlying topography could also potentially influence species distribution: two sites dominated by *Chloromonas krienitzii* were in runnels overlying ephemeral streams, which could influence nutrient availability where surface debris is deposited in the snow runnel.
+Given the aerial dispersal capabilities of microalgae [@tesson_airborne_2016] and genetic overlap between distant snow algae populations [@segawa_bipolar_2018] it seems unlikely that geographic distance is a barrier for snow algae distribution.
 
-Species composition is likely governed, at least in part, by different site habitat characteristics.
-The absence of *Sanguina* from low elevation sites could be indicative of habitat specificity: perhaps it is outcompeted at low elevation sites by taxa that are better adapted to the low light of the shade.
-Given the prevalence of airborne microalgae, it seems less likely that *Sanguina* is restricted in it's dispersal capabilities[@tesson_airborne_2016].
-Perhaps *Chloromonas krienitzii* dominates in low-angle runnels where moisture is sufficient; alternatively, we observed that many runnels overlay streams, which could relate to the dispersal mechanism of this species.
-We found *Raphidonema* only in high elevation sites that were sampled late in the growing season.
-@stibal_growth_2005 suggests that this is a slow-growing opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow.
-Perhaps sites late in the season were closer to bare ground containing source populations of this algae; alternately, perhaps this algae is slower to grow in cold temperatures on the snow, and is outcompeted by the better adapted *Sanguina*.
+Over multiple visits to sites S and H we observed green motile *Chloromonas krienitzii* cells below the snow surface develop into an orange bloom at the surface (Supplementary figure ???).
+Previous work has shown that this species can appear as distinct green and orange morphologies [@matsuzaki_revisiting_2015], but our observations are the first to our knowledge to describe this transition occurring in the field.
+The change over weeks suggests that this process is mediated by seasonal changes.
+Secondary pigments likely protect snow algae from the damaging effects of intense solar irradiation at the snow surface [@bidigare_evidence_1993], which could be why we only observed green cells below the snow surface.
 
-Characterization of the snow algae communities allows us to define the study system, and serves as a foundation for understanding the ecology and life history of this globally significant microbiome.
-[...]
-
-# Methods {-}
-
-## Field sampling and microscopy
-
-We collected snow algae from mountains near Vancouver, British Columbia, Canada throughout the summer of 2018.
-To capture the extent of snow algae diversity in our local mountains we collected as many samples as possible from different elevations, dates, mountains, and micro-habitats within the snow (S2).
-In total we collected 310 colored snow samples from 13 different mountains on 33 different dates from elevations between 880 m and 2150 m above sea level.
-In early season we sampled at lower elevations, moving uphill as the snow algae bloom progressed upwards in elevation.
-
-We scooped samples from visibly colored snow into 50 mL centrifuge tubes using sterile technique.
-To prevent melting en route to the lab we stored tubes in a bag of snow.
-Back in the lab, we melted each sample at room temperature on the bench and removed a 1 mL aliquot for light microscopy.
-Immediately after, samples were stored at -20 &deg;C for up to eight months until DNA extraction.
-
-We used light microscopy to characterize the cell morphologies in each sample.
-We prepared a slide of cells fixed in 2% gluteraldehyde, and at 400x magnification starting from the center of the slide counted up to 100 cells.
-We classified each cell as either *Sanguina nivaloides* [@Prochazkova2019a], *Chloromonas cf.
-nivalis* [@Matsuzaki2019;@Prochazkova2017], *Chloromonas cf.
-brevispina* [@Nedbalova2008a;@Matsuzaki2015], *Chloromonas krienitzii* [@Matsuzaki2015] *Chlainomonas rubra* [@Novis2008], or "Other" based on similarity to published photographs.
-
-## DNA extraction and amplicon library preparation
-
-We selected 33 out of 310 samples for rbcL and 18s high-throughput amplicon sequencing.
-We chose these samples to represent the variation in date, elevation, geographic location, snow color, micro-habitat, and cell morphology.
-We freeze-dried these samples for up to 48 hours until until samples appeared completely desiccated.
-To allow water vapor to escape sample tubes while freeze drying we poked holes in each lid; as a safeguard against cross-contamination from airborne snow algae powder we wrapped tubes individually with paper towels.
-We mini-pestled 5 mg of sample at room temperature to physically rupture cell walls.
-Based on our observation that samples dominated by *S.
-nivaloides* cell morphologies had lower DNA yields, and had many intact cells following various lysis methods, we used 20 mg of these samples to increase our DNA yield.
-
-To extract DNA from the crushed cells we added 800 mL 1x CTAB extraction buffer [@ColdSp2009], 1% B-mercaptoethanol, 5 $\mu$L each of proteinase K and RNAase A, and incubated these at 65 &deg;C for 30 minutes.
-We centrifuged samples at 10,000 g for 3 minutes to pellet cell debris, then added 700 $\mu$L of supernatant to an equal volume of 24:1 chloroform:isoamyl alcohol.
-We inverted samples to mix and centrifuged again at 12,000 g for 10 minutes.
-We precipitated DNA by transferring the top layer to 700 $\mu$L ice-cold ethanol, which we gently mixed, and spun over Qiagen DNA columns for 30 s at 15,000 g.
-We washed columns twice with 70% ethanol, and finally dissolved our DNA by spinning with 50 $\mu$L of sterile TE buffer.
-As a negative control we processed a sterile distilled water sample alongside each batch, treating it exactly the same as the other tubes; all negative controls did not contain Qubit-detectable levels of DNA.
-
-We used rbcL as a marker gene, which offers high resolution between microalgae species [@Zou2016], and has reference data for snow algae available on GenBank.
-We also sequenced each sample with 18S primers, due to it's coverage across a wide range of taxa, and its ubiquity in reference databases.
-We designed rbcL primers to target an approximately 400 bp section of this gene, based on 20 snow algae rbcL GenBank sequences from *Chloromonas* and *Chlainomonas* (GenBank accession numbers AB434272.1, LC012752.1, LC012747.1, AF517072.1, LC012738.1, LC012739.1, AB434267.1, EU030690.1, LC360494.1, AJ001878.1, AB022225.1, DQ885964.2, DQ885962.1, AJ001879.1, AB022226.1, AB022530.1, LC012751.1, AB504764.1, EU030689.1, AB101508.1).
-*Sanguina* sequences were not included because they were not available at the time.
-We designed primers using the primer design tool from the Eurofins Genomics website (www.eurofinsgenomics.eu/en/ecom/tools/pcr-primer-design/).
-The resultant primers were rbcL369F (5’-GAA CGT GAC AAA TTA AAC AAA-3’) and rbcL870R (5’-ACC WGA YAD ACG WAG AGC TT-3’).
-To target 18S we used Euk1181 (5'-TTA ATT TGA CTC AAC RCG GG-3') and Euk1624 (5'-CGG GCG GTG TGT ACA AAG G-3'), targeting the V7-V8 hypervariable regions[@Wang2014;@].
-
-We constructed our amplicon library using a two-step PCR [@Meyer2010].
-In the first PCR we amplified template DNA using our primers attached to a universal adapter, and in the second PCR we re-amplified that product to attach a 6 bp index to the universal adapter at the 3’ end.
-The first PCR total volume was 25 $\mu$L, consisting of 1 $\mu$L template, 12.5 $\mu$L Q5 high-fidelity 2X MM (New England BioLabs), 1.25 $\mu$L each of forward and reverse primer, and 9 $\mu$L of ddH2O.
-The second PCR was the same except we reduced our reaction volume to 20 $\mu$L by using only 5 $\mu$L of ddH2O.
-The cycling conditions were the same for both primer pairs for the first PCR, with an initial denaturation at 98 &deg;C for 30 s, followed by 30 cycles of 98 &deg;C for 5 s, 58 &deg;C for 10 s, and 72 &deg;C for 25 s, with a final extension at 72 &deg;C for 2 minutes.
-For the second indexing PCR we started with an initial denaturation at 98 &deg;C for 30 s, then 10 cycles of 98 &deg;C for 10 s, 65 &deg;C for 30 s, and 72 &deg;C for 30 s with a final denaturation of 72 &deg;C for 5 min.
-After each PCR, we purified using Agencourt AMPure XP kit (Beckman Coulter).
-We quantified final DNA concentration with Qubit (Thermo Fisher), and standardized sample concentration for pooling.
-The pooled library was then loaded and run on an Illumina MiSeq V3 kit.
-
-## Bioinformatic processing
-
-Reads were demultiplexing using CUTADAPT [@Martin2011].
-Samples were filtered and trimmed, errors removed, dereplicated, pair end reads merged, and chimeras removed following the default pipeline of DADA2 [@Callahan2017].
-<!-- singletons, failed samples removed, remove sequences based on negative controls, adjust for 0's, sequencing depth, -->
-
-We assigned taxonomy for both *rbcL* and 18S ASVs using IDTaxa with threshold set to 50 [@Murali2018].
-We made custom reference databases for each to include snow algae GenBank sequences.
-We made our rbcL database by downloading 30,865 algae rbcL sequences from GenBank with the query “rbcl[gene] AND (algae OR chlorophyta OR trebouxiophyceae) NOT (18s OR ribosomal OR psaB OR atpB) AND 300:2000[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion”.
-We removed ambiguous snow algae annotations based on the most recent snow algae phylogeny [@Matsuzaki2019].
-Because entries labelled as *Chloromonas cf.
-brevispina* and *Chloromonas cf.
-nivalis* are polyphyletic we re-labeled these as “unassigned *Chloromonas*”.
-To assign 18S reads we initially ran our ASVs against SILVA [@Quast2013], and then re-ran ASVs assigned to Chlorophyta on our custom 18S snow algae database based on the following GenBank query: "(18S OR ribosomal)[gene] AND (chloromonas OR chlainomonas OR sanguina OR raphidonema OR KMY-2018) NOT (rbcL OR psaB OR atpB) AND 300:3500[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion".
-This database was edited to reflect the most recent snow algae taxonomy and remove ambiguous annotations [@Matsuzaki2019;@Prochazkova2018c].
-
-As most OTU clustering algorithms are optimized for 18S data, we chose to assign high level OTUs based on the results of t-SNE clustering [@Graham-Rowe1999] (Fig.
-2a).
-We tested this with perplexity values ranging from 1 to 100, and found comparable results with perplexity ranging from 10 to 50.
-We compared sample similarity using UniFrac with default settings [@Lozupone2005].
-
-For software information, see Supplementary Table 9.
-All raw fastq files are freely available on the European Nucleotide Archive under the project accession PRJEB34539.
-All scripts are available at https://github.com/cengstro/bc_snow_algae_amplicon.
-
-##### pagebreak
+In conclusion, our study reveals substantial diversity within and between snow algae blooms.
+Species were differentially distributed along an elevational gradient, with distinct blooms dominated by *Chloromonas*, *Sanguina*, or *Chlainomonas*.
+Future studies will examine other taxa present in snow algae blooms, and their interactions in these globally important microbiomes.
 
@@
 
-We wish to thank Leah Tooman (Simon Fraser University) for assistance with sequencing, and Chris Rushton (Simon Fraser University) for assistance with bioinformatics.
+We thank Leah Tooman (Simon Fraser University) for assistance with sequencing, and Chris Rushton (Simon Fraser University) for assistance with bioinformatics.
 This project was funded with a Sector Innovation Grant from Genome BC (SIP016), and a NSERC Individual Discovery Grant, both awarded to LQ.
@@
 
+l
+
 ##### pagebreak
@@
 # Figure legends {-}
+
+<!-- formatting indentation? -->
+
+**Figure 1.** Representative photographs of snow algae.
+A.
+Red snow bloom above treeline at sample site G1.1.
+B.
+Microphotograph of red snow containing *Sanguina nivaloides* (Sn) and *Chlainomonas rubra* (Cr) cell morphologies.
+All scale bars 30 $\mu$m, all microphotos taken using 63x objectives with DIC.
+C.
+Snow runnels in a forest clearing containing snow algae (inset).
+D.
+Microphotograph of orange snow from the surface of runnel containing *Chloromonas krienitzii* (Ck) and *Chloromonas cf.
+nivalis* (Cn).
+E.
+Bronze coloured snow algae blooms below conifer canopy.
+F.
+Microphotograph of bronze snow containing Chloromonas cf.
+brevispina* (Cb) and *Chloromonas cf.
+nivalis* (Cn).
+
+**Figure 2.** Ordination plots of *rbcL* amplicon sequence variant (ASV) similarity.
+A.
+Multidimensional scaling (MDS) plot.
+Taxonomy (assigned by IDTaxa) indicated by color, and size of point is proportional to cumulative relative abundance.
+Dotted ellipses show OTU clusters.
+Stress = 0.13.
+B.
+t-Distributed Stochastic Neighbor Embedding (t-SNE) dimensionality reduction plot of snow algae *rbcL* ASVs.
+Perplexity = 30.
+
+**Figure 3.** Stacked bar plots of snow algae relative abundance.
+Each sample is arranged on the y axis in order of elevation, with each compositional bar plot representing the (left to right) cell morphology, 18S, and *rbcL* snow algae assemblages of the same sample.
+Colors below each bar plot indicate taxonomic categories.
+Morphospecies were identified by light microscopy, using similarity to published photographs as a guide.
+18S operational taxonomic units (OTUs) are reference-based, while *rbcL* OTUs were clustered *de novo*, and named using GenBank reference data.
+
+**Figure 4.** Non-metric multidimensional scaling (NMDS) showing *rbcL* UniFrac distances between samples. Each point represents a sample, which is labelled by sample ID and elevation (color).

### 2020-01-08 manuscript/manuscript.Rmd
final to ky

@@
 The highest diversity was contained within the genus *Chloromonas*, which included operational taxonomic units (OTUs) that likely represent novel species of snow algae.
-???
 
@@
 In recent years snow algae have received attention for their role in reducing snow surface albedo, which could substantially impact snow melt rates [@lutz_biogeography_2016;@ganey_role_2017].
-Thus, snow algae could impact the timing of spring melt and runoff, reducing water supplies in mountain snowpacks and glacier mass balance.
+Thus, snow algae could impact the timing of spring melt and runoff, depleting summer water supplies held in mountain snowpacks and reducing glacier mass balance.
 Despite their global ubiquity, we are only beginning to identify the species that cause snow algae blooms.
@@
 
-l
-
 ##### pagebreak
@@
 
-<!-- formatting indentation? -->
+<!-- formatting indentation??? -->
 
@@
 t-Distributed Stochastic Neighbor Embedding (t-SNE) dimensionality reduction plot of snow algae *rbcL* ASVs.
-Perplexity = 30.
+Perplexity = 30.l
 

### 2020-01-08 manuscript/manuscript.Rmd
fixed refs

@@
 For example, cultured *Chloromonas krienitzii* cells are small green biflagellates, but field samples cells of this species are larger orange spheres with thick walls and short spines [@matsuzaki_revisiting_2015].
-In the freshwater green algae *Haematococcus pluvialis* red pigment is produced in response to stress conditions [@shah_astaxanthin_2016], and @muller_persistant_2001 hypothesized that red Arctic snowfields develops from green snow, but thus far genetic studies have failed to support this hypothesis [@terashima_microbial_2017;@lutz_integrated_2015].
+In the freshwater green algae *Haematococcus pluvialis* red pigment is produced in response to stress conditions [@shah_astaxanthin-producing_2016], and @muller_persistent_2001 hypothesized that red Arctic snowfields develops from green snow, but thus far genetic studies have failed to support this hypothesis [@terashima_microbial_2017;@lutz_integrated_2015].
 Different species of snow algae can look nearly identical: recent phylogenetic work found that four distinct field samples dominated by similar-looking cells previously referred to as *Chloromonas cf.
@@
 
-# Methods {-}
+# Materials and Methods {-}
 

### 2020-01-10 manuscript/casey1.Rmd
renamed from 'manuscript.Rmd', so far re-wrote title and abstract to be more splashy

---
title: "Alpine snow algae blooms vary with elevation"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
output: 
  bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: templates/word-template.docx
  # pdf_document:
  #   fig_caption: yes
  #   number_sections: yes
csl: the-isme-journal.csl
bibliography: refs.bib

---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### *Keywords*: snow algae, microbiome, amplicon, *rbcL*, 18S, alpine, metabarcoding

### *Run title*: Alpine snow algae blooms

<!-- search for ??? before send out -->

##### pagebreak

# Abstract {-}

Snow algae blooms cover vast areas of summer snowfields worldwide, and can reduce albedo and increase snow melt. Despite their global prevalence, little is known about the algae species that comprise snow algae blooms. We used 18S and *rbcL* metabarcoding and light microscopy to characterise species composition of 33 snow algae samples from alpine and subalpine habitats in the Coast Range of British Columbia, Canada. We found blooms were dominated by the snow algae genera *Sanguina*, *Chloromonas*, and *Chlainomonas*. Above treeline blooms contained *Sanguina*, *Raphidonema*, and one OTU of *Chloromonas*, distinct from species assemblages below treeline predominated by *Chloromonas krienitzii*. Previously unrecognized as an important component of the snow algae microbiome, *Chlainomonas* were abundant in samples across all elevations. Our *rbcL* dataset detected diversity that was obscured by the low taxonomic resolution of traditional 18S metabarcoding, including OTUs that could represent novel species of snow algae. These three cross-referenced datasets show alpine snow algae blooms are more diverse than previously thought, and are highly variable from site to site along an elevational gradient.

<!-- need to add the few meters part -->

##### pagebreak

# Introduction {-}
<!-- global, vast blooms. albedo->snowmelt. what species? -->
Each summer in polar and alpine snowfields worldwide, vast areas of snow surface are colored red by snow algae blooms. Such blooms are reported from every continent [@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019] and Arctic sea ice [@gradinger_snow_1996]. They can be extensive: in Alaska, snow algae spectral signatures were detected in one third of a 1,900 km^2^ icefield [@ganey_role_2017]. In recent years snow algae have received attention for their role in reducing snow surface albedo, which could substantially impact snow melt rates [@lutz_biogeography_2016;@ganey_role_2017]. Thus, snow algae could impact the timing of spring melt and runoff, depleting summer water supplies held in mountain snowpacks and reducing glacier mass balance. Despite their global ubiquity, we are only beginning to identify the species that cause snow algae blooms.

<!-- why microscopy not reliable -->
Much of the previous work on snow algae relied on cell morphology for identification, but microscopy-based identification of snow algae can be unreliable. Different species of snow algae can look nearly identical: recent phylogenetic work found that four distinct field samples dominated by similar-looking cells previously referred to as *Chloromonas cf. nivalis* are actually genetically distinct clades [@matsuzaki_taxonomic_2019]. Furthermore, the same species can look completely different in different environmental conditions or life stages. For example, cultured *Chloromonas krienitzii* cells are small green biflagellates, but field samples cells of this species are larger orange spheres with thick walls and short spines [@matsuzaki_revisiting_2015]. In the freshwater green algae *Haematococcus pluvialis* red secondary pigments are produced in response to high light [@shah_astaxanthin-producing_2016], but genetic comparison of red and green snow found they contained different species [@terashima_microbial_2017;@lutz_integrated_2015].

<!-- species from snow algae blooms, what is currently known -->
A diversity of algae have been reported from snow algae blooms, with many new species only having been described recently. Genera from the class Chlorophyceae are predominant in many blooms, including *Sanguina*, *Chloromonas*, and *Chlainomonas*. The taxonomy of *Sanguina* was recently established, and currently contains just two species; however, many sequences from red snow samples form an unnamed sister clade to *Sanguina* [@prochazkova_sanguina_2019]. The same study found *Sanguina nivaloides* predominated in red snowfields worldwide, while *S. aurantia* occurred in just three orange snow samples in Svalbard. The genus *Chloromonas* is speciose, containing twelve species that were isolated from snow [@matsuzaki_taxonomic_2019]. Different species of *Chloromonas* can form green, orange, or brownish-red blooms on the snow surface, and this genus has been reported from both polar and mid-latitude snowfields [@remias_ecology_2018;@remias_insights_2013;@prochazkova_chloromonas_2018]. Blooms of *Chlainomonas* have been found in red snow overlying alpine lakes in central Europe, western USA, and New Zealand [@novis_two_2008;@remias_ecophysiology_2016;@prochazkova_ecophysiological_2018]. This genus, which currently contains two named species, is closely related to *Chloromonas*. Non-Chlorophyceaen snow algae include Chrysophyceae found in Antarctica, the Alps, and Svalbard [@soto_microbial_2020;@remias_two_2019], and Trebouxiophyceae found in green snow in Greenland [@lutz_integrated_2015].

<!-- metabarcoding -->
In contrast to the high diversity of snow algae species observed in culture and Sanger sequencing, metabarcoding suggests that snow algae blooms are highly similar across continents, and are dominated by few taxa. 33 red snow communities across the Arctic were remarkably similar to each other: all were dominated by two operational taxonomic units (OTUs) of "uncultured Chlamydomonadaceae 1 and 2" with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale* [@lutz_biogeography_2016]. Similarly, another study using ITS2 metabarcoding found 24 red snow sites across the Arctic and Antarctic contained similar assemblages, dominated by "uncultured Chlamydomonadaceae A and B" with secondary abundance of *Raphidonema* and *Chloromonadinia* [@segawa_bipolar_2018]. Other 18S metabarcoding studies of snow algae were limited to genus or family level taxonomy [@hamilton_primary_2017]---because 18S is highly conserved, this marker generally cannot distinguish between closely related taxa.

<!-- our goal, what we did, what we found -->
Virtually nothing is known about the regional variation in species composition of snow algae blooms. Our goal was to identify the algae in blooms on different mountains and elevations in the Coast Range of British Columbia, Canada. We used 18S and *rbcL* metabarcoding to assess species composition---the latter targeting a hypervariable region of *rbcL* (coding for large subunit of rubisco), thus targeting only photosynthetic species with high taxonomic resolution. Additionally, we used light microscopy to describe the relative abundance of different morphologies in each field sample. By using these three cross-referenced metrics of relative abundance we were able to describe the algal species assemblage with high accuracy. Using *rbcL* as a marker revealed much greater diversity than would have been found using 18S alone. We did not find any differences in bloom composition between mountains, but there were distinct algae species assemblages at high versus low elevations.
<!-- stronger hook, predictions??? cliff hangers -->

# Materials and Methods {-}

<!-- refer to each supp at least once in txt -->
<!-- add cities, states -->

## Field sampling and microscopy {-}

We collected snow algae from sites throughout the Coast Range near Vancouver, British Columbia, Canada over the summer of 2018 (Supplementary Figure S1). To capture the extent of snow algae diversity we sampled from different elevations, dates, mountains, and micro-habitats within the snow. In total, we collected 309 colored snow samples from a range of 880 - 2150 m elevation, including samples from 13 different mountains collected on 33 different dates (Supplementary Table S1). Our samples trended upwards in elevation over the course of the summer as the snow melted earliest at the lowest elevations (Supplementary Figure S2). We first detected snow algae above 1500 m on June 20. We scooped samples from visibly colored snow into sterile 50 mL centrifuge tubes. To prevent melting en route to the lab we stored tubes in a bag of snow. We melted each sample at room temperature on the bench and removed a 1 mL aliquot for light microscopy. Immediately after, samples were stored at -20 &deg;C for up to eight months until DNA extraction.

We used light microscopy to characterize the cell morphologies in each sample within 24 hours following collection. We fixed samples in 2% gluteraldehyde, and under 400x magnification observed the 100 cells closest to the center of the slide. We classified cell morphology based on similarity to published photographs of *Sanguina nivaloides* [@prochazkova_sanguina_2019], *Chloromonas cf. nivalis* [@prochazkova_chloromonas_2018], *Chloromonas cf. brevispina* [@matsuzaki_revisiting_2015], *Chloromonas krienitzii* [@matsuzaki_revisiting_2015], *Chlainomonas rubra* [@novis_two_2008], and for cells that did not fall into one of there categories as either "green cell" or "other".

<!-- how do select field and micro photos fit into this? -->

## DNA extraction and amplicon library preparation {-}

<!-- This part is non reproducible, selecting based on date, etc -->

We selected 33 out of our 310 samples for *rbcL* and 18S Illumina amplicon sequencing. We chose these samples to represent the variation in date, elevation, geographic location, snow color, micro-habitat, and cell morphology. We freeze-dried these samples for up to 48 hours until until samples appeared completely desiccated. We mini-pestled between 5 to 20 mg of sample at room temperature to physically rupture cell walls before incubation in CTAB lysis buffer [@noauthor_ctab_2009]. We extracted DNA using chloroform:isoamyl alcohol [@cubero_dna_1999], and spin columns (Qiagen, Hilden) (detailed protocol in Supplementary File S1). As a negative control we processed a sterile distilled water sample alongside each batch.

<!-- arent we following some extraction protocol here? cant we cite, and then have a supplementary protocol with our variation??? -->

<!-- To extract DNA from the crushed cells we added 800 mL  , 1% B-mercaptoethanol, 5 $\mu$L each of proteinase K and RNAase A, and incubated these at 65 &deg;C for 30 minutes. We spun samples at 10,000 g for 3 minutes to pellet cell debris, then added 700 $\mu$L of supernatant to an equal volume of 24:1 chloroform:isoamyl alcohol. We inverted samples to mix and centrifuged again at 12,000 g for 10 minutes. We precipitated DNA by transferring the top layer to 700 $\mu$L ice-cold ethanol, which we gently mixed, and spun over Qiagen DNA columns for 30 s at 15,000 g. We washed columns twice with 70% ethanol, and finally dissolved our DNA by spinning with 50 $\mu$L of sterile TE buffer.  -->

We designed custom primers to target a hypervariable region of snow algae *rbcL* (coding for the large subunit of rubisco). *rbcL* is more differentiated between snow algae than 18S (Supplementary Figure S3), and contains similar number of reference sequences on GenBank. Our *rbcL* primers target a 400 bp section of snow algae *rbcL* based off the consensus of 20 GenBank snow algae sequences (Supplementary Table S2).  We designed our primers with the Eurofins primer design tool (https://www.eurofinsgenomics.eu/en/ecom/tools/pcr-primer-design/). *Sanguina* sequences were not included because they were not available at the time. To target 18S, we used the universal primers Euk1181 and Euk1624 targeting the V7-V8 hypervariable regions [@wang_optimal_2014]. Primer sequences are available in Supplementary Table S3.

We constructed our 18S and *rbcL* amplicon libraries using a standard two-step PCR approach [@meyer_illumina_2010]. For detailed PCR conditions, see Supplementary File S2. After each PCR, we purified using Agencourt AMPure XP kit (Beckman Coulter, Brea, CA). We standardized sample DNA concentration using Qubit fluorometric quantification (Thermo Fisher, Waltham, MA). The pooled library was loaded and run on a MiSeq V3 kit (Illumina, San Diego, CA).

## Bioinformatic processing {-}

We demultiplexed reads using CUTADAPT [@martin_cutadapt_2011], and filtered and trimmed, removed errors, dereplicated, merged paired-end reads, and removed chimeras following the default pipeline of DADA2 [@callahan_dada2_2016]. We assigned taxonomy to amplicon sequence variants (ASVs) using IDTaxa [@murali_idtaxa_2018], only assigning names to ASVs that were assigned with high confidence (threshold set to 50). We assigned taxonomy using custom snow algae reference databases for both 18S and *rbcL*, in addition to re-assigning 18S with SILVA to classify non-snow algae [@quast_silva_2013]. Snow algae reference databases and the scripts to generate them are available on our github repository (see below).

<!-- should cite a github script here??? singletons, failed samples removed, remove sequences based on negative controls, adjust for 0's, sequencing depth, -->

<!-- make sure this is on GH: -->

<!-- We made our *rbcL* database by downloading 30,865 algae rbcL sequences from GenBank with the query “rbcl[gene] AND (algae OR chlorophyta OR trebouxiophyceae) NOT (18s OR ribosomal OR psaB OR atpB) AND 300:2000[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion”. We removed ambiguous snow algae annotations: because entries labelled as *Chloromonas cf. brevispina* and *Chloromonas cf. nivalis* are polyphyletic [@matsuzaki_taxonomic_2019] we re-labeled these as “unassigned *Chloromonas*”.  and then re-ran ASVs assigned to Chlorophyta on our custom 18S snow algae database based on the following GenBank query: "(18S OR ribosomal)[gene] AND (chloromonas OR chlainomonas OR sanguina OR raphidonema OR KMY-2018) NOT (rbcL OR psaB OR atpB) AND 300:3500[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion". -->

To generate OTUs we clustered ASVs using t-SNE [@maaten_visualizing_2008]. We compared t-SNE results with the perplexity parameter ranging from 1 to 100, and found comparable results from 10 to 50. We categorized OTUs by applying DBSCAN [@hahsler_dbscan_2019] to our t-SNE. We compared *rbcL* sample similarity using UniFrac with default settings [@lozupone_unifrac_2005]. To confirm that our clustering was not an artifact of t-SNE we created maximum likelihood phylogenetic trees of ASVs using IQTree [@nguyen_iq-tree:_2015] (Supplementary Figures S4-S5).

All raw sequence data are available under European Nucleotide Archive accession PRJEB34539. All scripts used in this study are available at https://github.com/cengstro/bc_snow_algae_amplicon.

# Results {-}

<!-- add read depth, etc? -->
<!-- make fig 5, ???, and make 18s otus -->

We observed morphologically distinct snow algae communities at different elevations (Figure 1). Red snow was prevalent in areas of high solar exposure above treeline, predominantly containing cells similar in appearance to *Sanguina nivaloides* [@prochazkova_sanguina_2019]. We occasionally observed green snow patches in alpine meadows receiving full sun, but more frequently observed green snow in well-shaded, forested sites. We found cells resembling *Chlainomonas rubra* [@novis_two_2008] at all elevations. Forested sites at low elevations often contained green or orange patches of snow, predominated by green or orange ovate cells with short spines resembling *Chloromonas cf. brevispina* [@matsuzaki_revisiting_2015].

Both 18S and *rbcL* amplicon libraries were dominated by ASVs that were assigned to Chlorophyta. Our 18S library contained 67 ASVs that were assigned to Chlorophyta and 7 ASVs to Ochrophyta, while our *rbcL* library contained 603 ASVs assigned to Chlorophyta and 41 to Trebouxiophyceae. Negative controls did not contain Qubit-detectable amounts of DNA. Our *rbcL* library contained higher ASV and OTU diversity than 18S: t-SNE clustering generated 9 *rbcL* OTUs, compared with just 4 18S OTUs (Supplementary Figure S6). The most frequent *rbcL* taxonomic assignments at genus level were *Chloromonas*, *Chlainomonas*, and *Sanguina*. Although most ASVs were not assigned to genus level, most ASVs clustered near one of these genera in NMDS ordination (Figure 2). Several *rbcL* OTUs were dominated by ASVs that were not assigned to genus level, including three OTUs closely related to *Chloromonas*, and one Chlamydomonadaceae OTU that was only assigned to the family level.

The composition of 18S and *rbcL* libraries varied with elevation (Figure 3; Figure 4). *Sanguina* was absent below 1500 m in both 18S and *rbcL*. High elevation samples contained an unannotated OTU of *Chloromonas* that was absent from low elevation sites (Figure 3, "Chloromonas C"). Additionally, the three highest elevation sites uniquely contained ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1)---interestingly, all of these samples were taken from snow overlying glacier. The two high alpine green snow samples were dominated by two different OTUs of *Chloromonas*, which were also both found in sites below treeline. *Chlainomonas* was prevalent at all elevations, and *Chloromonas krienitzii* was predominant in our low elevation sites.

We observed snow algae blooms dominated by *Chloromonas krienitzii* transition from green to orange from May to June (Figure 5). At two sites we observed snow melt channels or runnels containing high concentrations of snow algae. In May and early June these sites contained green snow hidden 2 to 5 cm below the white snow surface, but upon subsequent visits the same location contained orange snow at the surface, whose morphologies resembled published images of the *Chloromonas krienitzii* lineage [@matsuzaki_revisiting_2015]. We compared 18S and *rbcL* metabarcode profiles of green subsurface and orange surface snow, and found both contained high relative abundance of *Chloromonas krienitzii*. Although we did not detect *Chlainomonas* in cell counts, the surface samples contained nearly double the relative abundance of *Chlainomonas* from the subsurface samples.

<!-- include faiths diversity in supps??? thinking no  -->

# Discussion {-}

Here we present multiple data sets demonstrating elevational patterns in alpine snow algae blooms. *Sanguina* was dominant in red snow above treeline, while green and orange blooms of *Chloromonas krienitzii* were dominant in mid-elevation snow runnels. We found unexpected diversity within *Chloromonas* and *Chlainomonas* *rbcL* that we did not detect using 18S, including many potentially novel OTUs. In contrast to previous studies [@novis_two_2008;@prochazkova_ecophysiological_2018], our work suggests that *Chlainomonas* is widespread and abundant in alpine snow algae blooms.

Overall, our results show bloom composition can be highly variable on the same mountain, even in nearby patches. This is in contrast to previous papers that found snow algae assemblages were homogeneous across geographically distant sites [@lutz_biogeography_2016;@segawa_bipolar_2018]. This could be due to different sampling schemes: the aforementioned studies exclusively sampled high-latitude red snow, while our study included mid-latitude red, orange, and green snow from above and below treeline. Another recent study using metabarcoding found different snow algae communities in red and green snow algae blooms in Antarctica, including yellow blooms dominated by Chrysophyceae [@soto_microbial_2020]. Much of the diversity in our samples within *Chlainomonas* and *Chloromonas* was only detected with *rbcL* and not 18S, so the diversity within these genera may have been overlooked in previous metabarcoding studies.

Our *rbcL* data suggest *Chlainomonas* may be more common than previously thought. We could not differentiate between *Chlainomonas* and *Chloromonas* in our 18S data set, so this genus may have been overlooked in previous studies. *Chlainomonas* was previously thought to be restricted to waterlogged snow overlying mountain lakes [@novis_two_2008;@prochazkova_ecophysiological_2018], but we did not find any evidence that *Chlainomonas* was restricted in its habitat distribution. To the contrary, only one *Chlainomonas* dominant sample was located in waterlogged snow near the edge of a pool (sample S9); the other *Chlainomonas* dominant sites were not notably wetter than the surrounding snow, nor overlying frozen bodies of water. Interestingly, we found *Chlainomonas* was in higher relative abundance in our *rbcL* dataset than in cell counts (Figure 3)---possibly due to multiple chloroplasts per cell, which have been documented in this genus [@prochazkova_ecophysiological_2018].

Our findings highlight the remaining unexplored diversity in the snow algae microbiome. Many *rbcL* OTUs did not closely match any GenBank nucleotide archive sequences (Figure 2). "Chlamydomonadaceae E" highest BLAST percent identity was only 87%, and the top 10 BLAST matches contained 6 different algae genera---including two *Chloromonas* snow algae from Japan. This OTU did not correlate with any categories in our 18S or cell count data. One possible explanation is that it was lumped together with *Chloromonas* in 18S, and our crude quantification and categorization of cell counts did not detect this taxa. Other OTUs were also poorly matched to GenBank,  which could represent novel species of *Chloromonas* (Figure 2). The lack of clustering we observed in the remaining *Chloromonas* could be due to overlap in the levels of interspecific and intraspecific variation. *rbcL* diversity is likely higher within *Chloromonas* than in other algae genera: unlike *Sanguina* [@prochazkova_sanguina_2019], most species of *Chloromonas* lack a pyrenoid [@nozaki_differences_2002], which houses high concentrations of cross-linked rubisco. Likely related to its lack of pyrenoid, *Chloromonas* has many non-synonymous mutations in the region of *rbcL* the codes for binding rubisco together [@nozaki_differences_2002]. While *rbcL* is problematic in *Chloromonas* phylogeny [@nozaki_differences_2002], it nonetheless is highly differentiated and therefore an effective barcode for microalgae [@hall_assessment_2010].

<!-- *Sanguina* was found in the same samples, but relative abundance was generally higher in 18S and cell counts (Figure 3). Our *rbcL* primers were designed based on *Chloromonas* and *Chlainomonas* snow algae sequences (*Sanguina* sequence data was not available at the time), so primer bias could have resulted in over-representation of *Chloromonas* relative to *Sanguina*. -->

Variation in bloom composition could be due to a wide range of habitat features. *Sanguina* and *Chloromonas C* were limited to sites above 1500 m in full sunlight, but many low elevation sites also received full sunlight and did not contain these OTUs (Figure 3). Light intensity, snow moisture, or snow chemistry could all plausibly influence community composition. Intriguingly, we only observed *Raphidonema* at high-elevation glacier sites. In Svalbard, *Raphidonema nivale* abundance increased on glacier surface snow following wind storms, and the authors suggest that this is a soil algae that grows sub-optimally on the snow surface following wind deposition [@stibal_growth_2005]. Given the aerial dispersal capabilities of microalgae [@tesson_airborne_2016] and genetic overlap between distant snow algae populations [@segawa_bipolar_2018] it seems unlikely that geographic distance is a barrier to snow algae distribution. Underlying topography could potentially influence species distribution: two sites dominated by *Chloromonas krienitzii* were in runnels overlying ephemeral streams, which could influence nutrient availability where surface debris is deposited in the runnel.

Over multiple visits we observed motile green *Chloromonas krienitzii* cells below the snow surface develop into an orange bloom at the surface (Figure 5). Previous work has shown that this species can appear as distinct green and orange morphologies [@matsuzaki_revisiting_2015], but this paper is the first to document a bloom dominated by this species change color in the field. The transition over weeks suggests that this process is mediated by seasonal changes. Secondary pigments likely protect snow algae from the damaging effects of intense solar irradiation at the snow surface [@bidigare_evidence_1993], which could be why we only observed green cells below the snow surface.

In conclusion, our study reveals substantial diversity within and between snow algae blooms. We found greater diversity using *rbcL* metabarcoding than 18S, which suggests snow algae diversity may have been overlooked in studies that used only 18S. Species were differentially distributed along an elevational gradient, with distinct blooms dominated by *Chloromonas*, *Sanguina*, or *Chlainomonas*. Future studies will examine other taxa present in snow algae blooms, and their interactions in these globally important microbiomes.

# Acknowledgments {-}

We thank Leah Tooman (Simon Fraser University) for assistance with sequencing, and Chris Rushton (Simon Fraser University) for assistance with bioinformatics. This project was funded with a Sector Innovation Grant from Genome BC (SIP016), and a NSERC Individual Discovery Grant, both awarded to LQ.

# Author Contributions {-}

CE, LQ, and KY designed this study. Samples collected by CE with assistance from KY and LQ. KY and CE prepped samples for sequencing. All analyses were completed by CE and KY. All authors discussed the results and contributed to the final manuscript. CE wrote the manuscript with major input from LQ and KY.

# Additional information {-}

The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.

##### pagebreak

# References {-}
<div id="refs"></div>

##### pagebreak

# Figure legends {-}

<!-- formatting indentation??? -->

**Figure 1.** Representative photographs of snow algae. A. Red snow bloom above treeline at sample site G1.1. B. Microphotograph of red snow containing *Sanguina nivaloides* (Sn) and *Chlainomonas rubra* (Cr) cell morphologies. All scale bars 30 $\mu$m, all microphotos taken using 63x objectives with DIC. C. Snow runnels in a forest clearing containing snow algae (inset). D. Microphotograph of orange snow from the surface of runnel containing *Chloromonas krienitzii* (Ck) and *Chloromonas cf. nivalis* (Cn). E. Bronze coloured snow algae blooms below conifer canopy. F. Microphotograph of bronze snow containing Chloromonas cf. brevispina* (Cb) and *Chloromonas cf. nivalis* (Cn).

\setlength{\leftskip}{0pt}
**Figure 2.** Ordination plots of *rbcL* ASV similarity. A. Multidimensional scaling (MDS) plot. Taxonomy (assigned by IDTaxa) indicated by color, and size of point is proportional to cumulative relative abundance. Dotted ellipses show OTU clusters. Stress = 0.13. B. t-Distributed Stochastic Neighbor Embedding (t-SNE) dimensionality reduction of snow algae *rbcL* ASVs. Dotted lines indicate OTUs. Perplexity = 30.

**Figure 3.** Stacked bar plots of snow algae relative abundance. Each sample is arranged on the y axis in order of elevation, with each compositional bar plot representing the (left to right) cell morphology, 18S, and *rbcL* snow algae assemblages of the same sample. Colors below each bar plot indicate taxonomic categories. Morphospecies were identified by light microscopy, using similarity to published photographs as a guide. 18S OTUs are reference-based, while *rbcL* OTUs were clustered *de novo*, and named using GenBank reference data.

**Figure 4.** Non-metric multidimensional scaling (NMDS) showing *rbcL* UniFrac distances between samples. Each point represents a sample, which is labelled by sample ID and elevation (color).

**Figure 5.** ??? Need to make the figure first

### 2020-01-10 manuscript/etc/trash/Casey1.Rmd
directory cleanup

--- 2019-11-01 manuscript/Casey1.Rmd
@@
 output: 
-  word_document:
-    # toc: yes
+  bookdown::word_document2:
+    fig_caption: yes
     pandoc_args:
@@
   pdf_document: 
+    fig_caption: yes
     number_sections: yes
-    # toc: yes
-    # toc_depth: 4
     pandoc_args:
@@
       - '--lua-filter=templates/author-info-blocks.lua'
+csl: article.csl
 bibliography: /home/cengstro/ownCloud/mendeley_library/bibtex_by_grp/casey1.bib
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, root.dir = here::here())
 ```
@@
 
+Our *rbc*L amplicon library was dominated by sequences closely related to *Chloromonas*, *Chlainomonas*, and *Sanguina* (Fig. 2).
+
+60% and 30% of *rbc*L sequences were assigned to genus and species level, while only % and % of our 18S amplicon library was assigned respectively.
+However, our 18S library generally supported our rbcL findings: samples dominated by *Sanguina* were consistant between markers, however samples containing *Chlainomonas* rbcL were assigned to *Chloromonas* 18S, which could be due to lack of differentiation between the the two genera using 18S (Supp.
+Fig.
+X).
+
+*Chloromonas* was the most diverse
+
+Algae OTU8 has a best BLAST score of 88%, with top hits in both Trebouxiophyceae and Chlorophyceae.
+
+Currently there are just two species *Chlainomonas* in reference databases, but our results suggest there may be genetically distinct populations *Chlainomonas* at higher elevations (Fig.
+1, Supp.
+Fig.
+1, Supp.
+Fig 2).
+
+There were fewer *rbc*L ASVs in *Sanguina* than in *Chloromonas* and *Chlainomonas*, and there were no visible sub-clusters within *Sanguina* in our t-SNE or phylogenetic tree (Fig. 2).
+
 ### Snow algae community composition varies with date and elevation
@@
 
-### High elevation, high diversity
+Sites at higher elevation had higher *rbc*L diversity \@ref(fig:phylo-diversity).
 

### 2020-01-10 manuscript/etc/trash/_main.Rmd
directory cleanup

--- 2020-01-10 manuscript/etc/trash/Casey1.Rmd
@@
 ---
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, root.dir = here::here())
-```
-
-<br>
-<br>
-<br>
-
-### *Keywords*: snow algae rbcL 18S amplicon microbiome alpine community
-
-### *Run title*: Elevational patterns in snow algae communities
-
-### *To be submitted to*: 1) Nat Comm, 2) Frontiers in Micb 3) Env Micb 4) Microbial Ecology
-
-##### pagebreak
 
@@
 
-where we were: there are red and green blooms
-where we are: there are multiple types of red and green blooms
+Placeholder
 
-Snow algae are cold-adapted microalgae that bloom in melting summer snow throughout alpine and polar regions [@Leya2013].
-Red snow algae blooms can cover vast areas of snowfield, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017].
-Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], which can threaten water supplies [@Staudinger2014] and worsten drought [@Barnhart2016].
-Thus, snow algae could play an important role in snow melt regulation, but little is known about how these blooms form.
-We don't know the spatial extent of snow algae blooms worldwide, what conditions trigger snow algae bloom formation, or how cells arrive on the surface of the snow to "seed" the bloom.
-Critically, we lack basic understanding of the species composition and diversity in snow algae blooms, which is fundamental to answering these long-standing mysteries about snow algae blooms.
-
-Historically snow algae were classified by cell morphology, however, morphology can be deceptive when identifying microalgae.
-Until recently, all red cells from red snow were referred to as *Chlamydomonas cf.
-nivalis* [@Remias2013;@Segawa2018], desipte the fact that this morphology-based species was thought to contain multiple distinct species [@Kol1961].
-Recent genetic work designated found that some of these cells were not closely related to *Chlamydomonas* at all, and were designated as a new genus and species: *Sanguina nivaloides* [@Prochazkova2018c].
-Other genetic work shows that morphology-based species of snow algae may in fact contain multiple genetically unrelated clades: for example, *Chloromonas cf.
-nivalis* contains at least four distinct clades [@Matsuzaki2019].
-
-<!-- hammer this home: WE LACK BASIC UNDERSTANDING ABOUT SPECIES COMPOSITOIN AND DIVERSITY -->
-
-<!-- sequence cultures, and batch, and single cell, but this dosent tell us whats actually in the field -->
-More recently, snow algae taxonomy has
-
-<!-- high throughput: lacking ref data, no otu based, none in alpine environments (despite this tantalizing evidence of microscopy based, etc) -->
-
-<!-- But no one until now has used high throughput sequencing to compare snow algae blooms in alpine environments, or used higher resolution marker genes to do so -->
-
-<!-- so we did this... -->
-In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS (explain what these code for).
-Our aims were to: 1) see what species were present, 2) compare community composition between samples.
-We found ...
-
-<!-- the red green problem -->
-The same species can look completely different depending on the life stage [@Shah2016;@Salome2019]; conversely, different species can look nearly identical.
-For example, cell morphologies known as *Chloromonas cf.
-nivalis* appear in at least four different lineages.
-One of these lineages, *Chloromonas muramotoi* appear as green biflagellates in culture, but in the field the same species features thick cell walls with flanges [@Matsuzaki2019].
-However, life cycle changes in the the cosmopolitan and widespread genus *Sanguina* have yet to be documented: despite over a century of study [@Kol1961] no one has sucessfully cultured this genus.
-
-<!-- habitat differences -->
-Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013].
-In this micro-habitat, temperature is nearly constant, but snow moisture, pH, nutrients, and light can differ greatly within the snowpack depending on location of snow and time of year [@Loranger1988;@Kuhn2001;@Gorton2007;@Ganey2017].
-Differences in snow algae community composition could offer insight into the environmental factors that determine snow algae distribution.
-Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a].
-Two studies using microscopy found that red snow containing *Chlamydomonas cf.
-nivalis* was prevalent in the above timberline, while green and orange snows caused by *Chloromonas cf.
-brevispina* and *Chloromonas cf.
-nivalis* were prevalent in conifer forests below timberline [@Hoham1979a;@Nedbalova2008a].
-Differences in snow moisture could influence which species of snow algae grow; several species of snow algae are anecdotally reported to only inhabit very saturated, waterlogged snow [@Lutz2015;@Remias2013;@Novis2008;@Prochazkova2018c], while  *Chlamydomonas cf.
-nivalis* is hypothesized to be specially adapted to well-drained, drier snow [@Ganey2018].
-
-<!-- habitat differences, short new form -->
-  Red snow is also occasionally reported to be caused by another red-pigmented algae: *Chlainomonas* [@Novis2008].
-Little is known about the distribution of this genus; one study suggested that *Chlainomonas* is limited to high-moisture snowpack above alpine lakes [@Prochazkova2018].
-Green snow may contain a variety of algae, notably *Chloromonas* (refs), as well as algae from other families including Trebouxiophyceae and Chrysophyceae (refs).
-Morphology-based cell counts suggest that green snow caused by *Chloromonas* is prevalent in forested areas and red snow containing *Chlamydomonas cf.
-nivalis* is prevalent in alpine areas [@Hoham1979a;@Nedbalova2008a]; however, morphology can be misleading when identifying microalgae.
-
-Although much of the snow algae literature focusses on eye-catching red snow, different species of snow algae can also color snow green [@Lutz2015].
-
-<!-- hts -->
-Microscopic observation shows that snow algae blooms can contain a diversity of cell morphologies in the same sample [@Lutz2018], but the genetic community structure of snow algae may not correlate with what is seen under the microscope.
-18S amplicon high-throughput sequencing (HTS) shows that snow algae may contain more diversity than meets the eye: although red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a].
-18S is highly conserved and therefore suited to survey broadly across a variety of taxa; however, due to its high similarity between closely related species short segments of thie gene are poorly-differentiated among closely related taxa (ref).
-The *rbc*L plastid gene coding for the large subunit of RuBisCO is highly differentiated among photosynthetic organisms, and contains reference sequences for snow algae, making it an ideal complement to survey snow algae communities.
-
-Until recently, snow algae sequences were lacking in reference databases,
-
-# Results {-}
-
+### *Keywords*: snow algae rbcL 18S amplicon microbiome alpine community
+### *Run title*: Elevational patterns in snow algae communities
+### *To be submitted to*: 1) Nat Comm, 2) Frontiers in Micb 3) Env Micb 4) Microbial Ecology
+##### pagebreak
 ### Morphologically diverse snow algae blooms across a variety of habitats
-
-Snow algae was widespread in our study area, beginning on May _ when we first observed snow algae blooming in the trees.
-Over the course of the season, we collected 309 samples from 33 dates and 13 moutains.
-We did not observe pink snow above treeline until June 6, despite several visits above treeline before that (site data table).
-We observed snow algae until the first snows of winter, although by later in the season most of the snow had melted out at lower elevations.
-
-Visually, several patterns in snow algae site characteristics and cell morphology were apparent.
-Most frequently, we observed pink or red snow, typically in areas recieving high sunlight, dominated by cells that looked similar to *Sanguina nivaloides*, frequently containing green cells of varying shapes and sizes.
-However, red snow could also be dominated by *Chlainomonas* morphologies, in red snow that was visually indistinguishible from red snow containing primarily *S.
-nivaloides*.
-*Chlainomonas* was frequently mixed in with *S.
-nivaloides* as well (Fig.
-1d).
-We also occasionally observed green snow above treeline, which was dominated by green or yellow oval cells with flanges, similar to those described for *Chloromonas muramotoi*.
-Colored snow often appeared in snow runnels, frequently at elevations near or below treeline.
-These runnels containing colored snow formed linear streaks, often with orange on the surface and green snow just below the surface (Fig.
-1b).
-At two sites like this that we visited repeatedly, the runnel began with green snow hidden 1-2 cm below the surface, but on subsequent visits the surface snow had changed color to orange.
-The subsurface snow contained green, often flagellated algae, while surface snow was dominated by orange cells resembling *Chloro.
-krienitzii* (Fig.
-1e) [@Matsuzaki2015].
-Below treeline, orange and green snow was common, often dominated by orange oval cells with short spines, similar to *Chloromonas cf.
-brevispina* (Fig.
-1g).
-Snow dominated by *Chloromonas cf.
-brevispina* and *Chloromonas cf.
-nivalis* was abundnat in shady areas below treeline, often in distinctively bronze colored snow (supp data).
-
 ### Diverse *rbc*L reveals species-level snow algae taxonomy
-
-Our *rbc*L amplicon library was dominated by sequences closely related to *Chloromonas*, *Chlainomonas*, and *Sanguina* (Fig. 2).
-
-60% and 30% of *rbc*L sequences were assigned to genus and species level, while only % and % of our 18S amplicon library was assigned respectively.
-However, our 18S library generally supported our rbcL findings: samples dominated by *Sanguina* were consistant between markers, however samples containing *Chlainomonas* rbcL were assigned to *Chloromonas* 18S, which could be due to lack of differentiation between the the two genera using 18S (Supp.
-Fig.
-X).
-
-*Chloromonas* was the most diverse
-
-Algae OTU8 has a best BLAST score of 88%, with top hits in both Trebouxiophyceae and Chlorophyceae.
-
-Currently there are just two species *Chlainomonas* in reference databases, but our results suggest there may be genetically distinct populations *Chlainomonas* at higher elevations (Fig.
-1, Supp.
-Fig.
-1, Supp.
-Fig 2).
-
-There were fewer *rbc*L ASVs in *Sanguina* than in *Chloromonas* and *Chlainomonas*, and there were no visible sub-clusters within *Sanguina* in our t-SNE or phylogenetic tree (Fig. 2).
-
 ### Snow algae community composition varies with date and elevation
-
-This is supported by our 18S and cell count data. Although we found Chlainomonas
-
-Sites at higher elevation had higher *rbc*L diversity \@ref(fig:phylo-diversity).
-
 ### Green snow algae turns orange at the surface
-
-**A gradient between snow algae communities**.
-Variation in snow algae communities was primarily driven by the presence or absence of *Sanguina nivaloides*, *Chloromonas krienitzii*, *Chlainomonas rubra*, or *Chlainomonas* OTU3 (Fig.
-2a).
-However, most samples contained a mix of species; few samples were relatively "pure".
-However, sample comparison with weighted UniFrac [@Lozupone2005] found little difference between *Chlainomonas* and *Chloromonas* dominated sites, and the presence or absence of *Sanguina* was the driving factor, which was generally found in higher relative abundance at higher elevation (Fig.
-2b).
-
-Although *Chlainomonas* is reported to only grow in slushy snow above alpine tarns [@Novis2008;@Prochazkova2018c], we found *Chlainomonas* was present in low relative abundance at most sites sampled.
-Only one site high in Chlainomonas was near water (S9), the other Chlainomonas dominated sites were not near water, nor was the snow noticeably mushier than the surrounding snow.
-This does not rule out the possibility that these sites contained high moisture at some point; further sampling is needed to determine its habitat distribution.
-
-#
-
-**Snow algae community composition co-varies with date and elevation**.
-*Sanguina* was more abundant in alpine sites, as well as one OTU of *Chlainomonas* and two OTUs of *Chloromonas*, while *Chloromonas krienitzii* and Chlainomonas OTU6 were found in high relative abundance below treeline (Fig.
-2, 3).
-This consistent with the microscopy observations of @Hoham1979a and @Nedbalova2008a, who found primarily *Chlamydomonas cf.
-nivalis* above treeline and *Chloromonas* below treeline.
-However, in microscopy surveys we infrequently observed *Sanguina* below treeline, and *Chloromonas krienitzii* above treeline.
-We found higher diversity in samples from higher elevation, both at the ASV level and OTU level <!--Chao1 w fossil.-->, with several OTUs that were not found at lower elevations.
-*Raphidonema* and related ASVs were restricted to high alpine sites, with the highest relative abundance occurring in samples from August and September.
-This fits with the interpretation of @Stibal2005, who suggest that *Raphidonema* is an opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow.
-However, rare taxa should not be discounted as playing an important role in ecosystem functioning [@Jousset2017].
-
-Snowpack above and below treeline are very different habitats, with forest providing shade (and therefore less evaporation).
-The red pigment of *Sanguina* and *Chlainomonas* (astaxanthin) is likely an adaptation to high light [@Gorton2003;@Holzinger2006], and it has been suggested that *Sanguina* is adapted to the drier snowpacks that form at higher elevations [@Dial2018c].
-
-#
-
-**_Chloromonas krienitzii_ is green below, turns orange at surface**.
-On initial visits to Chloromonas krienitzii dominated sites, green snow was hidden 2 to 5 cm below the surface of the runnel (Supp.
-Fig.
-4).
-Green snow was noticeably moister than surrounding snow, in some locations containing ice lenses not found in the surrounding snowpack.
-By June 6 the surface of the runnel had turned orange while the snow below remained green, which persisted on three subsequent visits until the snow melted.
-Cell morphology at the surface closely resembled *Chloromonas cf.
-brevispina* “Trochiscia” [@Matsuzaki2015] (Fig.
-4), while green subsurface snow was dominated by green biflagellate or bipolar quadriflagellate cells (Fig.
-4).
-Both green and orange samples were dominated by reads assigned to *C.
-krienitzii*, with lower relative abundance of *Chlainomonas*, which was more abundant at the surface (Supp.
-Fig.
-3).
-
-One possibility is that C.
-krienitzii cells orange pigmentation is triggered or an adaptation to higher light levels at the surface.
-Green cells below the surface would be protected from UV radiation<!--cite -->.
-Possibly, the green cells maintain an optimal light level through phototaxis, similar to marine phytoplankton<!--cite-->.
-Perhaps higher elevations typically have fewer low angle runnels that would allow a moist enough environment for *C.
-krienitzii*.
-It seems unlikely that nutrients from conifer needles would explain this, as several of our *C.
-krienitzii* dominated sites were in full sunlight, far from any trees.
-
-#
-
-**Snow algae blooms harbor novel diversity**.
-Several OTUs were not assigned to species level due to low similarity (Fig.
-1).
-Two OTUs of Chloromonas, one OTU of Chlainomonas, and one OTU ("Algae OTU8") that had best BLAST matches to both Trebouxiophyceae and Chlorophyceae algae.
-Algae OTU8 has a best BLAST score of only 88%; this could represent a free-living algae found in snow, or possibly an endosymbiotic algae from a ciliate, as we noted many ciliates with green organelles in our microscopy samples.
-Sample X17 contained was dominated by spherical red cells superficially similar to *Sanguina*, this sample was dominated by reads of Chloromonas OTU7 and Algae OTU8.
-Cells in this sample had thicker cell wall, and some cells were slightly square-shaped (Fig.
-4).
-18S algae reads from this sample were equivocal, with best BLAST matches to both *Raphidonema* and *Chloromonas*.
-Currently there are just two species *Chlainomonas* in reference databases, but our results suggest there may be genetically distinct populations *Chlainomonas* at higher elevations (Fig.
-1, Supp.
-Fig.
-1, Supp.
-Fig 2).
-Although some of the high elevations ASVs were classified as *Chlainomonas rubra*, because we only included this species in our custom snow algae reference this could have biased our classifier, possibly causing over-classification [@Murali2018].
-
-Although our rbcL primers successfully detected distinct OTUs, this is likely an underestimate of the true diversity.
-We did not detect any clustering within *Sanguina* using either 18S or rbcL, however when we included the homologous rbcL fragment of *S.
-aurentia* in our analysis it did not form a distinct OTU <!--this is where I am quesy about out ghetto OTU method-->.
-Furthermore, rbcL is highly differentiated in *Chloromonas* relative to other microalgae [@Nozaki2002], which could partly explain why *Sanguina* is less differentiated than *Chloromonas* or *Chlainomonas* (Supp.
-Fig.
-1 and 2).
-
-# Discussion {-}
-
-We found snow algae communities were typically dominated by either *Sanguina*, *Chlainomonas* or *Chloromonas*, with secondary abundance of *Chlainomonas* or *Chloromonas* in most samples.
-Most samples contained a sizable proportion of secondary or tertiary OTUs, although some some contained nearly 100% *Chlainomonas rubra*, *Sanguina nivaloides*, or *Chloromonas krienitzii*.
-The drivers of these differences between communities remain unclear; it is likely that within the melting snowpack there may be distinct microhabitats that impacts snow algae distribution, with varying water content with topography, different light levels in shade or at depth [@Gorton2007], different daily freeze-thaw cycles, or possibly varying nutrient levels with proximity to vegetation [@Hoham1976] or bedrock [@Kol1968a].
-Further work needs to be done to uncover the micro-habitat niches within the snowpack that allow this differentiation, and what novel species the snow microbiome might contain.
-
-# Methods {-}
-
-**Field sampling and microscopy.** We collected snow algae from mountains near Vancouver, British Columbia throughout the summer of 2018.
-To capture the extent of snow algae diversity in our local mountains we collected as many samples as possible from different elevations, dates, mountains, and micro-habitats within the snow <!--insert suppl table-->.
-In total we collected 310 colored snow samples from 13 different mountains on 33 different dates from elevations between 880 m and 2150 m above sea level.
-In early season we sampled at lower elevations, moving uphill as the snow algae bloom progressed upwards in elevation [@Takeuchi2013].
-We observed the first small spots of red snow above treeline on June 20.
-
-We scooped samples from visibly colored snow into 50 mL centrifuge tubes using sterile technique.
-To prevent melting en route to the lab we stored tubes in a bag of snow.
-Back in the lab, we melted each sample at room temperature on the bench and removed a 1 mL aliquot for light microscopy.
-Immediately after, samples were stored at -20 &deg;C for up to eight months until DNA extraction.
-Most samples were viewed using light microscopy, characterizing the dominant cell morphology in each sample.
-We prepared a slide of cells fixed in 2% gluteraldehyde, and quantified the most common cell morphologies with a cell count to 100, classifying each cell as either *Sanguina*, *Chloromonas cf.
-nivalis*, *Chloromonas cf.
-brevispina*, *Chloromonas krienitzii* *Chlainomonas*, or "other".
-
-#
-
-**DNA extraction and amplicon library preparation.** We selected 35 out of 310 samples for rbcL and 18s high-throughput amplicon sequencing.
-We chose these samples to represent the variation in date, elevation, geographic location, snow color, micro-habitat, and cell morphology.
-We freeze-dried these samples for up to 48 hours until until samples appeared completely desiccated.
-To allow water vapor to escape sample tubes while freeze drying we poked holes in each lid; as a safeguard against cross-contamination from airborne snow algae powder we wrapped tubes individually with paper towels.
-We mini-pestled 5 mg of sample at room temperature to physically rupture cell walls.
-Based on our observation that samples dominated by *Sanguina* cell morphologies had lower DNA yields, and had many intact cells following various lysis methods, we used 20 mg of these samples to increase our DNA yield.
-
-To extract DNA from the crushed cells we added 800 mL 1x CTAB extraction buffer [@ColdSp2009], 1% B-mercaptoethanol, 5 $\mu$L each of proteinase K and RNAase A, and incubated these at 65 &deg;C for 30 minutes.
-We centrifuged samples at 10,000 g for 3 minutes to pellet cell debris, then added 700 $\mu$L of supernatant to an equal volume of 24:1 chloroform:isoamyl alcohol.
-We inverted samples to mix and centrifuged again at 12,000 g for 10 minutes.
-We precipitated DNA by transferring the top layer to 700 $\mu$L ice-cold ethanol, which we gently mixed, and spun over Qiagen DNA columns for 30 s at 15,000 g.
-We washed columns twice with 70% ethanol, and finally dissolved our DNA by spinning with 50 $\mu$L of sterile TE buffer.
-As a negative control we processed a sterile distilled water sample alongside each batch, treating it exactly the same as the other tubes.
-
-We used rbcL as a marker gene, which offers high resolution between microalgae species [@Zou2016], and has reference data for snow algae available on GenBank.
-We also sequenced each sample with 18S primers, due to it's coverage across a wide range of taxa, and its ubiquity in reference databases.
-We designed rbcL primers to target an approximately 400 bp section of this gene, based on 20 snow algae rbcL GenBank sequences from *Chloromonas* and *Chlainomonas* (GenBank accession numbers AB434272.1, LC012752.1, LC012747.1, AF517072.1, LC012738.1, LC012739.1, AB434267.1, EU030690.1, LC360494.1, AJ001878.1, AB022225.1, DQ885964.2, DQ885962.1, AJ001879.1, AB022226.1, AB022530.1, LC012751.1, AB504764.1, EU030689.1, AB101508.1).
-*Sanguina* sequences were not included because they were not available at the time.
-We designed primers with PRIME(+) from the GCG package.
-These primers were rbcL369F (5’-GAA CGT GAC AAA TTA AAC AAA-3’) and rbcL870R (5’-ACC WGA YAD ACG WAG AGC TT-3’).
-To target 18S we used Euk1181 (5'-TTA ATT TGA CTC AAC RCG GG-3') and Euk1624 (5'-CGG GCG GTG TGT ACA AAG G-3') [@Wang2014].
-
-We constructed our amplicon library using a two-step PCR [@Meyer2010].
-In the first PCR we amplified template DNA using our primers attached to a universal adapter, and in the second PCR we re-amplified that product to attach a 6 bp index to the universal adapter at the 3’ end.
-The first PCR total volume was 25 $\mu$L, consisting of 1 $\mu$L template, 12.5 $\mu$L Q5 high-fidelity 2X MM (New England BioLabs), 1.25 $\mu$L each of forward and reverse primer, and 9 $\mu$L of ddH2O.
-The second PCR was the same except we reduced our reaction volume to 20 $\mu$L by using only 5 $\mu$L of ddH2O.
-The cycling conditions were the same for both primer pairs for the first PCR, with an initial denaturation at 98 &deg;C for 30 s, followed by 30 cycles of 98 &deg;C for 5 s, 58 &deg;C for 10 s, and 72 &deg;C for 25 s, with a final extension at 72 &deg;C for 2 minutes.
-For the second indexing PCR we started with an initial denaturation at 98 &deg;C for 30 s, then 10 cycles of 98 &deg;C for 10 s, 65 &deg;C for 30 s, and 72 &deg;C for 30 s with a final denaturation of 72 &deg;C for 5 min.
-After each PCR, we purified using Agencourt AMPure XP kit (Beckman Coulter).
-We quantified final DNA concentration with Qubit (Thermo Fisher), and standardized sample concentration for pooling.
-The pooled library was then loaded and run on an Illumina MiSeq V3 kit.
-
-#
-
-**Bioinformatic processing.** Reads were demultiplexing using cutadapt [@Martin2011].
-Samples were filtered and trimmed, errors removed, dereplicated, pair end reads merged, and chimeras removed in DADA2 [@Callahan2017].
-<!-- singletons, failed samples removed, remove sequences based on negative controls, adjust for 0's, -->
-
-We assigned taxonomy using a custom reference databases for both rbcL and 18S, since snow algae are not well-represented in SILVA or PR2, and there is no curated database for rbcL.
-We made our rbcL database by downloading 30,865 algae rbcL sequences from GenBank with the query “rbcl[gene] AND (algae OR chlorophyta OR trebouxiophyceae) NOT (18s OR ribosomal OR psaB OR atpB) AND 300:2000[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion”.
-We removed ambiguous snow algae annotations based on the most recent snow algae phylogeny [@Matsuzaki2019].
-Because entries labelled as *Chloromonas cf.
-brevispina* and *Chloromonas cf.
-nivalis* are polyphyletic we re-labeled these as “unassigned *Chloromonas*”.
-<!--Where multiple sequences were present for a single species we generated a consensus sequence to avoid imbalance where some groups have far more representatives than others [@Murali2018] --><!--To  re-labeled GenBank accessions LC012733.1 and LC012735.1 as Chloromonas krienitzii, LC012734.1 as Chloromonas “Cryocystis”-type, and LC012744.1 to Chloromonas miwae, and LC012743 and LC012745 to “Chloromonas Group_B” 6`.--> Species designated by morphology alone that were not genetically different from other species in their genera were annotated only to genus level.
-<!-- all Chlainomonas to genus level only --> We assigned rbcL ASV taxonomy using IDTaxa with threshold = 50.
-
-To assign 18S reads we initially ran our ASVs against SILVA [@Quast2013], and selected ASVs assigned to Chlorophyta.
-We then ran this subset of ASVs on our custom 18S snow algae database based on the following GenBank query: "(18S OR ribosomal)[gene] AND (chloromonas OR chlainomonas OR sanguina OR raphidonema OR KMY-2018) NOT (rbcL OR psaB OR atpB) AND 300:3500[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion".
-This database was edited to reflect the most recent snow algae taxonomy and remove ambiguous annotations [@Matsuzaki2019;@Prochazkova2018c].
-We assigned taxonomy using IDTaxa with threshold = 50.
-
-We assigned OTUs with t-SNE, based on visually distinct clusters [@Graham-Rowe1999] <!--BibTeX header messed up-->(Fig.
-1).
-We split the largest cluster (containing the vast majority of the *Chloromonas* reads) into three clusters, one containing all ASVs assigned to *Chloromonas krienitzii*, and one containing all ASVs assigned to *Chloromonas muramotoi* and *Chloromonas hohamii*.
-To highlight our observation that one clade of *Chlainomonas* were only found above treeline (Suppl.
-Fig.
-1), we split *Chlainomonas* into two OTUs.
-The OTUs produced by this naive method were roughly consistent across different methods of visualization (Suppl.
-Fig.
-1 and 2).
-<!--Other OTU classifiers such swarm [@Mahe2014] are optimized for 18S data, and we found swarm did a poor job of clustering when visualised with MDS, t-SNE, and phylogenetic trees, with various values of d.
-<!--and CROP [@Chen2013].-->
-<!--OTU clustering with t-SNE: based on poor results with other classifiers, we used tsne.
-since tsne is known to find patterns where there are none depending on the perplexity, we ran this test with perplexity ranging from 5-100, and found the same large scale clusters each time.
--->
-
-**Software information.** All analysis was conducted in R 3.6.1 with tidyverse 1.2.1, with dada2 1.12.1, DECIPHER 2.12.0, Rtsne 0.15, GUniFrac 1.1, and vegan 2.5-6.
-Phylogenetic analysis was carried out in IQTree [@Nguyen2015] and visualized with ggtree [@Yu2017].
-
-**Data availability.** All raw fastq files are freely available on the European Nucleotide Archive under the project accession PRJEB34539.
-All scripts are available at cengstro.github.io/projects/sa_rbcl.
-
+##### pagebreak
+### *Supplementary information*.
+### *Competing financial interests.* The authors declare that the research was conducted in the absence of any
 ##### pagebreak
 
-# Acknowledgments {-}
+<!--chapter:end:Casey1.Rmd-->
 
-We wish to thank Leah Tooman (Simon Fraser University) for assistance with sequencing, and Chris Rushton (Simon Fraser University) for assistance with bioinformatics.
-<!--We also wish to thank Forest Rohwer (UCSD) for the inspiration that led to this project.--> This project was funded with a Sector Innovation Grant from Genome BC (SIP016), and a NSERC Individual Discovery Grant, both awarded to LQ.
+---
+title: "Distinct snow algae communities along an elevational gradient"
+author:
+  - name: Casey B. Engstrom
+    institute: SFU
+  - name: Kurt M. Yakimovich
+    institute: SFU
+  - name: Lynne M. Quarmby
+    institute: SFU
+    email: quarmby@sfu.ca
+institute:
+  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
+date: "`r format(Sys.time())`"
+output: 
+  bookdown::word_document2()
+    fig_caption: yes
+    pandoc_args:
+      - '--lua-filter=templates/scholarly-metadata.lua'
+      - '--lua-filter=templates/author-info-blocks.lua'
+    reference_docx: templates/word-template.docx
+  pdf_document: 
+    fig_caption: yes
+    number_sections: yes
+cls: templates/article.cls
+bibliography: casey1.bib
+abstract: |
+  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate.
+Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch on the same mountain, and varies with elevation and date.
+We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada.
+We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m).
+Samples from higher elevation had higher phylogenetic diversity, and included several potentially novel clades of algae.
+Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface.
+Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome.
+---
 
-# Author Contributions {-}
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
 
-<!--
+# header 1
+## header2
 
-When determining authorship the following criteria should be observed:
+```{r}
+plot(pressure)
+```
 
- - Substantial contributions to the conception or design of the work; or the
-   acquisition, analysis, or interpretation of data for the work; AND
- - Drafting the work or revising it critically for important intellectual
-   content; AND
- - Final approval of the version to be published ; AND
- - Agreement to be accountable for all aspects of the work in ensuring that
-   questions related to the accuracy or integrity of any part of the work are
-   appropriately investigated and resolved.
-
-Contributors who meet fewer than all 4 of the above criteria for authorship
-should not be listed as authors, but they should be acknowledged.
-(http://www.icmje.org/roles_a.html)
-
--->
-
-CE, LQ, and KY designed this study.
-Samples collected by CE with assistance from KY and LQ.
-KY and CE prepped samples for sequencing.
-All analyses were completed by CE and KY.
-All authors discussed the results and contributed to the final manuscript.
-CE wrote the manuscript with major input from LQ and KY.
-<!-- The statement about the authors and contributors can be up to several sentences -->
-<!-- long, describing the tasks of individual authors referred to by their initials -->
-<!-- and should be included at the end of the manuscript before the References -->
-<!-- section. -->
-
-# Additional information {-}
-
-### *Supplementary information*.
-
-### *Competing financial interests.* The authors declare that the research was conducted in the absence of any
-commercial or financial relationships that could be construed as a potential
-conflict of interest.
-
-##### pagebreak
-
-# References {-}
-<div id="refs"></div>
+<!--chapter:end:manuscript.Rmd-->

### 2020-01-10 manuscript/figs/etc/figures.Rmd
directory cleanup

--- 2019-11-16 manuscript/figures.Rmd
@@
 asv_similarity <- include_graphics(here("manuscript/figs/main/taxonomy/asvs2.png"))
-communities <- include_graphics(here("manuscript/figs/main/community/community_ggplot_polished_2.png"))
+communities <- include_graphics(here("manuscript/figs/stacked_barplot.png"))
 unifrac <- include_graphics(here("manuscript/figs/main/uf_plot.png"))
@@
 
-heatmap_cap <- "Relative abundance heatmap of snow algae community composition using cell morphology, 18S, and \\emph{rbc}L.
-Samples on the x-axis are arranged from low to high elevation.
-Taxonomic designations for \\emph{rbc}L are based on \\emph{de novo} OTU clustering, while taxonomic assignments for 18S are reference-based.
-Samples lacking data for a particular dataset are shown as blank (e.g.
-G1.4 18S)."
+heatmap_cap <- "Stacked barplot showing relative abundance community composition using cell morphology, 18S, and \\emph{rbc}L.
+Taxonomic designations for \\emph{rbcL} are based on \\emph{de novo} OTU clustering, while taxonomic assignments for 18S are reference-based.
+Samples on the y-axis are arranged in order of elevation."
 
@@
 
-```{r rel_abund, fig.cap = heatmap_cap}
+```{r rel_abund, fig.cap = heatmap_cap, fig.width=6, fig.height=5}
 communities

### 2020-01-10 manuscript/supp/etc/supplemental.Rmd
directory cleanup

--- 2019-11-16 manuscript/supplemental.Rmd
@@
 ```{r, include=F}
-map <- knitr::include_graphics(here::here("manuscript/figs/map.png")) # crop zoom, city font white, each site gets one point with box specifying elev range, n samples, dates, 
+map <- knitr::include_graphics(here::here("manuscript/figs/supp/map.png")) # crop zoom, city font white, each site gets one point with box specifying elev range, n samples, dates, 
 field_tbl <- read_csv(here::here("data/02_tidied/tidied_field.csv")) #
 field_photos <- knitr::include_graphics(here::here("manuscript/figs/supp/field_photos/field_collage.png"))
-micro_photos <- knitr::include_graphics(here::here("manuscript/figs/S4_site_micro_photos/micro.png"))
-cellct <- include_graphics(here("manuscript/figs/supp/cellct_scatterpie/cellct_scatterpie.png")) # make pies circles circular! they're shightly squished lengthwise
-tree_otu <- include_graphics(here("manuscript/supp/figs/otu_tree.png"))
+micro_photos1 <- knitr::include_graphics(here::here("manuscript/figs/supp/micro_photos/micro1.png"))
+micro_photos2 <- knitr::include_graphics(here::here("manuscript/figs/supp/micro_photos/micro2.png"))
+micro_photos3 <- knitr::include_graphics(here::here("manuscript/figs/supp/micro_photos/micro3.png"))
+
+cellct <- include_graphics(here("manuscript/figs/supp/cellct_scatterpie/cellct_scatterpie.png")) # try stacked barplots
+tree_otu <- include_graphics(here("manuscript/figs/supp/otu_tree.png"))
 tree_taxonomy <- include_graphics(here("manuscript/figs/supp/taxonomy_tree.png"))
@@
 field_photo_cap <- "Photos of snow algae field sites, arranged from high to low elevation from top to bottom. Alphanumeric code in upper left indicates sample ID."
-micro_photo_cap <- "Microphotos of snow algae samples included in sequencing run."
+micro_cap <- "Microphotos of snow algae samples included in sequencing run. All scale bars are approximately 20um."
 cellct_cap <- "Proportional abundance of snow algae cell morphologies, plotted against date and elevation."
@@
  3. Microphotos of snow algae samples
- 4. Plots of snow algae 18S community composition. A. All 18S, Silva assignments. B. Chlorophyta assignments with custom reference database.
- 5. Maximum likelihood tree of *rbc*L amplicon
- 6. Plot of base differences between Genbank 18S and *rbc*L segments
- 7. Phylogenetic diversity of snow algae.
+ 4. Piecharts of cell count relative abundance, plotted by sample elevation and date
+ 5. Maximum likelihood tree of *rbc*L amplicon, colored by taxonomic assignment
+ 6. ML tree of *rbc*L, colored by OTU
+ 7. Snow algae diversity over time and elevation.
+ 8. Photos of *Chloromonas krienitzii* above and below the surface, with relative abundance plots at surface and depth.
 
@@
 ```{r, fig.cap=micro_cap}
-#scale bar
-# # sizes the same (crop template)
-# arrange in grid
-micro_photos
+micro_photos1
 ```
 
-```{r s18_rel_abund, fig.cap=s18_cap}
-s18 # should this go in the main fig next to the rbcl assignments?
+```{r}
+micro_photos2
+```
+
+```{r}
+micro_photos3
+```
+
+```{r, (ref:cellct_cap)}
+cellct
 ```
@@
 ```{r tree, fig.cap=tree_cap}
-tree
+tree_otu
 ```
@@
 
-```{r, (ref:cellct_cap)}
-cellct
+```{r, fig.cap=depth_cap}
+depth
 ```
@@
 ```{r}
-field %>% 
+field_tbl %>% 
    filter(have_seq==T) %>% 

### 2020-01-10 code/misc/blast_treb_chloro_unknownE.R
directory cleanup

@@ -75,6 +75,39 @@ rbcl_otus %>%
   DNAStringSet() %>% 
   writeXStringSet(filepath = here("data/asv_seq_data/etc/unknown_E.fasta"))
 
+# ----------------------
+# check out Chloromonas D
+
+rbcl_otus %>% 
+  filter(otu_name == "Chloromonas D") %>% 
+  left_join(rbcl_key) %>% 
+  select(-otu_name) %>% 
+  deframe() %>% 
+  DNAStringSet() %>% 
+  writeXStringSet(filepath = here("data/asv_seq_data/etc/chloro_d.fasta"))
+
+# ----------------------
+# check out Chloromonas C
+
+rbcl_otus %>% 
+  filter(otu_name == "Chloromonas muramotoi C") %>% 
+  left_join(rbcl_key) %>% 
+  select(-otu_name) %>% 
+  deframe() %>% 
+  DNAStringSet() %>% 
+  writeXStringSet(filepath = here("data/asv_seq_data/etc/chloro_c.fasta"))
+
+# ----------------------
+# check out Chloromonas F
+
+rbcl_otus %>% 
+  filter(otu_name == "Chloromonas F") %>% 
+  left_join(rbcl_key) %>% 
+  select(-otu_name) %>% 
+  deframe() %>% 
+  DNAStringSet() %>% 
+  writeXStringSet(filepath = here("data/asv_seq_data/etc/chloro_f.fasta"))
+
 
 
 # check how many 18s asvs are algae

### 2020-01-11 code/rbcl_asv_clustering.Rmd
minor
assembled supplementary materials, moved scripts generating supp figs to manuscript/supp/code

@@
 
-# conflicted::conflict_scout()
+# conflicted::conflict_scout() # these packages cause rstudio to crash for me
 # conflicted::conflict_prefer("filter","dplyr")
@@
 
-```{r}
-# theme_set(dark_theme_classic())
-theme_set(dark_theme_light())
-```
+```{r read, include=F}
+rel_abund <- read_csv(here::here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
+taxonomy <- read_csv(here::here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
 
-```{r read, include=F}
-rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
-taxonomy <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
-
-aligns <- readDNAStringSet(here("output/alignments/rbcl_aligns_in_main.fasta"))
-tree <- read.tree(file=here("output/tree/rbcl/asv613.fasta.treefile")) # this is the tree with only the ASVs that appear in the filtered rel abund dataset (not including protocol stuff, etc)
+aligns <- readDNAStringSet(here::here("output/alignments/rbcl_aligns_in_main.fasta"))
+tree <- read.tree(file=here::here("output/tree/rbcl/asv613.fasta.treefile")) # this is the tree with only the ASVs that appear in the filtered rel abund dataset (not including protocol stuff, etc)
 
 # actual seq data
-key <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
+key <- read_csv(here::here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
 ```
@@
 ```{r output}
-write_csv(otus_asv_name, path = here("output/asv_otu_key.csv"))
+write_csv(otus_asv_name, path = here::here("output/asv_otu_key.csv"))
 
 # write the main ASV fig, w legend seperate
-ggsave(filename = here("manuscript/figs/main/taxonomy/asvs.pdf"), plot=asv_arranged, height = 6, width = 4)
-ggsave(filename = here("manuscript/figs/main/taxonomy/asv_legend.pdf"), plot=legend,  height = 4, width = 2.2)
+ggsave(filename = here::here("manuscript/figs/main/taxonomy/asvs.pdf"), plot=asv_arranged, height = 6, width = 4)
+ggsave(filename = here::here("manuscript/figs/main/taxonomy/asv_legend.pdf"), plot=legend,  height = 4, width = 2.2)
 
 # write black formatted versions for presentation
-ggsave(filename = here("presentation/omics/asvs.pdf"), plot=present_version, width = 7, height = 3.5)
-ggsave(filename = here("presentation/omics/asv_legend.pdf"), plot=legend,  height = 4, width = 2.2)
+ggsave(filename = here::here("presentation/omics/asvs.pdf"), plot=present_version, width = 7, height = 3.5)
+ggsave(filename = here::here("presentation/omics/asv_legend.pdf"), plot=legend,  height = 4, width = 2.2)
 
 # write out the trees
-ggsave(filename = here("manuscript/figs/otu_tree.png"), plot=otu_tree, width=7, height = 10)
-ggsave(filename = here("manuscript/figs/taxonomy_tree.png"), plot=taxonomy_tree, width=7, height = 10)
+ggsave(filename = here::here("manuscript/figs/otu_tree.png"), plot=otu_tree, width=7, height = 10)
+ggsave(filename = here::here("manuscript/figs/taxonomy_tree.png"), plot=taxonomy_tree, width=7, height = 10)
 
@@
   enframe() %>% 
-  write_csv(here("output/etc/otu_colors.csv"))
+  write_csv(here::here("output/etc/otu_colors.csv"))
 
 # sanguina fasta for breanna
-writeXStringSet(sang_seqs, here("output/taxa_filtered_fastas/sanguina_otu.fasta"))
-writeXStringSet(sang_taxa_seqs, here("output/taxa_filtered_fastas/sanguina_idtaxa.fasta"))
+writeXStringSet(sang_seqs, here::here("output/taxa_filtered_fastas/sanguina_otu.fasta"))
+writeXStringSet(sang_taxa_seqs, here::here("output/taxa_filtered_fastas/sanguina_idtaxa.fasta"))
 
 # chlainomonas fasta for jordan
-writeXStringSet(chlain_seqs, here("output/taxa_filtered_fastas/chlainomonas_otu.fasta"))
-writeXStringSet(chlain_taxa_seqs, here("output/taxa_filtered_fastas/chlainomonas_idtaxa.fasta"))
+writeXStringSet(chlain_seqs, here::here("output/taxa_filtered_fastas/chlainomonas_otu.fasta"))
+writeXStringSet(chlain_taxa_seqs, here::here("output/taxa_filtered_fastas/chlainomonas_idtaxa.fasta"))
 

### 2020-01-11 code/rbcl_community_compare.Rmd
assembled supplementary materials, moved scripts generating supp figs to manuscript/supp/code

@@
 
-ggsave(filename = here("manuscript/figs/main/uf_plot.png"), plot=uf_plot, width=5, height = 4)
+ggsave(filename = here("manuscript/figs/4/figure_4.tiff"), plot=uf_plot, width=85, height = 68, units="mm") # make 1 column wide, 5/4 aspect ratio
 ggsave(filename = here("manuscript/figs/diversity.png"), plot = diversity_plots, height = 10, width = 7)

### 2020-01-11 manuscript/supp/code/18s_asv_clust.Rmd
assembled supplementary materials, moved scripts generating supp figs to manuscript/supp/code

---
title: "18S algae ASV clustering"
output: pdf_document
---

```{r, include=F}
library(tidyverse)
library(DECIPHER)
library(ape)
library(ggtree)
```

```{r, include=F}
aligns <- readDNAStringSet(here::here("data/asv_seq_data/algae_18s_quality_aligns.fasta"))
# taxonomy w genbank (multiple copies of each designation, no editing)
taxonomy <- read_csv(here::here("output/algae_18s_assignments_snow_algae_genbank.csv"))
tree <- ape::read.tree(here::here("output/tree/algae_18s/algae_18s_seqs.fasta.treefile"))
rel_abund <- read_csv(here::here("data/03_filtered/18s_algae/18s_algae.filtered_cleaned_rel_abund.csv"))

# get total rel abund
total_rel_abund <- rel_abund %>% 
  group_by(asv_id) %>% 
  summarise(rel_abund = sum(rel_abund))

```

```{r, include=F}
BrowseSeqs(aligns, highlight = T)

dists <-  aligns %>% 
  DECIPHER::DistanceMatrix(type="matrix", correction = "JC")
```

# MDS
```{r mds, echo=F, message=F, warning=F}
set.seed(123)
mds_coords <- dists %>% 
  cmdscale() %>% # ok to use euclidian dist, no null values when comparing relatedness
  as_tibble() %>% 
  add_column(asv_id = colnames(dists), .before=1)

mds_coords_w_taxonomy <- mds_coords %>% 
  left_join(taxonomy, by="asv_id") %>% 
  left_join(rel_abund, by="asv_id")

colors <- c("Sanguina_nivaloides"="#ff534a",
            "Chloromonas_asteroidea"="#80c786", 
            "Chloromonas_nivalis"="#4aaf86", 
            "Chloromonas"="#4aaf53", 
            "Trebouxiophyceae"="#975c5c", 
            "Trebouxia"="#975c5c", 
            "Koliellaceae"="#975c5c",
            "Chlamydomonas_proboscigera"="grey60", 
            "Chlorogonium"="grey60", 
            "Chlamydomonadaceae"="grey60",
            "Chlamydomonadales"="grey60",
            "NA"="grey60")

mds <- ggplot(mds_coords_w_taxonomy, aes(x=V1, y=V2))+
  geom_point(aes(color=best_assignment, size=rel_abund), alpha=0.7, na.rm=TRUE) +
  scale_color_manual(values= colors) +
  guides(size="none",
         colour = guide_legend(override.aes = list(size=5, alpha=1))) +
  # include pts not assigned taxonomy
  geom_point(data = filter(mds_coords_w_taxonomy, is.na(best_assignment)), 
             color = "grey", alpha = 0.8) +
  labs(color="Taxonomic assignment")
mds
```

# t-SNE
```{r, echo=F,message=F, warning=F}
perplexity <- 20

set.seed(123)
tsne_full_output <- dists %>% 
  Rtsne::Rtsne(pca=F, perplexity = perplexity)

tsne_pts <- tsne_full_output$Y %>% 
  as_tibble()

tsne_pts_taxonomy <- tsne_pts %>% 
  add_column(asv_id = colnames(dists)) %>% # label each point by it's ASV ID
  left_join(taxonomy, by="asv_id") %>% 
  left_join(rel_abund, by="asv_id")

tsne <- ggplot(tsne_pts_taxonomy, aes(x=V1, y=V2)) +
  geom_point(data = tsne_pts_taxonomy, 
             aes(color = best_assignment, size=rel_abund), 
             alpha = 0.8, na.rm=TRUE) +
  # include points not assigned to genus or species
  geom_point(data = filter(tsne_pts_taxonomy, is.na(best_assignment)), 
             color = "grey", alpha = 0.8) +
  scale_color_manual(values= colors) +
  guides(size="none",
         colour = guide_legend(override.aes = list(size=5, alpha=1))) +
  labs(color="Taxonomic assignment")

tsne

```

\pagebreak

# Maximum likelihood tree (IQTree)

```{r, echo=F, message=F, warning=F}

# re-root tree
root <- taxonomy %>% 
  filter(class=="Trebouxiophyceae") %>% 
  head(1) %>% 
  pull(asv_id)

tree <- tree %>% root(outgroup = root, resolve.root=T)

ggtree_18s <- ggtree(tree)

# format bayesian/bootstrap confidence for each node
boots <- ggtree_18s$data %>% 
  filter(isTip == FALSE) %>% 
  separate(label, c("boot", "bayes"), sep = "/") %>% #split label into boots and bayes vals
  filter(boot > 80) %>%
  mutate(boot = round(as.numeric(boot), digits = 0),
         insignificant_edge = if_else(boot > 80, "black", "red"))# round boot to nearest integer

annotate <- ggtree_18s$data %>% 
  left_join(total_rel_abund, by=c("label"="asv_id")) %>% 
  left_join(taxonomy, by=c("label"="asv_id"))

colors <- c("Sanguina_nivaloides"="#ff534a",
            "Chloromonas_asteroidea"="#80c786", 
            "Chloromonas_nivalis"="#4aaf86", 
            "Chloromonas"="#4aaf53", 
            "Trebouxiophyceae"="#975c5c", 
            "Trebouxia"="#975c5c", 
            "Koliellaceae"="#975c5c",
            "Chlamydomonas_proboscigera"="grey60", 
            "Chlorogonium"="grey60", 
            "Chlamydomonadaceae"="grey60",
            "Chlamydomonadales"="grey60",
            "NA"="grey60")

tree_out <- ggtree(tree) +
  geom_text2(data=boots, aes(label=paste0(boot,";",bayes)),  vjust= 0.01, hjust = 1, size=2) +
  geom_tippoint(data = annotate, aes(color= best_assignment, size=rel_abund), alpha=1, na.rm=TRUE) + 
  theme(legend.position = c(0.89,0.7),
        legend.title = element_blank(),
        legend.key = element_blank()) +
  scale_color_manual(values= colors) +
  geom_tiplab(data = annotate, aes(label = best_assignment), size=2)+
  guides(size="none")

tree_out
```

### 2020-01-11 manuscript/supp/code/extraction_protocol.Rmd
assembled supplementary materials, moved scripts generating supp figs to manuscript/supp/code

--- 2020-01-10 manuscript/casey1.Rmd
@@
 ---
-title: "Alpine snow algae blooms vary with elevation"
-author:
-  - name: Casey B. Engstrom
-    institute: SFU
-  - name: Kurt M. Yakimovich
-    institute: SFU
-  - name: Lynne M. Quarmby
-    institute: SFU
-    email: quarmby@sfu.ca
-institute:
-  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
-output: 
-  bookdown::word_document2:
-    fig_caption: yes
-    pandoc_args:
-      - '--lua-filter=templates/scholarly-metadata.lua'
-      - '--lua-filter=templates/author-info-blocks.lua'
-    reference_docx: templates/word-template.docx
-  # pdf_document:
-  #   fig_caption: yes
-  #   number_sections: yes
-csl: the-isme-journal.csl
-bibliography: refs.bib
+title: "DNA extraction protocol"
+output: pdf_document
 
@@
 
-```{r include=FALSE}
+```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-### *Keywords*: snow algae, microbiome, amplicon, *rbcL*, 18S, alpine, metabarcoding
-
-### *Run title*: Alpine snow algae blooms
-
-<!-- search for ??? before send out -->
-
-##### pagebreak
-
-# Abstract {-}
-
-Snow algae blooms cover vast areas of summer snowfields worldwide, and can reduce albedo and increase snow melt.
-Despite their global prevalence, little is known about the algae species that comprise snow algae blooms.
-We used 18S and *rbcL* metabarcoding and light microscopy to characterise species composition of 33 snow algae samples from alpine and subalpine habitats in the Coast Range of British Columbia, Canada.
-We found blooms were dominated by the snow algae genera *Sanguina*, *Chloromonas*, and *Chlainomonas*.
-Above treeline blooms contained *Sanguina*, *Raphidonema*, and one OTU of *Chloromonas*, distinct from species assemblages below treeline predominated by *Chloromonas krienitzii*.
-Previously unrecognized as an important component of the snow algae microbiome, *Chlainomonas* were abundant in samples across all elevations.
-Our *rbcL* dataset detected diversity that was obscured by the low taxonomic resolution of traditional 18S metabarcoding, including OTUs that could represent novel species of snow algae.
-These three cross-referenced datasets show alpine snow algae blooms are more diverse than previously thought, and are highly variable from site to site along an elevational gradient.
-
-<!-- need to add the few meters part -->
-
-##### pagebreak
-
-# Introduction {-}
-<!-- global, vast blooms. albedo->snowmelt. what species? -->
-Each summer in polar and alpine snowfields worldwide, vast areas of snow surface are colored red by snow algae blooms.
-Such blooms are reported from every continent [@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019] and Arctic sea ice [@gradinger_snow_1996].
-They can be extensive: in Alaska, snow algae spectral signatures were detected in one third of a 1,900 km^2^ icefield [@ganey_role_2017].
-In recent years snow algae have received attention for their role in reducing snow surface albedo, which could substantially impact snow melt rates [@lutz_biogeography_2016;@ganey_role_2017].
-Thus, snow algae could impact the timing of spring melt and runoff, depleting summer water supplies held in mountain snowpacks and reducing glacier mass balance.
-Despite their global ubiquity, we are only beginning to identify the species that cause snow algae blooms.
-
-<!-- why microscopy not reliable -->
-Much of the previous work on snow algae relied on cell morphology for identification, but microscopy-based identification of snow algae can be unreliable.
-Different species of snow algae can look nearly identical: recent phylogenetic work found that four distinct field samples dominated by similar-looking cells previously referred to as *Chloromonas cf.
-nivalis* are actually genetically distinct clades [@matsuzaki_taxonomic_2019].
-Furthermore, the same species can look completely different in different environmental conditions or life stages.
-For example, cultured *Chloromonas krienitzii* cells are small green biflagellates, but field samples cells of this species are larger orange spheres with thick walls and short spines [@matsuzaki_revisiting_2015].
-In the freshwater green algae *Haematococcus pluvialis* red secondary pigments are produced in response to high light [@shah_astaxanthin-producing_2016], but genetic comparison of red and green snow found they contained different species [@terashima_microbial_2017;@lutz_integrated_2015].
-
-<!-- species from snow algae blooms, what is currently known -->
-A diversity of algae have been reported from snow algae blooms, with many new species only having been described recently.
-Genera from the class Chlorophyceae are predominant in many blooms, including *Sanguina*, *Chloromonas*, and *Chlainomonas*.
-The taxonomy of *Sanguina* was recently established, and currently contains just two species; however, many sequences from red snow samples form an unnamed sister clade to *Sanguina* [@prochazkova_sanguina_2019].
-The same study found *Sanguina nivaloides* predominated in red snowfields worldwide, while *S.
-aurantia* occurred in just three orange snow samples in Svalbard.
-The genus *Chloromonas* is speciose, containing twelve species that were isolated from snow [@matsuzaki_taxonomic_2019].
-Different species of *Chloromonas* can form green, orange, or brownish-red blooms on the snow surface, and this genus has been reported from both polar and mid-latitude snowfields [@remias_ecology_2018;@remias_insights_2013;@prochazkova_chloromonas_2018].
-Blooms of *Chlainomonas* have been found in red snow overlying alpine lakes in central Europe, western USA, and New Zealand [@novis_two_2008;@remias_ecophysiology_2016;@prochazkova_ecophysiological_2018].
-This genus, which currently contains two named species, is closely related to *Chloromonas*.
-Non-Chlorophyceaen snow algae include Chrysophyceae found in Antarctica, the Alps, and Svalbard [@soto_microbial_2020;@remias_two_2019], and Trebouxiophyceae found in green snow in Greenland [@lutz_integrated_2015].
-
-<!-- metabarcoding -->
-In contrast to the high diversity of snow algae species observed in culture and Sanger sequencing, metabarcoding suggests that snow algae blooms are highly similar across continents, and are dominated by few taxa.
-33 red snow communities across the Arctic were remarkably similar to each other: all were dominated by two operational taxonomic units (OTUs) of "uncultured Chlamydomonadaceae 1 and 2" with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale* [@lutz_biogeography_2016].
-Similarly, another study using ITS2 metabarcoding found 24 red snow sites across the Arctic and Antarctic contained similar assemblages, dominated by "uncultured Chlamydomonadaceae A and B" with secondary abundance of *Raphidonema* and *Chloromonadinia* [@segawa_bipolar_2018].
-Other 18S metabarcoding studies of snow algae were limited to genus or family level taxonomy [@hamilton_primary_2017]---because 18S is highly conserved, this marker generally cannot distinguish between closely related taxa.
-
-<!-- our goal, what we did, what we found -->
-Virtually nothing is known about the regional variation in species composition of snow algae blooms.
-Our goal was to identify the algae in blooms on different mountains and elevations in the Coast Range of British Columbia, Canada.
-We used 18S and *rbcL* metabarcoding to assess species composition---the latter targeting a hypervariable region of *rbcL* (coding for large subunit of rubisco), thus targeting only photosynthetic species with high taxonomic resolution.
-Additionally, we used light microscopy to describe the relative abundance of different morphologies in each field sample.
-By using these three cross-referenced metrics of relative abundance we were able to describe the algal species assemblage with high accuracy.
-Using *rbcL* as a marker revealed much greater diversity than would have been found using 18S alone.
-We did not find any differences in bloom composition between mountains, but there were distinct algae species assemblages at high versus low elevations.
-<!-- stronger hook, predictions??? cliff hangers -->
-
-# Materials and Methods {-}
-
-<!-- refer to each supp at least once in txt -->
-<!-- add cities, states -->
-
-## Field sampling and microscopy {-}
-
-We collected snow algae from sites throughout the Coast Range near Vancouver, British Columbia, Canada over the summer of 2018 (Supplementary Figure S1).
-To capture the extent of snow algae diversity we sampled from different elevations, dates, mountains, and micro-habitats within the snow.
-In total, we collected 309 colored snow samples from a range of 880 - 2150 m elevation, including samples from 13 different mountains collected on 33 different dates (Supplementary Table S1).
-Our samples trended upwards in elevation over the course of the summer as the snow melted earliest at the lowest elevations (Supplementary Figure S2).
-We first detected snow algae above 1500 m on June 20.
-We scooped samples from visibly colored snow into sterile 50 mL centrifuge tubes.
-To prevent melting en route to the lab we stored tubes in a bag of snow.
-We melted each sample at room temperature on the bench and removed a 1 mL aliquot for light microscopy.
-Immediately after, samples were stored at -20 &deg;C for up to eight months until DNA extraction.
-
-We used light microscopy to characterize the cell morphologies in each sample within 24 hours following collection.
-We fixed samples in 2% gluteraldehyde, and under 400x magnification observed the 100 cells closest to the center of the slide.
-We classified cell morphology based on similarity to published photographs of *Sanguina nivaloides* [@prochazkova_sanguina_2019], *Chloromonas cf.
-nivalis* [@prochazkova_chloromonas_2018], *Chloromonas cf.
-brevispina* [@matsuzaki_revisiting_2015], *Chloromonas krienitzii* [@matsuzaki_revisiting_2015], *Chlainomonas rubra* [@novis_two_2008], and for cells that did not fall into one of there categories as either "green cell" or "other".
-
-<!-- how do select field and micro photos fit into this? -->
-
-## DNA extraction and amplicon library preparation {-}
-
-<!-- This part is non reproducible, selecting based on date, etc -->
-
-We selected 33 out of our 310 samples for *rbcL* and 18S Illumina amplicon sequencing.
-We chose these samples to represent the variation in date, elevation, geographic location, snow color, micro-habitat, and cell morphology.
-We freeze-dried these samples for up to 48 hours until until samples appeared completely desiccated.
-We mini-pestled between 5 to 20 mg of sample at room temperature to physically rupture cell walls before incubation in CTAB lysis buffer [@noauthor_ctab_2009].
-We extracted DNA using chloroform:isoamyl alcohol [@cubero_dna_1999], and spin columns (Qiagen, Hilden) (detailed protocol in Supplementary File S1).
-As a negative control we processed a sterile distilled water sample alongside each batch.
-
-<!-- arent we following some extraction protocol here? cant we cite, and then have a supplementary protocol with our variation??? -->
-
-<!-- To extract DNA from the crushed cells we added 800 mL  , 1% B-mercaptoethanol, 5 $\mu$L each of proteinase K and RNAase A, and incubated these at 65 &deg;C for 30 minutes.
+To extract DNA from the crushed cells we added 800 mL CTAB (Cold Springs Harbour Protocols), 1% B-mercaptoethanol, 5 $\mu$L each of proteinase K and RNAase A, and incubated these at 65 &deg;C for 30 minutes.
 We spun samples at 10,000 g for 3 minutes to pellet cell debris, then added 700 $\mu$L of supernatant to an equal volume of 24:1 chloroform:isoamyl alcohol.
@@
 We washed columns twice with 70% ethanol, and finally dissolved our DNA by spinning with 50 $\mu$L of sterile TE buffer.
--->
-
-We designed custom primers to target a hypervariable region of snow algae *rbcL* (coding for the large subunit of rubisco).
-*rbcL* is more differentiated between snow algae than 18S (Supplementary Figure S3), and contains similar number of reference sequences on GenBank.
-Our *rbcL* primers target a 400 bp section of snow algae *rbcL* based off the consensus of 20 GenBank snow algae sequences (Supplementary Table S2).
-We designed our primers with the Eurofins primer design tool (https://www.eurofinsgenomics.eu/en/ecom/tools/pcr-primer-design/).
-*Sanguina* sequences were not included because they were not available at the time.
-To target 18S, we used the universal primers Euk1181 and Euk1624 targeting the V7-V8 hypervariable regions [@wang_optimal_2014].
-Primer sequences are available in Supplementary Table S3.
-
-We constructed our 18S and *rbcL* amplicon libraries using a standard two-step PCR approach [@meyer_illumina_2010].
-For detailed PCR conditions, see Supplementary File S2.
-After each PCR, we purified using Agencourt AMPure XP kit (Beckman Coulter, Brea, CA).
-We standardized sample DNA concentration using Qubit fluorometric quantification (Thermo Fisher, Waltham, MA).
-The pooled library was loaded and run on a MiSeq V3 kit (Illumina, San Diego, CA).
-
-## Bioinformatic processing {-}
-
-We demultiplexed reads using CUTADAPT [@martin_cutadapt_2011], and filtered and trimmed, removed errors, dereplicated, merged paired-end reads, and removed chimeras following the default pipeline of DADA2 [@callahan_dada2_2016].
-We assigned taxonomy to amplicon sequence variants (ASVs) using IDTaxa [@murali_idtaxa_2018], only assigning names to ASVs that were assigned with high confidence (threshold set to 50).
-We assigned taxonomy using custom snow algae reference databases for both 18S and *rbcL*, in addition to re-assigning 18S with SILVA to classify non-snow algae [@quast_silva_2013].
-Snow algae reference databases and the scripts to generate them are available on our github repository (see below).
-
-<!-- should cite a github script here??? singletons, failed samples removed, remove sequences based on negative controls, adjust for 0's, sequencing depth, -->
-
-<!-- make sure this is on GH: -->
-
-<!-- We made our *rbcL* database by downloading 30,865 algae rbcL sequences from GenBank with the query “rbcl[gene] AND (algae OR chlorophyta OR trebouxiophyceae) NOT (18s OR ribosomal OR psaB OR atpB) AND 300:2000[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion”.
-We removed ambiguous snow algae annotations: because entries labelled as *Chloromonas cf.
-brevispina* and *Chloromonas cf.
-nivalis* are polyphyletic [@matsuzaki_taxonomic_2019] we re-labeled these as “unassigned *Chloromonas*”.
-and then re-ran ASVs assigned to Chlorophyta on our custom 18S snow algae database based on the following GenBank query: "(18S OR ribosomal)[gene] AND (chloromonas OR chlainomonas OR sanguina OR raphidonema OR KMY-2018) NOT (rbcL OR psaB OR atpB) AND 300:3500[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion".
--->
-
-To generate OTUs we clustered ASVs using t-SNE [@maaten_visualizing_2008].
-We compared t-SNE results with the perplexity parameter ranging from 1 to 100, and found comparable results from 10 to 50.
-We categorized OTUs by applying DBSCAN [@hahsler_dbscan_2019] to our t-SNE.
-We compared *rbcL* sample similarity using UniFrac with default settings [@lozupone_unifrac_2005].
-To confirm that our clustering was not an artifact of t-SNE we created maximum likelihood phylogenetic trees of ASVs using IQTree [@nguyen_iq-tree:_2015] (Supplementary Figures S4-S5).
-
-All raw sequence data are available under European Nucleotide Archive accession PRJEB34539. All scripts used in this study are available at https://github.com/cengstro/bc_snow_algae_amplicon.
-
-# Results {-}
-
-<!-- add read depth, etc? -->
-<!-- make fig 5, ???, and make 18s otus -->
-
-We observed morphologically distinct snow algae communities at different elevations (Figure 1).
-Red snow was prevalent in areas of high solar exposure above treeline, predominantly containing cells similar in appearance to *Sanguina nivaloides* [@prochazkova_sanguina_2019].
-We occasionally observed green snow patches in alpine meadows receiving full sun, but more frequently observed green snow in well-shaded, forested sites.
-We found cells resembling *Chlainomonas rubra* [@novis_two_2008] at all elevations.
-Forested sites at low elevations often contained green or orange patches of snow, predominated by green or orange ovate cells with short spines resembling *Chloromonas cf.
-brevispina* [@matsuzaki_revisiting_2015].
-
-Both 18S and *rbcL* amplicon libraries were dominated by ASVs that were assigned to Chlorophyta.
-Our 18S library contained 67 ASVs that were assigned to Chlorophyta and 7 ASVs to Ochrophyta, while our *rbcL* library contained 603 ASVs assigned to Chlorophyta and 41 to Trebouxiophyceae.
-Negative controls did not contain Qubit-detectable amounts of DNA.
-Our *rbcL* library contained higher ASV and OTU diversity than 18S: t-SNE clustering generated 9 *rbcL* OTUs, compared with just 4 18S OTUs (Supplementary Figure S6).
-The most frequent *rbcL* taxonomic assignments at genus level were *Chloromonas*, *Chlainomonas*, and *Sanguina*.
-Although most ASVs were not assigned to genus level, most ASVs clustered near one of these genera in NMDS ordination (Figure 2).
-Several *rbcL* OTUs were dominated by ASVs that were not assigned to genus level, including three OTUs closely related to *Chloromonas*, and one Chlamydomonadaceae OTU that was only assigned to the family level.
-
-The composition of 18S and *rbcL* libraries varied with elevation (Figure 3; Figure 4).
-*Sanguina* was absent below 1500 m in both 18S and *rbcL*.
-High elevation samples contained an unannotated OTU of *Chloromonas* that was absent from low elevation sites (Figure 3, "Chloromonas C").
-Additionally, the three highest elevation sites uniquely contained ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1)---interestingly, all of these samples were taken from snow overlying glacier.
-The two high alpine green snow samples were dominated by two different OTUs of *Chloromonas*, which were also both found in sites below treeline.
-*Chlainomonas* was prevalent at all elevations, and *Chloromonas krienitzii* was predominant in our low elevation sites.
-
-We observed snow algae blooms dominated by *Chloromonas krienitzii* transition from green to orange from May to June (Figure 5).
-At two sites we observed snow melt channels or runnels containing high concentrations of snow algae.
-In May and early June these sites contained green snow hidden 2 to 5 cm below the white snow surface, but upon subsequent visits the same location contained orange snow at the surface, whose morphologies resembled published images of the *Chloromonas krienitzii* lineage [@matsuzaki_revisiting_2015].
-We compared 18S and *rbcL* metabarcode profiles of green subsurface and orange surface snow, and found both contained high relative abundance of *Chloromonas krienitzii*.
-Although we did not detect *Chlainomonas* in cell counts, the surface samples contained nearly double the relative abundance of *Chlainomonas* from the subsurface samples.
-
-<!-- include faiths diversity in supps??? thinking no  -->
-
-# Discussion {-}
-
-Here we present multiple data sets demonstrating elevational patterns in alpine snow algae blooms.
-*Sanguina* was dominant in red snow above treeline, while green and orange blooms of *Chloromonas krienitzii* were dominant in mid-elevation snow runnels.
-We found unexpected diversity within *Chloromonas* and *Chlainomonas* *rbcL* that we did not detect using 18S, including many potentially novel OTUs.
-In contrast to previous studies [@novis_two_2008;@prochazkova_ecophysiological_2018], our work suggests that *Chlainomonas* is widespread and abundant in alpine snow algae blooms.
-
-Overall, our results show bloom composition can be highly variable on the same mountain, even in nearby patches.
-This is in contrast to previous papers that found snow algae assemblages were homogeneous across geographically distant sites [@lutz_biogeography_2016;@segawa_bipolar_2018].
-This could be due to different sampling schemes: the aforementioned studies exclusively sampled high-latitude red snow, while our study included mid-latitude red, orange, and green snow from above and below treeline.
-Another recent study using metabarcoding found different snow algae communities in red and green snow algae blooms in Antarctica, including yellow blooms dominated by Chrysophyceae [@soto_microbial_2020].
-Much of the diversity in our samples within *Chlainomonas* and *Chloromonas* was only detected with *rbcL* and not 18S, so the diversity within these genera may have been overlooked in previous metabarcoding studies.
-
-Our *rbcL* data suggest *Chlainomonas* may be more common than previously thought.
-We could not differentiate between *Chlainomonas* and *Chloromonas* in our 18S data set, so this genus may have been overlooked in previous studies.
-*Chlainomonas* was previously thought to be restricted to waterlogged snow overlying mountain lakes [@novis_two_2008;@prochazkova_ecophysiological_2018], but we did not find any evidence that *Chlainomonas* was restricted in its habitat distribution.
-To the contrary, only one *Chlainomonas* dominant sample was located in waterlogged snow near the edge of a pool (sample S9); the other *Chlainomonas* dominant sites were not notably wetter than the surrounding snow, nor overlying frozen bodies of water.
-Interestingly, we found *Chlainomonas* was in higher relative abundance in our *rbcL* dataset than in cell counts (Figure 3)---possibly due to multiple chloroplasts per cell, which have been documented in this genus [@prochazkova_ecophysiological_2018].
-
-Our findings highlight the remaining unexplored diversity in the snow algae microbiome.
-Many *rbcL* OTUs did not closely match any GenBank nucleotide archive sequences (Figure 2).
-"Chlamydomonadaceae E" highest BLAST percent identity was only 87%, and the top 10 BLAST matches contained 6 different algae genera---including two *Chloromonas* snow algae from Japan.
-This OTU did not correlate with any categories in our 18S or cell count data.
-One possible explanation is that it was lumped together with *Chloromonas* in 18S, and our crude quantification and categorization of cell counts did not detect this taxa.
-Other OTUs were also poorly matched to GenBank,  which could represent novel species of *Chloromonas* (Figure 2).
-The lack of clustering we observed in the remaining *Chloromonas* could be due to overlap in the levels of interspecific and intraspecific variation.
-*rbcL* diversity is likely higher within *Chloromonas* than in other algae genera: unlike *Sanguina* [@prochazkova_sanguina_2019], most species of *Chloromonas* lack a pyrenoid [@nozaki_differences_2002], which houses high concentrations of cross-linked rubisco.
-Likely related to its lack of pyrenoid, *Chloromonas* has many non-synonymous mutations in the region of *rbcL* the codes for binding rubisco together [@nozaki_differences_2002].
-While *rbcL* is problematic in *Chloromonas* phylogeny [@nozaki_differences_2002], it nonetheless is highly differentiated and therefore an effective barcode for microalgae [@hall_assessment_2010].
-
-<!-- *Sanguina* was found in the same samples, but relative abundance was generally higher in 18S and cell counts (Figure 3).
-Our *rbcL* primers were designed based on *Chloromonas* and *Chlainomonas* snow algae sequences (*Sanguina* sequence data was not available at the time), so primer bias could have resulted in over-representation of *Chloromonas* relative to *Sanguina*.
--->
-
-Variation in bloom composition could be due to a wide range of habitat features.
-*Sanguina* and *Chloromonas C* were limited to sites above 1500 m in full sunlight, but many low elevation sites also received full sunlight and did not contain these OTUs (Figure 3).
-Light intensity, snow moisture, or snow chemistry could all plausibly influence community composition.
-Intriguingly, we only observed *Raphidonema* at high-elevation glacier sites.
-In Svalbard, *Raphidonema nivale* abundance increased on glacier surface snow following wind storms, and the authors suggest that this is a soil algae that grows sub-optimally on the snow surface following wind deposition [@stibal_growth_2005].
-Given the aerial dispersal capabilities of microalgae [@tesson_airborne_2016] and genetic overlap between distant snow algae populations [@segawa_bipolar_2018] it seems unlikely that geographic distance is a barrier to snow algae distribution.
-Underlying topography could potentially influence species distribution: two sites dominated by *Chloromonas krienitzii* were in runnels overlying ephemeral streams, which could influence nutrient availability where surface debris is deposited in the runnel.
-
-Over multiple visits we observed motile green *Chloromonas krienitzii* cells below the snow surface develop into an orange bloom at the surface (Figure 5).
-Previous work has shown that this species can appear as distinct green and orange morphologies [@matsuzaki_revisiting_2015], but this paper is the first to document a bloom dominated by this species change color in the field.
-The transition over weeks suggests that this process is mediated by seasonal changes.
-Secondary pigments likely protect snow algae from the damaging effects of intense solar irradiation at the snow surface [@bidigare_evidence_1993], which could be why we only observed green cells below the snow surface.
-
-In conclusion, our study reveals substantial diversity within and between snow algae blooms.
-We found greater diversity using *rbcL* metabarcoding than 18S, which suggests snow algae diversity may have been overlooked in studies that used only 18S.
-Species were differentially distributed along an elevational gradient, with distinct blooms dominated by *Chloromonas*, *Sanguina*, or *Chlainomonas*.
-Future studies will examine other taxa present in snow algae blooms, and their interactions in these globally important microbiomes.
-
-# Acknowledgments {-}
-
-We thank Leah Tooman (Simon Fraser University) for assistance with sequencing, and Chris Rushton (Simon Fraser University) for assistance with bioinformatics.
-This project was funded with a Sector Innovation Grant from Genome BC (SIP016), and a NSERC Individual Discovery Grant, both awarded to LQ.
-
-# Author Contributions {-}
-
-CE, LQ, and KY designed this study.
-Samples collected by CE with assistance from KY and LQ.
-KY and CE prepped samples for sequencing.
-All analyses were completed by CE and KY.
-All authors discussed the results and contributed to the final manuscript.
-CE wrote the manuscript with major input from LQ and KY.
-
-# Additional information {-}
-
-The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
-
-##### pagebreak
-
-# References {-}
-<div id="refs"></div>
-
-##### pagebreak
-
-# Figure legends {-}
-
-<!-- formatting indentation??? -->
-
-**Figure 1.** Representative photographs of snow algae.
-A.
-Red snow bloom above treeline at sample site G1.1.
-B.
-Microphotograph of red snow containing *Sanguina nivaloides* (Sn) and *Chlainomonas rubra* (Cr) cell morphologies.
-All scale bars 30 $\mu$m, all microphotos taken using 63x objectives with DIC.
-C.
-Snow runnels in a forest clearing containing snow algae (inset).
-D.
-Microphotograph of orange snow from the surface of runnel containing *Chloromonas krienitzii* (Ck) and *Chloromonas cf.
-nivalis* (Cn).
-E.
-Bronze coloured snow algae blooms below conifer canopy.
-F.
-Microphotograph of bronze snow containing Chloromonas cf.
-brevispina* (Cb) and *Chloromonas cf.
-nivalis* (Cn).
-
-\setlength{\leftskip}{0pt}
-**Figure 2.** Ordination plots of *rbcL* ASV similarity.
-A.
-Multidimensional scaling (MDS) plot.
-Taxonomy (assigned by IDTaxa) indicated by color, and size of point is proportional to cumulative relative abundance.
-Dotted ellipses show OTU clusters.
-Stress = 0.13.
-B.
-t-Distributed Stochastic Neighbor Embedding (t-SNE) dimensionality reduction of snow algae *rbcL* ASVs.
-Dotted lines indicate OTUs.
-Perplexity = 30.
-
-**Figure 3.** Stacked bar plots of snow algae relative abundance.
-Each sample is arranged on the y axis in order of elevation, with each compositional bar plot representing the (left to right) cell morphology, 18S, and *rbcL* snow algae assemblages of the same sample.
-Colors below each bar plot indicate taxonomic categories.
-Morphospecies were identified by light microscopy, using similarity to published photographs as a guide.
-18S OTUs are reference-based, while *rbcL* OTUs were clustered *de novo*, and named using GenBank reference data.
-
-**Figure 4.** Non-metric multidimensional scaling (NMDS) showing *rbcL* UniFrac distances between samples. Each point represents a sample, which is labelled by sample ID and elevation (color).
-
-**Figure 5.** ??? Need to make the figure first

### 2020-01-11 manuscript/supp/code/field_site_table.Rmd
assembled supplementary materials, moved scripts generating supp figs to manuscript/supp/code

---
title: "Snow algae sample sites"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=F}
library(tidyverse)
library(kableExtra)
```

```{r, include=F}
field <- read_csv(here::here("data/02_tidied/tidied_field.csv"))

table_data <- field %>% 
  filter(have_seq==T) %>% 
  select(alias, date, lat, lon, elev_m, color_snow, habitat) %>%
  arrange(date) %>% 
  dplyr::rename("Sample ID" = alias, "Date" = date, "Latitude"=lat, "Longitude"=lon, "Elevation (m)"=elev_m, "Color snow"=color_snow, "Habitat"=habitat)
```

```{r, echo=F}
table_data %>% 
  kable(booktabs=T, linesep="") %>% 
  kable_styling()
```

### 2020-01-11 manuscript/supp/code/map/drafts/2018_sample_map.Rmd
assembled supplementary materials, moved scripts generating supp figs to manuscript/supp/code

--- 2019-11-29 code/2018_sample_map.Rmd
@@
 ---
-
-<style>
-table {
-  background-color: white !important;
-  color: black !important;
-}
-</style>
-<!-- format the table black and white -->
 
@@
 
+* loads rda from the saved google map so I dont have to re-load it again and again... this could be tidier
+* Specify here::here to avoid conflict with lubridate::here
 ```{r include=F}
@@
 cities <- read_csv(here::here("output/map/city_coords.csv"))
-
 load(here::here("output/map/raw_google_map.rda")) # loads .rda called "algae_map"... not sure how avoid using.rda format, maybe png?
 ```
+
 ```{r}
@@
 ```{r}
-field_summary <-
-  field %>%
+field_summary <- field %>%
   group_by(mountain) %>% 
@@
   select(-date)
-
-# paste(field$date %>% month(label=TRUE), field$date %>%day())
 ```
@@
 
-ggsave(filename = here::here("manuscript/figs/map.png"), plot=full_map)
+ggsave(filename = here::here("manuscript/figs/map.tiff"), plot=full_map)
 
 # for the month by month maps
-save_map <- function(my_month, num){
-  map <- plot_map(my_month)
-
-  ggsave(filename = here::here("figs","map",paste0("map_",num,"_", my_month,".png")),
-       plot = map, height = 15, width = 20, units = "cm")
-}
-
-save_map("May",1)
-save_map("June",2)
-save_map("July",3)
-save_map("August",4)
-save_map("September",5)
+# save_map <- function(my_month, num){
+#   map <- plot_map(my_month)
+#   
+#   ggsave(filename = here::here("figs","map",paste0("map_",num,"_", my_month,".png")),
+#        plot = map, height = 15, width = 20, units = "cm")
+# }
+# 
+# save_map("May",1)
+# save_map("June",2)
+# save_map("July",3)
+# save_map("August",4)
+# save_map("September",5)
 

### 2020-01-11 manuscript/supp/code/map/drafts/map.Rmd
assembled supplementary materials, moved scripts generating supp figs to manuscript/supp/code

--- 2019-10-24 code/map.Rmd
@@
 title: "Snow algae 2018 map"
-
 ---
@@
 library(tidyverse)
+library(here)
 library(lubridate)
@@
 ```{r data_in}
-field <- read_csv(here("data","clean","field_clean.csv"))
+field <- read_csv(here::here("data/02_tidied/tidied_field.csv"))
 ```
@@
 register_google(key = key)
-theme_set(dark_theme_classic())
 ```
@@
 
+ggsave(filename = here::here("figs","map",paste0("full_map.png")),
+       plot = full_map, height = 15, width = 20, units = "cm")
+
 ```

### 2020-01-11 manuscript/supp/code/primer_table.Rmd
assembled supplementary materials, moved scripts generating supp figs to manuscript/supp/code

---
title: "Primers"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=F}
library(tidyverse)
library(kableExtra)
```

```{r, include=F}
# seq 5->3'
tbl <- tibble(name=c("rbcL369F","rbcL870R", "Euk1181 (Wang et al. 2014)","Euk1624  (Wang et al. 2014)"),
              seq=c("GAACGTGACAAATTAAACAAA","ACCWGAYADACGWAGAGCTT","TTAATTTGACTCAACRCGGG","CGGGCGGTGTGTACAAAGG"))
```

```{r, echo=F}
tbl %>% 
  rename(Name=name, `Sequence (5'-3')`=seq) %>% 
  kable(booktabs=T, linesep="") %>% 
  kable_styling()
```

References:
Wang Y, Tian RM, Gao ZM, Bougouffa S, Qian P-Y. Optimal Eukaryotic 18S and Universal 16S/18S Ribosomal RNA Primers and Their Application in a Study of Symbiosis. PLOS ONE 2014; 9: e90053.

### 2020-01-11 manuscript/supp/code/map/map.R

@@ -1,0 +1,60 @@
+# draw sample map
+# 1-10
+
+library(tidyverse)
+library(lubridate)
+library(ggmap)
+library(ggrepel)
+# library(kableExtra)
+# library(magick)
+# library(formattable)
+# library(ggdark)
+
+# it wouldve been easier to just download the map as png, and the city data instead of using the clumsy API
+# next time I will do this, but for now using the Rda map data
+
+field <- read_csv(here::here("data/02_tidied/tidied_field.csv"))
+cities <- read_csv(here::here("output/map/city_coords.csv"))
+load(here::here("output/map/raw_google_map.rda"))
+
+# only plot vancouver
+cities <- cities %>% 
+  filter(name == "Vancouver")
+
+# select field data to plot (only mtns, not individual sites)
+field_mtns <- field %>% 
+  mutate(mtn_letter = alias %>% str_sub(1,1))
+
+field_sites <- field_mtns %>% 
+  distinct(mtn_letter) 
+
+field_site_coords <- field_mtns %>% 
+  group_by(mtn_letter) %>% 
+  drop_na(lat, lon) %>% 
+  summarise(n=n(), 
+            mean_elev = mean(elev_m), 
+            first_visit=min(date) %>% format("%b %d"), 
+            last_visit= max(date) %>% format("%b %d"), 
+            n_visits= length(unique(date)),
+            lat=mean(lat),
+            lon=mean(lon),
+            date=min(date)) %>% 
+  arrange(date) %>% 
+  select(-date)
+
+my_map <- ggmap(algae_map, extent = "device") +
+  geom_point(data = field_site_coords, aes(x=lon, y=lat),
+             size = 1, alpha=1, 
+             color = "red") +
+  # geom_text(data=field_site_coords, aes(label = mtn_letter)) # dosen't map site A, too far south for map
+  geom_label_repel(data=field_site_coords, aes(label = mtn_letter),
+    size = 4, 
+    box.padding = 0.2, point.padding = 0.3,
+    segment.color = 'grey50') +
+  geom_point(data = cities, shape = 15, color="white") +
+  geom_text(data = cities, aes(label = name), hjust = 1.1, vjust = 2, size = color="white")
+
+my_map
+
+# save
+ggsave(filename = here::here("manuscript/supp/map_1-10.tiff"), plot=my_map, width = 200, height = 200, units="mm",dpi=300)

### 2020-01-11 manuscript/supp/code/pcr_conds.Rmd
assembled supplementary materials, moved scripts generating supp figs to manuscript/supp/code

---
title: "PCR Conditions"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We constructed our amplicon library using a two-step PCR. In the first PCR we amplified template DNA using our primers attached to a universal adapter, and in the second PCR we re-amplified that product to attach a 6 bp index to the universal adapter at the 3’ end. The first PCR total volume was 25 $\mu$L, consisting of 1 $\mu$L template, 12.5 $\mu$L Q5 high-fidelity 2X Master Mix (New England BioLabs), 1.25 $\mu$L each of forward and reverse primer, and 9 $\mu$L of ddH2O. The second PCR was the same except we reduced our reaction volume to 20 $\mu$L by using only 5 $\mu$L of ddH2O. The cycling conditions were the same for both primer pairs for the first PCR, with an initial denaturation at 98 $^\circ$C for 30 s, followed by 30 cycles of 98 $^\circ$C for 5 s, 58 $^\circ$C for 10 s, and 72 $^\circ$C for 25 s, with a final extension at 72 $^\circ$C for 2 minutes. For the second indexing PCR we started with an initial denaturation at 98 $^\circ$C for 30 s, then 10 cycles of 98 $^\circ$C for 10 s, 65 $^\circ$C for 30 s, and 72 $^\circ$C for 30 s with a final denaturation of 72 $^\circ$C for 5 min.

### 2020-01-11 protocol_s1.Rmd
assembled supplementary materials, moved scripts generating supp figs to manuscript/supp/code

--- 2019-10-31 manuscript/Casey1.Rmd
@@
 ---
-title: "Distinct snow algae communities along an elevational gradient"
-author:
-  - name: Casey B. Engstrom
-    institute: SFU
-  - name: Kurt M. Yakimovich
-    institute: SFU
-  - name: Lynne M. Quarmby
-    institute: SFU
-    email: quarmby@sfu.ca
-institute:
-  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
-date: "`r format(Sys.time())`"
-output: 
-  word_document:
-    # toc: yes
-    pandoc_args:
-      - '--lua-filter=templates/scholarly-metadata.lua'
-      - '--lua-filter=templates/author-info-blocks.lua'
-    reference_docx: /home/cengstro/projects/sa_rbcl/Casey1/tmp/word-template.docx
-  pdf_document: 
-    number_sections: yes
-    # toc: yes
-    # toc_depth: 4
-    pandoc_args:
-      - '--lua-filter=templates/scholarly-metadata.lua'
-      - '--lua-filter=templates/author-info-blocks.lua'
-bibliography: /home/cengstro/ownCloud/mendeley_library/bibtex_by_grp/casey1.bib
-abstract: |
-  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate.
-Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch on the same mountain, and varies with elevation and date.
-We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada.
-We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m).
-Samples from higher elevation had higher phylogenetic diversity, and included several potentially novel clades of algae.
-Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface.
-Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome.
+title: "PCR conditions"
+author: "Engstrom, Yakimovich, and Quarmby 2020"
+output: pdf_document
+---
 
----
 ```{r setup, include=FALSE}
@@
 
-<br>
-<br>
-<br>
-
-### *Keywords*: snow algae rbcL 18S amplicon microbiome alpine community
-
-### *Run title*: Elevational patterns in snow algae communities
-
-### *To be submitted to*: 1) Nat Comm, 2) Frontiers in Micb 3) Env Micb 4) Microbial Ecology
-
-##### pagebreak
-
-# Introduction {-}
-
-Snow algae are cold-adapted microalgae that bloom in melting summer snow throughout alpine and polar regions [@Leya2013].
-Red snow algae blooms can cover vast areas of snowfield, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017].
-Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], which can threaten water supplies [@Staudinger2014] and worsten drought [@Barnhart2016].
-Thus, snow algae could play an important role in snow melt regulation, but little is known about how these blooms form.
-We don't know the spatial extent of snow algae blooms worldwide, what conditions trigger snow algae bloom formation, or how cells arrive on the surface of the snow to "seed" the bloom.
-Critically, we lack basic understanding of the species composition and diversity in snow algae blooms, which is fundamental to answering these long-standing mysteries about snow algae blooms.
-
-Historically snow algae were classified by cell morphology, however, morphology can be deceptive when identifying microalgae.
-Until recently, all red cells from red snow were referred to as *Chlamydomonas cf.
-nivalis* [@Remias2013;@Segawa2018], desipte the fact that this morphology-based species was thought to contain multiple distinct species [@Kol1961].
-Recent genetic work designated found that some of these cells were not closely related to *Chlamydomonas* at all, and were designated as a new genus and species: *Sanguina nivaloides* [@Prochazkova2018c].
-Other genetic work shows that morphology-based species of snow algae may in fact contain multiple genetically unrelated clades: for example, *Chloromonas cf.
-nivalis* contains at least four distinct clades [@Matsuzaki2019].
-
-<!-- hammer this home: WE LACK BASIC UNDERSTANDING ABOUT SPECIES COMPOSITOIN AND DIVERSITY -->
-
-<!-- sequence cultures, and batch, and single cell, but this dosent tell us whats actually in the field -->
-More recently, snow algae taxonomy has
-
-<!-- high throughput: lacking ref data, no otu based, none in alpine environments (despite this tantalizing evidence of microscopy based, etc) -->
-
-<!-- But no one until now has used high throughput sequencing to compare snow algae blooms in alpine environments, or used higher resolution marker genes to do so -->
-
-<!-- so we did this... -->
-In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS (explain what these code for).
-Our aims were to: 1) see what species were present, 2) compare community composition between samples.
-We found ...
-
-<!-- the red green problem -->
-The same species can look completely different depending on the life stage [@Shah2016;@Salome2019]; conversely, different species can look nearly identical.
-For example, cell morphologies known as *Chloromonas cf.
-nivalis* appear in at least four different lineages.
-One of these lineages, *Chloromonas muramotoi* appear as green biflagellates in culture, but in the field the same species features thick cell walls with flanges [@Matsuzaki2019].
-However, life cycle changes in the the cosmopolitan and widespread genus *Sanguina* have yet to be documented: despite over a century of study [@Kol1961] no one has sucessfully cultured this genus.
-
-<!-- habitat differences -->
-Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013].
-In this micro-habitat, temperature is nearly constant, but snow moisture, pH, nutrients, and light can differ greatly within the snowpack depending on location of snow and time of year [@Loranger1988;@Kuhn2001;@Gorton2007;@Ganey2017].
-Differences in snow algae community composition could offer insight into the environmental factors that determine snow algae distribution.
-Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a].
-Two studies using microscopy found that red snow containing *Chlamydomonas cf.
-nivalis* was prevalent in the above timberline, while green and orange snows caused by *Chloromonas cf.
-brevispina* and *Chloromonas cf.
-nivalis* were prevalent in conifer forests below timberline [@Hoham1979a;@Nedbalova2008a].
-Differences in snow moisture could influence which species of snow algae grow; several species of snow algae are anecdotally reported to only inhabit very saturated, waterlogged snow [@Lutz2015;@Remias2013;@Novis2008;@Prochazkova2018c], while  *Chlamydomonas cf.
-nivalis* is hypothesized to be specially adapted to well-drained, drier snow [@Ganey2018].
-
-<!-- habitat differences, short new form -->
-  Red snow is also occasionally reported to be caused by another red-pigmented algae: *Chlainomonas* [@Novis2008].
-Little is known about the distribution of this genus; one study suggested that *Chlainomonas* is limited to high-moisture snowpack above alpine lakes [@Prochazkova2018].
-Green snow may contain a variety of algae, notably *Chloromonas* (refs), as well as algae from other families including Trebouxiophyceae and Chrysophyceae (refs).
-Morphology-based cell counts suggest that green snow caused by *Chloromonas* is prevalent in forested areas and red snow containing *Chlamydomonas cf.
-nivalis* is prevalent in alpine areas [@Hoham1979a;@Nedbalova2008a]; however, morphology can be misleading when identifying microalgae.
-
-Although much of the snow algae literature focusses on eye-catching red snow, different species of snow algae can also color snow green [@Lutz2015].
-
-<!-- hts -->
-Microscopic observation shows that snow algae blooms can contain a diversity of cell morphologies in the same sample [@Lutz2018], but the genetic community structure of snow algae may not correlate with what is seen under the microscope.
-18S amplicon high-throughput sequencing (HTS) shows that snow algae may contain more diversity than meets the eye: although red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a].
-18S is highly conserved and therefore suited to survey broadly across a variety of taxa; however, due to its high similarity between closely related species short segments of thie gene are poorly-differentiated among closely related taxa (ref).
-The *rbc*L plastid gene coding for the large subunit of RuBisCO is highly differentiated among photosynthetic organisms, and contains reference sequences for snow algae, making it an ideal complement to survey snow algae communities.
-
-Until recently, snow algae sequences were lacking in reference databases,
-
-# Results {-}
-
-**A gradient between snow algae communities**.
-Variation in snow algae communities was primarily driven by the presence or absence of *Sanguina nivaloides*, *Chloromonas krienitzii*, *Chlainomonas rubra*, or *Chlainomonas* OTU3 (Fig.
-2a).
-However, most samples contained a mix of species; few samples were relatively "pure".
-However, sample comparison with weighted UniFrac [@Lozupone2005] found little difference between *Chlainomonas* and *Chloromonas* dominated sites, and the presence or absence of *Sanguina* was the driving factor, which was generally found in higher relative abundance at higher elevation (Fig.
-2b).
-
-Although *Chlainomonas* is reported to only grow in slushy snow above alpine tarns [@Novis2008;@Prochazkova2018c], we found *Chlainomonas* was present in low relative abundance at most sites sampled.
-Only one site high in Chlainomonas was near water (S9), the other Chlainomonas dominated sites were not near water, nor was the snow noticeably mushier than the surrounding snow.
-This does not rule out the possibility that these sites contained high moisture at some point; further sampling is needed to determine its habitat distribution.
-
-#
-
-**Snow algae community composition co-varies with date and elevation**.
-*Sanguina* was more abundant in alpine sites, as well as one OTU of *Chlainomonas* and two OTUs of *Chloromonas*, while *Chloromonas krienitzii* and Chlainomonas OTU6 were found in high relative abundance below treeline (Fig.
-2, 3).
-This consistent with the microscopy observations of @Hoham1979a and @Nedbalova2008a, who found primarily *Chlamydomonas cf.
-nivalis* above treeline and *Chloromonas* below treeline.
-However, in microscopy surveys we infrequently observed *Sanguina* below treeline, and *Chloromonas krienitzii* above treeline.
-We found higher diversity in samples from higher elevation, both at the ASV level and OTU level <!--Chao1 w fossil.-->, with several OTUs that were not found at lower elevations.
-*Raphidonema* and related ASVs were restricted to high alpine sites, with the highest relative abundance occurring in samples from August and September.
-This fits with the interpretation of @Stibal2005, who suggest that *Raphidonema* is an opportunistic soil algae that blows into snow fields from surrounding dirt to colonize the snow.
-However, rare taxa should not be discounted as playing an important role in ecosystem functioning [@Jousset2017].
-
-Snowpack above and below treeline are very different habitats, with forest providing shade (and therefore less evaporation).
-The red pigment of *Sanguina* and *Chlainomonas* (astaxanthin) is likely an adaptation to high light [@Gorton2003;@Holzinger2006], and it has been suggested that *Sanguina* is adapted to the drier snowpacks that form at higher elevations [@Dial2018c].
-
-#
-
-**_Chloromonas krienitzii_ is green below, turns orange at surface**.
-On initial visits to Chloromonas krienitzii dominated sites, green snow was hidden 2 to 5 cm below the surface of the runnel (Supp.
-Fig.
-4).
-Green snow was noticeably moister than surrounding snow, in some locations containing ice lenses not found in the surrounding snowpack.
-By June 6 the surface of the runnel had turned orange while the snow below remained green, which persisted on three subsequent visits until the snow melted.
-Cell morphology at the surface closely resembled *Chloromonas cf.
-brevispina* “Trochiscia” [@Matsuzaki2015] (Fig.
-4), while green subsurface snow was dominated by green biflagellate or bipolar quadriflagellate cells (Fig.
-4).
-Both green and orange samples were dominated by reads assigned to *C.
-krienitzii*, with lower relative abundance of *Chlainomonas*, which was more abundant at the surface (Supp.
-Fig.
-3).
-
-One possibility is that C.
-krienitzii cells orange pigmentation is triggered or an adaptation to higher light levels at the surface.
-Green cells below the surface would be protected from UV radiation<!--cite -->.
-Possibly, the green cells maintain an optimal light level through phototaxis, similar to marine phytoplankton<!--cite-->.
-Perhaps higher elevations typically have fewer low angle runnels that would allow a moist enough environment for *C.
-krienitzii*.
-It seems unlikely that nutrients from conifer needles would explain this, as several of our *C.
-krienitzii* dominated sites were in full sunlight, far from any trees.
-
-#
-
-**Snow algae blooms harbor novel diversity**.
-Several OTUs were not assigned to species level due to low similarity (Fig.
-1).
-Two OTUs of Chloromonas, one OTU of Chlainomonas, and one OTU ("Algae OTU8") that had best BLAST matches to both Trebouxiophyceae and Chlorophyceae algae.
-Algae OTU8 has a best BLAST score of only 88%; this could represent a free-living algae found in snow, or possibly an endosymbiotic algae from a ciliate, as we noted many ciliates with green organelles in our microscopy samples.
-Sample X17 contained was dominated by spherical red cells superficially similar to *Sanguina*, this sample was dominated by reads of Chloromonas OTU7 and Algae OTU8.
-Cells in this sample had thicker cell wall, and some cells were slightly square-shaped (Fig.
-4).
-18S algae reads from this sample were equivocal, with best BLAST matches to both *Raphidonema* and *Chloromonas*.
-Currently there are just two species *Chlainomonas* in reference databases, but our results suggest there may be genetically distinct populations *Chlainomonas* at higher elevations (Fig.
-1, Supp.
-Fig.
-1, Supp.
-Fig 2).
-Although some of the high elevations ASVs were classified as *Chlainomonas rubra*, because we only included this species in our custom snow algae reference this could have biased our classifier, possibly causing over-classification [@Murali2018].
-
-Although our rbcL primers successfully detected distinct OTUs, this is likely an underestimate of the true diversity.
-We did not detect any clustering within *Sanguina* using either 18S or rbcL, however when we included the homologous rbcL fragment of *S.
-aurentia* in our analysis it did not form a distinct OTU <!--this is where I am quesy about out ghetto OTU method-->.
-Furthermore, rbcL is highly differentiated in *Chloromonas* relative to other microalgae [@Nozaki2002], which could partly explain why *Sanguina* is less differentiated than *Chloromonas* or *Chlainomonas* (Supp.
-Fig.
-1 and 2).
-
-# Discussion {-}
-
-We found snow algae communities were typically dominated by either *Sanguina*, *Chlainomonas* or *Chloromonas*, with secondary abundance of *Chlainomonas* or *Chloromonas* in most samples.
-Most samples contained a sizable proportion of secondary or tertiary OTUs, although some some contained nearly 100% *Chlainomonas rubra*, *Sanguina nivaloides*, or *Chloromonas krienitzii*.
-The drivers of these differences between communities remain unclear; it is likely that within the melting snowpack there may be distinct microhabitats that impacts snow algae distribution, with varying water content with topography, different light levels in shade or at depth [@Gorton2007], different daily freeze-thaw cycles, or possibly varying nutrient levels with proximity to vegetation [@Hoham1976] or bedrock [@Kol1968a].
-Further work needs to be done to uncover the micro-habitat niches within the snowpack that allow this differentiation, and what novel species the snow microbiome might contain.
-
-# Methods {-}
-
-**Field sampling and microscopy.** We collected snow algae from mountains near Vancouver, British Columbia throughout the summer of 2018.
-To capture the extent of snow algae diversity in our local mountains we collected as many samples as possible from different elevations, dates, mountains, and micro-habitats within the snow <!--insert suppl table-->.
-In total we collected 310 colored snow samples from 13 different mountains on 33 different dates from elevations between 880 m and 2150 m above sea level.
-In early season we sampled at lower elevations, moving uphill as the snow algae bloom progressed upwards in elevation [@Takeuchi2013].
-We observed the first small spots of red snow above treeline on June 20.
-
-We scooped samples from visibly colored snow into 50 mL centrifuge tubes using sterile technique.
-To prevent melting en route to the lab we stored tubes in a bag of snow.
-Back in the lab, we melted each sample at room temperature on the bench and removed a 1 mL aliquot for light microscopy.
-Immediately after, samples were stored at -20 &deg;C for up to eight months until DNA extraction.
-Most samples were viewed using light microscopy, characterizing the dominant cell morphology in each sample.
-We prepared a slide of cells fixed in 2% gluteraldehyde, and quantified the most common cell morphologies with a cell count to 100, classifying each cell as either *Sanguina*, *Chloromonas cf.
-nivalis*, *Chloromonas cf.
-brevispina*, *Chloromonas krienitzii* *Chlainomonas*, or "other".
-
-#
-
-**DNA extraction and amplicon library preparation.** We selected 35 out of 310 samples for rbcL and 18s high-throughput amplicon sequencing.
-We chose these samples to represent the variation in date, elevation, geographic location, snow color, micro-habitat, and cell morphology.
-We freeze-dried these samples for up to 48 hours until until samples appeared completely desiccated.
-To allow water vapor to escape sample tubes while freeze drying we poked holes in each lid; as a safeguard against cross-contamination from airborne snow algae powder we wrapped tubes individually with paper towels.
-We mini-pestled 5 mg of sample at room temperature to physically rupture cell walls.
-Based on our observation that samples dominated by *Sanguina* cell morphologies had lower DNA yields, and had many intact cells following various lysis methods, we used 20 mg of these samples to increase our DNA yield.
-
-To extract DNA from the crushed cells we added 800 mL 1x CTAB extraction buffer [@ColdSp2009], 1% B-mercaptoethanol, 5 $\mu$L each of proteinase K and RNAase A, and incubated these at 65 &deg;C for 30 minutes.
-We centrifuged samples at 10,000 g for 3 minutes to pellet cell debris, then added 700 $\mu$L of supernatant to an equal volume of 24:1 chloroform:isoamyl alcohol.
-We inverted samples to mix and centrifuged again at 12,000 g for 10 minutes.
-We precipitated DNA by transferring the top layer to 700 $\mu$L ice-cold ethanol, which we gently mixed, and spun over Qiagen DNA columns for 30 s at 15,000 g.
-We washed columns twice with 70% ethanol, and finally dissolved our DNA by spinning with 50 $\mu$L of sterile TE buffer.
-As a negative control we processed a sterile distilled water sample alongside each batch, treating it exactly the same as the other tubes.
-
-We used rbcL as a marker gene, which offers high resolution between microalgae species [@Zou2016], and has reference data for snow algae available on GenBank.
-We also sequenced each sample with 18S primers, due to it's coverage across a wide range of taxa, and its ubiquity in reference databases.
-We designed rbcL primers to target an approximately 400 bp section of this gene, based on 20 snow algae rbcL GenBank sequences from *Chloromonas* and *Chlainomonas* (GenBank accession numbers AB434272.1, LC012752.1, LC012747.1, AF517072.1, LC012738.1, LC012739.1, AB434267.1, EU030690.1, LC360494.1, AJ001878.1, AB022225.1, DQ885964.2, DQ885962.1, AJ001879.1, AB022226.1, AB022530.1, LC012751.1, AB504764.1, EU030689.1, AB101508.1).
-*Sanguina* sequences were not included because they were not available at the time.
-We designed primers with PRIME(+) from the GCG package.
-These primers were rbcL369F (5’-GAA CGT GAC AAA TTA AAC AAA-3’) and rbcL870R (5’-ACC WGA YAD ACG WAG AGC TT-3’).
-To target 18S we used Euk1181 (5'-TTA ATT TGA CTC AAC RCG GG-3') and Euk1624 (5'-CGG GCG GTG TGT ACA AAG G-3') [@Wang2014].
-
-We constructed our amplicon library using a two-step PCR [@Meyer2010].
 In the first PCR we amplified template DNA using our primers attached to a universal adapter, and in the second PCR we re-amplified that product to attach a 6 bp index to the universal adapter at the 3’ end.
-The first PCR total volume was 25 $\mu$L, consisting of 1 $\mu$L template, 12.5 $\mu$L Q5 high-fidelity 2X MM (New England BioLabs), 1.25 $\mu$L each of forward and reverse primer, and 9 $\mu$L of ddH2O.
+The first PCR total volume was 25 $\mu$L, consisting of 1 $\mu$L template, 12.5 $\mu$L Q5 high-fidelity 2X master mix (New England Biolabs), 1.25 $\mu$L each of forward and reverse primer, and 9 $\mu$L of ddH20.
 The second PCR was the same except we reduced our reaction volume to 20 $\mu$L by using only 5 $\mu$L of ddH2O.
@@
 For the second indexing PCR we started with an initial denaturation at 98 &deg;C for 30 s, then 10 cycles of 98 &deg;C for 10 s, 65 &deg;C for 30 s, and 72 &deg;C for 30 s with a final denaturation of 72 &deg;C for 5 min.
-After each PCR, we purified using Agencourt AMPure XP kit (Beckman Coulter).
-We quantified final DNA concentration with Qubit (Thermo Fisher), and standardized sample concentration for pooling.
-The pooled library was then loaded and run on an Illumina MiSeq V3 kit.
-
-#
-
-**Bioinformatic processing.** Reads were demultiplexing using cutadapt [@Martin2011].
-Samples were filtered and trimmed, errors removed, dereplicated, pair end reads merged, and chimeras removed in DADA2 [@Callahan2017].
-<!-- singletons, failed samples removed, remove sequences based on negative controls, adjust for 0's, -->
-
-We assigned taxonomy using a custom reference databases for both rbcL and 18S, since snow algae are not well-represented in SILVA or PR2, and there is no curated database for rbcL.
-We made our rbcL database by downloading 30,865 algae rbcL sequences from GenBank with the query “rbcl[gene] AND (algae OR chlorophyta OR trebouxiophyceae) NOT (18s OR ribosomal OR psaB OR atpB) AND 300:2000[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion”.
-We removed ambiguous snow algae annotations based on the most recent snow algae phylogeny [@Matsuzaki2019].
-Because entries labelled as *Chloromonas cf.
-brevispina* and *Chloromonas cf.
-nivalis* are polyphyletic we re-labeled these as “unassigned *Chloromonas*”.
-<!--Where multiple sequences were present for a single species we generated a consensus sequence to avoid imbalance where some groups have far more representatives than others [@Murali2018] --><!--To  re-labeled GenBank accessions LC012733.1 and LC012735.1 as Chloromonas krienitzii, LC012734.1 as Chloromonas “Cryocystis”-type, and LC012744.1 to Chloromonas miwae, and LC012743 and LC012745 to “Chloromonas Group_B” 6`.--> Species designated by morphology alone that were not genetically different from other species in their genera were annotated only to genus level.
-<!-- all Chlainomonas to genus level only --> We assigned rbcL ASV taxonomy using IDTaxa with threshold = 50.
-
-To assign 18S reads we initially ran our ASVs against SILVA [@Quast2013], and selected ASVs assigned to Chlorophyta.
-We then ran this subset of ASVs on our custom 18S snow algae database based on the following GenBank query: "(18S OR ribosomal)[gene] AND (chloromonas OR chlainomonas OR sanguina OR raphidonema OR KMY-2018) NOT (rbcL OR psaB OR atpB) AND 300:3500[slen] NOT plasmid NOT unverified NOT mRNA NOT bacteria NOT mitochondrion".
-This database was edited to reflect the most recent snow algae taxonomy and remove ambiguous annotations [@Matsuzaki2019;@Prochazkova2018c].
-We assigned taxonomy using IDTaxa with threshold = 50.
-
-We assigned OTUs with t-SNE, based on visually distinct clusters [@Graham-Rowe1999] <!--BibTeX header messed up-->(Fig.
-1).
-We split the largest cluster (containing the vast majority of the *Chloromonas* reads) into three clusters, one containing all ASVs assigned to *Chloromonas krienitzii*, and one containing all ASVs assigned to *Chloromonas muramotoi* and *Chloromonas hohamii*.
-To highlight our observation that one clade of *Chlainomonas* were only found above treeline (Suppl.
-Fig.
-1), we split *Chlainomonas* into two OTUs.
-The OTUs produced by this naive method were roughly consistent across different methods of visualization (Suppl.
-Fig.
-1 and 2).
-<!--Other OTU classifiers such swarm [@Mahe2014] are optimized for 18S data, and we found swarm did a poor job of clustering when visualised with MDS, t-SNE, and phylogenetic trees, with various values of d.
-<!--and CROP [@Chen2013].-->
-<!--OTU clustering with t-SNE: based on poor results with other classifiers, we used tsne.
-since tsne is known to find patterns where there are none depending on the perplexity, we ran this test with perplexity ranging from 5-100, and found the same large scale clusters each time.
--->
-
-**Software information.** All analysis was conducted in R 3.6.1 with tidyverse 1.2.1, with dada2 1.12.1, DECIPHER 2.12.0, Rtsne 0.15, GUniFrac 1.1, and vegan 2.5-6.
-Phylogenetic analysis was carried out in IQTree [@Nguyen2015] and visualized with ggtree [@Yu2017].
-
-**Data availability.** All raw fastq files are freely available on the European Nucleotide Archive under the project accession PRJEB34539.
-All scripts are available at cengstro.github.io/projects/sa_rbcl.
-
-##### pagebreak
-
-# Acknowledgments {-}
-
-We wish to thank Leah Tooman (Simon Fraser University) for assistance with sequencing, and Chris Rushton (Simon Fraser University) for assistance with bioinformatics.
-<!--We also wish to thank Forest Rohwer (UCSD) for the inspiration that led to this project.--> This project was funded with a Sector Innovation Grant from Genome BC (SIP016), and a NSERC Individual Discovery Grant, both awarded to LQ.
-
-# Author Contributions {-}
-
-<!--
-
-When determining authorship the following criteria should be observed:
-
- - Substantial contributions to the conception or design of the work; or the
-   acquisition, analysis, or interpretation of data for the work; AND
- - Drafting the work or revising it critically for important intellectual
-   content; AND
- - Final approval of the version to be published ; AND
- - Agreement to be accountable for all aspects of the work in ensuring that
-   questions related to the accuracy or integrity of any part of the work are
-   appropriately investigated and resolved.
-
-Contributors who meet fewer than all 4 of the above criteria for authorship
-should not be listed as authors, but they should be acknowledged.
-(http://www.icmje.org/roles_a.html)
-
--->
-
-CE, LQ, and KY designed this study.
-Samples collected by CE with assistance from KY and LQ.
-KY and CE prepped samples for sequencing.
-All analyses were completed by CE and KY.
-All authors discussed the results and contributed to the final manuscript.
-CE wrote the manuscript with major input from LQ and KY.
-<!-- The statement about the authors and contributors can be up to several sentences -->
-<!-- long, describing the tasks of individual authors referred to by their initials -->
-<!-- and should be included at the end of the manuscript before the References -->
-<!-- section. -->
-
-# Additional information {-}
-
-### *Supplementary information*.
-
-### *Competing financial interests.* The authors declare that the research was conducted in the absence of any
-commercial or financial relationships that could be construed as a potential
-conflict of interest.
-
-##### pagebreak
-
-# References {-}
-<div id="refs"></div>
-
-# Figures {-}
-
-<br>
-<br>
-
-![Figure 1.
-t-SNE clustering of snow algae rbcL amplicon sequence variants (ASVs).
-Colored points correspond to genus and species level assignment, grey points were not assigned at or below genus level.
-Labelled ellipses show OTUs.
-Perplexity = 30, theta = 0.5](/home/cengstro/projects/sa_rbcl/figs/asv_tsne_fig.png)
-
-<!-- todo: italics, underscores, ChlorKri5, ChlorMura5, put ChloroKri4 on top of Chloro3, so its obvious that these aren't part of both OTUs.-->
-
-##### pagebreak
-
-![Figure 2.
-Sample ordination plots.
-A.
-NMDS comparing sample similarity using OTUs, shaded by elevation.
-Vectors indicate relative contribution of taxa driving differences between samples.
-Stress = 0.183.
-B.
-NMDS showing weighted UniFrac distances between snow algae samples.
-Stress = 0.11.](/home/cengstro/projects/sa_rbcl/figs/sample_ordination.png){width=13.01cm, height=18.56cm}
-
-<!-- add dom morph groupings for UniFrac to connect with heatmap.  adjust size.  -->
-<!-- change axis labs of b, MDS to NMDS-->
-
-##### pagebreak
-
-![Figure 3.
-Relative abundance heatmap of snow algae OTUs in each sample.
-Weight of color corresponds to the relative abundance of that OTU within a sample.
-Samples (rows) are grouped by the dominant cell morphology within that sample as determined by light microscopy.
-Within each morphology group samples are ordered by elevation, from low on the bottom to high on top.
-Sample IDs follow the convention: letters represent the mountain, and the number represents the order in which it was collected.
-Samples collected on the same day are labelled as .1, .2, and so on.](/home/cengstro/projects/sa_rbcl/figs/heatmap.png){width=12.88cm, height=17.17cm}
-
-<!-- add elev range, date range. dom cell morph text to grey. lump chlainomonas categories (let it be a surprise)-->
-<!-- add treeline -->
-```{r, echo = F, include=F}
-# otu_heat2 %>% 
-#   group_by(bloom_type) %>% 
-#   summarise(min(elev_m), max(elev_m))
-# otu_heat2 %>% 
-#   group_by(bloom_type) %>% 
-#   summarise(min(date), max(date))
-```
-
-##### pagebreak
-
-![Figure 4.
-Select microphotographs of snow algae.
-All photos are taken with light microscopy, except (f) which was taken with DIC.
-Scale bars represent 30 $\mu$m.
-(a) H5 (b) S10.2.
-(c) S4.
-(d) K2.
-(e) S10.1.
-(f) G16.1.
-(g) N12.1.
-(h) P20.1.](/home/cengstro/projects/sa_rbcl/figs/fig4.png){width=13.51cm, height=17.49cm}
-<!-- order by elev, add sax1 -->
-
-<!-- ![Figure 5. Relative abundance barplots comparing community composition in surface orange snow and green snow below in three samples.] -->
-<!-- choose photo, microphotos, rethink colors (make consistant w heatmap) -->
-
-##### pagebreak
-
-# Supplementary Data {-}
-
-![Supplementary Figure 1.
-Maximum likelihood tree (IQTree) of snow algae rbcL ASVs.
-Colors correspond to OTU.
-Edges with bootstrap values < 80 are displayed as dotted lines.](/home/cengstro/projects/sa_rbcl/figs/ggtree.png)
-
-##### pagebreak
-
-![Supplementary Figure 2. MDS plot of rbcL ASVs, colored by taxa. Ellipses contain all ASVs within an OTU. Stress =  ](/home/cengstro/projects/sa_rbcl/figs/asv_mds.png)
-
-<!-- redo with NMDS, include stress vals -->
-##### pagebreak
-
-![Supplementary Figure 3.
-Relative abundance barplots comparing community composition in surface orange snow and subsurface green snow.
-* designates depth and surface samples collected on different days.](/home/cengstro/projects/sa_rbcl/figs/draft1/depth.png)
-
-##### pagebreak
-
-<!-- ![Supplementary Figure 4. Photos showing location of *Chloromonas krienitzii* dominated samples.]() -->
-
-<!-- word to template: *[:alpha:] to \*, and *[:space:] to } -->

### 2020-01-11 manuscript/supp/code/cellcount_stacked_barplot.R

@@ -1,0 +1,37 @@
+# cell count pie charts
+
+library(tidyverse)
+library(here)
+
+
+cellct <- read_csv(here("data/02_tidied/tidied.cell_count.csv"))
+field <- read_csv(here("data/02_tidied/tidied_field.csv"))
+
+cellct <- cellct %>% 
+  left_join(field, by="sample_id")
+
+
+colors <- c("Chlainomonas rubra"="#ac71b5", #purple
+            "Sanguina nivaloides"="#ff534a", #red
+            # Chloromonas: pastel greens / blues / yellows
+            "Chloromonas krienitzii"="#ffa54c",
+            "Chloromonas cf. brevispina"="#80c786", "Chloromonas cf. nivalis"="#4aaf86", 
+            #                 lt green                            teal
+            "Other Chloromonas"="#4aaf53", "Chloromonas"="#4aaf53", 
+            # greyish light green
+            "Green cell"="#4aaf53", "Other"="grey60") #greys
+# oops, ordered them backwards, too lazy to rewrite
+colors <- colors %>% rev()
+
+
+cellct_plot <- ggplot(cellct, aes(x=alias %>% fct_reorder(elev_m), 
+                     y=percent_of_sample)) +
+    geom_bar(aes(fill=morpho_sp %>% fct_relevel(names(colors))), position="stack", stat="identity") +
+    theme_classic() +
+    labs(x="Sample ID", y="Relative abundance", fill="Cell morphology")+
+    scale_fill_manual(values = colors) +
+    theme(text = element_text(size=8),
+          axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))  
+cellct_plot
+
+ggsave(filename = here::here("manuscript/supp/output/fig_S4_cellct.tiff"), width = 300, height = 100, units="mm", dpi=300)

### 2020-01-11 manuscript/supp/code/map/download_google_map_data.R

@@ -1,0 +1,45 @@
+# get map data
+# 1-10
+
+library(tidyverse)
+library(googleway)
+library(ggmap)
+library(png)
+
+# download data from googleway: do only once to avoid API overcharges
+key <- "AIzaSyBJsmyAKTkTg9LT-HdGCCdZdbE2Hjwm-L8"
+register_google(key = key)
+
+# get map
+googlemap <- get_googlemap(c(-123.004722, 49.850556), zoom = 7, maptype = "satellite")
+my_map <- ggmap(googlemap)
+my_map
+# get city coords -----------------------
+cities <- c("Vancouver, BC, Canada", "Squamish, BC, Canada")
+city_coords <- cities %>% 
+  map(~googleway::google_geocode(.x, key=key))
+glimpse(google_output) # ugly lists
+
+# functions to reach into the list and extract wanted elements
+city_name <- function(x) x$results$address_components[[1]][1,2]
+lat <- function(x) x$results$geometry$location[1,1]
+lon <- function(x) x$results$geometry$location[1,2]
+
+# make tbl with city and coords
+city_coords <- google_output %>% {
+  tibble(
+    name = map_chr(., .f=city_name),
+    lat = map_dbl(., .f=lat),
+    lon = map_dbl(., .f=lon)
+  )
+}
+
+city_coords # a pretty tbl
+
+# save data
+ggsave(filename = here::here("data/map_downloads/googlemap.png"), plot=my_map, 
+       width = 225, height = 225, units="mm", dpi = 300) # didnt end up using this
+save(googlemap, file=here::here("data/map_downloads/googlemap.rda")) # ended up using this... next time, dont bother with clunky API, download the map data directly as png ("do one thing and do it well")
+
+write_csv(city_coords, here::here("data/map_downloads/map_data_cities.csv"))
+

### 2020-01-11 manuscript/supp/code/tree.R

@@ -1,0 +1,80 @@
+# make tree for supp
+# 2020-01-10
+
+library(tidyverse)
+library(here)
+library(ape)
+library(ggtree)
+
+
+# ML phylogenetic tree (IQTree) of rbcL ASVs that appear in filtered relative abundance dataset (613 tips)
+tree <- read.tree(file=here::here("output/tree/rbcl/asv613.fasta.treefile")) 
+# Taxonomy tbl of each ASV (652; includes ASVs filtered out in samples not used in final analysis)
+taxonomy <- read_csv(here::here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
+# OTU of each ASV
+otus <- read_csv(here::here("output/asv_otu_key.csv"))
+
+
+# Misc edits of tree annotation data
+
+# rename one of the OTUs (should take care of this upstream)
+otus <- otus %>% 
+  mutate(otu_name= otu_name %>% str_remove("muramotoi "))
+
+taxonomy <- taxonomy %>% 
+  mutate(assignment = assignment %>% str_remove("root"))
+
+
+
+
+# Plot tree ---------
+
+# re-root tree
+root <- taxonomy %>% 
+  filter(class=="Trebouxiophyceae") %>% 
+  head(1) %>% 
+  pull(asv_id)
+
+rbcl.tree <- rbcl.tree %>% root(outgroup = root, resolve.root=T)
+
+rbcl_ggtree <- ggtree(rbcl.tree) # ignore warning message 
+
+# format bayesian/bootstrap confidence for each node
+boots <- rbcl_ggtree$data %>% 
+  filter(isTip == FALSE) %>% 
+  separate(label, c("boot", "bayes"), sep = "/") %>% #split label into boots and bayes vals
+  filter(boot > 80) %>%
+  mutate(boot = round(as.numeric(boot), digits = 0),
+         insignificant_edge = if_else(boot > 80, "black", "red"))# round boot to nearest integer
+
+# format taxonomy, otu for each node
+assignment <- taxonomy %>% 
+  select(asv_id, assignment)
+annotation <- rbcl_ggtree$data %>% 
+  left_join(assignment, by=c("label"="asv_id")) %>% 
+  left_join(otus, by=c("label"="asv_id"))
+
+# specify colors
+rbcl_otu_colors <- c("Chlainomonas A"="#ac71b5",
+                     "Sanguina B"="#ff534a", 
[…]

### 2020-01-11 manuscript/casey1.Rmd
responded to KY edits 1-10

@@
   #   number_sections: yes
-csl: the-isme-journal.csl
+csl: templates/the-isme-journal.csl
 bibliography: refs.bib
@@
 <!-- metabarcoding -->
-In contrast to the high diversity of snow algae species observed in culture and Sanger sequencing, metabarcoding suggests that snow algae blooms are highly similar across continents, and are dominated by few taxa.
-33 red snow communities across the Arctic were remarkably similar to each other: all were dominated by two operational taxonomic units (OTUs) of "uncultured Chlamydomonadaceae 1 and 2" with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale* [@lutz_biogeography_2016].
+In contrast to the high diversity of snow algae species observed in culture, metabarcoding suggests that snow algae blooms are highly similar across continents, dominated by few taxa.
+One study of 33 red snow communities across the Arctic found all samples were dominated by two OTUs of "uncultured Chlamydomonadaceae 1 and 2" with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale* [@lutz_biogeography_2016].
 Similarly, another study using ITS2 metabarcoding found 24 red snow sites across the Arctic and Antarctic contained similar assemblages, dominated by "uncultured Chlamydomonadaceae A and B" with secondary abundance of *Raphidonema* and *Chloromonadinia* [@segawa_bipolar_2018].
@@
 <!-- our goal, what we did, what we found -->
-Virtually nothing is known about the regional variation in species composition of snow algae blooms.
-Our goal was to identify the algae in blooms on different mountains and elevations in the Coast Range of British Columbia, Canada.
-We used 18S and *rbcL* metabarcoding to assess species composition---the latter targeting a hypervariable region of *rbcL* (coding for large subunit of rubisco), thus targeting only photosynthetic species with high taxonomic resolution.
-Additionally, we used light microscopy to describe the relative abundance of different morphologies in each field sample.
-By using these three cross-referenced metrics of relative abundance we were able to describe the algal species assemblage with high accuracy.
+Virtually nothing is known about the regional variation in species composition of snow algae blooms, particularly in alpine regions.
+We set out to identify the algae in blooms on different mountains and elevations in the Coast Range of British Columbia, Canada.
+We used 18S and *rbcL* metabarcoding to assess species composition---the latter targeting a hypervariable region of *rbcL* (coding for large subunit of Rubisco), thus targeting only photosynthetic species with higher taxonomic resolution than 18S.
+We also used light microscopy to describe the relative abundance of different morphologies in each field sample.
+Using these three cross-referenced metrics we described the algal species assemblage in our study area with high accuracy.
 Using *rbcL* as a marker revealed much greater diversity than would have been found using 18S alone.
 We did not find any differences in bloom composition between mountains, but there were distinct algae species assemblages at high versus low elevations.
-<!-- stronger hook, predictions??? cliff hangers -->
 
@@
 
+<!-- the problem with this is we cherry picked our samples.
+We weren't doing transects or anything like that.
+A reviewer could say, you couldve picked your samples that supported your hypothesis about different bloom types.
+which is why I brought in the cell count data.
+-->
+
 We collected snow algae from sites throughout the Coast Range near Vancouver, British Columbia, Canada over the summer of 2018 (Supplementary Figure S1).
@@
 In total, we collected 309 colored snow samples from a range of 880 - 2150 m elevation, including samples from 13 different mountains collected on 33 different dates (Supplementary Table S1).
-Our samples trended upwards in elevation over the course of the summer as the snow melted earliest at the lowest elevations (Supplementary Figure S2).
+Our samples trended upwards in elevation over the course of the summer as the snow melted earliest at the lowest elevations.
 We first detected snow algae above 1500 m on June 20.
@@
 We mini-pestled between 5 to 20 mg of sample at room temperature to physically rupture cell walls before incubation in CTAB lysis buffer [@noauthor_ctab_2009].
-We extracted DNA using chloroform:isoamyl alcohol [@cubero_dna_1999], and spin columns (Qiagen, Hilden) (detailed protocol in Supplementary File S1).
+We extracted DNA using chloroform:isoamyl alcohol [@cubero_dna_1999], and spin columns (Qiagen, Hilden) (details available in Supplementary Protocol S1).
 As a negative control we processed a sterile distilled water sample alongside each batch.
 
-<!-- arent we following some extraction protocol here? cant we cite, and then have a supplementary protocol with our variation??? -->
-
-<!-- To extract DNA from the crushed cells we added 800 mL  , 1% B-mercaptoethanol, 5 $\mu$L each of proteinase K and RNAase A, and incubated these at 65 &deg;C for 30 minutes.
-We spun samples at 10,000 g for 3 minutes to pellet cell debris, then added 700 $\mu$L of supernatant to an equal volume of 24:1 chloroform:isoamyl alcohol.
-We inverted samples to mix and centrifuged again at 12,000 g for 10 minutes.
-We precipitated DNA by transferring the top layer to 700 $\mu$L ice-cold ethanol, which we gently mixed, and spun over Qiagen DNA columns for 30 s at 15,000 g.
-We washed columns twice with 70% ethanol, and finally dissolved our DNA by spinning with 50 $\mu$L of sterile TE buffer.
--->
-
-We designed custom primers to target a hypervariable region of snow algae *rbcL* (coding for the large subunit of rubisco).
-*rbcL* is more differentiated between snow algae than 18S (Supplementary Figure S3), and contains similar number of reference sequences on GenBank.
-Our *rbcL* primers target a 400 bp section of snow algae *rbcL* based off the consensus of 20 GenBank snow algae sequences (Supplementary Table S2).
+We designed custom primers to target a hypervariable region of snow algae *rbcL*.
+*rbcL* is more differentiated between snow algae than 18S (Supplementary Figure S2), and contains similar number of reference sequences on GenBank.
+Our *rbcL* primers target a 400 bp section of snow algae *rbcL* based off the consensus of 20 GenBank snow algae sequences (Supplementary Figure S2).
 We designed our primers with the Eurofins primer design tool (https://www.eurofinsgenomics.eu/en/ecom/tools/pcr-primer-design/).
@@
 To target 18S, we used the universal primers Euk1181 and Euk1624 targeting the V7-V8 hypervariable regions [@wang_optimal_2014].
-Primer sequences are available in Supplementary Table S3.
+Primer sequences are available in Supplementary Table S2.
 
 We constructed our 18S and *rbcL* amplicon libraries using a standard two-step PCR approach [@meyer_illumina_2010].
-For detailed PCR conditions, see Supplementary File S2.
+For detailed PCR conditions, see Supplementary Protocol S2.
 After each PCR, we purified using Agencourt AMPure XP kit (Beckman Coulter, Brea, CA).
@@
 We assigned taxonomy using custom snow algae reference databases for both 18S and *rbcL*, in addition to re-assigning 18S with SILVA to classify non-snow algae [@quast_silva_2013].
-Snow algae reference databases and the scripts to generate them are available on our github repository (see below).
+Snow algae reference databases and the scripts to generate them are available on our github repository.
 
@@
 We compared *rbcL* sample similarity using UniFrac with default settings [@lozupone_unifrac_2005].
-To confirm that our clustering was not an artifact of t-SNE we created maximum likelihood phylogenetic trees of ASVs using IQTree [@nguyen_iq-tree:_2015] (Supplementary Figures S4-S5).
-
-All raw sequence data are available under European Nucleotide Archive accession PRJEB34539. All scripts used in this study are available at https://github.com/cengstro/bc_snow_algae_amplicon.
+To confirm that our clustering was not an artifact of t-SNE we created maximum likelihood phylogenetic trees of ASVs using IQTree [@nguyen_iq-tree:_2015] (Supplementary Figure S3).
 
 # Results {-}
+<!-- should we report the diversity data quickly? or tangential to our question... -->
 
@@
 
-We observed morphologically distinct snow algae communities at different elevations (Figure 1).
+We observed morphologically distinct snow algae communities at different elevations (Figure 1; Supplementary Figure S4).
 Red snow was prevalent in areas of high solar exposure above treeline, predominantly containing cells similar in appearance to *Sanguina nivaloides* [@prochazkova_sanguina_2019].
@@
 Negative controls did not contain Qubit-detectable amounts of DNA.
-Our *rbcL* library contained higher ASV and OTU diversity than 18S: t-SNE clustering generated 9 *rbcL* OTUs, compared with just 4 18S OTUs (Supplementary Figure S6).
-The most frequent *rbcL* taxonomic assignments at genus level were *Chloromonas*, *Chlainomonas*, and *Sanguina*.
+Our *rbcL* library contained higher ASV and OTU diversity than 18S, t-SNE clustering generated 9 *rbcL* OTUs, while we detected just 3 visually distinct clusters among 18S ASVs (Supplementary Figure S5).
+For *rbcL*, the most frequent taxonomic assignments at genus level were *Chloromonas*, *Chlainomonas*, and *Sanguina*.
 Although most ASVs were not assigned to genus level, most ASVs clustered near one of these genera in NMDS ordination (Figure 2).
-Several *rbcL* OTUs were dominated by ASVs that were not assigned to genus level, including three OTUs closely related to *Chloromonas*, and one Chlamydomonadaceae OTU that was only assigned to the family level.
+Several OTUs were dominated by ASVs that were not assigned to genus level, including three OTUs closely related to *Chloromonas*, and one Chlamydomonadaceae OTU that was only assigned to the family level.
 
@@
 
-Overall, our results show bloom composition can be highly variable on the same mountain, even in nearby patches.
-This is in contrast to previous papers that found snow algae assemblages were homogeneous across geographically distant sites [@lutz_biogeography_2016;@segawa_bipolar_2018].
-This could be due to different sampling schemes: the aforementioned studies exclusively sampled high-latitude red snow, while our study included mid-latitude red, orange, and green snow from above and below treeline.
-Another recent study using metabarcoding found different snow algae communities in red and green snow algae blooms in Antarctica, including yellow blooms dominated by Chrysophyceae [@soto_microbial_2020].
+Our results show bloom composition can be highly variable on the same mountain, even in nearby patches.
+In contrast, previous studies found snow algae assemblages were homogeneous across geographically distant sites [@lutz_biogeography_2016;@segawa_bipolar_2018].
+The aforementioned studies exclusively sampled high-latitude red snow, while our study included mid-latitude red, orange, and green snow from above and below treeline.
+Another recent study using 18S metabarcoding found different snow algae communities in red and green snow algae blooms in Antarctica, including yellow blooms dominated by Chrysophyceae [@soto_microbial_2020].
 Much of the diversity in our samples within *Chlainomonas* and *Chloromonas* was only detected with *rbcL* and not 18S, so the diversity within these genera may have been overlooked in previous metabarcoding studies.
 
-Our *rbcL* data suggest *Chlainomonas* may be more common than previously thought.
+Our *rbcL* data suggest *Chlainomonas* may be more widely distributed than previously thought.
 We could not differentiate between *Chlainomonas* and *Chloromonas* in our 18S data set, so this genus may have been overlooked in previous studies.
@@
 The lack of clustering we observed in the remaining *Chloromonas* could be due to overlap in the levels of interspecific and intraspecific variation.
-*rbcL* diversity is likely higher within *Chloromonas* than in other algae genera: unlike *Sanguina* [@prochazkova_sanguina_2019], most species of *Chloromonas* lack a pyrenoid [@nozaki_differences_2002], which houses high concentrations of cross-linked rubisco.
-Likely related to its lack of pyrenoid, *Chloromonas* has many non-synonymous mutations in the region of *rbcL* the codes for binding rubisco together [@nozaki_differences_2002].
+*rbcL* diversity is likely higher within *Chloromonas* than in other algae genera: unlike *Sanguina* [@prochazkova_sanguina_2019], most species of *Chloromonas* lack a pyrenoid [@nozaki_differences_2002], which houses high concentrations of cross-linked Rubisco.
+Likely related to its lack of pyrenoid, *Chloromonas* has many non-synonymous mutations in the region of *rbcL* the codes for binding Rubisco together [@nozaki_differences_2002].
 While *rbcL* is problematic in *Chloromonas* phylogeny [@nozaki_differences_2002], it nonetheless is highly differentiated and therefore an effective barcode for microalgae [@hall_assessment_2010].
@@
 In conclusion, our study reveals substantial diversity within and between snow algae blooms.
-We found greater diversity using *rbcL* metabarcoding than 18S, which suggests snow algae diversity may have been overlooked in studies that used only 18S.
+We found greater diversity using *rbcL* metabarcoding than 18S, which suggests snow algae diversity may be overlooked in studies using only 18S.
 Species were differentially distributed along an elevational gradient, with distinct blooms dominated by *Chloromonas*, *Sanguina*, or *Chlainomonas*.
 Future studies will examine other taxa present in snow algae blooms, and their interactions in these globally important microbiomes.
+
+<!-- This needs to be beefed up.
+Some sort of statement about how this work impacts the field, more specifics about future directions.
+How this work could inform albedo work that you mention in your intro.
+-->
+
+# Data availability {-}
+
+All raw sequence data are available under European Nucleotide Archive accession PRJEB34539.
+All scripts used in this study are available at https://github.com/cengstro/bc_snow_algae_amplicon.
+Supplementary materials are available at https://www.nature.com/ismej/.
 
@@
 
-# Author Contributions {-}
+# Author contributions {-}
 
@@
 
-# Additional information {-}
+# Conflict of interest {-}
 

### 2020-01-11 code/exploratory/negative_controls.Rmd
major reorg

---
title: "Negative controls"
output: html_notebook
---

```{r}
library(tidyverse)
library(here)
```

```{r}
neg_controls <- read_csv(here("data/03_filtered/rbcl/water_controls_rbcl_rel_abund.csv"))
taxonomy <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
otus <- read_csv(here("output/asv_otu_key.csv"))

# source(here("functions/ggplot_functions.R"))  ### not yet working
```

Which batches were these extracted alongside? check notebook

```{r}
joined <- neg_controls %>% 
  left_join(taxonomy, by="asv_id") %>% 
  left_join(otus, by="asv_id")

joined %>% distinct(sample_id)
```

```{r by_otu, echo=F}
otu_rel_abund <- joined %>% 
  group_by(otu_name, sample_id) %>% 
  summarise(rel_abund = sum(rel_abund))

ggplot(otu_rel_abund, aes(x=sample_id, y=rel_abund, fill=otu_name)) +
  geom_bar(stat="identity") +
  labs(x="Sample ID", y="Relative abundance", fill="OTU", tag="D")
```

### 2020-01-11 code/exploratory/whole_18s_community_compare.Rmd
major reorg

--- 2019-11-02 code/whole_18s_community_compare.Rmd
@@
 dir.create(here("manuscript/figs/S5_18s"))
-ggsave(filename = here("manuscript/figs/S5_18s/18s_rel_abund.pdf"), plot=rel_abund_phyla)
+# ggsave(filename = here("manuscript/figs/S5_18s/18s_rel_abund.pdf"), plot=rel_abund_phyla)
 ```

### 2020-01-11 code/misc/download_google_map_info.R

@@ -1,44 +1,0 @@
-# get google map
-
-library(tidyverse)
-library(here)
-library(googleway)
-library(ggmap)
-
-
-# get base map --------------
-key="AIzaSyBJsmyAKTkTg9LT-HdGCCdZdbE2Hjwm-L8"
-register_google(key = key) 
-
-algae_map <- get_googlemap(c(-123.004722, 49.850556), zoom = 7, maptype = "satellite")
-
-
-# get city data ------------------
-cities <- c("Vancouver, BC, Canada", "Squamish, BC, Canada")
-
-# get coordinates from google
-google_output <- cities %>% 
-  map(~googleway::google_geocode(.x, key=key))
-glimpse(google_output)
-
-# functions to reach into the list and extract wanted elements
-city_name <- function(x) x$results$address_components[[1]][1,2]
-lat <- function(x) x$results$geometry$location[1,1]
-lon <- function(x) x$results$geometry$location[1,2]
-
-# make tbl with city and coords
-city_coords <- google_output %>% {
-  tibble(
-    name = map_chr(., .f=city_name),
-    lat = map_dbl(., .f=lat),
-    lon = map_dbl(., .f=lon)
-  )
-}
-
-head(city_coords)
-
-# save data
-dir.create(here::here("output/map"))
-
-save(algae_map, file = here::here("output/map/raw_google_map.rda"))
-write_csv(city_coords, here::here("output/map/city_coords.csv"))

### 2020-01-11 functions/ggplot_functions.R

@@ -1,36 +1,0 @@
-library(tidyverse)
-
-##### none of this is working
-
-otus %>% distinct(otu_name)
-
-
-otu_ord <- c("Chlainomonas A","Sanguina B", "Other Chloromonas","Chloromonas krienitzii","Chloromonas muramotoi C","Chloromonas D", "Algae E","Chloromonas F", "Trebouxiophyceae G")
-
-barplot <- 
-
-
-plot_rel_abund_bar <- function(rel_abund_tbl, species, species_order, sample_order){
-  # input: long form tibble with
-  cust_theme <- theme(legend.position = "none",
-                      panel.grid.major.x = element_blank(), # remove vertical lines in background
-                      panel.grid.minor.x = element_blank(),
-                      axis.line=element_blank(), # remove x axis
-                      axis.title.x=element_blank(), 
-                      axis.text.x=element_blank(),
-                      axis.ticks.x=element_blank(),
-                      panel.spacing = unit(-0.01, "lines"), # remove space between facet panels
-                      strip.text.x = element_text(angle = 90, hjust=1.0, vjust = 0.8), # rotate facet labels 90
-                      strip.background = element_blank()) # remove grey boxes around facet labs
-  
-  ggplot(rel_abund_tbl, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
-    geom_bar(aes(fill=otu_name %>% fct_relevel(otu_ord)), stat="identity") + # classic stacked relative abundance plot
-    coord_flip() +
-    facet_grid(~otu_name %>% fct_relevel(otu_ord), switch = "x") +
-    labs( x="Sample ID", tag="B") +
-    cust_theme
-}
-
-plot_rel_abund_stacked_bar <- function(){
-  
-}

### 2020-01-11 reference_database/FlatToFastaIdTaxa.py

@@ -1,102 +1,0 @@
-#!/usr/bin/env python
-
-import argparse
-import os
-
-
-def is_valid_file(file, parser):
-    """
-    Checks to determine if the specified file exists
-
-    :param file: A string containing a filepath
-    :param parser: An argparse.ArgumentParser object
-    :return: file, if the file exists
-    :raises: parser.error, if the file does not exists
-    """
-
-    if os.path.exists(file):
-        return file
-    else:
-        raise parser.error("Unable to locate %s: No such file or directory" % file)
-
-
-def get_args():
-    """
-    Processes command line arguments
-
-    :return: A namespace containing command line arguments
-    """
-    parser = argparse.ArgumentParser(description="Converts one or more NCBI FLAT files into a single FASTA file")
-    parser.add_argument("-i", "--input", metavar="FLAT", required=True, nargs="+",
-                        type=lambda x: is_valid_file(x, parser), help="Input FLAT file(s)")
-    parser.add_argument("-o", "--output", metavar="FASTA", required=True, help="Output merged FASTA file")
-    return parser.parse_args()
-
-
-def main(args=None):
-    if args is None:
-        args = get_args()
-
-    # Process each FLAT file
-    with open(args.output, "w") as o:
-        for file in args.input:
-            with open(file) as f:
-
-                sequence = ""
-                taxonomy = ""
-                organism = ""
-                accession = ""
-
-                while True:
-                    line = f.readline()
-                    if line == "":  # There are no more lines in this file
-                        break
-                    line = line.strip("\n").strip("\n")  # Remove newline
-
-                    # Parse the required attributes from the FLAT file
-                    if line.startswith("ORIGIN"):  # Sequence
-
-                        while True:
-                            line = f.readline()
-                            line = line.strip("\r").strip("\n")  # Remove newline
-                            if line == "//":  # The // is a control character to end the FLAT entry
-                                # Format all the file features into a FASTA entry
-                                # The format is >taxonomy; Organism; Accession
-                                fasta_name = ">Root;" + taxonomy + organism.replace(" ", "_") + ";" + accession
-                                o.write(fasta_name)
-                                o.write(os.linesep)
-
-                                # Sequence
-                                o.write(sequence)
-                                o.write(os.linesep)
-
-                                # Reset variables
-                                sequence = ""
-                                taxonomy = ""
-                                organism = ""
-                                accession = ""
-                                break
-
-                            # Remove all spaces and numbers
-                            line = line.replace(" ", "")
[…]

### 2020-01-11 code/exploratory/community.Rmd
major reorg

--- 2019-12-28 code/community.Rmd
@@
 
-```{r rotated-stacked-bar, echo=F, fig.width=7, fig.height=5}
+```{r rotated-stacked-bar, echo=F}
 
@@
                 # greyish light green
-                "Trebouxiophyceae G"="#6a2d2f", "Trebouxiophyceae"="#6a2d2f", # maroon, standout
+                "Trebouxiophyceae G"="#975c5c", "Trebouxiophyceae"="#975c5c", # brown
                 "Unknown E"="grey60", "Green cell"="#4aaf53", "Other"="grey60") #greys
-
-# teal #69c095
-# medium blue #6994c0
 
@@
 
-# plot
+# plot  facet_grid(rows = vars(marker %>% fct_relevel(c("Cell morphology","18S","rbcL"))), scales="free_x", space="free_x") +
 stacked_bar <- ggplot(combined, aes(x=alias %>% fct_reorder(elev_m), y=rel_abund)) +
   geom_bar(aes(fill=species %>% fct_relevel(names(species_colors))), position="stack", stat="identity") +
-  coord_flip() +
-  facet_grid(cols = vars(marker %>% fct_relevel(c("Cell morphology","18S","rbcL"))), 
-             scales="free_y", space="free_y") +
+  # coord_flip() +
+  facet_grid(rows = vars(marker %>% fct_relevel(c("rbcL","18S", "Cell morphology")))) +
   theme_classic() +
-  labs(x="Sample ID", y="Relative abundance")+
-  scale_fill_manual(values = species_colors)
+  labs(x="Sample ID", y="Relative abundance", fill="Taxa")+
+  scale_fill_manual(values = species_colors) +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
 
@@
 
-ggsave(here::here("manuscript/figs/stack.pdf"), plot=stack_no_legend, width=7, height=5)
-ggsave(here::here("manuscript/figs/stack_legend.pdf"), plot=legend, width=7, height=5)
+ggsave(here::here("manuscript/figs/main/community/stack2.pdf"), plot=stack_no_legend, width=7, height=5)
+ggsave(here::here("manuscript/figs/main/community/stack2_legend.pdf"), plot=legend, width=7, height=5)
+ggsave(here::here("manuscript/figs/main/community/stack_full.pdf"), plot=stacked_bar, width=180, height=180, units = "mm", dpi=300)
 

### 2020-01-12 manuscript/supp/code/18s_asv_clust.Rmd
tweak figures

@@
 ```{r, include=F}
-aligns <- readDNAStringSet(here::here("data/asv_seq_data/algae_18s_quality_aligns.fasta"))
+aligns <- readDNAStringSet(here::here("data/fastas/asv_seq_data/algae_18s_quality_aligns.fasta"))
 # taxonomy w genbank (multiple copies of each designation, no editing)
-taxonomy <- read_csv(here::here("output/algae_18s_assignments_snow_algae_genbank.csv"))
-tree <- ape::read.tree(here::here("output/tree/algae_18s/algae_18s_seqs.fasta.treefile"))
-rel_abund <- read_csv(here::here("data/03_filtered/18s_algae/18s_algae.filtered_cleaned_rel_abund.csv"))
+taxonomy <- read_csv(here::here("data/taxonomy/algae_18s_assignments_snow_algae_genbank.csv"))
+tree <- ape::read.tree(here::here("data/tree/algae_18s/algae_18s_seqs.fasta.treefile"))
+rel_abund <- read_csv(here::here("data/rel_abund/3_filter/18s_algae/18s_algae_rel_abund.csv"))
 
@@
 
+Perplexity = 20.
+
 \pagebreak

### 2020-01-12 manuscript/figs/code/fig_4_unifrac_nmds.R

@@ -1,0 +1,74 @@
+# make UniFrac NMDS
+
+library(tidyverse)
+library(here)
+library(GUniFrac)
+library(vegan)
+library(ggrepel)
+library(picante)
+library(gridExtra)
+
+
+rel_abund <- read_csv(here::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv")) # not looking at the depth samples, controls, etc
+tax <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
+field <- read_csv(here::here("data/field/tidy_field.csv"))
+asv_seqs <- read_csv(here::here("data/fastas/asv_seq_data/rbcl_asv_sequence_key.csv"))
+otus <- read_csv(here::here("data/otus/otu_rbcl_asvs.csv"))
+rbcl.tree <- read.tree(here::here("data/tree/rbcl/asv613.fasta.treefile")) ### this tree should only contain the asvs we found in the "main" sites, not looking at controls, etc
+
+
+ra_wide <- rel_abund %>% 
+  select(sample_id, asv_id, rel_abund) %>% 
+  arrange(asv_id) %>% 
+  pivot_wider(names_from = asv_id, values_from = rel_abund, values_fill = list(rel_abund = 0)) %>% 
+  column_to_rownames("sample_id")
+
+# root tree with Trebouxiophyceae
+root <- tax %>% 
+  filter(class=="Trebouxiophyceae") %>% 
+  head(1) %>% 
+  pull(asv_id)
+rbcl.tree <- rbcl.tree %>% root(outgroup = root, resolve.root=T)
+
+
+# check for differences between asvs in tree and rel abund table (all names must appear in tree)
+# asvs_rel_abund <- colnames(ra_wide)
+# asvs_tree <- rbcl.tree$tip.label %>% sort()
+# 
+# length(asvs_rel_abund)
+# length(asvs_tree)
+# 
+# setdiff(asvs_rel_abund, asvs_tree) # output elts unique to asvs_rel_abund
+
+# run weighted UniFrac
+set.seed(123)
+unifracs <- ra_wide %>% 
+  GUniFrac::GUniFrac(rbcl.tree, alpha=1) # ignore error
+
+unifrac <- unifracs$unifracs[, , "d_1"]
+
+
+# run NMDS ordination on results 
+set.seed(123)
+nmds <- metaMDS(unifrac, autotransform = F)
+
+
+# plot
[…]

### 2020-01-12 manuscript/supp/code/cellcount_stacked_barplot.R

@@ -4,8 +4,8 @@ library(tidyverse)
 library(here)
 
 
-cellct <- read_csv(here("data/02_tidied/tidied.cell_count.csv"))
-field <- read_csv(here("data/02_tidied/tidied_field.csv"))
+cellct <- read_csv(here::here("data/cellct/tidied_cellct.csv"))
+field <- read_csv(here::here("data/field/tidy_field.csv"))
 
 cellct <- cellct %>% 
   left_join(field, by="sample_id")
@@ -28,10 +28,10 @@ cellct_plot <- ggplot(cellct, aes(x=alias %>% fct_reorder(elev_m),
                      y=percent_of_sample)) +
     geom_bar(aes(fill=morpho_sp %>% fct_relevel(names(colors))), position="stack", stat="identity") +
     theme_classic() +
-    labs(x="Sample ID", y="Relative abundance", fill="Cell morphology")+
+    labs(x="Sample ID (in order of increasing elevation)", y="Relative abundance", fill="Cell morphology")+
     scale_fill_manual(values = colors) +
     theme(text = element_text(size=8),
           axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))  
 cellct_plot
 
-ggsave(filename = here::here("manuscript/supp/output/fig_S4_cellct.tiff"), width = 300, height = 100, units="mm", dpi=300)
+# ggsave(filename = here::here("manuscript/supp/final/fig_S4_cellct.tiff"), width = 300, height = 100, units="mm", dpi=300)

### 2020-01-12 manuscript/supp/code/tree.R

@@ -1,4 +1,4 @@
-# make tree for supp
+# Make rbcl ASV tree
 # 2020-01-10
 
 library(tidyverse)
@@ -13,6 +13,7 @@ tree <- read.tree(file=here::here("output/tree/rbcl/asv613.fasta.treefile"))
 taxonomy <- read_csv(here::here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
 # OTU of each ASV
 otus <- read_csv(here::here("output/asv_otu_key.csv"))
+rel_abund <- read_csv("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv")
 
 
 # Misc edits of tree annotation data
@@ -47,17 +48,23 @@ boots <- rbcl_ggtree$data %>%
   mutate(boot = round(as.numeric(boot), digits = 0),
          insignificant_edge = if_else(boot > 80, "black", "red"))# round boot to nearest integer
 
+# get rel abund for each asv
+rel_abund_asv <- rel_abund %>% 
+  group_by(asv_id) %>% 
+  summarise(rel_abund=sum(rel_abund))
+
 # format taxonomy, otu for each node
 assignment <- taxonomy %>% 
   select(asv_id, assignment)
 annotation <- rbcl_ggtree$data %>% 
   left_join(assignment, by=c("label"="asv_id")) %>% 
-  left_join(otus, by=c("label"="asv_id"))
+  left_join(otus, by=c("label"="asv_id")) %>% 
+  left_join(rel_abund_asv, by=c("label"="asv_id"))
 
 # specify colors
 rbcl_otu_colors <- c("Chlainomonas A"="#ac71b5",
                      "Sanguina B"="#ff534a", 
-                      "Chloromonas krienitzii"="#ffa54c","Chloromonas C"="#87b1d4", "Chloromonas D"="#94cf92", "Chloromonas F"="#fdfd96", "Other Chloromonas"="#4aaf53", 
+                      "Chloromonas krienitzii"="#ffa54c","Chloromonas muramotoi C"="#87b1d4", "Chloromonas D"="#94cf92", "Chloromonas F"="#fdfd96", "Other Chloromonas"="#4aaf53", 
                      "Trebouxiophyceae G"="#975c5c",
                      "Unknown E"="grey60")
 
@@ -65,16 +72,17 @@ rbcl_otu_colors <- c("Chlainomonas A"="#ac71b5",
 # plot tree
 otu_tree <- rbcl_ggtree +
[…]

### 2020-01-12 manuscript/supp/code/diversity.Rmd
tweak figures

---
title: "Diversity"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=F}
library(tidyverse)
library(here)
library(vegan)
library(picante)
library(gridExtra)
library(ape)
library(ggrepel)
library(corrr)
library(lubridate)
```

```{r, include=F}
rel_abund <- read_csv(here::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv")) # not looking at the depth samples, controls, etc
tax <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
field <- read_csv(here::here("data/field/tidy_field.csv"))
asv_seqs <- read_csv(here::here("data/fastas/asv_seq_data/rbcl_asv_sequence_key.csv"))
otus <- read_csv(here::here("data/otus/otu_rbcl_asvs.csv"))
rbcl.tree <- read.tree(here::here("data/tree/rbcl/asv613.fasta.treefile")) ### this tree should only contain the asvs we found in the "main" sites, not looking at controls, etc
```

```{r, include=F}
# get relative abundance of each otu, in wide format, 
otu_rel_abund <- rel_abund %>% 
  left_join(otus, by="asv_id") %>% 
  group_by(sample_id, otu_name) %>% 
  summarise(rel_abund = sum(rel_abund)) %>%   
  ungroup() %>% 
  pivot_wider(names_from = otu_name, values_from = rel_abund, values_fill = list(rel_abund=0)) %>% 
  column_to_rownames("sample_id") # vegan::diversity() wants sample identifiers as rownames

# calculate shannon and simpson
shannon <- otu_rel_abund %>% 
  diversity(index = "shannon") %>% 
  enframe() %>% 
  dplyr::rename(sample_id = name, diversity = value) %>% 
  mutate(type="shannon")
# simpson <- otu_rel_abund %>% 
#   diversity(index = "simpson") %>% 
#   enframe() %>% 
#   dplyr::rename(sample_id = name, simpson = value)
# otu_diversity <- shannon %>% 
#   left_join(simpson, by="sample_id") %>% 
#   left_join(field, by="sample_id")

# plot diversity of samples
shannon_plot <- ggplot(shannon %>% left_join(field), aes(x=date, y=diversity))+
  geom_point(aes(color=elev_m), size=3) +
  scale_color_gradientn(colours = terrain.colors(8)[1:7]) +
  geom_text_repel(aes(label=alias)) +
  labs(y="Shannon diversity")+
  theme(axis.title.x = element_blank())
shannon_plot
```

```{r, include=F}
# format
ra_wide <- rel_abund %>% 
  select(sample_id, asv_id, rel_abund) %>% 
  arrange(asv_id) %>% 
  pivot_wider(names_from = asv_id, values_from = rel_abund, values_fill = list(rel_abund = 0)) %>% 
  column_to_rownames("sample_id")

ra_wide_mat <- ra_wide %>% as.matrix()

# root tree with Trebouxiophyceae
root <- tax %>% 
  filter(class=="Trebouxiophyceae") %>% 
  head(1) %>% 
  pull(asv_id)

rbcl.tree <- rbcl.tree %>% ape::root(outgroup = root, resolve.root=T)

faiths <- ra_wide_mat %>% 
  picante::pd(rbcl.tree) %>% 
  # attach elevation data
  rownames_to_column("sample_id") %>% 
  rename(diversity=PD) %>% 
  mutate(type="faiths") %>% 
  select(-SR)

faith_plot <- ggplot(faiths %>% left_join(field), aes(y=diversity, x=date))+
  geom_point(aes(color=elev_m), size=3) +
  scale_color_gradientn(colours = terrain.colors(8)[1:7]) +
  geom_text_repel(aes(label=alias))+
  labs(y="Faith's phylogenetic diversity (PD)",x="Date", color="Elevation (m)")

faith_plot
```

```{r diversity_plots, echo=F}
combined <- shannon %>% 
  bind_rows(faiths) %>%
  left_join(field, by="sample_id") %>% 
  mutate(type=str_to_title(type))

ggplot(combined, aes(x=date, y=diversity)) +
  geom_point(aes(color=elev_m), size=3) +
  scale_color_gradientn(colours = terrain.colors(8)[1:7]) +
  geom_text_repel(aes(label=alias), size=2) +
  facet_grid(rows = "type", scales="free_y") +
  labs(y="Diversity", x="Date", color="Elevation (m)")
```

```{r, include=F}
date <- field %>% 
  select(sample_id, date) %>% 
  mutate(date = date %>% lubridate::yday())

cor_df <- faiths %>% 
  rename(faiths = diversity) %>% 
  select(-type) %>% 
  left_join(shannon) %>% 
  rename(shannon = diversity) %>% 
  select(-type) %>% 
  left_join(date)

cor.test(cor_df$faiths, cor_df$date)

cor.test(cor_df$shannon, cor_df$date)

```

**Supplementary figure S6.** Faiths phylogenetic diversity (upper) and Shannon diversity (lower) of snow algae *rbcL* ASVs, plotted against date of sample collection. Samples colored by elevation, and labelled by sample ID. Faith's Pearson's *r*=0.36, p=0.04, Shannon's Pearson's *r*=0.17, p=0.35.

### 2020-01-12 manuscript/figs/code/incomplete_fig_2_rbcl_asvs.R

@@ -1,0 +1,22 @@
+# Fig 2 rbcL ASVs
+
+library(tidyverse)
+library(here)
+library(DECIPHER)
+library(ggforce)
+library(Rtsne)
+library(dbscan)
+library(ggrepel)
+
+
+rel_abund <- read_csv(here::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv"))
+taxonomy <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
+aligns <- readDNAStringSet(here::here("data/fastas/alignments/rbcl_aligns_in_main.fasta"))
+seq <- read_csv(here::here("data/fastas/asv_seq_data/asv_seq_key_18s.csv"))
+                           
+
+# make distance matrix
+BrowseSeqs(aligns, highlight = T)
+
+dists <-  aligns %>% 
+  DECIPHER::DistanceMatrix(type="matrix", correction = "JC")

### 2020-01-13 manuscript/casey1.Rmd
draft to lab, made krienitzii a small supplemental

@@
 ---
-title: "Alpine snow algae blooms vary with elevation"
+title: "Alpine snow algae species composition along an elevational gradient"
 author:
@@
 bibliography: refs.bib
-
 ---
@@
 
+The authors declare no conflicts of interest.
+
 <!-- search for ??? before send out -->
@@
 Despite their global prevalence, little is known about the algae species that comprise snow algae blooms.
-We used 18S and *rbcL* metabarcoding and light microscopy to characterise species composition of 33 snow algae samples from alpine and subalpine habitats in the Coast Range of British Columbia, Canada.
+We used 18S and *rbcL* metabarcoding and light microscopy to characterize species composition of 33 snow algae samples from alpine and subalpine habitats in the Coast Range of British Columbia, Canada.
 We found blooms were dominated by the snow algae genera *Sanguina*, *Chloromonas*, and *Chlainomonas*.
-Above treeline blooms contained *Sanguina*, *Raphidonema*, and one OTU of *Chloromonas*, distinct from species assemblages below treeline predominated by *Chloromonas krienitzii*.
-Previously unrecognized as an important component of the snow algae microbiome, *Chlainomonas* were abundant in samples across all elevations.
-Our *rbcL* dataset detected diversity that was obscured by the low taxonomic resolution of traditional 18S metabarcoding, including OTUs that could represent novel species of snow algae.
+This study is the first to thoroughly document regional variation between blooms, and we found that blooms were highly variable in species composition, with high elevation sites containing distinct species assemblages from forested sites.
+In contrast to previous work, the snow algae genus *Chlainomonas* was an abundant and widespread in snow algae blooms.
+Although we found few taxa using traditional 18S metabarcoding, *rbcL* revealed diversity with high taxonomic resolution, including OTUs that could represent novel species of snow algae.
 These three cross-referenced datasets show alpine snow algae blooms are more diverse than previously thought, and are highly variable from site to site along an elevational gradient.
@@
 <!-- global, vast blooms. albedo->snowmelt. what species? -->
-Each summer in polar and alpine snowfields worldwide, vast areas of snow surface are colored red by snow algae blooms.
-Such blooms are reported from every continent [@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019] and Arctic sea ice [@gradinger_snow_1996].
-They can be extensive: in Alaska, snow algae spectral signatures were detected in one third of a 1,900 km^2^ icefield [@ganey_role_2017].
-In recent years snow algae have received attention for their role in reducing snow surface albedo, which could substantially impact snow melt rates [@lutz_biogeography_2016;@ganey_role_2017].
-Thus, snow algae could impact the timing of spring melt and runoff, depleting summer water supplies held in mountain snowpacks and reducing glacier mass balance.
-Despite their global ubiquity, we are only beginning to identify the species that cause snow algae blooms.
+Each summer, vast areas of snow surface are coloured red by snow algae blooms in polar and alpine snowfields worldwide.
+Red snowfields have been found on every continent [@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019] and overlying Arctic sea ice [@gradinger_snow_1996].
+The distribution and extent of snow algae blooms are poorly understood, but they can be quite extensive: in Alaska, remote sensing detected snow algae in one third of a 1,900 km^2^ icefield [@ganey_role_2017].
+In recent years snow algae have received attention for their role in reducing snow surface albedo, which could substantially impact the rate of snow melt [@lutz_biogeography_2016;@ganey_role_2017].
+Potentially, snow algae could substantially impact the timing of spring melt, which in turn could deplete summer water supplies in mountain snowpack and reduce glacier mass balance.
+Snow algae blooms have been recorded throughout history since the time of the ancient Romans [@mayhoff_naturalis_1906;@darwin_voyage_1959], yet despite their global prevalence we are only beginning to identify the species that cause snow algae blooms.
 
-<!-- why microscopy not reliable -->
-Much of the previous work on snow algae relied on cell morphology for identification, but microscopy-based identification of snow algae can be unreliable.
-Different species of snow algae can look nearly identical: recent phylogenetic work found that four distinct field samples dominated by similar-looking cells previously referred to as *Chloromonas cf.
-nivalis* are actually genetically distinct clades [@matsuzaki_taxonomic_2019].
-Furthermore, the same species can look completely different in different environmental conditions or life stages.
-For example, cultured *Chloromonas krienitzii* cells are small green biflagellates, but field samples cells of this species are larger orange spheres with thick walls and short spines [@matsuzaki_revisiting_2015].
-In the freshwater green algae *Haematococcus pluvialis* red secondary pigments are produced in response to high light [@shah_astaxanthin-producing_2016], but genetic comparison of red and green snow found they contained different species [@terashima_microbial_2017;@lutz_integrated_2015].
+Microscopy reveals a diversity of cell morphologies in snow algae blooms, but different species can look nearly identical [@matsuzaki_taxonomic_2019], and the same species can look completely different under different environmental conditions.
+The snow algae *Chloromonas krienitzii* are small green biflagellates in culture, but in field samples cells of this species are nearly twice the diameter, orange-pigmented, with thick cell walls and short spines [@matsuzaki_revisiting_2015].
+The environmental cues that trigger this transformation are not well understood, but in the freshwater green algae *Haematococcus pluvialis* red secondary pigments are produced in response to high light [@shah_astaxanthin-producing_2016].
+Green snow algae blooms are less frequently reported, and microscopy-based studies have suggested that green snow develops into red snow [@muller_persistent_2001].
+However, metabarcoding studies to date have found that green snow and red snow contain different species [@terashima_microbial_2017;@lutz_integrated_2015].
 
@@
 
-<!-- metabarcoding -->
-In contrast to the high diversity of snow algae species observed in culture, metabarcoding suggests that snow algae blooms are highly similar across continents, dominated by few taxa.
-One study of 33 red snow communities across the Arctic found all samples were dominated by two OTUs of "uncultured Chlamydomonadaceae 1 and 2" with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale* [@lutz_biogeography_2016].
-Similarly, another study using ITS2 metabarcoding found 24 red snow sites across the Arctic and Antarctic contained similar assemblages, dominated by "uncultured Chlamydomonadaceae A and B" with secondary abundance of *Raphidonema* and *Chloromonadinia* [@segawa_bipolar_2018].
-Other 18S metabarcoding studies of snow algae were limited to genus or family level taxonomy [@hamilton_primary_2017]---because 18S is highly conserved, this marker generally cannot distinguish between closely related taxa.
+<!-- metabarcoding, red snow looks the same -->
+Although there are many species of snow algae, only a few OTUs seem to dominate red snow algae blooms worldwide.
+33 Arctic red snow samples were all dominated by "uncultured Chlamydomonadaceae 1 and 2" OTUs with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale* [@lutz_biogeography_2016].
+Similarly, another study using metabarcoding found 24 red snow sites across the Arctic and Antarctic contained similar assemblages, dominated by "uncultured Chlamydomonadaceae A and B" with secondary abundance of *Raphidonema* and *Chloromonadinia* [@segawa_bipolar_2018].
+Other 18S metabarcoding studies of snow algae were limited to genus or family level taxonomy [@hamilton_primary_2017]---because 18S is highly conserved, this marker cannot distinguish between closely related species or genera.
 
@@
 We set out to identify the algae in blooms on different mountains and elevations in the Coast Range of British Columbia, Canada.
-We used 18S and *rbcL* metabarcoding to assess species composition---the latter targeting a hypervariable region of *rbcL* (coding for large subunit of Rubisco), thus targeting only photosynthetic species with higher taxonomic resolution than 18S.
-We also used light microscopy to describe the relative abundance of different morphologies in each field sample.
-Using these three cross-referenced metrics we described the algal species assemblage in our study area with high accuracy.
-Using *rbcL* as a marker revealed much greater diversity than would have been found using 18S alone.
-We did not find any differences in bloom composition between mountains, but there were distinct algae species assemblages at high versus low elevations.
+We assessed species composition using light microscopy and 18S and *rbcL* metabarcoding.
+*rbcL* codes for the large subunit of Rubisco, thus targeting only the photosynthetic component of the microbiome.
+Using these three cross-referenced metrics we describe the algal species assemblage in our study area with high accuracy.
+*rbcL* revealed much greater diversity than would have been found using 18S alone.
+We did not find any differences in bloom composition between mountains, but there were distinct algae species assemblages at different elevations.
 
@@
 
-<!-- refer to each supp at least once in txt -->
-<!-- add cities, states -->
-
 ## Field sampling and microscopy {-}
 
-<!-- the problem with this is we cherry picked our samples.
-We weren't doing transects or anything like that.
-A reviewer could say, you couldve picked your samples that supported your hypothesis about different bloom types.
-which is why I brought in the cell count data.
--->
+We collected 309 snow algae samples over the summer of 2018 from alpine and subalpine sites in the Coast Range near Vancouver, British Columbia, Canada (Supplementary Figure S1).
+We collected red, orange, and green snow samples from 13 different mountains from sites ranging from 880 to 2150 m above sea level (Supplementary Table S1).
+Over the summer we collected samples from progressively higher elevations as snow melted at lower elevations, and we did not detect snow algae at higher elevations until later in the summer.
+We scooped samples from visibly coloured snow into sterile 50 mL centrifuge tubes.
+We kept samples cold during transport back to the lab by packing tubes in snow.
+We melted each sample at room temperature on the bench and removed a 1 mL aliquot for light microscopy.
+Samples were then stored at -20 &deg;C for up to eight months until DNA extraction.
 
-We collected snow algae from sites throughout the Coast Range near Vancouver, British Columbia, Canada over the summer of 2018 (Supplementary Figure S1).
-To capture the extent of snow algae diversity we sampled from different elevations, dates, mountains, and micro-habitats within the snow.
-In total, we collected 309 colored snow samples from a range of 880 - 2150 m elevation, including samples from 13 different mountains collected on 33 different dates (Supplementary Table S1).
-Our samples trended upwards in elevation over the course of the summer as the snow melted earliest at the lowest elevations.
-We first detected snow algae above 1500 m on June 20.
-We scooped samples from visibly colored snow into sterile 50 mL centrifuge tubes.
-To prevent melting en route to the lab we stored tubes in a bag of snow.
-We melted each sample at room temperature on the bench and removed a 1 mL aliquot for light microscopy.
-Immediately after, samples were stored at -20 &deg;C for up to eight months until DNA extraction.
-
-We used light microscopy to characterize the cell morphologies in each sample within 24 hours following collection.
-We fixed samples in 2% gluteraldehyde, and under 400x magnification observed the 100 cells closest to the center of the slide.
+We used light microscopy to characterize cell morphology within 24 hours after collection.
+We fixed samples in 2% gluteraldehyde, and under 400 x magnification observed the 100 cells closest to the center of the slide.
 We classified cell morphology based on similarity to published photographs of *Sanguina nivaloides* [@prochazkova_sanguina_2019], *Chloromonas cf.
@@
 
-<!-- how do select field and micro photos fit into this? -->
+<!-- collected a ton of samples, characterized morphology -->
 
@@
 
-<!-- This part is non reproducible, selecting based on date, etc -->
+<!-- arising implicit hypothesis 1) different morph species, different genetic species 2) different elevation, different species (our methods not ideal for this, should have been randomized sample if we wanted to test for elevation) -->
 
-We selected 33 out of our 310 samples for *rbcL* and 18S Illumina amplicon sequencing.
-We chose these samples to represent the variation in date, elevation, geographic location, snow color, micro-habitat, and cell morphology.
+We Illumina sequenced *rbcL* and 18S amplicons of 33 samples.
+We selected this subset for sequencing to compare the species composition of blooms containing distinct morphologies, and to compare the genetic composition of morphologically similar blooms from different mountains, elevations, and dates.
 We freeze-dried these samples for up to 48 hours until until samples appeared completely desiccated.
 We mini-pestled between 5 to 20 mg of sample at room temperature to physically rupture cell walls before incubation in CTAB lysis buffer [@noauthor_ctab_2009].
-We extracted DNA using chloroform:isoamyl alcohol [@cubero_dna_1999], and spin columns (Qiagen, Hilden) (details available in Supplementary Protocol S1).
-As a negative control we processed a sterile distilled water sample alongside each batch.
+We extracted DNA in batches of 5-6 samples using chloroform:isoamyl alcohol [@cubero_dna_1999], and purified DNA using ethanol and spin columns (Qiagen, Hilden) (Supplementary Protocol S1).
+As a negative control against cross-contamination we processed a tube of sterile distilled water alongside each batch, and tested this for DNA with Qubit fluorometric quantification (Thermo Fisher, Waltham, MA).
 
-We designed custom primers to target a hypervariable region of snow algae *rbcL*.
-*rbcL* is more differentiated between snow algae than 18S (Supplementary Figure S2), and contains similar number of reference sequences on GenBank.
-Our *rbcL* primers target a 400 bp section of snow algae *rbcL* based off the consensus of 20 GenBank snow algae sequences (Supplementary Figure S2).
+In addition to 18S, we designed custom snow algae *rbcL* primers.
+We chose this marker because *rbcL* is more differentiated between snow algae than 18S (Supplementary Figure S2), and there is snow algae *rbcL* reference data on GenBank.
+Our primers were designed to target a 400 bp section of snow algae *rbcL*, based off the consensus of 20 GenBank snow algae sequences (Supplementary Figure S2).
+*Sanguina* sequences were not included because they were not available at the time.
 We designed our primers with the Eurofins primer design tool (https://www.eurofinsgenomics.eu/en/ecom/tools/pcr-primer-design/).
-*Sanguina* sequences were not included because they were not available at the time.
 To target 18S, we used the universal primers Euk1181 and Euk1624 targeting the V7-V8 hypervariable regions [@wang_optimal_2014].
@@
 
-We constructed our 18S and *rbcL* amplicon libraries using a standard two-step PCR approach [@meyer_illumina_2010].
-For detailed PCR conditions, see Supplementary Protocol S2.
+We constructed our 18S and *rbcL* amplicon libraries using a standard two-step PCR approach [@meyer_illumina_2010] (Supplementary Protocol S2).
 After each PCR, we purified using Agencourt AMPure XP kit (Beckman Coulter, Brea, CA).
-We standardized sample DNA concentration using Qubit fluorometric quantification (Thermo Fisher, Waltham, MA).
-The pooled library was loaded and run on a MiSeq V3 kit (Illumina, San Diego, CA).
+We standardized sample DNA concentration with Qubit, pooled our library, and sequenced this on a MiSeq using the V3 kit following manufacturer instructions (Illumina, San Diego, CA).
 
@@
 
-We demultiplexed reads using CUTADAPT [@martin_cutadapt_2011], and filtered and trimmed, removed errors, dereplicated, merged paired-end reads, and removed chimeras following the default pipeline of DADA2 [@callahan_dada2_2016].
-We assigned taxonomy to amplicon sequence variants (ASVs) using IDTaxa [@murali_idtaxa_2018], only assigning names to ASVs that were assigned with high confidence (threshold set to 50).
-We assigned taxonomy using custom snow algae reference databases for both 18S and *rbcL*, in addition to re-assigning 18S with SILVA to classify non-snow algae [@quast_silva_2013].
-Snow algae reference databases and the scripts to generate them are available on our github repository.
+We demultiplexed reads with CUTADAPT [@martin_cutadapt_2011], and filtered, trimmed, denoised, dereplicated, merged paired-end reads, and removed chimeras following the default pipeline of DADA2 [@callahan_dada2_2016].
+We assigned taxonomy to amplicon sequence variants (ASVs) using IDTaxa [@murali_idtaxa_2018] using a high confidence threshold parameter (t=50).
+Because snow algae are not well represented on curated databases such as SILVA, we made custom databases using GenBank snow algae sequences (scripts and databases available on github).
+To ensure that we were not overlooking non-snow algae taxa we also ran our 18S library against SILVA [@quast_silva_2013].
 
@@
 
-<!-- make sure this is on GH: -->
+<!-- ??? make sure this is on GH: -->
 
@@
 
-To generate OTUs we clustered ASVs using t-SNE [@maaten_visualizing_2008].
-We compared t-SNE results with the perplexity parameter ranging from 1 to 100, and found comparable results from 10 to 50.
-We categorized OTUs by applying DBSCAN [@hahsler_dbscan_2019] to our t-SNE.
-We compared *rbcL* sample similarity using UniFrac with default settings [@lozupone_unifrac_2005].
-To confirm that our clustering was not an artifact of t-SNE we created maximum likelihood phylogenetic trees of ASVs using IQTree [@nguyen_iq-tree:_2015] (Supplementary Figure S3).
+To assign OTUs, we clustered ASVs using t-SNE [@maaten_visualizing_2008] and DBSCAN [@hahsler_dbscan_2019].
+To ensure that our t-SNE was not an artifact of parameter selection, we ran t-SNE with different values of "perplexity", and found the same clusters were generated from 10 to 50.
+To cross-validate this unconventional OTU clustering method, we overlaid these OTU clusters on phylogenetic trees using IQTree [@nguyen_iq-tree:_2015] (Supplementary Figure S3).
+
+<!-- mention swarm? would need to show this dint work on tree as supp, no time -->
 
 # Results {-}
-<!-- should we report the diversity data quickly? or tangential to our question... -->
 
 <!-- add read depth, etc? -->
-<!-- make fig 5, ???, and make 18s otus -->
+<!-- i think the cell counts needs to go into the main text -->
 
-We observed morphologically distinct snow algae communities at different elevations (Figure 1; Supplementary Figure S4).
+We observed morphologically distinct snow algae blooms at different elevations (Figure 1; Supplementary Figure S4).
 Red snow was prevalent in areas of high solar exposure above treeline, predominantly containing cells similar in appearance to *Sanguina nivaloides* [@prochazkova_sanguina_2019].
-We occasionally observed green snow patches in alpine meadows receiving full sun, but more frequently observed green snow in well-shaded, forested sites.
-We found cells resembling *Chlainomonas rubra* [@novis_two_2008] at all elevations.
-Forested sites at low elevations often contained green or orange patches of snow, predominated by green or orange ovate cells with short spines resembling *Chloromonas cf.
+We found cells resembling *Chlainomonas rubra* [@novis_two_2008] across all elevations.
+Green snow patches were infrequent in alpine meadows receiving full sun, and more common in shady forested sites.
+Forested sites often contained green or orange patches of snow, many of these sites contained green or orange ovate cells with short spines resembling *Chloromonas cf.
 brevispina* [@matsuzaki_revisiting_2015].
 
-Both 18S and *rbcL* amplicon libraries were dominated by ASVs that were assigned to Chlorophyta.
-Our 18S library contained 67 ASVs that were assigned to Chlorophyta and 7 ASVs to Ochrophyta, while our *rbcL* library contained 603 ASVs assigned to Chlorophyta and 41 to Trebouxiophyceae.
-Negative controls did not contain Qubit-detectable amounts of DNA.
-Our *rbcL* library contained higher ASV and OTU diversity than 18S, t-SNE clustering generated 9 *rbcL* OTUs, while we detected just 3 visually distinct clusters among 18S ASVs (Supplementary Figure S5).
-For *rbcL*, the most frequent taxonomic assignments at genus level were *Chloromonas*, *Chlainomonas*, and *Sanguina*.
-Although most ASVs were not assigned to genus level, most ASVs clustered near one of these genera in NMDS ordination (Figure 2).
-Several OTUs were dominated by ASVs that were not assigned to genus level, including three OTUs closely related to *Chloromonas*, and one Chlamydomonadaceae OTU that was only assigned to the family level.
+Both 18S and *rbcL* amplicon libraries were dominated reads assigned to Chlorophyta (Figure 2).
+Our 18S library assigned with SILVA contained 50 Chlorophyceae, 11 Trebouxiophyceae, and 7 Chrysophyceae ASVs, while our *rbcL* library contained 603 Chlorophyta and 41 Trebouxiophyceae ASVs.
+Both ASV and OTU diversity was higher in *rbcL* than 18S: nine *rbcL* OTUs, compared to just three in 18S (Supplementary Figure S5).
+The most frequent *rbcL* genera were *Chloromonas*, *Chlainomonas*, and *Sanguina*.
+Although the majority of ASVs were not assigned to genus level, most ASVs were genetically similar to one of these three genera, with one exception.
+Sequences in the OTU E were not assigned to genus level; the ten best BLAST matches included six different genera within Chlamydomonadaceae (86-87% sequence match), two of which were *Chloromonas* snow algae (LC012735).
+Many ASVs were similar to *Chloromonas*, including two distinct OTU clusters D and F which did not match any known species within this genus.
 
-The composition of 18S and *rbcL* libraries varied with elevation (Figure 3; Figure 4).
-*Sanguina* was absent below 1500 m in both 18S and *rbcL*.
-High elevation samples contained an unannotated OTU of *Chloromonas* that was absent from low elevation sites (Figure 3, "Chloromonas C").
-Additionally, the three highest elevation sites uniquely contained ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1)---interestingly, all of these samples were taken from snow overlying glacier.
+18S and *rbcL* taxonomic composition varied with elevation.
+Low elevation samples clustered together in *rbcL* UniFrac [@lozupone_unifrac_2005], forming a cluster that was distinct from most high elevation samples (Figure 3).
+We found the highest diversity in samples that were collected latest in the season, although we did not find a statistically significant trend between Shannon diversity and date there was a weak correlation between Faith's PD and date (Pearson's *r*=0.36, p=0.04) (Supplementary Figure S6).
+In both 18S and *rbcL*, *Sanguina* predominated many sites above 1500 m, but was absent below this elevation (Figure 4).
+Additionally, high elevation samples contained one OTU of *Chloromonas* that was absent from low elevation sites (Figure 3, OTU F).
+The three highest elevation sites uniquely contained ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1)---interestingly, all of these samples were taken from snow overlying glacier.
 The two high alpine green snow samples were dominated by two different OTUs of *Chloromonas*, which were also both found in sites below treeline.
@@
 
-We observed snow algae blooms dominated by *Chloromonas krienitzii* transition from green to orange from May to June (Figure 5).
-At two sites we observed snow melt channels or runnels containing high concentrations of snow algae.
-In May and early June these sites contained green snow hidden 2 to 5 cm below the white snow surface, but upon subsequent visits the same location contained orange snow at the surface, whose morphologies resembled published images of the *Chloromonas krienitzii* lineage [@matsuzaki_revisiting_2015].
-We compared 18S and *rbcL* metabarcode profiles of green subsurface and orange surface snow, and found both contained high relative abundance of *Chloromonas krienitzii*.
-Although we did not detect *Chlainomonas* in cell counts, the surface samples contained nearly double the relative abundance of *Chlainomonas* from the subsurface samples.
-
-<!-- include faiths diversity in supps??? thinking no  -->
+A green to orange transition occurred from May to June at two sites dominated by *Chloromonas krienitzii*.
+On initial visits to a site in May we observed green snow hidden 2 to 5 cm below the snow surface (sample S2.2), but on subsequent visits in June (samples S4.6, S6.1, S9.2, S11.2) the same runnel was orange on the surface overlying the green snow.
+Microscopy of orange snow revealed orange spherical cells with short spines resembling *Chloromonas krienitzii* [@matsuzaki_revisiting_2015], while green snow contained motile green flagellates.
+*rbcL* comparison of green and orange snow showed both were dominated by reads assigned to *Chloromonas krienitzii*, with the orange snow containing higher abundance of *Chlainomonas* (Supplementary Figure S7).
 
@@
 Here we present multiple data sets demonstrating elevational patterns in alpine snow algae blooms.
-*Sanguina* was dominant in red snow above treeline, while green and orange blooms of *Chloromonas krienitzii* were dominant in mid-elevation snow runnels.
+*Sanguina* was dominant in red snow above treeline, while green and orange blooms of *Chloromonas krienitzii* were dominant in runnels at lower elevations.
 We found unexpected diversity within *Chloromonas* and *Chlainomonas* *rbcL* that we did not detect using 18S, including many potentially novel OTUs.
-In contrast to previous studies [@novis_two_2008;@prochazkova_ecophysiological_2018], our work suggests that *Chlainomonas* is widespread and abundant in alpine snow algae blooms.
+Previously only reported from red snow overlying lakes [@prochazkova_ecophysiological_2018], *Chlainomonas* was widespread and abundant in alpine snow algae blooms.
 
 Our results show bloom composition can be highly variable on the same mountain, even in nearby patches.
-In contrast, previous studies found snow algae assemblages were homogeneous across geographically distant sites [@lutz_biogeography_2016;@segawa_bipolar_2018].
-The aforementioned studies exclusively sampled high-latitude red snow, while our study included mid-latitude red, orange, and green snow from above and below treeline.
-Another recent study using 18S metabarcoding found different snow algae communities in red and green snow algae blooms in Antarctica, including yellow blooms dominated by Chrysophyceae [@soto_microbial_2020].
-Much of the diversity in our samples within *Chlainomonas* and *Chloromonas* was only detected with *rbcL* and not 18S, so the diversity within these genera may have been overlooked in previous metabarcoding studies.
+In contrast, previous studies found red snow algae assemblages were homogeneous across geographically distant polar sites [@lutz_biogeography_2016;@segawa_bipolar_2018].
+This discrepancy could be because our study included green and orange snow samples, or it could be that mid-latitude mountains have higher snow algae diversity than polar sites.
+We noted most snow algae blooms above treeline were dominated by *Sanguina*; by only collecting a single sample as representative of a region could have meant that the previous studies could have overlooked less common snow algae communities.
+We detected more OTUs with *rbcL* than 18S, so previous 18S metabarcoding could have overlooked snow algae diversity at lower taxonomic levels.
 
 Our *rbcL* data suggest *Chlainomonas* may be more widely distributed than previously thought.
-We could not differentiate between *Chlainomonas* and *Chloromonas* in our 18S data set, so this genus may have been overlooked in previous studies.
 *Chlainomonas* was previously thought to be restricted to waterlogged snow overlying mountain lakes [@novis_two_2008;@prochazkova_ecophysiological_2018], but we did not find any evidence that *Chlainomonas* was restricted in its habitat distribution.
-To the contrary, only one *Chlainomonas* dominant sample was located in waterlogged snow near the edge of a pool (sample S9); the other *Chlainomonas* dominant sites were not notably wetter than the surrounding snow, nor overlying frozen bodies of water.
-Interestingly, we found *Chlainomonas* was in higher relative abundance in our *rbcL* dataset than in cell counts (Figure 3)---possibly due to multiple chloroplasts per cell, which have been documented in this genus [@prochazkova_ecophysiological_2018].
+To the contrary, only one *Chlainomonas* dominant sample was located in waterlogged snow near the edge of a pool (sample S9), while the other *Chlainomonas* dominant sites were not notably wetter than the surrounding snow, nor were they located over frozen lakes.
+We could not differentiate between *Chlainomonas* and *Chloromonas* in our 18S data set, so this genus may have remained undetected in previous 18S metabarcoding.
+Interestingly, *Chlainomonas* was in higher relative abundance in our *rbcL* dataset than in cell counts (Figure 3).
+One possible explanation is high *rbcL* copy number: the large cells of *Chlainomonas rubra* have multiple parietal chloroplasts per cell [@prochazkova_ecophysiological_2018], and *rbcL* is located in the plastid genome.
 
 Our findings highlight the remaining unexplored diversity in the snow algae microbiome.
-Many *rbcL* OTUs did not closely match any GenBank nucleotide archive sequences (Figure 2).
-"Chlamydomonadaceae E" highest BLAST percent identity was only 87%, and the top 10 BLAST matches contained 6 different algae genera---including two *Chloromonas* snow algae from Japan.
-This OTU did not correlate with any categories in our 18S or cell count data.
+Several *rbcL* OTUs did not closely match any GenBank nucleotide archive sequences, such as the OTU "Chlamydomonadaceae E" (Figure 2).
+This OTU was distinct in our *rbcL* data, but it did not correlate with any categories in 18S or cell count data.
 One possible explanation is that it was lumped together with *Chloromonas* in 18S, and our crude quantification and categorization of cell counts did not detect this taxa.
-Other OTUs were also poorly matched to GenBank,  which could represent novel species of *Chloromonas* (Figure 2).
-The lack of clustering we observed in the remaining *Chloromonas* could be due to overlap in the levels of interspecific and intraspecific variation.
+OTUs D and F were related to *Chloromonas*, but contained few ASVs that were assigned to genus (Figure 2).
+Possibly these OTUs represent novel species of *Chloromonas*.
+However, there were no discernible clusters within the OTU "Other *Chloromonas*", perhaps due to overlap in the levels of interspecific and intraspecific variation.
 *rbcL* diversity is likely higher within *Chloromonas* than in other algae genera: unlike *Sanguina* [@prochazkova_sanguina_2019], most species of *Chloromonas* lack a pyrenoid [@nozaki_differences_2002], which houses high concentrations of cross-linked Rubisco.
-Likely related to its lack of pyrenoid, *Chloromonas* has many non-synonymous mutations in the region of *rbcL* the codes for binding Rubisco together [@nozaki_differences_2002].
-While *rbcL* is problematic in *Chloromonas* phylogeny [@nozaki_differences_2002], it nonetheless is highly differentiated and therefore an effective barcode for microalgae [@hall_assessment_2010].
-
-<!-- *Sanguina* was found in the same samples, but relative abundance was generally higher in 18S and cell counts (Figure 3).
-Our *rbcL* primers were designed based on *Chloromonas* and *Chlainomonas* snow algae sequences (*Sanguina* sequence data was not available at the time), so primer bias could have resulted in over-representation of *Chloromonas* relative to *Sanguina*.
--->
+Lacking pyrenoid, *Chloromonas* have many non-synonymous mutations in the region of *rbcL* the codes for binding Rubisco together [@nozaki_differences_2002].
+While *rbcL* is a poor indicator of *Chloromonas* phylogeny [@nozaki_differences_2002], it nonetheless is highly differentiated and therefore an effective barcode for microalgae [@hall_assessment_2010].
 
 Variation in bloom composition could be due to a wide range of habitat features.
-*Sanguina* and *Chloromonas C* were limited to sites above 1500 m in full sunlight, but many low elevation sites also received full sunlight and did not contain these OTUs (Figure 3).
+*Sanguina* and *Chloromonas* C were limited to sites above 1500 m in full sunlight, but many low elevation sites also received full sunlight and did not contain these OTUs (Figure 3).
 Light intensity, snow moisture, or snow chemistry could all plausibly influence community composition.
 Intriguingly, we only observed *Raphidonema* at high-elevation glacier sites.
-In Svalbard, *Raphidonema nivale* abundance increased on glacier surface snow following wind storms, and the authors suggest that this is a soil algae that grows sub-optimally on the snow surface following wind deposition [@stibal_growth_2005].
+In Svalbard, *Raphidonema nivale* abundance increased on glacier surface snow following wind storms, and the authors concluded that *Raphidonema* was a soil algae that opportunistically colonizes snow following wind deposition, but being sub-optimally adapted for snow such blooms are short lived [@stibal_growth_2005].
 Given the aerial dispersal capabilities of microalgae [@tesson_airborne_2016] and genetic overlap between distant snow algae populations [@segawa_bipolar_2018] it seems unlikely that geographic distance is a barrier to snow algae distribution.
@@
 
-Over multiple visits we observed motile green *Chloromonas krienitzii* cells below the snow surface develop into an orange bloom at the surface (Figure 5).
-Previous work has shown that this species can appear as distinct green and orange morphologies [@matsuzaki_revisiting_2015], but this paper is the first to document a bloom dominated by this species change color in the field.
+<!-- As i presented it above, looks like Chlainomonas is causing the bloom. Need to explain the nuance -->
+Previous work has shown that *Chloromonas krienitzii* undergoes distinct green and orange morphologies [@matsuzaki_revisiting_2015], but this paper is the first to document this transition in the field.
 The transition over weeks suggests that this process is mediated by seasonal changes.
-Secondary pigments likely protect snow algae from the damaging effects of intense solar irradiation at the snow surface [@bidigare_evidence_1993], which could be why we only observed green cells below the snow surface.
+Secondary pigments likely protect snow algae from the damaging effects of intense solar irradiation at the snow surface [@bidigare_evidence_1993], which could be why green cells were most concentrated a few centimeters below the snow surface.
+*Chlainomonas* was more abundant in surface samples as measured by *rbcL*, but we microscopy suggested that *Chlainomonas* was only present in low relative abundance.
 
@@
 Future studies will examine other taxa present in snow algae blooms, and their interactions in these globally important microbiomes.
-
-<!-- This needs to be beefed up.
-Some sort of statement about how this work impacts the field, more specifics about future directions.
-How this work could inform albedo work that you mention in your intro.
--->
 
@@
 
-The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
+The authors declare this research was conducted in the absence of any conflict of interest.
 
@@
 # Figure legends {-}
-
-<!-- formatting indentation??? -->
 
@@
 
-\setlength{\leftskip}{0pt}
 **Figure 2.** Ordination plots of *rbcL* ASV similarity.
@@
 Multidimensional scaling (MDS) plot.
-Taxonomy (assigned by IDTaxa) indicated by color, and size of point is proportional to cumulative relative abundance.
+Taxonomy (assigned by IDTaxa) indicated by colour, and size of point is proportional to cumulative relative abundance.
 Dotted ellipses show OTU clusters.
@@
 Each sample is arranged on the y axis in order of elevation, with each compositional bar plot representing the (left to right) cell morphology, 18S, and *rbcL* snow algae assemblages of the same sample.
-Colors below each bar plot indicate taxonomic categories.
+colours below each bar plot indicate taxonomic categories.
 Morphospecies were identified by light microscopy, using similarity to published photographs as a guide.
@@
 
-**Figure 4.** Non-metric multidimensional scaling (NMDS) showing *rbcL* UniFrac distances between samples. Each point represents a sample, which is labelled by sample ID and elevation (color).
-
-**Figure 5.** ??? Need to make the figure first
+**Figure 4.** Non-metric multidimensional scaling (NMDS) showing *rbcL* UniFrac distances between samples. Each point represents a sample, which is labelled by sample ID and elevation (colour).

### 2020-01-13 code/exploratory/community.Rmd
organize

@@
 # define theme for all heatmaps
-my_heat_theme <- theme(axis.text.x = element_text(angle = 45, hjust = 1),
-        axis.title.x=element_blank())
-theme_set(dark_theme_light())
+# my_heat_theme <- theme(axis.text.x = element_text(angle = 45, hjust = 1),
+#         axis.title.x=element_blank())
+# theme_set(dark_theme_light())
 ```
@@
 ```{r input, include=F}
-rbcl_rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv")) # not looking at the depth samples, controls..
-rbcl_tax <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
-rbcl_asv_seqs <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
-rbcl_otus <- read_csv(here("output/asv_otu_key.csv"))
-otu_colors <- read_csv(here("output/etc/otu_colors.csv"))
+rbcl_rel_abund <- read_csv(here::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv")) # not looking at the depth samples, controls..
+rbcl_tax <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
+rbcl_asv_seqs <- read_csv(here::here("data/fastas/asv_seq_data/rbcl_asv_sequence_key.csv"))
+rbcl_otus <- read_csv(here::here("data/otus/otu_rbcl_asvs.csv"))
 
-s18_rel_abund <- read_csv(here("data/03_filtered/18s_algae/18s_algae.filtered_cleaned_rel_abund.csv"))
-s18_algae_taxonomy <- read_csv(here("output/algae_18s_assignments_snow_algae_genbank.csv"))
+s18_rel_abund <- read_csv(here::here("data/rel_abund/3_filter/18s_algae/18s_algae_rel_abund.csv"))
+s18_algae_taxonomy <- read_csv(here::here("data/taxonomy/algae_18s_assignments_snow_algae_genbank.csv"))
 # s18_algae_select_taxonomy <- read_csv(here("output/algae_18s_assignments_w_select_snow_algae.csv")) #### this didnt work very well, need to make consensus for each species
 
-field <- read_csv(here("data/02_tidied/tidied_field.csv"))
-cellct <- read_csv(here("data/02_tidied/tidied.cell_count.csv"))
+field <- read_csv(here::here("data/field/tidy_field.csv"))
+cellct <- read_csv(here::here("data/cellct/tidied_cellct.csv"))
 
-rbcl_tree <- read.tree(here("output/tree/rbcl/all_asvs/rbcl_aligns_all.fasta.treefile")) ### this tree should only contain the asvs we found in the "main" sites, not looking at controls, etc
-s18_tree <- read.tree(here("output/tree/algae_18s/algae_18s_seqs.fasta.treefile"))
+rbcl_tree <- read.tree(here::here("data/tree/rbcl/all_asvs/rbcl_aligns_all.fasta.treefile")) ### this tree should only contain the asvs we found in the "main" sites, not looking at controls, etc
+s18_tree <- read.tree(here::here("data/tree/algae_18s/algae_18s_seqs.fasta.treefile"))
 ```

### 2020-01-13 manuscript/figs/code/fig_4_rel_abund_barplots.R

@@ -1,0 +1,94 @@
+# fig 4 relative abundance plots: rbcl, 18s, cellct
+
+library(tidyverse)
+library(here)
+library(GUniFrac)
+library(vegan)
+library(ggrepel)
+
+
+
+rbcl_rel_abund <- read_csv(here::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv")) # not looking at the depth samples, controls..
+rbcl_tax <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
+rbcl_asv_seqs <- read_csv(here::here("data/fastas/asv_seq_data/rbcl_asv_sequence_key.csv"))
+rbcl_otus <- read_csv(here::here("data/otus/otu_rbcl_asvs.csv"))
+s18_rel_abund <- read_csv(here::here("data/rel_abund/3_filter/18s_algae/18s_algae_rel_abund.csv"))
+s18_algae_taxonomy <- read_csv(here::here("data/taxonomy/algae_18s_assignments_snow_algae_genbank.csv"))
+# s18_algae_select_taxonomy <- read_csv(here("output/algae_18s_assignments_w_select_snow_algae.csv")) #### this didnt work very well, need to make consensus for each species
+field <- read_csv(here::here("data/field/tidy_field.csv"))
+cellct <- read_csv(here::here("data/cellct/tidied_cellct.csv"))
+rbcl_tree <- read.tree(here::here("data/tree/rbcl/all_asvs/rbcl_aligns_all.fasta.treefile")) ### this tree should only contain the asvs we found in the "main" sites, not looking at controls, etc
+s18_tree <- read.tree(here::here("data/tree/algae_18s/algae_18s_seqs.fasta.treefile"))
+
+
+# combine rbcl, 18s, and cellct into single tbl ----------
+
+# rbcl, group by otu
+rbcl_ra <- rbcl_rel_abund %>%
+  left_join(rbcl_otus, by="asv_id") %>% 
+  group_by(sample_id, otu_name) %>% 
+  summarise(rel_abund = sum(rel_abund)) %>%   
+  ungroup() %>% 
+  mutate(marker="rbcL") %>% # add variable to facilitate joining tbls
+  dplyr::rename(species = otu_name) %>% # rename variable to facilitate joining lates
+  select(sample_id, marker, species, rel_abund) %>% 
[…]

### 2020-01-13 manuscript/supp/code/depth_krienitzii_runnels.R

@@ -1,0 +1,105 @@
+# C krienitzii depth vs surface sample figure
+
+library(tidyverse)
+library(here)
+library(gridExtra)
+
+
+
+rbcl_ra_depth <- read_csv(here::here("data/rel_abund/3_filter/rbcl/depth_rbcl_rel_abund.csv")) # only the "depth samples", see code/wrangle/filter_rbcl_rel_abund.R for selection criteria
+rbcl_ra_main <- read_csv(here:::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv"))
+s18_all_rel_abund <- read_csv(here::here("data/rel_abund/3_filter/18s/depth_18s_rel_abund.csv"))
+# rbcl_tax <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
+s18_tax <- read_csv(here::here("data/taxonomy/tidied_silva_18s_assignments.csv"))
+rbcl_otus <- read_csv(here::here("data/otus/otu_rbcl_asvs.csv"))
+field <- read_csv(here::here("data/field/tidy_field.csv"))
+
+
+# combine tbls, choose samples of interest
+rbcl <- rbcl_ra_depth %>%
+  # since there is overlap between this and the main one, select to avoid conflict
+  filter(!(sample_id %in% c("hol18.21","sey18.66"))) %>% 
+  bind_rows(rbcl_ra_main) %>%
+  left_join(field, by="sample_id") %>%  # annoying redundant
+  left_join(rbcl_otus, by="asv_id") %>% 
+  filter(alias %in% c("S9.2", "S9.1", "S6.7", "S2.2")) %>% # choose just six samples
+  # excluded H4.3, H4.4 because dominated by Chlainomonas, I want to focus on Chloro krienitzii samples
+  mutate(depth = case_when(alias == "S9.2"~"surface", # add new category, whether its surface or subsurface
+                           alias == "S9.1"~"depth",
+                           alias == "S6.7"~"surface",
+                           alias == "S2.2"~"depth"))
+
+# rbcl %>% distinct(alias, depth, depth_cm)
+
+# select variables of interest, group by otus
+rbcl_1 <- rbcl %>% 
+  group_by(alias, otu_name, depth) %>% 
+  summarise(rel_abund= sum(rel_abund)) %>% 
+  ungroup() %>% 
+  select(alias, depth, otu_name, rel_abund) %>% 
[…]

### 2020-01-13 code/exploratory/depth_comparison.Rmd
organize

--- 2019-11-02 code/depth_comparison.Rmd
@@
 ```{r include=F}
-rbcl_rel_abund <- read_csv(here("data/03_filtered/rbcl/depth_rbcl_rel_abund.csv"))
+rbcl_rel_abund <- read_csv(here("data/"))
 s18_all_rel_abund <- read_csv(here("data/03_filtered/18s/depth_18s_rel_abund.csv"))
@@
 ```{r output, include=FALSE}
-ggsave(filename = here("manuscript/figs/4_depth_comparison/rbcl.pdf"), plot=depth)
+#ggsave(filename = here("manuscript/figs/4_depth_comparison/rbcl.pdf"), plot=depth)
 ```

### 2020-01-13 code/exploratory/green-orange.Rmd
organize

--- 2020-01-13 code/exploratory/depth_comparison.Rmd
@@
 ---
-title: "Depth sample comparison"
+title: "Depth comparison Seymour + Hollyburn green -> orange runnels"
 output: html_notebook
 ---
-
 ```{r}
@@
 ```{r include=F}
-rbcl_rel_abund <- read_csv(here("data/"))
-s18_all_rel_abund <- read_csv(here("data/03_filtered/18s/depth_18s_rel_abund.csv"))
-
-rbcl_tax <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
-s18_tax <- read_csv(here("data/02_tidied/tidied_silva_18s_assignments.csv"))
-
-rbcl_otus <- read_csv(here("output/asv_otu_key.csv"))
-
-field <- read_csv(here("data/02_tidied/tidied_field.csv"))
+rbcl_ra_depth <- read_csv(here::here("data/rel_abund/3_filter/rbcl/depth_rbcl_rel_abund.csv")) # only the "depth samples", see code/wrangle/filter_rbcl_rel_abund.R for selection criteria
+rbcl_ra_main <- read_csv(here:::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv"))
+s18_all_rel_abund <- read_csv(here::here("data/rel_abund/3_filter/18s/depth_18s_rel_abund.csv"))
+# rbcl_tax <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
+s18_tax <- read_csv(here::here("data/taxonomy/tidied_silva_18s_assignments.csv"))
+rbcl_otus <- read_csv(here::here("data/otus/otu_rbcl_asvs.csv"))
+field <- read_csv(here::here("data/field/tidy_field.csv"))
 ```
@@
 ```{r rbcl_wrangle, include=F}
-rbcl <- rbcl_rel_abund %>% 
-  left_join(rbcl_tax, by="asv_id") %>% 
+rbcl <- rbcl_ra_depth %>%
+  bind_rows(rbcl_ra_main) %>%
   left_join(field, by="sample_id") %>%  
-  left_join(rbcl_otus, by="asv_id")
-glimpse(rbcl)
-rbcl %>% distinct(sample_id, alias, depth_cm)
+  left_join(rbcl_otus, by="asv_id") %>% 
+  filter(alias %in% c("S9.2", "S9.1", "S6.7", "S2.2","H2.10","H3.4")) %>% # choose just six samples
+  # excluded H4.3, H4.4 because dominated by Chlainomonas, I want to focus on Chloro krienitzii samples
+  mutate(depth = case_when(alias == "S9.2"~"surface", # add new category, whether its surface or subsurface
+                           alias == "S9.1"~"depth",
+                           alias == "S6.7"~"surface",
+                           alias == "S2.2"~"depth",
+                           alias == "H2.10"~"surface",
+                           alias == "H3.4"~"depth"))
 
-depth_sample_info <- rbcl %>% 
-  distinct(sample_id, alias, depth_cm) %>% 
-  mutate(depth_chr = if_else(is.na(depth_cm), "surface", "sub-surface"),
-         top_sample = case_when(alias == "H2.10"~"H2.10*", # add a star, same site, but different date
-                                alias == "H3.4"~"H2.10*",
-                                alias == "H4.3"~"H4.4",
-                                alias == "S9.2"~"S9.1",
-                                TRUE ~ alias)) %>% 
-  select(sample_id, top_sample, depth_chr)
-depth_sample_info
+# rbcl %>% distinct(alias, depth, depth_cm)
 
+# select variables of interest, group by otus
 rbcl_1 <- rbcl %>% 
-  left_join(depth_sample_info, by="sample_id") %>% 
-  group_by(top_sample, depth_chr, otu_name) %>% 
+  group_by(alias, otu_name, depth) %>% 
   summarise(rel_abund= sum(rel_abund)) %>% 
-  ungroup()
+  ungroup() %>% 
+  select(alias, depth, otu_name, rel_abund)
 
-rbcl_1
 ```

### 2020-01-13 manuscript/supp/code/cellcount_stacked_barplot.R

@@ -34,4 +34,4 @@ cellct_plot <- ggplot(cellct, aes(x=alias %>% fct_reorder(elev_m),
           axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))  
 cellct_plot
 
-# ggsave(filename = here::here("manuscript/supp/final/fig_S4_cellct.tiff"), width = 300, height = 100, units="mm", dpi=300)
+# ggsave(filename = here::here("manuscript/supp/final/fig_S4_cellct.pdf"), width = 300, height = 100, units="mm")

### 2020-01-16 manuscript/etc/make_word_output.R
response to LQ comments

@@ -1,21 +1,0 @@
-library(here)
-library(bookdown)
-library(rmarkdown)
-
-setwd(here("manuscript/")) # bookdown wants project root to be where the .Rmd files are
-
-# input
-file <- "manuscript.Rmd"
-
-# output
-render(input=file,
-       output_format = "bookdown::word_document2",
-       output_file = paste0("casey1_", format(Sys.Date(), ("%Y-%m-%d")), ".docx"),
-       output_dir = "drafts_outbox",
-       output_options = list(reference_docx = "templates/word-template.docx"))
-
-
-setwd(here()) # return to the project root
-
-
-# can use bookdown::preview_chapter(), https://eddjberry.netlify.com/post/writing-your-thesis-with-bookdown/

### 2020-01-20 code/exploratory/misc/download_google_map_info.R
remove API data

@@ -7,7 +7,7 @@ library(ggmap)
 
 
 # get base map --------------
-key="AIzaSyBJsmyAKTkTg9LT-HdGCCdZdbE2Hjwm-L8"
+key=
 register_google(key = key) 
 
 algae_map <- get_googlemap(c(-123.004722, 49.850556), zoom = 7, maptype = "satellite")

### 2020-01-20 manuscript/supp/code/map/download_google_map_data.R

@@ -7,7 +7,7 @@ library(ggmap)
 library(png)
 
 # download data from googleway: do only once to avoid API overcharges
-key <- "AIzaSyBJsmyAKTkTg9LT-HdGCCdZdbE2Hjwm-L8"
+key <-
 register_google(key = key)
 
 # get map

### 2020-01-22 manuscript/supp/code/18s_asv_clust.Rmd

@@
 library(ggtree)
+library(gridExtra)
+library(ggpubr)
+library(cowplot)
 ```
@@
 ```{r, include=F}
-BrowseSeqs(aligns, highlight = T)
+# BrowseSeqs(aligns, highlight = T)
 
@@
 
-mds <- ggplot(mds_coords_w_taxonomy, aes(x=V1, y=V2))+
+mds_1 <- ggplot(mds_coords_w_taxonomy, aes(x=V1, y=V2))+
   geom_point(aes(color=best_assignment, size=rel_abund), alpha=0.7, na.rm=TRUE) +
@@
              color = "grey", alpha = 0.8) +
-  labs(color="Taxonomic assignment")
-mds
+  labs(color="Taxonomic assignment", tag="A")
+
+legend <- mds_1 %>% 
+  cowplot::get_legend() %>% 
+  ggpubr::as_ggplot()
+
+mds <- mds_1 +
+  theme(legend.position = "none")
+
 ```
@@
          colour = guide_legend(override.aes = list(size=5, alpha=1))) +
-  labs(color="Taxonomic assignment")
-
-tsne
+  labs(color="Taxonomic assignment", tag="B") +
+  theme(legend.position = "none")
 
 ```
-
-Perplexity = 20.
-
-\pagebreak
-
-# Maximum likelihood tree (IQTree)
 
@@
   geom_tiplab(data = annotate, aes(label = best_assignment), size=2)+
-  guides(size="none")
+  guides(size="none") +
+  labs(tag="C") +
+  theme(legend.position = "none")
+```
 
-tree_out
+```{r, fig.width=8, fig.height=10}
+lay <- rbind(c(1,1,4),
+             c(2,2,4),
+             c(3,3,4),
+             c(3,3,4),
+             c(3,3,4))
+gridExtra::grid.arrange(mds, tsne, tree_out, legend, layout_matrix=lay)
+
 ```

### 2020-01-22 manuscript/supp/code/diversity.Rmd

@@
 ```{r, include=F}
-rel_abund <- read_csv(here::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv")) # not looking at the depth samples, controls, etc
+rbcl_rel_abund <- read_csv(here::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv")) # not looking at the depth samples, controls, etc
 tax <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
@@
 # get relative abundance of each otu, in wide format, 
-otu_rel_abund <- rel_abund %>% 
+otu_rel_abund <- rbcl_rel_abund %>% 
   left_join(otus, by="asv_id") %>% 
@@
 # format
-ra_wide <- rel_abund %>% 
+ra_wide <- rbcl_rel_abund %>% 
   select(sample_id, asv_id, rel_abund) %>% 

### 2020-01-22 manuscript/figs/code/fig_3_unifrac_nmds.R
removed ms

@@ -1,74 +1,0 @@
-# make UniFrac NMDS
-
-library(tidyverse)
-library(here)
-library(GUniFrac)
-library(vegan)
-library(ggrepel)
-library(picante)
-library(gridExtra)
-
-
-rel_abund <- read_csv(here::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv")) # not looking at the depth samples, controls, etc
-tax <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
-field <- read_csv(here::here("data/field/tidy_field.csv"))
-asv_seqs <- read_csv(here::here("data/fastas/asv_seq_data/rbcl_asv_sequence_key.csv"))
-otus <- read_csv(here::here("data/otus/otu_rbcl_asvs.csv"))
-rbcl.tree <- read.tree(here::here("data/tree/rbcl/asv613.fasta.treefile")) ### this tree should only contain the asvs we found in the "main" sites, not looking at controls, etc
-
-
-ra_wide <- rel_abund %>% 
-  select(sample_id, asv_id, rel_abund) %>% 
-  arrange(asv_id) %>% 
-  pivot_wider(names_from = asv_id, values_from = rel_abund, values_fill = list(rel_abund = 0)) %>% 
-  column_to_rownames("sample_id")
-
-# root tree with Trebouxiophyceae
-root <- tax %>% 
-  filter(class=="Trebouxiophyceae") %>% 
-  head(1) %>% 
-  pull(asv_id)
-rbcl.tree <- rbcl.tree %>% root(outgroup = root, resolve.root=T)
-
-
-# check for differences between asvs in tree and rel abund table (all names must appear in tree)
-# asvs_rel_abund <- colnames(ra_wide)
-# asvs_tree <- rbcl.tree$tip.label %>% sort()
-# 
-# length(asvs_rel_abund)
-# length(asvs_tree)
-# 
-# setdiff(asvs_rel_abund, asvs_tree) # output elts unique to asvs_rel_abund
-
-# run weighted UniFrac
-set.seed(123)
-unifracs <- ra_wide %>% 
-  GUniFrac::GUniFrac(rbcl.tree, alpha=1) # ignore error
-
-unifrac <- unifracs$unifracs[, , "d_1"]
-
-
-# run NMDS ordination on results 
-set.seed(123)
-nmds <- metaMDS(unifrac, autotransform = F)
-
-
-# plot
[…]

### 2020-01-22 manuscript/figs/code/fig_4_rel_abund_barplots.R

@@ -1,94 +1,0 @@
-# fig 4 relative abundance plots: rbcl, 18s, cellct
-
-library(tidyverse)
-library(here)
-library(GUniFrac)
-library(vegan)
-library(ggrepel)
-
-
-
-rbcl_rel_abund <- read_csv(here::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv")) # not looking at the depth samples, controls..
-rbcl_tax <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
-rbcl_asv_seqs <- read_csv(here::here("data/fastas/asv_seq_data/rbcl_asv_sequence_key.csv"))
-rbcl_otus <- read_csv(here::here("data/otus/otu_rbcl_asvs.csv"))
-s18_rel_abund <- read_csv(here::here("data/rel_abund/3_filter/18s_algae/18s_algae_rel_abund.csv"))
-s18_algae_taxonomy <- read_csv(here::here("data/taxonomy/algae_18s_assignments_snow_algae_genbank.csv"))
-# s18_algae_select_taxonomy <- read_csv(here("output/algae_18s_assignments_w_select_snow_algae.csv")) #### this didnt work very well, need to make consensus for each species
-field <- read_csv(here::here("data/field/tidy_field.csv"))
-cellct <- read_csv(here::here("data/cellct/tidied_cellct.csv"))
-rbcl_tree <- read.tree(here::here("data/tree/rbcl/all_asvs/rbcl_aligns_all.fasta.treefile")) ### this tree should only contain the asvs we found in the "main" sites, not looking at controls, etc
-s18_tree <- read.tree(here::here("data/tree/algae_18s/algae_18s_seqs.fasta.treefile"))
-
-
-# combine rbcl, 18s, and cellct into single tbl ----------
-
-# rbcl, group by otu
-rbcl_ra <- rbcl_rel_abund %>%
-  left_join(rbcl_otus, by="asv_id") %>% 
-  group_by(sample_id, otu_name) %>% 
-  summarise(rel_abund = sum(rel_abund)) %>%   
-  ungroup() %>% 
-  mutate(marker="rbcL") %>% # add variable to facilitate joining tbls
-  dplyr::rename(species = otu_name) %>% # rename variable to facilitate joining lates
-  select(sample_id, marker, species, rel_abund) %>% 
[…]

### 2020-01-22 manuscript/figs/code/incomplete_fig_2_rbcl_asvs.R

@@ -1,22 +1,0 @@
-# Fig 2 rbcL ASVs
-
-library(tidyverse)
-library(here)
-library(DECIPHER)
-library(ggforce)
-library(Rtsne)
-library(dbscan)
-library(ggrepel)
-
-
-rel_abund <- read_csv(here::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv"))
-taxonomy <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
-aligns <- readDNAStringSet(here::here("data/fastas/alignments/rbcl_aligns_in_main.fasta"))
-seq <- read_csv(here::here("data/fastas/asv_seq_data/asv_seq_key_18s.csv"))
-                           
-
-# make distance matrix
-BrowseSeqs(aligns, highlight = T)
-
-dists <-  aligns %>% 
-  DECIPHER::DistanceMatrix(type="matrix", correction = "JC")

### 2020-01-22 manuscript/supp/code/cellcount_stacked_barplot.R

@@ -1,37 +1,0 @@
-# cell count pie charts
-
-library(tidyverse)
-library(here)
-
-
-cellct <- read_csv(here::here("data/cellct/tidied_cellct.csv"))
-field <- read_csv(here::here("data/field/tidy_field.csv"))
-
-cellct <- cellct %>% 
-  left_join(field, by="sample_id")
-
-
-colors <- c("Chlainomonas rubra"="#ac71b5", #purple
-            "Sanguina nivaloides"="#ff534a", #red
-            # Chloromonas: pastel greens / blues / yellows
-            "Chloromonas krienitzii"="#ffa54c",
-            "Chloromonas cf. brevispina"="#80c786", "Chloromonas cf. nivalis"="#4aaf86", 
-            #                 lt green                            teal
-            "Other Chloromonas"="#4aaf53", "Chloromonas"="#4aaf53", 
-            # greyish light green
-            "Green cell"="#4aaf53", "Other"="grey60") #greys
-# oops, ordered them backwards, too lazy to rewrite
-colors <- colors %>% rev()
-
-
-cellct_plot <- ggplot(cellct, aes(x=alias %>% fct_reorder(elev_m), 
-                     y=percent_of_sample)) +
-    geom_bar(aes(fill=morpho_sp %>% fct_relevel(names(colors))), position="stack", stat="identity") +
-    theme_classic() +
-    labs(x="Sample ID (in order of increasing elevation)", y="Relative abundance", fill="Cell morphology")+
-    scale_fill_manual(values = colors) +
-    theme(text = element_text(size=8),
-          axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))  
-cellct_plot
-
-# ggsave(filename = here::here("manuscript/supp/final/fig_S4_cellct.pdf"), width = 300, height = 100, units="mm")

### 2020-01-22 manuscript/supp/code/depth_krienitzii_runnels.R

@@ -1,105 +1,0 @@
-# C krienitzii depth vs surface sample figure
-
-library(tidyverse)
-library(here)
-library(gridExtra)
-
-
-
-rbcl_ra_depth <- read_csv(here::here("data/rel_abund/3_filter/rbcl/depth_rbcl_rel_abund.csv")) # only the "depth samples", see code/wrangle/filter_rbcl_rel_abund.R for selection criteria
-rbcl_ra_main <- read_csv(here:::here("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv"))
-s18_all_rel_abund <- read_csv(here::here("data/rel_abund/3_filter/18s/depth_18s_rel_abund.csv"))
-# rbcl_tax <- read_csv(here::here("data/taxonomy/tidied.rbcl_taxonomy.csv"))
-s18_tax <- read_csv(here::here("data/taxonomy/tidied_silva_18s_assignments.csv"))
-rbcl_otus <- read_csv(here::here("data/otus/otu_rbcl_asvs.csv"))
-field <- read_csv(here::here("data/field/tidy_field.csv"))
-
-
-# combine tbls, choose samples of interest
-rbcl <- rbcl_ra_depth %>%
-  # since there is overlap between this and the main one, select to avoid conflict
-  filter(!(sample_id %in% c("hol18.21","sey18.66"))) %>% 
-  bind_rows(rbcl_ra_main) %>%
-  left_join(field, by="sample_id") %>%  # annoying redundant
-  left_join(rbcl_otus, by="asv_id") %>% 
-  filter(alias %in% c("S9.2", "S9.1", "S6.7", "S2.2")) %>% # choose just six samples
-  # excluded H4.3, H4.4 because dominated by Chlainomonas, I want to focus on Chloro krienitzii samples
-  mutate(depth = case_when(alias == "S9.2"~"surface", # add new category, whether its surface or subsurface
-                           alias == "S9.1"~"depth",
-                           alias == "S6.7"~"surface",
-                           alias == "S2.2"~"depth"))
-
-# rbcl %>% distinct(alias, depth, depth_cm)
-
-# select variables of interest, group by otus
-rbcl_1 <- rbcl %>% 
-  group_by(alias, otu_name, depth) %>% 
-  summarise(rel_abund= sum(rel_abund)) %>% 
-  ungroup() %>% 
-  select(alias, depth, otu_name, rel_abund) %>% 
[…]

### 2020-01-22 manuscript/supp/code/map/download_google_map_data.R

@@ -1,45 +1,0 @@
-# get map data
-# 1-10
-
-library(tidyverse)
-library(googleway)
-library(ggmap)
-library(png)
-
-# download data from googleway: do only once to avoid API overcharges
-key <- 
-register_google(key = key)
-
-# get map
-googlemap <- get_googlemap(c(-123.004722, 49.850556), zoom = 7, maptype = "satellite")
-my_map <- ggmap(googlemap)
-my_map
-# get city coords -----------------------
-cities <- c("Vancouver, BC, Canada", "Squamish, BC, Canada")
-city_coords <- cities %>% 
-  map(~googleway::google_geocode(.x, key=key))
-glimpse(google_output) # ugly lists
-
-# functions to reach into the list and extract wanted elements
-city_name <- function(x) x$results$address_components[[1]][1,2]
-lat <- function(x) x$results$geometry$location[1,1]
-lon <- function(x) x$results$geometry$location[1,2]
-
-# make tbl with city and coords
-city_coords <- google_output %>% {
-  tibble(
-    name = map_chr(., .f=city_name),
-    lat = map_dbl(., .f=lat),
-    lon = map_dbl(., .f=lon)
-  )
-}
-
-city_coords # a pretty tbl
-
-# save data
-ggsave(filename = here::here("data/map_downloads/googlemap.png"), plot=my_map, 
-       width = 225, height = 225, units="mm", dpi = 300) # didnt end up using this
-save(googlemap, file=here::here("data/map_downloads/googlemap.rda")) # ended up using this... next time, dont bother with clunky API, download the map data directly as png ("do one thing and do it well")
-
-write_csv(city_coords, here::here("data/map_downloads/map_data_cities.csv"))
-

### 2020-01-22 manuscript/supp/code/map/map.R

@@ -1,60 +1,0 @@
-# draw sample map
-# 1-10
-
-library(tidyverse)
-library(lubridate)
-library(ggmap)
-library(ggrepel)
-# library(kableExtra)
-# library(magick)
-# library(formattable)
-# library(ggdark)
-
-# it wouldve been easier to just download the map as png, and the city data instead of using the clumsy API
-# next time I will do this, but for now using the Rda map data
-
-field <- read_csv(here::here("data/02_tidied/tidied_field.csv"))
-cities <- read_csv(here::here("output/map/city_coords.csv"))
-load(here::here("output/map/raw_google_map.rda"))
-
-# only plot vancouver
-cities <- cities %>% 
-  filter(name == "Vancouver")
-
-# select field data to plot (only mtns, not individual sites)
-field_mtns <- field %>% 
-  mutate(mtn_letter = alias %>% str_sub(1,1))
-
-field_sites <- field_mtns %>% 
-  distinct(mtn_letter) 
-
-field_site_coords <- field_mtns %>% 
-  group_by(mtn_letter) %>% 
-  drop_na(lat, lon) %>% 
-  summarise(n=n(), 
-            mean_elev = mean(elev_m), 
-            first_visit=min(date) %>% format("%b %d"), 
-            last_visit= max(date) %>% format("%b %d"), 
-            n_visits= length(unique(date)),
-            lat=mean(lat),
-            lon=mean(lon),
-            date=min(date)) %>% 
-  arrange(date) %>% 
-  select(-date)
-
-my_map <- ggmap(algae_map, extent = "device") +
-  geom_point(data = field_site_coords, aes(x=lon, y=lat),
-             size = 1, alpha=1, 
-             color = "red") +
-  # geom_text(data=field_site_coords, aes(label = mtn_letter)) # dosen't map site A, too far south for map
-  geom_label_repel(data=field_site_coords, aes(label = mtn_letter),
-    size = 4, 
-    box.padding = 0.2, point.padding = 0.3,
-    segment.color = 'grey50') +
-  geom_point(data = cities, shape = 15, color="white") +
-  geom_text(data = cities, aes(label = name), hjust = 1.1, vjust = 2, size = color="white")
-
-my_map
-
-# save
-ggsave(filename = here::here("manuscript/supp/map_1-10.tiff"), plot=my_map, width = 200, height = 200, units="mm",dpi=300)

### 2020-01-22 manuscript/supp/code/tree.R

@@ -1,88 +1,0 @@
-# Make rbcl ASV tree
-# 2020-01-10
-
-library(tidyverse)
-library(here)
-library(ape)
-library(ggtree)
-
-
-# ML phylogenetic tree (IQTree) of rbcL ASVs that appear in filtered relative abundance dataset (613 tips)
-tree <- read.tree(file=here::here("output/tree/rbcl/asv613.fasta.treefile")) 
-# Taxonomy tbl of each ASV (652; includes ASVs filtered out in samples not used in final analysis)
-taxonomy <- read_csv(here::here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
-# OTU of each ASV
-otus <- read_csv(here::here("output/asv_otu_key.csv"))
-rel_abund <- read_csv("data/rel_abund/3_filter/rbcl/rbcl_rel_abund.csv")
-
-
-# Misc edits of tree annotation data
-
-# rename one of the OTUs (should take care of this upstream)
-otus <- otus %>% 
-  mutate(otu_name= otu_name %>% str_remove("muramotoi "))
-
-taxonomy <- taxonomy %>% 
-  mutate(assignment = assignment %>% str_remove("root"))
-
-
-
-
-# Plot tree ---------
-
-# re-root tree
-root <- taxonomy %>% 
-  filter(class=="Trebouxiophyceae") %>% 
-  head(1) %>% 
-  pull(asv_id)
-
-rbcl.tree <- rbcl.tree %>% root(outgroup = root, resolve.root=T)
-
-rbcl_ggtree <- ggtree(rbcl.tree) # ignore warning message 
-
-# format bayesian/bootstrap confidence for each node
-boots <- rbcl_ggtree$data %>% 
-  filter(isTip == FALSE) %>% 
-  separate(label, c("boot", "bayes"), sep = "/") %>% #split label into boots and bayes vals
-  filter(boot > 80) %>%
-  mutate(boot = round(as.numeric(boot), digits = 0),
-         insignificant_edge = if_else(boot > 80, "black", "red"))# round boot to nearest integer
-
-# get rel abund for each asv
-rel_abund_asv <- rel_abund %>% 
-  group_by(asv_id) %>% 
-  summarise(rel_abund=sum(rel_abund))
-
-# format taxonomy, otu for each node
-assignment <- taxonomy %>% 
-  select(asv_id, assignment)
-annotation <- rbcl_ggtree$data %>% 
[…]
